### 2024-03-05 aaai2024_neural_circuits/neural_circuits_rebuttal.tex
added neural circuit paper

%File: anonymous-submission-latex-2024.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai24}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
%
% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
\usepackage{algorithm}
\usepackage{algorithmic}

%
% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
\usepackage{newfloat}
\usepackage{listings}
\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
\lstset{%
	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
	aboveskip=0pt,belowskip=0pt,%
	showstringspaces=false,tabsize=2,breaklines=true}
\floatstyle{ruled}
\newfloat{listing}{tb}{lst}{}
\floatname{listing}{Listing}
%
% Keep the \pdfinfo as shown here. There's no need
% for you to add the /Title and /Author tags.
\pdfinfo{
/TemplateVersion (2024.1)
}

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.

% The file aaai24.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%

% Title

\title{}

%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \author {
%     Pedro Zuidberg Dos Martires
% }
% \affiliations{
%     Örebro University
% }

\author{}
\affiliations{}

\usepackage{mathtools} % amsmath with fixes and additions
% \usepackage{siunitx} % for proper typesetting of numbers and units
\usepackage{booktabs} % commands to create good-looking tables
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,calligraphy,calc,hobby,intersections,through}
\usetikzlibrary{spn}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fontsm
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.

\usepackage{multirow}
\usepackage{courier}
\usepackage{listings, lstautogobble,amsfonts}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mdframed}
\usepackage{mathtools}
\usepackage{xspace}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{multicol}
\usepackage{thmtools}
\usepackage{bm}
\usepackage{thm-restate}
\usepackage{todonotes}
\usepackage[inline]{enumitem}
\usepackage{soul}
\usepackage{physics}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{cancel}

\pgfplotsset{compat=1.18}

\input{macros}

%Commands definitions
\newcommand{\setbackgroundcolour}{\pagecolor[rgb]{0.19,0.19,0.19}}  
\newcommand{\settextcolour}{\color[rgb]{0.77,0.77,0.77}}    
\newcommand{\invertbackgroundtext}{\setbackgroundcolour\settextcolour}

%Command execution. 
%If this line is commented, then the appearance remains as usual.
% \invertbackgroundtext

\newcommand{\spacing}{\vspace{0.1cm}}
\newcommand{\question}[1]{\noindent{\bf\color{teal}{#1}}}

\newcommand{\answer}[1]{\noindent{\normalfont\color{black}{#1}}}

\begin{document}

% \maketitle
\noindent
We would like to thank the reviewers for their considerate and thoughtful comments. In order to adequately address these comments, we intend to purchase an extra page.

\vspace{0.1cm}
\noindent {\bf Reviewer 1}

\noindent We address the comment made by Reviewer 1 concerning marginalization in our responses to Reviewers 2 and 3.
\spacing
\question{
How do PNCs empirically compare to state-of-the-art (intractable) density estimators?
}

\answer{
State of the art neural density estimators usually deploy a plethora of tricks, such as the use of \textit{discretized logistic mixtures}~\Citep{salimans2017pixelcnn} to achieve a high performance. While we plan in future work to explore these ideas for PNCs, the main focus of the experiments was to show that giving up to a certain degree on tractability leads to more expressive models. Indeed, our experiments confirm this hypothesis, especially when comparing PNCs. PQCs, and state-of-the-art circuits.

Nevertheless, we also compare to less tractable models. In particular normalizing flows (IDF) and hierarchical latent variable models (BitSwap) (both using neural networks). In this regard we believe we make a fair effort in terms of comparisons. Especially when considering the parameter count of the models: The parameter count of PNCs is half the one of SHCLT and only a fraction of neural density estimators such as PixelCNN ($\approx56M$ parameters).

% In future work we plan on scaling up our approach (usually not trivial for PCs as parameter learning is tricky in massively overparametrized circuits). We would then compare to larger models, \eg PixelCNN++ ($\approx56M$ parameters). We would also deploy common tricks from the density estimation literature such as  in the leaves \Citep{salimans2017pixelcnn}.
}

\vspace{0.1cm}
\noindent {\bf Reviewer 2}

\noindent
\question{Relationship between CPCs and PNCs.}

\answer{
The advantage of introducing CPCs lies with the neat interpretation as deep mixture of Bayesian networks and the fact that we can, for instance, proof their validity (Proposition 3.9) using arguments similar to the case of PCs.

We agree, however, with the reviewer that simply stating the functional form of sum units for PNCs (3.10 Definition) obscures to a certain degree their relationship to CPCs. For the camera-ready version we will include an extended version of the following discussion, which gives us PNCs starting from CPCs.

Consider a sum unit of a CPC:
\begin{talign}
	\sum_{j\in\inputs(k)} \weight_{kj} \circuit_j(\xvars_n  {\mid}  \xvars_{\parents(n)})
\end{talign}
This can be rearranged into:
\begin{talign}
	\sum_{j\in\inputs(k)}
		\weight_{kj}
		\frac{
			\circuit_j(  \xvars_{\parents(n)} \mid \xvars_n  )
		}{
			\circuit_j(  \xvars_{\parents(n)  })
		}
	\circuit_j(  \xvars_n)
\end{talign}
We approximate the ratio with a neural network as follows:
\begin{talign}
	\nweight_{kj}(\xvars_{\ancestors(n)})\approx 		\weight_{kj}
	\frac{
		\circuit_j(  \xvars_{\parents(n)} \mid \xvars_n  )
	}{
		\circuit_j(  \xvars_{\parents(n)  })
	}.
\end{talign}
This gives us the expression of Definition 3.10. Constraining the $ \nweight_{kj} ( \xvars_{\ancestors(n)})$ to sum to one together with necessitating $\circuit_j (\xvars_n)$ being a valid circuit then gives us the validity of the PNC. This lets us conclude that PNCs are neural approximations of CPCs. We will make this explicit in the final version of the paper.
Note that it is this approximation that makes PNCs tractable: PNCs only need a single circuits for each $j$ while CPCs need a circuit for every $j$ and every instantiation of $\xvars_{\parents(n)}$.
}

\spacing
\question{
The marginal inference algorithm was a bit unclear. Does it require m number of passes through the circuit, or can it be done in a single pass? [...]
}

The marginalization method we describe in Proposition 3.13 walks through a circuit in a depth-first fashion. This requires only visiting each node at most once (assuming proper caching). This can indeed be translated into a breadth-first circuit traversal, which is then easily parallelized for layered PNCs. This would then only require a single pass through the layered computation graph.
As pointed out by the reviewer this would indeed boil down to replacing distributions at the leaves with $1$ and evaluating the circuit.
We will add a discussion on this to the paper as well as provide pseudocode.

% This is what we refer to as "the delicate point" in the paper (after the proof of Proposition 3.13).

\spacing
\question{
Also, I think PNCs are tractable for more general ordered conditionals than shown in Corollary 3.14. [...].
}

Yes, indeed thank you for the suggestion. We will add it to the paper. Showing this is quite straightforward by combining Proposition 3.13 (marginalizing $X_m$) and Corollary 3.14.

\spacing
\question{Minor Questions / Comments}

\answer{
Due to space constraints, we are, unfortunately, not able to comment on them here. We will address them in the camera-ready version of the manuscript.
}

% \begin{itemize}

% \item Definition 3.2: $pa(n)$ where $X_n$ is a set of variables was not previously defined. Is it just the union of parents of variables in $X_n$ (minus $X_n$ itself)?
% \item Eq (5): missing parenthesis for $w_{kj}$. Also, should the sum unit be $p_k(x_n | x_{an(n)})$ instead because the weights depend on $x_{an(n)}$, and should $p_j$ also be conditioned the same way for the recursive definition?
% \item Typo: Theorem 3.9 sum of $w_m$ -> $w_{nm}$
% \item Typo: Figure 5 caption “while on the left” -> “while on the right”
% \item Section 6.3: I did not follow “two components in the leaf units instead of 256”
% \item Are partition trees related to vtrees? (Kisa et al. "Probabilistic sentential decision diagrams." 2014)
% \item Definition 3.10 seems to assume Boolean variables. Can this be applied to continuous variables?

% \end{itemize}

\vspace{0.1cm}
\noindent {\bf Reviewer 3}

\noindent
Similarly to Reviewer 1, Reviewer 3 commented on tractability of marginals (in practice). In the final version of the manuscript we will include a more detailed discussion. Here we already answer some of the questions.

	\question{For the PNC structures used for 28x28 images, have the following questions:
	}

	\question{
	A. What is the missingness pattern that is allowed in the images so we still get tractable queries? Is there a intuitive way to see which missing patterns are allowed for the pixels?
	}
	\answer{Yes, there is an intuitive pattern. Consider the partition tree in Figure 3. We can then marginalize out the variables the right branch of the tree. For images this is similar but with a second dimension.}

	\question{
	 B. Is the missingness pattern allowed for MARs the same for PNCs learned on the different datasets? Or is it data dependent (assume still images of same size).
	}
	\answer{The missingness pattern is the same for different datasets (assuming the same PNC structure is used.)}

	\question{
	 C. In practice how often we get actual tractability for MARs for images.
	}
	\answer{In practice one could, for instance, marginalize the right side of an image and have a model for the left side. This means that one would need to know beforehand the queries posed to the model. An interesting question is whether this can be mitigated using approximate inference techniques such as sampling or adapting a form of loopy belief propagation.
	Alternatively, one could focus on tasks where only ordered marginalization (PNC) is needed instead of any-order marginalization (PC). As we suggest in our future work section, arithmetic coding seems to be a good candidate and there might be other such tasks of interest. 
	}

\bibliography{references}

\end{document}

### 2024-03-05 aaai2024_neural_circuits/neural_circuits.tex
added neural circuit paper

%File: anonymous-submission-latex-2024.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai24}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
%
% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
\usepackage{algorithm}
\usepackage{algpseudocode}

%
% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
\usepackage{newfloat}
\usepackage{listings}
\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
\lstset{%
	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
	aboveskip=0pt,belowskip=0pt,%
	showstringspaces=false,tabsize=2,breaklines=true}
\floatstyle{ruled}
\newfloat{listing}{tb}{lst}{}
\floatname{listing}{Listing}
%
% Keep the \pdfinfo as shown here. There's no need
% for you to add the /Title and /Author tags.
\pdfinfo{
/TemplateVersion (2024.1)
}

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.

% The file aaai24.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%

% Title

\title{Probabilistic Neural Circuits}

%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
\author {
    Pedro Zuidberg Dos Martires
}
\affiliations{
    Centre for Applied Autonomous Sensor Systems (AASS), Örebro University, Sweden
}

% \author{}
% \affiliations{}

\usepackage{mathtools} % amsmath with fixes and additions
% \usepackage{siunitx} % for proper typesetting of numbers and units
\usepackage{booktabs} % commands to create good-looking tables
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,calligraphy,calc,hobby,intersections,through}
\usetikzlibrary{spn}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fontsm
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.

\usepackage{multirow}
\usepackage{courier}
\usepackage{listings, lstautogobble,amsfonts}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mdframed}
\usepackage{mathtools}
\usepackage{xspace}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{multicol}
\usepackage{thmtools}
\usepackage{bm}
\usepackage{thm-restate}
\usepackage{todonotes}
\usepackage[inline]{enumitem}
\usepackage{soul}
\usepackage{physics}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{cancel}

\pgfplotsset{compat=1.18}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\cf}{cf.\xspace}
\newcommand{\eg}{e.g.\xspace}
\newcommand{\ie}{i.e.\xspace}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newtheorem{theorem}{Theorem}[section]
% \theoremstyle{theorem}
\newtheorem{definition}[theorem]{Definition}
% \theoremstyle{theorem}
\newtheorem{lemma}[theorem]{Lemma}
% \theoremstyle{theorem}
\newtheorem{corollary}[theorem]{Corollary}
% \theoremstyle{theorem}
\newtheorem{proposition}[theorem]{Proposition}
% \theoremstyle{theorem}
\newtheorem{example}[theorem]{Example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\expect}{\ensuremath{\mathbb{E}}}

\newenvironment{talign}
{\let\displaystyle\textstyle\align}
{\endalign}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\circuit}{\ensuremath{p}}
\newcommand{\inputs}{\ensuremath{\text{in}}}

\newcommand{\Xvars}{\ensuremath{\mathbf{X}}}
\newcommand{\xvars}{\ensuremath{\mathbf{x}}}
\newcommand{\Xvar}{\ensuremath{X}}
\newcommand{\xvar}{\ensuremath{x}}

\newcommand{\Zvars}{\ensuremath{\mathbf{Z}}}
\newcommand{\zvars}{\ensuremath{\mathbf{z}}}
\newcommand{\Zvar}{\ensuremath{Z}}
\newcommand{\zvar}{\ensuremath{z}}

\newcommand{\poset}{\ensuremath{\mathcal{O}}}

\newcommand{\weight}{\ensuremath{w}}
\newcommand{\nweight}{\ensuremath{\omega}}
\newcommand{\nparams}{\ensuremath{\mathbf{\theta}}}

\newcommand{\scope}{\ensuremath{\phi}}

\newcommand{\eff}{\ensuremath{{eff}}}
\newcommand{\con}{\ensuremath{{con}}}

\newcommand{\parents}{\ensuremath{{pa}}}
\newcommand{\ancestors}{\ensuremath{{an}}}

\newcommand{\component}{\ensuremath{{\kappa}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\smalllinewidth}{0.6pt}
\newcommand{\midlinewidth}{1.0pt}
\newcommand{\midlinewidthx}{2.0pt}
\newcommand{\largelinewidth}{1.7pt}
\newcommand{\middist}{24pt}
\newcommand{\middistt}{20pt}
\newcommand{\middisttt}{28pt}
\newcommand{\largedist}{30pt}
\newcommand{\hugedist}{50pt}
\newcommand{\smalldist}{20pt}
\newcommand{\smalldistt}{4pt}
\newcommand{\tinydist}{5pt}
\newcommand{\intermiddist}{30pt}
\newcommand{\sqintermiddist}{15.5pt}
\newcommand{\halfdist}{4pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Redefine the proof environment to modify margins
\makeatletter
\renewenvironment{proof}[1][\proofname]{\par
    \pushQED{\qed}%
    \normalfont\topsep0pt \partopsep0pt % Adjust the vertical spacing above
    \trivlist
    \item[\hskip\labelsep
                \itshape
                #1\@addpunct{.}]\ignorespaces
}{%
    \popQED\endtrivlist\@endpefalse
    \vskip 1ex  % Add some flexible glue for the bottom margin
}
\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Commands definitions
\newcommand{\setbackgroundcolour}{\pagecolor[rgb]{0.19,0.19,0.19}}  
\newcommand{\settextcolour}{\color[rgb]{0.77,0.77,0.77}}    
\newcommand{\invertbackgroundtext}{\setbackgroundcolour\settextcolour}

%Command execution. 
%If this line is commented, then the appearance remains as usual.
% \invertbackgroundtext

\begin{document}

\maketitle

\begin{abstract}
    Probabilistic circuits (PCs) have gained prominence in recent years as a versatile framework for discussing probabilistic models that support tractable queries and are yet expressive enough to model complex probability distributions.
    Nevertheless, tractability comes at a cost: PCs are less expressive than neural networks.
    In this paper we introduce probabilistic neural circuits (PNCs), which strike a balance between PCs and neural nets in terms of tractability and expressive power. Theoretically, we show that PNCs can be interpreted as deep mixtures of Bayesian networks. Experimentally, we demonstrate that PNCs constitute powerful function approximators.
    % In this paper we explore This paper explores the interplay between PCs and neural networks, presenting the concept of Probabilistic Neural Circuits (PNCs) as an intermediary model. PNCs are introduced as deep mixtures of Bayesian networks, offering a structured approach to capturing complex dependencies in data.
\end{abstract}

\section{Introduction}

In recent years probabilistic circuits (PCs) (also called sum-product networks)~\citep{darwiche2003differential,poon2011sum} have emerged as an assembly language to talk about tractable probabilistic models and inference therein~\citep{vergari2021compositional}. The core idea is quite simple: we start with a set of independent random variables and construct complex probability distribution by recursively adding and multiplying them together.
There are two common ways of interpreting PCs. Firstly, we can consider them to be hierarchical mixture models. Secondly, we look at them as neural nets consisting of sums, products, and atomic probability distributions.

Most of the recent advances in the field adhere to the second perspective: use an overparametrized probabilistic model and fit it to data using gradient based methods by leveraging discrete GPUs~\citep{peharz2019random,dang2021juice}. The computation units of such circuits are organized in a layered fashion. We give an example in Figure~\ref{fig:circuit}.

\begin{figure}[t]
    \centering
    \resizebox{0.85\columnwidth}{!}{
        \tikzset{point/.style={circle,inner sep=0pt,minimum size=3pt,fill=red}}

    \begin{tikzpicture}

        \varnode[line width=\midlinewidth]{v11}{$X_1$};
        \varnode[line width=\midlinewidth, right=\middist of v11]{v12}{$X_1$};
        \varnode[line width=\midlinewidth, right=\middist of v12]{v21}{$X_2$};
        \varnode[line width=\midlinewidth, right=\middist of v21]{v22}{$X_2$};

        \sumnode[line width=\midlinewidth, above=\smalldist of v11]{s11};
        \sumnode[line width=\midlinewidth, right=\halfdist of s11]{s12};
        \sumnode[line width=\midlinewidth, above=\smalldist of v12]{s1k};

        \sumnode[line width=\midlinewidth, above=\smalldist of v21]{s21};
        \sumnode[line width=\midlinewidth, right=\halfdist of s21]{s22};
        \sumnode[line width=\midlinewidth, above=\smalldist of v22]{s2k};

        \prodnode[line width=\midlinewidth, above=\smalldist of s1k]{p121};
        \prodnode[line width=\midlinewidth, right=\halfdist of p121]{p122};
        \prodnode[line width=\midlinewidth, above=\smalldist of s21]{p12k};

        \sumnode[line width=\midlinewidth, above=\smalldist of p121]{s121};
        \sumnode[line width=\midlinewidth, above=\smalldist of p122]{s122};
        \sumnode[line width=\midlinewidth, above=\smalldist of p12k]{s12k};

        % edges
        \edge[line width=\midlinewidth,left] {s11, s12, s1k} {v11, v12};
        \edge[line width=\midlinewidth,left] {s21, s22, s2k} {v21, v22};

        \edge[line width=\midlinewidth,dashed] {p121} {s11, s21};
        \edge[line width=\midlinewidth,dashed] {p122} {s12, s22};
        \edge[line width=\midlinewidth,dashed] {p12k} {s1k, s2k};

        \edge[line width=\midlinewidth,left] {s121, s122, s12k} {p121, p122, p12k};

        %%%%% V3 and V4
        \varnode[line width=\midlinewidth, right=\middist of v22]{v31}{$X_3$};
        \varnode[line width=\midlinewidth, right=\middist of v31]{v32}{$X_3$};
        \varnode[line width=\midlinewidth, right=\middist of v32]{v41}{$X_4$};
        \varnode[line width=\midlinewidth, right=\middist of v41]{v42}{$X_4$};

        \sumnode[line width=\midlinewidth, above=\smalldist of v31]{s31};
        \sumnode[line width=\midlinewidth, right=\halfdist of s31]{s32};
        \sumnode[line width=\midlinewidth, above=\smalldist of v32]{s3k};

        \sumnode[line width=\midlinewidth, above=\smalldist of v41]{s41};
        \sumnode[line width=\midlinewidth, right=\halfdist of s41]{s42};
        \sumnode[line width=\midlinewidth, above=\smalldist of v42]{s4k};

        \prodnode[line width=\midlinewidth, above=\smalldist of s3k]{p341};
        \prodnode[line width=\midlinewidth, right=\halfdist of p341]{p342};
        \prodnode[line width=\midlinewidth, above=\smalldist of s41]{p34k};

        \sumnode[line width=\midlinewidth, above=\smalldist of p341]{s341};
        \sumnode[line width=\midlinewidth, above=\smalldist of p342]{s342};
        \sumnode[line width=\midlinewidth, above=\smalldist of p34k]{s34k};

        % edges
        \edge[line width=\midlinewidth,left] {s31, s32, s3k} {v31, v32};
        \edge[line width=\midlinewidth,left] {s41, s42, s4k} {v41, v42};

        \edge[line width=\midlinewidth,dashed] {p341} {s31, s41};
        \edge[line width=\midlinewidth,dashed] {p342} {s32, s42};
        \edge[line width=\midlinewidth,dashed] {p34k} {s3k, s4k};

        \edge[line width=\midlinewidth,left] {s341, s342, s34k} {p341, p342, p34k};

        %%%% final root
        \prodnode[line width=\midlinewidth, above=90pt of s2k]{p12341};
        \prodnode[line width=\midlinewidth, right=\halfdist of p12341]{p12342};
        \prodnode[line width=\midlinewidth, above=90pt of s31]{p1234k};

        \sumnode[line width=\midlinewidth, above=\smalldist of p12342]{s12341};

        \edge[line width=\midlinewidth,dashed] {p12341} {s121, s341};
        \edge[line width=\midlinewidth,dashed] {p12342} {s122, s342};
        \edge[line width=\midlinewidth,dashed] {p1234k} {s12k, s34k};

        \edge[line width=\midlinewidth,left] {s12341} {p12341, p12342, p1234k};

        \draw [decorate, decoration = {brace, mirror, amplitude=10pt}, ultra thick] (6.2,4.7) --  (6.2,6.8) node[pos=0.5,right=10pt]{ \begin{tabular}{c}
                root layer \\(Layer 3)
            \end{tabular} };
        \draw [decorate, decoration = {brace, mirror, amplitude=10pt}, ultra thick] (9.1,2.2) --  (9.1,4.3) node[pos=0.5,right=10pt]{\begin{tabular}{c}
                sum-product layer \\(Layer 2)
            \end{tabular} };
        \draw [decorate, decoration = {brace, mirror, amplitude=10pt}, ultra thick] (10.5,-0.4) --  (10.5,1.7) node[pos=0.5,right=10pt]{\begin{tabular}{c}
                leaf layer \\(Layer 1)
            \end{tabular} };

        \node[draw=red, circle,dashed,ultra thick, minimum width=0.7cm] at (p121.center) (c1) {};
        \node[draw=red, circle,dashed,ultra thick, minimum width=0.7cm] at (p122.center) (c2) {};
        \node[draw=red, circle,dashed,ultra thick, minimum width=0.7cm] at (p12k.center) (c3) {};

        \node[draw=red, circle,dashed, ultra thick, minimum width=0.7cm, below=1.2 of p12342, inner sep=0pt]  (mid) {};

        \path (c1) edge[-, red,dashed, ultra thick, in=160, out=35] (mid);
        \path (c2) edge[-, red,dashed, ultra thick, in=175, out=45] (mid);
        \path (c3) edge[-, red,dashed, ultra thick, in=220, out=-20] (mid);

        \path (mid) edge[->, red,dashed, ultra thick, in=190, out=25] (s341);
        \path (mid) edge[->, red,dashed, ultra thick, in=220, out=10] (s342);
        \path (mid) edge[->, red,dashed, ultra thick, in=230, out=-10] (s34k);

    \end{tikzpicture}
    }
    \caption{Layered probabilistic circuit following the construction of~\citep{shih2021hyperspns}. Data (modeled as random variables) is first fed into the leaf layer at the bottom. The output of the leaf layer is a mixture of distributions produced by the sum units. In the sum-product layer (Layer 2)  mixtures of random variables are combined by taking pairwise products, these are then again mixed using sum units. Finally, the root layer (at the top) gives us the joint probability distribution. The red edges indicate functional dependencies not present in traditional probabilistic circuits but present in probabilistic neural circuits.}
    \label{fig:circuit}
\end{figure}

A major advantage of PCs is their ability to answer certain queries in polynomial time -- given that adequate restrictions are imposed on a circuit's structure~\citep{vergari2021compositional}. An example of such a tractable query would be the computation of conditional probabilities for so-called \textit{smooth and decomposable} PCs~\citep{darwiche2001decomposable,darwiche2003differential}.

This tractability, however, comes at a hefty price: the properties imposed on PCs in order to ensure polynomial time queries limit their expressive power~\citep{martens2014expressive,sharir2018sum,zhang2021probabilistic}.
This is in contrast to general neural networks and even sum-product networks with fewer structural constraints~\citep{delalleau2011shallow,kileel2019expressive}. \citet{martens2014expressive} have shown that decomposability is in fact a necessary condition for tractable marginal inference.

Nevertheless, using the concepts of \textit{conditional smoothness} and \textit{conditional decomposability}~\citep{sharir2018sum}, we study in this paper the space of models that lie in between probabilistic circuits and neural networks.
Concretely, we make the following contributions:
\begin{enumerate}
    \item We introduce conditional probabilistic circuits, from which we construct  probabilistic neural circuits (PNCs), which we interpret as deep mixtures of Bayesian nets.
    \item  We provide a prescription to construct layered PNCs.
    \item We provide an implementation of layered PNCs and experimentally study their expressive power.
\end{enumerate}

Our work is influenced by that of \citet{sharir2018sum}. We discuss the relationship to their approach (dubbed sum-product-quotient networks) in Section~\ref{sec:related}.

\section{Preliminaries}
\label{sec:prelim}

In the remainder of the paper we will denote random variables by uppercase $\Xvar$'s, the corresponding values are lowercase $\xvar$'s. Sets of random variables and their corresponding values are typed in boldface: $\Xvars$ and $\xvars$, respectively.
The definitions and notions we introduce in this section are loosely based on the work of~\citet{vergari2021compositional}.

\begin{definition}[Probabilistic Circuit]
    \label{def:circuit}
    A probabilistic circuit over random variables $\Xvars$ is a parametrized computational graph encoding a probability density function $\circuit(\Xvars{=}\xvars)$.
    The circuit consists of three kinds of computational units:
    \textit{leaf}, \textit{product}, and \textit{sum}.
    Each product or sum unit receives inputs from a set of input units denoted by $\inputs(n)$.
    Each unit $k$ encodes a function $\circuit_{k}(\cdot)$ as follows:
    \begin{align*}
        {\circuit}_k(\xvars_n)=
        \begin{cases}
            f_k({\xvars_n})                                             & \text{if $k$ leaf unit}    \\
            \circuit_{k_l}(\xvars_{n_l})  \circuit_{k_r} (\xvars_{n_r}) & \text{if $k$ product unit} \\
            \sum_{j\in\inputs(k)} \weight_{kj} \circuit_j(\xvars_n)     & \text{if $k$ sum unit}
        \end{cases}
        % \label{eq:def:circuit}
    \end{align*}
    where $f_k(\xvars_n)$  denotes a parametrized probability distribution having as support the sample space of the random variables in $\Xvars_n$.
\end{definition}

\begin{definition}[Scope]
    \label{def:scope:cond}
    The scope of a unit $k$, denoted by $\scope(k)$, is the set of random variables $\Xvars_n$ for which the function $\circuit_k (\cdot)$ encodes a probability distribution.
\end{definition}

Two important properties that are usually imposed on probabilistic circuits are smoothness and decomposability as they allow for tractable queries, \eg computing marginals~\citep{darwiche2001decomposable,darwiche2003differential}.
\begin{definition}[Smoothness]
    A circuit is smooth if for every sum unit $k$ its inputs encode distributions over the same random variables:
    $\forall j_1, j_2 {\in} \inputs(k)$ it holds that $\scope(j_1){=}\scope(j_2) $.
\end{definition}

\begin{definition}[Decomposability]
    A circuit is decomposable if the inputs of every product unit
    $k$ encode distributions over disjoint sets of random variables:
    $\scope (k_l) \cap \scope (k_r) = \emptyset$ with $\{k_l, k_r\}= \inputs (k)$.
\end{definition}

\begin{definition}[Valid Probabilsitic Circuit]
    We call a probabilistic circuit valid if for every unit $k$ we have that $p_k(\xvars_n){\geq} 0$ and $\int p_k(\xvars_n) \differential \xvars_n{=}1$.\footnote{Note that our notion of validity is slightly stricter than in its original definition, \cf \citep{poon2011sum}}
\end{definition}

As discussed by \citet{peharz2015theoretical}, probabilistic circuits are valid if they are smooth, decomposable, and that for the weights in the sum units we have $\weight_{kj} \in \mathbb{R}^+$ and $\sum_{j\in\inputs(k)} \weight_{kj} = 1$ for every $k$. Note that the notation in Definition~\ref{def:circuit} already suggests that the circuit is smooth as the inputs to the sum units are functions over the same set of variables $\Xvars_n$.

Furthermore, we can assume, without loss of generality, that sum and product units occur in an alternating fashion in the circuit~\citep{peharz2020einsum}.
This observation naturally leads to a layer-wise construction of probabilistic circuits where consecutive layers alternate between sum and product layers.
Such layered probabilistic circuits~\citep{peharz2019random} have the advantage that the computations within a layer can be trivially parallelized.
We can further abstract the layers in a circuit by fusing together alternating sums and products into a single sum-product layer~\citep{peharz2020einsum}.

In Figure~\ref{fig:circuit} we give a graphical representation of a layered circuit. Layers consist of  blocks of computational units that are processed sequentially in a bottom-up fashion. Layers are themselves constituted of so-called partitions. The circuit in Figure~\ref{fig:circuit} has four partitions in the leaf layer, two in the sum-product layer, and a single partition in the root layer. By construction, partitions in the same layer have disjoint scopes. Moreover, partitions are further subdivided into input components and output components, which constitute the elemental computing units. The circuit in Figure~\ref{fig:circuit} has, except at the very bottom and top, three such input and output components in each partition.

We can uniquely identify each computational unit (or component) in the circuit by specifying the layer, the partition, whether it is an input or an output, and its position within a partition. Counting layers from bottom to top, and partitions and units from left to right. Each component can be identified using 4 indices: $\component_{l,p,i,c}$. The first index $l$ identifies the layer, the second $p$ the partition, the third $i \in \{1,2 \}$ whether it is an input or output, and the fourth $c$ the horizontal position within a partition.
For instance, the symbol $\component_{2223}$ corresponds to  the upper-right unit in the sum-product layer.

\section{Conditional Probabilistic Circuits}
\label{sec:cpc}

We will first introduce the notion of posets (partially ordered sets) of random variables (Section~\ref{sec:poset}).
This will allow us to generalize probabilistic circuits to conditional probabilistic circuits, which we interpret as deep mixtures of Bayesian networks (Section ~\ref{sec:dmbn}).
We then introduce probabilistic neural circuits and their tractable queries (Section~\ref{sec:pnc}).

\subsection{Partially Ordered Random Variables}
\label{sec:poset}

Consider a set of random variables $\Xvars {=}\{\Xvar_1,\dots, \Xvar_N \}$ on which we impose the \textit{parents} relationship $\parents(\cdot)$.
The parents relationship induces a directed acyclic graph on the random variables $\Xvars$, where the nodes are the random variables themselves and an edge is present between $\Xvar_i$ and $\Xvar_j$ if $\Xvar_i {\in} \parents(\Xvar_j)$.
This gives us a partial ordering of the variables $\Xvars$.
We also define the ancestor relationship $\ancestors(\cdot)$ to be the transitive closure of $\parents(\cdot)$. That is, the ancestors of a random variable are its direct parents and recursively their parents.
Furthermore, we denote the poset for the random variables $\Xvars$ by $\mathcal{O}(\Xvars)$. We say that the relationship $\Xvars_i {\sqsubset} \Xvars_j$ between two sets holds if
$
    \forall \Xvar_r {\in} \Xvars_i, \Xvar_q {\in} \Xvars_j : \Xvar_q {\notin}  \ancestors(\Xvars_r)
$.
We also define the relation $\Xvar_i\sqsubset \Xvar_j$ on random variables as $\{\Xvar_i \}{\sqsubset} \{\Xvar_j \}$.

\begin{example}[Bayesian Network]
    Partially ordered random variables induce a factorization of a joint probability distribution. A prominent example of such a factorization are Bayesian networks (\cf Figure ~\ref{fig:bn}) where we have:
    \begin{talign}
        \label{eq:bn}
        p(\Xvars{=}\xvars) = \prod_{n:\Xvar_n \in \Xvars} p_n(X_n{=}x_n \mid \Xvars_{\parents(n)}{=}\xvars_{\parents(n)}).
    \end{talign}
\end{example}

\begin{figure}

    \begin{minipage}{0.50\linewidth}
        \begin{align*}
            X_1 \sqsubset X_2 &  & X_1 \sqsubset X_4 \\
            X_1 \sqsubset X_3 &  & X_3 \sqsubset X_1 \\
            X_3 \sqsubset X_2 &  & X_3 \sqsubset X_4 \\
            X_2 \sqsubset X_3 &  & X_2 \sqsubset X_4 \\
        \end{align*}

    \end{minipage}%
    \begin{minipage}{0.49\linewidth}
        \resizebox{0.9\linewidth}{!}{

            \centering

            \begin{tikzpicture}
                % Nodes
                \node[draw, circle,ultra thick] (x3) {$X_3$};
                \node[draw, circle, below=of x3,ultra thick] (x2) {$X_2$};
                \node[draw, circle,  right=2cm of $(x3)!0.5!(x2)$3,ultra thick] (x4) {$X_4$};
                \node[draw, circle, left=of x2,ultra thick] (x1) {$X_1$};

                % Edges
                \draw[->, ultra thick] (x1) -- (x2);
                \draw[->,ultra thick] (x2) -- (x4);
                \draw[->, ultra thick] (x3) -- (x4);
            \end{tikzpicture}
        }
    \end{minipage}
    \caption{Right: Bayesian network. Left: partial order relations that hold.}
    \label{fig:bn}
\end{figure}

In the example above we denote $\parents(X_n)$ by $\Xvars_{\parents(n)}$. This will allow us to omit the random variable $\Xvar_n$ when writing down a probability distribution and only use the instantiation $\xvar_n$ instead.

\subsection{Deep Mixtures of Bayesian Networks}
\label{sec:dmbn}

\begin{definition}[Conditinal Probabilsitic Circuit (CPC)]
    \label{def:con:circuit}
    A CPC $\circuit_k$ over a poset $\poset(\Xvars)$ is a parametrized computational graph encoding a probability density function $\circuit(\Xvars{=}\xvars)$.
    The  CPC consists of
    \textit{leaf}, \textit{product}, and \textit{sum} units.
    Each unit $k$ encodes a function $\circuit_{k}$ as follows:
    \begin{align}
        \label{eq:con:circuit}
         & {\circuit}_k(\xvars_n \mid  \xvars_{\parents(n)})
        \\
         & =
        \begin{cases}
            f_k({\xvars_n}\mid   \xvars_{\parents(n)})                                          & \text{if leaf}    \\
            \circuit_{k_l}(\xvars_{n_l}\mid   \xvars_{\parents(n_l)})
            \circuit_{k_r} (\xvars_{n_r}\mid \xvars_{\parents(n_r)})                            & \text{if product} \\
            \sum_{j\in\inputs(k)} \weight_{kj} \circuit_j(\xvars_n  \mid  \xvars_{\parents(n)}) & \text{if sum}
        \end{cases}
        \nonumber
        % \label{eq:def:circuitconditional}
    \end{align}
    where $f_k(\xvars_n)$ denotes a parametrized probability distribution having as support the sample space of the random variables in $\Xvars_n$.
    % Furthermore, we necessitate that $p_n(\xvars_n){\geq} 0$ and $\int p_n(\xvars_n) \differential \xvars_n{=}1$ for every unit $n$.
\end{definition}

\begin{definition}[Scope (CPC)]
    \label{def:scope}
    The scope  $\scope(k)$ of a unit $k$ encoding the probability distribution $\circuit_k(\xvars_n\mid \xvars_{\parents(n)} )$ is the set of random variables $\Xvars_n$.
\end{definition}

\begin{corollary} A conditional probabilistic circuit over an unordered set of random variables is a (non-conditional) probability circuit.
\end{corollary}
\begin{proof}
    Having no order means that $\parents(\Xvars)=\emptyset$ for every $\Xvar \in \Xvars$. This then means that the conditioning sets in Definition~\ref{def:con:circuit} are all empty and that we recover a (non-conditional) probabilistic circuit.
\end{proof}

We can now also introduce the notions of \textit{conditional smoothness} and \textit{conditional decomposability}:
\begin{definition}[Conditional Smoothness]
    A CPC is conditionally smooth if for every sum unit $k$ it holds that
    $\forall j_1, j_2 {\in} \inputs(k):  \scope(j_1){=}\scope(j_2) $
\end{definition}

\begin{definition}[Conditional Decomposability]
    A CPC is conditionally decomposable if for every product unit
    $k$ it holds that $\scope (k_l) \cap \scope (k_r) = \emptyset$ with $\{k_l, k_r\}= \inputs (k)$.
\end{definition}

\begin{corollary}
    A conditionally smooth and conditionally decomposable CPC is smooth and decomposable if its random variables $\Xvars$ are unordered.
\end{corollary}

\begin{definition}[Valid CPC]
    \label{def:valid_cpc}
    We call a CPC valid if for every unit $k$ we have that $p_k(\xvars_n {\mid} \xvars_{\parents(n)}){\geq} 0$ and $\int p_k(\xvars_n {\mid} \xvars_{\parents(n)}) \differential \xvars_n{=}1$.
\end{definition}

\begin{theorem}[Validity for CPCs]
    A CPC is valid if it is conditionally smooth, conditionally decomposable, and for every sum unit $n$ it holds that $\weight_{nm}{\geq} 0$ and $\sum_{m\in\inputs(n)} \weight_{nm}{=}1$.
\end{theorem}

\begin{proof}
    We start by rewriting the alternating sums and products of a CPC in its flat representation using the fact that products distribute over summations:
    \begin{align}
        \circuit_k(\xvars_n)
         & =  \sum_{\tau \in \mathcal{T}} \weight_\tau  \prod_{\rho \in \tau} \rho (\xvars_\rho)
        \label{eq:tree_decomp_PCP}
    \end{align}
    Here, $\mathcal{T}$ denotes the set of all products of leaf distributions in the CPC and $\rho \in \tau$ denotes a factor in one of these products
    (we refer to \citep{zhao2016unified} for a more detailed account).

    Invoking decomposability of the product units we have that each random variable $X_n$ only picks up a single factor $\rho (\xvar_\rho)$, which means that we can identify each $\rho (\xvar_\rho)$ with a specific $f_k(\xvar_n\mid \xvars_{\parents(n)} )$. This lets us rewrite Equation~\ref{eq:tree_decomp_PCP} as:
    \begin{align}
        \circuit_k(\xvars_n)
         & =  \sum_{\tau \in \mathcal{T}} \weight_\tau  \prod_{k: f_k \in \tau} f_k (\xvars_n \mid \xvars_{\parents(n)})
        \label{eq:f_tree_decomp_PCP}
    \end{align}
    Given that the $f_k$ are by definition (conditional) probability distributions their product forms a joint probability distribution as well. Next, we exploit smoothness, which states that the inputs to sum units have identical scope. This means that all the terms in the flat representation of $\circuit (\Xvars_n)$ mention the identical set of random variables and each term in the flat representation forms indeed a joint probability over the random variables $\Xvars_n$.

    Lastly, \citet{peharz2015theoretical} have shown that having normalized weights in the sum units of a circuit results in normalized weights $\weight_\tau$ in the flat representation. This lets us conclude that the circuit $p_k(\xvars_n)$ is a probability distribution. Note that we did not make any reference in our reasoning to any specific unit in the circuit. This means that our argument holds for all units in a conditionally smooth and conditionally decomposable circuit with normalized weights, which means in turn that such a circuit is valid.
\end{proof}

In light of Equation~\ref{eq:f_tree_decomp_PCP} and comparing it to Equation~\ref{eq:bn}, we can interpret CPCs as deep (or hierarchical) mixtures of Bayesian networks. This is analogous to interpreting probabilistic circuits as deep mixtures of fully factorized distributions~\citep{poon2011sum}.

\subsection{PNCs and Their Tractable Queries}
\label{sec:pnc}

The computational efficiency of probabilistic circuits stems from the fact the circuits evaluations are broken down into sub-evaluations, which are then cached and reused.
Inspecting, however, the functional form of the sum units in a CPC (\cf Equation~\ref{eq:con:circuit}), this is not the case:
each term in the sum over $j$ requires a separate conditional probability for each instantiation of the variables $\Xvars_{\parents(n)}$.
This means that we would need (assuming binary random variables) $2^{|\Xvars_{\parents(n)}|}$ functions to encode the conditional probabilities.
We alleviate this issue as follows. First, we rewrite the functional form of the sum units using Bayes' rule:
\begin{align}
    \sum_{j\in\inputs(k)}
    \weight_{kj}
    \frac{
        \circuit_j(  \xvars_{\parents(n)} \mid \xvars_n  )
    }{
        \circuit_j(  \xvars_{\parents(n)  })
    }
    \circuit_j(  \xvars_n)
\end{align}
Second, we make the following approximation:
\begin{align}
    \weight_{kj}
    \frac{
        \circuit_j(  \xvars_{\parents(n)} \mid \xvars_n  )
    }{
        \circuit_j(  \xvars_{\parents(n)  })
    }
    \approx
    \nweight_{kj}(\xvars_{\ancestors(n)}),
\end{align}
where $\nweight_{kj}(\cdot)$ is a neural network depending on the set of ancestors $\ancestors(\xvar_n)$. This now allows us to formally introduce probabilistic neural circuits.
\begin{definition}[Probabilistic Neural Circuit (PNC)]
    \label{def:pnc}
    A PNC is a conditionally smooth and conditionally decomposable  CPC where sum units take the following functional form:
    \begin{align}
        {\circuit}_k(\xvars_n \mid  \xvars_{\parents(n)})
        =
        \sum_{j\in\inputs(k)} \nweight_{kj} ( \xvars_{\ancestors(n)}) \circuit_j(\xvars_n),
        \label{eq:pnc}
    \end{align}
    with $\nweight_{kj}: \Omega( \ancestors(\xvar_n)) \rightarrow [0,1 ]$ being a neural network mapping from the sample space of the random variables $\ancestors(\xvar_n)$ to a real  between zero and one, and where it holds that $\sum_{j\in\inputs(k)} \nweight_{kj} = 1$.
\end{definition}
It can easily be seen that PNCs, already encode valid circuits (\cf. Definition~\ref{def:valid_cpc}) by construction.
Intuitively, we interpret PNCs as neural approximations of CPCs.
Note that it is this approximation that makes PNCs tractable: PNCs only need a single circuits for each $j$ while CPCs need a circuit for every $j$ and every instantiation of $\xvars_{\parents(n)}$.

\begin{proposition}
    Probabilistic circuits are PNCs.
\end{proposition}
\begin{proof}
    If the values $\nweight(\xvars_{\ancestors(n)})$ do not depend on $\xvars_{\ancestors(n)}$ we have $|\inputs(k)|$ constants that sum up to $1$. This means that the weights in the sum units do not depend on the conditioning sets from the antecedent product layer, and we can omit any conditioning sets. The definition of a PNC in this case is then equivalent to the definition of a probabilistic circuit.
\end{proof}

Given the definition of PNCs we can now determine tractable queries that we can perform.

\begin{proposition}[Density Evaluation]
    Given a probabilistic circuit $p_k$ over the random variables $\Xvars$. We can evaluate the circuit  at the instantiation $\xvars$ in linear time with respect to the size of the circuit.
\end{proposition}
\begin{proof}
    This follows simply from the fact that a circuit is a (non-recurrent) computation graph and that we can simply evaluate it by computing input units before output units.
\end{proof}

\begin{proposition}[Ordered Marginals]
    \label{prop:ord_marg}
    Consider a PNC $\circuit_k(\xvars_m, \xvars_e)$  we can then compute the marginal $\circuit_k(\xvars_e)$ in polynomial time if $\Xvars_e {\sqsubset} \Xvars_m$.
\end{proposition}
\begin{proof}
    We start by writing out explicitly the single elements in the set $\xvars_m$:
    \begin{align}
        \circuit_k(\xvars_m, \xvars_e) = \circuit_k(\xvar_\mu, \dots, \xvar_1, \xvars_e)
    \end{align}
    where the order of $\{ \xvar_\mu, \dots, \xvar_1\}= \xvars_m$ is arbitrary but respects the partial order $\poset(\Xvars_m \cup \Xvars_e )$.
    Given that the circuit is conditionally smooth and conditionally decomposable, we know that it encodes a proper probability distribution over its variables.
    We can hence obtain the marginal $\circuit(\xvars_e)$ by integrating over the possible values $\xvars_m$:
    \begin{align}
        \circuit_k(\xvars_e)
        =
        \int \cdots \int p_k(\xvar_\mu, \dots, \xvar_1,\xvars_e)  \differential  \xvar_\mu \cdots \differential \xvar_1
    \end{align}
    As $\xvar_\mu$ does not appear in any of the conditioning sets and as any product unit $q$ is decomposable we can simply push the summation to the input unit $r$ of $q$ for which we have  $\Xvar_\mu \in \scope (r)$.
    For summation units we exploit the linearity of the integral and distribute the integral over the terms in the sum.
    Performing this recursively brings the integral to the leaves where we have  $ \int f_i(x_\mu \mid x_{\parents(\mu)} ) \differential \xvar_\mu =1$.

    Up to this point the marginalization in CPC is identical to marginalization in probabilistic circuits. Contrary, to probabilistic circuits, however, we now need to propagate back up the result of the marginalization. Assuming, without loss of generality that leaf units feed into sum units, we then have
    \begin{align*}
         & \sum_{g\in\inputs(h)} \nweight_{hg} (\xvars_{\ancestors( \mu)})
        \int \circuit_g (\xvar_\mu  ) \differential \xvar_\mu
        % 
        {=} \sum_{g\in\inputs(h)} \nweight_{hg} (\xvars_{\ancestors( \mu)})
        {=}1
    \end{align*}
    The next product node that we encounter on our way up through the circuit is of the form:
    \begin{align}
         & \circuit_{l}(\xvars_{\mu-1}\mid   \xvars_{\parents(\mu-1)})
        \int  \circuit_{r} (\xvars_{\mu}\mid \xvars_{\parents(\mu)}) \differential \xvars_\mu
        \nonumber
        \\
         & =\circuit_{l}(\xvars_{\mu-1}\mid  \xvars_{\parents(\mu-1)})
        \nonumber
    \end{align}
    At this point we have integrated out the variable $\xvar_\mu$ from the circuit by traversing a number of units linear in the size of the circuit (assuming proper caching~\citep{vergari2021compositional}). Repeating this procedure for the remaining set of ordered random variables $\{\Xvar_{\mu-1},\dots,\Xvar_1 \}$ gives us the distribution $\circuit_k(\xvars_e)$ in polynomial time.
\end{proof}
The proof follows a similar reasoning to the case of probabilistic circuits. The delicate point was to show that in the product unit one of the factors drops out. This is important as $\Xvar_{\parents(\mu)}$ might include $\Xvar_{\mu-1}$. Retaining such a dependency would prevent us from performing tractable ordered marginalization.

\begin{corollary}[Ordered Conditionals]
    Assuming a PNC $\circuit_k (\xvars_m, \xvars_o, \xvars_e)$ where $\Xvars_e \sqsubset \Xvars_o \sqsubset \Xvars_m$ holds lets us compute the conditional $p(\xvars_o \mid \xvars_e)$ in polynomial time.
\end{corollary}
\begin{proof}
    We first apply the definition of the conditional probability:
    $p(\xvars_o \mid \xvars_e)=\nicefrac{p(\xvars_o , \xvars_e)}{p( \xvars_e)}$.
    Using the law of total probability we rewrite the ratio as
    $$
        p(\xvars_o \mid \xvars_e)
        =
        \frac
        {
            \int p(\xvars_m , \xvars_o, \xvars_e) \differential \xvars_m
        }
        {
            \int p(\xvars_m , \xvars_o, \xvars_e) \differential \xvars_m \differential \xvars_o
        }
    $$
    and Proposition~\ref{prop:ord_marg} tells us that we can perform both marginalizations in polytime.
\end{proof}

% \begin{definition}[Conditional Determinism]

% \end{definition}

\section{Layered Probabilistic Neural Circuits}

While Equation~\ref{eq:pnc} provides a generic functional expression to compute the sum units in PNCs, it is not clear how to construct a PNC in the first place. That is, how do we link up the individual computation units such that they form a (valid) CPC.
For (non-conditional) probabilistic circuits potent structure learning algorithms have been developed in recent years  (\eg  hidden Chow-Liu trees~\citep{liu2021tractable} or random probabilistic circuits~\citep{di2021random}). It is not entirely clear how to adapt these to the setting of conditional probabilistic circuits. For this reason we study problems where, informed by the structure of the data itself, a structure for a PNC can be constructed. Concretely, we will study PNC structures tailored towards image data: features, \ie pixels, that are close to each other should also be close to each other in the circuit -- a fact already exploited by~\citet{poon2011sum}.

The structure we propose in this paper is rather simple and inspired by simple feed-forward neural networks and also layered (non-conditional) probabilistic circuits~\citep{peharz2020einsum}. More concretely,
the computation units in the current layer only depend on computation units in the previous layer. Furthermore, we wish the units within each layer to be computable in parallel (given the previous layer). To construct such a probabilistic neural circuit we take the circuit structure introduced by \citet{shih2021hyperspns} as a backbone and add additional edges to the computation graph in order to obtain a PNC from a probabilistic circuit.
For ease of exposition we detail our approach using one-dimensional data instead of two-dimensional data.

\subsection{Structure for One-Dimensional Data}

In order to study PNC structures, we introduce the concept of a partition graph, which is a hypergraph of a probabilistic circuit with the partitions of a circuit being the nodes and edges encoding the sub-partitions\footnote{In the circuit structure introduced by ~\citet{shih2021hyperspns}, all the product nodes decompose in the same fashion, \ie for product nodes with the same scope it is the sames variables that come from the left and right inputs, respectively. This is also called structured decomposabiltiy~\citep{darwiche2011sdd}. Partition trees are related to the concept of \textit{variable trees} in \textit{(probabilistic) sentential decision diagrams}~\citep{darwiche2011sdd,kisa2014probabilistic}. However, nodes in a variable tree do not constitute abstractions of a specific group of sum and product units and are a more general concept. }.  We give an example of a partition tree in Figure~\ref{fig:partition_graph}.

\begin{figure}[t]
    \begin{minipage}[c]{\linewidth}
        \centering

        \begin{tikzpicture}[
                dot/.style = {circle, draw, fill, minimum size=4pt, inner sep=0pt},
                level distance=1cm,
                level 1/.style={sibling distance=3cm},
                level 2/.style={sibling distance=1.5cm},
                level 3/.style={sibling distance=0.75cm},
                scale=1
            ]

            \node[dot] (00) {}
            child {node[dot] (10) {}
                    child {node[dot] (20) {}
                            child {node[dot] (30) {}}
                            child {node[dot] (31) {}}
                        }
                    child {node[dot] (21) {}
                            child {node[dot] (32) {}}
                            child {node[dot] (33) {}}
                        }
                }
            child {node[dot] (11) {}
                    child {node[dot] (22) {}
                            child {node[dot] (34) {}}
                            child {node[dot] (35) {}}
                        }
                    child {node[dot] (23) {}
                            child {node[dot] (36) {}}
                            child {node[dot] (37) {}}
                        }
                };

            \node[below= 0.1 of 30, inner sep=0pt] (X1) {\tiny{$X_1$}};
            \node[below= 0.1 of 31, inner sep=0pt] (x2) {\tiny{$X_2$}};
            \node[below= 0.1 of 32, inner sep=0pt] (x3) {\tiny{$X_3$}};
            \node[below= 0.1 of 33, inner sep=0pt] (x4) {\tiny{$X_4$}};
            \node[below= 0.1 of 34, inner sep=0pt] (X5) {\tiny{$X_5$}};
            \node[below= 0.1 of 35, inner sep=0pt] (X6) {\tiny{$X_6$}};
            \node[below= 0.1 of 36, inner sep=0pt] (X7) {\tiny{$X_7$}};
            \node[below= 0.1 of 37, inner sep=0pt] (X8) {\tiny{$X_8$}};

            \draw[<-, dashed, red] (11) -- (10);
            \draw[<-, dashed, red] (23) -- (22);
            \draw[<-, dashed, red] (22) -- (21);
            \draw[<-, dashed, red] (21) -- (20);
            \draw[<-, dashed, red, out=135] (23) to[out=150,in=30]  (21);
            \draw[<-, dashed, red, out=135] (22) to[out=-150,in=-30]  (20);

            % \draw[->, dashed, red] (37) -- (36);
            % \draw[->, dashed, red] (36) -- (35);
            % \draw[->, dashed, red] (35) -- (34);
            % \draw[->, dashed, red] (34) -- (33);
            % \draw[->, dashed, red] (33) -- (32);
            % \draw[->, dashed, red] (32) -- (31);
            % \draw[->, dashed, red] (31) -- (30);

            % \draw[->, dashed, red] (37) to[out=150,in=30] (35);
            % \draw[->, dashed, red] (36) to[out=-150,in=-30] (34);

            % \draw[->, dashed, red] (35) to[out=150,in=30] (33);
            % \draw[->, dashed, red] (34) to[out=-150,in=-30] (32);

            % \draw[->, dashed, red] (33) to[out=150,in=30] (31);
            % \draw[->, dashed, red] (32) to[out=-150,in=-30] (30);

        \end{tikzpicture}
    \end{minipage}
    \caption{A balanced partition tree of a probabilistic neural circuit with eight variables. The partition tree (in black) describes how the variables decompose (in terms of the scope function). The edges in red indicate functional (neural) dependencies between partitions.}
    \label{fig:partition_graph}
\end{figure}

Ignoring for now neural dependencies in PNCs (\ie the red edges in the partition tree in Figure~\ref{fig:partition_graph}) we describe the layer-wise operations. Let us assume, for the sake of simplicity, that the number of variables $N$ is a power of $2$. For instance, the circuits in Figure~\ref{fig:circuit} has $N=4$ and the partition diagram in Figure~\ref{fig:partition_graph} has $N=8$. Given that we merge partitions pairwise at each layer via multiplication, we obtain for the number of layers in a circuit $N_L= \log_2 N+1$. Formally, we express the layer-wise product units as follows:
\begin{align}
    \component_{l,p,1,c} = \component_{l-1,2p,2,c} \times \component_{l-1,2p+1,2,c},
    \tag{ProductLayer}
\end{align}
which describes how to compute the value of a component given the components of the previous layer.
The meaning of the indices is described in Section~\ref{sec:prelim}. Using this notation we can also express the sum units at the leaves and at the root:
\begin{align}
    \component_{1,p,2,c}
     & =
    \sum_{c'=1}^{N_D} \weight_{1,p,c,c'} \times \component_{1,p,1,c'}
    \tag{SumLeaf}
    \\
    \component_{N_L,1,2,1}
     & =
    \sum_{c'=1}^{N_C} \weight_{N_L,1,1,c'} \times \component_{N_L,1,1,c'}
    \tag{SumRoot}
\end{align}
In the equations above, $N_D$ denotes the number of initial components in the leaves and $N_C$ is the number of components throughout the circuits.
For the circuit in Figure~\ref{fig:circuit} we have $N_D{=}2$ and $N_C{=}3$.
The weights $w_{l,p,c.c'}$ are real valued constants and normalized over the $c'$ dimension. Note that, in contrast to the formulation in Section~\ref{sec:cpc},
weights (and also computation units in general) are not identified by a single index (\eg $k$ in $\circuit_k(\cdot)$) but by four indices.
To make this distinction explicit we denote the computation units by $\kappa_{l,p,i,c}$ instead of $p_k(\cdot)$.
Observe also that the root layer has only a single component. Hence, the $1$ as the last index instead of $c$.

Next, we describe the neural sum units in layered PNCs. To gain some intuition, consider the partial computation graph in Figure~\ref{fig:neural_dependencies}, where we show in more detail, compared to Figures ~\ref{fig:circuit} and~\ref{fig:partition_graph}, the neural dependencies present at the sum units. Formally, we express the values of neural sum units as follows:
\begin{align}
     & \component_{l,p,2,c}     \tag{NeuralSumLayer}
    \\
     & =
    \sum_{c'=1}^{N_C} \nweight_{l,p,c,c'}(\component_{l, p-\nu_{l,p}:p-1, 1, 1:N_C})
    \times
    \component_{l,p,1,c'}
    \nonumber
\end{align}
\begin{figure}[t]
    \begin{minipage}[c]{\linewidth}
        \centering
        \resizebox{0.9\columnwidth}{!}{
            \begin{tikzpicture}

                \prodnode[line width=\midlinewidth]{p21};
                \prodnode[line width=\midlinewidth, right=\halfdist of p21]{p22};
                \prodnode[line width=\midlinewidth, right=\halfdist of p22]{p23};

                \prodnode[line width=\midlinewidth, right=2\halfdist of p23]{p31};
                \prodnode[line width=\midlinewidth, right=\halfdist of p31]{p32};
                \prodnode[line width=\midlinewidth, right=\halfdist of p32]{p33};

                \node[draw, dashed, draw=red, circle, above right=0.875 and 0.35 of p23, inner sep=1pt,line width=\midlinewidth, minimum width=0.5cm] (w2){};
                \node[draw, dashed, draw=red,circle, left= \halfdist of w2, inner sep=1pt,line width=\midlinewidth,minimum width=0.5cm] (w1) {};
                \node[draw, dashed, draw=red,circle, right= \halfdist of w2, inner sep=1pt,line width=\midlinewidth,minimum width=0.5cm] (w3) {};

                \draw[-,red,dashed,line width=\midlinewidth] (p21) to (w1);
                \draw[-,red,dashed,line width=\midlinewidth] (p21) to (w2);
                \draw[-,red,dashed,line width=\midlinewidth] (p21) to (w3);
                \draw[-,red,dashed,line width=\midlinewidth] (p22) to (w1);
                \draw[-,red,dashed,line width=\midlinewidth] (p22) to (w2);
                \draw[-,red,dashed,line width=\midlinewidth] (p22) to (w3);
                \draw[-,red,dashed,line width=\midlinewidth] (p23) to (w1);
                \draw[-,red,dashed,line width=\midlinewidth] (p23) to (w2);
                \draw[-,red,dashed,line width=\midlinewidth] (p23) to (w3);

                \draw[-,red,dashed,line width=\midlinewidth] (p31) to (w1);
                \draw[-,red,dashed,line width=\midlinewidth] (p31) to (w2);
                \draw[-,red,dashed,line width=\midlinewidth] (p31) to (w3);
                \draw[-,red,dashed,line width=\midlinewidth] (p32) to (w1);
                \draw[-,red,dashed,line width=\midlinewidth] (p32) to (w2);
                \draw[-,red,dashed,line width=\midlinewidth] (p32) to (w3);
                \draw[-,red,dashed,line width=\midlinewidth] (p33) to (w1);
                \draw[-,red,dashed,line width=\midlinewidth] (p33) to (w2);
                \draw[-,red,dashed,line width=\midlinewidth] (p33) to (w3);

                \prodnode[line width=\midlinewidth, right=2\halfdist of p33]{p11};
                \prodnode[line width=\midlinewidth, right=0.5 of p11]{p12};
                \prodnode[line width=\midlinewidth, right=0.5 of p12]{p13};

                \sumnode[line width=\midlinewidth, above=1.75 of p11]{s13};

                \draw[-,line width=\midlinewidth] (p12) to (s13);
                \draw[-,line width=\midlinewidth] (p11) to (s13);
                \draw[-,line width=\midlinewidth] (p13) to (s13);

                \node[circle,inner sep=0pt,dashed,red] (lw1) at ($(p11)!0.5!(s13)+(-0.3,0.2)$) {$\nweight_1$};
                \node[circle,inner sep=0pt,dashed,red] (lw2) at ($(p12)!0.5!(s13)+(-0.15,-0.3)$) {$\nweight_2$};
                \node[circle,inner sep=0pt,dashed,red] (lw3) at ($(p13)!0.5!(s13)+(0.3,0.2)$) {$\nweight_3$};

                \draw[->,red,dashed,line width=\midlinewidth] (w1) to[out=45,in=150] (lw1);
                \draw[->,red,dashed,line width=\midlinewidth] (w2) to[out=45,in=-180] (lw2);
                \draw[->,red,dashed,line width=\midlinewidth] (w3) to[out=-15,in=-110] (lw3);

            \end{tikzpicture}
        }
    \end{minipage}
    \caption{Detailed graphical representation of neural dependencies in a PNC. The sum unit at the top outputs the weighted sum of the three product units at the bottom right. The weights for the sum are the outputs of a neural network for which it holds that $\sum_{i=1}^3=1$. They are computed using a neural network that takes as input the values of the six product units at the bottom left.}
    \label{fig:neural_dependencies}

\end{figure}

\begin{algorithm}[t]
    \caption{Layer-wise circuit evaluation}
    \label{alg:layer}
    \hspace*{\algorithmicindent} \textbf{Input:} $\xvars_o$. $\Xvars_m$ \\
    \hspace*{\algorithmicindent} \textbf{Output:} $p(\xvars_o)$ \\
    \hspace*{\algorithmicindent} \textbf{Require:} $\Xvars_o {\cup} \Xvars_m {=} \Xvars$,  $\Xvars_o {\sqsubset} \Xvars_m$
    \begin{algorithmic}[1]
        \State $\component_{p,c} \gets init(p,c,\xvars_o)$
        \State $\component_{p,c} \gets
            \sum_{c'=1}^{N_D} \weight_{1,p,c,c'} \times \component_{p,c'}$ \Comment{LeafLayer}
        \State $l \gets 2$
        \While {$l< N_L$}
        \State $\component_{p,c} \gets \component_{2p,c} \times \component_{2p+1,c} $ \Comment{ProductLayer}
        \State
        $ \component_{p,c}
            \gets
            \sum_{c'=1}^{N_C} \nweight_{l,p,c,c'}
            \times
            \component_{p,c'}
        $ \Comment{NeuralSumLayer}
        \State $l \gets l+1$
        \EndWhile
        \State $\component_{p,c} \gets \component_{2p,c} \times \component_{2p+1,c} $ \Comment{ProductLayer}
        \State $\component_{1,1} \gets \sum_{c'=1}^{N_C} \weight_{N_L,1,c,c'} \times \component_{1,c'}$ \Comment{RootSum}
        \State \Return $\component_{1,1}$
    \end{algorithmic}
\end{algorithm}

Here, $\nweight_{l,p,c,c'}(\cdot)$ denotes a neural network and $N_C$ denotes again the number of components. The input to the neural net are the values of the set of units $\component_{l, p-\nu_{l,p}{:}p-1, 1, 1{:}N_C}$. The notation $1{:}N_C$ denotes the range of components $[1,\dots,N_C]$ (including the first and last element). Similarly, $p{-}\nu_p{:}p{-}1$ denotes the range of partitions $[p{-}\nu_{l,p},\dots,p{-}1]$. The parameter $\nu_{l,p}$ is a hyperparameter and describes how many partitions \textit{`to the left'} are taken into consideration when computing the neural weights. For instance, in the partial computation graph in Figure~\ref{fig:neural_dependencies}, we have $\nu_{l,p}{=}2$. When setting $\nu_{l,p}{=}0$ we recover the special case of layered (non-conditional) probabilistic circuits.

Observe that it is those \textit{intra-layer} neural dependencies that induce an order on the random variables: we can only perform the computations of the units in the right branch of a partition tree if the values in the left branch are known. This holds recursively.
\begin{definition}
    Given a poset of random variables $\poset (\Xvars)$, we call a neural sum layer valid if the (partial) variable order it induces respects $\poset(\Xvars)$.
\end{definition}

Using the equations above to compute the leaf, product, sum, and root layers, we can also write down the pseudocode for marginal inference in layered PNCs, which we give in Algorithm~\ref{alg:layer}.
For ease of exposition we assume again that the number of variables $\Xvars$ is a power of two, and we refer to our implementation for the general case\footnote{\url{https://github.com/pedrozudo/ProbabilisticNeuralCircuits.git}}.

The algorithm takes as input a set of random variable instantiation $\xvars_o$ and a set of random variables $\Xvars_m$. The latter ought to be marginalized out, and the circuit evaluation computes the probability $p(\xvars_o)$.

The first step is to initialize the $\component_{p,c}$, which we do with the following function:
\begin{align}
    init(p,c,\xvars_o)
    =
    \begin{cases}
        f_{p.c}(\xvar) & \text{if $\xvar \in \xvars_o$  } \\
        1              & \text{otherwise}
    \end{cases}
\end{align}
Each combination of $p$ and $c$ corresponds to an index $k$ in Equation~\ref{eq:con:circuit}. Note that we forego in Algorithm~\ref{alg:layer} the possibility of introducing conditional dependencies in the leaves. The $1$'s in the second case result from marginalizing out the probability distributions in the leaves for the variables in $\Xvars_m$.

The algorithm then proceeds by first performing the computations in the leaf, before looping through the internal sum and product layers, and finishes with computing the root layer.
In contrast to the $4$-index notation introduced in Section~\ref{sec:prelim} we only use two indices here. This is because the identification of the layers happens implicitly using the $l$ counter of the while loop.

\subsection{Implementation Using Convolutions}

\begin{figure}[t]
    \begin{minipage}[c]{0.49\columnwidth}
        \centering
        % \resizebox{0.7\columnwidth}{!}{

        \begin{tikzpicture}[scale=0.4]
            % Grid lines
            \draw[step=1cm, gray, thin] (0,0) grid (5,1);

            \fill[gray!50] (0,0) rectangle (1,1);
            \fill[gray!80] (1,0) rectangle (2,1);
            \path (0, -0.5) -- (0, -1.5);

        \end{tikzpicture}
        % }
    \end{minipage}
    \begin{minipage}[c]{0.49\columnwidth}
        \centering
        \begin{tikzpicture}[scale=0.4]
            % Grid lines
            \draw[step=1cm, gray, thin] (0,0) grid (3,3);

            % Color specific cells
            \fill[gray!30] (0,2) rectangle (1,3);
            \fill[gray!50] (0 ,1) rectangle (1,2);
            \fill[gray!80] (1,2) rectangle (2,3);

        \end{tikzpicture}

    \end{minipage}

    \caption{Graphical representation of \textit{half kernels} used for neural sum layers in layered PNCs. On the left we see a kernel used for one-dimensional data while on the right we have a $3\times3$ kernel for two-dimensional data. The gray blocks indicate the learnable parameters of the \textit{half kernels}, while a white square indicates a parameter fixed to zero. Effectively, the convolutional layer is blind with regard to the inputs for these zero elements of the kernel.}
    \label{fig:kernel}

\end{figure}

An efficient way of implementing neural sum layers is by means of convolutional neural networks.
We can readily see this if we interpret the components within a layer as the channels of the convolutional neural network and the partitions as the input dimension over which we perform the convolution.

Assuming that our one-dimensional data is ordered left to right, we can make sure that this order is respected throughout the neural sum layers by using convolutional networks with a \textit{half kernel} as depicted in the left of Figure~\ref{fig:kernel}. The convoluted channels can then be passed on to further layers. As a final activation function we use a softmax layer that normalizes the outputs such that they sum up to 1, per partition that is.
Importantly, the number of output channels has to match the number of components in the summation.

\section{Related Work}
\label{sec:related}

A first attempt at relaxing decomposability in probabilistic circuits was made by \citet{sharir2018sum} with the introduction of sum-product-quotient networks (SPQNs). SPQNs introduce the quotient unit (in addition to sum and products), which encode conditional probabilities within a circuit. We show that introducing extra units is unnecessary as the same effect can be obtained by generalizing sum and product units while retaining two types of computation units.

On a theoretical level this allows us to forego the introduction of expendable concepts such as conditional soundness~\cite[Definition 5]{sharir2018sum} or conditional and effective scopes of a unit~\citep[Section 3]{sharir2018sum}.
% This enables the definition of CPC using standard concepts present in the probabilistic circuit literature.
% Furthermore, our formulation does not force us to regard conditional mixing (in neural sum units for instance) as quotients but allows us to parametrize the conditional distribution as we wish. Using CPC instead of SPQNs also establishes a direct link between probabilistic circuits and Bayesian networks (in form of hierarchical mixtures).

On the practical side we will show in Section~\ref{sec:experiments} that neural sum layers outperform the \textit{conditional mixing operator} (CMO) proposed as a building block for SPQNs~\citep[Definition 6]{sharir2018sum}. As a matter of fact, PNCs strictly generalize SPQNs constructed with CMOs.

\begin{corollary} CMO-SPQNs are PNCs.
    \label{cor:spq_worse}
\end{corollary}
\begin{proof}
    This can trivially be shown by picking $\nweight(\cdot)$ such that
    \begin{align}
         & \component_{l,p,2,c}     \tag{QuotientSumLayer}
        \\
         & =
        \frac{
            \sum_{c'=1}^{N_C} \weight_{l,p,c,c'}
            \times
            (\prod_{\component' \in  \component_{l, p-\nu_{l,p}:p-1, 1, c'}}\kappa' )
            \times
            \component_{l,p,1,c'}}
        {
            \sum_{c'=1}^{N_C}
            \weight_{l,p,c,c'}
            \times
            (\prod_{\component' \in  \component_{l, p-\nu_{l,p}:p-1, 1, c'}}\kappa' )
        }
        \nonumber
    \end{align}
    where we use our notation from Section~\ref{sec:pnc} to write down the CMO of~\citet{sharir2018sum}.
\end{proof}
Guided by Corollary~\ref{cor:spq_worse} we implemented CMO-SPQNs using a convolutional layer by simply fixing the non-zero elements of the kernel (\cf~Figure~\ref{fig:kernel}) to one. Performing the convolution on probabilities in log-space then simply corresponds to multiplying them in linear space. In this fashion we easily obtain the product present in the quotient sum layer.
An important difference between CMO-SPQNs and PNCs is that for the former components (or channels) must not mix with each other, which hinders their performance in terms of function approximation.

The limited expressive power of sum units was also noted by \citet{shao2022conditional}, which led them to introduce \textit{conditional sum-product networks}. The idea is to condition the weights of the sum units in a probabilistic circuit on the value of a random value (using neural networks). The main difference to our work is that these random variables live out-side of the circuits itself. Speaking in terms of Bayesian networks the approach of~\citet{shao2022conditional} is only capable of encoding a single Bayesian network while CPC encode hierarchical mixtures of Bayesian networks.

\section{Experimental Evaluation}
\label{sec:experiments}

% In our experimental evaluation we answer three questions:
% \begin{description}
%     \item[Q1] How do PNCs fair against PQCs and PCs on density estimation benchmarks?
%     \item[Q2] How do PNCs fair against state-of-the-art methods on density estimation benchmarks?
%     \item[Q3] Can PNCs be used for discriminative learning?
% \end{description}

For our experimental evaluation we used the MNIST family of dataset. That is, the original MNIST~\citep{deng2012mnist}, FashionMNIST~\citep{xiao2017fashion}, and also EMNIST~\citep{cohen2017emnist}.
We implemented PNCs (and also SPQNs) in PyTorch and Lightning\footnote{\url{https://lightning.ai/}}, and ran all our experiments on a DGX-2 machine with V100 Nvidia cards.

\subsection{How Do PNCs Fair Against PQCs and PCs?}
\label{sec:q1}

\begin{table*}[t]
    \footnotesize
    \centering
    \begin{tabular}{lccccccccc}
                          & PNC             & PQC    & PSC    & SHCLT  & HCLT   & CMC    & RAT-SPN & IDF     & BitSwap \\
        % \midrule
        \cmidrule(lr){2-4}
        \cmidrule(lr){5-10}
        MNIST             & $\mathbf{0.87}$ & $1.20$ & $1.32$ & $1.14$ & $1.20$ & $1.28$ & $1.67$  & $1.90$  & $1.27$  \\
        FashionMNIST      & $\mathbf{2.51}$ & $3.47$ & $3.66$ & $3.27$ & $3.34$ & $3.55$ & $4.29$  & $3.47$  & $3.28$  \\
        EMNIST (mnist)    & $\mathbf{1.36}$ & $1.84$ & $2.07$ & $1.52$ & $1.77$ & --     & $2.56$  & $2.07$  & $1.88$  \\
        EMNIST (letters)  & $\mathbf{1.33}$ & $1.83$ & $2.07$ & $1.58$ & $1.80$ & --     & $2.73$  & $1.95$  & $1.84$  \\
        EMNIST (balanced) & $\mathbf{1.35}$ & $1.86$ & $2.16$ & $1.60$ & $1.82$ & --     & $2.78$  & $2.15$  & $1.96$  \\
        EMNIST (byclass)  & $\mathbf{1.27}$ & $1.76$ & $2.02$ & $1.54$ & $1.85$ & --     & $2.72$  & $1.98$  & $1.87$  \\
        \midrule
        {\#} parameters   & $2.8M$          & $2.6M$ & $2.6M$ & $7.0M$ & $7.0M$ & $0.1M$ & $7.0M$  & $24.1M$ & $2.8M$  \\
    \end{tabular}
    \caption{Test set bpd for MNIST datasets (lower is better). The last row shows the number of parameters for each model (the symbol $M$ stands for millions).}
    \label{tab:mnist}
\end{table*}

\paragraph*{Setup}
To answer the first question we trained a PNC, an SPQNs and a PC on the MNIST family of datasets and minimize the negative log-likelihood. In order to render the three models commensurable, all three have the same underlying circuit structure. That is, we start with a grid of $28\times 28$ pixels and merge, in an alternating fashion, rows and columns. This corresponds to product nodes in the circuits. Between each merge we perform a summation. For probabilistic (sum) circuits (PSCs) this is the usual sum unit, for probabilistic quotient circuits (PQCs or SPQNs) we use the conditional mixing operator, and for PNCs we use a neural sum layer. The kernel type used for PQCs and PNCs is the one depicted in the right of Figure~\ref{fig:kernel}.
For the leaves, which encode the $28\times 28$ pixels, we use one categorical distribution with $256$ categories for each pixel. This allows us to represent all possible pixel values.
Within the circuits we used $12$ components per partition.
Merging rows and columns in an alternating fashion and using the kernel from Figure~\ref{fig:kernel} induces a specific variable ordering on a 2-dimensional grid, which we graphically represent in Figure~\ref{fig:2dorder}.

All three models were trained for $100$ epochs using Adam~\citep{kingma2014adam} with a learning rate of $0.001$ and a batch size of $50$.
The best model was selected using a $90-10$ train-validation data split where we monitored the negative log-likelihood on the validation set. We refer to the configuration files of the experiments for more details.

\paragraph*{Results}
We compare the three architectures using \textit{bits per dimension}, which are calculated from the average negative log-likelihood ($\overline{NLL}$) as follows: $bpd = \nicefrac{\overline{NLL}}{(\log{2} \times D)}$, here $D=28^2$ for MNIST datasets.

The results are reported in Table~\ref{tab:mnist} in the first three columns. We see that quotient circuits outperform sum circuits when it comes to minimizing the negative log-likelihood (minimizing $bpd$). We also see that neural circuits, with their data dependent weights, outperform both other methods. Note that PNCs and PQCs allow for the same set of tractable queries, while PNCs are more performant.

% \question{For the PNC structures used for 28x28 images, have the following questions:
% }

% \question{
% A. What is the missingness pattern that is allowed in the images so we still get tractable queries? Is there a intuitive way to see which missing patterns are allowed for the pixels?
% }
% \answer{Yes, there is an intuitive pattern. Consider the partition tree in Figure 3. We can then marginalize out the variables the right branch of the tree. For images this is similar but with a second dimension.}

\begin{figure}[t]

    % \begin{minipage}[c]{0.49\columnwidth}
    %     \centering

    %     \begin{tikzpicture}[scale=0.5] % Adjust the scale factor as needed
    %         % Define grid dimensions
    %         \def\gridSize{4}
    %         \def\cellSize{1} % Adjust the size of each cell

    %         % Draw the grid
    %         \draw[step=\cellSize,gray,very thin] (0,0) grid (\gridSize*\cellSize, \cellSize);

    %         % Manually label the cells with centered text
    %         \foreach \x/\label in {0.5/1,1.5/2,2.5/3,3.5/4} {
    %                 \node at (\x, 0.5) {\label};
    %             }
    %         \path (0, 0) -- (0, -1.1);
    %     \end{tikzpicture}
    % \end{minipage}
    % \begin{minipage}[c]{\columnwidth}
    %     \centering
    \centering
    \begin{tikzpicture}[scale=0.6]
        % Define grid dimensions
        \def\gridSize{4}
        \def\cellSize{1} % Adjust the size of each cell

        % Draw the grid
        \draw[step=\cellSize,gray,very thin] (0,0) grid (\gridSize*\cellSize, \gridSize*\cellSize);

        % Manually label the cells
        \node at (0.5, 3.5) {1};
        \node at (1.5, 3.5) {2};
        \node at (2.5, 3.5) {5};
        \node at (3.5, 3.5) {6};

        \node at (0.5, 2.5) {3};
        \node at (1.5, 2.5) {4};
        \node at (2.5, 2.5) {7};
        \node at (3.5, 2.5) {8};

        \node at (0.5, 1.5) {9};
        \node at (1.5, 1.5) {10};
        \node at (2.5, 1.5) {13};
        \node at (3.5, 1.5) {14};

        \node at (0.5, 0.5) {11};
        \node at (1.5, 0.5) {12};
        \node at (2.5, 0.5) {15};
        \node at (3.5, 0.5) {16};
    \end{tikzpicture}

    % \end{minipage}

    \caption{For simplicity's sake, assume that we have a $4\times4$ grid of pixels (instead of the $28\times 28$ MNIST grid). Recursively merging rows and columns and using the kernel from Figure~\ref{fig:kernel} then gives us the  total pixel order as indicated in the grid. We can then marginalize out pixels with higher numbers before pixels with lower numbers.
        For the specific case here we could marginalize out the lower half of the $4\times 4$ image, while retaining a probability distribution for the upper half.
    }
    \label{fig:2dorder}

\end{figure}

\subsection{How Do PNCs Fair Against State of the Art?}

\paragraph*{Setup} We use again the same PNCs as in Section~\ref{sec:q1} and compare them to (decomposable) probabilistic circuits from the literature: hidden Chow-Liu trees (HCLT)~\citep{liu2021tractable}, sparse HCLT (SHCLT)~\citep{dang2022sparse}, random sum-product networks (RAT-SPN)~\citep{peharz2019random}, and continuous mixture circuits (CMC)~\citep{correia2023continuous}. For completeness, we also include the bpd for IDF (a flow-based approach)~\citep{hoogeboom2019integer} and for BitSwap (a hierarchical latent variable model)~\citep{kingma2019bit}. Note that the results reported for RAT-SPN were taken from~\citep{dang2022sparse}.

\paragraph*{Results}
We see again that PNCs outperform the other methods in terms of bpd. On the one hand this is due to the increased expressive power of PNCs obtained by relaxing decomposability (and thereby losing tractability). On the other hand, this is also due to the fact that PNCs use neural sum units. This can be seen by comparing the results for PQC and SHCLT. While PQC are in theory more expressive than SHCLT we don't see this in practice: the learned structure of SHCLTs overcomes this expressivity gap. However, we see that PNCs manage to outperform SHCLTs.
Interestingly our implementation of PCs (PSC) does produce lower bpd than RAT-SPNs, suggesting that the latter are a rather weak baseline when it comes to density estimation on image data.

\subsection{Can PNCs Perform Discriminative Learning?}
\paragraph*{Setup}
Assume that we have a data point $\xvars$ for the random variables in $\Xvars$ and a label $y$ for this data point. Using Bayes rule we can rewrite the discriminative probability using generative distributions:
$
    p(Y{=}y{\mid} \Xvars{=}\xvars )
    {=}
    \frac{
        p(x{\mid} y)
    }
    {
        \sum_{z\in \Omega(Y)} p(x{\mid} y)
    }
$,
where we assume that the class prior is identical for each of the classes belonging to the sample space $\Omega(Y)$. This means that for every class in $\Omega(Y)$ we have a separate distribution. That is, a separate circuit. In the case of MNIST and FashionMNIST we have ten classes. The resulting ten circuits are jointly optimized using cross-entropy~\citep{peharz2019random}.
Architecturally, we used again $12$ components per partition and the $10$ different circuits share all parameters but the parameters in the leaf layer and root layer.

Furthermore, instead representing pixels as categorical random variables (with sample space $\{0,\dots, 255 \}$) we model them as continuous random variables with samples belong to the interval $[0,1]$. We obtain this value $v$ by dividing the pixel by $255$. Each leaf has then two inputs: the corresponding value $v$ of of the pixel itself and $1-v$.
This follows the protocol of~\citep[]{liang2019learning}.
Apart from optimizing the cross-entropy instead of the log-likelihood the training protocol was identical to the one for density estimation.

\paragraph*{Results}

We report the comparison in terms of classification accuracy on the test set, which we show in Table~\ref{tab:discrim}. We compare PNCs and PQCs to logistic circuits (LCs)~\citep{liang2019learning} and RAT-SPNs~\Citep{peharz2019random}. We see that PNCs perform better than PQC. However, neither reaches the accuracies of LCs nor RAT-SPNs. We hypothesize that this is due to a lack of regularization. For instance, PNCs reach perfect train accuracy on MNIST and near perfect train accuracy on FashionMNIST.
Furthermore, the authors of LCs and RAT-SPNs reported having used aggressive regularization techniques -- for the former on their Github page\footnote{\url{https://github.com/UCLA-StarAI/LogisticCircuit}} and the latter in~\Citep[Section 4.2]{peharz2019random}.
While we experimented with various regularization techniques such as weight decay~\Citep{loshchilov2018fixing} or the stochastic delta rule~\citep{hanson1990stochastic}, we were not able to obtain consistent improvements. We leave the study of effective regularization techniques for discriminative learning with PNCs for future work.

\begin{table}[t]
    \footnotesize
    \centering
    \begin{tabular}{lcccc}
                     & PNC     & PQC     & LC     & RAT-SPN  \\
        \cmidrule(lr){2-3}
        \cmidrule(lr){4-5}
        MNIST        & $98.04$ & $97.38$ & $99.4$ & $98.29$  \\
        FashionMNIST & $88.84$ & $87.63$ & $91.3$ & $89.89 $ \\
    \end{tabular}
    \caption{Test accuracies for MNIST and FashionMNIST.}
    \label{tab:discrim}
\end{table}

\section{Conclusions \& Future Work}

We first introduced the concept of a conditional probabilistic circuit, from which we were then able to construct probabilistic neural circuits, which generalize probabilistic circuits and sum-product-quotient networks. Note that the construction of PNCs would not have been possible from the formulation of \citet{sharir2018sum}. Furthermore, our formulation allows us to intuitively interpret PNCs as neural approximations of deep mixtures of Bayesian networks.

Experimentally, we have shown that for density estimation PNCs deliver on the promise made by SPQNs. That is, giving up on tractability improves function approximation in practice. For the discriminative case the situation is more nuanced. While PNCs achieve perfect accuracy on the training set a lack of proper regularization techniques prevents them from matching accuracies obtained by competing methods. We would also like to note that more sophisticated architecture designs for PNCs could possibly further improve their performance. For instance, using different numbers of components per partition dramatically increased the performance of SHCLT when compared to HCTL.

In future work we would like to explore the potential of PNCs for sampling, a task that probabilistic models usually struggle with~\citep{lang2022elevating} as good likelihood estimates do not correlate with sample quality~\citep{theis2016note}.
This would also establish a tighter link to autoregressive models (ARM) (\eg PixelCNN~\citep{van2016conditional}) and we might use ideas developed there for PNCs. In this regard, any-order ARM~\citep{uria2014deep,shih2022training} seem to be of particular interest to PNCs as this could allow for arbitrary conditioning sets in PNCs.

Other open questions concern structure learning for PNCs and applying them to tabular data or finding applications of PNCs -- an obvious candidate would be lossless compression with circuits~\citep{liu2022lossless}, as any-order marginalization is not necessary.

\section*{Acknowledgements}

This project received funding from the Wallenberg AI, Autonomous Systems and Software Program (WASP) funded by the Knut and Alice Wallenberg-Foundation, as well as from the TAILOR Connectivity Fund (part of the TAILOR project funded by the EU Horizon 2020 research and innovation program under GA No 952215).
The author would like to thank the anonymous reviewers for the valuable comments.

\bibliography{references}

\end{document}

### 2024-07-01 202402_aaai_neural_circuits/tikzlibraryspn.code.tex
all

\usetikzlibrary{chains}
%\usetikzlibrary{arrows}
\usepackage{pgfplots}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\pgfmathdeclarefunction{gauss}{2}{%
  \pgfmathparse{1/(#2*sqrt(2*pi))*exp(-((x-#1)^2)/(2*#2^2))}%
}

% int node style
\tikzstyle{int} = [circle,fill=white,draw=black,inner sep=0pt,line width=0.6pt,
minimum size=16pt]

% sum node style
\tikzstyle{sum} = [circle,fill=white,draw=black,inner sep=0pt,line width=0.6pt,
minimum size=16pt, path picture={\draw [shorten >=3pt,shorten <=3pt] (path picture bounding box.south) -- (path picture bounding box.north); \draw [shorten >=3pt,shorten <=3pt] (path picture bounding box.west) -- (path picture bounding box.east);}, node distance=1]

% product node style
\tikzstyle{prod} = [circle,fill=white,draw=black,inner sep=0pt,line width=0.6pt,
minimum size=16pt, path picture={\draw [shorten >=6pt,shorten <=6pt] (path picture bounding box.south east) -- (path picture bounding box.north west); \draw [shorten >=6pt,shorten <=6pt] (path picture bounding box.south west) -- (path picture bounding box.north east);}, node distance=1]

% \intnode [tikz-options] {int node name} 
\newcommand{\intnode}[2][]{
  \node[int, #1] (#2) {$\int$};
}

% \sumnode [tikz-options] {sum node name} 
\newcommand{\sumnode}[2][]{
  \node[sum, #1] (#2) {};
}

% \prodnode [tikz-options] {product node name} 
\newcommand{\prodnode}[2][]{
  \node[prod, #1] (#2) {$\times$};
}
\newcommand{\wprodnode}[3][]{
  \node[prod, #1] (#2) {$\color{#3}\times$};
}

% \maxnode [tikz-options] {max node name} 
\newcommand{\maxnode}[2][]{
  \node[circle,fill=white,draw=black,inner sep=0pt,line width=0.6pt,
  minimum size=16pt, node distance=1, #1] (#2) {\scriptsize $max$};
}

% \emptynode [tikz-options] {generic node name}
\newcommand{\emptynode}[2][]{
  \node[circle,fill=white,draw=black,inner sep=0pt,line width=0.6pt,
  minimum size=16pt, node distance=1, #1] (#2) {};
}

\newcommand{\varnode}[3][]{
  \node[circle,fill=white,draw=black,inner sep=0pt,line width=0.6pt, minimum size=16pt, font=\small, #1] (#2) {#3};
}

\newcommand{\varsqnode}[4][]{
  \node[rectangle, rounded corners=3pt,fill=white,draw=#4,inner sep=0pt,line width=0.6pt, minimum size=19pt, #1] (#2) {#3};
}

\newcommand{\evnode}[4][]{
  \node[circle,fill=#4,text=white,inner sep=0pt, minimum size=16pt, #1] (#2) {#3};
}

\newcommand{\evnodevar}[5][]{
  \node[circle,fill=#4,text=white,inner sep=0pt, minimum size=16pt,
  #1] (#2) {#3};
  \node[below=1pt of #2, xshift=1pt] (#2_vn) {#5};
}

\newcommand{\bernode}[3][]{
  \node[circle,fill=white,draw=black,inner sep=0pt,line width=0.6pt,
  minimum size=16pt, #1] (#2) {\tikz\draw circle (3pt);};
  \node[below=1pt of #2, xshift=1pt] (#2_vn) {#3};
}

\newcommand{\contnode}[3][]{
  \node[circle,fill=white,draw=black,inner sep=0pt,line width=0.6pt,
  minimum size=16pt, node distance=1, #1] (#2) {           
          \begin{tikzpicture} %these distributions look really distorted, and are the ones I want to include, inside the minipage
                \begin{axis}[
                no markers, scale only axis, domain=-4:4,samples=50, smooth,
                axis lines=none,
                height=10pt]
                \addplot [mark=none] {gauss(0,1)};
                \end{axis}
            \end{tikzpicture}};
  \node[below=1pt of #2, xshift=1pt] (#2_vn) {#3};
}

\newcommand{\catnode}[3][]{
  \node[circle,fill=white,draw=black,inner sep=0pt,line width=0.6pt,
  minimum size=16pt, node distance=1, #1] (#2) {           
          \begin{tikzpicture} %these distributions look really distorted, and are the ones I want to include, inside the minipage
                \begin{axis}[
                no markers, scale only axis, domain=-4:4, smooth,
                axis lines=none,
                height=10pt]
                \addplot [mark=none] coordinates { (0,0) (0,10)};
                \addplot [mark=none] coordinates { (-1,0) (-1,7)};
                \addplot [mark=none] coordinates { (-2,0) (-2,3)};
                \addplot [mark=none] coordinates { (1,0) (1,5)};
                \addplot [mark=none] coordinates { (2,0) (2,8)};
                \end{axis}
            \end{tikzpicture}};
  \node[below=1pt of #2] (#2_vn) {#3};
}

% \edge [options] {inputs} {outputs}
\newcommand{\edge}[3][]{ %
  % Connect all nodes #2 to all nodes #3.
  \foreach \x in {#2} { %
    \foreach \y in {#3} { %
      \path (\x) edge [-,line width=0.6pt,#1] (\y) ;%
    } ;
  } ;
}

\newcommand{\upedge}[3][]{
  \path (#2) edge [line width=1pt,->,>=stealth,#1] (#3);
}

\newcommand{\ledge}[4][]{
  \path (#2) edge [line width=1pt,->,>=stealth,#1,edge label=#4] (#3);
}

\newcommand{\weigedge}[4][]{
  \draw[line width=0.6pt,-,#1] (#2) edge node [midway,line width=0.6pt,#1] {#4} (#3) ;
}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2024-07-01 202409_nesy_neural_circuits/main.tex
all

--- 2024-03-05 aaai2024_neural_circuits/neural_circuits.tex
@@
-%File: anonymous-submission-latex-2024.tex
-\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
-\usepackage{aaai24}  % DO NOT CHANGE THIS
-\usepackage{times}  % DO NOT CHANGE THIS
-\usepackage{helvet}  % DO NOT CHANGE THIS
-\usepackage{courier}  % DO NOT CHANGE THIS
-\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
-\usepackage{graphicx} % DO NOT CHANGE THIS
-\urlstyle{rm} % DO NOT CHANGE THIS
-\def\UrlFont{\rm}  % DO NOT CHANGE THIS
-\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
-\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
-\frenchspacing  % DO NOT CHANGE THIS
-\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
-\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
-%
-% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
-\usepackage{algorithm}
-\usepackage{algpseudocode}
+%% The first command in your LaTeX source must be the \documentclass command.
+%%
+%% Options:
+%% twocolumn : Two column layout.
+%% hf: enable header and footer.
+\documentclass[
+	% twocolumn,
+	% hf,
+]{ceurart}
 
-%
-% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
-\usepackage{newfloat}
+%%
+%% One can fix some overfulls
+\sloppy
+
+%%
+%% Minted listings support 
+%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
 \usepackage{listings}
-\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
-\lstset{%
-	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
-	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
-	aboveskip=0pt,belowskip=0pt,%
-	showstringspaces=false,tabsize=2,breaklines=true}
-\floatstyle{ruled}
-\newfloat{listing}{tb}{lst}{}
-\floatname{listing}{Listing}
-%
-% Keep the \pdfinfo as shown here. There's no need
-% for you to add the /Title and /Author tags.
-\pdfinfo{
-/TemplateVersion (2024.1)
-}
-
-% DISALLOWED PACKAGES
-% \usepackage{authblk} -- This package is specifically forbidden
-% \usepackage{balance} -- This package is specifically forbidden
-% \usepackage{color (if used in text)
-% \usepackage{CJK} -- This package is specifically forbidden
-% \usepackage{float} -- This package is specifically forbidden
-% \usepackage{flushend} -- This package is specifically forbidden
-% \usepackage{fontenc} -- This package is specifically forbidden
-% \usepackage{fullpage} -- This package is specifically forbidden
-% \usepackage{geometry} -- This package is specifically forbidden
-% \usepackage{grffile} -- This package is specifically forbidden
-% \usepackage{hyperref} -- This package is specifically forbidden
-% \usepackage{navigator} -- This package is specifically forbidden
-% (or any other package that embeds links such as navigator or hyperref)
-% \indentfirst} -- This package is specifically forbidden
-% \layout} -- This package is specifically forbidden
-% \multicol} -- This package is specifically forbidden
-% \nameref} -- This package is specifically forbidden
-% \usepackage{savetrees} -- This package is specifically forbidden
-% \usepackage{setspace} -- This package is specifically forbidden
-% \usepackage{stfloats} -- This package is specifically forbidden
-% \usepackage{tabu} -- This package is specifically forbidden
-% \usepackage{titlesec} -- This package is specifically forbidden
-% \usepackage{tocbibind} -- This package is specifically forbidden
-% \usepackage{ulem} -- This package is specifically forbidden
-% \usepackage{wrapfig} -- This package is specifically forbidden
-% DISALLOWED COMMANDS
-% \nocopyright -- Your paper will not be published if you use this command
-% \addtolength -- This command may not be used
-% \balance -- This command may not be used
-% \baselinestretch -- Your paper will not be published if you use this command
-% \clearpage -- No page breaks of any kind may be used for the final version of your paper
-% \columnsep -- This command may not be used
-% \newpage -- No page breaks of any kind may be used for the final version of your paper
-% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
-% \pagestyle -- This command may not be used
-% \tiny -- This is not an acceptable font size.
-% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
-% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference
-
-\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.
-
-% The file aaai24.sty is the style file for AAAI Press
-% proceedings, working notes, and technical reports.
-%
-
-% Title
-
-\title{Probabilistic Neural Circuits}
-
-%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
-\author {
-    Pedro Zuidberg Dos Martires
-}
-\affiliations{
-    Centre for Applied Autonomous Sensor Systems (AASS), Örebro University, Sweden
-}
-
-% \author{}
-% \affiliations{}
+%% auto break lines
+\lstset{breaklines=true}
 
@@
 
+%%
+%% end of the preamble, start of the body of the document source.
 \begin{document}
 
-\maketitle
+%%
+%% Rights management information.
+%% CC-BY is default license.
+\copyrightyear{2022}
+\copyrightclause{Copyright for this paper by its authors.
+	Use permitted under Creative Commons License Attribution 4.0
+	International (CC BY 4.0).}
 
+%%
+%% This command is for the conference information
+\conference{18th International conference on
+	Neural-Symbolic Learning and Reasoning}
+
+%%
+%% The "title" command
+\title{Autoregressive Connections in Probabilistic Circuits}
+
+%%
+%% The "author" command and its associated commands are used to define
+%% the authors and their affiliations.
+\author{Pedro {Zuidberg Dos Martires}}
+\address{Örebro University, Sweden}
+
+%%
+%% The abstract is a short summary of the work to be presented in the
+%% article.
 \begin{abstract}
@@
 	Nevertheless, tractability comes at a cost: PCs are less expressive than neural networks.
-    In this paper we introduce probabilistic neural circuits (PNCs), which strike a balance between PCs and neural nets in terms of tractability and expressive power.
-Theoretically, we show that PNCs can be interpreted as deep mixtures of Bayesian networks.
-Experimentally, we demonstrate that PNCs constitute powerful function approximators.
-    % In this paper we explore This paper explores the interplay between PCs and neural networks, presenting the concept of Probabilistic Neural Circuits (PNCs) as an intermediary model.
-PNCs are introduced as deep mixtures of Bayesian networks, offering a structured approach to capturing complex dependencies in data.
+	In this extended abstract we briefly summarize the concept of a probabilistic neural circuit (PNC), which strikes a balance between PCs and neural nets in terms of tractability and expressive power, and distinguishes itself from ordinary PCs by the presence of autoregressive connections.
 \end{abstract}
+
+%%
+%% Keywords. The author(s) should pick words that accurately describe
+%% the work being presented. Separate the keywords with commas.
+\begin{keywords}
+	LaTeX class \sep
+	paper template \sep
+	paper formatting \sep
+	CEUR-WS
+\end{keywords}
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
+\maketitle
 
@@
 
-In recent years probabilistic circuits (PCs) (also called sum-product networks)~\citep{darwiche2003differential,poon2011sum} have emerged as an assembly language to talk about tractable probabilistic models and inference therein~\citep{vergari2021compositional}.
+In recent years probabilistic circuits (PCs) (also called sum-product networks)~\citep{darwiche2003differential} have emerged as an assembly language to talk about tractable probabilistic models and inference therein~\citep{vergari2021compositional}.
 The core idea is quite simple: we start with a set of independent random variables and construct complex probability distribution by recursively adding and multiplying them together.
@@
 
-Most of the recent advances in the field adhere to the second perspective: use an overparametrized probabilistic model and fit it to data using gradient based methods by leveraging discrete GPUs~\citep{peharz2019random,dang2021juice}.
-The computation units of such circuits are organized in a layered fashion.
-We give an example in Figure~\ref{fig:circuit}.
+A major advantage of PCs is their ability to answer certain queries in polynomial time -- given that adequate restrictions are imposed on a circuit's structure~\citep{vergari2021compositional}.
+An example of such a tractable query would be the computation of conditional probabilities for so-called \textit{smooth and decomposable} PCs~\citep{darwiche2001decomposable,darwiche2003differential}.
+This tractability, however, comes at a hefty price: the properties imposed on PCs in order to ensure polynomial time queries limit their expressive power~\citep{martens2014expressive,sharir2018sum}.
+In this extended abstract we give a short exposition of the ideas presented in~\citep{zuidberg2024probabilistic} where we used the concepts of \textit{conditional smoothness} and \textit{conditional decomposability}~\citep{sharir2018sum}
+to develop probabilistic neural circuits -- a novel probabilistic model class that trades-off tractability and expressive power.
+
+\section{Layered Probabilistic Neural Circuits}
+
+The key difference between (layered) probabilistic circuits ~\citep{peharz2020einsum} and probabilistic neural circuits is the presence of áutoregressive connections in the circuit.
+We illustrate this in Figure~\ref{fig:partition_graph}.
+It is precisely these autoregressive connections that lead to a huge boost in expressive power.
+We show this in our experimental study in~\citep{zuidberg2024probabilistic}.
+Note that this gain in expressive power comes at the cost of tractability: not all marginals of the joint distributions are computable in poly-time anymore.
 
 \begin{figure}[t]
+	\begin{minipage}[c]{0.49\linewidth}
 		\centering
@@
 		}
-    \caption{Layered probabilistic circuit following the construction of~\citep{shih2021hyperspns}.
-Data (modeled as random variables) is first fed into the leaf layer at the bottom.
-The output of the leaf layer is a mixture of distributions produced by the sum units.
-In the sum-product layer (Layer 2)  mixtures of random variables are combined by taking pairwise products, these are then again mixed using sum units.
-Finally, the root layer (at the top) gives us the joint probability distribution.
-The red edges indicate functional dependencies not present in traditional probabilistic circuits but present in probabilistic neural circuits.}
-    \label{fig:circuit}
-\end{figure}
-
-A major advantage of PCs is their ability to answer certain queries in polynomial time -- given that adequate restrictions are imposed on a circuit's structure~\citep{vergari2021compositional}.
-An example of such a tractable query would be the computation of conditional probabilities for so-called \textit{smooth and decomposable} PCs~\citep{darwiche2001decomposable,darwiche2003differential}.
-
-This tractability, however, comes at a hefty price: the properties imposed on PCs in order to ensure polynomial time queries limit their expressive power~\citep{martens2014expressive,sharir2018sum,zhang2021probabilistic}.
-This is in contrast to general neural networks and even sum-product networks with fewer structural constraints~\citep{delalleau2011shallow,kileel2019expressive}.
-\citet{martens2014expressive} have shown that decomposability is in fact a necessary condition for tractable marginal inference.
-
-Nevertheless, using the concepts of \textit{conditional smoothness} and \textit{conditional decomposability}~\citep{sharir2018sum}, we study in this paper the space of models that lie in between probabilistic circuits and neural networks.
-Concretely, we make the following contributions:
-\begin{enumerate}
-    \item We introduce conditional probabilistic circuits, from which we construct  probabilistic neural circuits (PNCs), which we interpret as deep mixtures of Bayesian nets.
-    \item  We provide a prescription to construct layered PNCs.
-    \item We provide an implementation of layered PNCs and experimentally study their expressive power.
-\end{enumerate}
-
-Our work is influenced by that of \citet{sharir2018sum}. We discuss the relationship to their approach (dubbed sum-product-quotient networks) in Section~\ref{sec:related}.
-
-\section{Preliminaries}
-\label{sec:prelim}
-
-In the remainder of the paper we will denote random variables by uppercase $\Xvar$'s, the corresponding values are lowercase $\xvar$'s.
-Sets of random variables and their corresponding values are typed in boldface: $\Xvars$ and $\xvars$, respectively.
-The definitions and notions we introduce in this section are loosely based on the work of~\citet{vergari2021compositional}.
-
-\begin{definition}[Probabilistic Circuit]
-    \label{def:circuit}
-    A probabilistic circuit over random variables $\Xvars$ is a parametrized computational graph encoding a probability density function $\circuit(\Xvars{=}\xvars)$.
-    The circuit consists of three kinds of computational units:
-    \textit{leaf}, \textit{product}, and \textit{sum}.
-    Each product or sum unit receives inputs from a set of input units denoted by $\inputs(n)$.
-    Each unit $k$ encodes a function $\circuit_{k}(\cdot)$ as follows:
-    \begin{align*}
-        {\circuit}_k(\xvars_n)=
-        \begin{cases}
-            f_k({\xvars_n})                                             & \text{if $k$ leaf unit}    \\
-            \circuit_{k_l}(\xvars_{n_l})  \circuit_{k_r} (\xvars_{n_r}) & \text{if $k$ product unit} \\
-            \sum_{j\in\inputs(k)} \weight_{kj} \circuit_j(\xvars_n)     & \text{if $k$ sum unit}
-        \end{cases}
-        % \label{eq:def:circuit}
-    \end{align*}
-    where $f_k(\xvars_n)$  denotes a parametrized probability distribution having as support the sample space of the random variables in $\Xvars_n$.
-\end{definition}
-
-\begin{definition}[Scope]
-    \label{def:scope:cond}
-    The scope of a unit $k$, denoted by $\scope(k)$, is the set of random variables $\Xvars_n$ for which the function $\circuit_k (\cdot)$ encodes a probability distribution.
-\end{definition}
-
-Two important properties that are usually imposed on probabilistic circuits are smoothness and decomposability as they allow for tractable queries, \eg computing marginals~\citep{darwiche2001decomposable,darwiche2003differential}.
-\begin{definition}[Smoothness]
-    A circuit is smooth if for every sum unit $k$ its inputs encode distributions over the same random variables:
-    $\forall j_1, j_2 {\in} \inputs(k)$ it holds that $\scope(j_1){=}\scope(j_2) $.
-\end{definition}
-
-\begin{definition}[Decomposability]
-    A circuit is decomposable if the inputs of every product unit
-    $k$ encode distributions over disjoint sets of random variables:
-    $\scope (k_l) \cap \scope (k_r) = \emptyset$ with $\{k_l, k_r\}= \inputs (k)$.
-\end{definition}
-
-\begin{definition}[Valid Probabilsitic Circuit]
-    We call a probabilistic circuit valid if for every unit $k$ we have that $p_k(\xvars_n){\geq} 0$ and $\int p_k(\xvars_n) \differential \xvars_n{=}1$.\footnote{Note that our notion of validity is slightly stricter than in its original definition, \cf \citep{poon2011sum}}
-\end{definition}
-
-As discussed by \citet{peharz2015theoretical}, probabilistic circuits are valid if they are smooth, decomposable, and that for the weights in the sum units we have $\weight_{kj} \in \mathbb{R}^+$ and $\sum_{j\in\inputs(k)} \weight_{kj} = 1$ for every $k$.
-Note that the notation in Definition~\ref{def:circuit} already suggests that the circuit is smooth as the inputs to the sum units are functions over the same set of variables $\Xvars_n$.
-
-Furthermore, we can assume, without loss of generality, that sum and product units occur in an alternating fashion in the circuit~\citep{peharz2020einsum}.
-This observation naturally leads to a layer-wise construction of probabilistic circuits where consecutive layers alternate between sum and product layers.
-Such layered probabilistic circuits~\citep{peharz2019random} have the advantage that the computations within a layer can be trivially parallelized.
-We can further abstract the layers in a circuit by fusing together alternating sums and products into a single sum-product layer~\citep{peharz2020einsum}.
-
-In Figure~\ref{fig:circuit} we give a graphical representation of a layered circuit.
-Layers consist of  blocks of computational units that are processed sequentially in a bottom-up fashion.
-Layers are themselves constituted of so-called partitions.
-The circuit in Figure~\ref{fig:circuit} has four partitions in the leaf layer, two in the sum-product layer, and a single partition in the root layer.
-By construction, partitions in the same layer have disjoint scopes.
-Moreover, partitions are further subdivided into input components and output components, which constitute the elemental computing units.
-The circuit in Figure~\ref{fig:circuit} has, except at the very bottom and top, three such input and output components in each partition.
-
-We can uniquely identify each computational unit (or component) in the circuit by specifying the layer, the partition, whether it is an input or an output, and its position within a partition.
-Counting layers from bottom to top, and partitions and units from left to right.
-Each component can be identified using 4 indices: $\component_{l,p,i,c}$.
-The first index $l$ identifies the layer, the second $p$ the partition, the third $i \in \{1,2 \}$ whether it is an input or output, and the fourth $c$ the horizontal position within a partition.
-For instance, the symbol $\component_{2223}$ corresponds to  the upper-right unit in the sum-product layer.
-
-\section{Conditional Probabilistic Circuits}
-\label{sec:cpc}
-
-We will first introduce the notion of posets (partially ordered sets) of random variables (Section~\ref{sec:poset}).
-This will allow us to generalize probabilistic circuits to conditional probabilistic circuits, which we interpret as deep mixtures of Bayesian networks (Section ~\ref{sec:dmbn}).
-We then introduce probabilistic neural circuits and their tractable queries (Section~\ref{sec:pnc}).
-
-\subsection{Partially Ordered Random Variables}
-\label{sec:poset}
-
-Consider a set of random variables $\Xvars {=}\{\Xvar_1,\dots, \Xvar_N \}$ on which we impose the \textit{parents} relationship $\parents(\cdot)$.
-The parents relationship induces a directed acyclic graph on the random variables $\Xvars$, where the nodes are the random variables themselves and an edge is present between $\Xvar_i$ and $\Xvar_j$ if $\Xvar_i {\in} \parents(\Xvar_j)$.
-This gives us a partial ordering of the variables $\Xvars$.
-We also define the ancestor relationship $\ancestors(\cdot)$ to be the transitive closure of $\parents(\cdot)$.
-That is, the ancestors of a random variable are its direct parents and recursively their parents.
-Furthermore, we denote the poset for the random variables $\Xvars$ by $\mathcal{O}(\Xvars)$. We say that the relationship $\Xvars_i {\sqsubset} \Xvars_j$ between two sets holds if
-$
-    \forall \Xvar_r {\in} \Xvars_i, \Xvar_q {\in} \Xvars_j : \Xvar_q {\notin}  \ancestors(\Xvars_r)
-$.
-We also define the relation $\Xvar_i\sqsubset \Xvar_j$ on random variables as $\{\Xvar_i \}{\sqsubset} \{\Xvar_j \}$.
-
-\begin{example}[Bayesian Network]
-    Partially ordered random variables induce a factorization of a joint probability distribution.
-A prominent example of such a factorization are Bayesian networks (\cf Figure ~\ref{fig:bn}) where we have:
-    \begin{talign}
-        \label{eq:bn}
-        p(\Xvars{=}\xvars) = \prod_{n:\Xvar_n \in \Xvars} p_n(X_n{=}x_n \mid \Xvars_{\parents(n)}{=}\xvars_{\parents(n)}).
-    \end{talign}
-\end{example}
-
-\begin{figure}
-
-    \begin{minipage}{0.50\linewidth}
-        \begin{align*}
-            X_1 \sqsubset X_2 &  & X_1 \sqsubset X_4 \\
-            X_1 \sqsubset X_3 &  & X_3 \sqsubset X_1 \\
-            X_3 \sqsubset X_2 &  & X_3 \sqsubset X_4 \\
-            X_2 \sqsubset X_3 &  & X_2 \sqsubset X_4 \\
-        \end{align*}
-
-    \end{minipage}%
-    \begin{minipage}{0.49\linewidth}
-        \resizebox{0.9\linewidth}{!}{
-
-            \centering
-
-            \begin{tikzpicture}
-                % Nodes
-                \node[draw, circle,ultra thick] (x3) {$X_3$};
-                \node[draw, circle, below=of x3,ultra thick] (x2) {$X_2$};
-                \node[draw, circle,  right=2cm of $(x3)!0.5!(x2)$3,ultra thick] (x4) {$X_4$};
-                \node[draw, circle, left=of x2,ultra thick] (x1) {$X_1$};
-
-                % Edges
-                \draw[->, ultra thick] (x1) -- (x2);
-                \draw[->,ultra thick] (x2) -- (x4);
-                \draw[->, ultra thick] (x3) -- (x4);
-            \end{tikzpicture}
-        }
 	\end{minipage}
-    \caption{Right: Bayesian network. Left: partial order relations that hold.}
-    \label{fig:bn}
-\end{figure}
-
-In the example above we denote $\parents(X_n)$ by $\Xvars_{\parents(n)}$.
-This will allow us to omit the random variable $\Xvar_n$ when writing down a probability distribution and only use the instantiation $\xvar_n$ instead.
-
-\subsection{Deep Mixtures of Bayesian Networks}
-\label{sec:dmbn}
-
-\begin{definition}[Conditinal Probabilsitic Circuit (CPC)]
-    \label{def:con:circuit}
-    A CPC $\circuit_k$ over a poset $\poset(\Xvars)$ is a parametrized computational graph encoding a probability density function $\circuit(\Xvars{=}\xvars)$.
-    The  CPC consists of
-    \textit{leaf}, \textit{product}, and \textit{sum} units.
-    Each unit $k$ encodes a function $\circuit_{k}$ as follows:
-    \begin{align}
-        \label{eq:con:circuit}
-         & {\circuit}_k(\xvars_n \mid  \xvars_{\parents(n)})
-        \\
-         & =
-        \begin{cases}
-            f_k({\xvars_n}\mid   \xvars_{\parents(n)})                                          & \text{if leaf}    \\
-            \circuit_{k_l}(\xvars_{n_l}\mid   \xvars_{\parents(n_l)})
-            \circuit_{k_r} (\xvars_{n_r}\mid \xvars_{\parents(n_r)})                            & \text{if product} \\
-            \sum_{j\in\inputs(k)} \weight_{kj} \circuit_j(\xvars_n  \mid  \xvars_{\parents(n)}) & \text{if sum}
-        \end{cases}
-        \nonumber
-        % \label{eq:def:circuitconditional}
-    \end{align}
-    where $f_k(\xvars_n)$ denotes a parametrized probability distribution having as support the sample space of the random variables in $\Xvars_n$.
-    % Furthermore, we necessitate that $p_n(\xvars_n){\geq} 0$ and $\int p_n(\xvars_n) \differential \xvars_n{=}1$ for every unit $n$.
-\end{definition}
-
-\begin{definition}[Scope (CPC)]
-    \label{def:scope}
-    The scope  $\scope(k)$ of a unit $k$ encoding the probability distribution $\circuit_k(\xvars_n\mid \xvars_{\parents(n)} )$ is the set of random variables $\Xvars_n$.
-\end{definition}
-
-\begin{corollary} A conditional probabilistic circuit over an unordered set of random variables is a (non-conditional) probability circuit.
-\end{corollary}
-\begin{proof}
-    Having no order means that $\parents(\Xvars)=\emptyset$ for every $\Xvar \in \Xvars$.
-This then means that the conditioning sets in Definition~\ref{def:con:circuit} are all empty and that we recover a (non-conditional) probabilistic circuit.
-\end{proof}
-
-We can now also introduce the notions of \textit{conditional smoothness} and \textit{conditional decomposability}:
-\begin{definition}[Conditional Smoothness]
-    A CPC is conditionally smooth if for every sum unit $k$ it holds that
-    $\forall j_1, j_2 {\in} \inputs(k):  \scope(j_1){=}\scope(j_2) $
-\end{definition}
-
-\begin{definition}[Conditional Decomposability]
-    A CPC is conditionally decomposable if for every product unit
-    $k$ it holds that $\scope (k_l) \cap \scope (k_r) = \emptyset$ with $\{k_l, k_r\}= \inputs (k)$.
-\end{definition}
-
-\begin{corollary}
-    A conditionally smooth and conditionally decomposable CPC is smooth and decomposable if its random variables $\Xvars$ are unordered.
-\end{corollary}
-
-\begin{definition}[Valid CPC]
-    \label{def:valid_cpc}
-    We call a CPC valid if for every unit $k$ we have that $p_k(\xvars_n {\mid} \xvars_{\parents(n)}){\geq} 0$ and $\int p_k(\xvars_n {\mid} \xvars_{\parents(n)}) \differential \xvars_n{=}1$.
-\end{definition}
-
-\begin{theorem}[Validity for CPCs]
-    A CPC is valid if it is conditionally smooth, conditionally decomposable, and for every sum unit $n$ it holds that $\weight_{nm}{\geq} 0$ and $\sum_{m\in\inputs(n)} \weight_{nm}{=}1$.
-\end{theorem}
-
-\begin{proof}
-    We start by rewriting the alternating sums and products of a CPC in its flat representation using the fact that products distribute over summations:
-    \begin{align}
-        \circuit_k(\xvars_n)
-         & =  \sum_{\tau \in \mathcal{T}} \weight_\tau  \prod_{\rho \in \tau} \rho (\xvars_\rho)
-        \label{eq:tree_decomp_PCP}
-    \end{align}
-    Here, $\mathcal{T}$ denotes the set of all products of leaf distributions in the CPC and $\rho \in \tau$ denotes a factor in one of these products
-    (we refer to \citep{zhao2016unified} for a more detailed account).
-
-    Invoking decomposability of the product units we have that each random variable $X_n$ only picks up a single factor $\rho (\xvar_\rho)$, which means that we can identify each $\rho (\xvar_\rho)$ with a specific $f_k(\xvar_n\mid \xvars_{\parents(n)} )$.
-This lets us rewrite Equation~\ref{eq:tree_decomp_PCP} as:
-    \begin{align}
-        \circuit_k(\xvars_n)
-         & =  \sum_{\tau \in \mathcal{T}} \weight_\tau  \prod_{k: f_k \in \tau} f_k (\xvars_n \mid \xvars_{\parents(n)})
-        \label{eq:f_tree_decomp_PCP}
-    \end{align}
-    Given that the $f_k$ are by definition (conditional) probability distributions their product forms a joint probability distribution as well.
-Next, we exploit smoothness, which states that the inputs to sum units have identical scope.
-This means that all the terms in the flat representation of $\circuit (\Xvars_n)$ mention the identical set of random variables and each term in the flat representation forms indeed a joint probability over the random variables $\Xvars_n$.
-
-    Lastly, \citet{peharz2015theoretical} have shown that having normalized weights in the sum units of a circuit results in normalized weights $\weight_\tau$ in the flat representation.
-This lets us conclude that the circuit $p_k(\xvars_n)$ is a probability distribution.
-Note that we did not make any reference in our reasoning to any specific unit in the circuit.
-This means that our argument holds for all units in a conditionally smooth and conditionally decomposable circuit with normalized weights, which means in turn that such a circuit is valid.
-\end{proof}
-
-In light of Equation~\ref{eq:f_tree_decomp_PCP} and comparing it to Equation~\ref{eq:bn}, we can interpret CPCs as deep (or hierarchical) mixtures of Bayesian networks.
-This is analogous to interpreting probabilistic circuits as deep mixtures of fully factorized distributions~\citep{poon2011sum}.
-
-\subsection{PNCs and Their Tractable Queries}
-\label{sec:pnc}
-
-The computational efficiency of probabilistic circuits stems from the fact the circuits evaluations are broken down into sub-evaluations, which are then cached and reused.
-Inspecting, however, the functional form of the sum units in a CPC (\cf Equation~\ref{eq:con:circuit}), this is not the case:
-each term in the sum over $j$ requires a separate conditional probability for each instantiation of the variables $\Xvars_{\parents(n)}$.
-This means that we would need (assuming binary random variables) $2^{|\Xvars_{\parents(n)}|}$ functions to encode the conditional probabilities.
-We alleviate this issue as follows. First, we rewrite the functional form of the sum units using Bayes' rule:
-\begin{align}
-    \sum_{j\in\inputs(k)}
-    \weight_{kj}
-    \frac{
-        \circuit_j(  \xvars_{\parents(n)} \mid \xvars_n  )
-    }{
-        \circuit_j(  \xvars_{\parents(n)  })
-    }
-    \circuit_j(  \xvars_n)
-\end{align}
-Second, we make the following approximation:
-\begin{align}
-    \weight_{kj}
-    \frac{
-        \circuit_j(  \xvars_{\parents(n)} \mid \xvars_n  )
-    }{
-        \circuit_j(  \xvars_{\parents(n)  })
-    }
-    \approx
-    \nweight_{kj}(\xvars_{\ancestors(n)}),
-\end{align}
-where $\nweight_{kj}(\cdot)$ is a neural network depending on the set of ancestors $\ancestors(\xvar_n)$. This now allows us to formally introduce probabilistic neural circuits.
-\begin{definition}[Probabilistic Neural Circuit (PNC)]
-    \label{def:pnc}
-    A PNC is a conditionally smooth and conditionally decomposable  CPC where sum units take the following functional form:
-    \begin{align}
-        {\circuit}_k(\xvars_n \mid  \xvars_{\parents(n)})
-        =
-        \sum_{j\in\inputs(k)} \nweight_{kj} ( \xvars_{\ancestors(n)}) \circuit_j(\xvars_n),
-        \label{eq:pnc}
-    \end{align}
-    with $\nweight_{kj}: \Omega( \ancestors(\xvar_n)) \rightarrow [0,1 ]$ being a neural network mapping from the sample space of the random variables $\ancestors(\xvar_n)$ to a real  between zero and one, and where it holds that $\sum_{j\in\inputs(k)} \nweight_{kj} = 1$.
-\end{definition}
-It can easily be seen that PNCs, already encode valid circuits (\cf. Definition~\ref{def:valid_cpc}) by construction.
-Intuitively, we interpret PNCs as neural approximations of CPCs.
-Note that it is this approximation that makes PNCs tractable: PNCs only need a single circuits for each $j$ while CPCs need a circuit for every $j$ and every instantiation of $\xvars_{\parents(n)}$.
-
-\begin{proposition}
-    Probabilistic circuits are PNCs.
-\end{proposition}
-\begin{proof}
-    If the values $\nweight(\xvars_{\ancestors(n)})$ do not depend on $\xvars_{\ancestors(n)}$ we have $|\inputs(k)|$ constants that sum up to $1$.
-This means that the weights in the sum units do not depend on the conditioning sets from the antecedent product layer, and we can omit any conditioning sets.
-The definition of a PNC in this case is then equivalent to the definition of a probabilistic circuit.
-\end{proof}
-
-Given the definition of PNCs we can now determine tractable queries that we can perform.
-
-\begin{proposition}[Density Evaluation]
-    Given a probabilistic circuit $p_k$ over the random variables $\Xvars$. We can evaluate the circuit  at the instantiation $\xvars$ in linear time with respect to the size of the circuit.
-\end{proposition}
-\begin{proof}
-    This follows simply from the fact that a circuit is a (non-recurrent) computation graph and that we can simply evaluate it by computing input units before output units.
-\end{proof}
-
-\begin{proposition}[Ordered Marginals]
-    \label{prop:ord_marg}
-    Consider a PNC $\circuit_k(\xvars_m, \xvars_e)$  we can then compute the marginal $\circuit_k(\xvars_e)$ in polynomial time if $\Xvars_e {\sqsubset} \Xvars_m$.
-\end{proposition}
-\begin{proof}
-    We start by writing out explicitly the single elements in the set $\xvars_m$:
-    \begin{align}
-        \circuit_k(\xvars_m, \xvars_e) = \circuit_k(\xvar_\mu, \dots, \xvar_1, \xvars_e)
-    \end{align}
-    where the order of $\{ \xvar_\mu, \dots, \xvar_1\}= \xvars_m$ is arbitrary but respects the partial order $\poset(\Xvars_m \cup \Xvars_e )$.
-    Given that the circuit is conditionally smooth and conditionally decomposable, we know that it encodes a proper probability distribution over its variables.
-    We can hence obtain the marginal $\circuit(\xvars_e)$ by integrating over the possible values $\xvars_m$:
-    \begin{align}
-        \circuit_k(\xvars_e)
-        =
-        \int \cdots \int p_k(\xvar_\mu, \dots, \xvar_1,\xvars_e)  \differential  \xvar_\mu \cdots \differential \xvar_1
-    \end{align}
-    As $\xvar_\mu$ does not appear in any of the conditioning sets and as any product unit $q$ is decomposable we can simply push the summation to the input unit $r$ of $q$ for which we have  $\Xvar_\mu \in \scope (r)$.
-    For summation units we exploit the linearity of the integral and distribute the integral over the terms in the sum.
-    Performing this recursively brings the integral to the leaves where we have  $ \int f_i(x_\mu \mid x_{\parents(\mu)} ) \differential \xvar_\mu =1$.
-
-    Up to this point the marginalization in CPC is identical to marginalization in probabilistic circuits.
-Contrary, to probabilistic circuits, however, we now need to propagate back up the result of the marginalization.
-Assuming, without loss of generality that leaf units feed into sum units, we then have
-    \begin{align*}
-         & \sum_{g\in\inputs(h)} \nweight_{hg} (\xvars_{\ancestors( \mu)})
-        \int \circuit_g (\xvar_\mu  ) \differential \xvar_\mu
-        % 
-        {=} \sum_{g\in\inputs(h)} \nweight_{hg} (\xvars_{\ancestors( \mu)})
-        {=}1
-    \end{align*}
-    The next product node that we encounter on our way up through the circuit is of the form:
-    \begin{align}
-         & \circuit_{l}(\xvars_{\mu-1}\mid   \xvars_{\parents(\mu-1)})
-        \int  \circuit_{r} (\xvars_{\mu}\mid \xvars_{\parents(\mu)}) \differential \xvars_\mu
-        \nonumber
-        \\
-         & =\circuit_{l}(\xvars_{\mu-1}\mid  \xvars_{\parents(\mu-1)})
-        \nonumber
-    \end{align}
-    At this point we have integrated out the variable $\xvar_\mu$ from the circuit by traversing a number of units linear in the size of the circuit (assuming proper caching~\citep{vergari2021compositional}).
-Repeating this procedure for the remaining set of ordered random variables $\{\Xvar_{\mu-1},\dots,\Xvar_1 \}$ gives us the distribution $\circuit_k(\xvars_e)$ in polynomial time.
-\end{proof}
-The proof follows a similar reasoning to the case of probabilistic circuits.
-The delicate point was to show that in the product unit one of the factors drops out.
-This is important as $\Xvar_{\parents(\mu)}$ might include $\Xvar_{\mu-1}$.
-Retaining such a dependency would prevent us from performing tractable ordered marginalization.
-
-\begin{corollary}[Ordered Conditionals]
-    Assuming a PNC $\circuit_k (\xvars_m, \xvars_o, \xvars_e)$ where $\Xvars_e \sqsubset \Xvars_o \sqsubset \Xvars_m$ holds lets us compute the conditional $p(\xvars_o \mid \xvars_e)$ in polynomial time.
-\end{corollary}
-\begin{proof}
-    We first apply the definition of the conditional probability:
-    $p(\xvars_o \mid \xvars_e)=\nicefrac{p(\xvars_o , \xvars_e)}{p( \xvars_e)}$.
-    Using the law of total probability we rewrite the ratio as
-    $$
-        p(\xvars_o \mid \xvars_e)
-        =
-        \frac
-        {
-            \int p(\xvars_m , \xvars_o, \xvars_e) \differential \xvars_m
-        }
-        {
-            \int p(\xvars_m , \xvars_o, \xvars_e) \differential \xvars_m \differential \xvars_o
-        }
-    $$
-    and Proposition~\ref{prop:ord_marg} tells us that we can perform both marginalizations in polytime.
-\end{proof}
-
-% \begin{definition}[Conditional Determinism]
-
-% \end{definition}
-
-\section{Layered Probabilistic Neural Circuits}
-
-While Equation~\ref{eq:pnc} provides a generic functional expression to compute the sum units in PNCs, it is not clear how to construct a PNC in the first place.
-That is, how do we link up the individual computation units such that they form a (valid) CPC.
-For (non-conditional) probabilistic circuits potent structure learning algorithms have been developed in recent years  (\eg  hidden Chow-Liu trees~\citep{liu2021tractable} or random probabilistic circuits~\citep{di2021random}).
-It is not entirely clear how to adapt these to the setting of conditional probabilistic circuits.
-For this reason we study problems where, informed by the structure of the data itself, a structure for a PNC can be constructed.
-Concretely, we will study PNC structures tailored towards image data: features, \ie pixels, that are close to each other should also be close to each other in the circuit -- a fact already exploited by~\citet{poon2011sum}.
-
-The structure we propose in this paper is rather simple and inspired by simple feed-forward neural networks and also layered (non-conditional) probabilistic circuits~\citep{peharz2020einsum}.
-More concretely,
-the computation units in the current layer only depend on computation units in the previous layer.
-Furthermore, we wish the units within each layer to be computable in parallel (given the previous layer).
-To construct such a probabilistic neural circuit we take the circuit structure introduced by \citet{shih2021hyperspns} as a backbone and add additional edges to the computation graph in order to obtain a PNC from a probabilistic circuit.
-For ease of exposition we detail our approach using one-dimensional data instead of two-dimensional data.
-
-\subsection{Structure for One-Dimensional Data}
-
-In order to study PNC structures, we introduce the concept of a partition graph, which is a hypergraph of a probabilistic circuit with the partitions of a circuit being the nodes and edges encoding the sub-partitions\footnote{In the circuit structure introduced by ~\citet{shih2021hyperspns}, all the product nodes decompose in the same fashion, \ie for product nodes with the same scope it is the sames variables that come from the left and right inputs, respectively.
-This is also called structured decomposabiltiy~\citep{darwiche2011sdd}.
-Partition trees are related to the concept of \textit{variable trees} in \textit{(probabilistic) sentential decision diagrams}~\citep{darwiche2011sdd,kisa2014probabilistic}.
-However, nodes in a variable tree do not constitute abstractions of a specific group of sum and product units and are a more general concept.
-}.
-We give an example of a partition tree in Figure~\ref{fig:partition_graph}.
-
-\begin{figure}[t]
-    \begin{minipage}[c]{\linewidth}
-        \centering
-
-        \begin{tikzpicture}[
-                dot/.style = {circle, draw, fill, minimum size=4pt, inner sep=0pt},
-                level distance=1cm,
-                level 1/.style={sibling distance=3cm},
-                level 2/.style={sibling distance=1.5cm},
-                level 3/.style={sibling distance=0.75cm},
-                scale=1
-            ]
-
-            \node[dot] (00) {}
-            child {node[dot] (10) {}
-                    child {node[dot] (20) {}
-                            child {node[dot] (30) {}}
-                            child {node[dot] (31) {}}
-                        }
-                    child {node[dot] (21) {}
-                            child {node[dot] (32) {}}
-                            child {node[dot] (33) {}}
-                        }
-                }
-            child {node[dot] (11) {}
-                    child {node[dot] (22) {}
-                            child {node[dot] (34) {}}
-                            child {node[dot] (35) {}}
-                        }
-                    child {node[dot] (23) {}
-                            child {node[dot] (36) {}}
-                            child {node[dot] (37) {}}
-                        }
-                };
-
-            \node[below= 0.1 of 30, inner sep=0pt] (X1) {\tiny{$X_1$}};
-            \node[below= 0.1 of 31, inner sep=0pt] (x2) {\tiny{$X_2$}};
-            \node[below= 0.1 of 32, inner sep=0pt] (x3) {\tiny{$X_3$}};
-            \node[below= 0.1 of 33, inner sep=0pt] (x4) {\tiny{$X_4$}};
-            \node[below= 0.1 of 34, inner sep=0pt] (X5) {\tiny{$X_5$}};
-            \node[below= 0.1 of 35, inner sep=0pt] (X6) {\tiny{$X_6$}};
-            \node[below= 0.1 of 36, inner sep=0pt] (X7) {\tiny{$X_7$}};
-            \node[below= 0.1 of 37, inner sep=0pt] (X8) {\tiny{$X_8$}};
-
-            \draw[<-, dashed, red] (11) -- (10);
-            \draw[<-, dashed, red] (23) -- (22);
-            \draw[<-, dashed, red] (22) -- (21);
-            \draw[<-, dashed, red] (21) -- (20);
-            \draw[<-, dashed, red, out=135] (23) to[out=150,in=30]  (21);
-            \draw[<-, dashed, red, out=135] (22) to[out=-150,in=-30]  (20);
-
-            % \draw[->, dashed, red] (37) -- (36);
-            % \draw[->, dashed, red] (36) -- (35);
-            % \draw[->, dashed, red] (35) -- (34);
-            % \draw[->, dashed, red] (34) -- (33);
-            % \draw[->, dashed, red] (33) -- (32);
-            % \draw[->, dashed, red] (32) -- (31);
-            % \draw[->, dashed, red] (31) -- (30);
-
-            % \draw[->, dashed, red] (37) to[out=150,in=30] (35);
-            % \draw[->, dashed, red] (36) to[out=-150,in=-30] (34);
-
-            % \draw[->, dashed, red] (35) to[out=150,in=30] (33);
-            % \draw[->, dashed, red] (34) to[out=-150,in=-30] (32);
-
-            % \draw[->, dashed, red] (33) to[out=150,in=30] (31);
-            % \draw[->, dashed, red] (32) to[out=-150,in=-30] (30);
-
-        \end{tikzpicture}
-    \end{minipage}
-    \caption{A balanced partition tree of a probabilistic neural circuit with eight variables.
-The partition tree (in black) describes how the variables decompose (in terms of the scope function).
-The edges in red indicate functional (neural) dependencies between partitions.}
-    \label{fig:partition_graph}
-\end{figure}
-
-Ignoring for now neural dependencies in PNCs (\ie the red edges in the partition tree in Figure~\ref{fig:partition_graph}) we describe the layer-wise operations.
-Let us assume, for the sake of simplicity, that the number of variables $N$ is a power of $2$.
-For instance, the circuits in Figure~\ref{fig:circuit} has $N=4$ and the partition diagram in Figure~\ref{fig:partition_graph} has $N=8$.
-Given that we merge partitions pairwise at each layer via multiplication, we obtain for the number of layers in a circuit $N_L= \log_2 N+1$.
-Formally, we express the layer-wise product units as follows:
-\begin{align}
-    \component_{l,p,1,c} = \component_{l-1,2p,2,c} \times \component_{l-1,2p+1,2,c},
-    \tag{ProductLayer}
-\end{align}
-which describes how to compute the value of a component given the components of the previous layer.
-The meaning of the indices is described in Section~\ref{sec:prelim}. Using this notation we can also express the sum units at the leaves and at the root:
-\begin{align}
-    \component_{1,p,2,c}
-     & =
-    \sum_{c'=1}^{N_D} \weight_{1,p,c,c'} \times \component_{1,p,1,c'}
-    \tag{SumLeaf}
-    \\
-    \component_{N_L,1,2,1}
-     & =
-    \sum_{c'=1}^{N_C} \weight_{N_L,1,1,c'} \times \component_{N_L,1,1,c'}
-    \tag{SumRoot}
-\end{align}
-In the equations above, $N_D$ denotes the number of initial components in the leaves and $N_C$ is the number of components throughout the circuits.
-For the circuit in Figure~\ref{fig:circuit} we have $N_D{=}2$ and $N_C{=}3$.
-The weights $w_{l,p,c.c'}$ are real valued constants and normalized over the $c'$ dimension. Note that, in contrast to the formulation in Section~\ref{sec:cpc},
-weights (and also computation units in general) are not identified by a single index (\eg $k$ in $\circuit_k(\cdot)$) but by four indices.
-To make this distinction explicit we denote the computation units by $\kappa_{l,p,i,c}$ instead of $p_k(\cdot)$.
-Observe also that the root layer has only a single component. Hence, the $1$ as the last index instead of $c$.
-
-Next, we describe the neural sum units in layered PNCs.
-To gain some intuition, consider the partial computation graph in Figure~\ref{fig:neural_dependencies}, where we show in more detail, compared to Figures ~\ref{fig:circuit} and~\ref{fig:partition_graph}, the neural dependencies present at the sum units.
-Formally, we express the values of neural sum units as follows:
-\begin{align}
-     & \component_{l,p,2,c}     \tag{NeuralSumLayer}
-    \\
-     & =
-    \sum_{c'=1}^{N_C} \nweight_{l,p,c,c'}(\component_{l, p-\nu_{l,p}:p-1, 1, 1:N_C})
-    \times
-    \component_{l,p,1,c'}
-    \nonumber
-\end{align}
-\begin{figure}[t]
-    \begin{minipage}[c]{\linewidth}
+	\begin{minipage}[c]{0.49\linewidth}
 		\centering
@@
 	\end{minipage}
-    \caption{Detailed graphical representation of neural dependencies in a PNC.
+
+	\caption{
+		On the left we have probabilistic neural circuit. The par tof the circuit that is drawn in black indicates the structure of an ordinary probabilistic circuit.
+		The red structure indicates the (neural) autoregressive connection that Separates a PNCs from these ordinary PCs.
+		On the right we give a more detailed graphical representation of neural dependencies in a PNC.
 The sum unit at the top outputs the weighted sum of the three product units at the bottom right.
 The weights for the sum are the outputs of a neural network for which it holds that $\sum_{i=1}^3=1$.
-They are computed using a neural network that takes as input the values of the six product units at the bottom left.}
-    \label{fig:neural_dependencies}
+	}
 
+	\label{fig:partition_graph}
 \end{figure}
-
-\begin{algorithm}[t]
-    \caption{Layer-wise circuit evaluation}
-    \label{alg:layer}
-    \hspace*{\algorithmicindent} \textbf{Input:} $\xvars_o$. $\Xvars_m$ \\
-    \hspace*{\algorithmicindent} \textbf{Output:} $p(\xvars_o)$ \\
-    \hspace*{\algorithmicindent} \textbf{Require:} $\Xvars_o {\cup} \Xvars_m {=} \Xvars$,  $\Xvars_o {\sqsubset} \Xvars_m$
-    \begin{algorithmic}[1]
-        \State $\component_{p,c} \gets init(p,c,\xvars_o)$
-        \State $\component_{p,c} \gets
-            \sum_{c'=1}^{N_D} \weight_{1,p,c,c'} \times \component_{p,c'}$ \Comment{LeafLayer}
-        \State $l \gets 2$
-        \While {$l< N_L$}
-        \State $\component_{p,c} \gets \component_{2p,c} \times \component_{2p+1,c} $ \Comment{ProductLayer}
-        \State
-        $ \component_{p,c}
-            \gets
-            \sum_{c'=1}^{N_C} \nweight_{l,p,c,c'}
-            \times
-            \component_{p,c'}
-        $ \Comment{NeuralSumLayer}
-        \State $l \gets l+1$
-        \EndWhile
-        \State $\component_{p,c} \gets \component_{2p,c} \times \component_{2p+1,c} $ \Comment{ProductLayer}
-        \State $\component_{1,1} \gets \sum_{c'=1}^{N_C} \weight_{N_L,1,c,c'} \times \component_{1,c'}$ \Comment{RootSum}
-        \State \Return $\component_{1,1}$
-    \end{algorithmic}
-\end{algorithm}
-
-Here, $\nweight_{l,p,c,c'}(\cdot)$ denotes a neural network and $N_C$ denotes again the number of components.
-The input to the neural net are the values of the set of units $\component_{l, p-\nu_{l,p}{:}p-1, 1, 1{:}N_C}$.
-The notation $1{:}N_C$ denotes the range of components $[1,\dots,N_C]$ (including the first and last element).
-Similarly, $p{-}\nu_p{:}p{-}1$ denotes the range of partitions $[p{-}\nu_{l,p},\dots,p{-}1]$.
-The parameter $\nu_{l,p}$ is a hyperparameter and describes how many partitions \textit{`to the left'} are taken into consideration when computing the neural weights.
-For instance, in the partial computation graph in Figure~\ref{fig:neural_dependencies}, we have $\nu_{l,p}{=}2$.
-When setting $\nu_{l,p}{=}0$ we recover the special case of layered (non-conditional) probabilistic circuits.
-
-Observe that it is those \textit{intra-layer} neural dependencies that induce an order on the random variables: we can only perform the computations of the units in the right branch of a partition tree if the values in the left branch are known.
-This holds recursively.
-\begin{definition}
-    Given a poset of random variables $\poset (\Xvars)$, we call a neural sum layer valid if the (partial) variable order it induces respects $\poset(\Xvars)$.
-\end{definition}
-
-Using the equations above to compute the leaf, product, sum, and root layers, we can also write down the pseudocode for marginal inference in layered PNCs, which we give in Algorithm~\ref{alg:layer}.
-For ease of exposition we assume again that the number of variables $\Xvars$ is a power of two, and we refer to our implementation for the general case\footnote{\url{https://github.com/pedrozudo/ProbabilisticNeuralCircuits.git}}.
-
-The algorithm takes as input a set of random variable instantiation $\xvars_o$ and a set of random variables $\Xvars_m$.
-The latter ought to be marginalized out, and the circuit evaluation computes the probability $p(\xvars_o)$.
-
-The first step is to initialize the $\component_{p,c}$, which we do with the following function:
-\begin{align}
-    init(p,c,\xvars_o)
-    =
-    \begin{cases}
-        f_{p.c}(\xvar) & \text{if $\xvar \in \xvars_o$  } \\
-        1              & \text{otherwise}
-    \end{cases}
-\end{align}
-Each combination of $p$ and $c$ corresponds to an index $k$ in Equation~\ref{eq:con:circuit}.
-Note that we forego in Algorithm~\ref{alg:layer} the possibility of introducing conditional dependencies in the leaves.
-The $1$'s in the second case result from marginalizing out the probability distributions in the leaves for the variables in $\Xvars_m$.
-
-The algorithm then proceeds by first performing the computations in the leaf, before looping through the internal sum and product layers, and finishes with computing the root layer.
-In contrast to the $4$-index notation introduced in Section~\ref{sec:prelim} we only use two indices here.
-This is because the identification of the layers happens implicitly using the $l$ counter of the while loop.
-
-\subsection{Implementation Using Convolutions}
-
-\begin{figure}[t]
-    \begin{minipage}[c]{0.49\columnwidth}
-        \centering
-        % \resizebox{0.7\columnwidth}{!}{
-
-        \begin{tikzpicture}[scale=0.4]
-            % Grid lines
-            \draw[step=1cm, gray, thin] (0,0) grid (5,1);
-
-            \fill[gray!50] (0,0) rectangle (1,1);
-            \fill[gray!80] (1,0) rectangle (2,1);
-            \path (0, -0.5) -- (0, -1.5);
-
-        \end{tikzpicture}
-        % }
-    \end{minipage}
-    \begin{minipage}[c]{0.49\columnwidth}
-        \centering
-        \begin{tikzpicture}[scale=0.4]
-            % Grid lines
-            \draw[step=1cm, gray, thin] (0,0) grid (3,3);
-
-            % Color specific cells
-            \fill[gray!30] (0,2) rectangle (1,3);
-            \fill[gray!50] (0 ,1) rectangle (1,2);
-            \fill[gray!80] (1,2) rectangle (2,3);
-
-        \end{tikzpicture}
-
-    \end{minipage}
-
-    \caption{Graphical representation of \textit{half kernels} used for neural sum layers in layered PNCs.
-On the left we see a kernel used for one-dimensional data while on the right we have a $3\times3$ kernel for two-dimensional data.
-The gray blocks indicate the learnable parameters of the \textit{half kernels}, while a white square indicates a parameter fixed to zero.
-Effectively, the convolutional layer is blind with regard to the inputs for these zero elements of the kernel.}
-    \label{fig:kernel}
-
-\end{figure}
-
-An efficient way of implementing neural sum layers is by means of convolutional neural networks.
-We can readily see this if we interpret the components within a layer as the channels of the convolutional neural network and the partitions as the input dimension over which we perform the convolution.
-
-Assuming that our one-dimensional data is ordered left to right, we can make sure that this order is respected throughout the neural sum layers by using convolutional networks with a \textit{half kernel} as depicted in the left of Figure~\ref{fig:kernel}.
-The convoluted channels can then be passed on to further layers.
-As a final activation function we use a softmax layer that normalizes the outputs such that they sum up to 1, per partition that is.
-Importantly, the number of output channels has to match the number of components in the summation.
-
-\section{Related Work}
-\label{sec:related}
-
-A first attempt at relaxing decomposability in probabilistic circuits was made by \citet{sharir2018sum} with the introduction of sum-product-quotient networks (SPQNs).
-SPQNs introduce the quotient unit (in addition to sum and products), which encode conditional probabilities within a circuit.
-We show that introducing extra units is unnecessary as the same effect can be obtained by generalizing sum and product units while retaining two types of computation units.
-
-On a theoretical level this allows us to forego the introduction of expendable concepts such as conditional soundness~\cite[Definition 5]{sharir2018sum} or conditional and effective scopes of a unit~\citep[Section 3]{sharir2018sum}.
-% This enables the definition of CPC using standard concepts present in the probabilistic circuit literature.
-% Furthermore, our formulation does not force us to regard conditional mixing (in neural sum units for instance) as quotients but allows us to parametrize the conditional distribution as we wish.
-Using CPC instead of SPQNs also establishes a direct link between probabilistic circuits and Bayesian networks (in form of hierarchical mixtures).
-
-On the practical side we will show in Section~\ref{sec:experiments} that neural sum layers outperform the \textit{conditional mixing operator} (CMO) proposed as a building block for SPQNs~\citep[Definition 6]{sharir2018sum}.
-As a matter of fact, PNCs strictly generalize SPQNs constructed with CMOs.
-
-\begin{corollary} CMO-SPQNs are PNCs.
-    \label{cor:spq_worse}
-\end{corollary}
-\begin{proof}
-    This can trivially be shown by picking $\nweight(\cdot)$ such that
-    \begin{align}
-         & \component_{l,p,2,c}     \tag{QuotientSumLayer}
-        \\
-         & =
-        \frac{
-            \sum_{c'=1}^{N_C} \weight_{l,p,c,c'}
-            \times
-            (\prod_{\component' \in  \component_{l, p-\nu_{l,p}:p-1, 1, c'}}\kappa' )
-            \times
-            \component_{l,p,1,c'}}
-        {
-            \sum_{c'=1}^{N_C}
-            \weight_{l,p,c,c'}
-            \times
-            (\prod_{\component' \in  \component_{l, p-\nu_{l,p}:p-1, 1, c'}}\kappa' )
-        }
-        \nonumber
-    \end{align}
-    where we use our notation from Section~\ref{sec:pnc} to write down the CMO of~\citet{sharir2018sum}.
-\end{proof}
-Guided by Corollary~\ref{cor:spq_worse} we implemented CMO-SPQNs using a convolutional layer by simply fixing the non-zero elements of the kernel (\cf~Figure~\ref{fig:kernel}) to one.
-Performing the convolution on probabilities in log-space then simply corresponds to multiplying them in linear space.
-In this fashion we easily obtain the product present in the quotient sum layer.
-An important difference between CMO-SPQNs and PNCs is that for the former components (or channels) must not mix with each other, which hinders their performance in terms of function approximation.
-
-The limited expressive power of sum units was also noted by \citet{shao2022conditional}, which led them to introduce \textit{conditional sum-product networks}.
-The idea is to condition the weights of the sum units in a probabilistic circuit on the value of a random value (using neural networks).
-The main difference to our work is that these random variables live out-side of the circuits itself.
-Speaking in terms of Bayesian networks the approach of~\citet{shao2022conditional} is only capable of encoding a single Bayesian network while CPC encode hierarchical mixtures of Bayesian networks.
-
-\section{Experimental Evaluation}
-\label{sec:experiments}
-
-% In our experimental evaluation we answer three questions:
-% \begin{description}
-%     \item[Q1] How do PNCs fair against PQCs and PCs on density estimation benchmarks?
-%     \item[Q2] How do PNCs fair against state-of-the-art methods on density estimation benchmarks?
-%     \item[Q3] Can PNCs be used for discriminative learning?
-% \end{description}
-
-For our experimental evaluation we used the MNIST family of dataset. That is, the original MNIST~\citep{deng2012mnist}, FashionMNIST~\citep{xiao2017fashion}, and also EMNIST~\citep{cohen2017emnist}.
-We implemented PNCs (and also SPQNs) in PyTorch and Lightning\footnote{\url{https://lightning.ai/}}, and ran all our experiments on a DGX-2 machine with V100 Nvidia cards.
-
-\subsection{How Do PNCs Fair Against PQCs and PCs?}
-\label{sec:q1}
-
-\begin{table*}[t]
-    \footnotesize
-    \centering
-    \begin{tabular}{lccccccccc}
-                          & PNC             & PQC    & PSC    & SHCLT  & HCLT   & CMC    & RAT-SPN & IDF     & BitSwap \\
-        % \midrule
-        \cmidrule(lr){2-4}
-        \cmidrule(lr){5-10}
-        MNIST             & $\mathbf{0.87}$ & $1.20$ & $1.32$ & $1.14$ & $1.20$ & $1.28$ & $1.67$  & $1.90$  & $1.27$  \\
-        FashionMNIST      & $\mathbf{2.51}$ & $3.47$ & $3.66$ & $3.27$ & $3.34$ & $3.55$ & $4.29$  & $3.47$  & $3.28$  \\
-        EMNIST (mnist)    & $\mathbf{1.36}$ & $1.84$ & $2.07$ & $1.52$ & $1.77$ & --     & $2.56$  & $2.07$  & $1.88$  \\
-        EMNIST (letters)  & $\mathbf{1.33}$ & $1.83$ & $2.07$ & $1.58$ & $1.80$ & --     & $2.73$  & $1.95$  & $1.84$  \\
-        EMNIST (balanced) & $\mathbf{1.35}$ & $1.86$ & $2.16$ & $1.60$ & $1.82$ & --     & $2.78$  & $2.15$  & $1.96$  \\
-        EMNIST (byclass)  & $\mathbf{1.27}$ & $1.76$ & $2.02$ & $1.54$ & $1.85$ & --     & $2.72$  & $1.98$  & $1.87$  \\
-        \midrule
-        {\#} parameters   & $2.8M$          & $2.6M$ & $2.6M$ & $7.0M$ & $7.0M$ & $0.1M$ & $7.0M$  & $24.1M$ & $2.8M$  \\
-    \end{tabular}
-    \caption{Test set bpd for MNIST datasets (lower is better). The last row shows the number of parameters for each model (the symbol $M$ stands for millions).}
-    \label{tab:mnist}
-\end{table*}
-
-\paragraph*{Setup}
-To answer the first question we trained a PNC, an SPQNs and a PC on the MNIST family of datasets and minimize the negative log-likelihood.
-In order to render the three models commensurable, all three have the same underlying circuit structure.
-That is, we start with a grid of $28\times 28$ pixels and merge, in an alternating fashion, rows and columns.
-This corresponds to product nodes in the circuits.
-Between each merge we perform a summation.
-For probabilistic (sum) circuits (PSCs) this is the usual sum unit, for probabilistic quotient circuits (PQCs or SPQNs) we use the conditional mixing operator, and for PNCs we use a neural sum layer.
-The kernel type used for PQCs and PNCs is the one depicted in the right of Figure~\ref{fig:kernel}.
-For the leaves, which encode the $28\times 28$ pixels, we use one categorical distribution with $256$ categories for each pixel. This allows us to represent all possible pixel values.
-Within the circuits we used $12$ components per partition.
-Merging rows and columns in an alternating fashion and using the kernel from Figure~\ref{fig:kernel} induces a specific variable ordering on a 2-dimensional grid, which we graphically represent in Figure~\ref{fig:2dorder}.
-
-All three models were trained for $100$ epochs using Adam~\citep{kingma2014adam} with a learning rate of $0.001$ and a batch size of $50$.
-The best model was selected using a $90-10$ train-validation data split where we monitored the negative log-likelihood on the validation set.
-We refer to the configuration files of the experiments for more details.
-
-\paragraph*{Results}
-We compare the three architectures using \textit{bits per dimension}, which are calculated from the average negative log-likelihood ($\overline{NLL}$) as follows: $bpd = \nicefrac{\overline{NLL}}{(\log{2} \times D)}$, here $D=28^2$ for MNIST datasets.
-
-The results are reported in Table~\ref{tab:mnist} in the first three columns.
-We see that quotient circuits outperform sum circuits when it comes to minimizing the negative log-likelihood (minimizing $bpd$).
-We also see that neural circuits, with their data dependent weights, outperform both other methods.
-Note that PNCs and PQCs allow for the same set of tractable queries, while PNCs are more performant.
-
-% \question{For the PNC structures used for 28x28 images, have the following questions:
-% }
-
-% \question{
-% A. What is the missingness pattern that is allowed in the images so we still get tractable queries? Is there a intuitive way to see which missing patterns are allowed for the pixels?
-% }
-% \answer{Yes, there is an intuitive pattern.
-Consider the partition tree in Figure 3.
-We can then marginalize out the variables the right branch of the tree.
-For images this is similar but with a second dimension.}
-
-\begin{figure}[t]
-
-    % \begin{minipage}[c]{0.49\columnwidth}
-    %     \centering
-
-    %     \begin{tikzpicture}[scale=0.5] % Adjust the scale factor as needed
-    %         % Define grid dimensions
-    %         \def\gridSize{4}
-    %         \def\cellSize{1} % Adjust the size of each cell
-
-    %         % Draw the grid
-    %         \draw[step=\cellSize,gray,very thin] (0,0) grid (\gridSize*\cellSize, \cellSize);
-
-    %         % Manually label the cells with centered text
-    %         \foreach \x/\label in {0.5/1,1.5/2,2.5/3,3.5/4} {
-    %                 \node at (\x, 0.5) {\label};
-    %             }
-    %         \path (0, 0) -- (0, -1.1);
-    %     \end{tikzpicture}
-    % \end{minipage}
-    % \begin{minipage}[c]{\columnwidth}
-    %     \centering
-    \centering
-    \begin{tikzpicture}[scale=0.6]
-        % Define grid dimensions
-        \def\gridSize{4}
-        \def\cellSize{1} % Adjust the size of each cell
-
-        % Draw the grid
-        \draw[step=\cellSize,gray,very thin] (0,0) grid (\gridSize*\cellSize, \gridSize*\cellSize);
-
-        % Manually label the cells
-        \node at (0.5, 3.5) {1};
-        \node at (1.5, 3.5) {2};
-        \node at (2.5, 3.5) {5};
-        \node at (3.5, 3.5) {6};
-
-        \node at (0.5, 2.5) {3};
-        \node at (1.5, 2.5) {4};
-        \node at (2.5, 2.5) {7};
-        \node at (3.5, 2.5) {8};
-
-        \node at (0.5, 1.5) {9};
-        \node at (1.5, 1.5) {10};
-        \node at (2.5, 1.5) {13};
-        \node at (3.5, 1.5) {14};
-
-        \node at (0.5, 0.5) {11};
-        \node at (1.5, 0.5) {12};
-        \node at (2.5, 0.5) {15};
-        \node at (3.5, 0.5) {16};
-    \end{tikzpicture}
-
-    % \end{minipage}
-
-    \caption{For simplicity's sake, assume that we have a $4\times4$ grid of pixels (instead of the $28\times 28$ MNIST grid).
-Recursively merging rows and columns and using the kernel from Figure~\ref{fig:kernel} then gives us the  total pixel order as indicated in the grid.
-We can then marginalize out pixels with higher numbers before pixels with lower numbers.
-        For the specific case here we could marginalize out the lower half of the $4\times 4$ image, while retaining a probability distribution for the upper half.
-    }
-    \label{fig:2dorder}
-
-\end{figure}
-
-\subsection{How Do PNCs Fair Against State of the Art?}
-
-\paragraph*{Setup} We use again the same PNCs as in Section~\ref{sec:q1} and compare them to (decomposable) probabilistic circuits from the literature: hidden Chow-Liu trees (HCLT)~\citep{liu2021tractable}, sparse HCLT (SHCLT)~\citep{dang2022sparse}, random sum-product networks (RAT-SPN)~\citep{peharz2019random}, and continuous mixture circuits (CMC)~\citep{correia2023continuous}.
-For completeness, we also include the bpd for IDF (a flow-based approach)~\citep{hoogeboom2019integer} and for BitSwap (a hierarchical latent variable model)~\citep{kingma2019bit}.
-Note that the results reported for RAT-SPN were taken from~\citep{dang2022sparse}.
-
-\paragraph*{Results}
-We see again that PNCs outperform the other methods in terms of bpd.
-On the one hand this is due to the increased expressive power of PNCs obtained by relaxing decomposability (and thereby losing tractability).
-On the other hand, this is also due to the fact that PNCs use neural sum units.
-This can be seen by comparing the results for PQC and SHCLT.
-While PQC are in theory more expressive than SHCLT we don't see this in practice: the learned structure of SHCLTs overcomes this expressivity gap.
-However, we see that PNCs manage to outperform SHCLTs.
-Interestingly our implementation of PCs (PSC) does produce lower bpd than RAT-SPNs, suggesting that the latter are a rather weak baseline when it comes to density estimation on image data.
-
-\subsection{Can PNCs Perform Discriminative Learning?}
-\paragraph*{Setup}
-Assume that we have a data point $\xvars$ for the random variables in $\Xvars$ and a label $y$ for this data point.
-Using Bayes rule we can rewrite the discriminative probability using generative distributions:
-$
-    p(Y{=}y{\mid} \Xvars{=}\xvars )
-    {=}
-    \frac{
-        p(x{\mid} y)
-    }
-    {
-        \sum_{z\in \Omega(Y)} p(x{\mid} y)
-    }
-$,
-where we assume that the class prior is identical for each of the classes belonging to the sample space $\Omega(Y)$.
-This means that for every class in $\Omega(Y)$ we have a separate distribution.
-That is, a separate circuit.
-In the case of MNIST and FashionMNIST we have ten classes.
-The resulting ten circuits are jointly optimized using cross-entropy~\citep{peharz2019random}.
-Architecturally, we used again $12$ components per partition and the $10$ different circuits share all parameters but the parameters in the leaf layer and root layer.
-
-Furthermore, instead representing pixels as categorical random variables (with sample space $\{0,\dots, 255 \}$) we model them as continuous random variables with samples belong to the interval $[0,1]$.
-We obtain this value $v$ by dividing the pixel by $255$.
-Each leaf has then two inputs: the corresponding value $v$ of of the pixel itself and $1-v$.
-This follows the protocol of~\citep[]{liang2019learning}.
-Apart from optimizing the cross-entropy instead of the log-likelihood the training protocol was identical to the one for density estimation.
-
-\paragraph*{Results}
-
-We report the comparison in terms of classification accuracy on the test set, which we show in Table~\ref{tab:discrim}.
-We compare PNCs and PQCs to logistic circuits (LCs)~\citep{liang2019learning} and RAT-SPNs~\Citep{peharz2019random}.
-We see that PNCs perform better than PQC.
-However, neither reaches the accuracies of LCs nor RAT-SPNs.
-We hypothesize that this is due to a lack of regularization.
-For instance, PNCs reach perfect train accuracy on MNIST and near perfect train accuracy on FashionMNIST.
-Furthermore, the authors of LCs and RAT-SPNs reported having used aggressive regularization techniques -- for the former on their Github page\footnote{\url{https://github.com/UCLA-StarAI/LogisticCircuit}} and the latter in~\Citep[Section 4.2]{peharz2019random}.
-While we experimented with various regularization techniques such as weight decay~\Citep{loshchilov2018fixing} or the stochastic delta rule~\citep{hanson1990stochastic}, we were not able to obtain consistent improvements.
-We leave the study of effective regularization techniques for discriminative learning with PNCs for future work.
-
-\begin{table}[t]
-    \footnotesize
-    \centering
-    \begin{tabular}{lcccc}
-                     & PNC     & PQC     & LC     & RAT-SPN  \\
-        \cmidrule(lr){2-3}
-        \cmidrule(lr){4-5}
-        MNIST        & $98.04$ & $97.38$ & $99.4$ & $98.29$  \\
-        FashionMNIST & $88.84$ & $87.63$ & $91.3$ & $89.89 $ \\
-    \end{tabular}
-    \caption{Test accuracies for MNIST and FashionMNIST.}
-    \label{tab:discrim}
-\end{table}
-
-\section{Conclusions \& Future Work}
-
-We first introduced the concept of a conditional probabilistic circuit, from which we were then able to construct probabilistic neural circuits, which generalize probabilistic circuits and sum-product-quotient networks.
-Note that the construction of PNCs would not have been possible from the formulation of \citet{sharir2018sum}.
-Furthermore, our formulation allows us to intuitively interpret PNCs as neural approximations of deep mixtures of Bayesian networks.
-
-Experimentally, we have shown that for density estimation PNCs deliver on the promise made by SPQNs.
-That is, giving up on tractability improves function approximation in practice.
-For the discriminative case the situation is more nuanced.
-While PNCs achieve perfect accuracy on the training set a lack of proper regularization techniques prevents them from matching accuracies obtained by competing methods.
-We would also like to note that more sophisticated architecture designs for PNCs could possibly further improve their performance.
-For instance, using different numbers of components per partition dramatically increased the performance of SHCLT when compared to HCTL.
-
-In future work we would like to explore the potential of PNCs for sampling, a task that probabilistic models usually struggle with~\citep{lang2022elevating} as good likelihood estimates do not correlate with sample quality~\citep{theis2016note}.
-This would also establish a tighter link to autoregressive models (ARM) (\eg PixelCNN~\citep{van2016conditional}) and we might use ideas developed there for PNCs.
-In this regard, any-order ARM~\citep{uria2014deep,shih2022training} seem to be of particular interest to PNCs as this could allow for arbitrary conditioning sets in PNCs.
-
-Other open questions concern structure learning for PNCs and applying them to tabular data or finding applications of PNCs -- an obvious candidate would be lossless compression with circuits~\citep{liu2022lossless}, as any-order marginalization is not necessary.
 
@@
 This project received funding from the Wallenberg AI, Autonomous Systems and Software Program (WASP) funded by the Knut and Alice Wallenberg-Foundation, as well as from the TAILOR Connectivity Fund (part of the TAILOR project funded by the EU Horizon 2020 research and innovation program under GA No 952215).
-The author would like to thank the anonymous reviewers for the valuable comments.
 
@@
 \end{document}
+
+%%
+%% End of file

### 2024-09-10 202409_aij_dc_problog_aij_second/rebuttal.tex
aij dcproblog paper

\section*{ \Huge Rebuttal}
Firstly, we would like to thank all reviewers for taking their time and giving such detailed comments. We understand that our manuscript is rather long and that reviewing it takes considerable effort. We would like to stress that we found the reviews detailed and very helpful in improving the manuscript. Please find below our detailed rebuttal.

\section*{General Comments}

In the following sections we address the points raised by the reviewers one by one. We highlight in \fixedrebuttal{green} where we agree with the reviewers' opinion and where we made adequate changes in the manuscript. We highlight in \commentrebuttal{blue} parts where we do not necessarily agree with a certain opinion and give our justification. Within the paper there are passages that are highlighted in \new{purple}, which indicates new material not present in the previous version of the paper.

We would first like to highlight the three major changes made to the manuscript compared to the initial submission:
\begin{enumerate}
    \item As asked by Reviewer 1 we have added a proof that the $sigma$-algbera of the probability measure $P_F$ is indeed (as suggested by the reviewer) a cylinder measure. This let (in our opinion) to a more streamlined exposition of the semantics. In the draft of the paper this manifests itself by the three new Propositiona \ref{prop:omegaf} \ref{prop:pfsigma} and \ref{prop:pf}.
    \item We have relegated large parts of the syntactic sugar discussion to the appendix (as suggested by Reviewer 3). We hope this increases readability and accessibility of the manuscript. Specifically, we only introduce the syntactic sugar by example and discuss the technical details in the appendix.
    \item We have added experimental evidence (\cf Section~\ref{sec:experimental}) that our proposed algorithm for the discrete-continuous domain ( (Symbolic IALW)) results in improvements over state-of-the-art approaches.
\end{enumerate}

\section{Reviewer 1}

\subsection{Sigma Algebra}
Overall I believe the paper deserves publication but there is one point that should be clarified: in the proof of
proposition 3.20 that a valid DF-PLP program induces a unique probability distribution over Herbrand interpretations,
eq (C.3), it is not obvious that the set $\{F_\omega|M_\omega(V)\models \mu_1^{b_1} \wedge ... \wedge \mu_n^{b_n}\}$ is
measurable according to $P_F$. In proposition 3.15 the authors proved that $P_F$ exists and is unique but provide no
information on the $\sigma$-algebra of the probability measure. I think the algebra of $P_F$ is the sigma algebra of
cylinder sets of value assignments to $F$ but in this case this is not the powerset so it is not evident that
the set above belongs to the sigma algebra of $P_F$. This should be proved.

\fixedrebuttal{We agree with the reviewer that this was missing. We took the opportunity and restructure the construction of the probability space (using the new Definition~\ref{def:distDB}), which then led to a more explicit construction of the meaure $\probabilitymeasure_\comparisonfacts$. See also our first comment in the general comments above. }

\subsection{Minor comments}

\begin{enumerate}
    \item Example 5.3 and following: why not using a program clause instead of the mathematical definitions of $\nu_1$ and$ \nu_2$?
    \\
    \commentrebuttal{We strictly restricted the clause notation to writing explcit programs. We used math notation everywhere else. This is due to the fact that using clause notation makes it occasionally hard to read formal expression, especially when we make use of indexing.}

    \item Example 5.4: $\eta=1-> \eta=\top$
    \\
    \fixedrebuttal{Fixed.}
    \item Note 3: for a program to be range restricted, all the variables in the head should occur in POSITIVE literals in the body.
    \\
    \fixedrebuttal{Fixed.}
    \item Line 802: Definition 4.14 repeated
    \\
    \fixedrebuttal{Fixed.}
    \item Eq 7.3: I believe $N$ should be $|S|$
    \\
    \fixedrebuttal{Fixed.}
    \item Eq 7.10 $(s,m)->(t,m)$
    \\
    \fixedrebuttal{Fixed.}
    \item line 920 $n>m$ -> n
    \\
    \fixedrebuttal{Fixed.}
    \item line 966 $\phi$->$\varphi$
    \\
    \fixedrebuttal{Fixed.}
    \item line 999: this seems to be removed
    \\
    \fixedrebuttal{Fixed.}
    \item note 5: "(otherwise case in Definition 7.12)"; Def 7.12 does not speak of circuits, why is it mentioned here?
    \\
    \fixedrebuttal{Fixed, Indeed we referenced the wrong definition. We changed the footnote to referencing Definition~\ref{def:sample_labeling_function}.}
    \item Algorithm 7.19: it should return a probability but it actually returns an infinitesimal number
    \\
    \fixedrebuttal{Fixed. We also slightly adapted the text describing the final step of the algorithm (just above Algorithm 7.19.)}
    \item line 1000: computeS
    \\
    \fixedrebuttal{Fixed.}
    \item Figure 7.4: what is the meaning of 1 in \texttt{size\_1(1)} and \texttt{size\_0(1)}? the part in parenthesis does not seem necessary
    \\
    \fixedrebuttal{Fixed. Indeed the part in the parenthesis is not necessary. This was a remnant from a previous version. We removed the parenthesis and its argument.}
    \item Page 44:
    $Eval(5)=\alpha_IALW(1=1)$->$\alpha_{IALW}(size_0=0.4)$
    \\
    \fixedrebuttal{Fixed.}
    \item line 1029 DC-ProbLog->ProbLog
    \\
    \fixedrebuttal{Fixed.}
    \item page 46 Eval formulas: some formulas have equalities with random variables, other equalities with constants

    $Eval(4)=..\alpha_{SIALW}(1=1)$->$Eval(4)=..\alpha_{SIALW}(m=1)$

     $Eval(2)=..\alpha_{SIALW}(0.3)$->$Eval(2)=..\alpha_{SIALW}(M=1)$

     $Eval(5)=\alpha_{IALW}(1=1)$->$\alpha_{IALW}(size_0=0.4)$
     \\
     \fixedrebuttal{Fixed.}
     \item line 1069 inference task repeated
     \\
     \fixedrebuttal{Fixed.}
    \item line 1075 remove "views"
    \\
    \fixedrebuttal{Fixed.}
    \item eq C.3 $F_\omega->F_\omega(V)$
    \\
    \fixedrebuttal{Fixed.}
    \item eq F.11 in the numberator $|S|$ is missing from the limit, in the denominator $N$ should be replaced by $|S|$
    \\
    \fixedrebuttal{Fixed.}
\end{enumerate}

\section{Reviewer 2}

\subsection{Weaknesses}

\begin{enumerate}
    \item The contribution of this work is more conceptual with a focus on extending Problog. It would be much better and more complete if the paper provides concrete evidence of how the extension could benefit real-life applications, and, especially how DC-ProbLog with IALW performs in comparison with related approaches. For example, whether the proposed language can be applied to perform reasoning of mixed variables at a large scale such as images and their labels. There are several attempts with neurosymbolic approach which deserve to be discussed in this paper.

    \fixedrebuttal{We have added an experimental section demonstrating the benefits of the proposed infernce algorithm (\cf Section~\ref{sec:experimental})}

    \fixedrebuttal{In the context of neurosymbolic AI we have already deployed our semantics by developing the neurosymbolic programming language DeepSeaProblog \citep{desmet2023neural}. We discuss this now in the related work section. Note that without the work performed in this paper, developing DeepSeaProblog would have constituted a rather intricate affair.}

    The paper should clarify the complexity of the approach, e.g. whether it is scalable in terms of presentation (space complexity) and inference (time complexity). For example, in example 2.2, if the depth of the relation children-parents-ancestors increases the number of distributional facts would grow exponentially.

    \commentrebuttal{Indeed in Example 2.2, the number of distributional facts would grow exponentially. Hinting at the computational hardness of the problem. Note, however, that here we are intersted in defining the semantics of a Turing complete probabilistic programming language. Meaning that performing inference in \dcproblogsty is actually an undecidable problem. While we do not state this explicitly this follows immediately from \dcproblogsty being Turing-complete.

    In practice, however, we are concerned with programs that will become ground eventually. In this respect we already state explicitly that computing algebraic model counts is \#P-hard (\cf Section~\ref{sec:ALWviaKC}).

    With these two points we believe we have sufficiently delineated the complexity of inference in \dcproblogsty.
    }

    \item The paper's presentation needs to be significantly improved in order to make it more easily comprehensible for a wider range of readers.

    \commentrebuttal{Following the suggestion from Reviewer 3 concerning the syntactic sugar we moved a large part from the main body of the manuscript to the appendix. We hope this improves readability.}

    The numbering is not well-formatted, it is mixed between examples, definitions, propositions, theorems, etc.

    \commentrebuttal{The AIJ author guide does not provide any details on how to number environments (e.g, propositions and theorems)\footnote{\url{https://www.sciencedirect.com/journal/artificial-intelligence/publish/guide-for-authors}}. In our opinion having different environments sharing the same numbering helps to navigate the paper as everything is in order. We would like to note that this is not uncommon. For instance, the template for the "International Conference on Machine Learning" follows a similar convention. }

    Some definitions are not well-written. For example Definition 4.14 is more like a construction/procedure. There are two definitions (3.7 and 4.17) for Parent and Ancestor. Please check other definitions as well.

    \fixedrebuttal{We renamed Definition~\ref{def:df_ancestor} from (Parent, Ancestors) to (Parents and Ancestors of Random Variables). Furthermore, we renamed Definition 4.17 (now Definition~\ref{def:parentancestor2}) from (Parent, Ancestors) to (Parents and Ancestors of Random Terms)}

    \commentrebuttal{While we agree that Definition 4.14 (now Definitino~\ref{def:elim-ad}) and Definition 4.25 (now Definition~\ref{def:adfree-to-core}) have a procedural character we think this is acceptable. Especially, as they define rewrite rules to be applied to a program.}

    There are too many definitions.

    \commentrebuttal{We hope that moving parts of the syntactic sugar discussion to the appendix sufficiently reduced the number of definitions.}

    The paper may be difficult for those who do not have a deep background in probabilistic logic programming. It is a good idea to have the "A Panoramic Overview" section early. However, it goes straight to the CD-Problog concept without laying the ground for readers to understand the basis of probabilistic logic programming.

    \fixedrebuttal{Fixed: we now give an introductory example in Section~\ref{sec:introduction} (Example~\ref{example:intro}).}

\end{enumerate}

\subsection{Other comments}

\begin{enumerate}
    \item Abstract: DC-ProbLog --> DC has not been defined yet.
    \\
    % \commentrebuttal{\dcproblogsty is the name of the language and therefore a proper name. We do not believe that it necessistates stating in the abstract that this stands for  "distributional clauses probabilistic Prolog", where "Prolog" stands for the French expression "programmation en logique".}
    \fixedrebuttal{Fixed.}
    \item 5-10: can be used represent --> to represent
    \\
    \fixedrebuttal{Fixed.}
    \item 55-60: such e.g --> such as
    \\
    \fixedrebuttal{Fixed.}
    \item 120-125: distributional facts in disguise --> please make the statement more formal
    \\
    \fixedrebuttal{Fixed. We now say: Note how probabilistic facts are actually syntactic sugar for distributional facts.}
    \item 135-14: DF-PLP: first time introduced, please provide the detailed name.
    \\
    \fixedrebuttal{Fixed.}
    \item Example 4.26: T2a and T2b are only mentioned once, the purpose of these notations should be well explained.
    \\
    \fixedrebuttal{Fixed. This was indeed a remnant from an earlier draft and the T2a and T2b are referring to the rules in Definition~\ref{def:adfree-to-core}. We now refer to these rules as CR1 and CR2. Note this part of the manuscript moved from the main body to the appendix (Line ~\ref{line:comment_r2_t2a_rule}).}
    \item 555-560: comparison literals --> check the grammar
    \\
    \commentrebuttal{We believe the grammar is correct. Note this part has moved to the Appendix in Line \lineref{line:comment_grammar_r2}}
    \item 555-560: maintains measurability of the latter --> the measurability
    \\
    \fixedrebuttal{Fixed.}
    \item before 1060: lsited bellow --> listed
    \\
    \fixedrebuttal{Fixed.}
\end{enumerate}

\section{Reviewer 3}

\subsection{Weaknesses}

\begin{enumerate}
    \setcounter{enumi}{0}

\item The proposal seems very similar to an approach called Bayesian Logic
Program, but with a different syntax. The semantics look very similar, and
the basic results are inherited from that work. The authors should clarify
the similarities and differences between their approach and Bayesian Logic
Programs, and provide sufficient reasons for introducing a new language.
This comparison, as far as I can see is not included in the related work
section at the end of the paper.

\fixedrebuttal{We have added a discussion on the relationship to Bayesian logic programs to the related work section (Section~\ref{sec:blp}). Following the suggestion from Reviewer 1 we reworked the theory in Section~\ref{sec:semantics}. We now substantially generaliize BLPs and do not inherit any of the results from \Citet{kersting2000bayesian} anymore. We discuss the differences in Section~\ref{sec:blp}.}

\item The most novel (and interesting) part of the paper concerns the infinites-
imal semantics, which comes very late in the paper. I have some doubts
about the formalisation. If I’m correct this could be a problem. (see
comments below)

\commentrebuttal{we discuss the reviewer's doubt in more detail in the technical comments section below. While pointing out some minor issues, we were able to readily resolve them.}

\item I have found the part on “syntactic sugar” not very relevant. It’s ok to
put it in a manual. It does not introduce new important concepts, and it
distracts the reader from the main message of the paper.

\fixedrebuttal{We relegated a large portion from the "syntactic sugar" discussion to the Appendix. In the main body of the paper we only discuss the syntax, while studying the semantics and specific details in the appendix.}

\item No experimental evaluation is provided.

\fixedrebuttal{We have now included an experimental comparison (Section~\ref{sec:experimental}). Furthermore, we would like to point out the experiments  we have performed in a(n) (already published) follow-up paper in the field of neurosymbolic AI \citep{desmet2023neural}.}

\end{enumerate}

\subsection{Readability}
\begin{enumerate}
    \setcounter{enumi}{7}

\item The paper is not easy to read. I believe it is not well organised. The most
important part is at the end of the paper. I believe the paper should concentrate mostly on basic primitives and leave the rewriting of ”syntactic
sugar primitives” later.

\fixedrebuttal{as mentioned above we have relegated large parts of the syntactic sugar discussion to the appendix.}

\item the paper is too long. There are many examples, even for elementary
concepts. The reader gets lost. I think this style is good for a manual,
not a scientific paper.

\commentrebuttal{removing the syntactic sugar discussion from the main body has significantly reduced the paper's length. Concerning the examples, we refer the reader to our comment on the next point (\cf AIJ author guidelines)}

\item I have found the panoramic overview not very interesting. It’s good to have
it in a manual but not in a scientific article. A more compact example
would do the job.

\commentrebuttal{It seems that this is a question of taste. Reviewer 1 and 2 seem to have a different opinion. Furtheermore, AIJ demands from papers to be accessible to a wider audience. Quoting from the AIJ author guidelines\footnote{https://www.sciencedirect.com/journal/artificial-intelligence/publish/guide-for-authors}: "Papers that are heavily mathematical in content are welcome but should include a less technical high-level motivation and introduction that is accessible to a wide audience and explanatory commentary throughout the paper." We believe the overview section and the examples throughout the paper have this function.}

\item The notation sometimes is heavy, I think that some simplifications are
possible.

\commentrebuttal{
We put considerable work into keeping the notation as light as possible yet consistent throughout the paper and also precise. The latter two affect of course, occasionally, readability. However, if there are concrete suggestions on how to improve the notation we would eagerly incorporate them into the manuscript.
} 
\end{enumerate}

\subsection{Technical Issues}

\begin{enumerate}
    \setcounter{enumi}{11}

\item I don’t fully understand why you consider “countable” many random variables instead of just a finite set of random variables. It looks to me that
all the results. I’m not an expert in probability theory but in this way,
you can represent random processes as for instance $X_i \sim N (\mu = X_i , \sigma)$,
and this might introduce rather complex aspects. Perhaps you should
comment on the case of infinite random variables.

\commentrebuttal{We are not entirely sure what the reviewer is referring to: first the reviewer talks about "countably many" (which can be an infinite number but still countable) then in the last sentence the reviewer mentioned infinite random variables. We are not sure whether the reviewer means  countable or uncountable infinity here.

Concerning the distribution above ($X_i \sim N (\mu = X_i , \sigma)$). Here we have a cycle: $X_i$ is used as its own mean and we do not have a valid random variable to begin with. 
}

\item Definition 3.10 Well-Defined Distributional Database. This coincides with
the definition of Bayesian network. I might have missed some details but
I don’t see the difference when the set of random variables is finite. When
they are infinite, the restriction that every formula has a finite number of
ancestors

\commentrebuttal{We substantially altered the definition of a distritbutional database (Definition~\ref{def:distDB}) and the reviewer's comment does directly apply anymore. We point, however, to the new Section~\ref{sec:fintiedistdb}, which we hope clarifies possible remaining doubts.
}

\item Condition DC1 This condition is semantically defined. How is it possible
to check it at the syntactic level without running the program itself? The
conditions might depend on the entire part of the program e.g., Suppose
that you have two statements
\begin{align*}
    x \sim \delta_1 \lpif A.
    \\
    x \sim \delta_2 \lpif B.
\end{align*}
2and there are two subpart of the program that imposes conditions on A
and B may depend on other random variables. As a matter of fact, you
have to ensure that $P(A \land B)$ is equal to $0$. I believe you have to provide
some sufficient “syntactic” condition that guarantees such that $A \land B$ is
not satisfiable.

\commentrebuttal{Indeed, when implementing the language one could think about introducing syntactic sugar that would guarantee validity.
The problem with such syntactic constructs is that they might be sufficient but not necessary to guarantee validity and thereby they would limit the user.

In our opinion ensuring that a set of distributional clauses is well-defined lies therefore with the user of the language itself or with the programmer who implements the language. In this paper we are, however, concerned with the semantics and therefore refrain from making such restrictions. 
}

\item Definition 7.4 states that $e^\oplus = (0, 0)$ is the neutral element for $\oplus$, So I
expect that $(r, n) \oplus  e^\oplus  = e^\oplus  \oplus  (r, n) = (r, n)$, But according to Definition
7.2 when $n > 0$, $(r, n) \oplus  e^\oplus  = e^\oplus  \oplus  (r, n) = e^\oplus $ . From this definition it
looks like a neutral element for $\oplus $ is $(0, \infty)$ but you have to include $\infty$ to
the set $\mathbb{Z}$. However, this implies that the inverse w.r.t. the $\otimes$ does not
exist. I have probably missed something here, it looks like all the following
relies on the fact that this structure is a commutative semiring. \citet{jacobs2021paradoxes} introduces explicitly the operation of difference and inverse, without
considering a neutral elements. Perhaps, the fact that is a semiring is not
so important for the paper.

\fixedrebuttal{indeed this was a mistake and the neutral element should have been defined as $e^\oplus =(0, \infty)$. We have fixed this in the manuscript. Note, however, that this has no serious consequences on any other part of the paper, except minor for the inverse elements. Specifically, we strike (as the reviewer suggested the definition of the inverse elements and define immediately subtraction and division \cf Definition~\ref{def:subdiv}.)
}

\commentrebuttal{Having no explicit inverse for $\otimes$ and $\oplus$ is not a problem: semirings do not necessitate inverse elements (\cf Definition~\ref{def:comm_semiring}).}

\item The solution to the problem of conditioning with $0$-probability events are
solved only partially. I.e., only in the case in which the $0$ probable events
are explicitly stated in an atom by using the primitive delta interval.
However, there are many other situations of events that are still possible
but with $0$ probability. As a simple example, the conjunction of the two
constraints $v \leq 1$ and $v \geq 1$ is equivalent to the constraint $v = 1$, and
I think that in a pure declarative approach, like the one that is pursued
by ProbLog they should have the same semantics. While the latter is
interpreted in terms of infinitesimal number the latter is not. How would
the infinitesimal approach generalize to the other comparison atoms? A
possible solution to this problem can be obtained by rewriting $x \leq y$ into
$(x < y ) \overline{\vee} (x=y)$
˙ (where $ \overline{\vee}$ denotes the xor operator).

\commentrebuttal{Here we have to disagree with the following statement: "While the latter is
interpreted in terms of infinitesimal number the [former] is not". This is not correct. Neither of the two are interpreted as infinitesimal numbers!
We would like to stress here that $x=y$ is not interpreted as an infinitesimal interval but simply as an indicator function where the equality has to hold strictly. This means that the semantics align between $v \leq 1 \land v \geq 1$ and  $v = 1$ as they are equivalent. Note that both of these expressions are, however, not equivalent to $v \doteq 1$, which is the actual infinitesimal interval. We already discuss this issue in Example~\ref{ex:conditional_prob}.
}

\item I also think that the semantics based on infinitesimals introduces other
paradoxes, For instance, if $x \sim \mathcal{N} (0, 1)$ $P (x < 0 | x = 1)$ is equal to $(0, 0)$.
However, by infinitesimal semantics this is equal to $( \frac{2}{y}
, -1)$ where $y$ is
the value of $(0, 1)$ in 1 (which is different from $(0, 0)$).

\commentrebuttal{Indeed, following the definition of a conditional probability with a zero probability conditioning event (using a limit), gives us
$P (x < 0 | x \doteq 1)=0 = (0\epsilon^0) = (0,0)$. Note that here we use the "$\doteq$" notation to indicate that we condition on an infinitesimal interval.

However, using infinitesimal numbers does not yield $(0,-1)$. If we write the conditional probability in question using infinitesimal numbers we get the following result:
\begin{align}
    \frac{
        \sum_{i}^{|S|} \alpha_{SIALW}(\ive{x^{(i)}<0}) \otimes \alpha_{SIALW}(\ive{x^{(i)} \doteq 1})
    }
    {
        \sum_{i}^{|S|} \alpha_{SIALW}(\ive{x^{(i)} \doteq 1})
    }
\end{align}
Because $x^{(i)} \doteq 1$ denotes an observation, all the samples $x^{(i)}$ are forced to take the value $1$. Plugging this into above expression gives:
\begin{align}
    \frac{
        \sum_{i}^{|S|}  (0,0) \otimes (p(1),1)
    }
    {
        \sum_{i}^{|S|}   (p(1),1)
    }
    \label{eq:rev3idiot}
    &=
    \frac{
        \sum_{i}^{|S|} (0,1)    
    }{
        \sum_{i}^{|S|}  (p(1),1)
    }
    \\
    &=
    \frac{
        (0,1)
    }
    {
        (|S|  p(1),1)
    }
    \\
    &=
    (0,0)
    \label{eq:r3:misconception}
\end{align}
That means if we straightforwardly apply the correct expression (the one that we derived in the manuscript) to compute the conditional probability we do obtain the result we expect.
}

\item Theorem 7.11 and Proposition 7.14 are not well formulated because the
meaning of $(x, 0) \approx (y, n)$ when $n \neq 0$ is not defined. Consider the previous
example $P (x < 0 | x  1) = (0, 0)$, however it is possible that S contains
at least one sample $S^{(i)}$ such that $S^{(i)}_x = 1$. In this case, the summation
at the denominator will be some value $(y, 1)$ while at the numerator you
will have an empty summation, which I suppose it is equal to $(0, 0)$. The
$\frac{(0,0)}{(x,1)}$
is equal to $(0, -1)$.

\commentrebuttal{Let us consider the left-hand side of Equation~\ref{eq:rev3idiot} above:
\begin{align}
    \frac{
        \sum_{i}^{|S|}  (0,0) \otimes (p(1),1)
    }
    {
        \sum_{i}^{|S|}   (p(1),1)
    }
\end{align}
Instead of multiplying out the numerator we can also cancel out the $(p(1),1)$ expression to obtain:
\begin{align}
    \frac{
        \sum_{i}^{|S|}  (0,0) \otimes (p(1),1)
    }
    {
        \sum_{i}^{|S|}   (p(1),1)
    }
    &
    =
    \frac{
        (0,0) \otimes (p(1),1) \otimes \cancel{\sum_{i}^{|S|} (1,0)}  
    }
    {
        (p(1),1) \otimes  \cancel{\sum_{i}^{|S|} (1,0)}
    }
    \\
    &=
    \frac{
        (0,0) \otimes \cancel{(p(1),1) } 
    }
    {
        \cancel{(p(1),1) }
    }
    \\
    &=
    (0,0)
\end{align}

While here we show that the particular case (pointed out by the reviewer) does not cause any problems (if one correctly applies the defined algebra), we also prove this for the general case in Subsection~\ref{app:proof:alwapproximation}.
We conclude that the apparent misbehavior never arises when the definitions are applied correctly.

Also note that for the case that we do not condition on $x\doteq 1$ but on $x=1$, we almost surely compute the correct result. We state this explicitly in Proposition~\ref{prop:alw_consistency} already.
}

\end{enumerate}

\subsection{Minor Points}
\begin{enumerate}
    \setcounter{enumi}{18}
\item line 150 “on” → “one”
\\
\fixedrebuttal{Fixed.}
\item line 175 “mutually marginally independent” → “mutually independent”
why you use marginally.
\\
\fixedrebuttal{We removed "marginally."}

\item Definition 3.3: “regular ground term” → “regular ground atom”
\\
\commentrebuttal{No, these are indeed ground terms and not ground atoms.}

\item Paragraph 901 explains the order of sampling, and introduces the term
“ancestral sample”. This is confusing because every sample must be an
ancestral sample since you need the value of the parent variables in order
to sample a variable. I suggest not to use this term and to explain before
Property 7.1, that to obtain a sample for all the variables we have to
proceed according to the partial order on the variables.

\commentrebuttal{With using the term "ancestral sampling" we are stressing the fact we perform sampling using ancestral sampling. This is in order to distinguish the samples from other sampling strategies, such as Gibbs sampling or any other MCMC method. As this MCMC methods are ubiquitous in probabilistic programming we us "ancestral sample" to make the sample strategy explicit.}

\item line 510: $L > 5$ → $\Lambda \geq 5$
\\
\fixedrebuttal{Fixed.}
\item Definition 4.21. in the first bullet, you assume that only one random
variable occurs in a random term, however in general this is not the case.
\\
\commentrebuttal{We do not believe that we make this assumption. We actually explicitly state "a distribution term that involves exavtly $k$ different random terms". Note this definition has moved and is now called Definition~\ref{def:elim-dc}.}
\item Definition 4.28: Clarify what is $P^{DF,*}$.
\\
\fixedrebuttal{We now clearly state this at the end of Definitino~\ref{def:adfree-to-core}.}
\item Definition 7.9: revise English and “evaluate” → “denotes”. The defini-
tion is hard to read. The reference to literals and negated literals is a
bit confusing. For instance, what happens to the negation of the comparison atom $v =/= 3$? Using the notation $[[s_k = x]]$ could also improve
readability.
\\
\fixedrebuttal{We rewrote the Definiton~\ref{def:sample_labeling_function}, hopefully improving the readability. We also use the Iverson bracket $\ive{\cdot}$ as suggested.}
\item Lines 1029-1031 it seems that you compare DC problog with itself. Is the
system introduced in [Fierence et al. 2015] also called DC-ProbLog?
\\
\fixedrebuttal{The language introduced by \citep{fierens2015inference} is called \problogsty. This was a typo on our side, and we change "\dcproblogsty" to "\problogsty",}
\item Definition 7.24: “given is” → “given”
\\
\fixedrebuttal{Fixed.}
\item It’s not completely clear if $\neg(x =/= 2)$ should be considered a negative
literal, or $x =/= 2$ is a negative literal of the literal $x=2$.
\\
\commentrebuttal{We discuss the issue of how to interpret negation in detail in Appendix~\ref{sec:non-mixture-dc} and Appendix~\ref{sec:dcproblog-dc}. Especially as \dcproblogsty and \dcsty have different semantics for negation. In short: when distributional clauses are present $x =/= 2$ is not necessriöy equivalent to $\neg(x=:=2)$ otherwise they are equivalent.}
\item Paragraph 185: is very obscure at this point of the paper. I believe it
could be put in a proper related work section.
\\
\fixedrebuttal{We removed the discussion in this paragraph and give now forward pointers to the related work Section~\ref{sec:related} and to Appendix~\ref{sec:dcproblog-dc}.}
\item The definition of infinitesimal interval is not very intuitive. In order to
grasp it I had to read the paper \citep{jacobs2021paradoxes}. I suggest reporting his explanation provided below the definition 5.2 of that paper.
\\
\fixedrebuttal{We added a reference to \Citet[Section 5.2]{jacobs2021paradoxes} just before Definition~\ref{def:inf_number}.}
\item Bibliography format is not correct.

\fixedrebuttal{Fixed.}

\end{enumerate}

\subsection{Reccomendations}

\begin{enumerate}
    \setcounter{enumi}{32}

\item The paper is potentially interesting but I think it needs to be improved in
the readability, i.e., make it shorter by concentrating on the major points,
and optimizing the usage of examples. The technical part should also
be double-checked. I might be wrong with my comments, in such a case
the authors should better explain the technical part, or if I’m right the
authors have to fix and revise (if necessary) the proofs. Since the paper
does not contain major technical theoretical results, most of the results
are applications/simple generalization of existing results, the paper should
provide some experiments that show the potential of the approach. I think
that to address all the above observations the paper needs a major revision,
which is my suggestion.

\comment{We hope that the changes we made to the paper addresses the concerns raised by the reviewer. Specifically,
\begin{enumerate}
    \item we substantially reduced the section on syntactic sugar
    \item we corrected our mistake regarding the neutral elements for the addition in our semiring and adapted the manuscript accordingly,
    \item we added experimental evidence that the developed SIALW algorithm exhibits advantageous behavior when compares to similar algorithms in the presence of low probability events. 
\end{enumerate}

We would also argue that our technical contribution does not constitute a simple extension of existing ideas. This is exemplified by the fact that a series of papers \citep{kersting2000bayesian,gutmann2010extending,gutmann2011magic,azzolini2021semantics} have indeed tackled the problem of unifying discrete and continuous random variables for probabilistic logic programming but none have succeeded in linking these discrete-continuous languages to the inference method of knowledge compilation -- the de facto standard for exact inference in the discrete domain. We achieve this using our IALW semiring. While we do deploy proof techniques developed in other papers, considerable effort has gone into correctly formulating them for the discrete-continuous domain. We are of the strong opinion that our theoretical contributions are not applications but strong generalizations of existing work.
}
\end{enumerate}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_appendix1/amc.tex
aij dcproblog paper

\section{Weighted Model Counting and Algebraic Model Counting} \label{app:amc}

Formal logic can be used to describe problems that arise in artificial intelligence. Here we will focus on propositional logic formulas, which we define as follows
\begin{definition}\label{def:preplogic}
	Let $\bvars$ be a set of $M$ Boolean variables.
	We then define propositional logic formulas as Boolean combinations (by means of the standard Boolean operators $\{\neg, \land, \lor, \rightarrow, \leftrightarrow \}$) of {\bf Boolean variables} $b_i \in \bvars$.
\end{definition}

A fundamental question to ask is whether there exists an assignment to the Boolean variables present in the logic formula that satisfies the formula.
% \begin{definition}[Satisfying interpretation of propositional logic formula]
% \label{def:sat_interpretation}
% 	Let ${\bf j}$ and ${\bf k}$ be two disjoint sets of variables and $\phi({\bf j}, {\bf k})$ be a propositional formula over ${\bf j}$ and ${\bf k}$.
% 	The set of total interpretations (or total assignments) that satisfy $\phi$ is the set of assignments to the elements in ${\bf j}$ and ${\bf k}$ that satisfy $\exists {\bf j}, ${\bf k}$:\phi(${\bf j}$,${\bf k}$)$.
% 	We denote the set of total satisfying interpretations (or models) by $\mathcal{I}_{{\bf j},{\bf k}}(\phi)$.
% 	The set of partial interpretations is denoted by $\mathcal{I}_{{\bf j}}(\phi)$, which is the set of assignments to ${\bf j}$ that satisfy $\exists {\bf k}:\phi({\bf j},{\bf k})$.
% 	The set of total assignments to a partially interpreted formula is denoted by $\mathcal{I}_{{\bf j}} (\phi^{\bf k})$, which denotes the set of assignments to the elements in ${\bf j}$ that satisfy $\phi({\bf j},{\bf k}_\inter)$, with ${\bf k}_\inter \in \mathcal{I}_{{\bf k}}(\phi)$. 
% \end{definition}
Determining whether a propositional logic formula is satisfiable is a computational hard problem and falls in the NP-complete complexity class. Nevertheless, a plethora of practical solvers exists (\eg MiniSAT~\citep{sorensson2005minisat}, CryptoSAT~\citep{lafitte2018cryptosat}) that tackle the Boolean satisfiability problem and perform astoundingly well in practice by exploiting structure present in all problems but the most intricate ones.

We can also ask an other questions: how many distinct assignments satisfy a formula? For propositional logic formulas this simply amounts to counting the number of satisfying assignment to the Boolean variables --  a \#P-complete problem~\citep{valiant1979complexity}. The number of distinct satisfying assignments is dubbed the \textit{model count} and the task of computing the model count is also referred to as $\#SAT$.

Weighted model counting further generalizes \#SAT tasks. Instead of simply counting the number of satisfying assignments, one performs a weighted sum over models.
\begin{definition}[Weighted model counting (WMC)] \label{def:wmc} Given a set $\bvars$ of $M$ Boolean variables, a weight function $w: \mathbb{B}^M \rightarrow \mathbb{R}_{\geq 0}$, and a propositional formula $\support$ over $\bvars$, the {\bf weighted model count} is
	\begin{align}\label{eqn:wmc}
	 \wmc(\support,\weight {\mid} \bvars) = \sum_{\omega \in \mathcal{M}(\phi)} \weight(\omega)
	\end{align}
	$\mathcal{M}(\phi)$ is the set of models of $\support$, \ie the set of interpretations of the Boolean $\bvars$ that satisfy $\support$.
\end{definition}
Traditionally, WMC is used when the weight function $\weight$ factorizes as a product of weights of literals:
\begin{equation}
 \wmc(\support, \weight |\bvars) {=} \sum_{\omega \in \mathcal{M}(\phi)} \prod_{\ell \in \omega} \weight(\ell)
\end{equation}
When performing probabilistic inference, we take
\begin{equation}
0 {\leq} w(b_i){\leq} 1 \quad\text{and}\quad w(b_i){+}w(\neg b_i)=1
\end{equation}
The resulting sum over products is then a computation in the probability semiring~\citep{kimmig2017algebraic}. Effectively, this computes the probability that a propositional logic formula is satisfied.

\citet{kimmig2017algebraic} further generalize probabilistic inference over propositional formulas, \ie over satisfying models, to arbitrary commutative semirings. The general setting of model counting over commutative semirings is called algebraic model counting.
\begin{definition}\label{def:comm_semiring} 
	A {\bf  commutative semiring} is an algebraic structure $(\mathcal{A},\oplus,\otimes,\allowbreak e^{\oplus},e^\otimes)$ equipping a set of elements $\mathcal{A}$ with addition and multiplication such that
	\begin{enumerate}
		\item addition $\oplus$ and multiplication $\otimes$ are binary operations $\mathcal{A}\times \mathcal{A}\rightarrow \mathcal{A}$
		\item addition $\oplus$ and multiplication $\otimes$ are  associative and commutative binary operations over the set $\mathcal{A}$
		\item $\otimes$ distributes over $\oplus$
		\item  $e^\oplus \in \mathcal{A}$ is the neutral element of $\oplus$
		\item  $e^\otimes \in \mathcal{A}$ is the neutral element of $\otimes$
		\item $e^\oplus \in \mathcal{A}$ is an annihilator for $\otimes$
	\end{enumerate}
\end{definition}

\begin{definition}\label{def:amc} (Algebraic model counting (AMC)) \citep{kimmig2017algebraic} Given:
	\begin{itemize}
		\item a propositional logic theory $\phi$ over a set of variables $\bvars$
		\item a commutative semiring $(\mathcal{A},\oplus,\otimes,e^{\oplus},e^\otimes)$
		\item a labeling function $\alpha:\mathcal{L}\rightarrow \mathcal{A}$, mapping literals $\mathcal{L}$ from the variables in $B$ to values from the semiring set $\mathcal{A}$
	\end{itemize} 
	The algebraic model count of a theory $\phi$ is then defined as:
	\begin{align}
	\textstyle \amc(\phi,\alpha | \bvars) = \bigoplus_{b \in \mathcal{I}_\bvars(\phi)} \bigotimes_{b_i \in b} \alpha (b_i)&  \nonumber
	\end{align}
\end{definition}
We use $\alpha$ instead of $w$ and the term label rather than weight to reflect that the elements of the semiring cannot always be interpreted as weights.

Algebraic model counting constitutes a general framework for many common inference tasks in artificial intelligence. Defining an appropriate semiring and labeling functions allows one, for instance, to perform sensitivity analysis, compute gradients or determine the provenance of queries in databases~\citep{kimmig2017algebraic}.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_appendix1/compilation.tex
aij dcproblog paper

\section{Knowledge Compilation and Algebraic Model Counting} \label{app:compilation}

As mentioned in the previous section, model counting is a computationally hard problem, \#P-hard to be precise. Nevertheless practically useful model counters exist. State-of-the-art techniques for solving model counting problems, are based on exhaustive DPLL algorithms~\citep{birnbaum1999good}, which count the number of satisfying assignments to a formula. These solvers can be divided into two classes: the ones that build up a trace of the DPLL search, and the ones that do not. The latter return immediately the model count. The former builds up a diagrammatic representation of the propositional formula over which the model count can be obtained efficiently. By keeping a trace, such \#SAT solvers~\citep{huang2005dpll,oztok2018exhaustive} constitute, in fact, top-down {\em knowledge compilation} schemes.

Knowledge compilation~\citep{darwiche2002knowledge} has emerged as the go-to technique for dealing with the computational intractability of propositional reasoning (\#P-hard~\citep{valiant1979complexity}). The key idea is to split up inference on logical formulas into an \textit{off-line} and an \textit{on-line} step. In the off-line step, a propositional formula is compiled from its source representation into a target representation, in which repeated on-line poly-time inference is available.

\#SAT can also be performed by compiling formulas bottom-up~\citep{choi2013dynamic}. However, it has been shown ~\citep{huang2005dpll,oztok2018exhaustive} that top-down compilation, \ie knowledge compilation through exhaustive DPLL search, outperforms bottom-up compilers.

Similar to model counting, weighted model counting has as well been performed via knowledge compilation, for instance for probabilistic inference in Bayesian networks~\citep{chavira2008probabilistic} or probabilistic programming~\citep{fierens2015inference}.

A popular and well-known target language for the knowledge compilation of propositional formulas are {\em deterministic decomposable negation normal
form} (d-DNNF) formulas~\citep{darwiche2001tractable}
and variations thereof~\citep{bryant1986graph,darwiche2011sdd}.
Boolean formulas in {\em negation normal form}~\citep{darwiche1999compiling} are nested conjunctions and disjunctions that allow negations only to be applied directly to the atoms in the formula.
Determinism and decomposability are two additional properties that we need to impose in order to guarantee tractable probabilistic inference using algebraic model counting.

\begin{definition}[Determinism] An NNF formula is deterministic if and only if for every disjuntion the disjuncts are pairwise logical inconsistent.
\end{definition}

\begin{definition}[Decomposability] An NNF formula is decomposable if and only if for every conjunction the conjuncts do not share any variables.
\end{definition}

\begin{definition}[Smoothness] 
An NNF is smooth if and only if for each disjunction the disjuncts mention the same variables.
\end{definition}

\begin{algorithm}[t]
    \SetKwFunction{EvalFn}{Eval}
    \SetKwProg{Fn}{function}{}{}
    \SetKwProg{ElseComment}{function}{}{}

	\caption{Evaluating a d-DNNF circuit $\Gamma$ for a
		commutative semiring $\mathcal{S}$ and labeling function $\alpha$~\citep{kimmig2017algebraic}.}
	\label{alg:eval}
\Fn{\EvalFn{$\Gamma$,$\mathcal{S}$,$\alpha$}}{
		\If{  $\Gamma$ is a literal node $l$}{ \Return $\alpha(l)$}
		\ElseIf{$\Gamma$ is a disjunction $\bigvee_{i=1}^m \Gamma_i$}
		{\Return $\bigoplus_{i=1}^m$ \EvalFn{($\Gamma_i$,$\mathcal{S}$,$\alpha$}}
		\Else( \tcp*[f]{$\Gamma$ is a conjunction $\bigwedge_{i=1}^m \Gamma_i$}){
		 \Return $\bigotimes_{i=1}^m$ \EvalFn{($\Gamma_i$,$\mathcal{S}$,$\alpha$}}
	}
\end{algorithm}

% 		{$N$ is a conjunction $\bigwedge_{i=1}^mN_i$}

We will also need the notions of {\em sum-neutrality} and {\em consistency preserving}.

%In order to compute algebraic/weighted model counts on a d-DNNF, we additionally need the neutral-sum property.	
\begin{definition}[Neutral-sum property~\citep{kimmig2017algebraic}]\label{lem:nsp} A semiring addition and labeling function pair $(\oplus,\alpha)$ is neutral \emph{if and only if} $\forall b \in \bvars:  \alpha(b)\oplus \alpha(\neg b)=e^\otimes$.
\end{definition}

\begin{definition}[Consistency-preserving property~\citep{kimmig2017algebraic}]\label{lem:con_prev} A semiring multiplication and labeling function pair $(\otimes,\alpha)$ is consistency-preserving \emph{if and only if} $\forall b \in \bvars:  \alpha(b)\otimes \alpha(\neg b)=e^\oplus$.
\end{definition}

\begin{theorem}[AMC on d-DNNF~\citep{kimmig2017algebraic}]
	\label{theo:amc_ddnnf}
Evaluating an sd-DNNF representation of the propositional theory $\phi$, using Algorithm~\ref{alg:eval}
%, for a semiring and labeling function with neutral tuple $(\oplus, \alpha)$
is a correct computation (cf. \citep[Definition 10]{kimmig2017algebraic}) of the algebraic model count.
\end{theorem}

Under the premise that the $\otimes$ and $\oplus$ operations in Algorithm~\ref{alg:eval} can be perform in constant time and that intermediate results are cached, the computation of an algebraic model count on an sd-DNNF is performed in poly-time~\citep{darwiche2002knowledge}.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_appendix1/lp.tex
aij dcproblog paper

\section{Logic Programming Basics}
\label{app:lp}
This appendix summarizes key concepts of the syntax and semantics of logic programming; for a full introduction, we refer to \citep{lloyd2012foundations}.

The basic building blocks of logic programs are \emph{variables} (denoted by strings starting with upper case letters), \emph{constants}, \emph{functors} and
\emph{predicates} (all denoted by strings starting with lower case letters).  A \emph{term} is a variable, a constant, or a functor
\texttt{f} of \emph{arity} $n$ followed by $n$ terms $\mathtt{t_i}$, \ie,
$\mathtt{f(t_1,...,t_n)}$. 
An \emph{atom} is a predicate $p$ of arity $n$ followed by $n$ terms $\mathtt{t_i}$, \ie,
$\mathtt{p(t_1,...,t_n)}$. A predicate $p$ of arity $n$ is also written as $p/n$. A \emph{literal} is an
atom or a negated atom $\mathtt{not(p(t_1,...,t_n))}$. A \emph{definite
  clause} is a universally quantified expression of the form $h
:- b_1, ... , b_n$ where $h$ and the $b_i$ are atoms.
$h$ is called the \emph{head} of the clause, and $b_1, ... , b_n$ its
\emph{body}. Informally, the meaning of such a clause is that if all
the $b_i$ are true, $h$ has to be true as well. 
 A
\emph{normal clause}  is a universally quantified expression of the form $h
:- b_1, ... , b_n$ where $h$ is an atom and the $b_i$ are
literals.
If $n=0$, a clause is called \emph{fact} and simply written
as $h$. A \emph{definite clause program} or \emph{logic program} for
short is a finite set of definite clauses. A \emph{normal logic
  program} is a finite set of normal clauses. 
A \emph{substitution} $\theta$ is an expression of the form
$\{V_1/t_1,...,V_m/t_m\}$ where the $V_i$ are different variables and
the $t_i$ are terms. Applying a substitution $\theta$ to an expression
$e$ (term or clause) yields the \emph{instantiated} expression $e\theta$
where all variables $V_i$ in $e$ have been simultaneously replaced by
their corresponding terms $t_i$ in $\theta$. If an expression does not
contain variables it is \emph{ground}. Two expressions $e_1$ and $e_2$ can be \emph{unified} if and only if there are substitutions $\theta_1$ and $\theta_2$ such that $e_1\theta_1 = e_2\theta_2$. In Prolog, unification is written using $=$ as an infix predicate.

The \emph{Herbrand universe} of a logic program is the set of ground terms that can be constructed using the functors and constants occurring in the program.\footnote{If the program does not contain constants, one arbitrary constant is added.} 
The \emph{Herbrand base} of a logic program is the set of ground atoms that can be constructed from the predicates in the program and the terms in its Herbrand universe. 
A truth value assignment to all atoms in the 
Herbrand base is called \emph{Herbrand interpretation}, and is also represented as the set of atoms that are true according to the assignment. 
A Herbrand interpretation is a \emph{model} of a clause $\mathtt{h ~~:- ~~b_1,\ldots ,b_n\ldotp}$ if for every substitution $\theta$ such that all $b_i\theta$ are in the interpretation, $h\theta$ is in the interpretation as well. It is a model of a logic program if it is a model of all clauses in the program. The model-theoretic semantics of a definite clause program is given by its smallest Herbrand model with respect to set inclusion, the so-called \emph{least Herbrand model} (which is unique). We say that a logic program $\dcpprogram$ \emph{entails} an atom $a$, denoted $\dcpprogram\models a$, if and only if $a$ is true in the least Herbrand model of $\dcpprogram$.

Normal logic programs use the notion of \emph{negation as failure}, that is, for a ground atom $a$, $not(a)$ is true exactly if $a$ cannot be proven in the program. They are not guaranteed to have a unique minimal Herbrand model. Various ways to define the canonical model of such programs have been studied; see, \eg, \citep[Chapter 3]{lloyd2012foundations} for an overview. We use the well-founded semantics here~\citep{van1991well}.

The main inference task of a logic programming system is to determine
whether a given atom, also called \emph{query} (or \emph{goal}), is true in the canonical
 model of a  program. If the answer is yes (or no), we also say that the query \emph{succeeds} (or \emph{fails}). If such a query is not ground, inference asks for the existence of  \emph{answer substitutions}, that is,  substitutions that ground the query into atoms that are part of the canonical model.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_appendix2/app_table.tex
aij dcproblog paper

\section{Table of Notations}\label{app:table}
\begin{center}
    \rowcolors{2}{gray!15}{white}

\begin{tabular}{c|p{0.5\textwidth}|c}
symbol & meaning & for details, see \\ \hline

  $\distributionfunctors$  & set of distribution functors  & Definition~\ref{def:reserved_vocabulary}\\
  $\arithmeticfunctors$   &  set of arithmetic functors & Definition~\ref{def:reserved_vocabulary}\\
  $\comparisonpredicates$   &  set of comparison  predicates & Definition~\ref{def:reserved_vocabulary}\\
  \hline
  $\samplespace_\nu$ & sample space of random variable $\nu$ & Definition~\ref{def:samplespace}\\
  $\samplefunction(\cdot)$ & value assignment function & Definition~\ref{def:samplespace}\\
  $\distdb$ & distributional database & Definition~\ref{def:distDB}\\
  $\randomvariableset$ & set of random variables & Definition~\ref{def:distDB}\\
  $(\samplespace_\distdb, \sigmaalgebra_\distdb, \probabilitymeasure_\distdb)$ & probability space induced by $\distdb$ & Definition~\ref{def:well-distdb}\\
  % $\measurerandomvariableset$ & probability measure over $\randomvariableset$ defined by $\distdb$ & Proposition~\ref{prop:pv}\\
  \hline
  $\comparisonfacts$ & set of Boolean comparison atoms & Definition~\ref{def:comparison-atoms-set}\\
  $\samplespace_\comparisonfacts$ & sample space induced by $\comparisonfacts$ & Proposition~\ref{prop:omegaf}\\
  $\sigmaalgebra_\comparisonfacts$ & sigma algebra induced by $\comparisonfacts$ & Proposition~\ref{prop:pfsigma}\\
  $\probabilitymeasure_\comparisonfacts$ & probability measure induced by $\comparisonfacts$ & Proposition~\ref{prop:pf}\\
  \hline
  $\dfprogram=\distdb  \cup \logicprogram$ & \dfplpsty program & Definition~\ref{def:core-prog}\\
  $\comparisonfacts_{\samplefunction(\randomvariableset)}$ & consistent comparison database induced by $\samplefunction$ on the random variables in $\randomvariableset$ & Definition~\ref{def:consistent-fact-db}\\
  $\probabilitymeasure_\dfprogram$ & probability measure over Herbrand interpretations defined by $\dfprogram$ & Proposition~\ref{prop:pp}\\
  \hline
  $\program$  & \dcproblogsty program & Definition~\ref{def:fullprog}\\
  \adfreeprogram & AD-free \dcproblogsty program & Definition~\ref{def:ad_free_program}\\
  $\headsdc_\adfreeprogram$ & set of heads of distributional clauses in \adfreeprogram & Definition~\ref{def:ad_free_program}\\
  $\randomtermset_\adfreeprogram$ & random terms in $\headsdc _\adfreeprogram$ & Definition~\ref{def:ad_free_program}\\
  $\dclauses_{\adfreeprogram}$ & set of distributional clauses in $\adfreeprogram$ & Definition~\ref{def:dc-df-well-def}\\
  $\contextfunc(\cdot)$ &  contextualization function & Definition~\ref{def:context_function}\\
  $\dfadfreeprogram$ &  \dfplpsty program providing the semantics of $\adfreeprogram$ & Definition~\ref{def:adfree-to-core}\\
  \hline
  $MOD(\dcpprogram)$ & models of a program $\dcpprogram$ & Theorem~\ref{theo:model_equivalence} \\
  $ENUM(\phi)$ & models of a propositional formula $\phi$ & Theorem~\ref{theo:model_equivalence} \\
  $\alpha(\cdot)$ & labeling function of a propositional literal & Definition~\ref{def:labeling_function} \\
  $\ive{\cdot}$ & Iverson bracket denoting an indicator function & Definition~\ref{def:labeling_function} \\
  $\E[\cdot]$ & expected value & Theorem~\ref{theo:label_equivalence} \\
  \hline
  $\mathbb{I}$ & set of infinitesimal numbers & Equation~\ref{def:inf_number} \\
  $\mathcal{S}$ & set of ancestral samples & Equation~\ref{eq:ancestral_samples}
\end{tabular}
\end{center}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_appendix2/detailed_comp_to_GutmanNittiDC.tex
aij dcproblog paper

\section{Relation to the DC language}\label{sec:dcproblog-dc}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Distributional clauses were first introduced in the language of the same name  by \cite{gutmann2011magic}, which at that point did not support negation. For negation-free programs, our interpretation of distributional clauses exactly corresponds to theirs, and \dcproblogsty thus generalizes both ProbLog (with negation) and the original (definite) distributional clause language.

% Indeed, our stochastic $W_P$-operator is a direct generalization of 
% Gutmann et al.'s stochastic $T_P$-operator, and also  incorporates the checks on when a comparison can use a random variable into the definition of the operator executing the program.
% Our declarative view, on the other hand, incorporates these into the transformed program itself, thus making the meaning of a program independent of its execution mechanism.

In the following, we first discuss how the semantics of \dcproblogsty differs from \cite{nitti2016probabilistic}'s procedural view on negated comparison atoms, and then how \dcproblogsty's acyclicity conditions imposed on valid programs differ from those of \cite{gutmann2011magic}.

\subsection{Non-exhaustive sets of DCs}\label{sec:proc-nitti}
\cite{nitti2016probabilistic}  have extended the procedural view of the stochastic $T_P$ operator to locally stratified programs with negation under the perfect models semantics.\footnote{Local stratification is a necessary condition for perfect models semantics, and a sufficient one for well-founded semantics. On this class of programs, both semantics agree~\citep{van1991well}} In their view, a distributional clause \probloginline{x~d :- body} is informally interpreted  as "if \probloginline{body} is true, define a random variable \probloginline{x} with distribution \probloginline{d}". They then use the principle that "any comparison involving a non-defined variable will fail; therefore, its negation will succeed", \ie, they apply negation as failure to comparison atoms. 
In contrast, as already illustrated in Section~\ref{sec:non-mixture-dc}, we take a purely declarative view here, where all random variables are defined up front, independently of logical reasoning, and distributional clauses serve as syntactic sugar to compactly talk about a group of random variables. Then, truth values of comparison atoms are fully determined by their external interpretation, and do not involve reasoning about whether a random variable is defined or not. That is, we apply classical negation to comparison atoms, and restrict negation as failure to atoms defined by the logic program itself.

The following example adapted from \cite{nitti2016probabilistic} illustrates the difference.
\begin{example}\label{ex:dc-vs-dcproblog}
	Consider the following program about the color of certain objects, where the number of objects is given by the random variable \probloginline{n}:
	\begin{problog}
n ~ uniform([1,2,3]).
color(1) ~ uniform([red,green,blue]) :- 1=<n .
color(2) ~ uniform([red,green,blue]) :- 2=<n .
color(3) ~ uniform([red,green,blue]) :- 3=<n .
not_red :- not color(2)=:=red .
not_red_either :- color(2)=\=red.
	\end{problog}
	The \dcproblogsty semantics is given by the transformed program:
	\begin{problog}
v0 ~ uniform([1,2,3]).
v1 ~ uniform([red,green,blue]).
v2 ~ uniform([red,green,blue]).
v3 ~ uniform([red,green,blue]).

rv(n,v0).
rv(color(1),v1) :- rv(n,v0), 1=<v0.
rv(color(2),v2) :- rv(n,v0), 2=<v0.
rv(color(3),v3) :- rv(n,v0), 3=<v0.

not_red :- rv(color(2),v2), not v2=:=red.
not_red_either :- rv(color(2),v2), v2=\=red.
	\end{problog}
	If $\mathprobloginline{n}=1$ (i.e., $\mathprobloginline{v0}=1$), neither \probloginline{color(2)} nor \probloginline{color(3)} are associated with a distribution.  Thus, \probloginline{rv(color(2),v2)} fails, and both \probloginline{not_red} and \probloginline{not_red_either} therefore fail as well, independently of the values of the comparison literals.
	In contrast, under 
	the procedural semantics of \cite{nitti2016probabilistic}, \probloginline{color(2)=:=red} fails in this case, and \probloginline{not_red} thus succeeds. Similarly, \probloginline{color(2)=\=red} fails, and  \probloginline{not_red_either} thus fails.  
	Both views agree for $n>1$.
\end{example}

This example again illustrates that DC-ProbLog's semantics clearly follows the spirit of the distribution   semantics of defining a distribution over interpretations of basic facts (comparison atoms in this case) independently of the logic program rules. 
%In this view, all random variables "exist" and  "are defined" in all possible worlds, and  truth values of 
%comparison atoms thus exclusively depend on the external evaluation with respect to the current values of these random variables. This also implies that the negation of a comparison atom is equivalent to the comparison atom using the negated comparison operator, \eg, \probloginline{color(2)=:=red} and \probloginline{not color(2)=\=red} are equivalent, as are \probloginline{n>=2} and \probloginline{not n<2}.  \ak{we should probably stress these declarative aspects earlier already}
We note that the expressive power of logic programs allows the programmer to explicitly model the procedural view of "failure through undefined variable" in the program if desired, as illustrated in the following example.

\begin{example}
	The following DC-ProbLog program is equivalent to the procedural interpretation of the program in  Example~\ref{ex:dc-vs-dcproblog}:
	\begin{problog*}{linenos}
n ~ uniform([1,2,3]).
color(1) ~ uniform([red,green,blue]) :- 1=<n .
color(2) ~ uniform([red,green,blue]) :- 2=<n .
color(3) ~ uniform([red,green,blue]) :- 3=<n .
not_red :- not color(2)=:=red.     
not_red :- not 2=<n.               
not_red_either :- 2=<n, color(2)=\=red.
	\end{problog*}
	We explicitly model that \probloginline{not_red} is true if either \probloginline{color(2)} can be interpreted and \probloginline{color(2)=:=red} is false (line 5, which is how \dcproblogsty interprets the first clause), or \probloginline{color(2)} cannot be interpreted (line 6, negating the body of the DC in line 3). Similarly, \probloginline{not_red_either} is true if and only if \probloginline{color(2)} can be resolved and \probloginline{color(2)=\=red} is true (line 7, repeating the body of the DC in line 3).
\end{example}

\subsection{Program validity}
To define valid programs, \citet{gutmann2011magic} impose acyclicity criteria based on the structure of the clauses in the program, whereas we use the ancestor relation between random variables in \dcproblogsty. This means that \dcproblogsty accepts certain cycles in the logic program structure that are rejected by \dcsty, as illustrated in the following example.

\begin{example}
	We model a scenario where a property of a node in a network is either initiated locally with probability $0.1$, or propagated from a neighboring node that has the property with probability $0.3$. We consider a two node network with directed edges from each of the nodes to the other one, and directly ground the program for this situation.
	\begin{problog}
local(n1) ~ flip(0.1).
local(n2) ~ flip(0.1).
transmit(n1,n2) ~ flip(0.3) :- active(n1).
transmit(n2,n1) ~ flip(0.3) :- active(n2).
active(n1) :- local(n1)=:=1.
active(n2) :- local(n2)=:=1.
active(n1) :- transmit(n2,n1)=:=1.
active(n2) :- transmit(n1,n2)=:=1.
	\end{problog}
	This program is not distribution-stratified based on \citep{gutmann2011magic}, where in order to avoid cyclic probabilistic dependencies 1) DC heads have to be of strictly higher rank than any of their body atoms, 2) heads of regular clauses have to have at least the same rank as each body atom, and 3) atoms involving random terms have to have at least the same rank as the head of the DC introducing the random term. This is impossible with our program due to the cyclic dependency between \probloginline{active}-atoms and \probloginline{transmit}-random terms. The \dcproblogsty semantics, in contrast,  is clearly specified through the mapping:
	\begin{problog}
x1 ~ flip(0.1).
x2 ~ flip(0.1).
x3 ~ flip(0.3).
x4 ~ flip(0.3).
rv(local(n1),x1).
rv(local(n2),x2).
rv(transmit(n1,n2),x3) :- active(n1).
rv(transmit(n2,n1),x4) :- active(n2).
active(n1) :- rv(local(n1),x1), x1=:=1.
active(n2) :- rv(local(n2),x2), x2=:=1.
active(n1) :- rv(transmit(n2,n1),x4), x4=:=1.
active(n2) :- rv(transmit(n1,n2),x3), x3=:=1.
	\end{problog}
	We have four independent random variables, and a definite clause program whose meaning is well-defined despite of the cyclic dependencies between derived atoms. We can equivalently rewrite the logic program part to avoid deterministic auxiliaries:
	\begin{problog}
x1 ~ flip(0.1).
x2 ~ flip(0.1).
x3 ~ flip(0.3).
x4 ~ flip(0.3).
active(n1) :- x1=:=1.
active(n2) :- x2=:=1.
active(n1) :- active(n2), x4=:=1.
active(n2) :- active(n1), x3=:=1.
	\end{problog}
	Furthermore, \dcproblogsty agrees with the ProbLog formulation of the original program, \ie,
	\begin{problog}
0.1::local_cause(n1).
0.1::local_cause(n2).
0.3::transmit_cause(n1,n2) :- active(n1).
0.3::transmit_cause(n2,n1) :- active(n2).
active(n1) :- local_cause(n1).
active(n2) :- local_cause(n2).
active(n1) :- transmit_cause(n2,n1).
active(n2) :- transmit_cause(n1,n2).
	\end{problog}
	The AD-free program is
	\begin{problog}
v1 ~ flip(0.1).
local_cause(n1) :- v1=:=1.
v2 ~ flip(0.1).
local_cause(n2) :- v2=:=1.
v3 ~ flip(0.3).
transmit_cause(n1,n2) :- v3=:=1, active(n1).
v4 ~ finite(0.3).
transmit_cause(n2,n1) :- v4=:=1, active(n2).
active(n1) :- local_cause(n1).
active(n2) :- local_cause(n2).
active(n1) :- transmit_cause(n2,n1).
active(n2) :- transmit_cause(n1,n2).
	\end{problog}
	As this already is a \dfplpsty program, we can skip the further rewrites. While the definite clauses are factored differently compared to the earlier variants, their meaning is the same. 
\end{example}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_appendix2/labeled_propositional_to_SMT.tex
aij dcproblog paper

\section{The Equivalence of Labeled Propositional Formulas and SMT Formulas}
\label{app:labeled_prop_to_SMT}

%\pedro{ to be adapted
%First, we rewrite the equation in terms of weighted model integration, which follows from our exposition in Section~\ref{sec:dc2smt}:
%\begin{align}
%    P_\dcpprogram(Q=1|\evidenceset=e) = \frac{\lwmi(\phi_{q\land e},w)}{\lwmi(\phi_{e},w)}
%\end{align}
%The formula $\phi_{e}$ denotes the SMT formula for the ground program with the evidence asserted and $\phi_{q\land e}$ its conjunction with the SMT variable for the query. The symbol $w$ denotes again the joint probability distributions for all the random variables present in $\phi_{q\land e}$ and $\phi_e$, which we refer to as $\{X_1, \dots,  X_V  \}$. This means that $\int w d(\lambda\times \xi \times \mu)=1$. We continue by expressing the conditional probability in terms of a ratio of expectations:
%\begin{align}
%    P_\dcpprogram(Q=1|\evidenceset=e)
%    &= \frac{\int_{\mathcal{M}(\phi_{q\land e})} w d(\lambda\times \xi \times \mu) }{\int_{\mathcal{M}(\phi_{e})} w d(\lambda\times \xi \times \mu)} \\
%    &= \frac{\int \ive{ \mathcal{M}(\phi_{q\land e})} w d(\lambda\times \xi \times \mu) }{\int \ive{\mathcal{M}(\phi_{e})} w d(\lambda\times \xi \times \mu)} \\
%    &= \frac{\E_{w} \left[  \ive{\mathcal{M}(\phi_{q\land e})} \right] }{\E_{w} \left[ \ive{\mathcal{M}(\phi_{e})}  \right]}\\
%    &= \frac{\E_{w} \left[ \sum_{\omega\in ENUM(\phi^\downarrow_{q\land e}) } \alpha(\omega) \right] }{\E_{w} \left[ \sum_{\omega\in ENUM(\phi^\downarrow_{e}) } \alpha(\omega)  \right]}
%\end{align}
%In the last line the $(\omega\in ENUM (\phi^\downarrow_{q\land e})$ denotes an abstracted version of the SMT formula $\phi_{q\land e}^\downarrow$ and similarly for $\phi_e^\downarrow$ (see Example~\ref{ex:abstraction} for an instance of an abstraction).
%Moreover, the last line expresses the initial conditional probability as the ratio of the expected values of $\phi_{q\land e}^\downarrow$ and $\phi_{e}^\downarrow$, respectively, with regards to the variables in the joint probability $w$.
%}

%\ak{this section needs to start with a sentence that provides context -- why are we discussing this? where are we headed?}

In Section~\ref{sec:dc2smt} where we describe the conversions performed by the inference algorithm on a queried \dcproblogsty program we used labeled propositional formulas as an intermediate representation. This, however, ignored the arithmetic meaning inherent to the comparison predicates. We will now refine this our understanding of labeled propositional logic formulas by making the link with weighted model integration~\citep{belle2015probabilistic}.

On the logic formula level we have been treating comparisons, \eg, \probloginline{(temp<25.0)}, as purely logical, with no inherent meaning. For instance, we never exploited facts like $(\text{\probloginline{temp<25}} \land \text{\probloginline{temp>30}}) \leftrightarrow \bot $. Only through the labeling function (cf. Definition~\ref{def:labeling_function}) have comparisons been assigned an arithmetic meaning. For instance, the labeling function {\em interprets} the predicate \probloginline{</2} as an arithmetic comparison and the two arguments (\probloginline{temp}, \probloginline{25.0}) as arithmetic expressions.\footnote{Note that logic programming languages, such as \problogsty, often include a non-logical external arithmetics engine. Through such engines logic programming languages are capable of interpreting arithmetic expressions and comparison predicates, \eg \probloginline{25>5+2} and assign a logical meaning to the comparison~\citep[Section 8]{sterling1994art}. In a sense, probabilistic logic programming languages extend logic programming with probabilistic arithmetics engines.}

Interpreting comparisons can also be performed on the formula level, without the use of labeled atoms. Such formulas are called satisfiability modulo theory (SMT) formulas and extend propositional logic with interpreted first-order predicates. 
% We give a formal definition of SMT formulas in Appendix~\ref{app:wmi} and refer the reader to~\citet{barrett2009handbook} for an in-depth treatment.
% Using the so-called refinement operation one can obtain SMT formulas from Boolean formulas~\citep{barrett2009handbook}.
We refer the reader to~\citet{barrett2009handbook} for a formal definition and in-depth treatment of SMT formulas.
Using the so-called refinement operation one can obtain SMT formulas from Boolean formulas.

\begin{example}[Refinement of a Boolean Formula]\label{example:refinment}
Consider the Boolean formula $\support_g $ from Example~\ref{example:dc2bool}, \ie,:
	\begin{align}
		%\support_g  \nonumber
		%&\leftrightarrow
		&\Big( \phi_{\text{\probloginline{works(1)}}} \leftrightarrow \phi_{\text{\probloginline{cool(1)}}} 
		\lor \phi_\text{\probloginline{hot}} \land  \phi_\text{\probloginline{temp(hot)<25.0}}  
		\lor\neg \phi_\text{\probloginline{hot}}  \land \phi_\text{\probloginline{temp(not_hot)<25.0}}   \Big)  \nonumber \\
	\land ~~	& \Big(
		  \phi_{\text{\probloginline{cool(1)}}} \leftrightarrow  \phi_{\text{\probloginline{rv_cool(1)=:=1}}} \Big) \nonumber \\
	\land ~~ 	& \Big(
		  \phi_\text{\probloginline{hot}} \leftrightarrow \phi_{\text{\probloginline{rv_hot}=:=1}} \Big)
	\end{align}
Refining the Boolean literals that encode a comparison in \dcproblogsty yields the following SMT formula $\support_g ^\uparrow$:
\begin{align}
	%	\support_g ^\uparrow \nonumber
	%	&\leftrightarrow
		&\Big( \phi_{\text{\probloginline{works(1)}}} \leftrightarrow \phi_{\text{\probloginline{cool(1)}}} 
		\lor \phi_\text{\probloginline{hot}} \land  (temp_{hot}{<}25.0)  
		\lor\neg \phi_\text{\probloginline{hot}}  \land  (temp_{not\_hot}{<}25.0)    \Big) \nonumber \\
	\land ~~ 		& \Big( 
		  \phi_{\text{\probloginline{cool(1)}}} \leftrightarrow (rv\_cool_1=1)\Big) \nonumber \\
		\land ~~ 	& \Big( 
		  \phi_\text{\probloginline{hot}} \leftrightarrow  (rv\_hot=1)\Big)
\end{align}
Here the $\uparrow$ in $\support_g ^\uparrow$ indicates that the formula is the refinement of $\support_g $. The formula $\support_g ^{\uparrow}$ is now an SMT formula over the propositional variables 
$\phi_{\text{\probloginline{works(1)}}}$, $\phi_{\text{\probloginline{cool(1)}}}$  and $\phi_\text{\probloginline{hot}}$, the binary SMT variables $rv\_cool_1$ and $rv\_hot$, and the real-valued SMT variables $temp_{hot}$ and $temp_{not\_hot}$.
\end{example}

\begin{example}[Abstraction of SMT Formula]\label{ex:abstraction}
The process of going from an SMT formula back to a propositional logic formula is called abstraction and we denote it by $\downarrow$. If we have, for instance, the SMT formula $\phi$ defined as 
\begin{align*}
    \phi_{\text{\probloginline{cool(1)}}} \lor \phi_\text{\probloginline{hot}} \land  (temp_{hot}{<}25.0)
\end{align*}
we obtain the corresponding propositional formula $\phi^\downarrow$ as:
\begin{align*}
		 \phi_{\text{\probloginline{cool(1)}}} 
		\lor \phi_\text{\probloginline{hot}} \land  \phi_\text{\probloginline{temp(hot)<25.0}}  
\end{align*}
\end{example}

An obvious query to pose is that of the satisfiability of an SMT formula, \ie to figure out whether there is a satisfying assignment to the variables in an SMT formula (\eg~\citep{demoura2009z3}). An equally important query, however, is the weighted counting problem on SMT formulas, which was dubbed {\em weighted model integration}~\citep{belle2015probabilistic}.
% We include a primer on weighted model integration in Appendix~\ref{app:wmi} and refer the reader to the survey paper of~\citet{morettin2021hybrid} for an overview of the field.
We refer the reader to the survey paper of~\citet{morettin2021hybrid} for an overview of the field.
% Recently, \citet{miosic2021measure} also gave a stringent measure theoretic formulation of weighted model integration.

Under the assumption that the values of the variables in an SMT formula are distributed according to a given probability distribution, the weighted model integral of an SMT formula is identical to the expected value of the SMT formula with  respect to the probability distributions of its variables.
This observation gives us the following Lemma~\ref{lemma:expt_label2wmi}.

\begin{lemma}[The Expected Label as a WMI Problem] \label{lemma:expt_label2wmi}
Let $\phi$ denote the logic propositional formula derived from a relevant ground program $\dcpprogram_g$ with evidence asserted, and let $\alpha$ be the labeling function as defined in Definition~\ref{def:labeling_function}. Computing the expected value of the label of $\phi$ is a weighted model integration problem.
\end{lemma}
\begin{proof}[Proof.] We start by expressing the expected value of the label of a specific $\omega$ in terms of Iverson brackets (Equation~\ref{proof:label_wmieq:iverson}). We then rewrite the integral by putting the Iverson brackets in the boundary of the integral (Equation~\ref{proof:label_wmieq:boundary1}) and slightly rewrite the boundary (Equation~\ref{proof:label_wmieq:boundary2}).
\begin{align}
    &\E_{vars(\omega) \sim  \dcpprogram_g} [\alpha( \omega )] \nonumber \\
    &=
    \int
        \left( \prod_{i: a_i \in CA^+(\omega)} \ive{  c_i(vars(a_i)) } \right)
        \left( \prod_{i: a_i \in CA^-(\omega)} \ive{  \neg c_i(vars(a_i)) } \right) 
    dP_{vars(\omega)}(\omega) \label{proof:label_wmieq:iverson} \\
    &= \int_{\{ a_i^\uparrow: a_i \in CA^+(\omega) \} \cup \{a_i^\uparrow : a_j \in CA^-(\omega)\} }     dP_{vars(\omega)}(\omega)   \label{proof:label_wmieq:boundary1} \\
    &= \int_{ \{  a_i^\uparrow : a_i \in ( CA^+(\omega) \cup CA^-(\omega) )   \} } dP_{vars(\omega)}(\omega)  \label{proof:label_wmieq:boundary2}
\end{align}

To get the expected value of the label of $\phi$ we sum over all $\omega \in ENUM(\phi)$ (Equation~\ref{proof:label_wmieq:expected_label_formula}). We then rewrite the expected label in terms of the Lebesgue measure $\lambda$ and a counting measure $\xi$ for integer variables (Equation~\ref{proof:label_wmieq:product_measure}). We refer to~\citep[Section 4.2]{,miosic2021measure} for a stringent measure theoretic formulation of weighted model integration. Note how we now explicitly integrate over the function $w_\omega$ representing the probability distributions and which was hidden before in the measure $P_{vars(\omega)}$. Referring again to~\citep{miosic2021measure}, we equate Equation~\ref{proof:label_wmieq:product_measure} with the measure theoretic formulation of weighted model integration (Equation~\ref{proof:label_wmieq:wmi}).
\begin{align}
    &\sum_{\omega \in ENUM(\phi) } \int_{ \{  a_i^\uparrow : a_i \in ( CA^+(\omega) \cup CA^-(\omega) )   \} } dP_{vars(\omega)}(\omega) \label{proof:label_wmieq:expected_label_formula}   \\
    &= \sum_{\omega \in ENUM(\phi) } \int_{ \{  a_i^\uparrow : a_i \in ( CA^+(\omega) \cup CA^-(\omega) )   \} } \weight_{\omega} d( \lambda \times \xi) \label{proof:label_wmieq:product_measure}\\
    &=\int_{\mathcal{M}(\phi^\uparrow)} w d(\lambda\times \xi \times \mu)\\
    &= \lwmi(\phi^\uparrow,w) \label{proof:label_wmieq:wmi}
\end{align}
In the second but last line the integration goes over all models of the SMT formula $\phi^\uparrow$ using the product measure $\lambda\times \xi \times \mu$, where $\mu$ is a counting measure for the Boolean variables in $\phi^\uparrow$ (cf.~\citep[Definition 8]{,miosic2021measure}). This then corresponds to the definition of the measure theoretic formulation of WMI, denoted in Equation~\ref{proof:label_wmieq:wmi} by $\lwmi$ (the `L' stands for Lebesgue integration).
\end{proof}

Lemma~\ref{lemma:expt_label2wmi} has the delightful property of  allowing us to compute the expected value of the label of a Boolean formula by adopting inference algorithms that have been developed as approaches to weighted model integration~\citep{sanner2011symbolic,morettin2017efficient,kolb2018efficient,zuidbergdosmartires2019exact,kolb2019exploit,zeng2019efficient,derkinderen2020ordering,zeng2020probabilistic}.

% This means that we can reduce inference in \dcproblogsty to weighted model integration.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_appendix2/lp_new.tex
aij dcproblog paper

\section{Logic Programming}
\label{app:lp_new}

We briefly summarize key concepts of the syntax and semantics of logic programming; for a full introduction, we refer to \citep{lloyd2012foundations}.

\subsection{Building Blocks}
The basic building blocks of logic programs are \emph{variables} (denoted by strings starting with upper case letters), \emph{constants}, \emph{functors} and
\emph{predicates} (all denoted by strings starting with lower case letters).  A \emph{term} is a variable, a constant, or a functor
$f$ of \emph{arity} $n$ followed by $n$ terms $t_i$, \ie,
$f(t_1,\dots,t_n)$. 
An \emph{atom} is a predicate $p$ of arity $n$ followed by $n$ terms $t_i$, \ie,
$p(t_1,\dots,t_n)$. A predicate $p$ of arity $n$ is also written as \mathpredicate{p}{n}. A \emph{literal} is an
atom or a negated atom $ not(p(t_1,\dots,t_n))$.

\subsection{Logic Programs}

A \emph{definite clause} is a universally quantified expression of the form $h
\lpif b_1, \dots, b_n$ where $h$ and the $b_i$ are atoms.
$h$ is called the \emph{head} of the clause, and $b_1, \dots, b_n$ its
\emph{body}. Informally, the meaning of such a clause is that if all
the $b_i$ are true, $h$ has to be true as well. 
 A \emph{normal clause}  is a universally quantified expression of the form $h
\lpif l_1, \dots, l_n$ where $h$ is an atom and the $l_i$ are
literals.
If $n=0$, a clause is called \emph{fact} and simply written
as $h$.
A \emph{definite clause program} or \emph{logic program} for
short is a finite set of definite clauses. A \emph{normal logic
  program} is a finite set of normal clauses.

\subsection{Substitutions}

A \emph{substitution} $\theta$ is an expression of the form
$\{V_1/t_1,\dots,V_m/t_m\}$ where the $V_i$ are different variables and
the $t_i$ are terms. Applying a substitution $\theta$ to an expression
$e$ (term or clause) yields the \emph{instantiated} expression $e\theta$
where all variables $V_i$ in $e$ have been simultaneously replaced by
their corresponding terms $t_i$ in $\theta$. If an expression does not
contain variables it is \emph{ground}. Two expressions $e_1$ and $e_2$ can be \emph{unified} if and only if there are substitutions $\theta_1$ and $\theta_2$ such that $e_1\theta_1 = e_2\theta_2$.

\subsection{Herbrand Universe}

The \emph{Herbrand universe} of a logic program is the set of ground terms that can be constructed using the functors and constants occurring in the program.
% \footnote{If the program does not contain constants, one arbitrary constant is added.} 
The \emph{Herbrand base} of a logic program is the set of ground atoms that can be constructed from the predicates in the program and the terms in its Herbrand universe. 
A truth value assignment to all atoms in the 
Herbrand base is called \emph{Herbrand interpretation}, and is also represented as the set of atoms that are true according to the assignment. 
A Herbrand interpretation is a \emph{model} of a clause $h \lpif b_1,\ldots ,b_n\ldotp$ if for every substitution $\theta$ such that all $b_i\theta$ are in the interpretation, $h\theta$ is in the interpretation as well. It is a model of a logic program if it is a model of all clauses in the program. The model-theoretic semantics of a definite clause program is given by its smallest Herbrand model with respect to set inclusion, the so-called \emph{least Herbrand model} (which is unique). We say that a logic program $\dcpprogram$ \emph{entails} an atom $a$, denoted $\dcpprogram\models a$, if and only if $a$ is true in the least Herbrand model of $\dcpprogram$.

% Normal logic programs use the notion of \emph{negation as failure}, that is, for a ground atom $a$, $not(a)$ is true exactly if $a$ cannot be proven in the program. They are not guaranteed to have a unique minimal Herbrand model. Various ways to define the canonical model of such programs have been studied; see, \eg, \citep[Chapter 3]{lloyd2012foundations} for an overview. We use the well-founded semantics here~\citep{van1991well}.

% The main inference task of a logic programming system is to determine
% whether a given atom, also called \emph{query} (or \emph{goal}), is true in the canonical
%  model of a  program. If the answer is yes (or no), we also say that the query \emph{succeeds} (or \emph{fails}). If such a query is not ground, inference asks for the existence of  \emph{answer substitutions}, that is,  substitutions that ground the query into atoms that are part of the canonical model.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_appendix2/syntactic_sugar_semantics.tex
aij dcproblog paper

\section{Detailed Discussion on \dcproblogsty}
\label{sec:detaileddcproblog}

\subsection{Syntactic Sugar Semantics}
\label{sec:semantics_syntactic_sugar}

We now formalize the declarative semantics of \dcproblogsty, \ie \dfplpsty extended with probabilistic facts, annotated disjunctions and distributional clauses,
The idea is to define program transformations that eliminate these three modelling constructs from a \dcproblogsty program, resulting in a \dfplpsty program for which we have defined the semantics in Section~\ref{sec:semantics}.

Throughout this section, we will treat distributional facts as distributional clauses with empty bodies, and we will only consider ground programs for ease of notation. As usual, a non-ground program is shorthand for its Herbrand grounding.

\begin{definition}[Statement]
    A \emph{\dcproblogsty statement} is either a probabilistic fact, an annotated disjunction, a distributional clause, or a normal clause.
\end{definition}

\begin{definition}[\dcproblogsty program] \label{def:fullprog}
    A \dcproblogsty program $\program$ is a countable set of ground \dcproblogsty statements.
\end{definition}

\subsubsection{Eliminating Probabilistic Facts and Annotated Disjunctions}

\begin{example}\label{ex:running-sugar-full}
	We use the following \dcproblogsty program as running example.
	\begin{problog*}{linenos}
p ~ beta(1,1).
p::a.
b ~ normal(3,1) :- a.
b ~ normal(10,1) :- not a.
c ~ normal(b,5).
0.2::d; 0.5::e; 0.3::f :- not b<5, b < 10.
g :- a, not f, b+c<15.
	\end{problog*}
\end{example}

\begin{definition}[Eliminaton Rules for Probabilistic Facts and ADs]\label{def:elim-ad}
	Let $\program$ be  a \dcproblogsty program. We define the following elimination rules (ER) to eliminate probabilistic facts and annotated disjunctions.
	\begin{itemize}
		\item[ER1:] replace each probabilistic fact $p\prob \mu$ in $\program$ by
		      \begin{align*}
			       & \nu \sim flip(p).  \\
			       & \mu \lpif \nu=:=1.
		      \end{align*}
		      with a fresh random variable $\nu$ for each probabilstic fact.
		\item[ER2:] replace each AD $p_1\prob\mu_1;\dots;p_n \prob \mu_n \lpif \beta$ in $\program$ by
		      \begin{align*}
			       & \nu \sim finite([p_1 : 1, \dots, p_n : n]) \\
			       & \mu_1 \lpif \nu=:=1, \beta.                \\
			       & \dots                                      \\
			       & \mu_n \lpif \nu=:=n, \beta.
		      \end{align*}
		      with a fresh random variable $\nu$ for each AD.
	\end{itemize}
\end{definition}
Note that if the probability label(s) of a fact or AD include random terms, as in the case of \probloginline{p::a} in the Example~\ref{ex:running-sugar-full},  then these are  parents of the newly introduced random variable. However, the new random variable will not be a parent of other random variables, as they are only used locally within the new fragments. They thus introduce neither cycles nor infinite ancestor sets into the program.

\begin{definition}[AD-Free Program] \label{def:ad_free_program}
	An \emph{AD-free} \dcproblogsty program \adfreeprogram is a \dcproblogsty program that contains neither probabilistic facts nor annotated disjunctions. We denote by $\headsdc_\adfreeprogram$ the set of atoms $\tau \sim \delta$ that appear as head of a distributional clause in $\adfreeprogram$, and by $\randomtermset_\adfreeprogram$ the set of random terms in $\headsdc_\adfreeprogram$.
\end{definition}

\begin{example}\label{ex:running-sugar-adfree}
	Applying Definition~\ref{def:elim-ad} to Example~\ref{ex:running-sugar-full} results in
	\begin{problog*}{linenos}
p ~ beta(1,1).
x ~ flip(p).
a :- x =:= 1.
b ~ normal(3,1) :- a.
b ~ normal(10,1) :- not a.
c ~ normal(b,5).
y ~ finite([0.2:1,0.5:2,0.3:3]).
d :- y =:= 1, not b<5, b < 10.
e :- y =:= 2, not b<5, b < 10.
f :- y =:= 3, not b<5, b < 10.
g :- a, not f, b+c<15.
	\end{problog*}
	We have $\headsdc_\adfreeprogram=\{ \, \text{\probloginline{p~beta(1,1)}},\,\allowbreak \text{\probloginline{x~flip(p)}},\,\allowbreak  \text{\probloginline{b~normal(3,1)}},\,\allowbreak \text{\probloginline{b~normal(10,1)}},\allowbreak \text{\probloginline{c~normal(b,5)}},\,\allowbreak \text{\probloginline{y~finite[0.2:1,0.5:2.0.3:3])}} \, \}$. Furthermore, we also have $\randomtermset_\adfreeprogram = \{ \mathprobloginline{p}, \mathprobloginline{x}, \mathprobloginline{b}, \mathprobloginline{c}, \mathprobloginline{y} \}$.
\end{example}

\subsubsection{Eliminating Distributional Clauses}
\label{sec:eliminate_dc}

While eliminating probabilistic facts and annotated disjunctions is a rather straightforward local transformation, eliminating distributional clauses is more involved. The reason is that a distributional clause has a global effect in the program, as it defines a condition under which a random term has to be \emph{interpreted} as a specific random variable when mentioned in a distributional clause or comparison atom. Therefore, eliminating a distributional clause involves both introducing the relevant random variable explicitly to the program and pushing the condition from the body of the distributional clause to all the places in the logic program that interpret the original random term.

Before delving into the mapping from an AD-free \dcproblogsty to a \dfplpsty program, we introduce some relevant terminology.

\begin{definition}[Parents and Ancestors for Random Terms]
	\label{def:parentancestor2}
	Given an AD-free program \adfreeprogram with $\tau_p$ and $\tau_c$ in $\randomtermset_\adfreeprogram$. We call $\tau_p$   a \emph{parent} of  $\tau_c$ if and only if  $\tau_p$ appears in the distribution term $\delta_c$ associated with $\tau_c$  in $\headsdc_\adfreeprogram$ ($\tau_c\sim \delta_c \in \headsdc_\adfreeprogram$).
	We define \emph{ancestor} to be the transitive closure of \emph{parent}.
\end{definition}

\begin{figure}[h]
	\centering
	\tikzstyle{node}=[circle, text centered, draw,thick]
	\begin{tikzpicture}[remember picture]

		\node[node] (p) {\probloginline{p}};
		\node[node] (x) [below=of p] {\probloginline{x}};

		\node[node] (b) [right=of p] {\probloginline{b}};
		\node[node] (c) [below=of b] {\probloginline{c}};

		\node[node] (y) [right=of b] {\probloginline{y}};

		\draw[->,thick] (p) to  (x);
		\draw[->,thick] (b) to  (c);

	\end{tikzpicture}
	\caption{Directed acyclic graph representing the ancestor relationship between the random variables in Example~\ref{def:ad_free_program}.
		The random terms \probloginline{p}, \probloginline{b} and \probloginline{y} have the empty set as their ancestor set. The ancestor set of \probloginline{x} is $\{ \mathprobloginline{p} \}$ and \probloginline{c} is $\{ \mathprobloginline{b} \}$.}
\end{figure}

For random terms, we distinguish  \emph{interpreted occurrences} of the term that need to be resolved to the correct random variable from other occurrences where the  random term is treated as any other term in a logic program, \eg, as an argument of a logical atom.
\begin{definition}[Interpreted Occurrence]
	\label{def:interpreted_occ}
	An \emph{interpreted occurrence} of a random term $\tau$ in an AD-free program \adfreeprogram is one of the following:
	\begin{itemize}
		\item the use of $\tau$ as parameter of a distribution term in the head of a distributional clause in \adfreeprogram
		\item the use of $\tau$  in a comparison literal in the body of a (distributional or normal) clause in \adfreeprogram
	\end{itemize}
	We say that a clause \emph{interprets} $\tau$ if there is at least one interpreted occurrence of $\tau$ in the clause.
\end{definition}

\begin{definition}[Well-Defined AD-free Program]\label{def:dc-df-well-def}
	Given an AD-free program \adfreeprogram with $\dclauses_\adfreeprogram$ the set of distributional clauses in \adfreeprogram, we call $\dclauses_\adfreeprogram$ \emph{well-defined} if the following conditions hold:
	\begin{description}
		\item[DC1] For each random term $\tau \in \randomtermset_\adfreeprogram$,  the number of  distributional clauses $\tau \sim \delta \lpif \beta$ in $\adfreeprogram$ is finite, and these clauses all have mutually exclusive bodies. This means that only a single rule can be true at once.
		\item[DC2] All distribution terms in $\dclauses_\adfreeprogram$ are well-defined for all possible values of the random terms they interpret.
		\item[DC3] Each random term has a finite set of ancestors.
		\item[DC4] The ancestor relation is acyclic.
	\end{description}
\end{definition}

% Note that the arguments of a predicate \mathfunctor{\mu}{k} can contain elements of $\distdb$ and $\comparisonfacts$. In this case they will have a purely logical meaning. We discuss this in more detail in Definition~\ref{def:interpreted_occ}.

We now discuss how to reduce a (valid) \dcproblogsty program to a \dfplpsty program. This happens in two steps. First, we eliminate distributional clauses and introduce appropriate distributional facts instead (see Definition~\ref{def:elim-dc}). Second, we {\em contextualize} interpreted occurrences of random terms in clause bodies (see Definition~\ref{def:core_encoding}).

The first step introduces a new built-in predicate \predicate{rv}{2} that associates random terms in a well-defined AD-free program with explicit random variables in the \dfplpsty program it is transformed into. This predicate is used in the bodies of clauses that interpret random terms (\cf Definition~\ref{def:interpreted_occ}) to appropriately contextualize those.

The idea behind the built-in \predicate{rv}{2} predicate is to restrict the applicability of a clause to contexts where all the random terms can be interpreted, \ie to contexts where the random terms are random variables. This implies that in contexts where such a random term cannot be interpreted, the \emph{entire} body evaluates to false.

\begin{definition}[Eliminating Distributional Clauses]\label{def:elim-dc}
	Let $\dclauses_{\adfreeprogram}$ be a well-defined set of distributional clauses.
	We denote by $\delta_{\rho_1, \dots, \rho_k}$ a distribution term that involves exactly $k$ different random terms $\rho_1,\ldots,\rho_k$.
	For each ground random term $\tau\in \randomtermset_{\adfreeprogram}$ we simultaneously define the following sets:
	\begin{itemize}
		\item the set of distributional facts for $\tau$
		      \begin{align*}
			      \distdb(\tau) =
			      \{
			      \tau^\beta_{\nu_1,\ldots,\nu_k}
			       & \sim \delta^\beta_{\nu_1,\ldots,\nu_k}
			      \\
			       & \mid
			      (\tau \sim \delta_{\rho_1,\ldots,\rho_k} \lpif \beta \in \dclauses_\adfreeprogram
			      ,
			      v_1\in \randomvariableset(\rho_1), \ldots, v_k\in \randomvariableset(\rho_k)
			      \}
		      \end{align*}
		\item    the set of (fresh) random variables for $\tau$
		      \begin{equation*}
			      \randomvariableset(\tau) = \{\nu \mid \nu\sim \delta \in \distdb(\tau)\}
		      \end{equation*}
		\item the set of context clauses for $\tau$
		      \begin{align*}
			       & \logicprogram^{c}(\tau)
			      =
			      \\
			       &
			      \quad \biggl\{\text{\probloginline{rv}}(\tau,\tau^\beta_{\nu_1,\ldots,\nu_k})
			      \lpif
			      \text{\probloginline{rv}}(\rho_1,\nu_1),\ldots,\text{\probloginline{rv}}(\rho_k,\nu_k), \beta
			      \\
			       &
			      \quad \bigm|
			      \tau \sim \delta_{\rho_1,\ldots,\rho_k} \lpif \beta \in \dclauses_{\adfreeprogram}, \nu_1\in \randomvariableset(\rho_1), \ldots, \nu_k\in \randomvariableset(\rho_k)
			      \biggr\}
		      \end{align*}
	\end{itemize}
\end{definition}

At first glance, Definition~\ref{def:elim-dc} seems to contain a mutual recursion involving $\distdb(\cdot)$ and $\randomvariableset(\cdot)$.
However, if we recall that for a well-defined set of distributional clauses $\dclauses_\adfreeprogram$ the ancestor relationship between random terms constitutes an acyclic directed graph, the apparent mutual recursion evaporates. We can now define the distributional facts encoding of the distributional clauses, which will give rise to a \dfplpsty program instead of \dcproblogsty program.

\begin{definition}[Distributional Facts Encoding]
	\label{def:core_encoding}
	Let $\adfreeprogram$ be an AD-free \dcproblogsty program and $\dclauses_{\adfreeprogram}$ its set of distributional clauses. We define the distributional facts encoding of $\dclauses_{\adfreeprogram}$  as
	$\dclauses_{\adfreeprogram}^{DF}\coloneqq \distdb\cup \logicprogram^{c}$, with
	\begin{align*}
		\distdb = \bigcup_{\tau\in \randomtermset_{\adfreeprogram}}\distdb(\tau)
		 &  &
		\logicprogram^{c} = \bigcup_{\tau\in \randomtermset_{\adfreeprogram}} \logicprogram^{c}(\tau)
	\end{align*}
	using $\distdb(\cdot)$ and $\logicprogram^{c}(\cdot)$ from Definition~\ref{def:elim-dc}.
\end{definition}

\begin{example}[Eliminating Distributional Clauses]
	\label{ex:running-sugar-core-encoding}
	We demonstrate the elimination of distributional clauses using the DCs in Example~\ref{ex:running-sugar-adfree}, \ie
	\begin{problog*}{linenos}
p ~ beta(1,1).
x ~ flip(p).
b ~ normal(3,1) :- a.
b ~ normal(10,1) :- not a.
c ~ normal(b,5).
y ~ finite([0.2:1,0.5:2,0.3:3]).
	\end{problog*}
	Here,  the distribution terms in Line 2 and Line 5 (\probloginline{flip(p)} and \probloginline{normal(b,5)}) contain one parent random term each (\probloginline{p} and \probloginline{b}, respectively), whereas all others have no parents. As \probloginline{b} is defined by two clauses, we get fresh random variables for each of them, which in turn  introduces different fresh random variables for the child \probloginline{c}.
	This gives us:
	\begin{problog*}{linenos}
v1 ~ beta(1,1).
rv(p,v1).
v2 ~ flip(v1).
rv(x,v2) :- rv(p,v1).
v3 ~ normal(3,1).
rv(b,v3) :- a.
v4 ~ normal(10,1).
rv(b,v4) :- not a.
v5 ~ normal(v3,5).
rv(c,v5) :- rv(b,v3).
v6 ~ normal(v4,5).
rv(c,v6) :- rv(b,v4).
v7 ~ finite([0.2:1,0.5:2,0.3:3]).
rv(y,v7).
	\end{problog*}
\end{example}

Eliminating distributional clauses (following Definition~\ref{def:elim-dc}) introduces the distributional facts and context rules necessary to encode the original distributional clauses. To complete the transformation to a \dfplpsty program, we further transform the logical rules. Prior to that, however, we need to define the {\em contextualization function}.

\begin{definition}[Contextualization Function] \label{def:context_function}
	Let $\beta$ be a conjunction of atoms and let its comparison literals interpret the random terms $\tau_1,\ldots,\tau_n$.
	Furthermore, let $\Lambda_i$ be a special logical variable associated to a random term $\tau_i \in \randomtermset_{\adfreeprogram}$ for each $\tau_i$.
	We define $\contextfunc(\beta)$ to be the  conjunction of literals obtained by replacing the interpreted occurrences of the $\tau_i$ in $\beta$ by their corresponding $\Lambda_i$
	and conjoining to this modified conjunction $\text{\probloginline{rv}}(\tau_i,\Lambda_i)$ for each $\tau_i$.
	We call $\contextfunc(\cdot)$ the contextualization function.
\end{definition}

\begin{definition}[Contextualized Rules]\label{def:adfree-to-core}
	Let \adfreeprogram be an AD-free program with logical rules $\logicprogram^\adfreeprogram$ and distributional clauses $\dclauses_\adfreeprogram$,
	and let $\dclauses_\adfreeprogram^{DF}=\distdb\cup \logicprogram^{c}$ be the distributional facts encoding of $\dclauses_\adfreeprogram$. We define the contextualization of the bodies of the rules $\logicprogram^\adfreeprogram \cup \logicprogram^{DF}$ as the sequential application of the following contextualization rules:
	\begin{enumerate}[label=\alph*.]
		\item[CR1:] apply the contextualization function $\contextfunc$ to all bodies in $\logicprogram^\adfreeprogram\cup \logicprogram^{c}$ and obtain:
		      \begin{align*}
			      \logicprogram^\Lambda = \{\eta \lpif \contextfunc(\beta) \mid \eta \lpif \beta \in \logicprogram^\adfreeprogram \cup \logicprogram^{c} \}
		      \end{align*}
		\item[CR2:] obtain the set of ground logical rules $\logicprogram$ by grounding each logical variable $\Lambda_i$ in $\logicprogram^\Lambda$  with random variables $\nu_i\in \randomvariableset(\tau_i)$ in all possible ways.
	\end{enumerate}
	We call $\logicprogram$ the contextualized logic program of $\adfreeprogram$, \fixed{Furthermore, we define $\dfadfreeprogram\coloneqq\distdb \cup \logicprogram $ to be the corresponding \dcplpsty program.}
\end{definition}

The contextualization function $\contextfunc(\cdot)$ creates non-ground comparison atoms, \eg $\Lambda_i>5$. Contrary to (ground) random terms, non-ground logical variables in such a comparison atom are not interpreted occurrences (\cf~Definition~\ref{def:interpreted_occ}) and the comparison itself only has a logical meaning. By grounding out the freshly introduced logical variables we obtain a purely logical program where the comparison atoms contain either arithmetic expressions or random variables (instead of random terms).

\begin{example}[Contextualizing Random Terms]
	\label{example:contextualization}
	Let us now study the effect of the second transformation step.
	Consider again the AD-free program in Example~\ref{ex:running-sugar-adfree} and the set of rules and distributional clauses obtained in Example~\ref{ex:running-sugar-core-encoding}.
	The contextualization rule CR1 (\cf Definition~\ref{def:adfree-to-core})  rewrites the logical rules in the AD-free input program to
	% \linelabel{line:comment_r2_t2a_rule}
	\begin{problog*}{linenos}
a :- rv(x,Lx), Lx =:= 1.
d :- rv(y,Ly), rv(b,Lb), Ly =:= 1, not Lb<5, Lb < 10.
e :- rv(y,Ly), rv(b,Lb), Ly =:= 2, not Lb<5, Lb < 10.
f :- rv(y,Ly), rv(b,Lb), Ly =:= 3, not Lb<5, Lb < 10.
g :- rv(b,Lb), rv(c,Lc), a, not f, Lb+Lc < 15.
	\end{problog*}
	These rules then get instantiated (rule CR2) to
	\begin{problog*}{linenos}
a :- rv(x, v2), v2 =:= 1.
d :- rv(y,v7), rv(b,v3), v7 =:= 1, not v3<5, v3 < 10.
e :- rv(y,v7), rv(b,v3), v7 =:= 2, not v3<5, v3 < 10.
f :- rv(y,v7), rv(b,v3), v7 =:= 3, not v3<5, v3 < 10.
d :- rv(y,v7), rv(b,v4), v7 =:= 1, not v4<5, v4 < 10.
e :- rv(y,v7), rv(b,v4), v7 =:= 2, not v4<5, v4 < 10.
f :- rv(y,v7), rv(b,v4), v7 =:= 3, not v4<5, v4 < 10.
g :- rv(b,v3), rv(c,v5), a, not f, v3+v5<15.
g :- rv(b,v3), rv(c,v6), a, not f, v3+v6<15.
g :- rv(b,v4), rv(c,v5), a, not f, v4+v5<15.
g :- rv(b,v4), rv(c,v6), a, not f, v4+v6<15.
	\end{problog*}
	Together with the distributional facts and rules obtained in Example~\ref{ex:running-sugar-core-encoding}, this last block of rules forms the \dcplpsty program that specifies the semantics of the AD-free \dcproblogsty program, and thus the semantics of the \dcproblogsty program in Example~\ref{ex:running-sugar-full}.
\end{example}

We note that the mapping from an AD-free program to a set of distributional facts and contextualized rules as defined here is purely syntactical, and written to avoid case distinctions.
Therefore, it usually produces overly verbose programs.
For instance, for random terms introduced by a distributional fact, the indirection via \probloginline{rv} is only needed if there is a parent term in the distribution that has context-specific interpretations.
The grounding step may introduce rule instances whose conjunction of \probloginline{rv}-atoms is inconsistent.
This is for example the case for the last three rules for \probloginline{g} in the Example~\ref{example:contextualization}, which we illustrate in the example below.

\begin{example}\label{ex:running-sugar-simplified}
	The following is a (manually) simplified version of the \dfplpsty program for the running example, where we propagated definitions of  \probloginline{rv}-atoms:
	\begin{problog*}{linenos}
v1 ~ beta(1,1).
v2 ~ flip(v1).
v3 ~ normal(3,1).
v4 ~ normal(10,1).
v5 ~ normal(v3,5).
v6 ~ normal(v4,5).
v7 ~ finite([0.2:1,0.5:2,0.3:3]).

a :- v2 =:= 1.
d :- a,     v7 =:= 1, not v3<5, v3 < 10.
e :- a,     v7 =:= 2, not v3<5, v3 < 10.
f :- a,     v7 =:= 3, not v3<5, v3 < 10.
d :- not a, v7 =:= 1, not v4<5, v4 < 10.
e :- not a, v7 =:= 2, not v4<5, v4 < 10.
f :- not a, v7 =:= 3, not v4<5, v4 < 10.
g :- a,     a,     a, not f, v3+v5<15.
g :- a,     not a, a, not f, v3+v6<15.  % inconsistent
g :- not a, a,     a, not f, v4+v5<15.  % inconsistent
g :- not a, not a, a, not f, v4+v6<15.  % inconsistent
	\end{problog*}
	In the bodies of the last three rules we have, inter alia, conjunctions of \probloginline{a} and \probloginline{not a}. This can never be satisfied and renders the bodies of these rules inconsistent.
\end{example}

\begin{definition}[Semantics of AD-free \dcproblogsty Programs]
	\label{def:semantics_adfree}
	The semantics of an AD-free \dcproblogsty program \adfreeprogram is the semantics of the \dfplpsty program $ \dfadfreeprogram = \distdb \cup \logicprogram$. We call $\adfreeprogram$ valid if and only if $\dfadfreeprogram$ is valid.
\end{definition}

\begin{definition}[Semantics of $\dcproblogsty$ Programs]
	The semantics of a $\dcproblogsty$ program $\program$ is the semantics of the AD-free \dcproblogsty program $\adfreeprogram$. We call $\program$ valid if and only if $\adfreeprogram$ is valid.
\end{definition}

Programs with distributional clauses can make programs with combinatorial structures more readable by grouping random variables with the same role  under the same random term. However, the programmer needs to be aware of the fact that distributional clauses have non-local effects on the program, as they affect the interpretation of their random terms also outside the distributional clause itself. This can be rather subtle, especially  if the bodies of the distributional clauses with the same random term are not exhaustive.
We discuss this issue in more detail in Appendix~\ref{sec:non-mixture-dc}.

\subsection{Syntactic Sugar: Validity}
\label{sec:dcvalidity}

As stated above, a DC-ProbLog program $\program$  is syntactic sugar for an AD-free program $\adfreeprogram$ (Definition~\ref{def:elim-ad}),
and is valid if $\dfadfreeprogram$ as specified in Definition~\ref{def:semantics_adfree} is a valid \dfplpsty program, \ie the distributional database is well-defined,
the comparison literals are measurable,
and each consistent fact database results in a two-valued  well-founded  model if added to the logic program (Definition~\ref{def:core-valid}).
For the distributional database to be well-defined (Definition~\ref{def:well-defd-facts}), it suffices to have $\dclauses_\adfreeprogram$ well-defined  (Definition~\ref{def:dc-df-well-def}), as can be verified by comparing the relevant definitions. Indeed, a well-defined $\dclauses_\adfreeprogram$ is a precondition for the transformation as stated in the definition.

The transformation changes neither distribution terms nor comparison literals, and thus maintains the measurability of the latter.
% \linelabel{line:comment_grammar_r2}
As far as the logic program structure is concerned, the transformation to a \dfplpsty adds rules for \probloginline{rv} based on the bodies of all distributional clauses, and uses positive \probloginline{rv} atoms in the bodies of all clauses  that interpret random terms to ensure that all interpretations of random variables are anchored in the appropriate parts of the distributional database. This level of indirection does not affect the logical reasoning for programs that only interpret random terms in appropriate contexts. It is the responsibility of the programmer to ensure that this is the case and indeed results in appropriately defined models.

\subsection{Syntactic Sugar: Additional Constructs}
\label{sec:Additionalsyntacticsugar}

\subsubsection{User-Defined Sample Spaces}
The semantics of \dcproblogsty as presented in the previous sections only alllows for random variables with numerical sample spaces, \eg normal distributions, or Poisson distributions. For categorical random variables, however, one might like to give a specific meaning to the elements in the sample space instead of a numerical value.
\begin{example}
	Consider the following program:
	\begin{problog*}{linenos}
color ~ uniform([r,g,b]).
red:- color=:=r.
	\end{problog*}
	Here we discribe a categorical random variable (uniformaly distributed) whose sample space is the set of expressions $\{\text{\probloginline{r}},\text{\probloginline{b}},\text{\probloginline{g}}\}$. By simply associating a natural number to each element of the sample space we can map the program back to a program whose semantics we already defined:
	\begin{problog*}{linenos}
color ~ uniform([1,2,3]).
r:- color=:=1,
red:- r.
	\end{problog*}
\end{example}

Swapping out the sample space of discrete random variables with natural numbers is always possible as the cardinality of such a sample space is either smaller (finite categorical) or equal (infinite) to the cardinality of the natural numbers.

\subsubsection{Multivariate Distributions}

Until now we have restricted the syntax and semantics of \dcproblogsty to univariate distributions, \eg the univariate normal distribution. At first this might seem to severely limit the expressivity of \dcproblogsty, as probabilistic modelling with multivariate random variables is a common task in modern statistics and probabilistic programming. However, this concern is voided by realizing that multivariate random variables can be decomposed into {\em combinations} of independent univariate random variables. We will illustrate this on the case of the bivariate normal distribution.

\begin{example}[Constructing the Bivariate Normal Distribution]
	Assume we would like to construct a random variable distributed according to a bivariate normal distribution:
	\begin{align*}
		\begin{pmatrix} \nu_1 \\ \nu_2 \end{pmatrix}
		\sim
		\mathcal{N}\left(
		\begin{pmatrix}
				\mu_1 \\  \mu_2
			\end{pmatrix}
		,
		\begin{pmatrix}
				\sigma_{11} & \sigma_{12} \\
				\sigma_{21} & \sigma_{22} \\
			\end{pmatrix}
		\right)
	\end{align*}
	The equation above can be rewritten as:
	\begin{align*}
		\begin{pmatrix} \nu_1 \\ \nu_2 \end{pmatrix}
		\sim
		\begin{pmatrix}
			\mu_1 \\  \mu_2
		\end{pmatrix}
		+
		\begin{pmatrix}
			\eta_{11} & \eta_{12} \\
			\eta_{21} & \eta_{22} \\
		\end{pmatrix}
		\begin{pmatrix}
			\mathcal{N}(0, \lambda_1) \\
			\mathcal{N}(0, \lambda_2)
		\end{pmatrix}
	\end{align*}
	where it holds that
	\begin{align*}
		\begin{pmatrix}
			\sigma_{11} & \sigma_{12} \\
			\sigma_{21} & \sigma_{22} \\
		\end{pmatrix}
		=
		\begin{pmatrix}
			\eta_{11} & \eta_{12} \\
			\eta_{21} & \eta_{22} \\
		\end{pmatrix}
		\begin{pmatrix}
			\lambda_1 & 0         \\
			0         & \lambda_2
		\end{pmatrix}
		\begin{pmatrix}
			\eta_{11} & \eta_{21} \\
			\eta_{12} & \eta_{22} \\
		\end{pmatrix}
	\end{align*}
	It can now be shown that the bivariate distributions can be expressed as:
	\begin{align*}
		\begin{pmatrix} \nu_1 \\ \nu_2 \end{pmatrix}
		\sim
		\begin{pmatrix}
			\mathcal{N}(\mu_{\nu_1}, \sigma_{\nu_1}) \\
			\mathcal{N}(\mu_{\nu_2}, \sigma_{\nu_2}) \\
		\end{pmatrix}
	\end{align*}
	where $\mu_{\nu_1}$, $\mu_{\nu_2}$, $\sigma_{\nu_1}$ and $\sigma_{\nu_2}$ can be expressed as:
	\begin{align*}
		\mu_{\nu_1} & = \mu_1 \qquad
		            & \sigma_{\nu_1} = \sqrt{ \eta_{11}\lambda_1^2 + \eta_{12}\lambda_2^2 } \\
		\mu_{\nu_2} & = \mu_2 \qquad
		            & \sigma_{\nu_2} = \sqrt{ \eta_{21}\lambda_1^2 + \eta_{22}\lambda_2^2 }
	\end{align*}
	We conclude from this that a bivariate normal distribution can be modeled using two univariate normal distributions that have a shared set of parameters and is thereby semantically defined in \dcproblogsty.

\end{example}

Expressing multivariate random variables in a user-friendly fashion in a probabilistic programming language is simply a matter of adding syntactic sugar for combinations of univariate random variables once the semantics are defined for the latter.

\begin{example}[Bivariate Normal Distribution]
	Possible syntactic sugar to declare a bivariate normal distribution in \dcproblogsty, where the mean of the distribution in the two dimensions is $0.5$ and $2$, and the covariance matrix is
	$
		\begin{bmatrix}
			2   & 0.5 \\
			0.5 & 1
		\end{bmatrix}
	$.
	\begin{problog*}{linenos}
(x1,x2) ~ normal2D([0.5,2], [[2, 0.5],[0.5,1]])
q:- x1<0.4, x2>1.9.
	\end{problog*}
\end{example}

On the inference side, the special syntax might then additionally be used to deploy dedicated inference algorithms. This is usually done in probabilistic programming languages that cater towards inference with multivariate (and often continuous) random variables~\citep{carpenter2017stan,bingham2019pyro}. Note that probability distributions are usually constructed by applying transformations to a set of independent uniform distribution. From this viewpoint the builtin-in \probloginline{normal/2}, denoting the univariate normal distribution, is syntactic sugar for such a transformation as well.

\subsection{Beyond Mixtures}\label{sec:non-mixture-dc}
\label{sec:beyondmixtures}

By definition, we impose on distributional clauses mutual exclusivity of their bodies when they share a random term (\cf Definition~\ref{def:dc-df-well-def}). That is, if we have a set of distributional clauses: $\{ \tau \sim \delta_i\lpif \beta_1 \dots, \tau \sim \delta_n\lpif \beta_n  \}$ we impose that the conjunction of two distinct bodies $\beta_i$ and $\beta_j$ ($i \neq j$) is false.

A further condition that we might impose, which is, however, not necessary to define a valid distributional clause, is exhaustiveness. Let us consider again the set of distributional clauses  $\{ \tau \sim \delta_i\lpif \beta_1 \dots, \tau \sim \delta_n\lpif \beta_n  \}$. We call this set exhaustive if the disjunction of all the $\beta_i$'s is equivalent to true.

A set of exhaustive distributional clauses can be interpreted as a mixture models as they assign a unique distribution to the random term in any possible context. When, the bodies of such distributional clauses are not exhaustive, however, they may
interact with the logic program in rather subtle ways, especially if negation is involved. We demonstrate this in the examples below.

\begin{example}
	Consider the  following program fragments
	\begin{problog}
q :- not (x=:=1).
	\end{problog}
	and
	\begin{problog}
aux :- x=:=1.
q :- not aux.
	\end{problog}
	and now assume \probloginline{x} follows a mixture distribution, \eg,
	\begin{problog}
0.2::b.
x~flip(0.5) :- b.
x~flip(0.9) :- not b.
	\end{problog}
	With such a mixture model, as in the case of a distributional fact, "\probloginline{x} has an associated distribution" is always true, and both fragments agree on the truth value of \probloginline{q}.
\end{example}
In general, however, only the first of these two conditions is necessary, and it is thus possible to  associate a distribution with a random term in \emph{some} contexts only.

\begin{example}
	Consider again the two program fragments above with the following non-exhaustive definition of \probloginline{x}:
	\begin{problog}
0.2::b.
x~flip(0.5) :- b.
	\end{problog}
	With this definition, "\probloginline{x} has an associated distribution" is true if and only if \probloginline{b} is true, and the two fragments therefore no longer agree on the truth values of \probloginline{q}, as we more easily see after eliminating the distributional clause. We omit the transformation of the probabilistic fact for brevity. The fragment defining \probloginline{x} transforms to
	\begin{problog}
v1~flip(0.5).
rv(x,v1) :- b.
	\end{problog}
	The first program fragment maps to
	\begin{problog}
q :- rv(x,v1), not (v1=:=1).
	\end{problog}
	and the second one to
	\begin{problog}
aux :- rv(x,v1), v1=:=1.
q :- not aux.
	\end{problog}
	which clearly exposes the difference in how the negation is interpreted.
\end{example}
As this example illustrates, if random variables are defined through non-exhaustive sets of DCs, we can no longer refactor the logic program independently of the definition of the random variables in general, as it interacts with the context structure. The reason is that \dcproblogsty's declarative semantics builds upon the principle that the distributional database is declared \emph{independently} of the logic program, and can thus be combined modularly and declaratively with \emph{any} logic program over its comparison atoms. This is no longer the case with such arbitrary sets of DCs, which intertwine the definition of the two parts of a \dfplpsty program. We note that this differs from the procedural view on the existence of random variables taken in the \dcsty language~\citep{nitti2016probabilistic}, as we
discuss in more detail in Appendix~\ref{sec:dcproblog-dc}.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_appendix2/transformation_proofs.tex
aij dcproblog paper

\section{Proofs of Theorems and Propositions in Section~\ref{sec:dc2smt} and Section~\ref{sec:alw}}
\label{app:transformation_proofs}

\subsection{Proof of Theorem~\ref{theo:rgp}} \label{app:proof:rgp}

\theorgp*

\begin{proof}
    %The proof is analogous  to the one for Theorem~1 in~\citep[Appendix A]{fierens2015inference}, as it only concerns the logic part of the program.
    The semantics of $\dcpprogram$ is given by the ground program that is obtained by first grounding $\dcpprogram$ with respect to its Herbrand base and reducing it to a \dfplpsty program as specified in Section~\ref{sec:dcproblog}. The resulting program consists of distributional facts and ground normal clauses only, and includes clauses defining \probloginline{rv}-atoms as well as calls to those atoms in clause bodies. However, as the definitions of these atoms are acyclic, and each ground instance is defined by a single rule (with the body of the DC that introduced the new random variable), we can eliminate all references to such atoms by recursively applying the well-known \emph{unfolding} transformation, which replaces atoms in clause bodies by their definition. The result is an equivalent ground program using only predicates from \dcpprogram, but where rule bodies have been expanded with the contexts of the random variables they interpret. We know from Theorem~1 in \citep{fierens2015inference} that for given query and evidence, it is sufficient to use the part of this logic program that is encountered during backward chaining from those atoms. We note that in our case, this also includes the distributional facts providing the distributions for relevant random variables, \ie, random variables in relevant comparison atoms as well as their ancestors.
\end{proof}

\subsection{Proof of Theorem~\ref{theo:label_equivalence}}
\label{app:proof:label_equivalence}

\Labelequivalence*

\begin{proof} The probability of a model $\varphi$ of the relevant ground program $\dcpprogram_g$ is, according to the distribution semantics (cf. Appendix~\ref{app:proof:pf}), given by:
    \begin{align}
        P_{\dcpprogram_g}(\varphi) = \int \mathbf{1}_{[\mu_1=\boolval_1\wedge\ldots\wedge \mu_n=\boolval_n]}(\samplefunction(\randomvariableset)) \differential{\probabilitymeasure_{\randomvariableset}}
    \end{align}
    where the $\mu_i$ are the comparison atoms that appear (positively or negatively) in $\dcpprogram_g$ and the $b_i$ the truth values these atoms take in $\varphi$.
    We can manipulate the probability into:
    \begin{align}
        P_{\dcpprogram_g}
         & =
        \int \left( \prod_{i =1}^n \mathbf{1}_{[\mu_i=b_i]}(\samplefunction(\randomvariableset)) \right) \differential{\probabilitymeasure_{\randomvariableset}}                                                                                                         \\
         & =
        \int \left( \prod_{i: b_i=\bot} \mathbf{1}_{[\mu_i=b_i]}(\samplefunction(\randomvariableset)) \right)  \left( \prod_{i: b_i=\top} \mathbf{1}_{[\mu_i=b_i]}(\samplefunction(\randomvariableset)) \right)  \differential{\probabilitymeasure_{\randomvariableset}} \\
         & =
        \int \left( \prod_{i: b_i=\bot} \ive{\neg c_i(vars(\mu_i))} \right)  \left( \prod_{i: b_i=\top}  \ive{ c_i(vars(\mu_i))}  \right)  \differential{\probabilitymeasure_{\randomvariableset}}
        \label{eq:program_prob_label_equi}
    \end{align}
    Turning our attention now to the expected value of $\alpha(\varphi)$ we have:
    \begin{align}
        \E_{\randomvariableset \sim  \dcpprogram_g} [\alpha( \varphi )]
        =
        \int \alpha \left( \bigwedge_{\ell_i \in \varphi} \ell_i \right) \differential{P_\randomvariableset}
        =
        \int  \left( \prod_{\ell_i \in \varphi} \alpha \left(  \ell_i \right) \right)  \differential{P_\randomvariableset}
    \end{align}
    The literals $\ell_i \in \varphi$ fall into four groups: atoms whose predicate is a comparison and that are true in $\varphi$ (denoted by $CA^+(\varphi)$), non-comparison atoms that are true in $\varphi$ (denoted $NA^+(\varphi)$), and similarly the atoms that are false in $\varphi$ (denoted by $CA^-(\varphi)$ and $NA^-(\varphi)$). This yields:
    \begin{align}
         & \E_{\randomvariableset \sim  \dcpprogram_g} [\alpha( \varphi )] \\
         & =
        \int
        \left( \prod_{\ell_i \in CA^+(\varphi)} \alpha (  \ell_i ) \right)
        \left( \prod_{\ell_i \in CA^-(\varphi)} \alpha (  \neg \ell_i ) \right)
        \left( \prod_{\ell_i \in NA^+(\varphi)} \alpha (  \ell_i ) \right)
        \left( \prod_{\ell_i \in NA^-(\varphi)} \alpha (  \neg \ell_i ) \right)
        \nonumber
        \differential{P_\randomvariableset}
    \end{align}
    Plugging in the definition of the labeling function the last two products reduce to $1$ and we obtain for the remaining expression:
    \begin{align}
        \E_{\randomvariableset \sim  \dcpprogram_g} [\alpha( \varphi )]
        =
        \int
        \left( \prod_{i: \ell_i \in CA^+(\varphi)} \ive{  c_i(vars(\ell_i)) } \right)
        \left( \prod_{i: \ell_i \in CA^-(\varphi)} \ive{  \neg c_i(vars(\ell_i)) } \right)
        \differential{P_\randomvariableset}
        \label{eq:formula_prob_lable_equi}
    \end{align}
    Identifying now the set $\{ i: \ell_i \in CA^+(\varphi) \}$ with the set $\{i : \mu_i=\top   \} $ and the set  $\{ i: \ell_i \in CA^-(\varphi) \}$ with the set $\{i : \mu_i=\bot   \} $ proves the theorem, as this equates Equation~\ref{eq:program_prob_label_equi} and Equation~\ref{eq:formula_prob_lable_equi}
\end{proof}

\subsection{Proof of Proposition~\ref{prop:mcapproxconditional}}
\label{app:proof:mcapproxconditional}

\mcapproxconditional*

\begin{proof}
    First we write the conditional probability as a ratio of expected values invoking Theorem~\ref{thm:inference-by-expectation}, on which we then use Defintion~\ref{def:label_bool_formula}:
    \begin{align}
        P_\dcpprogram(\mu=q\mid\evidenceset = e)
         & = \frac{\E_{\randomvariableset \sim  \dcpprogram_g} [\alpha( \phi \wedge \phi_q)] }{\E_{\randomvariableset \sim  \dcpprogram_g} [\alpha( \phi )] }
        \\
         & =
        \frac{
            \E_{\randomvariableset \sim  \dcpprogram_g} \left[ \sum_{\varphi\in ENUM(\phi \land \phi_{q}) } \prod_{\ell \in \varphi} \alpha(\ell) \right]
        }
        {
            \E_{\randomvariableset \sim  \dcpprogram_g} \left[ \sum_{\varphi\in ENUM(\phi) } \prod_{\ell \in \varphi} \alpha(\ell)  \right]
        }
        \\
         & =
        \frac{
            \E_{\randomvariableset \sim  \dcpprogram_g} \left[ \sum_{\varphi\in ENUM(\phi \land \phi_{q}) } \alpha(\varphi) \right]
        }
        {
            \E_{\randomvariableset \sim  \dcpprogram_g} \left[ \sum_{\varphi\in ENUM(\phi) } \alpha(\varphi)  \right]
        }
    \end{align}

    We can now express the conditional probability in terms of the sampled values $\mathcal{S}$:
    \begin{align}
        P_\dcpprogram(\mu=q\mid\evidenceset=e)
           & =
        \frac{
        \lim_{ \lvert \mathcal{S} \rvert \rightarrow \infty} \nicefrac{1}{\lvert \mathcal{S} \rvert} \sum_{i=1}^{\lvert \mathcal{S} \rvert}  \sum_{\varphi\in ENUM(\phi \land \phi{q}) } \alpha^{(i)}(\varphi)
        }
        {
        \lim_{\lvert \mathcal{S} \rvert \rightarrow \infty} \nicefrac{1}{\lvert \mathcal{S} \rvert} \sum_{i=1}^{\lvert \mathcal{S} \rvert} \sum_{\varphi\in ENUM(\phi) } \alpha^{(i)}(\varphi)
        }                                                                   \\
           & \approx
        \frac{
        \sum_{i=1}^{\lvert \mathcal{S} \rvert}  \sum_{\varphi\in ENUM(\phi \land \phi_{q}) } \alpha^{(i)}(\varphi)
        }
        {
        \sum_{i=1}^{\lvert \mathcal{S} \rvert} \sum_{\varphi\in ENUM(\phi) } \alpha^{(i)}(\varphi)
        }, & \quad \lvert \mathcal{S} \rvert<\infty \label{eq:mc_prob_cond}
    \end{align}
\end{proof}

\subsection{Proof of Proposition~\ref{prop:alw_consistency}}
\label{app:proof:alw_consistency}

\alwconsistency*
\begin{proof}
    First we manipulate the expected value on the left hand side of Equation~\ref{eq:ALW}:
    \begin{align}
         & \E \left[ \sum_{\varphi \in ENUM(\phi)} \prod_{\ell \in \varphi}  \alpha \left(\ell \right) \bigg| \mathcal{S} \right]                                                           \\
         & =
        \lim_{ \lvert \mathcal{S} \rvert\rightarrow\infty} \sum_{i=1}^{ \lvert \mathcal{S} \rvert}  \sum_{\varphi \in ENUM(\phi)} \prod_{\ell \in \varphi}  \alpha^{(i)} \left(\ell \right) \\
         & =
        \lim_{ \lvert \mathcal{S} \rvert \rightarrow\infty} \sum_{i=1}^{ \lvert \mathcal{S} \rvert}  \sum_{\varphi \in ENUM(\phi)}
        \left(\prod_{\ell \in \varphi \setminus DI(\varphi)}  \alpha^{(i)} \left(\ell \right) \prod_{\ell \in DI(\varphi)}  \alpha^{(i)} \left(\ell \right)   \right)
    \end{align}
    As the samples are ancestral samples, they satisfy by construction the delta invervals appearing in the second product. This means that $\prod_{\ell \in DI(\varphi)}  \alpha^{(i)} \left(\ell \right) =1$ and that we can write the expected value in function of non delta interval atoms only:
    \begin{align}
        \E \left[ \sum_{\varphi\in ENUM(\phi)} \prod_{\ell \in \varphi}  \alpha \left(\ell \right) \bigg| \mathcal{S} \right]
        =
        \E \left[ \underbrace{\sum_{\varphi\in ENUM(\phi)} \prod_{\ell \in \varphi \setminus DI(\varphi)}  \alpha \left(\ell \right)}_{\coloneqq f(\phi)} \bigg| \mathcal{S} \right]  \label{eq:ALW_exp_simplified}
    \end{align}
    Let us now manipulate the expression in the numerator on the right hand side of Equation~\ref{eq:ALW}:
    \begin{align}
         & \bigoplus_{i=1}^{ \lvert \mathcal{S} \rvert}  \bigoplus_{\varphi \in ENUM(\phi)} \bigotimes_{\ell \in \varphi}  \alpha_{IALW}^{(i)} \left(\ell \right) \\
         & =
        \bigoplus_{i=1}^{ \lvert \mathcal{S} \rvert}  \bigoplus_{\varphi \in ENUM(\phi)}
        \underbrace{\left(  \bigotimes_{\ell \in \varphi \setminus DI(\varphi)}  \alpha_{IALW}^{(i)} \left(\ell \right) \right) }_{\left( r_\varphi^{(i)} , 0  \right)}  \otimes
        \underbrace{\left( \bigotimes_{\ell \in DI(\varphi)}  \alpha_{IALW}^{(i)} \left(\ell \right) \right) }_{\left( t_\varphi^{(i)} , m_\varphi^{(i)}  \right)} \label{eq:alw_proof_intermediate_1}
    \end{align}
    The expressions $\left( r_\varphi^{(i)} , 0  \right)$ and $\left( t_\varphi^{(i)} , m_\varphi^{(i)}  \right)$ denote infinitesimal numbers. Note how only the latter of the two picks up a non-zero second part.

    From the definition of the addition of two infinitesimal numbers we can see that only those infinitesimal numbers with the smallest integer in the second part {\em survive} the addition. This also means that in Equation~\ref{eq:alw_proof_intermediate_1} only those terms that have the smallest integer in their second part among all terms will contribute. We denote this smallest integer by:
    \begin{align}
        m^* =  \min_{ \substack{i\in \{ 1, \dots,  \lvert \mathcal{S} \rvert \} \\ \varphi\in ENUM(\phi) }} m^{(i)}_\varphi
    \end{align}
    We rewrite Equation~\ref{eq:alw_proof_intermediate_1} in function of $m^*$:
    \begin{align}
         & \bigoplus_{i=1}^{ \lvert \mathcal{S} \rvert}  \bigoplus_{\varphi \in ENUM(\phi)}
        \left(  \ive{m_{\varphi}^{(i)}{=} m^* } , 0  \right) \otimes
        \left( r_\varphi^{(i)} , 0  \right) \otimes
        \left( t_\varphi^{(i)} , m^*  \right)                                                                              \\
         & =
        \bigoplus_{i=1}^{ \lvert \mathcal{S} \rvert}  \bigoplus_{\varphi \in ENUM(\phi)}
        \left( \ive{m_{\varphi}^{(i)}{=} m^* } r_\varphi^{(i)}  t_\varphi^{(i)} , 0  \right) \otimes \left( 1, m^* \right) \\
         & =
        \left( \sum_{i=1}^{ \lvert \mathcal{S} \rvert}  \sum_{\varphi \in ENUM(\phi)}  \ive{m_{\varphi}^{(i)}{=} m^* } r_\varphi^{(i)}  t_\varphi^{(i)} , 0   \right)  \otimes \left( 1, m^* \right) \label{eq:ALW_num_simplified}
    \end{align}
    Similarly we get for the denominator in Equation~\ref{eq:ALW}:
    \begin{align}
         & \bigoplus_{i=1}^{ \lvert \mathcal{S} \rvert}  \bigoplus_{\varphi \in ENUM(\phi)} \bigotimes_{\ell \in  DI(\varphi)}  \alpha_{IALW}^{(i)} \left(\ell \right) \\
         & =
        \left( \sum_{i=1}^{ \lvert \mathcal{S} \rvert}  \sum_{\varphi\in ENUM(\phi)}  \ive{m_{\varphi}^{(i)}{=} m^* }  t_\varphi^{(i)} , 0   \right)  \otimes \left( 1, m^* \right)  \label{eq:ALW_den_simplified}
    \end{align}
    We can now plug Equation~\ref{eq:ALW_exp_simplified}, Equation~\ref{eq:ALW_num_simplified} and Equation~\ref{eq:ALW_den_simplified} back into Equation~\ref{eq:ALW} and obtain:

    \begin{align}
         & \phantom{{}\Leftrightarrow{}}  \left( \E \left[  f(\phi) | \mathcal{S} \right],  0 \right)
        =
        \left(
        \frac
        { \sum_{i=1}^{ \lvert \mathcal{S} \rvert} \sum_{\varphi \in ENUM(\phi)}  \ive{m_{\varphi}^{(i)}{=} m^* } r_\varphi^{(i)}  t_\varphi^{(i)} }
        { \sum_{i=1}^{ \lvert \mathcal{S} \rvert}  \sum_{\varphi\in ENUM(\phi)}  \ive{m_{\varphi}^{(i)}{=} m^* }  t_\varphi^{(i)}   }
        , 0
        \right) \quad ( \lvert \mathcal{S} \rvert\rightarrow \infty)                                  \\
         & \Leftrightarrow \E \left[  f(\phi) | \mathcal{S} \right]
        =
        \frac
        { \sum_{i=1}^{ \lvert \mathcal{S} \rvert}  \sum_{\varphi \in ENUM(\phi)}  \ive{m_{\varphi}^{(i)}{=} m^* } r_\varphi^{(i)}  t_\varphi^{(i)} }
        { \sum_{i=1}^{ \lvert \mathcal{S} \rvert}  \sum_{\varphi\in ENUM(\phi)}  \ive{m_{\varphi}^{(i)}{=} m^* }  t_\varphi^{(i)}   } \label{eq:llw}
    \end{align}
    We realize that $r_\varphi^{(i)}$ is actually $f(\phi)$ evaluated at the $i$-th sample at the instantiation $\varphi$ and evoke~\citep[Theorem 4.1]{wu2018discrete} to prove Equation~\ref{eq:llw}, which also finishes this proof.
\end{proof}

\subsection{Proof of Proposition~\ref{prop:alwapproximation}}
\label{app:proof:alwapproximation}

\alwapproximation*

\begin{proof}
    We start the proof by invoking Theorem~\ref{thm:inference-by-expectation}, which expresses the conditional probability as a ratio of expectations. In the numerator and the denominator we then write the label of the propositional logic formulas as the sum of the labels of the respective possible worlds.
    \begin{align}
        P_\dcpprogram(\mu=q|\evidenceset=e)  \label{eq:cond_amc_1}
         & =\frac{\E_{\randomvariableset \sim  \dcpprogram_g} [\alpha( \phi \wedge \phi_q)] }{\E_{\randomvariableset \sim  \dcpprogram_g} [\alpha( \phi )] } \\
         & =
        \frac
        {
            \E_{\randomvariableset \sim  \dcpprogram_g} \left[ \sum_{\varphi\in ENUM(\phi \land \phi_{q e}) } \alpha(\varphi) \right]
        }
        {
            \E_{\randomvariableset\sim  \dcpprogram_g} \left[ \sum_{\varphi\in ENUM(\phi) } \alpha(\varphi)  \right]
        }  \label{eq:cond_amc_2}
    \end{align}
    Next, we approximate the expectation using a set of ancestral samples $\mathcal{S}$, followed by pulling out the query from the summation index in the numerator:
    \begin{align}
         & \frac
        {
            \E_{\randomvariableset \sim  \dcpprogram_g} \left[ \sum_{\varphi\in ENUM(\phi \land \phi_{q}) } \alpha(\varphi) \right]
        }
        {
            \E_{\randomvariableset \sim  \dcpprogram_g} \left[ \sum_{\varphi\in ENUM(\phi) } \alpha(\varphi)  \right]
        }
        \\
         & \approx
        \frac
        {
            \E \left[ \sum_{\varphi\in ENUM(\phi \land \phi_{q})   } \alpha(\varphi) |  \mathcal{S}  \right]
        }
        {
            \E \left[ \sum_{\varphi\in ENUM(\phi) } \alpha(\varphi) |  \mathcal{S}   \right]
        }  \label{eq:cond_amc_approx}
        \\
         & \approx
        \frac{
            \E \left[ \sum_{\varphi\in ENUM(\phi)   } \ive{\varphi \models \phi_q} \alpha(\varphi) |  \mathcal{S}  \right]
        }
        {
            \E \left[ \sum_{\varphi\in ENUM(\phi) } \alpha(\varphi) |  \mathcal{S}   \right]
        }
        \label{eq:cond_amc_expanded}
    \end{align}
    We now rewrite the fraction of two real numbers in Equation~\ref{eq:cond_amc_expanded} as the fraction of two infinitesimal numbers and plug in the definition of the infinitesimal algebraic likelihood weight (cf. Definition~\ref{def:alw}):
    \begin{align}
         & \frac{
            \E \left[ \sum_{\varphi\in ENUM(\phi)   } \ive{\varphi \models \phi_q} \alpha(\varphi) |  \mathcal{S}  \right]
        }
        {
            \E \left[ \sum_{\varphi\in ENUM(\phi) } \alpha(\varphi) |  \mathcal{S}   \right]
        }
        \\
         & =
        \frac{
            \left(\E \left[ \sum_{\varphi\in ENUM(\phi)   } \ive{\varphi \models \phi_q} \alpha(\varphi) |  \mathcal{S}  \right], 0 \right)
        }
        {
            \left( \E \left[ \sum_{\varphi\in ENUM(\phi) } \alpha(\varphi) |  \mathcal{S}   \right],0 \right)
        }          \\
         & \approx
        \frac
        {
        \displaystyle \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}  \bigoplus_{\varphi\in ENUM(\phi)} \ive{\varphi \models \phi_q} \bigotimes_{\ell \in \varphi}   \alpha_{IALW}^{(i)} \left(\ell \right)}
        {
        \displaystyle  \cancel{ \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}  \bigoplus_{\varphi\in ENUM(\phi)} \bigotimes_{\ell \in  DI(\varphi)}  \alpha_{IALW}^{(i)} \left(\ell \right) }}
        \otimes
        \frac
        {
        \displaystyle \cancel{\bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}  \bigoplus_{\varphi\in ENUM(\phi)} \bigotimes_{\ell \in  DI(\varphi)}  \alpha_{IALW}^{(i)} \left(\ell \right) } }
        {
        \displaystyle \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}  \bigoplus_{\varphi\in ENUM(\phi)} \bigotimes_{\ell \in \varphi}  \alpha_{IALW}^{(i)} \left(\ell \right)
        } \label{eq:cond_amc_canel}
    \end{align}

    In the last line the first factor corresponds to the numerator of the previous equation and the second factor corresponds to the reciprocal of the denominator.
    Note that the consistency of the infinitesimal algebraic likelihood weight of the numerator (first factor) is guaranteed by defining a new labeling function $\alpha^q(\varphi)\coloneqq \ive{\varphi \models \phi_q} \alpha(\varphi)$ and evoking Proposition~\ref{prop:alw_consistency} with $\alpha^q$.

    Finally, we push the expression $\ive{\varphi \models \phi_q}$ in the numerator back into the index of the summation ($\oplus$), which yields the following expression:
    \begin{align}
        \frac
        { \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}  \bigoplus_{\varphi\in ENUM(\phi \land \phi_{q})} \bigotimes_{\ell \in \varphi}  \alpha_{ALWI}^{(i)} \left(\ell \right)}
        { \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}  \bigoplus_{\varphi\in ENUM(\phi)} \bigotimes_{\ell \in \varphi}  \alpha_{IALW}^{(i)} \left(\ell \right)}
        \label{eq:cond_alw_last}
    \end{align}
    which proves the proposition. \end{proof}

\subsection{Proof of Proposition~\ref{prop:alwonddnnf}}
\label{app:proof:alwonddnnf}

\alwonddnnf*
\begin{proof}
    Algorithm~\ref{alg:prob_via_alw_kc} first compiles both propositional formulas into equivalent d-DNNF representations, cf. Lines~\ref{alg:prob_via_alw_kc:kc_qe} and~\ref{alg:prob_via_alw_kc:kc_e}.
    In Lines~\ref{alg:prob_via_alw_kc:alw_qe} and~\ref{alg:prob_via_alw_kc:alw_e} it then computes the (unnormalized) infinitesimal algebraic likelihood weight for both formulas by calling Algorithm~\ref{alg:unormalize_alw}. In other words, we compute the numerator and denominator in Equation~\ref{eq:cond_alw_last}. We observe that Algorithm~\ref{alg:unormalize_alw} evaluates a given d-DNNF formula for each conditioned topological sample using the \texttt{Eval} function, which evaluates a d-DDNF formula given a labeling function, cf. Algorithm~\ref{alg:eval}~\citep{kimmig2017algebraic}. The correctness of Algorithm~\ref{alg:prob_via_alw_kc} now hinges on the correctness of the \texttt{Eval} function, which was proven by~\citet{kimmig2017algebraic} for the evaluation of a d-DNNF formula using a semiring and labeling function pair that adheres to the properties described in Lemmas~\ref{lem:non_idem} to~\ref{lem:non_cons}. Effectively, Algorithm~\ref{alg:unormalize_alw} correctly computes the algebraic model count for each ancestral sample, adds up the results, and returns the unnormalized algebraic model count to Algorithm~\ref{alg:prob_via_alw_kc}. Line~\ref{alg:prob_via_alw_kc:conditional} finally return the ratio of the two unnormalized algebraic likelihood weights, which corresponds to the conditional probability $P_\dcpprogram(\mu=q|\evidenceset=e)$, as proven in Equations~\ref{eq:cond_amc_1} to~\ref{eq:cond_alw_last}.
\end{proof}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_excluded/implementation.tex
aij dcproblog paper

\section{Implementing \dcproblogsty}
\label{sec:implementation}

Having developed a probabilistic inference approach for \dcproblogsty based on knowledge compilation and similar to the one of \problogsty has the advantage of allowing us to leverage existing implementations of \problogsty (and its inference algorithms) in order to implement \dcproblogsty.
This entails that our implementation of \dcproblogsty, which we dub \dcproblogsys, inherits its general architecture from \problogsys and performs inference via a number of deterministic program transformations~\citep{dries2015problog2}. These program transformations reflect the inference steps described in Section~\ref{sec:dc2smt} and Section~\ref{sec:alw}.
We graphically summarize the program transformations in Figure~\ref{figure:dcproblog_sys} and describe further details in Appendix~\ref{sec:system_blocks}.
In Appendix~\ref{sec:system_query_predicates} we introduce built-in system predicates that allow a user to query and condition on atoms of interest, whose use we then briefly exhibit in Appendix~\ref{sec:usage}, where we present the usage of our implementation.

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{../figures/dcproblog_sys.pdf}
	\caption[Overview of the program transformation steps in the \dcproblogsys system]{Overview of the program transformation steps in the \dcproblogsys system. The nodes represent the intermediate representations and the annotations of the arrows show which transformations is performed on them. Each transformation is implemented in a seperate system block. (Figure inspired by~\protect\citep[Figure 2]{zuidbergdosmartires2019transforming} and \citep[Figure 1]{dries2015problog2}.}
	\label{figure:dcproblog_sys}
\end{figure}

\subsection{System Blocks}
\label{sec:system_blocks}

The system blocks that constitute \dcproblogsys are alterations of the existing system blocks already present in the \problogsys implementation, which is mainly written in Python\footnote{\url{https://github.com/ML-KULeuven/problog}}. Consequently, \dcproblogsys is written in Python as well and effectively extends \problogsys with continuous random variables but also discrete random variables with infinite support (cf. Poisson distribution). We would like to point out that in the absence distributional facts, \dcproblogsty programs in \dcproblogsys are equivalent to \problogsty programs in \problogsys.

\paragraph{\bf Grounding}
Inference in \dcproblogsys starts with a non-ground probabilistic program, a query set and an evidence set. In a first instance, the program is grounded with respect to the query set, which results in the (relevant) ground program. Grounding is performed using the resolution rules described in Appendix~\ref{app:resolution}.
\dcproblogsys reuses the grounder of \problogsys by adding additional resolution rules to the ones that are already implemented. More specifically, \dcproblogsys additionally has resolution rules for the cases {\bf C3} to {\bf C6} (cf. Definition~\ref{def:resolution_rules}), while the resolution rules for the cases {\bf C1} and {\bf C2} are already present in \problogsys's grounder. Note that \problogsys has special resolution rules for probabilistic facts and annotated disjunctions, which means that the \dcproblogsys implementation can also handle these. In the absence of distributional facts grounding in \dcproblogsys is identical to grounding in \problogsys as none of the rules for {\bf C3} to {\bf C6} are used.

\paragraph{\bf Breaking Cycles}
Once we have the relevant ground program we would like to obtain a cycle-free representation of the program, as such a cycle-free program is trivially mapped onto a Boolean formula (cf. Section~\ref{sec:relprog2boolfrom}). Concerning the cycle breaking step, \dcproblogsys reuses the cycle breaking algorithm deployed in \problogsys with minor bookkeeping extensions in order to ensure consistency between the names of random variables (stemming from distributional facts) before and after performing cycle breaking.

\paragraph{\bf Knowledge Compiling}
Next, \dcproblogsys takes, a labeled propositional formula, \ie the acyclic representation of the program, and knowledge compiles it into a so-called sentential decision diagram (SDD)~\citep{darwiche2011sdd}.
Identically to \problogsys, the compilation of SDDs, which are a strict subset of d-DNNF formulas, is performed using bottom-up compilation\footnote{\url{http://reasoning.cs.ucla.edu/sdd/}}~\citep{choi2013dynamic}.

Currently, the compilation of labeled propositional formulas is agnostic towards information present in the labels of the atoms. This can lead to situations where infeasible labeled formula are not discarded. Take for instance the formula $a \land b$, where the labels are $\alpha(a)=\ive{x<0}$ and $\alpha(b)=\ive{x>1}$. Clearly there is no instantiation of $x$ that would satisfy the conditions in the labels. However, as the compiler is label-agnostic, it does not take this information into account and does not disregard such a formula by setting it equivalent to $\bot$. This might results in needless computations. To prevent this from happening one could take into account the information in the labels and effectively perform knowledge compilation with SMT formulas, which has been studied in the WMI literature~\citep{sanner2011symbolic,kolb2018efficient,kolb2019exploit,derkinderen2020ordering}.
We leave it for future research to adapt and further develop these techniques in the context of probabilistic programming.\footnote{We establish the link between computing expected labels of propositional logic formulas and weighted model integration in  Appendix~\ref{app:labeled_prop_to_SMT}.}

\paragraph{\bf Evaluating} Finally, we obtain the probability of a query to a (hybrid) probabilistic logic program by evaluating the compiled logic formula using algebraic likelihood weighting.
Note that the arithmetic circuit that corresponds to the compiled formula does not contain any program branches, in other words there are no if-else statements presents. Effectively, knowledge compilation transforms a query to a probabilistic query into a branchless program that can trivially be run on a graphics processing unit. Furthermore, we observe that each iteration in the loop of Algorithm~\ref{alg:unormalize_alw} uses the same (branchless) circuit. This means that the loop in Algorithm~\ref{alg:unormalize_alw} is  embarrassingly parallelizable, as already pointed out by~\citet{zuidbergdosmartires2019exact}.
\dcproblogsys implements the circuit evaluations for the different samples by using a vectorized version of Algorithm~\ref{alg:unormalize_alw} implemented in PyTorch~\citep{paszke2019pytorch} and Pyro~\citep{bingham2019pyro}.

\subsection{System Predicates to Query \dcproblogsty Programs}
\label{sec:system_query_predicates}

Implementations of probabilistic programming languages usually define a set of system predicates that allow users to declare queries and conditioning sets within a probabilistic program.
% A key issue of these system predicates is that they should encode unambiguous inference tasks, \eg encode conditional probabilities that exist and do not succumb to the Borel-Kolmogorov paradox. \ak{didn't you say in an earlier comment that syntax cannot ensure this? I agree about the unambiguous part, but not the rest.}
We are going to introduce such system predicates for \dcproblogsty based on \problogsty's system predicates.

% Many probabilistic programming languages are often quite permissive with the use of system predicates for querying programs. This can then result in possibly ill-defined queries, such as in Distributional Clauses~\citep[Section 3.2]{nitti2016probabilistic}.
% Other languages, such as BLOG, leave it up to the user and simply assume the well-definedness of conditional probability queries~\citep{wu2018discrete}.
% A more sophisticated method to tackle the non-uniqueness of the conditional probability is {\em disintegration}~\citep{shan2017exact}, where, instead of returning a conditional probability, a family of conditional probabilities is returned (representing an uncountable number of limits).
% However, this does not resolve the issue of the observation/measuring process being inherently underspecified.

\begin{definition}[Query Predicates]
	\label{def:query_predicates}
	Let $\herbrandbase$ be the Herbrand base of a given \dcproblogsty program \dcpprogram.
	Let us assume that we are given the evidence $\evidenceset=e$ (with $\evidenceset \subset \herbrandbase$).
	and a set $\queryset \subset \herbrandbase$ of atoms of interest, called
	query atoms.
	We encode these sets directly as facts in \dcpprogram using the following system predicates:
	\begin{itemize}
		\item The unary predicate \probloginline{query/1} declares a set of Boolean atoms of interest $\queryset = \{ \text{\probloginline{q}} \mid \text{\probloginline{query(q)}} \in \dcpprogram     \}$.
		% 
		\item The binary predicate \probloginline{evidence/2}, where the second argument is either the \probloginline{true} symbol or the \probloginline{false} symbol, encodes a set of Boolean atoms that we condition on. It produces an ordered set of evidence atoms:
		$$
		\evidenceset =
		\{ E_\text{\probloginline{e@$_i$@}}  \mid \text{\probloginline{evidence(e@$_i$@,false)} } \in \dcpprogram \}
		\cup
		\{   E_\text{\probloginline{e@$_i$@}} = \mid \text{\probloginline{evidence(e@$_i$@,true)}} \in \dcpprogram \} $$
		and the corresponding ordered multiset of truth values for the elements of $\evidenceset$:
		$$
		e =
		\{  0   \mid \text{\probloginline{evidence(e@$_i$@,false)} } \in \dcpprogram \}
		\cup
		\{ 1 \mid \text{\probloginline{evidence(e@$_i$@,true)}} \in \dcpprogram \}
		$$
		The probability $P_\dcpprogram(\evidenceset=e)$ now corresponds to the denominator in Equation~\ref{eq:conditional_prob}.
		The unary version \probloginline{evidence/1} is shorthand for the \probloginline{true} case and omits the second argument.

		\item The binary predicate \probloginline{observation/2} is used to condition random variables on observed values they take. The first argument denotes a random term and the second one the observed value. We define the semantics of the \probloginline{observation/2} predicate in terms of  \probloginline{evidence/1} and a novel Boolean comparison predicate \probloginline{delta_interval/2}. The atom \probloginline{observation(x,v)} is syntactic sugar for two clauses of the form:
		\begin{problog}
obs:- delta_interval(x,v).
evidence(obs).
		\end{problog}
		where \probloginline{obs} is a fresh ground atom. \probloginline{delta_interval/2} is the built-in comparison predicate describing infinitesimally small intervals and which we introduce in Definition~\ref{def:delta_interval}.
	\end{itemize}
\end{definition}

The \probloginline{query} and \probloginline{evidence} predicates are inherited from \problogsty, contrary to \probloginline{observation}, which we introduce in order to handle point observations of both discrete and continuous random variables via the \probloginline{delta_interval/2} predicate.
%Note that the definition of the \probloginline{observation} predicate can equally be used with discrete and continuous random variables. This will allow us to condition queries to \dcproblogsty programs on mixed discrete-continuous random variables, cf. Example~\ref{ex:indian_gpa}.

\subsection{System Usage}
\label{sec:usage}

We provide two modes of use for \dcproblogsys. Firstly, through the command line  and secondly, as \dcproblogsys is implemented in Python, through Python scripts. To illustrate the command line interface, consider a file named \mintinline{bash}{coin_flip.pl} containing the program described in Example~\ref{example:bayesian_learning}.

\begin{example}\label{example:bayesian_learning}
	We model a coin flip scenario where the probability of the coin turning up heads is defined through two distributional clauses. We then flip the coin three times and get as evidence that the first and third flip turn up heads, and the second coin turns up tails. We then query for the probability of \probloginline{b>0.5.}
	\begin{problog*}{linenos}
0.2::a.
b~beta(1,1):- a.
b~beta(1,2):- not a.
b::coin_flip(N).

evidence(coin_flip(1), true).
evidence(coin_flip(2), false).
evidence(coin_flip(3), true).

q: b>0.5.
query(q).
	\end{problog*}
\end{example}
Having \dcproblogsys installed on a machine, which we provide as an extension of \problogsys, we can execute the query with the following command line instruction:
\begin{minted}{bash}
problog dc simple_coin.pl -n_samples=20000
\end{minted}
The instruction simply expresses our desire to run the \mintinline{bash}{simple_coin.pl} file using \problogsys with the \dcproblogsys extension and to furthermore use $20000$ samples (per random variable) to answer the query.

Alternatively, \dcproblogsys can also be used within Python scripts directly, which enhances its flexible as Python code can be mixed with ProbLog programs. For instance, take again the program from Example~\ref{example:bayesian_learning}. We can now extend \dcproblogsys with a new builtin \probloginline{samples} predicate that returns sampled values of random variables instead of querying for probabilities, cf. Example~\ref{example:bayesian_learning_python}.

\begin{example}[Using ProbLog from Python]
\label{example:bayesian_learning_python}
We write the program in Example~\ref{example:bayesian_learning} within a Python script and also use the \probloginline{samples} predicates to obtain samples.

\begin{minted}[linenos]{python}
#import necessary classes from core ProbLog
from problog.program import PrologString

# import classes from DC-ProbLog
from problog.tasks.dcproblog.solver import InferenceSolver
from problog.tasks.dcproblog.parser import DCParser

# set up the configuration for inference
configuration = {
    "n_samples":20000, # set the number of samples to be used
    "device":"gpu", # use the GPU (alterntively CPU)
}

# put the ProbLog program in a string
str_program = """
    0.2::a.
    b~beta(1,1):- a.
    b~beta(1,2):- \+a.
    B::coin_flip(N):- B is b.

    evidence(coin_flip(1), true).
    evidence(coin_flip(2), false).
    evidence(coin_flip(3), true).

    samples(b).  % we are interested in samples
"""

# initiliaze the solver for DC-ProbLog inference
solver = InferenceSolver(**configuration)

# obtain a ProbLog program from a string 
# using the DCParser instead of the ProbLog parser
program = PrologString(str_program, parser=DCParser())

# query the program
probabilities = solver.probability(program, **configuration)
\end{minted}

The \mintinline{python}{probabilities} variable now contains the samples for the two beta distributions in the \dcproblogsty program. Retrieving the samples allows us to plot them as a histogram. We do this in Figure~\ref{fig:coins}.
\end{example}

\begin{figure}[h]
	\begin{center}
	\includegraphics[width=0.49\linewidth]{../figures/prior.pdf}
	\includegraphics[width=0.49\linewidth]{../figures/posterior.pdf}
	\end{center}
	\caption[Bayesian learning of coins]{On the left we have the histogram for the unconditioned samples (prior) from the program in Example~\ref{example:bayesian_learning} and the histogram on the right demonstrates the effect of conditioning on the outcome of the coin flips (posterior). The green part of the histogram corresponds to the first distributional clause (\probloginline{b~beta(1,1)}) and the red part to the second (\probloginline{b~beta(1,2)}).}
	\label{fig:coins}
\end{figure}

% Figure~\ref{fig:coins} shows the prior and posterior after learning with the dataset {\probloginline{coin_flip(1)=true}, \probloginline{coin_flip(2)=false}, \probloginline{coin_flip(3)=true}}.

% \subsection{old section intro}
% We make two observations:
% \begin{enumerate}
%     \item Every implementation of \prologsty  (or any other logic programming language)  reserves a set of {\em  system predicates} executing code not written in pure \prologsty~\citep[Chapter 8]{sterling1994art}.
%     System predicates complement pure \prologsty implementations and give users access to efficient algorithms that are cumbersome to implement in pure logic programming -- for instance algorithms for numeric computations.
%     \item In Section~\ref{sec:semantics} we have seen that we can separate a probabilistic logic program into a set of independent random choices and a logic program. The independent choices and the logic program communicate to each other via comparison predicates. Comparison predicates are an instance of system predicates.
% \end{enumerate}

% These observations hint at the possibility of implementing \dcproblogsty as a deterministic logic programming language with access to an external engine with support for random variables.
% On an implementation level, the key difference between a probabilistic logic programming language and a (deterministic) logic programming language is the availability of probability distributions in the external engine of the former.

% Decoupling independent choices (random variables) and the deterministic system (logic program) does also mean that we should be able to take a probabilistic logic programming language that only supports finite discrete probability distributions and extend it with continuous probability distributions.
% We will demonstrate how to achieve this by adding continuous random variables to the ProbLog2 system~\citep{dries2015problog2} (the most recent implementation of \problogsty). We dub our implementation \dcproblogsys.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_excluded/procedural_semantics.tex
aij dcproblog paper

\section{Procedural Semantics of \dcproblogsty}\label{sec:swp}
We now provide a procedural view on the semantics that extends \citet{van1991well}'s $W_\dcpprogram$-operator for well-founded models to deal with comparison literals following the approach of the stochastic $T_\dcpprogram$-operator for negation-free distributional clauses \citep{gutmann2011magic}. As in the latter, we exclude comparison atoms from interpretations, and instead introduce a function \textsc{ReadTable}$(\cdot)$ that keeps track of the sampled values of random variables to evaluate comparison literals. Given a comparison literal \probloginline{q}, \textsc{ReadTable(\probloginline{q})}  returns the truth value of \probloginline{q} based on the values of the random terms \probloginline{x} in \probloginline{q}, which are either retrieved from the table or sampled according to their definition \probloginline{x~D} as included in the interpretation and stored in the table in case they are not yet available.

\subsection{Unfounded Sets}

We first extend the notion of \emph{unfounded set}, which is used to draw negative conclusions, to the stochastic case.
\begin{definition}[Stochastic unfounded set]
	Let $\dcpprogram$ be a DC-ProbLog program with Herbrand base $H$. Let $I$ be a partial interpretation of  $H\setminus F$, \ie, not including comparison atoms. 
	We say that  $A\subseteq H\setminus F$ is a stochastic unfounded set (of $\dcpprogram$) with respect to $I$ if each atom $a\in A$ satisfies the following condition: For each instantiated rule $R$ of $\dcpprogram$ whose head is $a$, (at least) one of the following holds:
	\begin{enumerate}
		\item Some (positive or negative) non-comparison subgoal $q$ of the body is false in $I$.
		\item For some comparison literal $q$ in the body, the following both hold
		\begin{itemize}
			\item $I$ contains  a true atom \probloginline{x~D} for every random term \probloginline{x} in \probloginline{q}.
			\item $\textsc{ReadTable(\probloginline{q})}=true$
		\end{itemize}
		\item Some positive subgoal of the body occurs in $A$.
	\end{enumerate}
	The greatest stochastic unfounded set (of $\dcpprogram$) with respect to $I$, denoted $SU_\dcpprogram(I)$, is the union of all sets that are unfounded with respect to $I$.
\end{definition}
We note that comparison literals can invalidate external support for an unfounded set only if all involved random variables have associated distributions in the partial interpretation. \todo{example}

\begin{definition}[Stochastic $W_\dcpprogram$-operator]\label{def:swp}
	We define the transformations $ST_\dcpprogram$, $SU_\dcpprogram$, and $SW_\dcpprogram$ as follows:
	\begin{itemize}
		\item $p\in T_\dcpprogram(I)$ if and only if there is some instantiated rule $R$ of $\dcpprogram$ such that $R$ has head $p$, and
		\begin{itemize}
			\item each non-comparison  literal in the body of $R$ is true in $I$,
			\item $I$ contains a true atom \probloginline{x~D} for every random term \probloginline{x} that appears inside a comparison literal in the body, and 
			\item for each comparison literal \probloginline{q} in the body, $\textsc{ReadTable(\probloginline{q})}=true$
		\end{itemize}
		\item $SU_\dcpprogram(I)$ is the greatest unfounded set of $\dcpprogram$ with respect to $I$.
		\item $SW_\dcpprogram(I) = ST_\dcpprogram(I) \cup \neg SU_\dcpprogram(I)$
	\end{itemize}
	where negation applied to a set negates each element of the set.
\end{definition}
That is, comparison literals in rule bodies are evaluated based on the table, but can only decide on whether a rule is fired (in $ST_\dcpprogram$) or blocked (in $SU_\dcpprogram$) if all their random variables have a distributional atom in the (partial) interpretation.
However, for valid core \dcproblogsty programs, the first iteration of fixpoint computation adds all distributional facts to the interpretation. Thus, from the second iteration onwards, all rules with comparison atoms will have their necessary \probloginline{x~D} atoms available, and may access the table to evaluate comparisons. The procedural view thus exactly mirrors the three layers of the declarative view: it first introduces all random variables and their distributions, which then makes it possible to evaluate comparison atoms based on the sampled joint assignment maintained by the table, and thus enables derivation of logical consequences through the computation of the well-founded model.

\todo{too much context assumed -- expand / illustrate}

\subsection{Procedural view}
As stated above, the semantics of a full \dcproblogsty program is given by the corresponding core program, which explicitly encodes in the program how each random term should be interpreted based on logical context. 
In Section~\ref{sec:swp}, we provided a procedural view on the semantics through the stochastic $W_\dcpprogram$ operator, which, as the procedural semantics of the distributional clause language, incorporates these context checks into the execution mechanism rather than the program itself. While the explicit encoding makes the program and its semantics independent of the execution mechanism, this generality comes at the price of a potentially significant blowup in program size, as illustrated in the following example.

\begin{example}
	Consider the following program fragment modeling a robot moving along a one-dimensional corridor under different conditions.
	\begin{problog*}{linenos}
0.2::cond(1); 0.5::cond(2); 0.3::cond(3).
pos(0) ~ normal(20,3).
pos(1) ~ normal(pos(0)+0.5, 0.1) :- cond(1). @\label{line:cond1pos1}@
pos(1) ~ normal(pos(0), 0.5) :- cond(2).
pos(1) ~ normal(pos(0), 2) :- cond(3).
pos(2) ~ normal(pos(1)+0.5, 0.1) :- cond(1). @\label{line:cond1pos2}@
pos(2) ~ normal(pos(1), 0.5) :- cond(2).
pos(2) ~ normal(pos(1), 2) :- cond(3).
...
	\end{problog*}
	Given any interpretation of the AD, only one of the three conditions applies, and determines the distribution to associate with every position. For instance, if \probloginline{cond(1)} holds, \probloginline{pos(1)} is determined by line~\ref{line:cond1pos1} only, and \probloginline{pos(2)} by line~\ref{line:cond1pos2}.  The transformed program, however, has to summarize \emph{all} situations, and thus needs a correspondingly larger set of random variables. We omit the transform of the annotated disjunction for brevity:
	\begin{problog*}{linenos}
v1 ~ normal(20,3).
refers2rv(pos(0),v1).

v2 ~ normal(v1+0.5, 0.1).
refers2rv(pos(1),v2) :- refers2rv(pos(0),v1),cond(1).
v3 ~ normal(v1, 0.5).
refers2rv(pos(1),v3) :- refers2rv(pos(0),v1),cond(2).
v4 ~ normal(v1, 2).
refers2rv(pos(1),v4) :- refers2rv(pos(0),v1),cond(3).

v5 ~ normal(v2+0.5, 0.1).
refers2rv(pos(2),v5) :- refers2rv(pos(1),v2), cond(1).
v6 ~ normal(v3+0.5, 0.1).
refers2rv(pos(2),v6) :- refers2rv(pos(1),v3), cond(1).
v7 ~ normal(v4+0.5, 0.1).
refers2rv(pos(2),v7) :- refers2rv(pos(1),v4), cond(1).
v8 ~ normal(v2, 0.5).
refers2rv(pos(2),v8) :- refers2rv(pos(1),v2), cond(2).
v9 ~ normal(v3, 0.5).
refers2rv(pos(2),v9) :- refers2rv(pos(1),v3), cond(2).
v10 ~ normal(v4, 0.5).
refers2rv(pos(2),v10) :- refers2rv(pos(1),v4), cond(2).
v11 ~ normal(v2, 2).
refers2rv(pos(2),v11) :- refers2rv(pos(1),v2), cond(3).
v12 ~ normal(v3, 2).
refers2rv(pos(2),v12) :- refers2rv(pos(1),v3), cond(3).
v13 ~ normal(v4, 2).
refers2rv(pos(2),v13) :- refers2rv(pos(1),v4), cond(3).
...
	\end{problog*}
	As the transform is purely syntactical, in this specific example, where conditions do not change from one timestep to the next, many of the random variables it introduces are irrelevant, as they correspond to impossible combinations of contexts. For instance, \probloginline{v6} is relevant only if \probloginline{v3} is relevant and \probloginline{cond(1)} holds (line 14), but \probloginline{v3} is relevant only if \probloginline{cond(2)} holds (line 7), which is inconsistent with \probloginline{cond(1)}. 
\end{example}

It is worth noting that, for a valid AD-free program $\dcpprogram$, the fixpoint of  the $SW_\dcpprogram$-operator (Definition~\ref{def:swp}) applied on $\dcpprogram$ itself realizes the same distribution over the non-comparison atoms' truth values as that of applying the operator to the transform of $\dcpprogram$, despite the fact that the transform uses different names for random variables. The $SW_\dcpprogram$-operator controls the application of rules with comparison literals in the body through the presence of suitable distributional atoms in the partial interpretation, which for distributional clauses can only appear in the interpretation if the body of the clause is satisfied. In the transformed program, the same control mechanism is explicitly implemented in the program itself through \dcplpinline{refers2rv}-atoms, whose defining rules mirror the original DC bodies.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_excluded/smart_grounding.tex
aij dcproblog paper

\section{Grounding via Resolution}
\label{app:resolution}

In Section~\ref{sec:ALWviaKC} we describe the conversion of a queried \dcproblogsty program to a labeled propositional formula through a sequence of deterministic steps. The first of which is obtaining the relevant ground program with respect to a set of query atoms and evidence atoms.

The goal of this first step of the conversion algorithm is to identify a ground core \dcproblogsty program $\dcpprogram_g$ that defines the same joint distribution over the query and evidence atoms as the original program, and contains only logical rules and distributional facts that are relevant for the query and evidence atoms.

To achieve this goal, we define a procedure that constructs a ground core program for each potential logical derivation of an atom. We use this procedure to construct the full ground program by iterating over all such derivations for the query and evidence atoms. The procedure itself consists of two steps. First, it constructs a derivation using a set of dedicated resolution rules, where each application of a resolution rule produces a tag consisting of a set of logical rules and a set of rule manipulation instructions. If the first step succeeds, the sequence of tags is then translated to a ground program in the second step by applying the rule manipulation instructions of the tags.

The overall approach 
%as well as the two rules for  atoms defined by logical rules are 
is borrowed from the corresponding procedure introduced by \citet{fierens2015inference} for \problogsty. To generalize this approach to \dcproblogsty, we introduce rules to deal with comparison atoms, distributional clauses, and the context-dependent random variables they introduce, in addition to the original rules intriduced by~\citet{fierens2015inference}.\footnote{\citet{nitti2016probabilistic} also introduced modified resolution rules for hybrid probabilistic logic programs but for an inference algorithm that interleaves grounding and sampling, which does not require keeping track of rule groundings.}

%To obtain the relevant ground program for a \problogsty program, \citet{fierens2015inference} introduce a backward chaining procedure that applies SLD resolution~\citep{kowalski1974predicate,sterling1994art}, but ignores negation symbols, as the goal is to \emph{identify}   ground rules that are (potentially) relevant for proving query or evidence atoms rather than to actually  \emph{prove} those atoms.

%We extend this procedure to \dcproblogsty, where we additionally need to 
%identify the \emph{relevant distributional facts}, \ie, the random variables that are interpreted in the relevant ground program and their distributions. As we have seen in Sections~\ref{sec:semantics} and~\ref{sec:dcproblog}, random variables can be interpreted both within comparison literals and within distributions of other random variables, and can be context-dependent, \ie, the same random variable term may be used to refer to several random variables introduced by a set of distributional clauses. We therefore  introduce additional  resolution rules to handle these cases.\footnote{\citet{nitti2016probabilistic} also introduced modified resolution rules for hybrid probabilistic logic programs but for an inference algorithm that interleaves grounding and sampling.}

%In order to obtain the relevant ground program we deploy, similarly to \problogsty, {\em selective linear definite clause} resolution~\citep{kowalski1974predicate,sterling1994art} with {\em negation as failure}~\citep{Clark1977NegationAF} (SLDNF). SLDNF resolution allows for computing the relevant ground program for a specific query to the logic program, while avoiding the grounding of unnecessary predicates~\citep{kersting2000bayesian,fierens2015inference}. In the case of range-restricted programs SLDNF resolution is guaranteed to terminate. We refer the reader to \citet{fierens2015inference} for a more detailed account of grounding procedure. However, in order to handle comparison predicates and distributional clauses, we will need to modify the SLDNF-resolution rules\footnote{\citet{nitti2016probabilistic} also introduced modified resolution rules for hybrid probabilistic logic programs but for an inference algorithm that interleaves grounding and sampling.}.

%Before we describe the the adapted SLDNF  resolution rules, which we dub {\em SLDNF-DC resolution}, we need specify two helper functions and a builtin predicate that are used during SLDNF-DC resolution:
In contrast to the declarative transformation used to define the semantics of full \dcproblogsty, cf.~Definition~\ref{def:adfree-to-core}, which introduces additional atoms and rules to describe the logical contexts in which specific instances of a random variable term are used, the grounding procedure directly incorporates the original context definitions into the bodies of ground rules, without introducing further atoms.  
Our resolution rules further rely on two global helper functions and an auxiliary predicate:
\begin{itemize}
	\item The injective helper function `$\mathrm{createRVname}$' maps a pair consisting of a ground random variable term and a logical context\footnote{We syntactically represent the context as a tuple of terms, which is interpreted as their logical conjunction (and thus independently of order) by the function.} to a unique constant. This constant serves as the name for the random variable term in the specific context, thus providing an external implementation of the \probloginline{refers2rv/2} predicate.
	\item The function `$\mathrm{symbols}$' maps a term~$t$  to the set of random variable terms in~$t$.  For instance, $\mathrm{symbols}(\text{\probloginline{a(N)+b(1)*2}})$ returns the set $\{ \text{\probloginline{a(N)}}, \text{\probloginline{b(1)}} \}$. This function is crucial to identify relevant random variable terms. 
	\item The auxiliary predicate \probloginline{rv/3}  takes as first argument a random variable term, as second argument a distribution term, and as third argument a context for the random variable term, which is given as a tuple of terms. \probloginline{rv/3} atoms are inserted into the current conjunctive goal when new random variables are identified, and trigger the creation of the corresponding random variable once its full ground context, which combines the body of the defining distributional clause with the contexts of the parent random variables, is determined.
\end{itemize}

We describe each resolution rule using three elements: the condition under which the rule applies (\cposlit to \crvone), the actual resolution step performed by the rule (\rposlit to \rrvonea and \rrvoneb), and the tag that labels this step (\tposlit to \trvone). We overload substitution notation to also allow for substitution of a random variable term by a newly introduced context dependent random variable name. The context-dependent rewrite operator $\twoheadleftarrow$ introduced in the last tag (\trvone) will be defined in Definition~\ref{def:context-substitution}.

\begin{definition}[Resolution Rules]
	\label{def:resolution_rules}

	%Let $T_{RV}$ be a hash table of $(key, value)$ pairs that keeps track of random variables (key) and their distributions (value), with insertion function $\mathrm{INSERT}(T_{RV},key,value)$. 
	Let $T_{RV}$ be a hash table that keeps track of random variables, with insertion function $\mathrm{INSERT}(T_{RV},var)$. 
	The six resolution rules for grounding \dcproblogsty programs with respect to a given goal are defined as follows.

	\begin{enumerate}
		\item[\cposlit] If the goal is $(l_1, l_2, \dots, l_n)$ and  $\exists \theta$ such that  $h \lpif b_1, \dots, b_m \in \dcpprogram$, $\theta  = mgu(l_1, h)$ then:
		\begin{align*}
			&{\text{\bf \rposlit:}} \quad (l_1, l_2, \dots, l_n) \vdash (b_1, \dots, b_m, l_2, \dots, l_n) \theta
			\\
			&{\text{\bf \tposlit:}} \quad \tau = (\{ h \lpif b_1, \dots, b_m \}, \theta)
		\end{align*}

		\item[\bf \cneglit]  If the goal is $(not(l_1), l_2, \dots, l_n)$ and  $\exists \theta$ such that  $h \lpif b_1, \dots, b_m \in \dcpprogram$, $\theta  {=} mgu(l_1, h)$ then:
		\begin{align*}
			&{\text{\bf \rneglit:}} \quad (not(l_1), l_2, \dots, l_n) \vdash (b_1, \dots, b_m, l_2, \dots, l_n) \theta
			\\
			&{\text{\bf \tneglit:}} \quad \tau = (\{ h \lpif b_1, \dots, b_m \}, \theta)
		\end{align*}

		\item[\ccompone]  If the goal is $( l_{comp}, l_1, \dots, l_n )$ where  $l_{comp}$ is a comparison literal, and  there exists a substitution $\theta$ such that for some $s \in symbols(l_{comp})\setminus T_{RV}$ it holds that $h\sim d\lpif b_1, \dots, b_m\in \dcpprogram$, with $\theta = mgu(s,h)$, then:
		\begin{align*}
			&{\text{\bf \rcompone:}} \quad ( l_{comp}, l_1, \dots, l_n ) \vdash (b_1, \dots, b_m, rv(h, d, (b_1, \dots, b_m)),  l_{comp}, l_1,\dots ,l_n)\theta
			\\
			&{\text{\bf \tcompone:}} \quad \tau = (\emptyset, \theta)
		\end{align*}

		\item[\bf \ccomptwo]  If the goal is $( l_{comp}, l_1, \dots, l_n )$  where $l_{comp}$ is a comparison literal, and   $symbols(l_{comp})\subseteq T_{RV}$ then:
		\begin{align*}
			&{\text{\bf \rcomptwo:}} \quad  ( l_{comp}, l_1, \dots, l_n ) \vdash  (l_1, \dots, l_n )
			\\
			&{\text{\bf \tcomptwo:}} \quad \tau = (\emptyset, \emptyset)
		\end{align*}

		\item[\bf \crvtwo]  If the goal is $(rv(x, d, (b_1, \dots, b_{m})), l_1,\dots, l_n)$ and  there exists a substitution $\theta$ such that for some $s \in symbols(d)\setminus T_{RV}$ it holds that $h\sim d' \lpif b'_1, \dots, b'_{k}\in \dcpprogram$, $\theta = mgu(s,h)$, then:
		\begin{align*}
			{\text{\bf \rrvtwo:}} \quad 
			&(rv(x, d, (b_1, \dots, b_{m})), l_1, \dots, l_n) \\
			&~~~\vdash (b'_1, \dots, b'_{k}, rv(h, d', (b'_1, \dots, b'_{k})), rv(x, d, (b_1, \dots, b_{m})), l_1, \dots, l_n)\theta
			\\
			{\text{\bf \trvtwo:}} \quad &\tau =  (\emptyset, \theta)
		\end{align*}

		\item[\bf \crvone]  If the goal is $(rv(x, d, (b_1, \dots, b_{m})), l_1,\dots, l_n)$ and   $symbols(d)\subseteq T_{RV}$ then we further  distinguish two cases.
		$\text{\bf R6a}$ applies if $l_1 = rv(y, d_y, (b'_1, \dots, b'_k))$ and $x\in symbols(d_y)$. $\text{\bf R6b}$ applies in all other cases.  The tag is identical for both.
		\begin{align*}
			{\text{\bf \rrvonea:}} \quad
			&x_{new} = \mathrm{createRVname}(x,(b_1,\dots, b_m)) \\
			& \mathrm{INSERT}(T_{RV}, x_{new}) \\
			&\phi = \{ x/ x_{new} \} \\
			&(rv(x, d, (b_1, \dots, b_{m})), rv(y, d_y, (b'_1, \dots, b'_k)) , l_2, \dots, l_n) \\
			&~~~    \vdash (rv(y, d_y, (b_1, \dots, b_{m}, b'_1, \dots, b'_k) , l_2, \dots, l_n)\phi
			\\
			{\text{\bf \rrvoneb:}} \quad
			&x_{new} = \mathrm{createRVname}(x,(b_1,\dots, b_m)) \\
			& \mathrm{INSERT}(T_{RV}, x_{new}) \\
			&\phi = \{ x/ x_{new} \} \\
			&(rv(x, d, (b_1, \dots, b_{m})), l_1, \dots, l_n) \vdash (l_1, \dots, l_n)\phi
			\\
			{\text{\bf \trvone:}} \quad &\tau =  (\{x\sim d \}, \{x\twoheadleftarrow (x_{new},(b_1,\dots, b_m))\})
		\end{align*}
	\end{enumerate}
	% AK: for R6a, note that: R5 always inserts the parent rv/3-atom to the left of the child rv/3-atom, so if we need to propagate context a generation down, it is always to the goal that directly follows. However, not all pairs of neighboring rv/3-atoms are parent and child (we could also get an unrelated rv/3 from the last subgoal of the parent's context directly before the parent's rv/3), and we thus also need to check the child's distribution mentions the parent. 
\end{definition}
Intuitively, the first two rules (\rposlit and \rneglit) apply SLD resolution just as for \problogsty, and record both the rule and substitution used in the tag, while the remaining four rules are responsible for collecting information related to random variables and their distributions. These latter rules use the `$\mathrm{symbols}$' function to extract the interpreted random variable terms from the current subgoal, and compare that set against the hash table storing known random variables to distinguish between cases where all interpretations are already available and cases where the interpretation of some symbol still needs to be determined. Rules \rcompone and \rcomptwo are concerned with comparison literals, which can simply be dropped from the goal once all their random variables are resolved (\rcomptwo), and otherwise are re-considered after resolving the body of a relevant distributional clause and inserting the corresponding random variable via the auxiliary predicate (\rcompone). The last two rules ensure that the parents of a relevant random variable are included recursively using the same principle (\rrvtwo), and that all random variables are recorded in the hash table once their ancestors and context have been fully evaluated (\rrvonea and \rrvoneb). Rules \rcompone--\rrvtwo only record their substitutions in the tag, whereas \rrvonea and \rrvoneb record a new distributional fact as well as contextual rewrite information for its random variable term.  Additionally, \rrvonea propagates the parent random variable's context into the context of the child.

%Given a query or literal to be proven, the resolution rules in Definition~\ref{def:resolution_rules} allow us to construct a resolution tree from a specific non-ground program with respect to the query. 
%The conversion algorithm uses these rules to construct a resolution tree for each query or evidence atom, using a shared hash map for the random variables.

\begin{example}[Derivations and tag sequences]
	\label{ex:resolution_tree}
	%Consider the AD-free program obtained in Example~\ref{ex:eliminating_pf}. 
	%We can construct an SLDNF-DC tree using the resolution rules from Definition~\ref{def:resolution_rules}. We depict the resolution tree in
	Figures~\ref{fig:sld_tree_left} and~\ref{fig:sld_tree_right} graphically depict the derivations obtained by applying the resolution rules from Definition~\ref{def:resolution_rules} to the AD-free program in Example~\ref{ex:eliminating_pf} as a resolution tree, where nodes represent goals and edges are labeled with the context number of the applied rule. We represent empty context conjunctions resulting from distributional facts using the term \probloginline{true} in \probloginline{rv/3}-atoms and labels.

	The tag sequence for the leftmost derivation is\\
	$(\{$\probloginline{works(N):-cool(N)}$\}, \emptyset)$, \\
	$(\{$\probloginline{cool(1):-rv_cool(1)=:=1}$\}, \{N/1\})$,\\
	$(\emptyset, \emptyset)$,\\
	$(\{$\probloginline{rv_cool(1)~flip(0.99)}$\}, \{$\probloginline{rv_cool(1)}$\twoheadleftarrow ($\probloginline{rv_cool_1_1}, \probloginline{true}$)\})$,\\
	$(\emptyset, \emptyset)$.

	The tag sequence for the derivation in the middle is\\
	$(\{$\probloginline{works(N):-temp(N)<25.0}$\}, \emptyset)$, \\
	$(\emptyset, \{N/1\})$,\\
	$(\{$\probloginline{hot :- rv_hot =:=1}$\}, \emptyset)$, \\
	$(\emptyset, \emptyset)$,\\
	$(\{$\probloginline{rv_hot~flip(0.2)}$\}, \{$\probloginline{rv_hot}$\twoheadleftarrow ($\probloginline{rv_hot_1}, \probloginline{true}$)\})$,\\
	$(\emptyset, \emptyset)$,\\
	$(\{$\probloginline{temp(1)~normal(27,5)}$\}, \{$\probloginline{temp(1)}$\twoheadleftarrow ($\probloginline{temp_1_1}, \probloginline{hot}$)\})$,\\
	$(\emptyset, \emptyset)$.

	The tag sequence for the rightmost derivation is\\
	$(\{$\probloginline{works(N):-temp(N)<25.0}$\}, \emptyset)$, \\
	$(\emptyset, \{N/1\})$,\\
	$(\{$\probloginline{hot :- rv_hot =:=1}$\}, \emptyset)$, \\
	$(\emptyset, \emptyset)$,\\
	$(\{$\probloginline{rv_hot~flip(0.2)}$\}, \{$\probloginline{rv_hot}$\twoheadleftarrow ($\probloginline{rv_hot_1}, \probloginline{true}$)\})$,\\
	$(\emptyset, \emptyset)$,\\
	$(\{$\probloginline{temp(1)~normal(20,5)}$\}, \{$\probloginline{temp(1)}$\twoheadleftarrow ($\probloginline{temp_1_2}, \probloginline{not hot}$)\})$,\\
	$(\emptyset, \emptyset)$.

	\begin{figure}[h]
		\begin{center}
			\begin{tikzpicture}
				{\small
					\node {(\probloginline{works(N)})} [sibling distance = 0.5\linewidth]{
						child [] { node {(\probloginline{cool(N)}) }
							child [] { node {(\probloginline{rv_cool(1)=:=1})}
								child [] { node {(\probloginline{rv(rv_cool(1),flip(0.99),true)}, \probloginline{rv_cool(1)=:=1})}
									child [] { node  {( \probloginline{rv_cool_1_1=:=1})}
										child [] {node {()} edge from parent [] node [left] { \scriptsize \ccomptwo } }
										edge from parent [] node [left] { \scriptsize \crvone }
									}
									edge from parent [] node [left] { \scriptsize \ccompone }
								}
								edge from parent [] node [left] { \scriptsize \cposlit }
							}
							edge from parent [] node [left,above] { \scriptsize \cposlit }
						}
						child [] { node {(\probloginline{temp(N)<25.0})} [sibling distance = 0.4\linewidth]
							child [] { node {} edge from parent [dashed] edge from parent [] node [left, above] { \scriptsize \ccompone   } }
							child [align=center] { node {} edge from parent [dashed] edge from parent [] node [right, above] { \scriptsize \ccompone  } }
							edge from parent [] node [right, above] {  \scriptsize \cposlit  }
						}
					};
				}
			\end{tikzpicture}
		\end{center}
		\caption{Left subtree of the resolution tree for the program in Example~\ref{ex:eliminating_pf} with respect to \probloginline{query(works(N))}; dashed lines continue in Figure~\ref{fig:sld_tree_right}.}
		\label{fig:sld_tree_left}
	\end{figure}

	\begin{figure}
		\begin{center}
			\begin{tikzpicture}
				{\small

					\node {} [sibling distance = 0.55\linewidth]
					child [] {node {(\probloginline{temp(N)<25.0})} edge from parent [dashed,level distance = 2.5cm] 
						child [align=center] {node { \probloginline{(hot},\\\probloginline{rv(temp(1),normal(27,5),hot)},\\ \probloginline{temp(1)<25.0}) } edge from parent [solid] 
							child [align=center] { node {(\probloginline{rv_hot=:=1},\\ \probloginline{rv(temp(1),normal(27,5),hot)},\\ \probloginline{temp(1)<25.0})}
								child [align=center] { node {(\probloginline{rv(rv_hot, flip(0.2), true) },\\  \probloginline{rv_hot=:=1}, \\ \probloginline{rv(temp(1),normal(27,5),hot)}, \\ \probloginline{temp(1)<25.0}) }
									child [align=center] { node {( \probloginline{rv_hot_1=:=1}, \\ \probloginline{rv(temp(1),normal(27,5),hot)}, \\ \probloginline{temp(1)<25.0}) }
										child [align=center, ] { node {( \probloginline{rv(temp(1),normal(27,5),hot)}, \\\probloginline{temp(1)<25.0} ) }
											child [ level distance = 1.5cm] { node {(\probloginline{temp_1_1<25.0})}
												child { node {()}  edge from parent [] node [left,] {  \scriptsize \ccomptwo  } }
												edge from parent [] node [left,] {\scriptsize \crvone } 
											}
											edge from parent [] node [left,] { \scriptsize \ccomptwo }
										}
										edge from parent [] node [left,] { \scriptsize \crvone }
									}
									edge from parent [] node [left,] { \scriptsize \ccompone }
								}
								edge from parent [] node [left,] { \scriptsize \cposlit }
							}
							edge from parent [] node [left,above] { \scriptsize \ccompone }
						}
						child [align=center] {node { (\probloginline{not hot},\\\probloginline{rv(temp(1),normal(20,5),not hot)},\\ \probloginline{temp(1)<25.0} } edge from parent [solid] 
							child [align=center] { node {(\probloginline{rv_hot=:=1},\\ \probloginline{rv(temp(1),normal(20,5),not hot)},\\ \probloginline{temp(1)<25.0})}
								child [align=center] { node {(\probloginline{rv(rv_hot, flip(0.2), true) }, \\ \probloginline{rv_hot=:=1}, \\ \probloginline{rv(temp(1),normal(20,5),not hot)}, \\ \probloginline{temp(1)<25.0}) } 
									child [align=center] { node {( \probloginline{rv_hot_1=:=1}, \\ \probloginline{rv(temp(1),normal(20,5),not hot)},\\\probloginline{temp(1)<25.0}) }
										child [align=center] { node {( \probloginline{rv(temp(1),normal(20,5),not hot)}, \\\probloginline{temp(1)<25.0}) }
											child [ level distance = 1.5cm] { node {(\probloginline{temp_1_2<25.0})}
												child { node {()}  edge from parent [] node [right,] { \scriptsize \ccomptwo } }
												edge from parent [] node [right,]  { \scriptsize \crvone  }
											}
											edge from parent [] node [right]  { \scriptsize \ccomptwo  }
										}
										edge from parent [] node [right]  { \scriptsize \crvone  }
									}
									edge from parent [] node [right] { \scriptsize \ccompone  }
								}
								edge from parent [] node [right] { \scriptsize \cneglit }
							}
							edge from parent [] node [right,above] { \scriptsize \ccompone}
						}
						edge from parent [] node [right] {\scriptsize \cposlit }
					}
					;
				}
			\end{tikzpicture}
		\end{center}
		\caption{Right subtree of the resolution tree for the program in Example~\ref{ex:eliminating_pf} with respect to \probloginline{query(works(N))}; dashed line continued from Figure~\ref{fig:sld_tree_left}}\label{fig:sld_tree_right}
	\end{figure}
\end{example}

\newpage
\clearpage

%The resolution tree functions as an intermediate representation for obtaining the relevant ground program. Defining a tagging function for each edge in the resolution tree will allow us to specify an algorithms that maps a path in a tree (also called derivation) to its corresponding relevant ground program.

%\begin{definition}[Edge Tagging in  Resolution Tree]
%\label{def:sld_edge_tagging_rules}
%Let $\mathcal{T}$ be a resolution tree. We tag each edge in the tree with a tuple $\tau=(\tau_r,\tau_s)$, where $\tau_r$ represents a set of logic rules and $\tau_s$ a substitution. The tag of an edge of the tree $\mathcal{T}$ depends on the condition ({\bf C1-C6}) that held when the edge was created.
%We now define the six tagging rules for edges in a resolution tree in function of the conditions that apply.

%\begin{enumerate}
%    \item[\cposlit] If the goal is $(l_1, l_2, \dots, l_n)$ and if  $\exists \theta$ such that  $h \lpif b_1, \dots, b_n \in \dcpprogram$, $\theta  = mgu(l_1, h)$ then:
%   \begin{align*}
	%      {\text{\bf \tposlit:}} \quad \tau = (\{ h \lpif b_1, \dots, b_n \}, \theta)
	% \end{align*}

% \item[\bf \cneglit]  If th goal is $(not(l_1), l_2, \dots, l_n)$ and if $\exists \theta$ such that  $h \lpif b_1, \dots, b_n \in \dcpprogram$, $\theta  {=} mgu(l_1, h)$ then:
%\begin{align*}
%   {\text{\bf \tneglit:}} \quad \tau = (\{ h \lpif b_1, \dots, b_n \}, \theta)
%\end{align*}

%\item[\ccompone]  If the goal is $( l_{comp}, l_1, \dots, l_n )$ and if $l_{comp}$ is a comparison atom (or its negation) and if there exists a substitution $\theta$ such that for an arbitrary symbol $s \in symbols(l_{comp})\setminus keys(T_{RV})$ it holds that $h\sim d\lpif b_1, \dots, b_n\in \dcpprogram$, with $\theta = mgu(s,h)$, then:
%\begin{align*}
%   {\text{\bf \tcompone:}} \quad \tau = (\emptyset, \theta)\\
%\end{align*}

%\item[\bf \ccomptwo]  If the goal is $( l_{comp}, l_1, \dots, l_n )$  and if $l_{comp}$ is a comparison atom (or its negation) and if $symbols(l_{comp})\subseteq keys(T_{RV})$ then:
%\begin{align*}
%   {\text{\bf \tcomptwo:}} \quad \tau = (\emptyset, \emptyset)
%\end{align*}

%  \item[\bf \crvtwo]  If the goal is $(rv(x, d, (b_1, \dots, b_{n})), l_1,\dots, l_m)$ and if there exists a substitution $\theta$ such that for an arbitrary symbol $s \in symbols(d)\setminus keys(T_{RV})$ it holds that $h\sim d^h \lpif b^h_1, \dots, b^h_{n^h}\in \dcpprogram$, $\theta = mgu(s,h)$, then:
%  \begin{align*}
	%      {\text{\bf \trvtwo:}} \quad \tau =  (\emptyset, \theta)\\
	%  \end{align*}

%  \item[\bf \crvone]  If the goal is $(rv(x, d, (b_1, \dots, b_{n})), l_1,\dots, l_m)$ and if  $symbols(d)\subseteq keys(T_{RV})$ then:
%  \begin{align*}
	%          {\text{\bf \trvone:}} \quad \tau =  (\{x\sim d \}, \{ x/ x_{b_1,\dots,b_n} \})
	%   \end{align*}
%\end{enumerate}
%\end{definition}

%\begin{definition}[Derivation]
%Let $\dcpprogram$ be a \dcproblogsty program, and $Q$ a query atom. We call a path in the   resolution tree of $Q$ for \dcpprogram a derivation of $Q$. A derivation consists of a set of edges $\{e^0,\dots, e^N  \}$.
%Edges in a derivation are ordered, starting at the root.
%We call a derivation tagged if the rules in Definition~\ref{def:sld_edge_tagging_rules} have been applied.
%\end{definition}

In order to define the ground program corresponding to a tag sequence, we first need to define the rewrite operation introduced by \trvone.
\begin{definition}[RV substitution]\label{def:context-substitution}
	Let $\theta=\{x\twoheadleftarrow (v, (b_1, \dots ,b_m))\}$ be a random variable substitution. Then,
	\begin{align*}
		(h \lpif l_1\dots, l_n)\theta &= (h\lpif b_1, \dots, b_m,l_1, \dots, l_n)\{x/v\}, &\text{~if~x~is~interpreted~in~some~}l_i\\
		s\theta &= s\{x/v\}, &\text{~otherwise}
	\end{align*}
	where we overload regular substitution notation for random variable terms, \ie, $c\{x/v\}$ is $c$ with every occurrence of $x$ replaced by $v$.
\end{definition}
That is, we add the  context to the bodies of all rules that interpret $x$, \ie, whose body contains a comparison literal involving $x$, and replace all occurrences of $x$ by the new variable $v$.
%replace the old random variable term $x$ by the context-dependent random variable $v$ in all statements,  and  
% we do not need to substitute parents in distribution terms, as (1) R5 ensures that R6 is applied to all parents before it is applied to the child, (2) the moment R6 is applied to a parent, there is a rv/3 atom for the child in the rest of the goal, and R6 substitutes the new parent name there, thus (3) by the time the T6 tag of the child is created, the distribution contains the parent's new name
Algorithm~\ref{alg:tags2gp} uses this rewrite operation, together with regular substitutions included in other types of tags, to extract a relevant ground program from a tag sequence.

\begin{algorithm}[h!]
	\SetKwInput{KwOutput}{Output}              
	\DontPrintSemicolon
	\KwInput{Derivation tag sequence $S=((P_1,\theta_1),\ldots,(P_N,\theta_N))$}
	\KwOutput{Corresponding Relevant Ground Program}

	$\dcpprogram_S \leftarrow \emptyset$\;
	\For{$i \in \{0,\dots, N\}$  }{
		$\dcpprogram_S \leftarrow \dcpprogram_S \cup P_i$\;
		$\dcpprogram_S \leftarrow  \dcpprogram_S\theta_i$
	}

	\Return $\dcpprogram_S$
	\caption{Relevant Ground Program for a Derivation Tag Sequence}
	\label{alg:tags2gp}
\end{algorithm}
%Note that applying a tag's substitutions directly after adding its rules is crucial, as several tags may substitute the same random variable terms

%\begin{algorithm}[h!]
% \SetKwFor{For}{for (}{}
%
%\newlength\mylen
%\newcommand\myinput[1]{%
	%  \settowidth\mylen{\KwIn{}}%
	%  \setlength\hangindent{\mylen}%
	%  \hspace*{\mylen}#1\\}
%
%
%\SetKwInput{KwOutput}{output}              
%\DontPrintSemicolon
%\KwInput{Tagged derivation $\mathcal{R}$ (of depth $N$)}
%\myinput{Table $tags$ storing the tags of the edges in $\mathcal{R}$}
%\KwOutput{Relevant Ground Program corresponding to $\mathcal{R}$}
%
%
%$RelGroundProgram \leftarrow \emptyset$\;
%\For{$i \in \{0,\dots, N\}$  }{
	%    $\tau_r^i, \tau_s^i \leftarrow tags[i] $\;
	%    $RelGroundProgram \leftarrow RelGroundProgram \cup \tau_r^i$\;
	%    $RelGroundProgram \leftarrow  RelGroundProgram/\tau_s^i$
	%}
%
%\Return $RelGroundProg$
%\caption{Relevant Ground Program with Respect to a Derivation}
%\label{alg:tags2gp}
%\end{algorithm}

%\begin{definition}[Relevant Ground Program]
%Let \dcpprogram be an AD-free program, $Q$ a query atom, and  $\mathcal{T}$ the resolution tree obtained from applying the resolution rules in Definition~\ref{def:resolution_rules} to  \dcpprogram and $Q$, where all derivations have been tagged. We denote these derivations by  $\{\mathcal{R}_0, \dots,  \mathcal{R}_M \}$  and their ground programs as obtained from Algorithm~\ref{alg:tags2gp} by $\dcpprogram
%_g^{\mathcal{R}} =\{\dcpprogram_g^{\mathcal{R}_0}, \dots, \dcpprogram_g^{\mathcal{R}_M}  \}$. The relevant ground program of \dcpprogram with respect to the atom of interest $Q$, denoted by $\dcpprogram_g^Q$, is then given by: $ \dcpprogram_g^Q = \bigcup_{i\in \{0,\dots,M  \}} \dcpprogram_g^{\mathcal{R}_i}  $.
%\end{definition}

\begin{example}
	Algorithm~\ref{alg:tags2gp} constructs the following program for the rightmost tagging sequence in the example:
	\begin{problog*}{linenos}
works(1) :- not hot, temp_1_2<25.0.
hot :- true, rv_hot_1 =:= 1.
rv_hot_1~flip(0.2).
temp_1_2~normal(27,5).
	\end{problog*}
	Note that all random variable terms have been replaced by the new random variables, and appropriate contexts added to the rules interpreting them. 
\end{example}

\begin{definition}[Relevant ground program for an atom]
	Let \dcpprogram be an AD-free program, $A$ an atom and $\mathcal{S}$ be the set of all derivation tag sequences for $A$ in \dcpprogram. The \emph{relevant ground program with respect to $A$}, denoted by $\dcpprogram_g^A$, is given by $\dcpprogram_g^A = \bigcup_{S\in\mathcal{S}}\dcpprogram_S$.
\end{definition}

\begin{definition}[Relevant ground program]
	Let \dcpprogram be an AD-free program, $\queryset$ a set of queries of interest, and  $\evidenceset=e$ the evidence. The \emph{relevant ground program}  is $\dcpprogram_g = \bigcup_{A\in (\queryset\cup\evidenceset)}\dcpprogram_g^A$.
\end{definition}

\begin{example}[Relevant Ground Program] 
	%Applying the edge tagging and the definition of a relevant ground program  to the resolution tree in Example~\ref{ex:resolution_tree} results in the following ground program.
	The full ground program obtained from the derivations in Example~\ref{ex:resolution_tree} is
	\begin{problog*}{linenos}
rv_hot_1 ~ flip(0.2).
hot :- true, rv_hot_1=:=1.
rv_cool_1_1 ~ flip(0.99).
cool(1) :- true, rv_cool_1_1=:=1.

temp_1_1 ~ normal(27,5).
temp_1_2 ~ normal(20,5).

works(1):- cool(1).
works(1):- hot, temp_1_1<25.0.
works(1):- not hot, temp_1_2<25.0.
	\end{problog*}
	Note that lines 1 and 2 are generated by both the middle and the rightmost derivation, while all other statements stem from a single derivation in this example.	
	%Note how the `$\mathrm{createRVname}$' function created unique names for the random variables in the program, which are then also inserted into the arguments of the comparison terms. For instance \probloginline{temp_1_2} in the the comparison \probloginline{temp_1_2<25.0}. 
\end{example}

In order to implement an efficient resolving algorithm, one ought to avoid processing the same atom twice. This is usually done through tabling, which also avoids going into an infinite loop when cyclical rules are present in the (logic part of the) program. 
%Note also that if we had had multiple query atoms, instead of a single one, our approach would not change dramatically. The main difference would lie in having a multi-rooted resolution tree instead of a single root tree.
Note also that while the relevant ground program computed by this approach does not contain any rules that are disconnected from the reasoning process related to queries and evidence, it is not necessarily the smallest such program. It may for instance include rules whose bodies are logically inconsistent because of added conjunctions of contexts, or invalidated by the actual values of the evidence.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_excluded/types.tex
aij dcproblog paper

\section{Typing \dcproblogsty}
\label{app:types}

\subsection{Type System}\label{sec:type_system}

A distinctive property of system predicates, which are implemented using an external engine, is their mandatory type signature.
An instance of such system predicates are comparison predicates.
Consider for example the comparison (\probloginline{x>20}).
The right-hand side is an integer and if the \probloginline{x} denotes a numeric random variable we have a well-typed comparison predicate inducing a measurable set.
Consider for the sake of contrast also the comparison predicate (\probloginline{a>2}), where \probloginline{a} is an atomic formula. This comparison predicate is ill-typed as atomic formulas cannot be compared to (integer) numbers. In other words, the external engine, which implements the system predicates does not know how to commensurate an atomic formula and a number.
A type system lets us specify which combinations of types are allowed as arguments of comparison predicates, \ie which comparisons encode measurable sets and are supported by the external engine.

\begin{samepage}
\begin{definition}\label{def:type_system}
	The \dcproblogsty type system is a hierarchical type system with each type being a disjoint union of its subtypes.
	The base type for \dcproblogsty is the \type{Term} type. We subtype terms with a \type{Distribution} type and a \type{Symbol} type. 
    A diagrammatic overview of the \dcproblogsty type system is given in Figure~\ref{fig:type_system}.
	\begin{enumerate}
		\item[] \type{Term} 
		\begin{enumerate}
			\item[1.] \type{Distribution}
			\begin{itemize}
				\item \type{NumericD}
				\item \type{CategorialD}
			\end{itemize}
			\item[2.] \type{Symbol}
			\begin{itemize}		
				\item \type{RandomSymbol}
				\begin{itemize}
				    \item \type{NumericRV}
				    \item \type{CategoricalRV}
				\end{itemize}
				\item \type{DeterministicSymbol}
				\begin{itemize}
				    \item \type{Number}
				    \item \type{Atom}
				\end{itemize}	
			\end{itemize}
		\end{enumerate}
	\end{enumerate}
\end{definition}
\end{samepage}

\begin{figure}[t]
	\begin{center}
		\includegraphics[width=\linewidth]{../figures/type_hierarchy_reduced.pdf}
	\end{center}
	\caption[Diagrammatic representation of the hierarchical \dcproblogsty type system.]{Diagrammatic representation of the hierarchical \dcproblogsty type system.}
	\label{fig:type_system}
\end{figure}

\begin{definition}[\type{Distribution}]
	Terms of type \type{Distribution} are constructed by using a reserved random function symbol and applying it to a (possible empty) tuple of input arguments, which maps the input tuple to a distribution term.
	Distribution terms are only allowed to be invoked as a second argument to the reserved distribution predicate \probloginline{~}\lstinline[columns=fixed]|/2|.
	This also means that random function symbols are only used there as well. 
% 	\dcproblogsty subtypes the \type{Distribution} type into \type{NumericD} and \type{CategoricalD}.
% 	\type{DiscreteD}, \type{ContinuousD}, and \type{MixtureD} types.
\end{definition}

\type{Distribution} terms determine the probability distribution according to which a random variable is distributed.
\probloginline{poisson(6)} and \probloginline{normal(20,5)} are for instance of type \type{NumericD}.

% Note that distributions of type \type{ContinuousD} denote {\em absolutely continuous} probability distributions\footnote{\dcproblogsty does not support {\em singular continuous} probability distributions such as the Cantor distribution.}. \ak{would be nice to have a hint on why this restriction is there} We give an example of a distribution of type \type{MixtureD} in Example~\ref{ex:indian_gpa_function}.
%\begin{example} \probloginline{mixed} is a random variables that is a mixture of a continuous part (the truncated normal between $0$ and $4$ with mean $3$ and standard deviation $1$) and a discrete part. 
%\begin{problog}
%mixted~Mix((0.998,TruncNormal(3,1,0,4)),(0.001,4),(0.001,4))
%\end{problog}
%\end{example}

%For and example, see Example~\ref{example:dcproblog_poisson_population}: we have the random function symbol \probloginline{poisson}\lstinline[columns=fixed]{/1}, apply it to the input tuple \probloginline{(6)}, and obtain a term of type \type{Distribution}.

\begin{definition}[\type{Symbol}]
  \type{Symbol} terms are the building blocks of the discrete structure that makes up a logic program and as such determine also the deterministic system of a probabilistic logic program. Beyond the reserved syntax, which is used to define the distribution terms, the user is free to use any identifier (following \prologsty conventions) to declare \type{Symbol} terms (this includes standard \prologsty terms such as lists).  
\end{definition}

%The subtyping of the \type{Symbol} type (shown in Figure~\ref{fig:type_system}) follows the convention given in~\citep[Chapter 9]{sterling1994art}.

% \begin{definition}[\type{Arithmetic}]
% 	Terms of type \type{Symbol} for which arithmetic operations are defined are of type \type{Arithmetic}. \type{Arithmetic} terms can be used as second argument to the builtin \probloginline{is}\lstinline[columns=fixed]|/2| functor. \type{Arithmetic} terms are subtyped with \type{RandomVariable} terms and \type{Number} terms.
%	\ak{Def 6.3 Arithmetic terms: if (as stated in def 6.1) each type is the disjoint union of its subtypes, that means that "arithmetic" is made up exclusively of numbers and RVs, \ie, terms like 3-2 or min(x,5) for a RV x are not allowed, whereas the first two sentences of def 6.3 suggests they might be allowed (only having numbers and RVs directly on the rhs of is/2 seems quite boring...). Which is it?
%		Related to this, defs 6.8 and 6.10 talk about RVs and Numbers, why not simply "arithmetic"?}
% \end{definition}

\begin{definition}[\type{RandomSymbol}] \label{def:random_variable}
	\type{RandomSymbol} terms are named by the programmer using distributional facts with the binary infix predicate \probloginline{~}\lstinline[columns=fixed]|/2| and inherit their subtype from the distribution term. All random variables present in a \dcproblogsty program  explicitly named.
\end{definition}

Examples of terms of type \type{RandomSymbol} would be \probloginline{temperature} and \probloginline{n_people} in Examples~\ref{example:dcproblog_machine} and~\ref{example:dcproblog_poisson_population}, where we declared the random variables through distributional facts: 
\begin{problog}
temperature ~ normal(20,5).
n_people ~ poisson(6).
\end{problog}
As mentioned in Defintion~\ref{def:random_variable}, \type{RandomSymbol} terms inherit their type from the associated \type{Distribution} term, \eg a \type{NumericD} term induces a random variable term of type \type{NumericRV}.

Note that not all probabilistic programming languages enforce the naming of random variables. In the functional probabilistic programming language \anglicansty~\citep{wood2014approach}, for instance, the user can declare anonymous random variables, much like anonymous functions, \ie \textlambda-functions. 
\ak{Discussion of named/unnamed between Defs 6.4 and 6.5: would be nice to have a brief hint on why you made that choice} \pedro{usually in logic programming everything is global. an unamed random variable would mean that it is only available locally. we could allow that, I guess but would have to introduce new syntax. moreover, internally we would give the unnamed random variable a name anyway.}

\begin{definition}[\type{Atom}]
	An atom is of the form \probloginline{p(t@$_1$@,@\dots @, t@$_n$@ )} where \dcplpinline{p} is a predicate with arity $n$ and the \probloginline{t@$_i$@} are \type{Symbol} terms. Functors used to construct \type{Atom} terms must not clash with the builtin predicates and reserved functors used to declare \type{Distribution} terms.
% 	Terms of type \type{Atom} constitute the building blocks of the discrete structure of a probabilistic logic program.
\end{definition}

Examples of \type{Atom} terms would be \probloginline{foo} or \probloginline{bar(foo,42)}. The latter is constructed using the predicate \lstinline{bar/2} and the tuple of \type{Symbol} terms \probloginline{(foo,42)}. \probloginline{foo} is of type \type{Atom} and \probloginline{42} is a \type{Number} term.

\begin{definition}[\type{Number}] Terms of type \type{Number} are used to express numerical data types such as integers and real numbers.	
\end{definition}

\paragraph{Strongly Typed \prologsty}

While \prologsty is traditionally not strongly/statically typed, there have been efforts to extend \prologsty with such a type system. Arguably, the most interesting effort is presented in~\citep{Schrijvers2008TowardsTP}, where the authors develop the idea of partially typing a \prologsty program, \ie typing is optional and untyped code is interpreted instead of compiled. Introducing predicates with strongly typed signatures can be regarded as adding predicates with a specific meaning to the \prologsty implementation but the {\em `evaluation'} of these predicates still takes place within \prologsty.
Our efforts of introducing a type system for system predicates within a probabilistic programming context are orthogonal and complimentary to the efforts of~\citet{Schrijvers2008TowardsTP}.

% \dcproblogsty's type system refines the one of DC-PLP to instantiate a conrete version of the abstract language.  Specifically, we have the following refinements:
% \begin{enumerate}
%     \item distribution functors are further partitioned into \emph{discrete distribution functors}. \emph{continuous distribution functors} and \emph{mixture distribution functors}
%     \item arithmetic functors are further partitioned into \emph{operation functors} and \emph{number functors}, with the latter in turn partitioned into \emph{integers} and \emph{floats} \todo{do we have / want a complete list of operations? should we at least give some examples?}
%     \item we use Boolean query functors 	$\bowtie=\{ \text{\probloginline{=:=}}, \text{\probloginline{=\=}}, \text{\probloginline{<}}, \text{\probloginline{>}},  \text{\probloginline{=<}}, \text{\probloginline{>=}} \}$
% 	in infix notation (and lift them to predicate level as discussed before)
% \end{enumerate}

% \ak{I don't think we need to sub-type random variable functors here, as rv terms inherit the \emph{output} type of the distribution only, which should be integer, float etc, not "discrete distribution"}

\subsection{External Engine}\label{sec:external_engine}

In the previous subsection we have introduced an abstract type system.
Abstract types describe the structure/hierarchy of a type system and cannot be instantiated. Concrete types, which are subtypes of abstract types are used for instantiations. This means that abstract types relate collections of concrete types, \ie implementations, to each other. For example, the \type{Integer} \probloginline{3} could be instantiated as an \type{Int32} or also an \type{Int64}, which are both concrete types. The choice taken depends on the external engine used to evaluate arithmetic expressions.

The only condition that we necessitate for an implementation of \dcproblogsty is that the external engine provides concrete types for the abstract types. Additionally, a practical external engine should also implement operations on these types. For example, an engine should be able to multiply a floating point value and an integer value: (\probloginline{3.12*2}). The type for floating points could be dubbed \type{Float32} and would be a concrete type subtyped from the abstract type \type{Real}. The type of the integer value could be \type{Int32} subtyped from \type{Integer}. The return type could then again be \type{Float32}.

\begin{example}\label{example:julia_addition}
We show two functions that perform addition of numbers but with different types. These different number could be invoked on either side of a comparison predicate, \eg (\probloginline{1<3+5} ) or (\probloginline{1<3+5.0} ). We write these functions using Julia like syntax~\citep{bezanson2017julia}. 
\begin{samepage}

\begin{minted}[
    breaklines,
    escapeinside=||,
    mathescape=true, 
    linenos,
    xleftmargin=20pt, 
    % numbersep=3pt, 
    % gobble=2, 
    % framesep=2mm
]{julia}
function +(a::Int32, b::Int32)::Int32
    return int32_addition(a,b)
end;
\end{minted}

\end{samepage}

\begin{samepage}
\begin{minted}[
    breaklines,
    escapeinside=||,
    mathescape=true, 
    linenos,
    xleftmargin=20pt, 
    % numbersep=3pt, 
    % gobble=2, 
    % framesep=2mm
]{julia}
function +(a::Int32, b::Float32)::Float32
    float_a = convert(Float32, a)
    return float32_addition(float_a,b)
end;
\end{minted}

\end{samepage}
\end{example}

The programming paradigm followed in Example~\ref{example:julia_addition} is called {\em multiple dispatch} \citep{keene1989object,castagna1995calculus,bezanson2017julia}.
Multiple dispatching is the process of dynamically selecting at run time a specific implementation of a function based on the types of the arguments.
Functions with the same functor but a different type signature are called multimethods.

We implement probability distributions using the external engine. As such, the corresponding random funtions symbols are necessarily typed. We recede again to multiple dispatch to implement probability distributions.
Consider, for instance, the declaration of the normal distribution in Example~\ref{example:dcproblog_machine}.
The definition of the function is the following:
$$ \text{\probloginline{normal}}: \type{Number} \times \type{Number} \rightarrow \type{ContinuousD}$$
However, the arguments of a normal distribution (its mean and standard deviation) might as well be random variables themselves.
Multiple dispatching then allows us to {\em overload} the \probloginline{normal} function symbol by using arguments of a different type\footnote{For simplicity we omit that the standard deviation has to be a strictly positive (real) number, which can be enforced by extending the type system.}:
\begin{align*}
	\text{\probloginline{normal}}&: \type{NumericRV} \times \type{Number} \rightarrow \type{ContinuousD} \\
	\text{\probloginline{normal}}&: \type{Number} \times \type{NumericRV} \rightarrow \type{ContinuousD} \\
	\text{\probloginline{normal}}&: \type{NumericRV} \times \type{NumericRV} \rightarrow \type{ContinuousD}
\end{align*}

Reusing the idea of multimethods in the context of probabilistic logic programming with random function symbols results in a simple and neat syntax.
The exact implementation of these random function symbols is left to to the actual implementation of the language by providing an external engine capable of performing adequate arithmetic evaluations.

Currently, \dcproblogsys, our implementation of \dcproblogsty, does only emulate multimethods. This is due to the fact that the \dcproblogsys is written in Python, which does not follow the multiple dispatch programming paradigm. The emulation of multimethods is performed by using the \verb|isinstance| function in Python, which checks whether an object is of a certain type or not. This allows us to branch on the type of the arguments passed to a function and call different implementation given differently typed arguments.

An advantage of decoupling the deterministic system from the independent choices and implementing probability distribution using an external engine is that we can a single language with two different external engines, which might perform probabilstic inference differently.
\dcproblogsys, for instance, provides an interface to two different external engines. The first one uses \pyrosty~\citep{bingham2019pyro}, a probabilistic programming language built on top of the PyTorch~\citep{paszke2019pytorch} deep learning framework.
The second option uses the probabilistic programming language PSI~\citep{gehr2016psi}. PSI is a computer probabilistic computer algebra system. The main difference between the two different external engines becomes apparent when performing inference. The first one performs approximate inference while the second one exact inference. We will discuss this in more detail in Section~\ref{sec:inference}.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_excluded/wmi.tex
aij dcproblog paper

\section{Weighted Model Integration}
\label{app:wmi}

Standard weighted model counting only supports discrete probability distributions. To repair this omission, WMC has recently been extended towards weighted model integration (WMI)~\citep{belle2015probabilistic}, supporting additionally continuous variables. Contrary to WMC, which is formulated using weighted propositional logic formulas, WMI uses so-called {\em weighted  satisfiability modulo theory} (SMT) formulas. We refer the reader to~\citep{barrett2009handbook} for a treatise on (unweighted) SMT formulas.

\begin{example}\label{example:wmi_intro}
    Consider the theory $\mathtt{broken}$
	\begin{align}
	\textstyle{\texttt{broken} \leftrightarrow ( \texttt{no\_cool} \land (\texttt{t}> 20) )  \lor (\texttt{t}> 30)} \label{eqn:broken}
	\end{align}
	where \texttt{no\_cool} is a Boolean variable and \texttt{t} a real-valued variable.
    Let us assume that \texttt{t} is distributed  according to $\texttt{t} \sim \mathcal{N}_\texttt{t}(20,5)$ and that the probability for $\texttt{no\_cool}$ being true is $0.01$. Determining the probability of the formula being true/satisfied is the weighted model integration problem.
\end{example}

We see in Example~\ref{example:wmi_intro} that SMT formulas generalize traditional propositional logic formulas by additionally allowing the use of expressions formulated in a background theory. More formally, following \cite{zuidbergdosmartires2019exact}:

\begin{definition}[SMT($\mathcal{RA}$) (real arithmetics)] \label{def:smtra}
	Let $\bvars$ be a set of $M$ Boolean and $\xvars$ a set of $N$ real variables.
	An {\bf atomic formula} is an expression of the form $g(X) \bowtie c$, where $c \in \mathbb{Q}$, $\bowtie \in \{ =,\neq, \geq,\leq,>,< \}$, and  $g:\mathbb{R}^{N} \rightarrow \mathbb{R}$. The symbols $\mathbb{Q}$  and $\mathbb{R}$ denote the rational number and real numbers, respectively. 
	We then define SMT($\mathcal{RA}$) formulas as Boolean combinations (by means of the standard Boolean operators $\{\neg, \land, \lor, \rightarrow, \leftrightarrow \}$) of {\bf Boolean variables} $b_i \in \bvars$ and of {\bf atomic formulas} over $\xvars$.

	We distinguish two special cases:
	\begin{itemize}
		\item SMT($\mathcal{NRA}$) (non-linear real arithmetics): atomic formulas take the form $\sum_i c_i \cdot x_i^{p_i}\bowtie c$, where the $x_i \in \xvars$ and $c_i,c, p_i \in \mathbb{Q}$.  
		\item SMT($\mathcal{LRA}$) (linear real arithmetics): atomic formulas take the form $\sum_i c_i \cdot x_i \bowtie c$, where the $x_i \in \xvars$ and $c_i,c \in \mathbb{Q}$.
	\end{itemize}
\end{definition}

We define weighted model integration in terms of an indicator function over an SMT formula, cf.~\citep{kolb2019exploit}.
\begin{definition}[{\bf WMI}]\label{def:wmi}
Given a set $\bvars$ of $M$ Boolean variables,  $\xvars$ of $N$ real variables, a weight function $\pweight:
\mathbb{B}^M {\times} \mathbb{R}^N {\rightarrow} \mathbb{R}_{\geq 0}$, and a support $\support$, in the form of an SMT formula, over $\bvars$ and  $\xvars$, the weighted model integral is given by:
\begin{align}
  \wmi(\support, \pweight {\mid} \xvars, \bvars) = \sum_{\bvars} \int \ive{\support(\xvars, \bvars)} \pweight({\xvars},{\bvars}) d{\xvars} \label{eqn:def_wmi}
\end{align}
where we use the Iverson bracket notation in $\ive{\support(\xvars, \bvars)}$ to denote the indicator function of $\support(\xvars, \bvars)$. If the set of real variables $\xvars$ is empty the WMI task reduces to WMC (cf.~\ref{app:amc}).
\end{definition}

\citet{kolb2019exploit} further manipulate the expression in Equation~\ref{eqn:def_wmi} in order to
% get rid of the Boolean dependencies in the weights function $\pweight$ and to 
transform the $\lor$-$\land$ structure present in $\ive{\support(\xvars, \bvars)}$ into a simple conjunction of SMT atoms.
\begin{align}
\wmi(\support, \pweight {\mid} \xvars, \bvars)&=\sum_{\wsupport_i \in \wtuples} \sum_{\bvars} \int \llbracket  \wsupport_i  (\xvars, \bvars) \rrbracket \pwweight(\xvars, \bvars) d\xvars \label{eqn:wmi_sum_ive_weight} \\
&=\sum_{\wsupport_i \in \wtuples} \sum_{\bvars} \int \Bigg\llbracket  \bigwedge_{j} \wsupport_{ij} (\xvars, \bvars) \Bigg\rrbracket \pwweight(\xvars, \bvars) d\xvars 
\end{align}
The weighted model integral in Equation~\ref{eqn:def_wmi} over a support with disjunctions and conjunctions, has become now a sum of multiple weighted model integrals whose supports are simple conjunctions of SMT atoms. $\wtuples$ denotes the set of mutually exclusive SMT formulas whose disjunction is equivalent to the original SMT formula $\support$, \ie $\support\leftrightarrow \lor_i \support_i \leftrightarrow \lor_i \land_j  \support_{ij}$.

% For the purely Boolean case, they correspond to the possible worlds of $\support$.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_main/conclusions.tex
aij dcproblog paper

\section{Conclusions}\label{sec:conclusions}

We introduced \dcproblogsty, a hybrid PLP language for the discrete-continuous domain and its accompanying measure semantics.
\dcproblogsty strictly extends the discrete \problogsty language \citep{de2007problog,fierens2015inference} and the negation-free \dcsty~\citep{gutmann2011magic} language.
In designing the language and its semantics we adapted ~\citet{poole2010probabilistic}'s design principle of percolating probabilistic logic programs into two separate layers: the random variables and the logic program.
% , and applied it to both the discrete and the continuous random variables. 
 Boolean comparison atoms then form the  link between the two layers.
It is this clear separation between the random variables and the logic program that has allowed us to use simpler language constructs and to write programs using a  more concise and intuitive syntax than alternative hybrid PLP approaches \citep{gutmann2010extending,nitti2016probabilistic,speichert2019learning,azzolini2021semantics}.

Separating random variables from the logic program also allowed us to develop the IALW algorithm to perform inference in the hybrid domain. 
IALW is the first algorithm based on knowledge compilation and algebraic model counting for hybrid probabilistic programming languages
and as such it generalizes the standard knowledge compilation based approach for PLP.
It is noteworthy that IALW correctly computes conditional probabilities in the discrete-continuous domain using the newly introduced infinitesimal numbers semiring.

Interesting future research directions include adapting ideas from functional probabilistic programming (the other declarative programming style besides logic programming) in the context of probabilistic logic programming. For instance, extending \dcproblogsty with a type system~\citep{Schrijvers2008TowardsTP} or investigating more recent advances, such as {\em quasi-Borel spaces}~\citep{heunen2017convenient} in the context of the measure semantics.

% We presented \dcproblogsty, a PLP language for the discrete-continuous domain. \dcproblogsty strictly extends the discrete \problogsty language \citep{de2007problog,fierens2015inference} and the negation-free \dcsty~\citep{gutmann2011magic} language.
% In designing the language we followed the approach of percolating probabilistic logic programs into two layers 1) random variables 2) logic program. We adapted this paradigm from~\citet{poole2010probabilistic} and applied it to the discrete-continuous setting. Boolean comparison atoms form the link between the two layers.

% When compared to other PLP languages following Sato's distribution semantics \citep{gutmann2010extending,nitti2016probabilistic,speichert2019learning,azzolini2021semantics}, the clear separation between the random variables and the logic program has allowed us to eliminate bloated language constructs and to write programs using a concise and intuitive syntax.

% With IALW, we have also presented the first algorithm based on knowledge compilation and algebraic model counting for hybrid probabilistic programming languages. It is noteworthy that IALW correctly computes conditional probabilities in the discrete-continuous domain using the newly introduced infinitesimal numbers semiring.

% Besides providing an efficient implementation of the \dcproblogsty language together with the IAWL inference algorithm, 
% interesting future directions of research might include incorporating ideas from functional probabilistic programming (the second declarative programming style besides logic programming). For instance, extending \dcproblogsty with a type system~\citep{Schrijvers2008TowardsTP} or investigating more recent advances, such as {\em quasi-Borel spaces}~\citep{heunen2017convenient} in the context of the distribution semantics.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_main/dc2smt.tex
aij dcproblog paper

\section{Inference via Computing Expectations of Labeled Logic Formulas} \label{sec:dc2smt}

In the previous sections we have delineated the semantics of \dcproblogsty programs and described the PROB task that defines conditional probability queries on \dcproblogsty programs. The obvious next step is to actually perform the inference. We will follow an approach often found in implementations of PLP languages in the discrete domain: reducing inference in probabilistic programs to 
performing inference on labeled Boolean formulas that encode relevant parts of the logic program.
Contrary to languages in the discrete domain that follow this approach~\citep{fierens2015inference,riguzzi2011pita}, we will face the additional complication of handling random variables with infinite sample spaces. We refer the reader to~\citep[Section 5]{riguzzi2018foundations} for a broader overview of this approach.

Specifically, we are going to define a reduction from \dcproblogsty inference to the task of computing the expected label of a propositional formula. The formula is a propositional encoding of the  relevant part of the logic program (relevant with respect to a query), where atoms become propositional variables, and the labels of the basic facts in the distribution database are derived from the probabilistic part of the program.
At a high level, we extend \problogsty's inference algorithm such that Boolean comparison atoms over (potentially correlated) random variables are correctly being kept track of. The major complication, with regard to \problogsty and other systems such as PITA~\citep{riguzzi2011pita}, is the presence of context-dependent random variables, which are denoted by the same ground random term. For instance, the random term \probloginline{size} in the program in Example~\ref{example:dcproblog:observation} denotes two different random variables but is being referred to by one and the same term in the program.

Inference algorithms for PLP languages often consider only a fragment of the language for which the semantics have been defined. A common restriction for inference algorithms is to only consider range-restricted programs\footnote{We call a \dcproblogsty program range-restricted if it holds that for every statement all logic variables occurring in the head also occur as positive literals in the body. This guarantees that all terms will become ground during backward chaining. Note that range-restrictedness implies that all facts (including probabilistic and distributional ones) are ground.}. 
% , the finite support assumption is usually presumed. That is, the queries and evidence depend only on a finite subset of Boolean comparison atoms (and thus a finite set of random variables),
Furthermore, we consider, without loss of generality only AD-free programs, cf.~Definition~\ref{def:ad_free_program}, as annotated disjunctions or probabilistic facts can be eliminated up front by means of {\em local} transformations that  solely affect the annotated disjunctions (or probabilistic facts).\footnote{For non-ground ADs, we adapt Definition~\ref{def:elim-ad} to include all logical variables as arguments of the new random variable. As this introduces non-ground distributional facts, which are not range-restricted, we also move the comparison atom to the end of the rule bodies of the AD encoding to ensure those local random variables are ground when reached in backward chaining.}

% \begin{example}[Eliminating Probabilistic Facts]
% \label{ex:eliminating_pf}
% Consider the non-ground \dcproblogsty program below, which is range-restricted but not AD-free.
% %containing ground probabilistic fact, ground distributional clauses, and a non-ground logic program. 
% 	\begin{problog*}{linenos}
% 0.2::hot.
% 0.99::cool(1).

% temp(1) ~ normal(27,5):- hot.
% temp(1) ~ normal(20,5):- not hot.

% works(N):- cool(N).
% works(N):- temp(N)<25.0.
% 	\end{problog*}
% We can eliminate the probabilistic facts independently form the rest of the program as this is a local transformation, which does not affect any part of the program but a probabilistic fact itself:
% 	\begin{problog*}{linenos}
% rv_hot ~ flip(0.2).
% hot:- rv_hot=:=1.
% rv_cool(1) ~ flip(0.99).
% cool(1):- rv_cool(1)=:=1.

% temp(1) ~ normal(27,5):- hot.
% temp(1) ~ normal(20,5):- not hot.

% works(N):- cool(N).
% works(N):- temp(N)<25.0.
% 	\end{problog*}
% \end{example}

The high level steps for converting a \dcproblogsty program to a labeled propositional formula closely follow the corresponding conversion for \problogsty programs provided by \citet[Section 5]{fierens2015inference}, \ie, given a \dcproblogsty program \dcpprogram,
		evidence $\evidenceset =e$ and a set of query atoms $\queryset$, the conversion algorithm performs the following steps:
\begin{enumerate}
    \item Determine the { relevant ground program} $\dcpprogram_g$ with respect to the atoms in  $\queryset\cup\evidenceset$ and obtain the corresponing \dfplpsty program. 
    \item Convert $\dcpprogram_g$  to an { equivalent propositional formula} $\phi_g$ and $\evidenceset =e$ to a propositional conjunction $\phi_e$.
%    \item {\bf Assert the evidence} $\evidenceset=e$.
    \item Define the { labeling function} for all atoms in $\phi_g$.
\end{enumerate}
Step 1 exploits the fact that ground clauses that have no influence on the truth values of  query or evidence atoms are irrelevant for inference and can thus be omitted from the ground program. Step 2 performs the conversion from logic program semantics to propositional logic, generating a formula that encodes \emph{all} models of the relevant ground program as well as a formula that serves to assert the evidence by conjoining both formulas. 
Step 3 completes the conversion by defining the labeling function. 
In the following, we discuss the three steps in more detail and prove correctness of our approach (cf. Theorem~\ref{thm:inference-by-expectation}).

%Below we give the outline of the conversion algorithm to map a \dcproblogsty program to a weighted propositional logic formula, this is heavily inspired by~\citet[Section 5]{fierens2015inference},

%	\begin{algo}[\dcproblogsty to Weighted Propositional Formula] \label{algo:inference_dc_problog}
%		The algorithm takes as input a \dcproblogsty program \dcpprogram,
%		evidence $\evidenceset =e$ and a set of \emph{query atoms} $\queryset$.
%		The algorithm maps the queries to $\dcpprogram$ through the following steps to a labeled propositional  formula.
%		\begin{enumerate}
%			\item Ground out $\dcpprogram$ with respect to every query $Q\in \queryset$ and obtain the ground program $\dcpprogram_g$.
 %           Note that it is unnecessary to consider the full grounding of the program, as we only need the part that is relevant to the queries  (given the evidence). That is, the part that captures the probability $P(Q=1 \mid \evidenceset=e)$ for every $Q\in \queryset$. We refer to the resulting program $\dcpprogram_g$ as the {\bf relevant ground program} with respect to $\queryset$ and $\evidenceset=e$.

  %          \item Convert the ground rules in $\dcpprogram_g$ to an {\bf equivalent propositional formula} $\phi_g$.

   %         \item {\bf Assert the evidence} by defining the formula $\phi$ as the conjunction of the formula $\phi_g$ for the rules and for the evidence $\phi_e$. The formula $\phi_e$ is obtained by setting the conjunction of the evidence atoms as the query of interests and perform steps 1. and 2., ie. ground the program \dcpprogram with respect to the conjunction of evidence atoms and obtain a propositional formula $\phi_e$ from that ground program.

    %        \item Define a {\bf labeling function} for all atoms in $\phi$.
	%	\end{enumerate}
	%\end{algo}

%We will now describe the four steps of Algorithm~\ref{algo:inference_dc_problog}  in more detail and prove its correctness by proving that each of the four steps does not alter the probability of the queried atoms. We will limit ourselves to a single query atom for the sake of exposition. Note, however, that the discussion below also holds for multiple queries by realizing that $\dcpprogram_g= \bigcup_{Q\in \queryset} \dcpprogram_g^Q$, where $\dcpprogram_g^Q$ is the ground program with respect to a single query atom $Q$.

\subsection{The Relevant Ground Program}

The first step in the conversion of a non-ground \dcproblogsty program to a labeled  Boolean formula consists of grounding the program with respect to a query set $\queryset$ and the evidence $\evidenceset=e$.
For each ground atom in $\queryset$ and $\evidenceset$ we construct its dependency set. That is, we collect the set of ground atoms and ground rules that occur in any of the proofs of an atom in $\queryset\cup \evidenceset$. The union of all dependency sets for all the ground atoms in $\queryset\cup\evidenceset$ is the dependency set of the \dcproblogsty with respect to the sets $\queryset$ and $\evidenceset$. This dependency set, consisting of ground rules and ground atoms, is called the relevant ground program (with respect to a set of queries and evidence).

\begin{example}
	\label{ex:grounded_nicely}
	Consider the non-ground (AD-free) \dcproblogsty program below.     
	\begin{problog*}{linenos}
rv_hot ~ flip(0.2).
hot:- rv_hot=:=1.
rv_cool(1) ~ flip(0.99).
cool(1):- rv_cool(1)=:=1.

temp(1) ~ normal(27,5):- hot.
temp(1) ~ normal(20,5):- not hot.

works(N):- cool(N).
works(N):- temp(N)<25.0.
	\end{problog*}
If we ground it with respect to the query \probloginline{works(1)} and subsequently apply the rewrite rules from Section~\ref{sec:eliminate_dc} we obtain:
	\begin{problog*}{linenos}
rv_hot ~ flip(0.2). 
hot:- rv_hot=:=1.
rv_cool(1) ~ flip(0.99).
cool(1):- rv_cool(1)=:=1.

temp(hot) ~ normal(27,5).
temp(not_hot) ~ normal(20,5).

works(1):- cool(1).
works(1):- hot, temp(hot)<25.0,
works(1):- not hot, temp(not_hot)<25.0.
	\end{problog*}
\end{example}

A possible way, as hinted at in Example~\ref{ex:grounded_nicely}, of obtaining a ground \dfplpsty program from a non-ground \dcproblogsty program is to first ground out all the logical variables. Subsequently, one can apply Definition~\ref{def:elim-ad} to eliminate annotated disjunctions and probabilistic facts, and Definition~\ref{def:adfree-to-core} in order to obtain a \dfplpsty program with no distributional clauses.
A possible drawback of such a two-step approach (grounding logical variables followed by obtaining a \dcproblogsty program) is that it might introduce spurious atoms to the relevant ground program. A more elegant but also more challenging approach is to interleave the grounding of logical variables and distributional clause elimination. We leave this for future research.

\begin{restatable}[Label Equivalence]{theorem}{theorgp}
	\label{theo:rgp}
	Let $\dcpprogram$ be a \dcproblogsty program and let $\dcpprogram_g$ be the relevant ground program for $\dcpprogram$ with respect to a query $\mu$ and the evidence $\evidenceset=e$ obtained by first grounding out logical variables and subsequently applying transformation rules from Section~\ref{sec:dcproblog}. The programs $\dcpprogram$ and $\dcpprogram_g$ specify the same probability:
	\begin{align}
		P_{\dcpprogram}(\mu=\top \mid \evidenceset=e)
		=
		P_{\dcpprogram_g}(\mu=\top \mid \evidenceset=e)
	\end{align}
    \end{restatable}

    \begin{proof}
        See Appendix~\ref{app:proof:rgp}.
        \end{proof}

\subsection{The Boolean Formula for the Relevant Ground Program}
\label{sec:relprog2boolfrom}

Converting a ground logic program, \ie a set of ground rules, into an equivalent Boolean formula is a purely logical problem and well-studied in the non-probabilistic logic programming literature. We refer the reader to \citet{janhunen2004representing} for an account of the transformation to Boolean formula in the non-probabilistic setting and to \citet{mantadelis2010dedicated} and \citet{fierens2015inference} in the probabilistic setting,
including correctness proofs. We will only illustrate the most basic case with an example here.

\begin{example}[Mapping \dcproblogsty to Boolean Formula] \label{example:dc2bool}
Consider the ground program in Example~\ref{ex:grounded_nicely}. 
To highlight the move from logic programming to propositional logic,	
	%We map  the logic program to a Boolean logic formula $\support_g$, where 
	we introduce for every atom \probloginline{a} in the program a %fresh and equivalent
	corresponding propositional variable  $\phi_\text{\probloginline{a}}$.
	As the program does not contain cycles, we can use Clark's completion for the transformation, \ie, a derived atom is true if and only if the disjunction of the bodies of its defining rules is true. The propositional formula $\phi_g$ corresponding to the program is then the conjunction of the following three formulas:
	\begin{align*}
	    \phi_\mathprobloginline{works(1)} &\leftrightarrow
		\Big( \phi_{\text{\probloginline{cool(1)}}} 
		\lor \phi_\text{\probloginline{hot}} \land  \phi_\text{\probloginline{temp(hot)<25.0}}  
		\lor\neg \phi_\text{\probloginline{hot}}  \land \phi_\text{\probloginline{temp(not_hot)<25.0}}   \Big) \\
		  \phi_{\text{\probloginline{cool(1)}}} &\leftrightarrow  \phi_{\text{\probloginline{rv_cool(1)=:=1}}}\\
		\phi_\text{\probloginline{hot}} &\leftrightarrow \phi_{\text{\probloginline{rv_hot=:=1}}}
	\end{align*}
% AK: 	I'm not quite sure what was going on in the rest of the complicated formula, so I couldn't fix it to fit the fix above, but it didn't seem relevant to the story, so I dropped it (and adapted subsequent examples)
% 	\begin{align}
% 		\support_q
% 		&\leftrightarrow \phi_{\text{\probloginline{cool(1)}}} \land \phi_{\text{\probloginline{rv_cool(1)=:=1}}}  \lor \nonumber \\
% 		&\phantom{{}\leftrightarrow{}}   \phi_\text{\probloginline{hot}} \land  \phi_\text{\probloginline{temp(hot)<25.0}} \land \phi_{\text{\probloginline{hot}=:=1}} \lor \nonumber \\
% 		&\phantom{{}\leftrightarrow{}} \phi_\text{\probloginline{machine(1)}} \land \phi_\text{\probloginline{temp(not_hot)<25.0}} \land  \neg \phi_\text{\probloginline{hot=:=1}} 
% % 		&\leftrightarrow (\text{\dcplpinline{cool(1)}}=1) \lor \nonumber \\
% % 		&\phantom{{}\leftrightarrow{}} (\text{\dcplpinline{temp(hot)}}<25.0) \land (\text{\dcplpinline{hot}}=1) \lor \nonumber \\
% % 		&\phantom{{}\leftrightarrow{}} (\text{\dcplpinline{temp(not_hot)}}<25.0) \land  (\text{\dcplpinline{hot}}=0)  
% 	\end{align}
%	\begin{align}
%		&\phantom{{}\leftrightarrow{}} \support_g  \nonumber \\
%		&\leftrightarrow
%		\Big( \phi_{\text{\probloginline{cool(1)}}} 
%		\lor \phi_\text{\probloginline{hot}} \land  \phi_\text{\probloginline{temp(hot)<25.0}}  
%		\lor\neg \phi_\text{\probloginline{hot}}  \land \phi_\text{\probloginline{temp(not_hot)<25.0}}   \Big) \land  \nonumber \\
%		&\phantom{{}\leftrightarrow{}} 
%		  \phi_{\text{\probloginline{cool(1)}}} \leftrightarrow  \phi_{\text{\probloginline{rv_cool(1)=:=1}}} \land \nonumber \\
%		&\phantom{{}\leftrightarrow{}}
%		  \phi_\text{\probloginline{hot}} \leftrightarrow \phi_{\text{\probloginline{rv_hot=:=1}}} \\
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    	&\leftrightarrow
%    	 \left( \phi_{\text{\probloginline{cool(1)}}}  \land \phi_{\text{\probloginline{rv_cool(1)=:=1}}} \land  \phi_\text{\probloginline{hot}} \land \phi_{\text{\probloginline{rv_hot=:=1}}} \right) \lor \nonumber \\
%		&\phantom{{}\leftrightarrow{}}
 %   	\left( \phi_{\text{\probloginline{cool(1)}}}  \land \phi_{\text{\probloginline{rv_cool(1)=:=1}}} \land  \neg \phi_\text{\probloginline{hot}} \land \neg \phi_{\text{\probloginline{rv_hot=:=1}}} \right) \lor \nonumber \\
  %      &\phantom{{}\leftrightarrow{}} 
%		\left( \phi_\text{\probloginline{hot}} \land  \phi_{\text{\probloginline{rv_hot=:=1}}} \land  \phi_\text{\probloginline{temp(hot)<25.0}} \land \phi_{\text{\probloginline{cool(1)}}}  \land \phi_{\text{\probloginline{rv_cool(1)=:=1}}}  \right) \lor \nonumber \\
 %       &\phantom{{}\leftrightarrow{}} 
%		\left( \phi_\text{\probloginline{hot}} \land  \phi_{\text{\probloginline{rv_hot=:=1}}} \land  \phi_\text{\probloginline{temp(hot)<25.0}} \land \neg \phi_{\text{\probloginline{cool(1)}}}  \land \neg \phi_{\text{\probloginline{rv_cool(1)=:=1}}}  \right) \lor \nonumber \\
 %       &\phantom{{}\leftrightarrow{}} 
%		\left( \neg \phi_\text{\probloginline{hot}} \land \neg \phi_{\text{\probloginline{rv_hot=:=1}}} \land  \phi_\text{\probloginline{temp(not_hot)<25.0}} \land \phi_{\text{\probloginline{cool(1)}}}  \land
%		\phi_{\text{\probloginline{rv_cool(1)=:=1}}}  \right) \lor \nonumber \\
 %       &\phantom{{}\leftrightarrow{}} 
%		\left( \neg \phi_\text{\probloginline{hot}} \land \neg \phi_{\text{\probloginline{rv_hot=:=1}}} \land  \phi_\text{\probloginline{temp(not_hot)<25.0}} \land \neg \phi_{\text{\probloginline{cool(1)}}}  \land
%		\neg \phi_{\text{\probloginline{rv_cool(1)=:=1}}}  \right) \label{eq:example_dc2bool_last}
    % 	
%	\end{align}

\end{example}

Note that the formula obtained by converting the relevant ground program still admits \emph{any} model of that program, including ones that are inconsistent with the evidence. In order to use that formula to compute conditional probabilities, we still need to assert  the evidence into the formula by conjoining the corresponding propositional literals. 
The following theorem then directly applies to our case as well.
\begin{theorem}[Model Equivalence~\citep{fierens2015inference} (Theorem 2, part 1)]
\label{theo:model_equivalence}
Let $\dcpprogram_g$ be the relevant ground program for a \dcproblogsty program $\dcpprogram$ with respect to  query set \queryset\ and  evidence $\evidenceset=e$. Let $MOD_{\evidenceset=e}(\dcpprogram_g)$ be those models in $MOD(\dcpprogram_g)$ that are consistent with the evidence.
Let $\phi_g$ denote the  propositional formula derived from $\dcpprogram_g$, and set $\phi \leftrightarrow \phi_g \land \phi_e$, where  $\phi_e$ is the conjunction of literals that corresponds to the observed truth values of the atoms in $\evidenceset$. We then have {\bf model equivalence}, \ie, 
\begin{align}
    MOD_{\evidenceset=e}(\dcpprogram_g)
    =
    ENUM(\phi)
\end{align}
where $ENUM(\phi)$ denotes the set of models of $\phi$.
\end{theorem}

\subsection{Obtaining a Labeled Boolean Formula}

In contrast to a \problogsty program, a \dcproblogsty program does not explicitly provide independent probability labels for the basic facts in the distribution semantics, and we thus need to suitably adapt the last step of the conversion. 
We will first define the labeling function on propositional atoms and will then show that the probability of the label of a propositional formula is the same as the probability of the relevant ground program under the measure semantics from Section~\ref{sec:semantics}. We call this {\em label equivalence} and prove it in Theorem~\ref{theo:label_equivalence}.

\begin{definition}[Label of Literal] \label{def:labeling_function}
The label $\alpha(\phi_{\rho})$ of a propositional atom $\phi_{\rho}$ (or its negation) is given by:
\begin{align}
    \alpha( \phi_{\rho })=
    \begin{cases} 
    \ive{c(vars(\rho)},  &  \text{if $\rho$ is a  comparison atom} \\
    1, & \text{otherwise}
% 
    \end{cases}
\end{align}
and for the negated atom:
\begin{align}
    \alpha(\neg \phi_{\rho})=
    \begin{cases} 
    \ive{\neg c(vars(\rho))},  &  \text{if $\rho$ is a  comparison atom} \\
    1, & \text{otherwise}
% 
    \end{cases}
\end{align}
We use Iverson brackets $\ive{\cdot}$~\citep{iverson1962programming} to denote an indicator function. Furthermore, $vars(\rho)$ denotes the random variables that are present in the arguments of the atom $\rho$ and $c(\cdot)$ encodes the constraint given by $\rho$.
\end{definition}

\begin{example}[Labeling function]
Continuing Example~\ref{example:dc2bool}, we obtain, inter alia, the following labels:
\begin{align*}
    & \alpha(\phi_\text{\probloginline{rv_hot=:=1}} ) = \ive{rv\_hot=1}  \\
    & \alpha(\neg \phi_\text{\probloginline{rv_hot=:=1}} ) = \ive{\neg(rv\_hot=1)} = \ive{rv\_hot=0} \\
    & \alpha(\phi_\text{\probloginline{hot}} ) = 1\\  
    & \alpha(\neg \phi_\text{\probloginline{hot}} ) = 1 
\end{align*}
\end{example}

\begin{definition}[Label of Boolean Formula]
\label{def:label_bool_formula}
Let $\phi$ be a Boolean formula and $\alpha(\cdot)$ the labeling function for the variables in $\phi$ as given by Definition~\ref{def:labeling_function}. We define the label of $\phi$ as
\begin{align*}
    \alpha(\phi) &= \sum_{\varphi\in ENUM(\phi)}\prod_{\ell\in\varphi}\alpha(\ell) 
\end{align*}
i.e. as the sum of the labels of all its models, which are in turn defined as the product of the labels of their literals.
\end{definition}

\begin{example}[Labeled Boolean Formula]
The label of the conjunction 
$$\neg \phi_\text{\probloginline{hot}} \land \neg \phi_{\text{\probloginline{rv_hot=:=1}}} \land  \phi_\text{\probloginline{temp(not_hot)<25.0}} \land \neg \phi_{\text{\probloginline{cool(1)}}}  \land
		\neg \phi_{\text{\probloginline{rv_cool(1)=:=1}}} \land \phi_{\text{\probloginline{works(1)}}}$$
which describes one model of the example formula, is computed as follows:
\begin{align*}
% 
    &\phantom{{}={}}
    \alpha( \neg \phi_\text{\probloginline{hot}}
    \land
    \neg \phi_{\text{\probloginline{rv_hot=:=1}}}
    \land
    \phi_\text{\probloginline{temp(not_hot)<25.0}}\\
    &\phantom{{}={}}\land
    \neg \phi_{\text{\probloginline{cool(1)}}}
    \land
	\neg \phi_{\text{\probloginline{rv_cool(1)=:=1}}}
	\land \phi_{\text{\probloginline{works(1)}}} )  \nonumber \\
	&=
    \alpha( \neg \phi_\text{\probloginline{hot}})
    \times \alpha( \neg \phi_{\text{\probloginline{rv_hot=:=1}}})
    \times  \alpha(\phi_\text{\probloginline{temp(not_hot)<25.0}}) \nonumber \\
    &\phantom{{}={}}
    \times \alpha(\neg \phi_{\text{\probloginline{cool(1)}}})
    \times \alpha(\neg \phi_{\text{\probloginline{rv_cool(1)=:=1}}} \times \alpha(\phi_{\text{\probloginline{works(1)}}}) )  \nonumber \\
    &= 1 \times \ive{rv\_hot=0} \times \ive{temp(not\_hot)<25} \times 1 \times \ive{rv\_cool(1)=0}\times 1\nonumber \\
    &= \ive{rv\_hot=0} \times \ive{temp(not\_hot)<25} \times \ive{rv\_cool(1)=0} 
\end{align*}

\end{example}

\begin{restatable}[Label Equivalence]{theorem}{Labelequivalence}
\label{theo:label_equivalence}

Let $\dcpprogram_g$ be the relevant ground program for a \dcproblogsty program $\dcpprogram$ with respect to a query $\mu$ and the evidence $\evidenceset=e$. Let $\phi_g$ denote the  propositional formula derived from $\dcpprogram_g$ and let $\alpha$ be the labeling function as defined in Definition~\ref{def:labeling_function}. We then have {\bf label equivalence}, \ie
\begin{align}
    \forall \varphi \in ENUM(\phi_g):  \E_{\randomvariableset \sim  \dcpprogram_g} [\alpha( \varphi )] = P_{\dcpprogram_g}(\varphi)
\end{align}
In other words, for all models $\varphi$ of $\phi_g$, the expected value ($\E_\cdot [\cdot]$) of the label of $\varphi$ is equal to the probability of $\varphi$ according to the probability measure of relevant ground program $\dcpprogram_g$.
\end{restatable}

\begin{proof}
See Appendix~\ref{app:proof:label_equivalence}.
\end{proof}

Theorem~\ref{theo:label_equivalence} states that we can reduce inference in hybrid probabilistic logic programs to computing the expected value of  labeled Boolean formulas, as summarized in the following theorem.
\begin{theorem}
\label{thm:inference-by-expectation}
Given a \dcproblogsty program \dcpprogram, a set \queryset\ of queries, and evidence $\evidenceset = e$, for every $\mu\in\queryset$, we obtain the conditional probability of $\mu = q$ ($q\in \{\bot,\top \}$) given $\evidenceset = e$ as
\begin{align*}
    P(\mu=q\mid\evidenceset = e) = \frac{\E_{vars(\phi) \sim  \dcpprogram_g} [\alpha( \phi \wedge \phi_q)] }{\E_{vars(\phi) \sim  \dcpprogram_g} [\alpha( \phi )] }
\end{align*}
where $\phi$ is the formula encoding the relevant ground program $\dcpprogram_g$ with  the evidence asserted  (cf.~Theorem~\ref{theo:model_equivalence}), and $\phi_q$ the propositional atom for $\mu$.
\end{theorem}
\begin{proof}
This directly follows from model and label equivalence together with the definition of conditional probabilities. 
\end{proof}

We have shown that the probability of a query to a \dcproblogsty program can be expressed as the expected label of a propositional logic formula.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_main/dcproblog.tex
aij dcproblog paper

\section{\dcproblogsty} \label{sec:dcproblog}

\fixed{
While the previous section has focused on the core elements of the \dcproblogsty language, we now introduce syntactic sugar to ease modeling.
We  consider three kinds of modeling patterns in \dfplpsty, and introduce a more compact notation for each of them.
Here we focus on examples and intuitions and relegate the more technical details to Appendix~\ref{sec:detaileddcproblog}.
Specifically, we discuss the precise semantics of the syntactic sugar introduced below and how to map it onto \dfplpsty language constructs (Appendix~\ref{sec:semantics_syntactic_sugar} and~\ref{sec:dcvalidity}).
In Appendix~\ref{sec:Additionalsyntacticsugar} we introduce additional syntactic sugar constructs for user-defined sample spaces and multivariate distributions, and in Appendix~\ref{sec:beyondmixtures} we study the intricacies introduced by negation when using syntactic sugar. 
}

\subsection{Boolean Random Variables}

The first modelling pattern concerns Boolean random variables, which we already encountered in Section~\ref{sec:panorama_semantics} as probabilistic facts (in \dcproblogsty) or as a combination of a Bernoulli random variable, a comparison atom, and a logic rule (in \dfplpsty). Below we give a more concise example.

\begin{example}
    We model, in \dfplpsty, an alarm that goes off for different reasons.
    \begin{problog*}{linenos}
issue1 ~ flip(0.1).
issue2 ~ flip(0.6).
issue3 ~ flip(0.3).

alarm :- issue1=:=1, not issue2=:=1.
alarm :- issue3=:=1, issue1=:=0.
alarm :- issue2=:=1.
    \end{problog*}
\end{example}
To make such programs more readable, we borrow the well-known notion of \emph{probabilistic fact} from discrete PLP, which directly introduces a logical atom as alias for the comparison of a random variable with the value \probloginline{1}, together with the probability of that value being taken.

\begin{definition}[Probabilistic Fact]
    A probabilistic fact is of the form $p\prob\mu$, where  $p$ is an arithmetic term that evaluates to a real number in the interval $[0,1]$ and $\mu$ is a regular ground atom.
\end{definition}
\begin{example}
    We use probabilistic facts to rewrite the previous example.
    \begin{problog*}{linenos}
0.1::problem1.
0.6::problem2.
0.3::problem3.

alarm :- problem1, not problem2.
alarm :- problem3, not problem1.
alarm :- problem2.
    \end{problog*}
\end{example}

\subsection{Probabilistically Selected Logical Consequences}
The second pattern concerns situations where a random variable with a finite domain is used to model a choice between several logical consequences:
\begin{example}
    We use a random variable to model a choice between whom to call upon hearing the alarm.
    \begin{problog*}{linenos}
call ~ finite([0.6:1,0.2:2,0.1:3]).
alarm.
call(mary) :- call=:=1, alarm.
call(john) :- call=:=2, alarm.
call(police) :- call=:=3, alarm.
    \end{problog*}
\end{example}
To more compactly specify such situations, we borrow the concept of an \emph{annotated disjunction} from discrete PLP~\citep{vennekens2004logic}.

\begin{definition}[Annotated Disjunction]
    An annotated disjunction (AD) is a rule of the form
    $$
        p_1\prob \mu_1; \dots; p_n \prob \mu_n \lpif \beta.
    $$
    where the  $p_i$'s are arithmetic terms each evaluating to a number in $[0,1]$ with a total sum of at most $1$. The $\mu_i$'s are regular ground atoms, and $\beta$ is a (possibly empty) conjunction of literals.
\end{definition}
The informal meaning of such an AD is "if $\beta$ is true, it probabilistically causes one of the $\mu_i$ (or none of them, if the probabilities sum to less than one) to be true as well".

\begin{example}
    We now use an AD for the previous example.
    \begin{problog}
alarm.
0.6::call(mary); 0.2::call(john); 0.1::call(police) :- alarm.
    \end{problog}
\end{example}

It is worth noting that the same head atom may appear in multiple ADs, whose bodies may be non-exclusive, \ie, be true at the same time. That is, while a single AD \emph{can} be used to model a multi-valued random variable, \emph{not all} ADs  encode such variables.
\begin{example}
    The following program models the effect of two kids throwing stones at a window.
    \begin{problog}
0.5::throws(suzy).
throws(billy).

0.8::effect(broken); 0.2::effect(none) :- throws(suzy).
0.6::effect(broken); 0.4::effect(none) :- throws(billy).
    \end{problog}
    Here, we have $P(\text{\probloginline{effect(broken)}})=0.76$ and $P(\text{\probloginline{effect(none)}})=0.46$, as there are worlds where both \probloginline{effect(broken)} and \probloginline{effect(none)} hold. The two ADs do hence not encode a categorical distribution.
    This is explicit in the \dfplpsty program, which contains two  random variables (\probloginline{x1} and \probloginline{x2}):
    \begin{problog}
x0 ~ flip(0.5).
throws(suzy) :- x0=:=1.
throws(billy).

x1 ~ finite([0.8:1,0.2:2]).
effect(broken) :- x1=:=1, throws(suzy).
effect(none) :- x1=:=2, throws(suzy).
x2 ~ finite([0.6:1,0.4:2]).
effect(broken) :- x2=:=1, throws(billy).
effect(none) :- x2=:=2, throws(billy).
    \end{problog}
    % From a modeling perspective, it is good practice to prefer explicit random variables over ADs when encoding multi-valued random variables, \eg, 
    % \begin{problog}
    % 0.5::throws(suzy).
    % throws(billy).

    % hit(suzy) ~ flip(0.8).
    % hit(billy) ~ flip(0.6).

    % effect(broken) :- throws(X), hit(X)=:=1.
    % effect(none) :- not effect(broken).
    % \end{problog}
\end{example}

\subsection{Context-Dependent Distributions}\label{sec:sugar-dc}
The third pattern is concerned with situations where the same conclusion is based on random variables with different distributions depending on specific properties of the situation, as illustrated by the following example.
\begin{example}
    We use two separate random variables to model that whether a machine works depends on the temperature being below or above a threshold. The temperature follows different distributions based on whether it is a hot day or not, but the threshold is independent of the type of day.
    \begin{problog*}{linenos}
0.2::hot.

temp_hot ~ normal(27,5).
temp_not_hot ~ normal(20,5).

works :- hot, temp_hot<25.0.
works :- not hot, temp_not_hot<25.0.
    \end{problog*}
\end{example}
To more compactly specify such situations, we borrow the syntax of \emph{distributional clauses} from the DC  language~\citep{gutmann2011magic}, which we already encountered in Section~\ref{sec:panorama_semantics}.

\begin{definition}[Distributional Clause]
    A distributional clause (DC) is a rule of the form
    $$
        \tau \sim \delta \lpif \beta.
    $$
    where $\tau$ is a regular ground expression, the functor of $\delta$ is in $\distributionfunctors$, and $\beta$ is a conjunction of literals.
\end{definition}

We call the left-hand side of the \mathpredicate{\sim}{2} prediate in a distributional clause a {\em random term} and the right-hand side a {\em distribution term}. Note that random terms cannot always be interpreted as random variables, which we discuss now.

The informal meaning of a distributional clause is "if $\beta$ is true, then the random term $\tau$ refers to a random variable that follows a distribution given by the distribution term $\delta$". Here, the distinction between \emph{refers to} a random variable and \emph{is} a random variable becomes crucial, as we will often have several distributional clauses for the same random term. This is also the case in the following example.
\begin{example}
    Using distributional clauses, we can rewrite the previous example with a single random term \probloginline{temp} as
    \begin{problog*}{linenos}
0.2::hot.

temp ~ normal(27,5) :- hot.
temp ~ normal(20,5) :- not hot.

works :- temp < 25.0.
    \end{problog*}
    The idea is that we still have two underlying random variables, one for each distribution, but the logic program uses the same term to refer to both of them depending on the logical context. The actual comparison facts are on the level of these implicit random variables, and \probloginline{temp<0.25} refers to one of them depending on context, just as in the original example.
\end{example}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_main/evaluating.tex
aij dcproblog paper

\section{Computing Expected Labels via Algebraic Model Counting}
\label{sec:alw}

In this section we will adapt the approach taken by~\citet{zuidbergdosmartires2019exact}, dubbed {\em Sampo} to compute the expected value of labeled propositional Boolean formulas.
The method approximates intractable integrals that appear when computing expected labels using Monte Carlo estimation. The main difference between Sampo and our approach, which we dub {\em infinitesimal algebraic likelihood weighting} (IALW) is that IALW can also handle infinitesimally small intervals, which arise when conditioning on zero probability events.

\subsection{Monte Carlo Estimate of Conditional Query}

In Definition~\ref{def:conditional_prob} we defined the conditional probability as:
\begin{align}
    P_\dcpprogram(\mu=\top\mid\evidenceset=e)= \frac{P_{\dcpprogram}(\mu=\top, \evidenceset=e)}{P_\dcpprogram(\evidenceset=e)} 
\end{align}
and we also saw in Definition~\ref{def:conditional_prob_zero_event} that using infinitesimal intervals allows us to consider zero probability events, as well. Computing the probabilities in the numerator and denominator in the equation above is, in general, computationally hard. We resolve this using a Monte Carlo approximation.

\begin{restatable}[Monte Carlo Approximation of a Conditional Query]{proposition}{mcapproxconditional}
\label{prop:mcapproxconditional}
Let the set 
\begin{align}
    \mathcal{S} = \left\{ \left(s_1^{(1)}, \dots, s_M^{(1)} \right), \dots , \left(s_1^{(\lvert \mathcal{S} \rvert)}, \dots, s_M^{(\lvert \mathcal{S} \rvert)} \right)  \right\} \label{eq:rejection_samples}
\end{align}
denote $\lvert \mathcal{S} \rvert$ i.i.d. samples for each random variable in $ \dcpprogram_g$.
A conditional probability query to a  \dcproblogsty program \dcpprogram can be approximated as: 
\begin{align}
P_\dcpprogram(\mu = q \mid \evidenceset = e) 
&\approx  \frac{ \sum_{i=1}^{\lvert \mathcal{S} \rvert}  \sum_{\varphi \in ENUM(\phi \land \phi_q) } \alpha^{(i)}(\varphi) } { \sum_{i=1}^{\lvert \mathcal{S} \rvert} \sum_{\varphi \in ENUM(\phi) } \alpha^{(i)}(\varphi) }, & \quad |\mathcal{S}|<\infty
\end{align}
The index $(i)$ on $\alpha^{(i)}(\varphi)$ indicates that the label of $\varphi$ is evaluated at the $i$-th ordered set of samples $ \left(s_1^{(i)}, \dots, s_M^{(i)} \right)$. 
\end{restatable}

\begin{proof}
    See Appendix~\ref{app:proof:mcapproxconditional}.
\end{proof}

In the limit $\lvert \mathcal{S} \rvert\rightarrow \infty$ this sampling approximation scheme is perfectly valid. However, in practice, with only limited resources available, such a rejection sampling strategy will perform poorly (in the best case) or even give completely erroneous results. After all, the probability of sampling a value from the prior distribution that falls exactly into an infinitesimally small interval given in the evidence tends to zero.
To make the computation of a conditional probability, using Monte Carlo estimates, feasible, we are going to introduce {\em infinitesimal algebraic likelihood weighting}. But first, we will need to introduce the concept of infinitesimal numbers.

\subsection{Infinitesimal Numbers}
Remember that infinitesimal intervals arise in zero probability conditioning events and describe an infinitesimally small interval around a specific observed value, \eg $\nu \in [w-\nicefrac{\Delta w}{2}, w+\nicefrac{\Delta w}{2} ]$ for a continuous random variable $\nu$ that was observed to take the value $w$ (cf. Definition~\ref{def:conditional_prob_zero_event}).
We will describe these infinitesimally small intervals using so-called {\em infinitesimal numbers}, which were first introduced by~\citet{nitti2016probabilistic} and further formalized in~\citet{wu2018discrete}, \citep{zuidberg2020atoms} and~\citep{jacobs2021paradoxes}. The latter work also coined the term {\em `infinitesimal number'} \fixed{and we refer the reader specifically to \citet[Section 5.2]{jacobs2021paradoxes} for an intuitive exposition of infinitesimal numbers.}

\begin{definition}[Infinitesimal Numbers]
    \label{def:inf_number}
An infinitesimal number is a pair \fixed{$(r, n) \in \mathbb{R} \times (\mathbb{N} \cup +\infty )$}, also written as $r\epsilon^n$, and which corresponds to a real number when $n=0$. We denote the set of all infinitesimal numbers by $\mathbb{I}$.
\end{definition}

\begin{definition}[Operations in $\mathbb{I}$]
\label{def:inf_number_ops}
Let $(r, n)$ and $(t,m)$ be two numbers in $\mathbb{I}$. We define the addition and multiplication as binary operators:
\begin{align}
	(r,n) \oplus
	(t,m)
	&\coloneqq
	\begin{cases}
		(r+t,n) &\quad  \text{if $n=m$} \\
		(r,n) &\quad  \text{if $n<m$} \\
		(t,m) &\quad  \text{if $n>m$}
	\end{cases} 
    \label{eq:infininumber_plus}
    \\
	(r,n) \otimes
	(t,m)
	&\coloneqq (r \times t , n+m)  &
    \label{eq:infininumber_times}
\end{align}
The operations $+$ and $\times$ on the right hand side denote the usual addition and multiplication operations for real and integer numbers.
\end{definition}

\begin{definition}[Neutral Elements]
\label{def:inf_number_neutral_elem}
The neutral elements of the addition and multiplications in $\mathbb{I}$ are, respectively, defined as:
\begin{align}
	e^\oplus  \coloneqq (0,+\infty)  &&
	e^\otimes  \coloneqq (1,0)
\end{align}
\end{definition}

Probabilistic inference and generalization thereof can often be cast as performing computations using commutative semirings~\citep{kimmig2017algebraic}. We will follow a similar strategy.

\begin{definition}\label{def:comm_semiring} 
	A {\bf  commutative semiring} is an algebraic structure $(\mathcal{A},\oplus,\otimes,\allowbreak e^{\oplus},e^\otimes)$ equipping a set of elements $\mathcal{A}$ with addition and multiplication such that
	\begin{enumerate}
		\item addition $\oplus$ and multiplication $\otimes$ are binary operations $\mathcal{A}\times \mathcal{A}\rightarrow \mathcal{A}$
		\item addition $\oplus$ and multiplication $\otimes$ are  associative and commutative binary operations over the set $\mathcal{A}$
		\item $\otimes$ distributes over $\oplus$
		\item  $e^\oplus \in \mathcal{A}$ is the neutral element of $\oplus$
		\item  $e^\otimes \in \mathcal{A}$ is the neutral element of $\otimes$
		\item $e^\oplus \in \mathcal{A}$ is an annihilator for $\otimes$
	\end{enumerate}
\end{definition}

\begin{lemma}
The structure $(\mathbb{I}, \oplus, \otimes, e^\oplus , e^\otimes )$ is a commutative semiring.
\end{lemma}
\begin{proof}
This follows trivially from the operations defined in Definition~\ref{def:inf_number_ops} and the neutral elements in Definition~\ref{def:inf_number_neutral_elem}.
\end{proof}
We will also need to perform subtractions and divisions in $\mathbb{I}$,

% \st{
% \fixed{
% \begin{definition}[Inverse Elements]
% Let $(r, n)$ be a number in $\mathbb{I}$. We define its inverse with respect to the addition $-(r,n)$, also called negation, as:
% \begin{align}
%     -(r,n) \coloneqq (-r, n)
% \end{align}
% Moreover, we define its inverse with respect to the multiplication $(r,n)^{-1}$, also called the reciprocal, as:
% \begin{align}
%     (r,n)^{-1}\coloneqq
%     \begin{cases}
%         (r^{-1}, -n) &\quad \text{if $r\neq 0$} \\
%         \text{undefined} &\quad \text{if $r=0$}
%     \end{cases}
% \end{align}
% \end{definition}
%     }
% }

\begin{definition}[Subtraction and Division in $\mathbb{I}$]
    \label{def:subdiv}
    Let $(r, n)$ and $(s,m)$ be two numbers in $\mathbb{I}$. We define the subtraction and division as:
    \begin{align}
            (r,n) \ominus (t,m) &\coloneqq  (r,n) \oplus (-t,m) \\
            (r,n) \oslash (t,m) &\coloneqq 
            \begin{cases}
                \text{undefined} &\quad \text{if $|n|=|m|=\infty$ and $sign(n) \neq sign(m)$} \\
                (\nicefrac{r}{t}, n-m)  &\quad \text{if $t\neq 0$} \\
                \text{undefined} &\quad \text{if $t=0$}
            \end{cases}
    \end{align}

\end{definition}

We would like to note that similar algebraic structures have been used for counting optimal variable assignments in graphical models~\citep{marinescu2019counting} and probabilistic inference in generating circuits~\citep{harviainen2023inference}.

\subsection{Infinitesimal Algebraic Likelihood Weighting}
The idea behind IALW is that we do not sample random variables that fall within an infinitesimal small interval, encoded as a delta interval (cf. Definition~\ref{def:delta_interval}), but that we force, without sampling, the random variable to lie inside the infinitesimal interval. 
To this end, assume again that we have $\lvert \mathcal{S} \rvert$ i.i.d. samples for each random variable. That means that we have again a set of ordered sets of samples:
\begin{align}
    \label{eq:ancestral_samples}
    \mathcal{S} = \left\{ \left(s_1^{(1)}, \dots, s_M^{(1)} \right), \dots , \left(s_1^{(\lvert \mathcal{S} \rvert)}, \dots, s_M^{(\lvert \mathcal{S} \rvert)} \right)  \right\}
\end{align}

This time the samples are drawn with the infinitesimal delta intervals taken into account. For example, assume we have a random variable $\nu_1$ distributed according to a normal distribution $\mathcal{N}(5,2)$ and we have an atom \probloginline{delta_interval(@$\nu_1$@,4)} in the propositional formula $\phi$. Each sampled value of $s_1^{(i)}$ will then equal $4$ ( $1\leq i\leq \lvert \mathcal{S} \rvert$). Furthermore, when sampling, we sample the parents of a random variable prior to sampling the random variable itself. For instance, take the random variable $\nu_2\sim \mathcal{N}(\nu_3=w,2)$, where $\nu_3$ is itself a random variable. We first sample $\nu_3$ and once we have a value for $\nu_3$ we plug that into the distribution for $\nu_2$, which we sample subsequently. In other words, we sample according to the ancestor relationship between the random variables.
We call the ordered set of samples $\varset{s}^{(i)} \in \mathcal{S}$ an {\em ancestral sample}.

\begin{definition}[IALW Label] \label{def:sample_labeling_function}
    \fixed{
Let $\delta_k$ denote the probability distribution of a random variable $\nu_k$.
Given an ancestral sample $\varset{s}^{(i)}= (s_1^{(i)}, \dots,  s_M^{(i)} ) $ for the random variables $\randomvariableset = (\nu_1,\dots, \nu_M)$, we denote by $\delta_k( \varset{s}^{(i)} )$ the evaluation of the density $\delta_k$ at $\varset{s}^{(i)}$, where $i$ specifies the $i$-th sample.
}
The IALW label of a positive literal $\ell$ is an infinitesimal number given by:
\begin{align}
    &\alpha_{IALW}^{(i)}( \ell) \nonumber \\
    &=\begin{cases}
    (\delta_k(\varset{s}^{(i)}) , 1),  &  \text{if $\ell$ is a \probloginline{delta_interval} whose first argument } \\
    & \text{is a continuous random variable} \\
    % (p_k(\varset{s}^{(i)}) , 0),  &  \text{if $\ell$ is a \probloginline{delta_interval} having as first argument the random} \\
                        % & \text{variable $\nu_k$, which is countable} \\
    ( \ive{ c_\ell(\varset{s}^{(i)}) }, 0), & \text{if $\ell$ is any comparison atom}
    \\
    (1, 0), & \text{otherwise} 
    \end{cases}
    \nonumber
\end{align}
The expression $\ive{ c_\ell(\varset{s}^{(i)}) }$ denotes the indicator function on the constraint that corresponds to the literal $\ell$ and which is evaluated using the samples $\varset{s}^{(i)}$.

For the negated literals we have the following labeling function:
\begin{align}
    &\alpha_{IALW}^{(i)}( \neg \ell) \nonumber \\
    &=\begin{cases}
    (1 , 0),  &  \text{if $\ell$ is a \probloginline{delta_interval} whose first argument } \\
                        & \text{is a continuous random variable} \\
    % (1{-}w_k(\varset{s}^{(i)}) , 0),  &  \text{if $\ell$ is a \probloginline{delta_interval} having as first argument the random} \\
                        % & \text{variable $X_k$, which is countable} \\
    (1{-}\ive{ c_\ell(\varset{s}^{(i)}) }, 0), & \text{if $\ell$ is any other comparison atom}
    \\
    (1, 0), & \text{otherwise} 
    \end{cases}
    \nonumber
\end{align}
\end{definition}

Intuitively speaking and in the context of probabilistic inference,
the first part of an infinitesimal number accumulates (unnormalized) likelihood weights, while the second part counts the number of times we encounter a \probloginline{delta_interval} atom. This counting happens with $\oplus$ operation of the infinitesimal numbers (Equation~\ref{eq:infininumber_plus}). The $\oplus$ operation tells us that for two infinitesimal numbers $(r,n)$ and $(t,m)$ with $n<m$, the event corresponding to the first of the two infinitesimal numbers is infinitely more probable to happen and that we drop the likelihood weight of the second infinitesimal number (Equation~\ref{eq:infininumber_plus}). 
In other words, an event with fewer \probloginline{delta_interval}-atoms is infinitely more probable than an event with more such intervals.

\begin{example}[IALW Label of \probloginline{delta_interval} with Continuous Random Variable]
Let us consider a random variable \probloginline{x}, which is normally distributed: $p(\mathprobloginline{x}|\mu, \sigma)=\nicefrac{1}{(\sigma \sqrt{2 \pi})} \exp \left( - \nicefrac{(\mathprobloginline{x}-\mu)^2}{2 \sigma^2 } \right) $), 
where $\mu$ and $\sigma>0$ are real valued parameters that we can choose freely.
The atom \probloginline{delta_interval(x,3)} gets the label
$$
\left( \frac{1}{(\sigma \sqrt{2 \pi})} \exp \left( - \nicefrac{(\mathprobloginline{3}-\mu)^2}{2 \sigma^2 } \right)  , 1\right)
$$
The first element of the infinitesimal number is the probability distribution evaluated at the observation, in this case \probloginline{3}. As this is a zero probability event, the label also picks up a non-zero second element.

The label of $\neg \mathprobloginline{delta_interval(x,3)}$ is $(1,0)$. The intuition here being that the complement of an event with zero probability of happening will happen with probability $1$. As the complement event is not a zero probability event the second element of the label is $0$ instead of $1$. 
\end{example}

\begin{example}[IALW Label of \probloginline{delta_interval} with Discrete Random Variable]
    Let us consider a discrete random variable \probloginline{k}, which is Poisson distributed: 
    $$
    p(\mathprobloginline{k}|\lambda)=\nicefrac{\lambda^\mathprobloginline{k} e^{-\lambda}}{\mathprobloginline{k}!}
    $$
    where $\lambda>0$ is a real-valued parameter that we can freely choose.

    As a \probloginline{delta_interval} with a discrete random variable is equivalent to a \probloginline{=:=} comparison (\cf Definition~\ref{def:delta_interval}), we get for the label of the atom \probloginline{delta_interval(k,3)}:
    $(\ive{s_x^{(i)}=3}, 0)$, where $s_\mathprobloginline{k}^{(i)}$ is the $i$-th sample for \probloginline{k}.  
    % The atom \probloginline{delta_interval(x,3)} gets the label $( \nicefrac{\lambda^3 e^{-\lambda}}{3!}  , 0)$. The first element of the infinitesimal number is the probability of \probloginline{x} taking the the value 3. Because this is a non-zero probability event, the label does not pick up a non-zero second element. The label of $\neg \text{\probloginline{delta_interval(x,3)}}$ is now simply $(1-  \nicefrac{\lambda^3 e^{-\lambda}}{3!}  , 0)$.
\end{example}

% \begin{align}
%     P\Bigl(\text{\probloginline{american}}{=}\top \mid (\text{\probloginline{gpa(student)}}\doteq4)=\top \Bigr)&=1 
%     \nonumber \\
%     P \Bigl(\text{\probloginline{indian}}{=}\top \mid (\text{\probloginline{gpa(student)}}\doteq4)=\top \Bigr)&=0 \nonumber
%     \end{align}

\begin{definition}[Infinitesimal Algebraic Likelihood Weighting]
\label{def:alw}
Let $\mathcal{S}$ be a set of ancestral samples and let $ DI(\varphi)$ denote the subset of literals in $\varphi$ that are delta intervals. We then define IALW as expressing the expected value of the label of a propositional formula (given a set of ancestral samples) in terms of a fraction of two infinitesimal numbers:
\begin{align}
    \left( \E \left[ \sum_{\varphi\in ENUM(\phi)} \prod_{\ell \in \varphi}  \alpha \left(\ell \right) \bigg| \mathcal{S} \right] ,0 \right)
    \approx
    \frac
    {\displaystyle \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}  \bigoplus_{\varphi\in ENUM(\phi)} \bigotimes_{\ell \in \varphi}  \alpha_{IALW}^{(i)} \left(\ell \right)}
    {\displaystyle \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert} \bigoplus_{\varphi\in ENUM(\phi)} \bigotimes_{\ell \in  DI(\varphi)}  \alpha_{IALW}^{(i)} \left(\ell \right)  } \label{eq:ALW}
\end{align}
The left hand side expresses the expected value as an infinitesimal number.
\end{definition}

% \begin{proposition}[Consistency of ALW]
\begin{restatable}[Consistency of IALW]{proposition}{alwconsistency}
\label{prop:alw_consistency}
Infinitesimal algebraic likelihood weighting is consistent, that is, the approximate equality in Equation~\ref{eq:ALW} is almost surely an equality for $\lvert \mathcal{S} \rvert \rightarrow \infty$.
% \begin{align}
%     \left( \E \left[ \sum_{\omega\in ENUM(\phi)} \prod_{\ell \in \omega}  \alpha \left(\ell \right) \right] ,0 \right)
%     =
%     \frac{\displaystyle \sum_{i=1}^N  \sum_{\omega\in ENUM(\phi)} \prod_{\ell \in \omega}  \alpha^{(i)} \left(\ell \right)}
%     {\displaystyle \sum_{i=1}^N  \sum_{\omega\in ENUM(\phi)} \prod_{\ell \in  DI(\omega)}  \alpha^{(i)} \left(\ell \right)  } 
% \end{align}
% holds almost surely for $N\rightarrow \infty$.
\end{restatable}
\begin{proof}
    See Appendix~\ref{app:proof:alw_consistency}.
\end{proof}

Likelihood weighting, the core idea behind IALW, is a well known technique for inference in Bayesian networks~\citep{fung1990weighing} and probabilistic programming~\citep{milch2005approximate,nitti2016probabilistic}, and falls within the broader class of self-normalized importance sampling~\citep{kahn1950random,kloek1978bayesian,casella1998post}.
Just like IALW, the inference approaches proposed by \citet{nitti2016probabilistic}, \citet{wu2018discrete}, and \citet{jacobs2021paradoxes}  generalize the idea of likelihood weighting to the setting with infinitesimally small intervals. What sets IALW apart from these methods is its semiring formulation. The semiring formulation will allow us to seamlessly combine IALW with knowledge compilation~\citep{darwiche2002knowledge}, a technique underlying state-of-the art probabilistic inference algorithms in the discrete setting. We examine this next.

Having proven the consistency of IALW, we can now express the probability of a conditional query to a \dcproblogsty program in terms of semiring operations for infinitesimal numbers $\mathbb{I}$.

\begin{restatable}{proposition}{alwapproximation}
\label{prop:alwapproximation}
% \begin{proposition}
A conditional probability query to a  \dcproblogsty program \dcpprogram can be approximated as: 
    \begin{align}
    P_\dcpprogram(\mu=q|\evidenceset=e) 
    \approx    
    \frac
    { \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}  \bigoplus_{\varphi\in ENUM(\phi \land \phi_{q})} \bigotimes_{\ell \in \varphi}  \alpha_{IALW}^{(i)} \left(\ell \right)}
    { \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}  \bigoplus_{\varphi\in ENUM(\phi)} \bigotimes_{\ell \in \varphi}  \alpha_{IALW}^{(i)} \left(\ell \right)}  
    \label{eq:prop:alwapproximation}
    \end{align}

% \end{proposition}
\end{restatable}

\begin{proof}
    See Appendix~\ref{app:proof:alwapproximation}.
\end{proof}

% In Example~\ref{ex:indian_gpa} we showed how to write down the Indian GPA problem in \dcproblogsty and also agave the probabilities for the conditional queries. After having introduced infinitesimal numbers we can now also explicitly perform the computation of the probabilities.

% \begin{example}[Computing the Indian GPA Problem]
%     \label{example:compute_indian}
%     Let us first write the relevant ground program for the query
%     \begin{align*}       
%     P \Bigl(\mathprobloginline{american}{=}\top \mid (\text{\probloginline{gpa(student)}}\doteq4)=\top \Bigr)
%     \end{align*}
%     where we also transform distributional clauses into distributional facts and introduce an atom \probloginline{ev} for the evidence.

%     \begin{problog*}{linenos}
% rva ~ flip(1/4).
% american:- rva=:=1.
% rvisd ~ flip(19/20).
% isdensity(a):-  rvisd=:=1.
% rvp ~ flip(17/29).
% perfect_gpa(a):- rvp=:=1.

% gpa1~uniform(0,4).
% gpa2~delta(4.0).  
% gpa3~delta(0.0)

% ev:- american, delta_interval(gp1,4), isdensity(a).
% ev:- american, delta_interval(gp2,4), not isdensity(a),
%     perfect_gpa(a). 
% ev:- american, delta_interval(gp3,4), not isdensity(a),
%     not perfect_gpa(a).
% \end{problog*}
% The query for the transformed program is now:
% \begin{align*}
% P\Bigl(\text{\probloginline{american}}{=}\top \mid \mathprobloginline{ev} =\top\Bigr)
%     =
%     \frac{
%         P\Bigl(\mathprobloginline{american}{=}\top, \mathprobloginline{ev}=\top \Bigr)
%     }
%     {
%         P \Bigl(\mathprobloginline{ev}=\top \Bigr)
%     }
% \end{align*}

% \end{example}

\subsection{Infinitesimal Algebraic Likelihood Weighting via Knowledge Compilation}
\label{sec:ALWviaKC}

Inspecting Equation~\ref{eq:prop:alwapproximation} we see that we have to evaluate expressions of the following form in order to compute the probability of a conditional query to a \dcproblogsty program.
\begin{align}
    \bigoplus_{i=1}^{\lvert \mathcal{S} \rvert} \underbrace{\bigoplus_{\omega\in ENUM(\varphi)} \bigotimes_{\ell \in \varphi}  \alpha_{IALW}^{(i)} \left(\ell \right)}_{= \text{algebraic model count}} \label{eq:alw_show}
\end{align}
In other words, we need to compute $\lvert \mathcal{S} \rvert$ times a sum over products -- each time with a different ancestral sample. Such a sum over products is also called the algebraic model count of  a formula $\phi$~\citep{kimmig2017algebraic}. 
Subsequently, we then add up the $\lvert \mathcal{S} \rvert$ results from the different algebraic model counts giving us the final answer.

Unfortunately, computing the algebraic model count is in general a computationally hard problem~\citep{kimmig2017algebraic} -- \#P-hard to be precise~\citep{valiant1979complexity}.
A popular technique to mitigate this hardness is to use a technique called knowledge compilation~\citep{darwiche2002knowledge}, which splits up the computation into a hard step and a subsequent easy step. The idea is to take the propositional Boolean formula underlying 
an algebraic model counting problem (cf. $\varphi$ in Equation~\ref{eq:alw_show}) and compile it into a logically equivalent formula that allows for the tractable computation of algebraic model counts. The compilation constitutes the computationally hard part (\#P-hard). Afterwards, the algebraic model count is performed on the compiled structure, also called {\em algebraic circuit}~\citep{zuidbergdosmartires2019transforming}. Intuitively speaking, knowledge compilation takes the sum of products and maps it to recursively nested sums and products. Effectively, finding a dynamic programming scheme~\citep{bellman1957dynamic} to compute the initial sum of products.

Different circuit classes have been identified as valid knowledge compilation targets~\citep{darwiche2002knowledge} -- all satisfying different properties.
Computing the algebraic model count on an algebraic circuit belonging to a specific target class is only correct if the properties of the circuit class match the properties of the deployed semiring.
The following three lemmas will help us determining which class of circuits we need to knowledge-compile our propositional formula $\phi$ into.

\begin{lemma}
\label{lem:non_idem}
The operator $\oplus$ (c. Definition~\ref{def:inf_number_ops}) is not idempotent. That is, it does not hold for every $a \in \mathbb{I}$ that $a\oplus a =a$. 
\end{lemma}
\begin{lemma} The pair  $(\oplus, \alpha_{\ialw})$ is not neutral. That is, it does not hold that $\alpha_{\ialw}(\ell)\oplus \alpha_{\ialw}(\neg \ell) = e^{\otimes}$ for arbitrary $\ell$.
\end{lemma}
\begin{lemma} 
\label{lem:non_cons}
The pair  $(\otimes, \alpha_{\ialw})$ is not consistency-preserving. That is, it does not hold that $\alpha_{\ialw}(\ell)\otimes \alpha_{\ialw}(\neg \ell) = e^{\oplus}$ for arbitrary $\ell$.
\end{lemma}

From~\citep[Theorem 2 and Theorem 7]{kimmig2017algebraic} and the three lemmas above, we can conclude that we need to compile our propositional logic formulas into so-called smooth, deterministic and decomposable negation normal form (sd-DNNF) formulas~\citep{darwiche2001tractable}.\footnote{Note that we only require smoothness over derived atoms (otherwise case in Definition~\ref{def:sample_labeling_function}), as for the other cases the neutral sum property holds. Certain encodings of logic programs eliminate derived atoms. For such encodings the smoothness property can be dropped~\citep{vlasselaer2014compiling}. A more detailed discussion on the smoothness requirement of circuits in a PLP context can be found in \citep[Appendix C]{fierens2015inference}.}

% \begin{proposition}[ALW on d-DNNF]
\begin{restatable}[ALW on d-DNNF]{proposition}{alwonddnnf}
\label{prop:alwonddnnf}
    We are given the propositional formulas $\phi$ and $\phi_q$ and a set $\mathcal{S}$ of ancestral samples, we can use Algorithm~\ref{alg:prob_via_alw_kc} to compute the conditional probability $P_\dcpprogram(\mu=q|\evidenceset=e)$.
\end{restatable}

\begin{proof}
    See Appendix~\ref{app:proof:alwonddnnf}.
\end{proof}

Algorithm~\ref{alg:prob_via_alw_kc} takes as input a two propositional logic formulas $\phi$ and $\phi_q$, and a set of ancestral samples. It then knowledge-compiles the formulas $\phi \land \phi_q$ and $\phi$ into circuits $\Gamma_q$ and $\Gamma$. These circuits are then evaluated using Algorithm~\ref{alg:unormalize_alw}. The variables $ialw_q$ and $ialw$ hold infinitesimal numbers.
\fixed{
The ratio of these two numbers, which corresponds to the ratio in Equation~\ref{eq:prop:alwapproximation}, is an infinitesimal number having as second argument $0$ and as first argument the conditional probability.}
\begin{algorithm}[h]
    \SetKwFunction{ProbALW}{ProbALW}
    \SetKwFunction{KC}{KC}
    \SetKwFunction{IALW}{IALW}

    \SetKwProg{Fn}{function}{}{}
    \SetKwProg{ElseComment}{function}{}{}

    \caption{Conditional Probability via IALW and KC}
	\label{alg:prob_via_alw_kc}
\Fn{\ProbALW{$\phi$, $\phi_q$, $\mathcal{S}$ }}{
    $\Gamma_{q}$ $\leftarrow$ \KC{$\phi \land \phi_q$}\;
     \label{alg:prob_via_alw_kc:kc_qe} 
    $\Gamma$ $\leftarrow$ \KC{$\phi$}\;
    \label{alg:prob_via_alw_kc:kc_e}
    $ialw_{q}$ $\leftarrow$ \IALW{$\Gamma_{q}$,$\mathcal{S}$} \label{alg:prob_via_alw_kc:alw_qe}
    \tcp*[r]{cf. Algorithm~\ref{alg:unormalize_alw}} 
    $ialw$ $\leftarrow$ \IALW{$\Gamma$,$\mathcal{S}$}   \label{alg:prob_via_alw_kc:alw_e} 
    \tcp*[r]{cf. Algorithm~\ref{alg:unormalize_alw}}
    $(p,0)$ $\leftarrow$ $ialw_{q} \oslash ialw$\;
    \Return $p$ \label{alg:prob_via_alw_kc:conditional}
	}
\end{algorithm}

\begin{algorithm}[h]
    \SetKwFunction{IALW}{IALW}
    \SetKwFunction{Eval}{Eval}
    \SetKwProg{Fn}{function}{}{}
    \SetKwProg{ElseComment}{function}{}{}
    \caption{Computing the IALW}
	\label{alg:unormalize_alw}
\Fn{\IALW{$\Gamma$, $\mathcal{S}$}}{
    $ialw$ $\leftarrow$ (0,0) \;
    \For{$\varset{s}^{(i)} \in \mathcal{S}$  }{
        $ialw$ $\leftarrow$ $ialw$ $\oplus$ \Eval{$\Gamma$, $\varset{s}^{(i)}$}
        \tcp*[r]{cf. Algorithm~\ref{alg:eval}}  
    }
    \Return $ialw$
	}
\end{algorithm}

Algorithm~\ref{alg:unormalize_alw} computes the IALW given as input a circuit $\Gamma$ and a set of ancestral samples.
The loop evaluates the circuit (using Algorithm~\ref{alg:eval}) for each ancestral sample $\varset{s}^{(i)}$ and accumulates the result, which is then returned once the loop terminates. The accumulation inside the loop corresponds to the  $\bigoplus_{i=1}^{\lvert \mathcal{S} \rvert}$ summation in Equation~\ref{eq:alw_show}.
Algorithm~\ref{alg:eval} evaluates a circuit $\Gamma$ for a single ancestral sample $\varset{s}^{(i)}$ and is a variation of the circuit evaluation algorithm presented by~\citet{kimmig2017algebraic}.

\begin{algorithm}[h]
    \SetKwFunction{EvalFn}{Eval}
    \SetKwProg{Fn}{function}{}{}
    \SetKwProg{ElseComment}{function}{}{}

	\caption{Evaluating an sd-DNNF circuit $\Gamma$ for labeling function $\alpha^{(i)}$ (Definition~\ref{def:sample_labeling_function}) and semiring operations $\oplus$  and $\otimes$ (Definition~\ref{def:inf_number_ops})}
	\label{alg:eval}
\Fn{\EvalFn{$\Gamma$,$\varset{s}^{(i)}$}}{
		\If{  $\Gamma$ is a literal node $l$}{ \Return $\alpha^{(i)}(l)$}
		\ElseIf{$\Gamma$ is a disjunction $\bigvee_{j=1}^m \Gamma_j$}
		{\Return $\bigoplus_{j=1}^m$ \EvalFn{$\Gamma_j$, $\varset{s}^{(i)}$}}
		\Else( \tcp*[f]{$\Gamma$ is a conjunction $\bigwedge_{j=1}^m \Gamma_j$}){
		 \Return $\bigotimes_{i=j}^m$ \EvalFn{$\Gamma_j$, $\varset{s}^{(i)}$}}
	}
\end{algorithm}

\begin{example}[IALW on Algebraic Circuit]
\label{example:eval_observation}

Consider a version of the program in Example~\ref{example:dcproblog:observation} where the annotated disjunction has been eliminated and been replaced with a binary random variable $m$ and a \probloginline{flip} distribution.
	\begin{problog*}{linenos}
m~flip(0.3).

size~beta(2,3):- m=:=0.
size~beta(4,2):- m=:=1.
	\end{problog*}
We query the program for the conditional probability $P((\mathprobloginline{m=:=1}) =\top \mid \mathprobloginline{size}\doteq 4/10 )$.
Following the program transformations introduced in Section~\ref{sec:dc2smt} and then compiling the labeled propositional formula, we obtain a circuit representation of the queried program. Evaluating this circuit yields the probability of the query. To be precise, we actually obtain two circuits, one representing the probability of relevant program with the evidence enforced and with additionally having the value of the query atom set. 
In Figure~\ref{fig:circuit:ialw} we show the circuit where only the evidence is enforced.

	\begin{figure}[h]
	\resizebox{\linewidth}{!}{%

		\tikzstyle{distribution}=[rectangle, text centered, fill=white, draw, dashed,thick]
		\tikzstyle{leaf}=[rectangle, text centered, fill=gray!10, draw,thick]

		\tikzstyle{negate}=[
			rectangle split,
			rectangle split parts=3, 
			rectangle split horizontal,
			text centered,
			rectangle split part fill={gray!10,white,gray!10},
			draw,
			rectangle split draw splits=false,
			anchor=center,
			align=center,
			thick
		]
		\newcommand{\minus}{  ${\bm e^\otimes}$ \nodepart{second} ${{\bm \ominus}}$ \nodepart{third}  \phantom{${\bm e^\otimes}$}}

		\tikzstyle{sumproduct}=[
			rectangle split,
			rectangle split parts=3, 
			rectangle split horizontal,
			text centered,
			rectangle split part fill={gray!10,white,gray!10},
			draw,
			rectangle split draw splits=false,
			anchor=center,
			align=center,
			thick
		]
		\newcommand{\supr}[1]{  \phantom{${\bm e^\otimes}$} \nodepart{second} ${{\bm #1}}$ \nodepart{third} \phantom{${\bm e^\otimes}$}}

		\tikzstyle{circuitedge}=[ultra thick, thick,->]
		\tikzstyle{distributionedge}=[thick,->, dashed, in=-90]

		\tikzstyle{indexnode}=[draw,circle, inner sep=1pt]

		\begin{tikzpicture}[remember picture]

			\node[sumproduct] (14) at (200.54bp,378.0bp) {\supr{\oplus}};
			\draw[ultra thick, thick,->] (14.90) to  ([shift={(0,1)}]14.90);

			\node[sumproduct] (9) [below left = of 14] {\supr{\otimes}};
			\node[sumproduct] (13) [below right = of 14]  {\supr{\otimes}};

			\node[negate] (m1) [below=of 9] {\minus};
			\node[sumproduct] (12)  [below  = of 13] {\supr{\oplus}};

			\node[sumproduct] (8)  [below=of 12]  {\supr{\otimes}};

			\node[indexnode, left=of 14, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$6$}};			
			\node[indexnode, left=of 13, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$5$}};			
			\node[indexnode, left=of 8, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$2$}};			
			\node[indexnode, left=of 12, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$4$}};			
			\node[indexnode, left=of 9, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$3$}};			
			\node[indexnode, left=of m1, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$1$}};

			\node[leaf, below= of 8, xshift=-0.2cm] (4)   {$\subnode{var_m0}{$m$}=1$};
			\node[leaf, below= of 8, xshift=1.8cm] (2) {$\subnode{var_m1}{$m$}=0$};				
			\node[leaf] (size1obs) [left=of 4, xshift=0cm]  {$\subnode{var_s_11_un}{size_{1}} \doteq 0.4$};
			\node[leaf] (size0obs) [left=of 4, xshift=-4cm] {$\subnode{var_s_01_un}{size_{0}} \doteq 0.4$};

			\node[distribution] (size0)  [below=of size0obs] {\probloginline{beta(2,3)}};
			\node[distribution] (size1)  [below=of size1obs] {\probloginline{beta(4,2)}};
			\node[distribution] (flip)  [below= of 4, xshift=1cm] {\probloginline{flip(0.3)}};

			\draw[distributionedge,out=90]  (size1) to (var_s_11_un);
			\draw[distributionedge,out=90] (size0) to  (var_s_01_un);
			\draw[distributionedge,out=160] (flip) to  (var_m0);
			\draw[distributionedge] (flip) to  (var_m1);

			%https://tex.stackexchange.com/questions/447989/anchor-node-names-for-tikz-rectangle-split-horizontal			
			\draw[circuitedge] (9) to  (14.mid);
			\draw[circuitedge] (13) to  (14.three south |- 14.mid);
			\draw[circuitedge] (m1) to  (9.mid);
			\draw[circuitedge] (12) to  (13.three south |- 13.mid);
			\draw[circuitedge] (size0obs) to   (m1.three south |- m1.mid);	
			\draw[circuitedge] (2) to  (12.three south |- 12.mid);				
			\draw[circuitedge] (size1obs) to   (8.mid);
			\draw[circuitedge] (4) to  (8.three south |- 8.mid);	
			\draw[circuitedge] (8) to  (9.three south |- 9.mid);	
			\draw[circuitedge] (8) to   (12.mid);
			\draw[circuitedge] (size0obs) to  (13.mid);

		\end{tikzpicture}
	}
    \captionof{figure}{At the bottom of the circuit we see the distributions feeding in. The \probloginline{flip} distribution feeds into its two possible (non-zero probability) outcomes. The two \probloginline{beta} distributions feed into an observation statement each. We use the `$\doteq$' symbol to denote such an observation. Note how we identify each of the two random variables for the size by a unique identifier in their respective subscripts. The circled numbers next to the internal nodes, \ie the sum and product nodes, will allow us to reference the nodes later on and do not form a part of the algebraic circuit.}
    \label{fig:circuit:ialw}
\end{figure}

The probability of the query (given the evidence) can now be obtained by evaluating recursively the internal nodes in the algebraic circuit using Algorithm~\ref{alg:eval}.
We perform the evaluation  of the circuit in Figure~\ref{fig:circuit:ialw} for a single iteration of the loop in Algorithm~\ref{alg:unormalize_alw}, and we assume that we have sampled the value $m=0$ from the \probloginline{flip(0.3)} distribution.

\begin{minipage}{0.49\linewidth}
    \begin{align*}
        &\mathtt{Eval}(\footcircled{$1$})\\
        &=
        e^\otimes \ominus \alpha_{IALW}\big(size_0\doteq 0.4\big) \\
        &=
        (1,0) \ominus (1.728,1) \\
        &= (1,0)
        \\
        \hfill
        \\
        % 
        &\mathtt{Eval}(\footcircled{$2$})\\
        &=
        \alpha_{IALW}\big( size_1\doteq 0.4) \otimes \alpha_{IALW}\big(  m=1 \big) \\
        &=
        (0.768,1) \otimes (0,0) \\
        &=
        (0,1)\\
        \hfill
        \\
        % 
        &\mathtt{Eval}(\footcircled{$3$})\\
        &=
        \mathtt{Eval}(\footcircled{$1$}) \otimes \mathtt{Eval}(\footcircled{$2$}) \\
        &=
        (1,0) \otimes (0,1) \\
        &=
        (0,1)\\
    \end{align*}
\end{minipage}
\vline
\begin{minipage}{0.49\linewidth}
    \begin{align*}
        &\mathtt{Eval}(\footcircled{$4$})\\
        &=
        \mathtt{Eval}(\footcircled{$2$}) \oplus \alpha_{IALW}\big(  m=0 \big) \\
        &=
        (0,1) \oplus (1,0) \\
        &=
        (1,0)\\
        \hfill
        \\
        % 
        &\mathtt{Eval}(\footcircled{$5$})\\
        &=
        \alpha_{IALW}\big(  size_0\doteq 0.4 \big) \otimes \mathtt{Eval}(\footcircled{$2$})  \\
        &=
        (1.728,1) \otimes (1,0) \\
        &=
        (1.728,1)\\
        \hfill
        \\
        % 
        &\mathtt{Eval}(\footcircled{$6$})\\
        &=
        \mathtt{Eval}(\footcircled{$3$}) \oplus \mathtt{Eval}(\footcircled{$5$})  \\
        &=
        (0,1) \oplus (1.728,1) \\
        &=
        (1.728,1)\\
    \end{align*}
\end{minipage}

If we evalute the circuit for a sample $m=1$ we obtain in a similar fashion the result $\mathtt{Eval}(\footcircled{$6$})= (0.768,1)$. Moreover, if we evaluate the circuit multiple times we obtain (in the limit) 70\% of the time the outcome $(1.728,1)$ and 30\% of the time the value $(0.768,1)$. This yields an average of $(0.7\times 1.728, 1)\oplus (0.3\times 0.768, 1)= (1.440,1)$ and represents the unnormalized infinitesimal algebraic likelihood weight of the evidence.
The unnormalized infinitesimal algebraic likelihood weight of the query conjoined with the evidence is obtain again in a similar fashion but with the samples for $m=0$ being discarded. This then yields the result $(0.3\times 1.728, 1)$.
Dividing these two (unnormalized) infinitesimal algebraic likelihood weights by each other gives the probability of the query.
\begin{align*}
    &P((\mathprobloginline{m=:=1}) =\top \mid \mathprobloginline{size}\doteq 4/10 )\\
    &=(0.3\times 1.728, 1) \oslash \Big( (0.7\times 0.768,1 ) \oplus (0.3\times 1.728,1) \Big) \\
    &= (0.2304/1.440 , 1{-}1) \\
    &= (0.16,0)
\end{align*}
\end{example}

\subsection{Partial Symbolic Inference}

Evaluating circuits using binary random variables is quite wasteful: on average half of the samples are unused for one of the two possible outcomes ($0$ or $1$). We can remedy this by performing (exact) symbolic inference on binary random variables and replace the comparisons where they appear with their expectation. For instance, we replace \probloginline{m=:=1} by the infinitesimal number $(0.3,0)$ instead of sampling a value for \probloginline{m} and testing whether the sample satisfies the constraint. This technique is also used by other probabilistic programming languages such as \problogsty~\citep{fierens2015inference} and Dice~\citep{holtzen2020dice}. The main difference to \dcproblogsty is that those languages only support binary random variables (and by extension discrete random variables with finite support), while \dcproblogsty interleaves discrete and continuous random variables.

In a sense, the expectation gets pushed from the root of the algebraic circuit representing a probability to its leaves. This is, however, only possible if the circuit respects specific properties. Namely, the ones respected by \mbox{d-DNNF} formulas (cf. Section~\ref{sec:ALWviaKC}), which we use as our representation language for the probability.

\begin{definition}[Symbolic IALW Label of a Literal] \label{def:sample_probability_labeling_function}
Given an ancestral sample $\varset{s}^{(i)}= (s_1^{(i)}, \dots,  s_M^{(i)} ) $ for the random variables $\randomvariableset = (\nu_1,\dots, \nu_M)$.
The  Symbolic IALW (SIALW) label of a positive literal $\ell$ is an infinitesimal number given by:
\begin{align}
    \alpha_{SIALW}^{(i)}( \ell)
    =\begin{cases}
    (p_{\ell}, 0), & \text{if $\ell$ encodes a probabilistic fact} \\
    \alpha^{(i)}_{IALW}(\ell), & \text{otherwise} 
    \end{cases}
    \nonumber
\end{align}
For the negated literals we have the following labeling function:
\begin{align}
    \alpha_{SIALW}^{(i)}( \neg \ell)
    =\begin{cases}
    (1{-}p_{\ell}, 0), & \text{if $\ell$ encodes a probabilistic fact} \\
    \alpha^{(i)}_{IALW}(\neg \ell), & \text{otherwise}
    \end{cases}
    \nonumber
\end{align}
The number $p_\ell$ is the label of the probabilistic fact in a \dcproblogsty program.
\end{definition}

In the definition above we replace the label of a comparison that corresponds to a probabilistic fact with the probability of that fact being satisfied. This has already been shown to be beneficial when performing inference, both in terms of inference time and accuracy of Monte Carlo estimates~\citep{zuidbergdosmartires2019exact}. Following the work of~\citep{kolb2019exploit} one could also develop more sophisticated methods to detect which comparison in the leaves can be replaced with their expectation. We leave this for future work.

\begin{example}[Symbolic IALW on Algebraic Circuit]
\label{example:eval_observation_marginalized}

Symbolic inference for the random variable $m$ from the circuit in Example~\ref{example:eval_observation} results in annotating the leaf nodes for the different outcomes of the random variable $m$ with the probabilities of the respective outcomes. This can be seen in the red dashed box in the bottom right of Figure~\ref{fig:circuit:sialw}.

	\begin{figure}[h]
		\resizebox{\linewidth}{!}{%

			\tikzstyle{distribution}=[rectangle, text centered, fill=white, draw, dashed,thick]
			\tikzstyle{leaf}=[rectangle, text centered, fill=gray!10, draw,thick]

			\tikzstyle{negate}=[
			rectangle split,
			rectangle split parts=3, 
			rectangle split horizontal,
			text centered,
			rectangle split part fill={gray!10,white,gray!10},
			draw,
			rectangle split draw splits=false,
			anchor=center,
			align=center,
			thick
			]
			\newcommand{\minus}{  ${\bm e^\otimes}$ \nodepart{second} ${{\bm \ominus}}$ \nodepart{third}  \phantom{${\bm e^\otimes}$}}

			\tikzstyle{sumproduct}=[
			rectangle split,
			rectangle split parts=3, 
			rectangle split horizontal,
			text centered,
			rectangle split part fill={gray!10,white,gray!10},
			draw,
			rectangle split draw splits=false,
			anchor=center,
			align=center,
			thick
			]
			\newcommand{\supr}[1]{  \phantom{${\bm e^\otimes}$} \nodepart{second} ${{\bm #1}}$ \nodepart{third} \phantom{${\bm e^\otimes}$}}

			\tikzstyle{circuitedge}=[ultra thick, thick,->]
			\tikzstyle{distributionedge}=[thick,->, dashed, in=-90]

	    	\tikzstyle{indexnode}=[draw,circle, inner sep=1pt]

			\begin{tikzpicture}[remember picture]

				\node[sumproduct] (14) at (200.54bp,378.0bp) {\supr{\oplus}};
				\draw[ultra thick, thick,->] (14.90) to  ([shift={(0,1)}]14.90);

				\node[sumproduct] (9) [below left = of 14] {\supr{\otimes}};
				\node[sumproduct] (13) [below right = of 14]  {\supr{\otimes}};

				\node[negate] (m1) [below=of 9] {\minus};
				\node[sumproduct] (12)  [below  = of 13] {\supr{\oplus}};

				\node[sumproduct] (8)  [below=of 12]  {\supr{\otimes}};

    			\node[indexnode, left=of 14, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$6$}};			
    			\node[indexnode, left=of 13, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$5$}};			
    			\node[indexnode, left=of 8, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$2$}};			
    			\node[indexnode, left=of 12, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$4$}};			
    			\node[indexnode, left=of 9, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$3$}};			
    			\node[indexnode, left=of m1, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$1$}};

				\node[leaf, below= of 8, xshift=-0.2cm] (4)   {$3/10$};
				\node[leaf, below= of 8, xshift=1.8cm] (2) {$7/10$};
				\draw[red,ultra thick,dashed] ($(4.north west)+(-0.2,0.2)$)  rectangle ($(2.south east)+(0.2,-0.2)$);

				\node[leaf] (size1obs) [left=of 4, draw,  xshift=0cm]  {$\subnode{var_s_11}{size_{1}} \doteq 0.4$};
				\node[leaf]  (size0obs) [left =of 4, xshift=-4cm] {$\subnode{var_s_01}{size_{0}} \doteq 0.4$};

		    	\node[distribution] (size0)  [below=of size0obs] {\probloginline{beta(2,3)}};
		    	\node[distribution] (size1)  [below=of size1obs] {\probloginline{beta(4,2)}};

				\draw[distributionedge,out=90]  (size1) to (var_s_11);
				\draw[distributionedge,out=90] (size0) to  (var_s_01);

				%https://tex.stackexchange.com/questions/447989/anchor-node-names-for-tikz-rectangle-split-horizontal			
				\draw[circuitedge] (9) to  (14.mid);
				\draw[circuitedge] (13) to  (14.three south |- 14.mid);
				\draw[circuitedge] (m1) to  (9.mid);
				\draw[circuitedge] (12) to  (13.three south |- 13.mid);
				\draw[circuitedge] (size0obs) to   (m1.three south |- m1.mid);	
				\draw[circuitedge] (2) to  (12.three south |- 12.mid);				
				\draw[circuitedge] (size1obs) to   (8.mid);
				\draw[circuitedge] (4) to  (8.three south |- 8.mid);	
				\draw[circuitedge] (8) to  (9.three south |- 9.mid);	
				\draw[circuitedge] (8) to   (12.mid);
				\draw[circuitedge] (size0obs) to  (13.mid);

			\end{tikzpicture}
		}
        \captionof{figure}{Circuit representation of the SIALW algorithm for the probability $P(\mathprobloginline{size}\doteq 4/10 )$.}
        \label{fig:circuit:sialw}

	\end{figure}

Evaluating the marginalized circuit now returns immediately the unnormalized algebraic model count for the evidence without the need to draw samples and consequently without the need to sum over the samples.

\begin{minipage}{0.49\linewidth}
    \begin{align*}
        &\mathtt{Eval}(\footcircled{$1$})\\
        &=
        e^\otimes \ominus \alpha_{SIALW}\big(size_0\doteq 0.4\big) \\
        &=
        (1,0) \ominus (1.728,1) \\
        &= (1,0)
        \\
        \hfill
        \\
        % 
        &\mathtt{Eval}(\footcircled{$2$})\\
        &=
        \alpha_{SIALW}\big( size_1\doteq 0.4) \otimes \alpha_{SIALW}\big(  m=1 \big) \\
        &=
        (0.768,1) \otimes (0.3,0) \\
        &=
        (0.2304,1)\\
        \hfill
        \\
        % 
        &\mathtt{Eval}(\footcircled{$3$})\\
        &=
        \mathtt{Eval}(\footcircled{$1$}) \otimes \mathtt{Eval}(\footcircled{$2$}) \\
        &=
        (1,0) \otimes (0.2304,1) \\
        &=
        (0.2304,1)\\
    \end{align*}
\end{minipage}
\vline
\begin{minipage}{0.49\linewidth}
    \begin{align*}
        &\mathtt{Eval}(\footcircled{$4$})\\
        &=
        \mathtt{Eval}(\footcircled{$2$}) \oplus \alpha_{SIALW}\big(  m=0 \big) \\
        &=
        (0.2304,1) \oplus (0.7,0) \\
        &=
        (0.7,0)\\
        \hfill
        \\
        % 
        &\mathtt{Eval}(\footcircled{$5$})\\
        &=
        \alpha_{SIALW}\big(  size_0\doteq 0.4 \big) \otimes \mathtt{Eval}(\footcircled{$2$})  \\
        &=
        (1.728,1) \otimes (0.7,0) \\
        &=
        (1.2096,1)\\
        \hfill
        \\
        % 
        &\mathtt{Eval}(\footcircled{$6$})\\
        &=
        \mathtt{Eval}(\footcircled{$3$}) \oplus \mathtt{Eval}(\footcircled{$5$})  \\
        &=
        (0.2304,1 \oplus (1.2096,1) \\
        &=
        (1.440,1)\\
    \end{align*}
\end{minipage}

\end{example}

\subsection{Experimental Evaluation}
\label{sec:experimental}

\new{

In order to demonstrate the benefits of adapting the technique of knowledge compilation to the discrete-continuous domain with zero-probability conditioning events, we model a machine that runs either in operating \probloginline{mode1} or \probloginline{mode2}; (with probability $0.2$ and $0.8$ respectively). Furthermore, the machine can be faulty with a small probability of $10^{-5}$. In this case we would like to switch the machine of and repair it.

If the machine is not faulty the temperature measurements we perform on the machine are distributed according to two Gaussian (Lines~\ref{line:ex:faulty:temp1} and \ref{line:ex:faulty:temp2}).
If the machine is faulty, however, we get a deterministic temperature reading of $2.0$ (Line \ref{line:ex:faulty:temp3}).

}

\begin{problog*}{linenos}
0.00001::faulty.
0.2::mode1;0.7::mode2.

temperature ~ normal(0.5,1.0) :- \+faulty, mode1. @\label{line:ex:faulty:temp1}@
temperature ~ normal(2.0,2.0) :- \+faulty, mode2. @\label{line:ex:faulty:temp2}@
temperature ~ delta(2.0) :- faulty. @\label{line:ex:faulty:temp3}@
\end{problog*}

\new{
We are now interested in computing $p( \mathprobloginline{faulty}{=}\top {\mid} \mathprobloginline{temperature}{\doteq} 2.0 )$. That is, what is the probability that the machine is faulty given that the temperature measurement is $2.0$.

In our experiment we compared the performance of SIALW (\cf Definition~\ref{def:sample_probability_labeling_function}) to the inference algorithm of \dcsty~\citep{nitti2016probabilistic}. The latter is equivalent to the algorithms presented by \Citet{wu2018discrete} and \Citet{jacobs2021paradoxes} as all three perform, in essence, likelihood weighting with infinitesimal numbers.   
Specifically, we study the sensitivity of the algorithms with regard to the probability of the machine being faulty.

In Figure~\ref{fig:experiment:faultVScorrectness} we see that using SIALW computes the correct probability regardless of the fault probability $\{10^{-5}, 10^{-4}, 10^{-3}, 10^{-2}, 10^{-1} \}$. We also see that the naivc likelihood weighting algorithm (without the exact symbolic inference of SIALW) needs a substantial amount of samples to infer the correct probability. Most notably, for a fault probability of $10^{-5}$ not even a sample size of $10^5$ is sufficient.

The large discrepancy between SIALW and the competing approach by \Citet{nitti2016probabilistic} is explained as follows: in order to correctly infer the queried probability one of the samples drawn from the Bernoulli distribution \probloginline{faulty ~ flip(0.00001)} needs to be true. This would then trigger the rule for \probloginline{temperature ~ delta(2.0) :- faulty}. As this is, however, extremely unlikely the crucial rule needed to perform correct likelihood weighting with infinitesimal numbers is never triggered and the returned probability is incorrect.
SIALW, in constrast, does not sample \probloginline{faulty ~ flip(0.00001)} but performs exact inference using knowledge compilation. As a result SIALW always computes the correct posterior probability.

\begin{figure}
    \begin{center}
        \includegraphics[width=\linewidth]{experiment_dc/faultVScorrectness.pdf}
    \end{center}
    \caption{
        \new{
        We queried SIALW and (non-symbolci) likelihood weighting each $100$ times for the probability $p( \mathprobloginline{faulty}{=}\top {\mid} \mathprobloginline{temperature}{\doteq} 2.0 )$. On the $y$-axis we give the ratio $\nicefrac{\text{\#correct runs}}{\text{\#runs}}$.
        Die to the use of knowledge compilation, SIALW is insensitive to the probability of fault (on the x-axis).
        This is in contrast to the log-likelihood weighting (LLW) algorithm presented by~\Citet{nitti2016probabilistic}, which necessitates a considerable number of samples to compute the queried probability reliably. The different dotted lines indicate settings with varying sample sizes ($\{10^1, 10^2, 10^3, 10^4, 10^5 \}$).  
        }
    }
    \label{fig:experiment:faultVScorrectness}
\end{figure}
}

% \begin{figure}
%     \begin{center}
%         \includegraphics[width=\linewidth]{experiment_dc/samplesVScorrectness.pdf}
%     \end{center}
%     \caption{test}
%     \label{fig:experiment:samplesVScorrectness}
% \end{figure}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_main/introduction.tex
aij dcproblog paper

\section{Introduction}
\label{sec:introduction}

Probabilistic logic programming (PLP) is at the crossroads of two parallel developments in artificial intelligence and machine learning.
On the one hand, there are the probabilistic programming languages with built-in support for machine learning. These languages can be used to represent very expressive -- Turing equivalent
-- probabilistic models, and they provide primitives for inference and learning.
On the other hand, there is the longstanding open question for integrating the two main frameworks
for reasoning, that is logic and probability, within a common framework \citep{russell2015unifying,deraedt:mc16}.
Probabilistic logic programming~\citep{de2015probabilistic,riguzzi2018foundations} fits both paradigms and goes back to at least the early 90s
with seminal works by \citet{sato1995statistical} and \citet{poole1993probabilistic}. 
\citeauthor{poole1993probabilistic} introduced ICL, the Independent Choice Logic, an elegant
extension of the Prolog programming language, and \citeauthor{sato1995statistical} introduced
the {\em distribution semantics} for probabilistic logic programs in conjunction with a learning algorithm based on expectation maximization (EM).
The PRISM language \citep{sato1995statistical}, which utilizes the distribution semantics and the EM learning algorithm constitutes, to the best of the authors' knowledge, the very first probabilistic programming language with support for machine learning.

Today, there is a plethora of probabilistic logic programming languages, most of which are based
on extensions of the ideas by \citeauthor{sato1995statistical} and~\citeauthor{poole1993probabilistic} ~\citep{sato1997prism,kersting2000bayesian,vennekens2004logic,de2007problog}. However, the vast majority of them is restricted to discrete, and more precisely finite categorical, random variables. 
When merging logic with probability, the restriction to discrete random variables
is natural and allowed Sato to elegantly extend the logic program semantics
into the celebrated distribution semantics.

\begin{example}[Probabilistic Logic Program]
  \label{example:intro}

  \new{
  Consider the probabilistic logic program below (written in \problogsty syntax~\citep{fierens2015inference}), where we model the behavior of two machines. We first state that there are two machines (Line~\ref{line:exintro:1}). Subsequently, we say that the temperature has a probability of $0.8$ to be low (Line \ref{line:exintro:temp}) and that the cooling of the machines works with probability $0.99$ and $0.95$ respectively (Lines \ref{line:exintro:cool1} and \ref{line:exintro:cool2}) These labeled facts are called \textit{probabilistic facts}. We also model that the machines themselves work: either if the cooling is working (Line~\ref{line:exintro:work1}) or if the temperature is low (Line~\ref{line:exintro:work2}).
  }
  \begin{problog*}{linenos}
machine(1). machine(2). @\label{line:exintro:1}@
0.8::temperature(low). @\label{line:exintro:temp}@
0.99::cooling(1).@\label{line:exintro:cool1}@
0.95::cooling(2).@\label{line:exintro:cool2}@

works(N):- machine(N), cooling(N). @\label{line:exintro:work1}@
works(N):- machine(N), temperature(low). @\label{line:exintro:work2}@

    \end{problog*}
    \new{
We can now, for instance, ask for the conditional probability of the first machine working given that the second one works:
    $$
    P(\mathprobloginline{works(1)}=\top \mid \mathprobloginline{works(2)}=\bot).
    $$
The (exact) inference algorithm currently implemented in \problogsys~\citep{fierens2015inference,dries2015problog2} then returns as answer the probability $\approx 0.998$.
}
\end{example}

While \citeauthor{sato1995statistical}'s extension of logic programming to the probabilistic domain is elegant, it also imposes an important restriction to random variables with countable sample spaces. This raises the question of how to extend the distribution semantics towards hybrid, \ie discrete-continuous, random variables.

Defining the semantics of probabilistic programming language with support for random variables with infinite and possibly uncountable sample spaces is a much harder task. This can be observed when looking at the development of important imperative and functional probabilistic programming languages~\citep{goodman2008church,mansinghka2014venture} that support  continuous random variables. 
These works initially focused on inference, typically using  a particular Monte Carlo approach, yielding  an operational or procedural semantics. It is only follow-up work that started to address a declarative semantics for such hybrid probabilistic programming languages.
~\citep{staton2016semantics,wu2018discrete}.
%\footnote{Aside from some subtle differences (\cf~\citep{kimmig2017probabilistic}), in non-logical probabilistic programming languages declarative semantics are usually called denotational and procedural semantics are referred to as operational.}

The PLP landscape has experienced similar struggles. First approaches for  hybrid PLP languages were achieved by restricting the language~\citep{gutmann2010extending,gutmann2011magic,islam2012inference} or via recourse to procedural semantics~\citep{nitti2016probabilistic}.  The key contributions of this paper are:

\begin{enumerate}[label={\bf C\arabic*}]
  \setcounter{enumi}{0}
\item We introduce the {\em measure semantics} for mixed discrete-continuous probabilistic logic programming.
Our {\em measure semantics} (based on measure theory) extends \citeauthor{sato1995statistical}'s distribution semantics and supports:
\begin{itemize}
    \item  \label{item:k1} a countably infinite number of random variables,
     \item a uniform treatment of discrete and continuous random variables,
      \item a clear separation between probabilistic dependencies and logical dependencies by extending the ideas of \citet{poole2010probabilistic} to the hybrid domain.
    \end{itemize}
\item \label{item:k3} We introduce \dcproblogsty, an expressive PLP language in the discrete-continuous domain,
which incorporates the{\em measure semantics}. 
 \dcproblogsty has standard discrete PLP, \eg ProbLog~\citep{fierens2015inference}, as a special case (unlike other hybrid PLP languages~\citep{gutmann2011magic,nitti2016probabilistic}).
    \item \label{item:k2}  We introduce a novel inference algorithm, {\em infinitesimal algebraic likelihood weighting} (IALW), for hybrid PLPs,
which extends the standard knowledge compilation approach used in PLP towards mixed discrete continuous distributions, and  which 
provides an operational semantics for hybrid PLP.
\end{enumerate}

In essence, our contributions ~\ref{item:k1} and ~\ref{item:k3} generalize both  Sato's distribution semantics and discrete PLP such that in the absence of random variables with infinite sample spaces we recover the \problogsty language and declarative semantics. It is noteworthy that our approach of disentangling probabilistic dependencies and logical ones, allows us to express more general distributions than state-of-the-art approaches such as~\citep{gutmann2011magic,nitti2016probabilistic,azzolini2021semantics}. 
Contribution \ref{item:k2} takes this generalization to the inference level: in the exclusive presence of finite random variables our IALW algorithm reduces to \problogsty's current inference algorithm~\citep{fierens2015inference}.

% Probabilistic logic programming (PLP) constitutes a well-studied field within computer science, with a rich tradition dating back to the early 1990s. In particular, \citet{dantsin1990probabilistic}, \citet{ng1992probabilistic}, and \citet{poole1993probabilistic} generalized earlier ideas of \citet{nilsson1986probabilistic} and \citet{pearl1988probabilistic} on probabilistic logic towards probabilistic logic programming. In \citeyear{sato1995statistical}, \citeauthor{sato1995statistical} then presented his seminal work on {\em distribution semantics} for probabilistic logic programs. All these early works on PLP restrict random variables to finite categorical (usually binary) random variables.

% An advantage of restricting a probabilistic programming language to finite random variables is the relative straightforwardness of defining a declarative semantics for programs, i.e giving meaning to programs regardless of the underlying inference algorithm used in the implementation of the language.
% In PLP, this is often done by falling back on \citeauthor{sato1995statistical}'s distribution semantics~\citep{sato1997prism,kersting2000bayesian,vennekens2004logic,de2007problog}.

% Defining declarative semantics for a language including random variables with infinite and possibly uncountable sample spaces is a much harder task. This can be observed when looking at the development of non-logic probabilistic programming languages~\citep{goodman2008church,mansinghka2014venture}, whose focus is on continuous random variables and which were only defined with regard to a specific Monte Carlo inference algorithm. Only follow-up works started tackling the issue of fully declarative semantics for non-logic probabilistic programming languages in discrete-continuous domains~\citep{staton2016semantics,wu2018discrete}.\footnote{Aside from some subtle differences (\cf~\citep{kimmig2017probabilistic}), in non-logical probabilistic programming languages declarative semantics are usually called denotational and procedural semantics are referred to as operational.}

% The PLP landscape has experienced similar struggles. First attempts of extending the traditionally finite sample space to discrete-continuous domains (also called hybrid) were achieved by restricting the language~\citep{gutmann2010extending,gutmann2011magic,islam2012inference} or via recourse to procedural semantics~\citep{nitti2016probabilistic}. This brings us to our first key contribution:

% \begin{enumerate}[label={\bf C\arabic*}]
%   \setcounter{enumi}{0}
% \item
% We introduce \dcproblogsty, a fully expressive PLP language in the discrete-continuous domain, which we equip with fully declarative semantics based on \citeauthor{sato1995statistical}'s distribution semantics. \dcproblogsty is a PLP language with:
%     \begin{itemize}
%         \item \label{item:k1} a countably infinite number of random variables
%         \item a uniform treatment of discrete and continuous random variables
%         \item a clear separation between probabilistic dependencies and logical dependencies by extending the ideas of \citet{poole2010probabilistic} to the hybrid domain
%         \item standard discrete PLP, \eg ProbLog~\citep{fierens2015inference}, as a special case (unlike other hybrid PLP languages~\citep{gutmann2011magic,nitti2016probabilistic}).
%     \end{itemize}
% \end{enumerate}
% Our second key contribution concerns extending probabilistic inference in discrete PLP, \eg \problogsty~\citep{fierens2015inference}, to discrete-continuous domains. 
% \begin{enumerate}[label={\bf C\arabic*}]
%   \setcounter{enumi}{1}
%     \item \label{item:k2} We present {\em infinitesimal algebraic likelihood weighting} (IALW), a novel inference algorithm that allows for the combination of knowledge compilation~\citep{darwiche2002knowledge} and sampling based approaches in hybrid domains. 
% \end{enumerate}

% In essence, our contribution ~\ref{item:k1} generalizes Sato's distribution semantics such that in the absence of random variables with infinite sample spaces we recover the \problogsty language.
% It is noteworthy that our approach of disentangling probabilistic dependencies and logical ones, allows us to express more general distributions than state-of-the-art approaches, \eg~\citep{gutmann2011magic,nitti2016probabilistic,azzolini2021semantics}. 
% Contribution \ref{item:k2} takes this generalization to the inference level: in the exclusive presence of finite random variables our ALW algorithm reduces to \problogsty's current inference algorithm~\citep{fierens2015inference}.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_main/panorama.tex
aij dcproblog paper

\section{A Panoramic Overview}
\label{sec:panorama}

Before diving into the technical details of the paper we first give a high-level overview of the \dcproblogsty language. This will also serve us as roadmap to the remainder of the paper.  
We will first introduce, by example, the \dcproblogsty language (Section~\ref{sec:panorama_semantics}). The formal syntax and semantics of which are discussed in Section~\ref{sec:semantics} and Section~\ref{sec:dcproblog}.
In Section~\ref{sec:panorama_inference} we demonstrate how to perform probabilistic inference in \dcproblogsty by translating a queried \dcproblogsty program to an algebraic circuit~\citep{zuidbergdosmartires2019transforming}. 
Before giving the details of this transformation in Section~\ref{sec:dc2smt} and Section~\ref{sec:alw}, we define conditional probability queries on \dcproblogsty programs (Section~\ref{sec:inference-tasks}).
The paper ends with a discussion on related work (Section~\ref{sec:related}) and concluding remarks in Section~\ref{sec:conclusions}.

Throughout the paper, we assume that the reader is familiar with basic concepts from logic programming and probability theory. We provide, however, a brief refresher of basic logic programming concepts in Appendix~\ref{app:lp_new}. In Appendix~\ref{app:table} we give a tabular overview of notations used, and in the remaining sections of the appendix we give proofs to propositions and theorems or discuss in more detail some of the more subtle technical issues.

\subsection{Panorama of the Syntax and Semantics}
\label{sec:panorama_semantics}

\begin{example}
    \label{example:sweets_dc}
    A shop owner creates random bags of  sweets with two independent random binary properties (\emph{large} and \emph{balanced}).  He first picks the number of red sweets from a Poisson distribution whose parameter is 20 if the bag is large and 10 otherwise, and then the number of yellow sweets from a Poisson whose parameter is the number of red sweets if the bag is balanced and twice that number otherwise. His favorite type of bag contains more than 15 red sweets and no less than 5 yellow ones. We model this in \dcproblogsty as follows:
\begin{problog*}{linenos}
0.5::large.
0.5::balanced.

red ~ poisson(20) :- large.
red ~ poisson(10) :- not large.

yellow ~ poisson(red) :- balanced.
yellow ~ poisson(2*red) :- not balanced.

favorite :- red > 15, not yellow < 5.
\end{problog*}

In the first two lines we encounter {\em probabilistic facts}, a well-known modelling construct in discrete PLP languages (\eg~\citep{de2007problog}). Probabilistic facts, written as logical facts labeled with a probability, express Boolean random variables that are true with the probability specified by the label. For instance, \probloginline{0.5::large} expresses that \probloginline{large} is true with probability \probloginline{0.5} and false with probability \probloginline{1-0.5}.

In Lines 4 to 8, we use {\em distributional clauses} (DCs); introduced by~\citet{gutmann2011magic} into the PLP literature. DCs are of the syntactical form \probloginline{v~d:-b} and define random variables \probloginline{v} that are distributed according to the distribution \probloginline{d}, given that \probloginline{b} is true.
For example, Line 4 specifies that when \probloginline{large} is true, \probloginline{red} is distributed according to a Poisson distribution. 
We call the left-hand argument of a \predicate{~}{2} predicate in infix notation a {\em random term}. The random terms in the program above are \probloginline{red} and \probloginline{yellow}.

Note how random terms reappear in three distinct places in the \dcproblogsty program. First, we can use them as parameters to other distributions, \eg \probloginline{yellow ~ poisson(red)}. 
Second, we can use them within arithmetic expression, such as \probloginline{2*red} in Line 8.
Third, we can use them in comparison atoms (\probloginline{red>15}) in Line 10. The comparison atoms appear in the bodies of logical rules that express logical consequences of probabilistic event, for example having more than 15 red sweets and less than 5 yellow ones.
\end{example}

Probabilistic facts and distributional clauses are the main modelling constructs to define random variables in probabilistic logic programs.
As they are considered to be fundamental building blocks of a PLP language, the semantics of a language are defined in function of these syntactical constructs (\cf.~\citep{fierens2015inference,gutmann2011magic}). 
We now make an important observation: probabilistic facts and distributional clauses can be deconstructed into a much more fundamental concept, which we call the {\em distributional fact}.  
Syntactically, a distributional fact is of the form \probloginline{v~d}. That is, a distributional clause with an empty body.
As a consequence, probabilistic facts and distributional clauses do not constitute fundamental concepts in PLP but are merely special cases, \ie while helpful for writing concise programs, they are only of secondary importance when it comes to semantics.

\begin{example}\label{ex:sweets_df}
We now rewrite the program in Example~\ref{example:sweets_dc} using distributional facts only. Note how probabilistic facts are actually syntactic \fixed{sugar for distributional facts}. The random variable is now distributed according to a Bernoulli distribution (\probloginline{flip}) and the atom of the probabilistic fact is the head of a rule with a probabilistic comparison in its body (\eg Lines 1 and 2 in the program below). Rewriting distributional facts is more involved. The main idea is to introduce a distinct random term for each distributional clause. Take for example, the random term \probloginline{red} in Example~\ref{example:sweets_dc}. This random term encodes, in fact, two distinct random variables, which we denote in the program below \probloginline{red_large} and \probloginline{red_small}. We now have to propagate this rewrite through the program and replace every occurrence of \probloginline{red} with \probloginline{red_large} and \probloginline{red_small}. This is why we get instead of two distributional clauses for \probloginline{yellow}, four distributional facts. It explains also why we get instead of one rule for \probloginline{favorite} in Example~\ref{example:sweets_dc} four rules now.

    \begin{problog*}{linenos}
rv_large ~ flip(0.5).
large :- rv_large=:=1.
rv_balanced ~ flip(0.5).
balanced :- rv_balanced=:=1.

red_large ~ poisson(20).
red_small ~ poisson(10).

yellow_large_balanced ~ poisson(red_large).
yellow_large_unbalanced ~ poisson(2*red_large).
yellow_small_balanced ~ poisson(red_small).
yellow_small_unbalanced ~ poisson(2*red_small).

favorite :- large, red_large > 15, 
              balanced, not yellow_large_balanced < 5.
favorite :- large, red_large > 15, 
              not balanced, not yellow_large_unbalanced < 5.
favorite :- not large, red_small > 15, 
              balanced, not yellow_small_balanced < 5.
favorite :- not large, red_small > 15, 
              not balanced, not yellow_small_unbalanced < 5.
\end{problog*}
\end{example}

The advantage of using probabilistic facts and distributional clauses is clear. They allow us to write much more compact and readable programs. However, as they do not really constitute fundamental building blocks of PLP, defining the semantics of a PLP language is much more intricate. For this reason we adapt a two-stage approach to define the semantics of \dcproblogsty. We first define the semantics of \dfplpsty (distributional fact PLP), a bare-bones language with no syntactic sugar only relying on distributional facts to define random variables. This happens in Section~\ref{sec:semantics}. During the second stage we define the program transformations to rewrite syntactic sugar (\eg distributional clauses) as distributional facts. The semantics of \dfplpsty and the program transformations then give us the \dcproblogsty language (\cf Section~\ref{sec:dcproblog}).

\subsection{Panorama of the Inference}
\label{sec:panorama_inference}

The part of the paper concerning inference consists of three sections. First, we start in Section~\ref{sec:inference-tasks} to define a query to a \dcproblogsty program.
For instance, we might be interested in the probability
$$
P(\mathprobloginline{favorite}=\top, \mathprobloginline{large}=\bot)
$$
In other words, the joint probability of \probloginline{favorite} being true and \probloginline{large} being false. While the example query above is simply a joint probability, we generalize this in Section~\ref{sec:inference-tasks} to conditional probabilities (possible with zero-probability events in the conditioning set).

Second, we map the queried ground program to a labeled Boolean formula. Contrary to the approach taken by~\citet{fierens2015inference} the labels are not probabilities (as usual in PLP) but indicator functions. This mapping to a labeled Boolean formula happens again in a series of program transformations, which we describe in Section~\ref{sec:dc2smt}. One of these steps is obtaining the relevant ground program to a query. For example, for the query above 
only the last two rules for  \probloginline{favorite} matter.
\begin{problog*}{}
favorite :- not large, rs > 15, balanced, not ysb < 5.
favorite :- not large, rs > 15, not balanced, not ysu < 5.
\end{problog*}
Here, we abbreviated \probloginline{red_small} as \probloginline{rs} and \probloginline{yellow_small_balanced} and \probloginline{yellow_small_unbalanced} as \probloginline{ysb} and \probloginline{ysu}, respectively. We can further rewrite these rules by replacing \probloginline{large} and \probloginline{balanced} with equivalent comparison atoms and pushing the negation into the comparisons:
\begin{problog*}{}
favorite :- rvl=:=0, rs > 15, rvb=:=1, ysb >= 5.
favorite :- rvl=:=0, rs > 15, rvb=:=0, ysu >= 5.
\end{problog*}
Again using abbreviations: \probloginline{rvl} for \probloginline{rv_large} and \probloginline{rvb} for \probloginline{rv_balanced}.

In Section~\ref{sec:alw} we then show how to compute the expected value of the labeled propositional Boolean formula corresponding to these rules by compiling it into an algebraic circuit, which is graphically depicted in Figure~\ref{fig:circuit:panorama}. In order to evaluate this circuit and obtain the queried probability (expected value), we introduce the IALW algorithm.

The idea of IALW is the following: sample the random variables dangling at the bottom of the circuit by sampling parents before children. For instance, we first sample from $\mathprobloginline{poisson}(10)$ (at the very bottom) before sampling from $\mathprobloginline{poisson}(rs)$ using the sampled value of the parent as the parameter of the child. Once we reach the comparison atoms (\eg $ysb\geq5$) we stick in the sampled values for the mentioned random variables. This evaluates the comparisons to either $1$ or $0$, for which we then perform the sums and products as prescribed by the circuit. We get a Monte Carlo estimate of the queried probability by averaging over multiple such evaluations of the circuit.

The method, as sketched here, is in essence the probabilistic inference algorithm Sampo presented in~\citep{zuidbergdosmartires2019exact}. The key contribution of IALW, which we discuss in Section~\ref{sec:alw}, is to extend Sampo such that conditional inference with zero probability events is performed correctly.

\begin{figure}
	\resizebox{\linewidth}{!}{%

		\tikzstyle{distribution}=[rectangle, text centered, fill=white, draw, dashed,thick]
		\tikzstyle{leaf}=[rectangle, text centered, fill=gray!10, draw,thick]

		\tikzstyle{negate}=[
			rectangle split,
			rectangle split parts=3, 
			rectangle split horizontal,
			text centered,
			rectangle split part fill={gray!10,white,gray!10},
			draw,
			rectangle split draw splits=false,
			anchor=center,
			align=center,
			thick
		]
		\newcommand{\minus}{  ${\bm e^\otimes}$ \nodepart{second} ${{\bm \ominus}}$ \nodepart{third}  \phantom{${\bm e^\otimes}$}}

		\tikzstyle{sumproduct}=[
			rectangle split,
			rectangle split parts=3, 
			rectangle split horizontal,
			text centered,
			rectangle split part fill={gray!10,white,gray!10},
			draw,
			rectangle split draw splits=false,
			anchor=center,
			align=center,
			thick
		]
		\newcommand{\supr}[1]{  \phantom{${\bm e^\otimes}$} \nodepart{second} ${{\bm #1}}$ \nodepart{third} \phantom{${\bm e^\otimes}$}}

		\tikzstyle{circuitedge}=[ultra thick, thick,->]
		\tikzstyle{distributionedge}=[thick,->, dashed, in=-90]

		\tikzstyle{indexnode}=[draw,circle, inner sep=1pt]

		\begin{tikzpicture}[remember picture]

			\node[sumproduct] (r) at (200.54bp,378.0bp) {\supr{\otimes}};
			\draw[ultra thick, thick,->] (r.90) to  ([shift={(0,1)}]r.90);

			\node[sumproduct] (l1) [below left = of r] {\supr{\oplus}};
			\node[sumproduct] (r1) [below right = of r,  yshift=-5.5cm]  {\supr{\otimes}};

			\node[leaf] (r21) [below right=of r1,  xshift=-1.3cm] {$\subnode{var_rvl}{rvl}=0$};
			\node[leaf] (r22) [below left=of r1, xshift=1.3cm] {$\subnode{var_rs}{rs}>15$};

			\node[sumproduct] (l21) [below left=of l1, xshift=0.5cm] {\supr{\otimes}};
			\node[sumproduct] (l22) [below right=of l1, xshift=-0.5cm] {\supr{\otimes}};

			\node[leaf] (l31) [below right=of l21,  xshift=-1.5cm] {$\subnode{var_rvb}{rvb}=1$};
			\node[leaf] (l32) [below left=of l21,  xshift=1.5cm] {$\subnode{var_ysb}{ysb}\geq5$};

			\node[distribution] (rvb)  [below=of l1,  yshift=-3.5cm,xshift=-0.5cm] {$\mathprobloginline{flip}(0.5)$};
			\node[distribution] (ysb)  [left=of rvb, xshift=0.4cm ]{$\mathprobloginline{poisson}(\subnode{poisson_rs_ysb}{rs})$};
			\node[distribution] (ysu)  [right=of rvb, xshift=-0.4cm] {$\mathprobloginline{poisson}(2\times \subnode{poisson_rs_ysu}{rs})$};

			\node[leaf] (l33) [below left=of l22,  xshift=1.5cm] {$\subnode{var_rvbbis}{rvb}=0$};
			\node[leaf] (l34) [below right=of l22,  xshift=-1.5cm] {$\subnode{var_ysu}{ysu}\geq5$};

			\node[distribution] (rvl)  [below=of r21] {$\mathprobloginline{flip}(0.5)$};
			\node[distribution] (rs)  [below=of rvb, yshift=-2cm] {$\mathprobloginline{poisson}(10)$};

			\draw[circuitedge] (l1) to  (r.mid);
			\draw[circuitedge] (r1) to  (r.three south |- r.mid);

			\draw[circuitedge] (l21) to  (l1.mid);
			\draw[circuitedge] (l22) to  (l1.three south |- l1.mid);

			\draw[circuitedge] (l21) to  (l1.mid);
			\draw[circuitedge] (l22) to  (l1.three south |- l1.mid);

			\draw[circuitedge] (l32) to  (l21.mid);
			\draw[circuitedge] (l31) to  (l21.three south |- l21.mid);

			\draw[circuitedge] (l33) to  (l22.mid);
			\draw[circuitedge] (l34) to  (l22.three south |- l22.mid);

			\draw[circuitedge] (r22) to  (r1.mid);
			\draw[circuitedge] (r21) to  (r1.three south |- r1.mid);

			\draw[distributionedge,out=90]  (ysb) to (var_ysb);
			\draw[distributionedge,out=90]  (ysu) to (var_ysu);
			\draw[distributionedge,out=90]  (rvb) to (var_rvb);
			\draw[distributionedge,out=90]  (rvb) to (var_rvbbis);

			\draw[distributionedge,out=90]  (rs) to (poisson_rs_ysb);
			\draw[distributionedge,out=90]  (rs) to (poisson_rs_ysu);
			\draw[distributionedge,out=90]  (rs) to (var_rs);

			\draw[distributionedge,out=90]  (rvl) to (var_rvl);

			% \draw[circuitedge] (9) to  (14.mid);
			% \draw[circuitedge] (13) to  (14.three south |- 14.mid);

			% \node[leaf] (l41) [below left=of l31,  xshift=1.3cm] {$\subnode{var_l_rv}{rv\_b}=0$};
			% \node[leaf] (l42) [below right=of l32, xshift=-1.3cm] {$\subnode{var_rs}{rs}>15$};

			% \node[sumproduct] (12)  [below  = of 13] {\supr{\oplus}};

			% \node[sumproduct] (8)  [below=of 12]  {\supr{\otimes}};

			% \node[indexnode, left=of 14, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$6$}};			
			% \node[indexnode, left=of 13, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$5$}};			
			% \node[indexnode, left=of 8, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$2$}};			
			% \node[indexnode, left=of 12, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$4$}};			
			% \node[indexnode, left=of 9, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$3$}};			
			% \node[indexnode, left=of m1, xshift=0.9cm, yshift=0.4cm] {\footnotesize {$1$}};

			% \node[leaf, below= of 8, xshift=-0.2cm] (4)   {$\subnode{var_m0}{$m$}=1$};
			% \node[leaf, below= of 8, xshift=1.8cm] (2) {$\subnode{var_m1}{$m$}=0$};				
			% \node[leaf] (size1obs) [left=of 4, xshift=0cm]  {$\subnode{var_s_11_un}{size_{1}(1)} \doteq 0.4$};
			% \node[leaf] (size0obs) [left=of 4, xshift=-4cm] {$\subnode{var_s_01_un}{size_{0}(1)} \doteq 0.4$};

			% \node[distribution] (size0)  [below=of size0obs] {\probloginline{beta(2,3)}};
			% \node[distribution] (size1)  [below=of size1obs] {\probloginline{beta(4,2)}};
			% \node[distribution] (flip)  [below= of 4, xshift=1cm] {\probloginline{flip(0.3)}};

			% \draw[distributionedge,out=90]  (size1) to (var_s_11_un);
			% \draw[distributionedge,out=90] (size0) to  (var_s_01_un);
			% \draw[distributionedge,out=160] (flip) to  (var_m0);
			% \draw[distributionedge] (flip) to  (var_m1);

			% %https://tex.stackexchange.com/questions/447989/anchor-node-names-for-tikz-rectangle-split-horizontal			
			% \draw[circuitedge] (9) to  (14.mid);
			% \draw[circuitedge] (13) to  (14.three south |- 14.mid);
			% \draw[circuitedge] (m1) to  (9.mid);
			% \draw[circuitedge] (12) to  (13.three south |- 13.mid);
			% \draw[circuitedge] (size0obs) to   (m1.three south |- m1.mid);	
			% \draw[circuitedge] (2) to  (12.three south |- 12.mid);				
			% \draw[circuitedge] (size1obs) to   (8.mid);
			% \draw[circuitedge] (4) to  (8.three south |- 8.mid);	
			% \draw[circuitedge] (8) to  (9.three south |- 9.mid);	
			% \draw[circuitedge] (8) to   (12.mid);
			% \draw[circuitedge] (size0obs) to  (13.mid);

		\end{tikzpicture}
	}
    \captionof{figure}{Graphical representation of the computation graph (\ie algebraic circuit) used to compute the  probability $(\mathprobloginline{favorite}=\top, \mathprobloginline{large}=\bot)$ using the IALW algorithm introduced in Section~\ref{sec:alw}.}
    \label{fig:circuit:panorama}
\end{figure}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_main/related_work.tex
aij dcproblog paper

\section{\dcproblogsty and the Probabilistic Programming Landscape}\label{sec:related}

In recent years a plethora of different probabilistic programming languages have been developed. 
We discuss these by pointing out key features present in \dcproblogsty (listed below), which are missing in specific related works.
We organize these features along the three key contributions stated in Section \ref{sec:introduction}. Our first key contribution is the introduction of the measure semantics with the following features:

\begin{enumerate}[label=C1.\arabic*, leftmargin=2\parindent]
    \item \new{possibly infinite number (even uncountable) number of random variables} 
    \label{f:infinite_number_of rv}
    \item random variables with (possibly) infinite sample spaces
    \label{f:infinite_samplespaces}
    \item functional dependencies between random variables
    \label{f:rv_dependencies}
    \item uniform treatment of discrete and continuous random variables
    \label{f:dc_rv_uniform}
    \item negation
    \label{f:declarative_negation}
\end{enumerate}

\noindent Our second contribution is the  introduction of  the \dcproblogsty language, which  
\begin{enumerate}[label=C2.\arabic*, leftmargin=2\parindent]
    \item has purely discrete PLPs and their semantics as a special case,
    \label{f:discrete_special}
    \item supports a rich set of comparison predicates, and
    \label{f:comparison_predicates}
    \item is a Turing complete language (\dcplpsty)
    \label{f:turing_complete}
\end{enumerate}

\noindent Our last contributions concern inference, which includes 
\begin{enumerate}[label=C3.\arabic*, leftmargin=2\parindent]
    \item a formal definition of the hybrid probabilistic inference task,
    \label{f:definition_inference}
    \item an inference algorithm called IALW, 
    \label{f:inference_algorithm}
    \item that uses standard knowledge compilation in the hybrid domain.
    \label{f:inference_kc}
\end{enumerate}

% Our focus will lie with probabilistic logic programming languages, and we only touch lightly upon other paradigms of probabilistic programming
\subsection{\problogsty and \dcsty}

The \dcproblogsty language is a generalization of \problogsty, both in terms of syntax and semantics. A \dcproblogsty program that does not use distributional clauses (or distributional facts) is also a \problogsty program, and both define the same distribution over the logical vocabulary of the program. \dcproblogsty properly generalizes \problogsty to include random variables with infinite sample spaces (\ref{f:infinite_samplespaces}).

On a syntactical level, \dcproblogsty is closely related to the \dcsty (DC) language, with which it shares the  \predicate{~}{2} predicate used in infix notation. 
In Appendix~\ref{sec:dcproblog-dc} we discuss in more detail the relationship between \dcproblogsty and the \dcsty language. Concretely, we point out that 
\dcproblogsty generalizes the original and negation-free version of DC~\citep{gutmann2011magic} (\ref{f:declarative_negation}).
However, \dcproblogsty differs in its declarative interpretation of negation from the procedural interpretation as introduced to DC by~\citet{nitti2016probabilistic}.
As a consequence, the semantics of DC and \problogsty differ in the absence of continuous random variables, while \dcproblogsty is a strict generalization of \problogsty (\ref{f:discrete_special}).

\subsection{Bayesian Logic Programs}
\label{sec:blp}

\new{
Bayesian logic programs (BLPs)~\citep{kersting2000bayesian} can be seen as a special case of \dcplpsty: while the semantics of \dcplpsty allow for a (possibly uncountable) infinite number of random variables~\ref{f:infinite_number_of rv}, BLPs are limited to finite distributional databases (expressed as Bayesian networks).

Moreover, using the construct of distributional clauses we introduce syntax to interleave logic programming statements and the declaration of the distributional database. This is not supported in the BLP language.
Lastly, IAWL equips \dcproblogsty with a sound inference algorithm for the discrete-continuous space. This is again in contrast to BLP's inference algorithm that only handles (discrete mixtures of) continuous random variables. 
}

\subsection{\extendedprismsty}

An early attempt of equipping a probabilistic logic programming language with continuous random variables can be found in~\citep{islam2012inference}, which was dubbed {\em \extendedprismsty}. 
Similar to \dcproblogsty, \extendedprismsty's semantics are based again on \citeauthor{sato1995statistical}'s distribution semantics.
However, \extendedprismsty assumes, just like \dcsty , pairwise mutually exclusive proofs (we refer again to Appendix~\ref{sec:dcproblog-dc} for details on this).
On the expressivity side, \extendedprismsty only supports linear equalities -- in contrast to \dcproblogsty, where also inequalities are included in the semantics of the language (\ref{f:comparison_predicates}). 
An advantage of restricting possible constraints to equalities is the possibility of performing exact symbolic inference. In this regard, \extendedprismsty, together with its symbolic inference algorithm, can be viewed as a logic programming language that has access to a computer algebra system. Swapping out the approximate Sampo-inspired inference algorithm in \dcproblogsty by an exact inference algorithm using symbolic expression manipulations would result in an inference approach closely related to that of \extendedprismsty. One possibility would be to use the Symbo algorithm presented in~\citep{zuidbergdosmartires2019exact}, which uses the PSI-language~\citep{gehr2016psi} as its (probabilistic)  computer algebra system.

\subsection{Probabilistic Constraint Logic Programming}

Impressive work on extending probabilistic logic programs with continuous random variables was presented by~\citet{michels2015new} with the introduction of Probabilistic Constraint Logic Programming (PCLP). The semantics of PCLP are again based on \citeauthor{sato1995statistical}'s distribution semantics and the authors also presented an approximate inference algorithm for hybrid probabilistic logic programs.
Interestingly, the algorithm presented in~\citep{michels2015new} to perform (conditional) probabilistic inference extends weighted model counting to continuous random variables using imprecise probabilities, and more specifically credal sets.

A shortcoming of PCLP's semantics is the lack of direct support for generative definitions of random variables, \ie, random variables can only be interpreted within constraints, but not within distributions of other random variables as is possible in \dcproblogsty (\ref{f:rv_dependencies}).
\citet{azzolini2021semantics} define a non-credal version of this semantics using a product measure over a space that explicitly separates discrete and continuous random variables, assuming that a measure over the latter is given as part of the input without further discussion of how this part of the measure is specified in a program. Furthermore, they do not define any inference tasks (\ref{f:definition_inference}), \eg computing conditional probabilities (\cf Section~\ref{sec:inference-tasks}), nor
do they provide an inference algorithm (\ref{f:inference_algorithm}).

A later proposal for the syntax of such programs~\citep{azzolini:iclp21} combines two classes of terms (logical and continuous ones) with typed predicates and functors, and defines mixture variables as well as arithmetic expressions over random variables through logical clauses. In other words, user-defined predicates define families of random variables through the use of typed arguments of the predicate identifying a specific random variable, arguments providing parameters for the distribution, and one argument representing the random variable itself.
In contrast, the syntax of \dcproblogsty clearly identifies all  random variables through explicit terms introduced through distributional facts or distributional clauses, explicitly exposes the probabilistic dependency structure by using random variable terms inside distribution terms, and avoids typing through argument positions.
Moreover, \dcproblogsty takes a uniform view on all random variables in terms of semantics, thereby avoiding treating discrete and continuous random variables separately (\ref{f:dc_rv_uniform}).

\subsection{\blogsty}

Notable in the domain of probabilistic logic programming is also the BLOG language~\citep{milch2005blog,wu2018discrete}. Contrary to the aforementioned probabilistic logic programming languages, BLOG's semantics are not specified using \citeauthor{sato1995statistical}'s distribution semantics but via so-called {\em measure-theoretic Bayesian networks} (MTBN), which were introduced in~\citep{wu2018discrete}. MTBNs can be regarded as the assembly language for BLOG: every BLOG program is translated or compiled to an MTBN.
With \dcproblogsty we follow a similar pattern: every \dcproblogsty program with syntactic sugar (\eg annotated disjunctions)  is transformed into \dfplpsty program. The semantics are defined on the bare-bones program. Note that the assembly language for \dcproblogsty (\dfplpsty) is Turing complete. This is not the case for MTBNs (\ref{f:turing_complete}).

% On the practical side, the default inference algorithm deployed by BLOG is similar in spirit to IALW inference  but does also provide alternatives, such as Metropolis-Hastings MCMC~\citep{milch2006general}. An important difference of \blogsty's inference algorithms with regard to \dcproblogsys is the absence of knowledge compilation. BLOG will, hence, struggle with problems that exhibit rich combinatorial structures, as shown by~\citet{zuidbergdosmartires2019exact}.

\subsection{Non-logical Probabilistic Programming}

As first pointed out by~\citet{russell2015unifying} and later on elaborated upon by~\citet{kimmig2017probabilistic}, probabilistic programs fall either into the {\em possible worlds semantics} category or the {\em probabilistic execution traces semantics} category. The former is usually found in logic based languages, while the latter is the prevailing view in imperative and functional probabilistic languages.

While, the probabilistic programming languages discussed so far follow the possible worlds paradigm,
many languages follow the execution traces paradigm, either as a probabilistic functional language~\citep{goodman2008church,wood2014approach} or as a imperative probabilistic language~\citep{gehr2016psi,salvatier2016probabilistic,carpenter2017stan,bingham2019pyro,ge2018turing}. Generally speaking, functional and imperative probabilistic programming languages target first and foremost continuous random variables, and discrete random variables are only added as an afterthought. A notable exception is the functional probabilistic programming language Dice~\citep{holtzen2020dice}, which targets discrete random variables exclusively.

Concerning inference in probabilistic programming, we can observe general trends in logical and non-logical probabilistic languages. While the latter are interested in adapting and speeding up approximate inference algorithms, such as Markov chain Monte Carlo sampling schemes or variational inference, the former type of languages are more invested in exploiting independences in the probabilistic programs, mainly by means of knowledge compilation. Clearly, these trends are not strict. For instance, \citet{obermeyer2019functional} proposed so-called {\em funsors} to express and exploit independences in \pyrosty~\citep{bingham2019pyro}, an imperative probabilistic programming language, and \citet{gehr2016psi} developed a computer algebra system to perform exact symbolic probabilistic inference.

\subsection{Representation of Probabilistic Programs at Inference Time}

Lastly, we would like to point out a key feature of the IALW inference algorithm that sets it apart from any other inference scheme for probabilistic programming in the hybrid domain. But first, let us briefly talk about computing probabilities in probabilistic programming. Roughly speaking, probabilities are computed summing and multiplying weights. These can for example be represented as floating point numbers or symbolic expressions. The collection of all operations that were performed to obtain the probability of a query to a program is called the computation graph. Now, the big difference between IALW and other inference algorithms lies in the structure of the computation graph. IALW represents the computation graph as a directed acyclic graph (DAG), while all other languages, except some purely discrete languages~\citep{fierens2015inference,holtzen2020dice}, use a tree representation. IALW is the first inference algorithm in the discrete-continuous domain that uses DAGs (\ref{f:inference_kc}). In cases where the computation graph can be represented as a DAG the size of the representation might be exponentially smaller compared to tree representations, which leads to faster inference times.

Note that \citet{gutmann2010extending} and more recently~\citet{saad2021sppl} presented implementations of hybrid languages where the inference algorithm leverages directed acyclic graphs, as well. However, the constraints that may be imposed on random variables are limited to univariate equalities and inequalities. In the weighted model integration literature it was shown that such probability computations can be mapped to probability computations of discrete random variables only~\citep{zeng2019efficient}.

\subsection{Probabilistic Neurosymbolic AI}

\new{
As noted by \Citet{desmet2023neural} a shortcoming of many neurosymbolic AI systems~\citep{garcez2019neural,marra2024statistical}, \ie systems that combine the function approximation power of neural networks with logic reasoning, is their restriction to only allowing discrete random variables.
Based on the semantics of \dcplpsty, \citet{desmet2023neural} extended distributional facts to so-called neural-distributional facts. 
That is, they allowed for neural networks to estimate the parameters of the distribution in the distributional fact.
Importantly, they showed that endowing a neurosymbolic system in the discrete-continuous domain with proper probabilistic semantics is advantageous when comparing to systems that exhibit, for instance, a fuzzy logic semantics~\citep{badreddine2022logic}.
}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_main/semantics.tex
aij dcproblog paper

\section{\dfplpsty}\label{sec:semantics}

Sato's distribution semantics~\citep{sato1995statistical} start from a probability measure over a countable set of facts $\comparisonfacts$, the so-called \emph{basic distribution}, which he then extends this to a probability measure over the Herbrand interpretations of the full program.
It is worth noting that the basic distribution is defined independently of the logical rules and that the random variables are mutually independent.

In our case, the set $\comparisonfacts$ consists of ground Boolean comparison atoms over the random variables, for which we drop the mutual independence assumption. These comparison atoms form an interface between the random variables (represented as terms) and the logical layer (clauses) that reasons about truth values of atoms.
\fixed{This is inspired by the work of \citet{gutmann2011magic} and \citet{nitti2016probabilistic} on the \dcsty language. We discuss this relationship in more detail in the related work Section~\ref{sec:related} and in Appendix~\ref{sec:dcproblog-dc}.
}
% While~\citet{gutmann2011magic} used the same principle to define the distribution semantics for \dcsty, they did not support negation. \citep{nitti2016probabilistic} extended the fixed point semantics for hybrid probabilistic logic programs (also introduced by \citet{gutmann2011magic}) to stratified programs with negation . However, by doing so \citet{nitti2016probabilistic} introduced a procedural element to the semantics.

In this section we introduce the syntax and declarative semantics of \dfplpsty -- a probabilistic programming language with a minimal set of built-in predicates and functors. 
We do this in three steps.
Firstly, we discuss distributional facts and the probability measure over random variables they define (Section~\ref{sec:dist_facts_rand_var}).
Secondly, we introduce the Boolean comparison atoms that form the interface layer between random variables and a logic program (Section~\ref{sec:boolean_comparison_atoms}).
Thirdly, we add the logic program itself (Section~\ref{sec:log_cons_bool_comp}).
Fourht, we discuss practical considerations for constructing sets of distributional facts (Section~\ref{sec:fintiedistdb}).
An overview table of the notation related to semantics is provided in Appendix~\ref{app:table}.

\begin{definition}[Reserved Vocabulary]
    \label{def:reserved_vocabulary}
We use the following {\em reserved} vocabulary (built-ins), whose intended meaning is fixed across programs:
\begin{itemize}
    \item a finite set $\distributionfunctors$ of \emph{distribution functors}.
    \item a finite set $\arithmeticfunctors$ of \emph{arithmetic functors}.
    % \footnote{
    %     As logic programming generally operates on a countable universe, we restrict the arithmetic functors to the rational numbers (constructed from the natural numbers and appropriate functors). Similarly to implementations of \prologsty, an implementation of \dcproblogsty can support integers and floating point numbers using external arithmetic engines~\cite[Chapter 9]{sterling1994art}. We will use those in our examples for ease of readability.
    % }
    \item A finite set $\comparisonpredicates$ of binary \emph{comparison predicates}, 
    \item the binary predicate \predicate{~}{2} (in infix notation).
\end{itemize}
\end{definition}

Examples of distribution functors that we have already seen in Section~\ref{sec:panorama} are \functor{poisson}{1} and \functor{flip}{1} but may also include further functors such as \functor{normal}{2} to denote normal distributions. Possible arithmetic functors are \functor{*}{2} (\cf Example~\ref{example:sweets_dc}) but also \functor{max}{2}, \functor{+}{2}, \functor{abs}{1}, \etc.
Binary comparison predicates (in Prolog syntax and infix notation) are \predicate{<}{2}, \predicate{>}{2}, \predicate{=<}{2}, \predicate{>=}{2}, \predicate{=:=}{2}, \predicate{=\=}{2}.
The precise definitions of $\distributionfunctors$, $\arithmeticfunctors$ and $\comparisonpredicates$ are left to system designers implementing the language.

\begin{definition}[Regular Vocabulary] \label{def:regular_vocabulary}
    We call an atom $\mu(\rho_1, \dots, \rho_k)$ whose predicate \mathfunctor{\mu}{k} is not part of the reserved vocabulary a \emph{regular} atom. The set of all regular atoms constitutes the regular vocabulary. 
\end{definition}

As a  brief comment on notation: in the remainder of the paper we will usually denote logic program expressions in teletype font (\eg \probloginline{4>x}) when giving examples. When defining new concepts or stating theorems and propositions, we will use the Greek alphabet.

\subsection{Distributional Facts and Random Variables}\label{sec:dist_facts_rand_var}

\begin{definition}[Distributional Fact]\label{def:distfact}
A distributional fact is of the form $\nu \sim \delta$, with $\nu$ a regular ground term, and $\delta$ a ground term whose functor is in $\distributionfunctors$. 
The distributional fact states that the ground term $\nu$ is interpreted as a random variable distributed according to $\delta$.
\end{definition}

\begin{definition}[Sample Space]\label{def:samplespace}
    Let $\nu$ be be a random variable distributed according to $\delta$. The set of possible samples (or values) for $\nu$ is the sample space denoted by $\samplespace_\nu$ and which is determined by $\delta$. We denote a sample from $\samplespace_\nu$ by $\samplefunction(\nu)$, where $\samplefunction$ is the {\em sampling} or {\em value} function. 
\end{definition}

\begin{definition}[Distributional Database]\label{def:distDB}
A \emph{distributional database} is a \new{(not necessarily countable)} set $\distdb=\{\nu_1 \sim \delta_1,\nu_2 \sim \delta_2, \ldots \}$ of distributional facts, with distinct $\nu_i$. We let $\randomvariableset= \{\nu_1, \nu_2, \dots \}$ denote the set of random variables.
\end{definition}

\begin{example}\label{ex:dist_db} 
The following distributional database encodes a Bayesian network with  normally distributed random variables, two of which serve as parameters in the distribution of another one. We thus have $\randomvariableset= \{ \text{\probloginline{x}}, \text{\probloginline{y}},\text{\probloginline{z}} \}$.
\begin{problog*}{linenos}
% distributional facts $\distdb$
x ~ normal(5,2).
y ~ normal(x,7).
z ~ normal(y,1).
\end{problog*}
\end{example}

\new{
\begin{definition}[Well-Defined Distributional Database]
    \label{def:well-distdb}
    We call a distributional database $\distdb$ well-defined if and only if the product of probability spaces induced by the random variables in $\distdb$ is unique. We denote this product probability space by
    $\probabilityspace_\distdb=(\samplespace_\distdb, \sigmaalgebra_\distdb, \probabilitymeasure_\distdb)$, where $\samplespace_\distdb$ is the sample space, $\sigmaalgebra_\distdb$ the sigma-algebra, and  $\probabilitymeasure_\distdb$ the probability measure. 
\end{definition}

\new{
Note that in Definition~\ref{def:distDB} we do not impose the restriction of having a countable set of random variables. This allows for modelling a rich class of probabilistic models, including random processes that are described via uncountable sets of random variables. We discuss the construction of well-defined databases in Section~\ref{sec:fintiedistdb}.
}

}

% \begin{restatable}{proposition}{proppv}
% 	\label{prop:pv}
% A well-defined distributional database~$\distdb$  defines a unique probability measure $\measurerandomvariableset$ on value assignments $\samplefunction(\randomvariableset)$.
% \end{restatable}

% \begin{proof}
%     See Appendix~\ref{app:proof:pv}.
% \end{proof}

%%%%%% comparison atom layer %%%%%%%%%%

\subsection{Boolean Comparison Atoms over Random Variables} \label{sec:boolean_comparison_atoms}

Starting from a well-defined distributional database, we now introduce the \new{probability space $\probabilityspace_\comparisonfacts$ induced by Boolean comparison atoms}, which corresponds to the basic (discrete) distribution in Sato's distribution semantics.

\begin{definition}[Boolean Comparison Atoms]\label{def:comparison-atoms-set}
Let $\distdb$ be a well-defined distributional database. 
A binary \emph{comparison atom} $\gamma_1 {\bowtie} \gamma_2$ over $\distdb$ is a ground atom with predicate $\bowtie \in \comparisonpredicates$. The ground terms $\gamma_1$  and $\gamma_2$ are either random variables in $\randomvariableset$ or terms whose functor is in $\arithmeticfunctors$.
We denote by $\comparisonfacts$ a countable set of \new{$\probabilitymeasure_\distdb$-measurable} 
Boolean comparison atoms over $\distdb$.
\end{definition}

\begin{example}
    Examples of Boolean comparison atoms over the distributional database of Example~\ref{ex:dist_db} include   \probloginline{z>10},  \probloginline{x<y}, \probloginline{abs(x-y)=<1},  and \probloginline{7*x=:=y+5}. 
\end{example}

\new{
\begin{restatable}{proposition}{propomegaf}
    \label{prop:omegaf}
The Boolean comparison atoms $\comparisonfacts$ induce a product sample space $\samplespace_\comparisonfacts$.
\end{restatable}

\begin{proof}
    See Appendix~\ref{app:proof:omegaf}.
\end{proof}

}

\new{

\begin{restatable}{proposition}{proppfsigma}
    \label{prop:pfsigma}
    The Boolean comparison atoms $\comparisonfacts$ induce a sigma-algebra $\sigmaalgebra_\comparisonfacts{\subseteq}\sigmaalgebra_\distdb$.
\end{restatable}

\begin{proof}
    See Appendix~\ref{app:proof:pfsigma}.
\end{proof}

}

\new{

\begin{restatable}{proposition}{proppf}
    \label{prop:pf}
Let $\distdb$ be a well-defined distributional database, the function $\probabilitymeasure_\comparisonfacts$ defined via
$
\probabilitymeasure_\comparisonfacts(A)
=
\frac
{
    \probabilitymeasure_\distdb(A)
}
{
    \probabilitymeasure_\distdb(\samplespace_\comparisonfacts)
}
$ defines a unique probability measure over the sample space $\samplespace_\comparisonfacts$ and the sigma algebra $\sigmaalgebra_\comparisonfacts$.
\end{restatable}

\begin{proof}
    See Appendix~\ref{app:proof:pf}.
\end{proof}

}

\new{

\begin{proposition}
    The triplet $\probabilityspace_\comparisonfacts= (\samplespace_\comparisonfacts, \sigmaalgebra_\comparisonfacts, \probabilitymeasure_\comparisonfacts)$ forms a probability space.
\end{proposition}

\begin{proof}
    This follows immediately from Propositions \ref{prop:omegaf}, \ref{prop:pfsigma}, and \ref{prop:pf}.
\end{proof}
Note that, while \citeauthor{sato1995statistical} refers to $\probabilitymeasure_{\comparisonfacts}$ as the \textit{basic distribution}, we use the term  \textit{basic measure}. This is due to the fact that we do not necessarily have a distribution.
}

%%%%%% logic program layer %%%%%%%%%%

\subsection{Logical Consequences of Boolean Comparisons}\label{sec:log_cons_bool_comp}
We now define the semantics of a \dfplpsty program, \ie, extend the basic measure $\measurecomparisonfacts$ over the comparison atoms to a measure over the Herbrand interpretations of a logic program.

\begin{definition}[\dfplpsty Program]\label{def:core-prog}
A \dfplpsty program $\dfprogram= \distdb \cup \logicprogram $ consists of a well-defined distributional database $\distdb$ (Definition~\ref{def:well-distdb}), comparison atoms $\comparisonfacts$ (Definition~\ref{def:comparison-atoms-set}), and a normal logic program $\logicprogram$ where clause heads belong to the regular vocabulary (cf. Definition~\ref{def:regular_vocabulary}), and which can use comparison atoms from $\comparisonfacts$ in their bodies.
\end{definition}

\begin{example}\label{ex:random_choice_dependency}
We further extend the running example.
\begin{problog*}{linenos}
% distributional facts $\distdb$
x ~ normal(5,2).
y ~ normal(x,7).
z ~ normal(y,1).
% logic program $\logicprogram$
a :- abs(x-y) =< 1. 
b :- not a, z>10.
\end{problog*}
The logic program defines two logical consequences of Boolean comparisons, where \probloginline{a} is true if the absolute difference between random variables \probloginline{x} and \probloginline{y} is at most one, and \probloginline{b} is true if \probloginline{a} is false, and the random variable \probloginline{z} is greater than $10$. 
\end{example}

In order to extend the basic measure to logical consequences, \ie logical rules, we require the notion of a {\em consistent comparisons database} (CCD).
The key idea is that samples of the random variables in $\distdb$ jointly determine a truth value assignment to the comparison atoms in $\comparisonfacts$.

\begin{definition}
    A {\em value assignment}
    $\samplefunction(\randomvariableset)$ is a combined value assignment to all random variables $\randomvariableset=\{\nu_1,\nu_2,... \}$, \ie,  $\samplefunction(\randomvariableset)=(\samplefunction(\nu_1),\samplefunction(\nu_2),\ldots)$.
\end{definition}

\begin{definition}[Consistent Comparisons Database]\label{def:consistent-fact-db}
    Let $\distdb$ be a well-defined distributional database,  $\comparisonfacts=\{\kappa_1, \kappa_2,\ldots\}$ a set of measurable Boolean comparison atoms, and  $\samplefunction(\randomvariableset)$ a  value assignment to all random variables $\randomvariableset=\{ \nu_1,\nu_2,... \}$. 
    We define $I_{\samplefunction(\randomvariableset)}(\kappa_i)=\top$ if $\kappa_i$ is true after setting all random variables to their values under $\samplefunction(\randomvariableset)$, and $I_{\samplefunction(\randomvariableset)}(\kappa_i) = \bot$ otherwise. $I_{\samplefunction(\randomvariableset)}$ induces the consistent comparisons database $\comparisonfacts_{\samplefunction(\randomvariableset)}=\{\kappa_i \mid I_{\samplefunction(\randomvariableset)}(\kappa_i) = \top\}$.
\end{definition}

To define the semantics of a \dfplpsty program $\dfprogram$, we now require that, given a CCD $\comparisonfacts_{\samplefunction(\randomvariableset)}$, the logical consequences in $\dfprogram$ are uniquely defined.
As common in the PLP literature, we achieve this by requiring the program to have a two-valued well-founded model~\citep{van1991well} for each possible value assignment $\samplefunction(\randomvariableset)$.

\begin{definition}[Valid \dfplpsty Program]\label{def:core-valid}
A \dfplpsty program $\dfprogram=\distdb \cup\logicprogram $ is called \emph{valid} if and only if for each CCD $\comparisonfacts_{\samplefunction(\randomvariableset)}$, the logic program $\comparisonfacts_{\samplefunction(\randomvariableset)} \cup  \logicprogram$ has a two-valued well-founded model. 
\end{definition}

We follow the common practice of defining the semantics with respect to ground programs; the semantics of a program with non-ground $\logicprogram$ is defined as the semantics of its grounding with respect to the  Herbrand universe.

\begin{restatable}{proposition}{proppp}
\label{prop:pp}
A valid \dfplpsty program \dfprogram  induces a unique probability measure $\probabilitymeasure_\dfprogram$ over Herbrand interpretations.
\end{restatable}

\begin{proof}
    See Appendix~\ref{app:proof:pp}.
\end{proof}

\begin{definition}
    We define the declarative semantics of a \dfplpsty program \dfprogram to be the probability measure $\probabilitymeasure_\dfprogram$, and we call this the {\em measure semantics}.  
\end{definition}

In contrast to the imperative semantics of~\citet{nitti2016probabilistic}, in \dfplpsty the connection between comparison atoms and the logic program is purely declarative. That is, logic program negation on comparison atoms negates the (interpreted) comparison. For example, if we have a random variable \probloginline{n}, then \probloginline{n>=2} is equivalent to \probloginline{not n<2}. Such equivalences do not hold in the stratified programs introduced by~\citet{nitti2016probabilistic}.
This then allows the programmer to refactor the logic part as one would expect.

\subsection{Constructing Distributional Databases}
\label{sec:fintiedistdb}

\new{
In the definition of the measure semantics, we simply assumed that the distributional database $\distdb$ was valid, and we forewent an explicit construction of such databases.
When implementing \dcplpsty we would, however, ideally have a prescription for constructing $\distdb$.
A practical choice that is often made in probabilistic logic programming~\citep{kersting2000bayesian,fierens2015inference} restricts the distributional database to be a finite (and consequently countable) set.
For such distributional databases with a countable number of random variables to be meaningful, they have to encode a unique joint distribution over the variables~$\randomvariableset$.
}

We will now provide the conditions under which a finite distributional database is valid, but first we define the concepts of parents and ancestors that, which also hold for non-finite distributional databases.
\begin{definition}[Parents and Ancestors of Random Variables]
    \label{def:df_ancestor}
Let $\distdb$ be a distributional database. For facts $\nu_p\sim\delta_p$ and $\nu_c\sim\delta_c$  in $\distdb$.  The random variable $\nu_p$  is a \emph{parent} of the child random variable $\nu_c$ if and only if  $\nu_p$ appears in $\delta_c$. We define \emph{ancestor} to be the transitive closure of \emph{parent}. A node's ancestor set is the set of all its ancestors.
\end{definition}

\begin{example}[Ancestor Set]
    We graphically depict the ancestor set of the distributional database in Example~\ref{ex:dist_db} in Figure~\ref{fig:ex:dist_db_ancestor}.

    \begin{figure}[h]
        \centering
    \tikzstyle{node}=[circle, text centered, draw,thick]
        \begin{tikzpicture}[remember picture]

            \node[node] (x) {\probloginline{x}};
            \node[node] (y) [right=of x] {\probloginline{y}};
            \node[node] (z) [right=of y] {\probloginline{z}};

			\draw[->,thick] (x) to  (y);
			\draw[->,thick] (y) to  (z);

        \end{tikzpicture}
        \caption{Directed acyclic graph representing the ancestor relationship between the random variables in Example~\ref{ex:dist_db}. The ancestor set of \probloginline{x} is the empty set, the one of \probloginline{y} is $\{\mathprobloginline{x} \}$ and the one of \probloginline{z} is $\{\mathprobloginline{x}, \mathprobloginline{y} \}$.}
        \label{fig:ex:dist_db_ancestor}
    \end{figure}

\end{example}

% \new{
% \begin{definition}[Ancestor Chain]
%     \label{def:df_ancestor_chain}
% An ancestor chain is an ordered set of random variables $\nu_1, \nu_2, \dots$ for which we have that $\nu_{i+1}$ is a parent of $\nu_{i}$. 
% \end{definition}
% }

\begin{definition}[Well-Defined Finite Distributional Database]\label{def:well-defd-facts}
A finite distributional database~$\distdb$  is called \emph{well-defined} if and only if it satisfies the following criteria:
\begin{description}
    \item[W!] Each $\nu \in \randomvariableset$ has a finite set of ancestors.
    \item[W1] The ancestor relation on the variables~$\randomvariableset$ is acyclic.
\item[W3] If $\nu \sim \delta \in \distdb$ and the parents of $\nu$ are $\{ \nu_1, \dots, \nu_m \}$, then replacing each occurrence of $\nu_i$ in $\delta$ by a sample $\samplefunction(\nu_i)$ always results in a well-defined distribution for~$\nu$. 
\end{description}  
\end{definition}

The distributional database in Example~\ref{ex:dist_db} is well-defined: the ancestor relation is acyclic and finite, and as normally distributed random variables are real-valued, using such a variable as the mean of another normal distribution is always well-defined. The database would no longer be well-defined after adding \probloginline{w ~ poisson(x)}, as not all real numbers can be used as a parameter of a Poisson distribution.

\new{
Constructing valid finite distributional databases can be viewed as building Bayesian networks over a finite set of variables, where nodes represent (conditional) random variables and where each node's distribution is parameterized by the node's parents.
This approach was, for instance, taken by~\citep{kersting2000bayesian}.
More recently, \citet{wu2018discrete} extended these finite Bayesian network to allow also (under certain conditions) for infinite and uncountable numbers of nodes in these Bayesian networks. They dub this extension {\em measure theoretic Bayesian network} (MTBNs).

While our measure semantics allow for MTBNs as the distributional database, we will restrict ourselves in the remainder of the paper to distributional databases that are finite. We leave it as future work on how to effectively use MTBNs in probabilistic logic programming. This would necessitate developing novel syntax and inference algorithms able to handle, for example, stochastic processes.
}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_main/tasks.tex
aij dcproblog paper

\section{Probabilistic Inference Tasks} \label{sec:inference-tasks}

In Section~\ref{sec:log_cons_bool_comp} we defined the probability distribution induced by a \dfplpsty program by extending the basic measure to logical consequences (expressed as logical rules). The joint distribution is then simply the joint distribution over all (ground) logical consequences. We obtain marginal probability distributions by marginalizing out specific logical consequences.
This means that marginal and joint probabilities of atoms in \dfplpsty programs are well-defined. 
Defining the semantics of probabilistic logic programs using an extension of Sato's distribution semantics gives us the semantics of probabilistic queries: the probability of an atom of interest is given by the probability induced by the joint probability of the program and marginalizing out all atoms one is not interested in.

The situation is more involved with regard to conditional probability queries. In contrast to unconditional queries, not all conditional queries are well-defined under the measure semantics (as well ass the distribution semantics). We will now give the formal definition of the PROB task, which lets us compute the (conditional) marginal probability of probabilistic events and which has so far not yet been defined in the PLP literature for hybrid domains under a declarative semantics (\eg \citep{azzolini2021semantics}).

After defining the task of computing conditional marginal probabilities, we will study how to compute these probabilities in the hybrid domain. Before defining the PROB task, we will first need to formally introduce the notion of a conditional probability with respect to a  \dcproblogsty program.

\begin{definition}[Conditional Probability]
\label{def:conditional_prob}
Let $\herbrandbase$ be 
% the Herbrand base, i.e,
the set of all ground atoms in a given \dcproblogsty program \dcpprogram.
Let $\evidenceset = \{\eta_1,\ldots,\eta_n\} \subset \herbrandbase$ be a set of observed atoms, and  $e=\langle e_1,\ldots, e_n\rangle$ a vector of corresponding  observed truth values, with $e_i\in \{\bot, \top \}$.
We refer to $(\eta_1=e_1) \wedge \ldots\wedge (\eta_n=e_n)$  as the evidence and write more compactly  $\evidenceset = e$.
Let  $\mu \in\herbrandbase$ be an atom of interest called the query.
If the probability of $\evidenceset = e$ is greater than zero, then the conditional probability of $\mu=\top$ given $\evidenceset=e$ is defined as:
\begin{align}
    P_\dcpprogram(\mu=\top \mid \evidenceset=e)= \frac{P_{\dcpprogram}(\mu=\top, \evidenceset=e)}{P_\dcpprogram(\evidenceset=e)} \label{eq:conditional_prob}
\end{align}
\end{definition}

\begin{definition}[PROB Task]
\label{def:prob_task}
Let $\herbrandbase$ be the set of all ground atoms of a given \dcproblogsty program \dcpprogram.
We are given the (potentially empty) evidence $\evidenceset=e$ (with $\evidenceset \subset \herbrandbase$)
and a set $\queryset \subset \herbrandbase$ of atoms of interest, called
query atoms.
The {\bf PROB task} consists of computing the conditional probability of the truth value of every atom in $\queryset$ given the evidence, \ie compute the conditional probability $P_{\dcpprogram} (\mu {=} \top \mid \evidenceset {=} e)$ for each $\mu \in \queryset$.
\end{definition}

\begin{example}[Valid Conditioning Set]
    Assume two random variables $\nu_1$ and $\nu_2$, where $\nu_1$ is distributed according to a normal distribution and $\nu_2$ is distributed according to a Poisson distribution. Furthermore, assume the following conditioning set $\evidenceset= \{\eta_1=\top, \eta_2=\top \}$, where $\eta_1 \leftrightarrow (\nu_1>0)$ and $\eta_2 \leftrightarrow (\nu_2=5)$. This is a valid conditioning set as none of the events has a zero probability of occurring, and we can safely perform the division in Equation~\ref{eq:conditional_prob}.
\end{example}

\subsection{Conditioning on  Zero-Probability Events}

A prominent class of conditional queries, which are not captured by Definition~\ref{def:conditional_prob}, are so-called  zero probability conditional queries. For such queries the probability of the observed event happening is actually zero but the event is still possible. Using Equation~\ref{eq:conditional_prob} does not work anymore as a division by zero would now occur.

\begin{example}[Zero-Probability Conditioning Set]
    Assume that we have a random variable $\nu$ distributed according to a normal distribution and that we have the conditioning set $\evidenceset= \{ \eta=\top \} $, with $\eta\leftrightarrow (\nu=20)$. In other words, we condition the query on the observation that the random variable $\nu$ takes the value $20$ -- for instance in a distance measuring experiment. This is problematic as the probability of any specific value for a random variable with uncountably many outcomes is in fact zero and applying Equation~\ref{eq:conditional_prob} leads to a division-by-zero. Consequently,  an ill-defined conditional probability arises.
\end{example}

In order to sidestep divisions by zero when conditioning on zero-probability (but possible) events, we modify Definition~\ref{def:conditional_prob}. Analogously to~\citet{nitti2016probabilistic}, we follow the approach taken in~\citep{kadane2011principles}.

\begin{definition}[Conditional Probability with Zero-Probability Events]
\label{def:conditional_prob_zero_event}
Let $\nu$ be a continuous random variable in the \dcproblogsty program \dcpprogram with ground atoms \herbrandbase. Furthermore, let us assume that the evidence consists of $\evidenceset = \{ \eta_0 = \top  \}$ with $\eta_0\leftrightarrow (\nu=w)$ and $w\in \samplespace_\nu$.
The conditional probability of an atom of interest $\mu\in \herbrandbase$ is now defined as:
\begin{align}
    P_\dcpprogram (\mu= \top \mid \eta_0= \top)
    &= \lim_{\Delta w \rightarrow 0}
    \frac{P_\dcpprogram(\mu=\top, \nu \in [w-\nicefrac{\Delta w}{2}, w+\nicefrac{\Delta w}{2} ]) }{P_\dcpprogram(\nu \in [w-\nicefrac{\Delta w}{2}, w+\nicefrac{\Delta w}{2} ] )} \label{eq:condition_prob_zero_ev}
\end{align}
\end{definition}

To write this limit more compactly, we introduce an 
infinitesimally small constant $\delta w$ and two new comparison atoms  $\eta_1 \leftrightarrow ( w-\nicefrac{\delta w}{2} \leq  \nu)$ and $\eta_2 \leftrightarrow (\nu \leq w+\nicefrac{\delta w}{2} )$ that together encode the limit interval.
Using these, we rewrite Equation~\ref{eq:condition_prob_zero_ev} as
\begin{align}
    P_\dcpprogram (\mu = \top \mid \eta_0 = \top) =  \frac{P_\dcpprogram(\mu=\top,  \eta_1=\top, \eta_2=\top )}{P_\dcpprogram( \eta_1=\top, \eta_2=\top )}
\end{align}

Applying the definition recursively, allows us to have multiple zero probability conditioning events. More specifically, let us assume an additional continuous random variable $\nu'$ that takes the value $w'$ for which we define: 
$\eta^{\prime}_1 \leftrightarrow ( w'-\nicefrac{\delta w'}{2} \leq  \nu')$ and $\eta^{\prime}_2 \leftrightarrow (\nu' \leq w'+\nicefrac{\delta w'}{2} )$.
This then leads to the following conditional probability:
\begin{align}
    P_\dcpprogram(\mu=\top \mid \nu=w, \nu'=w') 
    &= \frac{P_\dcpprogram(\mu=\top, \eta_1=\top, \eta_2=\top \mid \nu'=w' )}{P_\dcpprogram(\eta_1=\top, \eta_2=\top \mid \nu'=w')} \nonumber \\
    &= \frac{ \frac{P_\dcpprogram(\mu=\top, \eta_1=\top, \eta_2=\top, \eta'_1=\top, \eta'_2=\top)}{ \cancel{P_\dcpprogram(\eta'_1=\top, \eta'_2=\top)}} }{\frac{ P_\dcpprogram(\eta_1=\top, \eta_2=\top, \eta'_1=\top, \eta'_2=\top)}{\cancel{ P_\dcpprogram(\eta'_1=\top, \eta'_2=\top) }}} \nonumber \\
    &= \frac{P_\dcpprogram(\mu=\top,\eta_1=\top, \eta_2=\top, \eta'_1=\top, \eta'_2=\top)}{P_\dcpprogram(\eta_1=\top, \eta_2=\top, \eta'_1=\top, \eta'_2=\top)} 
\end{align}
Here we first applied the definition of the conditional probability for the observation of the random variable $\nu$ and then for the observation of the random variable $\nu'$. Finally, we simplified the expression.

\begin{proposition}
\label{prop:existence_cond_prob_zero_evidence}
The conditional probability as defined in Definition~\ref{def:conditional_prob_zero_event} exists.
\end{proposition}
\begin{proof}
See ~\citep[Equation 6]{nitti2016probabilistic}.
\end{proof}

In order to express zero-probability events in \dcproblogsty we add a new built-in comparison predicate to the finite set of comparison predicates $\comparisonpredicates =\{ \mathprobloginline{<}, \mathprobloginline{>},\allowbreak \mathprobloginline{=<}, \mathprobloginline{>=}, \mathprobloginline{=:=},\allowbreak \mathprobloginline{=\=} \}$ (cf. Definition~\ref{def:reserved_vocabulary}).

\begin{definition}[Delta Interval Comparison]\label{def:delta_interval}
    For a random variable \probloginline{v} and a rational number \probloginline{w}, we define \probloginline{delta_interval(v,w)} (with $\text{\predicate{delta_interval}{2}} \in \comparisonpredicates$) as follows. If \probloginline{v} has a countable sample space, then  \probloginline{delta_interval(v,w)} is equivalent to \probloginline{v=:=w}. Otherwise, \probloginline{delta_interval(v,w)} is equivalent to the conjunction of
    the two comparison atoms \probloginline{w-@$\delta$@w=<v} and \probloginline{v=<w+@$\delta$@w},
    where \probloginline{@$\delta$@w} is an infinitesimally small number.
\end{definition}

The delta interval predicate lets us express conditional probabilities with zero probability conditioning events as defined in Definition~\ref{def:conditional_prob_zero_event}.

Zero probability conditioning events are often abbreviated as $P_\dcpprogram (\mu=\top\mid \nu=w)$.
This can be confusing as it does not convey the intent of conditioning on an infinitesimally small interval. To this end, we introduce the symbol `$\doteq$' (equal sign with a dot on top). We use this symbol to explicitly point out an infinitesimally small conditioning set. For instance, we abbreviate the limit
\begin{align*}
    \lim_{\Delta w \rightarrow 0}
    \frac{P_\dcpprogram(\mu=\top, \nu\in [w-\nicefrac{\Delta w}{2}, w+\nicefrac{\Delta w}{2} ]) }{P_\dcpprogram(\nu \in [w-\nicefrac{\Delta w}{2}, w+\nicefrac{\Delta w}{2} ] )}
\end{align*}
in Definition~\ref{def:conditional_prob_zero_event} as:
\begin{align}
    P_\dcpprogram(\mu=\top \mid \nu\doteq w )
\end{align}

More concretely, if we measure the height $h$ of a person to be $180cm$ we denote this by $h\doteq 180$. This means that we measured the height  of the person to be in an infinitesimally small interval around $180cm$. Note that the $\doteq$ sign has slightly different semantics for random variables with a countable support. For discrete random variables the $\doteq$ is equivalent to the $equal$ sign.

\begin{example} \label{ex:conditional_prob_zero}
    Assume that we have a random variable $\nu$ distributed according to a normal distribution and that we have the evidence set $\evidenceset= \{ \eta=\top \} $, with $\eta\leftrightarrow (\nu\doteq 20)$. This is a valid conditional probability defined through Definition~\ref{def:conditional_prob_zero_event}.
\end{example}

\begin{example}
    Assume that we have a random variable $\nu$ distributed according to a normal distribution and that we have the conditioning set $\evidenceset= \{ \eta=\top, \eta' = \top \} $, with $\eta_1\leftrightarrow (\nu\doteq 20)$ and $\eta'\leftrightarrow (\nu\doteq 30)$. This does not encode a conditional probability as the conditioning event is not a possible event: one and the same random variable cannot be observed to have two different outcomes.
\end{example}

The notation used to condition on zero probability events (even when using `$\doteq$') hides away the limiting process that is used to define the conditional probability. This can lead to situations where seemingly equivalent conditional probabilities have diametrically opposed meanings.

\begin{example} \label{ex:conditional_prob}
    Let us consider the conditioning set $\evidenceset = \{ \eta=\top, \eta'=\top \}$, with $\eta \leftrightarrow (\nu\leq 20)$ and $\eta' \leftrightarrow (20\leq \nu)$, which we use again to condition a continuous random variable $\nu$. In contrast to Example~\ref{ex:conditional_prob_zero}, where we directly observed $\nu\doteq 20$, here, Definition~\ref{def:conditional_prob} applies, which %leads to a division by zero 
    states that the conditional probability is undefined as $P(\nu\leq 20, 20\leq \nu )=0$.
\end{example}

\subsection{Discussion on the Well-Definedness of a Query}

The probability of an unconditional query to a valid \dcproblogsty program is always well-defined, as it is simply a marginal of the distribution represented by the program.  
This stands in stark contrast to conditional probabilities: an obvious issue are  divisions by zero occurring when the conditioning event does not belong to the set of possible outcomes of the conditioned random variable. Similarly to~\citet{wu2018discrete} we will assume for the remainder of the paper that conditioning events are always possible events, \ie events that have a non-zero probability but possibly an infinitesimally small probability  of occurring. This allows us to bypass potential issues caused by zero-divisions.\footnote{In general, deciding whether a conditioning event is possible or not is undecidable. This follows from the undecidability of general logic programs under the well-founded semantics~\citep{cherchago2007decidability}. A similar discussion is also presented in the thesis of Brian Milch~\citep[Proposition 4.8]{milch2006probabilistic} for the \blogsty language, which also discusses decidable language fragments~\citep[Section 4.5]{milch2006probabilistic}.}

Even when discarding impossible conditioning events, conditioning a probabilistic event on a zero probability (but possible) event remains inherently ambiguous~\citep{jaynes2003probability} and might lead to the Borel-Kolmogorov paradox.
Problems arise when the limiting process used to define the conditional probability with zero probability events (cf. Definition~\ref{def:conditional_prob_zero_event}) does not produce a unique limit. 
For instance, a conditional probability  $P(\mu=\top \mid  2\nu \doteq \nu' )$, where $\nu$ and $\nu'$ are two random variables, depends on the parametrization used. We refer the reader to~\citep{shan2017exact} and \citep{jacobs2021paradoxes} for a more detailed discussion on ambiguities arising with zero probability conditioning events in the context of probabilistic programming.
We will sidestep such ambiguities completely by limiting observations of zero probability events to direct comparisons between random variables and numbers. This makes also sense from an epistemological perspective: we interpret a conditioning event as the outcome of an experiment, which produces a number, for instance the reading of a tape measure.

\subsection{Conditional Probabilities by Example}

\begin{example}\label{example:problog_machine}
The following ProbLog program models the conditions under which machines work. There are two machines (Line \ref{program:problog_machines_dec}), and three (binary) random terms, which we interpret as random variables as the bodies of the probabilistic facts are empty. The random variables are: the outside temperature (Line \ref{program:problog_machines_temp}) and  whether the cooling of each machine works (Lines \ref{program:problog_machines_cool1} and \ref{program:problog_machines_cool2}). Each machine works if its cooling works or if the temperature is low (Lines \ref{program:problog_machines_work_cool} and \ref{program:problog_machines_work_temp}).

	\begin{problog*}{linenos}
machine(1). machine(2). @\label{program:problog_machines_dec}@

0.8::temperature(low). @\label{program:problog_machines_temp}@
0.99::cooling(1). @\label{program:problog_machines_cool1}@
0.95::cooling(2). @\label{program:problog_machines_cool2}@

works(N):- machine(N), cooling(N). @\label{program:problog_machines_work_cool}@
works(N):- machine(N), temperature(low). @\label{program:problog_machines_work_temp}@
	\end{problog*}
We can query this program for the probability of \dcplpinline{works(1)} given that we have as evidence that \dcplpinline{works(2)} is true:
$$
P(\text{\dcplpinline{works(1)}}{=}1\mid\text{\dcplpinline{works(2)}}{=}1)\approx 0.998
$$
\end{example}

\begin{example}\label{example:dcproblog_machine}
In the previous example there are only Boolean random variables (encoded as probabilistic facts) and the \dcproblogsty program is equivalent to an identical \problogsty program. An advantage of \dcproblogsty is that we can now use an almost identical program to model the temperature as a continuous random variable.

	\begin{problog*}{linenos}
machine(1). machine(2).

temperature ~ normal(20,5). @\label{program:dcproblog_machines_temp}@
0.99::cooling(1).
0.95::cooling(2).

works(N):- machine(N), cooling(N).
works(N):- machine(N), temperature<25.0. @\label{program:dcproblog_machines_work_temp}@
	\end{problog*}
We can again ask for the probability of \probloginline{works(1)} given that we have as evidence that \probloginline{works(2)} is true but now the program also involves a continuous random variable:
$$
P(\text{\probloginline{works(1)}}{=}\top\mid\text{\probloginline{works(2)}}{=}\top)\approx 0.998
$$
\end{example}

In the two previous examples we were interested in a conditional probability where the conditioning event has a non-zero probability of occurring. However, \dcproblogsty programs can also encode conditional probabilities where the conditioning event has a zero probability of happening, while still being possible.
\begin{example}\label{example:dcproblog:observation}
	We model the size of a ball as a mixture of  different beta distributions, depending on whether the ball is made out of wood or metal (Line~\ref{program:dcproblog_machines_observation:ad}).
	We would now like to know the probability of the ball being made out of wood given that we have a measurement of the size of the ball.
	\begin{problog*}{linenos}
3/10::material(wood);7/10::material(metal).@\label{program:dcproblog_machines_observation:ad}@

size~beta(2,3):- material(metal)@\label{example:dcproblog:observation:beta23}@.
size~beta(4,2):- material(wood).
	\end{problog*}
Assume that we measure the size of the ball and we find that it is $0.4cm$, which means that we have a measurement (or observation) infinitesimally close to $0.4$. Using the `$\doteq$' notation, we write this conditional probability as:
	\begin{align}
		P\Bigl(\text{\probloginline{material(wood)}}{=}\top \mid \left(\text{\probloginline{size@$\doteq$@4/10}}\right) {=}\top \Bigr)
	\end{align}
\end{example}

The {\em Indian GPA problem} was initially proposed by Stuart Russell as an example problem to showcase the intricacies of mixed random variables.
Below we express the Indian GPA problem in \dcproblogsty.

\begin{example} \label{ex:indian_gpa}
The Indian GPA problem models US-American and Indian students and their GPAs. Both receive scores on the continuous domain, namely from 0 to 4 (American) and from 0 to 10 (Indian), cf. Line~\ref{ex:gpa:dens_am} and \ref{ex:gpa:dens_in}. With non-zero probabilities both student groups can also obtain marks at the extremes of the respective scales (Lines~\ref{ex:gpa:max_am}, \ref{ex:gpa:min_am}, \ref{ex:gpa:max_in}, \ref{ex:gpa:min_in}).

\begin{problog*}{linenos}
1/4::american;3/4::indian.

19/20::isdensity(a).
99/100::isdensity(i).

17/20::perfect_gpa(a).
1/10::perfect_gpa(i).

gpa(a)~uniform(0,4):- isdensity(a). @\label{ex:gpa:dens_am}@
gpa(a)~delta(4.0):- not isdensity(a), perfect_gpa(a).  @\label{ex:gpa:max_am}@
gpa(a)~delta(0.0):- not isdensity(a), not perfect_gpa(a). @\label{ex:gpa:min_am}@

gpa(i)~uniform(0,10):- isdensity(i). @\label{ex:gpa:dens_in}@
gpa(i)~delta(10.0):- not isdensity(i), perfect_gpa(i).  @\label{ex:gpa:max_in}@
gpa(i)~delta(0.0):- not isdensity(i), not perfect_gpa(i).  @\label{ex:gpa:min_in}@

gpa(student)~delta(gpa(a)):- american.
gpa(student)~delta(gpa(i)):- indian.
\end{problog*}
Note that in order to write the probability distribution of \probloginline{gpa(a)} and \probloginline{gpa(i)} we used uniform and Dirac delta distributions. This allowed us to distribute the random variables \probloginline{gpa(a)} and \probloginline{gpa(i)} according to a discrete-continuous mixture distribution.
We then observe that a student has a GPA of $4$ and we would like to know the probability of this student being American or Indian. 
\begin{align}
P\Bigl(\text{\probloginline{american}}{=}\top \mid (\text{\probloginline{gpa(student)}}\doteq4)=\top \Bigr)&=1 
\nonumber \\
P \Bigl(\text{\probloginline{indian}}{=}\top \mid (\text{\probloginline{gpa(student)}}\doteq4)=\top \Bigr)&=0 \nonumber
\end{align}
\end{example}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/arithmetics_old_20210120.tex
aij dcproblog paper

--- 2024-09-10 202409_aij_dc_problog_aij_second/files_excluded/types.tex
@@
-\section{Typing \dcproblogsty}
-\label{app:types}
+%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Type System}\label{sec:type_system}
+\section{Arithmetic Expressions}\label{sec:arithmetic}
+\ak{I think this should be a part of implementation section, and should be about comparison predicates rather than about arithmetic expressions}
 
-A distinctive property of system predicates, which are implemented using an external engine, is their mandatory type signature.
-An instance of such system predicates are comparison predicates.
-Consider for example the comparison (\probloginline{x>20}).
-The right-hand side is an integer and if the \probloginline{x} denotes a numeric random variable we have a well-typed comparison predicate inducing a measurable set.
-Consider for the sake of contrast also the comparison predicate (\probloginline{a>2}), where \probloginline{a} is an atomic formula.
-This comparison predicate is ill-typed as atomic formulas cannot be compared to (integer) numbers.
-In other words, the external engine, which implements the system predicates does not know how to commensurate an atomic formula and a number.
-A type system lets us specify which combinations of types are allowed as arguments of comparison predicates, \ie which comparisons encode measurable sets and are supported by the external engine.
+Implementations of (probabilistic) logic programming languages, such as \prologsty and \problogsty, reserve a set of predicate names for system-related procedures.
+These system predicates (or built-in predicates) are usually deployed when the goal is to perform efficient arithmetic operations on numbers.
+Either by using specialized algorithms or specialized hardware.
+In \dcproblogsty we are going to use an external arithmetic engine to evaluate expressions within comparison predicates.
 
-\begin{samepage}
+\begin{example}\label{ex:eval_comp} A \dcproblogsty code snippet demonstrating arithmetic evaluations within a comparison predicate.\ak{how is this example relevant here?}
+	\begin{problog*}{linenos}
+0.5::a.
+x ~ normal(0,3):- a.
+x ~ normal(1,2):- \+a.
+
+q:- x+3>3.
+query(q).
+	\end{problog*}
+Using the \probloginline{map}\lstinline[columns=fixed]|/2| predicate this program gets transformed into the program below.
+	\begin{problog*}{linenos}
+0.5::a.
+x1 ~ normal(0,3).
+map(x1,x):- a.
+
+x2 ~ normal(1,2).
+map(x2,x):- \+a.
+
+q:- map(X,x), X+3>3.
+query(q).
+	\end{problog*}
+
+\end{example}
+
+\luc{technical problem with DC; do we need this?}
+
+\subsection{Type System} \label{sec:type_system}
+
+While we have stated that arithmetic expressions can appear on both sides of comparison predicates, we have not yet mentioned what they evaluate to.
+To this end we will regard expressions to be evaluated as unnamed functions, instead of arithmetic predicates\ak{we do not have arithmetic predicates, only arithmetic functors}.
+Let us consider, for instance, the expression \probloginline{3+5}.
+This can can be viewed as a term whose predicate is \probloginline{+}\lstinline[columns=fixed]|/2| in infix notation.
+However, we can also regard the expression as a function with two inputs and an output.
+Furthermore, we can equip the function with a type signature.
+The input values are of type \type{Integer} and the output value is of \type{Integer} as well.
+We will now introduce a type system for the evaluation of arithmetic expressions in \dcproblogsty.
+
+The base type for (arithmetic evaluations in) \dcproblogsty is the \type{Term} type. We subtype terms with a \type{Distribution} type and a \type{Symbol} type. 
+Dividing \dcproblogsty terms into \justify\type{Distribution} terms and \type{Symbol} terms reflects the characterization of~\citet{poole2010probabilistic} of probabilistic programs as being constituted of independent choices (\type{Distribution}) and a deterministic system (\type{Symbol}).
+
 \begin{definition}\label{def:type_system}
 	The \dcproblogsty type system is a hierarchical type system with each type being a disjoint union of its subtypes.
-	The base type for \dcproblogsty is the \type{Term} type. We subtype terms with a \type{Distribution} type and a \type{Symbol} type. 
 A diagrammatic overview of the \dcproblogsty type system is given in Figure~\ref{fig:type_system}.
@@
 \end{definition}
-\end{samepage}
 
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{../figures/type_hierarchy_reduced.pdf}
+		\includegraphics[width=\linewidth]{../figures/type_hierarchy.pdf}
 	\end{center}
-	\caption[Diagrammatic representation of the hierarchical \dcproblogsty type system.]{Diagrammatic representation of the hierarchical \dcproblogsty type system.}
+	\caption[Diagrammatic representation of the hierarchical \dcproblogsty type system.]{Diagrammatic representation of the hierarchical \dcproblogsty type system.
+Compared to Definition~\ref{def:type_system},  the \type{Number} type is further subtyped into \type{Integer} and \type{Real} types and the \type{RandomVariable} type is subtyped into \type{DiscreteRV}, \type{ContinuousRV}, and \type{MixtureRV}.
+These subtypes of \type{RandomVariable} mimic the subtypes of the \type{Distribution} type.\luc{do we want mixtureRV?}}
 	\label{fig:type_system}
@@
 	Distribution terms are only allowed to be invoked as a second argument to the reserved distribution predicate \probloginline{~}\lstinline[columns=fixed]|/2|.
-	This also means that random function symbols are only used there as well. 
-% 	\dcproblogsty subtypes the \type{Distribution} type into \type{NumericD} and \type{CategoricalD}.
+	This also means that random function symbols are only used there as well. \dcproblogsty subtypes the \type{Distribution} type into \type{NumericD} and \type{CategoricalD}.
 % 	\type{DiscreteD}, \type{ContinuousD}, and \type{MixtureD} types.
@@
 \type{Distribution} terms determine the probability distribution according to which a random variable is distributed.
-\probloginline{poisson(6)} and \probloginline{normal(20,5)} are for instance of type \type{NumericD}.
-
-% Note that distributions of type \type{ContinuousD} denote {\em absolutely continuous} probability distributions\footnote{\dcproblogsty does not support {\em singular continuous} probability distributions such as the Cantor distribution.}.
+\probloginline{poisson(6)} and \probloginline{normal(20,5)} are for instance of types \type{DiscreteD} and  \type{ContinuousD}, respectively.
+Note that distributions of type \type{ContinuousD} denote {\em absolutely continuous} probability distributions\footnote{\dcproblogsty does not support {\em singular continuous} probability distributions such as the Cantor distribution.}.
 \ak{would be nice to have a hint on why this restriction is there} We give an example of a distribution of type \type{MixtureD} in Example~\ref{ex:indian_gpa_function}.
@@
 
-\begin{definition}[\type{Symbol}]
 \type{Symbol} terms are the building blocks of the discrete structure that makes up a logic program and as such determine also the deterministic system of a probabilistic logic program.
-Beyond the reserved syntax, which is used to define the distribution terms, the user is free to use any identifier (following \prologsty conventions) to declare \type{Symbol} terms (this includes standard \prologsty terms such as lists).
-\end{definition}
-
+Beyond the reserved syntax, which is used to define the distribution terms, the user is free to use any identifier (following \prologsty conventions) to declare \type{Symbol} terms -- this includes standard \prologsty terms such as lists.
 %The subtyping of the \type{Symbol} type (shown in Figure~\ref{fig:type_system}) follows the convention given in~\citep[Chapter 9]{sterling1994art}.
@@
 	\type{RandomSymbol} terms are named by the programmer using distributional facts with the binary infix predicate \probloginline{~}\lstinline[columns=fixed]|/2| and inherit their subtype from the distribution term.
-All random variables present in a \dcproblogsty program  explicitly named.
+All random variables present in a \dcproblogsty program are named.
 \end{definition}
@@
 \end{problog}
-As mentioned in Defintion~\ref{def:random_variable}, \type{RandomSymbol} terms inherit their type from the associated \type{Distribution} term, \eg a \type{NumericD} term induces a random variable term of type \type{NumericRV}.
+As mentioned in Defintion~\ref{def:random_variable}, \type{RandomSymbol} terms inherit their type from the associated \type{Distribution} term, \eg a \type{ContinuousD} term induces a random variable term of type \type{ContinuousRV}.
 
@@
 In the functional probabilistic programming language \anglicansty~\citep{wood2014approach}, for instance, the user can declare anonymous random variables, much like anonymous functions, \ie \textlambda-functions.
-\ak{Discussion of named/unnamed between Defs 6.4 and 6.5: would be nice to have a brief hint on why you made that choice} \pedro{usually in logic programming everything is global.
-an unamed random variable would mean that it is only available locally.
-we could allow that, I guess but would have to introduce new syntax.
-moreover, internally we would give the unnamed random variable a name anyway.}
-
-\begin{definition}[\type{Atom}]
-	An atom is of the form \probloginline{p(t@$_1$@,@\dots @, t@$_n$@ )} where \dcplpinline{p} is a predicate with arity $n$ and the \probloginline{t@$_i$@} are \type{Symbol} terms.
-Functors used to construct \type{Atom} terms must not clash with the builtin predicates and reserved functors used to declare \type{Distribution} terms.
-% 	Terms of type \type{Atom} constitute the building blocks of the discrete structure of a probabilistic logic program.
-\end{definition}
-
-Examples of \type{Atom} terms would be \probloginline{foo} or \probloginline{bar(foo,42)}.
-The latter is constructed using the predicate \lstinline{bar/2} and the tuple of \type{Symbol} terms \probloginline{(foo,42)}.
-\probloginline{foo} is of type \type{Atom} and \probloginline{42} is a \type{Number} term.
+\ak{Discussion of named/unnamed between Defs 6.4 and 6.5: would be nice to have a brief hint on why you made that choice}
 
@@
 
-\paragraph{Strongly Typed \prologsty}
+\ak{terminology mix-up; predicates never form terms} \pedro{not sure what this comment refers to.}
 
-While \prologsty is traditionally not strongly/statically typed, there have been efforts to extend \prologsty with such a type system.
-Arguably, the most interesting effort is presented in~\citep{Schrijvers2008TowardsTP}, where the authors develop the idea of partially typing a \prologsty program, \ie typing is optional and untyped code is interpreted instead of compiled.
-Introducing predicates with strongly typed signatures can be regarded as adding predicates with a specific meaning to the \prologsty implementation but the {\em `evaluation'} of these predicates still takes place within \prologsty.
-Our efforts of introducing a type system for system predicates within a probabilistic programming context are orthogonal and complimentary to the efforts of~\citet{Schrijvers2008TowardsTP}.
+\begin{definition}[\type{Atom}]
+	An atom is of the form \probloginline{p(t@$_1$@,@\dots @, t@$_n$@ )} where \probloginline{p} is a predicate of arity $n$ and the \probloginline{t@$_i$@} are \type{Symbol} terms~\citep{fierens2015inference}.
+Predicates used to construct \type{Atom} terms must not clash with the builtin predicates and reserved random function symbols used to declare \type{Distribution} terms.
+% 	Terms of type \type{Atom} constitute the building blocks of the discrete structure of a probabilistic logic program.
+\end{definition}
+
+Examples of atoms terms would be \probloginline{foo} or \probloginline{bar(foo,42)}.
+The latter is constructed using the predicate \lstinline{bar/2} and the tuple of \type{Symbol} terms \probloginline{(foo,42)}.
+\probloginline{foo} is of type \type{Atom} and \probloginline{42} is a \type{Number} term.
 
@@
 
-\subsection{External Engine}\label{sec:external_engine}
+\subsection{Multiple Dispatch} \label{ref:multiple_dispatch}
+\ak{I think there's a lot in here that misses context to be understandable; and I'm not sure all of it is really needed for the paper}
+The attentive reader might have noticed that, for example in Example~\ref{ex:dc-vs-dcproblog}, we used the same comparison predicates, for \type{RandomSymbol} terms as well as \type{Number} terms.
+In \dcproblogsty we resolve this clash through {\em multiple dispatching}~\citep{castagna1995calculus}.
+Multiple dispatching is the process of dynamically selecting at run time\footnote{In the context of logic programming we understand as run time the grounding of the logic program part of a \dcproblogsty program.} a specific implementation of a function based on the types of the arguments that are fed as input to the corresponding function symbol.
+Such functions are referred to as {\em multimethods}. 
+An early example of multiple dispatch is the {\em Common Lisp Object System}~\citep{keene1989object}.
+Multiple dispatch is also the programming paradigm on which the \juliasty programming language is built~\citep{bezanson2017julia}.
 
-In the previous subsection we have introduced an abstract type system.
+We also find multimethods in logic programming languages such as \prologsty.
+While, for groundings of relations \prologsty simply checks whether the grounding is present in the database (agnostic towards the types of the arguments), this is not the case for built-in predicates such as comparison predicates, which make a call to an external arithmetic engine in order to evaluate an expression.
+For instance, (\probloginline{1<3+5}) calls an integer addition procedure, whereas (\probloginline{1<3.0+5}) might call a floating point procedure.
+Calling the external arithmetic engine with an ill-typed expression might even cause a run time error.
+Something that is not possible in {\em pure} \prologsty without an external arithmetic engine.
+
+\luc{implementatin details}
+Compared to \problogsty and \problogsty, \dcproblogsty does not only have type signatures for the built-in comparison predicates but also for random function symbols, \eg \probloginline{normal/2}.
+Consider, for instance, the declaration of the normal distribution in Example~\ref{example:dcproblog_machine}.
+The definition of the function is the following:
+$$ \text{\probloginline{normal}}: \type{Number} \times \type{Number} \rightarrow \type{ContinuousD}$$
+However, the arguments of a normal distribution (its mean and standard deviation) might as well be random variables themselves.
+Multiple dispatching then allows us to {\em overload} the \probloginline{normal} function symbol by using arguments of a different type\footnote{For simplicity we omit that the standard deviation has to be a strictly positive (real) number, which can be enforced by extending the type system.}:
+\begin{align*}
+	\text{\probloginline{normal}}&: \type{NumericRV} \times \type{Number} \rightarrow \type{ContinuousD} \\
+	\text{\probloginline{normal}}&: \type{Number} \times \type{NumericRV} \rightarrow \type{ContinuousD} \\
+	\text{\probloginline{normal}}&: \type{NumericRV} \times \type{NumericRV} \rightarrow \type{ContinuousD}
+\end{align*}
+
+Reusing the idea of multimethods in the context of probabilistic logic programming with random function symbols results in a simple and neat syntax.
+The exact implementation of these random function symbols is left to the designer of an implementation of the language by providing an external engine that performs arithmetic evaluations.
+
+%In the next Subsection~\ref{sec:arithmetic} we will encounter further usages of multiple dispatching in \dcproblogsty.
+
+% We can say that interpreted predicates and functors, that is predicates and functors that have a system internal meaning, possess a type signature. This results in a type dependent calculus.
+
+\paragraph{Strongly Typed \prologsty}
+
+While \prologsty is traditionally not strongly/statically typed, there have been efforts to extend \prologsty with such a type system.
+Arguably, the most interesting effort is presented in~\citep{Schrijvers2008TowardsTP}, where the authors develop the idea of partially typing a \prologsty program, \ie typing is optional and untyped code is interpreted instead of compiled.
+Introducing predicates with strongly typed signatures can be regarded as adding interpreted predicates to the \prologsty implementation but the {\em `evaluation'} of these predicates still takes place within \prologsty.
+Our efforts to utilize multiple dispatch for arithmetic expressions with an external engine are orthogonal and complimentary to the efforts of~\citet{Schrijvers2008TowardsTP}.
+
+\ak{move to related work?}
+
+\subsection{External Engine}
+
+The key difference between the external arithmetic engine of \prologsty (or \problogsty) and the external arithmetic engine of \dcproblogsty is that the external engine of the latter is capable of performing arithmetic operations (perform evaluations) with random variables.
+Combining this innovation with multiple dispatch allows us to retain the same syntax (same functor) to perform arithmetic operations on \type{SymbolicRV} terms and/or \type{Number} terms.
+
+In Subsection~\ref{sec:type_system} we introduced an abstract type system.
 Abstract types describe the structure/hierarchy of a type system and cannot be instantiated.
@@
 
-The only condition that we necessitate for an implementation of \dcproblogsty is that the external engine provides concrete types for the abstract types.
-Additionally, a practical external engine should also implement operations on these types.
+The only condition that we necessitate for an implementation of \dcproblogsty is that the external arithmetic back-end provides concrete types for the abstract types that should be used in arithmetic expression.
+Additionally, a practical arithmetic engine should implement arithmetic operations on these types.
 For example, an engine should be able to multiply a floating point value and an integer value: (\probloginline{3.12*2}).
@@
 
-\begin{example}\label{example:julia_addition}
+\begin{example}
 We show two functions that perform addition of numbers but with different types.
 These different number could be invoked on either side of a comparison predicate, \eg (\probloginline{1<3+5} ) or (\probloginline{1<3+5.0} ).
-We write these functions using Julia like syntax~\citep{bezanson2017julia}.
+We write these functions using Julia like syntax, as Julia follows the multiple dispatch paradigm.
+\luc{too much}
 \begin{samepage}
@@
 \end{samepage}
+
 \end{example}
 
-The programming paradigm followed in Example~\ref{example:julia_addition} is called {\em multiple dispatch} \citep{keene1989object,castagna1995calculus,bezanson2017julia}.
-Multiple dispatching is the process of dynamically selecting at run time a specific implementation of a function based on the types of the arguments.
-Functions with the same functor but a different type signature are called multimethods.
+\subsubsection*{Eager Evaluation}
 
-We implement probability distributions using the external engine.
-As such, the corresponding random funtions symbols are necessarily typed.
-We recede again to multiple dispatch to implement probability distributions.
-Consider, for instance, the declaration of the normal distribution in Example~\ref{example:dcproblog_machine}.
-The definition of the function is the following:
-$$ \text{\probloginline{normal}}: \type{Number} \times \type{Number} \rightarrow \type{ContinuousD}$$
-However, the arguments of a normal distribution (its mean and standard deviation) might as well be random variables themselves.
-Multiple dispatching then allows us to {\em overload} the \probloginline{normal} function symbol by using arguments of a different type\footnote{For simplicity we omit that the standard deviation has to be a strictly positive (real) number, which can be enforced by extending the type system.}:
-\begin{align*}
-	\text{\probloginline{normal}}&: \type{NumericRV} \times \type{Number} \rightarrow \type{ContinuousD} \\
-	\text{\probloginline{normal}}&: \type{Number} \times \type{NumericRV} \rightarrow \type{ContinuousD} \\
-	\text{\probloginline{normal}}&: \type{NumericRV} \times \type{NumericRV} \rightarrow \type{ContinuousD}
-\end{align*}
+Let us revisit the comparison predicates, \eg (\probloginline{1<2}), (\probloginline{x>20}).
+According to Definition~\ref{def:semantics_dcproblog} these get wrapped into a \dcplpinline{holds/3} predicate when \dcproblogsty is transformed into \dcplpsty.
+For the deterministic case we can eagerly evaluate these \dcplpinline{holds/3} predicates.
+That is, we can replace the atom with its corresponding Boolean value.
+For the comparison (\probloginline{1<2}) we would get \probloginline{true}.
 
-Reusing the idea of multimethods in the context of probabilistic logic programming with random function symbols results in a simple and neat syntax.
-The exact implementation of these random function symbols is left to to the actual implementation of the language by providing an external engine capable of performing adequate arithmetic evaluations.
+Eager evaluation of arithmetic expressions makes most sense when this leads to program simplifications.
+More concretely, when a comparison predicate is not satisfied this results in a \probloginline{false} atom and proving a goal during grounding can be stopped pre-maturely.\ak{we haven't talked about grounding before...}
+Eager evaluation of the \dcplpinline{holds/3} predicate allows \dcproblogsty to retain the usual behavior of proving in \problogsty where a proof fails if a (deterministic) comparison is not satisfied.
 
-Currently, \dcproblogsys, our implementation of \dcproblogsty, does only emulate multimethods.
-This is due to the fact that the \dcproblogsys is written in Python, which does not follow the multiple dispatch programming paradigm.
-The emulation of multimethods is performed by using the \verb|isinstance| function in Python, which checks whether an object is of a certain type or not.
-This allows us to branch on the type of the arguments passed to a function and call different implementation given differently typed arguments.
+% \subsubsection{Arithemtics of Random Variables}
 
-An advantage of decoupling the deterministic system from the independent choices and implementing probability distribution using an external engine is that we can a single language with two different external engines, which might perform probabilstic inference differently.
-\dcproblogsys, for instance, provides an interface to two different external engines.
-The first one uses \pyrosty~\citep{bingham2019pyro}, a probabilistic programming language built on top of the PyTorch~\citep{paszke2019pytorch} deep learning framework.
-The second option uses the probabilistic programming language PSI~\citep{gehr2016psi}.
-PSI is a computer probabilistic computer algebra system.
-The main difference between the two different external engines becomes apparent when performing inference.
-The first one performs approximate inference while the second one exact inference.
-We will discuss this in more detail in Section~\ref{sec:inference}.
+% Evaluating arithmetic expressions that contain random variables should not to be confused with performing arithmetic operations on random variables.
+Writing \probloginline{x+y} in a comparison predicate, where \probloginline{x} and \probloginline{y} are random variables has the semantics of evaluating both random variables (for instance by drawing a sample) and adding up both evaluations.
+
+% In order to perform arithmetic operations on random variables directly, we would need to introduce syntax and an arithmetic engine for random variables.
+This could, for example, be achieved in the following fashion:
+% \begin{problog*}{linenos}
+%  x ~ normal(20,5).
+%  y ~ normal(25,6).
+%  z ~ x+y.
+% \end{problog*}
+% In this case z would be a normal distribution with mean ($20{+}25$) and standard deviation ($5{+}6$).
+We leave arithmetic operations on random variables for future research.\pedro{maybe add citation of somebody who has done something like this already?}
+
+%%%%%%%%%%%%%%%%%%%%

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/dcplp.tex
aij dcproblog paper

\section{\dcproblogsty: Core syntax and semantics}\label{sec:semantics}

\todo{better guide the reader through this part; potentially reconsider subsection headings}

In this section, we define the semantics of  \dcproblogsty programs that combine normal logic programs with ground \emph{distributional facts} only, which we also call \emph{core} \dcproblogsty. 
This restriction is for clarity only, 
as we will see in Section~\ref{sec:dcproblog}, where 
we will introduce additional modeling constructs inherited from the ProbLog and DC languages as syntactic sugar defined in terms of this core language.

\dcproblogsty's semantics is an instance of  Sato's distribution semantics~\citep{sato1995statistical}, which starts from a probability measure over a countable set of facts $F$, the so called \emph{basic distribution},  and then extend this to a probability measure over the Herbrand interpretations of the full program. It is worth noting that the basic distribution is independent of the logical rules. 
In our case, the set $F$ consists of ground Boolean comparison atoms over the random variables defined by the distributional facts. These comparison atoms form an interface between the random variables (represented as terms) and the logical layer that reasons about truth values of atoms. 
While \cite{gutmann2011magic} used the same principle to define the declarative semantics for Distributional Clauses without negation, their focus was on the procedural semantics, and only the latter has been extended to include negation \citep{nitti2016probabilistic}; cf.~also Section~\ref{sec:dcproblog-dc}.
\ak{it seems appropriate to clearly attribute the idea to DC here already, but not sure how to best do this "in passing", without raising too many questions}

\subsection{Syntax}
In the following, we assume the following reserved vocabulary:
\begin{itemize}
    \item A finite set $F_D$ of \emph{distribution functors}, such as \verb|normal/2|, \verb|poisson/1| etc, used to specify the distributions of random variables.
    \item A finite set $F_A$ of \emph{arithmetic functors}, such as \verb|max/2|, (infix) \verb|+/2| etc, as in Prolog arithmetic.
    \item The countable set $F_N=\{0,1,2,\ldots\}$ of constants denoting natural numbers.
    \item $F_M = F_A\cup F_N$
    \item A finite set $C$ of \emph{comparison predicates}, which we further fix to the binary 
    operators \verb|<,>,=<,>=,=:=,=\=| (in Prolog syntax and infix notation) for the sake of concreteness.
\end{itemize}
The precise definitions of $F_D$ and $F_A$, as well as the syntax for defining queries and observations for inference, are left to system designers. \ak{decide names of sets later, when we know all of them...}
As logic programming generally operates on a countable universe, we restrict arithmetic to the rational numbers (constructed from the natural numbers and appropriate functors) here. As for Prolog arithmetic, an implementation can of course support floating point numbers as syntactic sugar in programs, and we also use those in our examples for ease of readability.

\begin{definition}[Distributional Fact]
A \emph{distributional fact} is of the form \probloginline{x ~ d}, with \probloginline{x} a ground term whose functor is not part of the reserved vocabulary, and \probloginline{d} a ground term with a functor in $F_D$. Such a fact declares \probloginline{x} to be a random variable following the distribution denoted by \probloginline{d}. 
\end{definition}

\begin{definition}[(Core) DC-ProbLog program]
A \dcproblogsty program $P=(D,R)$ consists of
a countable set $D$=\{\probloginline{x1 ~ d1}, \probloginline{x2 ~ d2}, \ldots\} of distributional facts, where the \probloginline{xi} are all different, and a normal logic program $R$ where clause heads do not use reserved predicates. $P$ defines the set of random variables $V=\{\text{\probloginline{x}}~|~\exists \text{\probloginline{d}}. \text{\probloginline{x~d}}\in D\}$. 
\end{definition}

\begin{example}\label{ex:random_choice_dependency}
Consider the  following \dcproblogsty program $P=(D,R)$.
\begin{problog*}{linenos}
% distributional facts D
x ~ normal(5,2).
y ~ normal(x,7).
z ~ normal(y,1).
% logic program R
a :- abs(x-y) =< 1. 
b :- not a, z>10.
\end{problog*}
The distributional facts encode a Bayesian network with  normally distributed random variables, two of which serve as parameters in the distribution of another one. We thus have $V=\{$\dcplpinline{x}, \dcplpinline{y}, \dcplpinline{z}$\}$.
The logic program defines two logical consequences of Boolean comparisons over the BN, where \probloginline{a} is true if the absolute difference between random variables \probloginline{x} and \probloginline{y} is at most one, and \probloginline{b} is true if \probloginline{a} is false and random variable \probloginline{z} is greater than ten. 
Here, $F$ contains the comparison atoms \probloginline{abs(x-y) =< 1} and \probloginline{z>10} explicitly mentioned in the logic program, but also all other meaningful comparison atoms that can be built from the numbers in the program and the vocabulary for arithmetic and comparisons, \eg, \probloginline{x<y} or \probloginline{7*x=:=y+5}. 
\end{example}
\ak{What about eg x/y<5 which divides by zero in some worlds?}
In Section~\ref{sec:dcproblog}, we will extend the syntax to also include probabilistic facts, annotated disjunctions and distributional clauses.

\subsection{Validity}\label{sec:dcplpvalidity}
We now discuss validity criteria under which a 
DC-ProbLog program defines a unique probability distribution over its Herbrand interpretations.

The semantics of a \dcproblogsty program involves two kinds of assignments. First, each random variable takes a value according to the associated distribution. Second, once fixed, the values of the random variables jointly determine a truth value assignment to the comparison atoms, which we call a \emph{consistent fact database} to distinguish it from arbitrary interpretations of these atoms that may not respect their intended meaning, and which is extended to a Herbrand interpretation of the program via the logical rules. The following definition  introduces corresponding terminology and notation.
\begin{definition}[Consistent fact database]
Let $P=(D,R)$ be a \dcproblogsty program. For a distributional fact \probloginline{x~d} $\in D$, we denote by  $\omega$(\probloginline{x}) any value that \probloginline{x} can take according to \probloginline{d}.\todo{proper terminology for "can take"?} We write $\omega = (\omega(v_1),\omega(v_2),\ldots)$ for a combination of value assignments to all random variables $V=v_1,v_2,...$. 
We denote by $F=\{q_1,q_2,\ldots\}$ the set of all well-typed ground Boolean comparison atoms  that can be built from the reserved vocabulary $F_A\cup C$ together with $V$. \todo{what exactly is "well-typed"?} We define $I_{\omega}(q_i)=1$ if $q_i$ is true after setting all random variables to their values under $\omega$, and $I_{\omega}(q_i)=0$ otherwise. $I_{\omega}$ induces the  \emph{consistent fact database} $F_{\omega}=\{q_i~|~I_{\omega}(q_i)=1\}$. 
\end{definition}

Informally, in order to obtain a probability distribution over Herbrand interpretations, we  need to ensure that
\begin{itemize}
    \item The distributional facts $D$ define a unique joint distribution over the random variables they introduce, which will be our starting point.
    \item The comparison atoms are measurable under that distribution, such that we can use it to define the basic distribution.
    \item For any consistent fact database, adding the rules $R$ results in a program whose canonical model is two-valued, \ie, assigns a truth value to every atom in the Herbrand base. This is needed to extend the basic distribution to all atoms.
\end{itemize}

We further elaborate on the first point in the following. The key idea is to view the set of random variables as the nodes in a Bayesian network, where each node's distribution is parameterized by the parents of that node, as already illustrated in Example~\ref{ex:random_choice_dependency}.  
\begin{definition}[parent, ancestor]
Given a program $P$ with facts \probloginline{vp~dp} and \probloginline{vc~dc}  in $D$, the random variable \probloginline{vd}  is a \emph{parent} of the child random variable \probloginline{vc} if and only if  \probloginline{vd} appears in \probloginline{dc}. We define \emph{ancestor} to be the transitive closure of \emph{parent}.
\end{definition}

\begin{example}\label{ex:random_choice_dependency_parents}
In Example~\ref{ex:random_choice_dependency}, \probloginline{x} is the parent of \probloginline{y}, and \probloginline{y} is the parent of \probloginline{z}. Thus, \probloginline{x} does not have ancestors, \probloginline{y}'s only ancestor is \probloginline{x}, and both \probloginline{x} and \probloginline{y} are ancestors of \probloginline{z}.
\end{example}

The next definition formalizes the BN idea.
\begin{definition}\label{def:well-defd-facts}
The set $D$ of distributional facts in a program $P=(D,R)$ is called \emph{well-defined} if and only if it satisfies the following criteria:
\begin{labeling}{W3}
\item[W1] Each $v_i\in V$ has a finite set of ancestors.
\item[W2] The ancestor relation on the variables~$V$ is acyclic.
\item[W3] For each atom $v_i \sim d_i$ in $D$, and for each combination of possible values the parents of $v_i$ can take given the distributions associated to them in $D$, 
$d_i$ is a well-defined distribution term. 
\end{labeling}  
\end{definition}

%\pedro{do we need finite support condition?}\ak{finite support is something else, and not needed here (not part of DS, only needed for inference)}

\begin{definition}\label{def:core-valid}
A \dcproblogsty program $P=(D,R)$ is called \emph{valid} if and only if it satisfies the following criteria:
\begin{labeling}{V3}
\item[V1] The set $D$ is well-defined. 
\item[V2] All Boolean queries in $F$ are Lebesgue-measurable.
\item[V3] For each consistent fact database $F_{\omega}$, the logic program $F_{\omega}\cup R$ has a two-valued well-founded model~\citep{van1991well}.  
\end{labeling}
\end{definition}

\subsection{Semantics}
We follow the common practice of defining the semantics with respect to ground programs; the semantics of a program with non-ground $R$ is defined as the semantics of its grounding with respect to the  Herbrand universe.

Let $P$ be a valid \dcproblogsty program as defined above.
If $D$ is empty, \ie, $P$ does not define any random variables,  the semantics of $P$ is the well-founded model of $R$. Thus, normal logic programs are a special case of \dcproblogsty.\footnote{Note that in the absence of random variables, all well-typed Boolean queries, and thus all facts in $F$, have deterministic truth values.} In the following, we assume that $D$ is not empty.

Our definition of the semantics closely follows Sato's distribution semantics \citep{sato1995statistical}, \ie, we first establish a probability measure over interpretations of the countable set of facts $F$, and then  extend this to one over Herbrand interpretations. The key differences are that (1) to support continuous distributions, our measure over facts is in itself induced by a measure over value assignments to the random  variables and (2) to support negation, we use normal logic programs with the well-founded model as canonical model instead of definite clause programs with least Herbrand models.

\subsubsection{Probability measure over random variables}
We first show that a valid DC-ProbLog program $P=(D,F)$ defines a probability measure $P_V$ on truth value assignments $\omega$ to its random variables $V$. To do so, we borrow ideas from Bayesian Logic Programs (BLPs)~\citep{kersting2000bayesian}\footnote{Note that while BLPs also use LP syntax to define the random variables and structure of a Bayesian network, the way they use that syntax is fundamentally different from ours.}, exploiting the analogy between the ancestor structure on $V$ and the structure of a Bayesian network. 
Validity condition V1 states that $D$ is well-defined, \ie, each random variable has a finite set of ancestors, and the ancestor relation is acyclic. There thus exists an enumeration $v_1,v_2,...$ of the random variables such that $i<j$ for all pairs $(v_i,v_j)$ where $v_i$ is an ancestor of $v_j$.
V1 further ensures that each variable's distribution is well-defined for every combination of parent values. For each finite subset of random variables closed under the ancestor relation, the joint distribution on that set  thus has the form of a Bayesian network, and factorizes into the product of the individual variables' distributions. It can be shown that under these conditions, there is a unique probability measure $P_V$ over the space of possible assignments to $V$; we refer to \cite[Theorem 4.9]{kersting2000bayesian} for the technical details.

\luc{not formal enough?}
\ak{should be a bit more formal than before; I don't know how to make it more formal without reproducing Kristian's work...}

\subsubsection{Probability measure over (consistent) fact databases}
We now show that the probability measure $P_V$ over assignments to the random variables uniquely induces a probability measure $P_F$ over truth value assignments to the facts $F$; this is the \emph{basic distribution} in the distribution semantics.

We fix an arbitrary enumeration  $A_1,A_2,...$ of the atoms in $F$. 
Each $A_i$ \emph{depends} on a finite set $V_i$ of random variables, namely  those mentioned in $A_i$ as well as all their ancestors. We write $V_{\leq n}=\bigcup_{1\leq j\leq n}V_j $ for the union of random variables the first $n$ atoms in the enumeration depend on, and $P_{V_{\leq n}}$ for $P_V$ restricted to this set.

By validity criterion {\bf V2}, all queries $q_i$ are  Lebesgue-measurable, and we thus get a family of distributions
\begin{align}
P_F^{(n)}(A_1=x_1,\ldots,A_n=x_n) = \int_{\Omega(V_{\leq n})} \mathbf{1}_{[q_1=x_1\wedge\ldots\wedge q_n=x_n]}(\omega)dP_{V_{\leq n}}(\omega)
\end{align}
where the $x_i$ are $0$ or $1$, $P_{V_{\leq n}}$ factorizes over the random variables in $V_{\leq n}$ as discussed above, $\Omega(V_{\leq n})$ denotes the space of possible assignments for variables in $V_{\leq n}$, and $\mathbf{1}_{[\varphi]}$ is the indicator function, \ie, equals $1$ if $\varphi$ is true and $0$ otherwise. The definition in terms of an indicator function and the measurability of the underlying Boolean queries ensures that this family of distributions is of the form required for the distribution semantics, \ie, they are well-defined probability distributions satisfying the compatibility condition\footnote{I.e., $P_F^{(n)}$ can be obtained from $P_F^{(n+1)}$ by summing out $A_{n+1}$, for any $n$.}. There thus exists a completely additive probability measure $P_F$ over the space of truth value assignments to $F$ such that for any $n$, we have
\begin{align}
P_F(A_1=x_1,\ldots,A_n=x_n) =P_F^{(n)}(A_1=x_1,\ldots,A_n=x_n)
\end{align}
We note that by construction, this measure assigns mass to consistent fact databases only.

\subsubsection{Probability measure over Herbrand interpretations}
In the last step, we follow Sato's construction to obtain a probability measure $P_P$ over Herbrand interpretations from $P_F$. To do so, we fix an enumeration $A_1,A_2,\ldots$ of all atoms in the Herbrand base, which includes those in $F$. 
Validity criterion {\bf V3} ensures that for every consistent fact database $F_{\omega}$, the logic program $F_{\omega}\cup R$ has a total well-founded model $M_{\omega}$, and we can thus define   
\begin{align}
P_P^{(n)}(A_1=x_1,\ldots,A_n=x_n) := P_F(\{F_{\omega}~|~ M_{\omega}\models A_1^{x_1}\wedge\ldots\wedge A_n^{x_n}\})
\end{align}

\pedro{is the $P_F$(of a set) equal to sum of $P_F$(not a set)}\ak{what do you mean? note that the elements of the set are mutually exclusive}

where $A_i^1=A_i$ and $A_i^0=\neg A_i$, and it again follows that there is a completely additive probability measure $P_P$ over Herbrand interpretations, which completes the definition of the semantics.

\pedro{$P_p^{n}$ never introduced}\ak{(3) introduces it? we could explicitly write the analogue of (2) for $P_p$ as well if needed}

\subsection{Inference tasks}\label{sec:inference-tasks}
As in any probabilistic programming language, key inference tasks include computing the marginal probability of a query, \ie, a ground atom, being true in a given program, or the conditional probability of a query being true given observed truth values of one or more other logical atoms.
\ak{do we need formal details?}

\subsubsection{Conditional Queries}
\ak{this is mostly adapted from text moved from elsewhere; I am not fully sure whether all of it should stay here (some of it could be related work or details of inference algorithms)}

Queries conditioned  on standard logic atoms only  can be answered as usual by dividing the marginal probabilities of the conjunction of query and evidence by the marginal probability of just the evidence.

As the probability of a random variable with uncountably many outcomes taking a specific value is zero, conditioning on a random variable taking a specific value is slightly more complicated, as the above approach would result in  a division by zero and consequently an undefined conditional probability. 
This problem is related to the Borel-Kolomogorov paradox~\citep{kolmogorov1950foundations,gyenis2017conditioning}, which is again caused by a zero-division.

We will, analogously to \citet{nitti2016probabilistic},   follow~\citet{kadane2011principles} to resolve the division by zero issue.
We make the explicit distinction between the probability $P$ and the probability density function $p:\frac{d}{dx}P(X \leq x)=p(x)$. The conditional probability is defined as:\todo{align syntax with rest of semantics section}
\begin{align}
	P(\support_q| \support_e)
	&=P(\support_q|rv=v) & (\support_e \leftrightarrow rv=v )\\
	&=\lim_{\Delta v\rightarrow 0} \frac{P(\support_q,rv \in [v-\nicefrac{\Delta v}{2},v+\nicefrac{\Delta v}{2}])}{ P(rv \in [v-\nicefrac{\Delta v}{2},v+\nicefrac{\Delta v}{2}])} \label{equ:borel_limit} \\
	&=\lim_{\Delta v\rightarrow 0} \frac{\int \ive{ \support_q(x,rv=v)} p(x,rv=v)dx \cancel{\Delta v}}{p(rv=v)\cancel{\Delta v}} \\
	&=\frac{\int \ive{ \support_q(x, rv=v)} p(x,rv=v)dx}{p(rv=v)}
\end{align}
This means that we do not divide anymore by zero but by the number obtained when evaluating the probability distribution at the observed point. For example, when evaluating \probloginline{beta(2,3)} (cf. Example~\ref{example:dcproblog:observation}, Line~\ref{example:dcproblog:observation:beta23}) at $4/10$  we get $1.7280$ instead of zero.
% Note that this is not a probability between zero and one.

In order to guarantee that the limit in Equation~\ref{equ:borel_limit} exists, we restrict this kind of conditioning to equality between a {\em named} random variable and a given constant, cf. Example~\ref{example:dcproblog:observation}.
Relaxing this restriction might lead to the non-existence of the limit. Exactly this situation gives rise to the Borel-Kolmogorov paradox emerges.

Nevertheless, some languages lift this restriction, which results in possibly semantically ill-defined programs, such as in Distributional Clauses~\citep[Section 3.2]{nitti2016probabilistic}. A more sophisticated method to tackle the non-uniqueness of the conditional probability is {\em disintegration}~\citep{shan2017exact}, where, instead of returning a conditional probability, a family of conditional probabilities is returned (representing an uncountable number of limits). Other languages, such as BLOG, simply assume the existence of the limit~\citep{wu2018discrete}.

The limitation to direct equality statements might seem too restrictive for the purpose of probabilistic programming. Take for example the following conditional probability~\citep{nitti2016probabilistic}:
\begin{align}
	P(nationality| height {=} 160 \lor height {=} 180) \label{eqn:disj_conditioning}
\end{align}
We cannot express this query. However, if we think about what a conditioning set represents in the context of probabilistic programming and machine learning, the example above appears quite odd. The conditioning set tells us which observations we have made about the world, \ie data that we collected using a physical device (\eg tape measure for the height of a person). The example in Equation~\ref{eqn:disj_conditioning} now tells us that our measuring device returned two numbers for a single measurement, which means that the measuring protocol is underspecified.

We just stated that a measurement consists of retrieving a number from a measuring device. This is not entirely correct. In the natural sciences a measurement consists always of a number and an upper and lower error bound, which effectively eliminates the problem of zero probability events. However, given the ubiquity of single point measurement, we feel that a probabilistic programming language that is not able to handle single point measurements is greatly hurt in its expressivity.

To conclude,  \dcproblogsty allows a user to condition on zero probability events, while at the same time the imposed restrictions ensure that the encoded probability distribution is well defined.

\ak{conditional probabilities: just wondering: is it possible to condition on tests, \eg, observing that someone's height is at most 180? related to that, in (17), what if someone with bad handwriting measured the height and recorded it on a piece of paper, and I can't tell whether it is 160 or 180? :) }
\pedro{in this case the measurement is the pixels and you would have yes/no for each pixel}

We introduce syntax to specify inference tasks when discussing our implementation of \dcproblogsty in Section~\ref{sec:implementation}.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/dcplp_old.tex
aij dcproblog paper

\section{DC-PLP}\label{sec:semantics}

In this section we will briefly sketch the main ideas for defining the semantics of \dcproblogsty.
The semantics of \dcproblogsty are defined through the semantics of DC-PLP, a probabilistic logic programming language with both discrete and continuous random variables following \citeauthor{sato1995statistical}'s distribution semantics~\citep{sato1995statistical}. DC-PLP is an assembly language with minimal syntax, whose main purpose is to define the declarative semantics of the language independently of existing syntax, inference algorithms or systems.

\citet{poole2010probabilistic} has characterised probabilistic programming as independent choices plus deterministic systems, with (discrete) probabilistic logic programming being the case where the deterministic system is given by a logic program~\citep{poole2010probabilistic}. DC-PLP builds upon this idea, but relaxes the independence assumption by explicitly distinguishing two types of dependencies: 1) those expressed by the logic program (i.e.  deterministic system), and 2) those expressed by using random variables as parameters of distributions for other random variables.

DC-PLP serves as a stepping stone towards \dcproblogsty: the semantics of \dcproblogsty are defined through the semantics of DC-PLP and an additional unique deterministic program transformation that takes \dcproblogsty programs and maps them to DC-PLP programs.

%which will combine the semantics defined here with a higher level modeling language and a system combining that modeling language with inference algorithms developed in the context of related languages from the \problogsty family.

The key idea behind DC-PLP is to build the semantics of a program in three steps:
\begin{enumerate}
	\item Define a {\bf countable set of random variables}
	%	(denoted by terms in the logic program)
	and associate a unique parametric distribution
	%	(denoted by a term from a language $P$ in the logic program)
	to each random variable through a logic program that defines terms of the form \probloginline{rv(name,dist)}. We do not use the \probloginline{is}\lstinline[columns=fixed]|/2| infix notation for distributional clauses to make explicit that this is restricted to fully deterministic bodies. We do not introduce syntax for probabilistic facts, as these can be modeled through Boolean random variables and rules.
	For valid programs, this will give us a 
	unique joint probability measure over these random variables that factors into the individual distributions.
	\item Define a {\bf countable set of measurable Boolean queries}
	%	(denoted by terms from a language $Q$)
	over the random variables. These correspond to the type of tests that are written with comparison predicates (\probloginline{<}, \probloginline{=<}, \probloginline{>}, \probloginline{>=}, \probloginline{=:=}, \probloginline{=\=}). However, we here define these on the term level and wrap them into a single predicate \probloginline{test}\lstinline[columns=fixed]|/1| for ease of uniform notation.
	\item Define a {\bf logic program} that computes consequences of the Boolean queries. To include a  Boolean query \probloginline{q} in the body of a clause, it gets wrapped into a term of the form \probloginline{test(q)}. 
\end{enumerate}

The first step defines all random choices, while the second and third step form the deterministic system. 
As the second and third step are deterministic functions of the random variables, the measure defined in the first step extends to the Boolean queries and from there to the terms in the logic program.

%Note that DC-PLP is parameterized by two sets: the  language $P$ of parameterized distribution terms  and the language $Q$ of basic Boolean queries.
%We sometimes use DC-PLP(P,Q) to highlight this.

%Note further that while we implicitly assume  ``atomic'' random variables here, the principles also apply to ``compound'' random variables (such as multivariate Gaussians), provided usual laws of probability are respected and suitable syntax for such variables is introduced.

In contrast to the Distributional Clauses probabilistic programming language~\citep{gutmann2011magic}, where several conditional distributional clauses can jointly define the distribution of a single random variable through a complex logical factorization, DC-PLP uses a single fact to define a random variable unconditionally and associates a single parameterized distribution to it. As we still have the full power of normal logic programs to model the deterministic system, this does not restrict expressivity, but simplifies the formal definition of the semantics. When modeling conditional distributions, users have a range of choices, from encoding the entire conditional probability distribution into a single term within the defining \probloginline{rv}\lstinline[columns=fixed]|/2| fact, all the way to a set of individually named choice random variables with minimal use of random variables as parameters tied together by a deterministic system, or any intermediate factorisation.
As already discussed by~\citet{poole2010probabilistic}, not all choices work equally well with all inference algorithms.
DC-PLP provides these choices within the same semantical framework, leaving it to system developers to focus the syntax towards a subset of choices and/or to exploit the different choices when integrating different inference approaches into their system.

\ak{example 6.7: I'm surprised that the test/1 atoms explicitly appear as facts in the DC-PLP program, as I've always imagined them as a component defined implicitly (like the Herbrand universe in LP semantics). Is there a reason to make them explicit?}

\begin{example}[DC-PLP Program] \label{ex:dcproblog_program_without_dc_plpdc}
	This example program models the correct working of a machine.

\begin{problog*}{linenos}
rv(hot, flip(0.2)). 
rv(cooling(1), flip(0.99)). 
rv(temperature(hot), normal(27,5)).
rv(temperature(not_hot),normal(20,5)).

test(cooling(N)=:=1)
test(hot=:=1).
test(hot=:=0).
test(temperature(hot)<25.0).
test(temperature(not_hot)<25.0).

machine(1).
works(N):- machine(N), test(cooling(N)=:=1).
works(N):-
	machine(N),
	test(temperature(hot)<25.0),
	test(hot=:=1). 
works(N):-
	machine(N),
	test(temperature(not_hot)<25.0),
	test(hot=:=0).

query(works(1)).
\end{problog*}
	We are now able to clearly identify the three components of a DC-PLP program. In the first part we see the random variables (representing independent choices). Here we also also included the probabilistic facts from the intial \dcproblogsty program. DC-PLP encodes probabilistic facts using the \probloginline{flip}\lstinline[columns=fixed]|/2| random function symbol, which returns either $1$ or $0$ with the probability given as the argument. In the second block of code we see the countable set of measurable Boolean queries, and the last part contains the logic program.
\end{example}

\subsubsection*{Multiple Dispatch and Comparison Predicates}

As mentioned earlier, multiple dispatching is of outstanding usefulness when applying comparison predicates to different type signatures that result in semantically different terms. We will illustrate this by example.

\begin{example} Consider the following \dcproblogsty code snippet with no random variables.
	\begin{problog*}{linenos}
people(N):- N is 8.
large_group:- people(N), N>6.
	\end{problog*}
	During grounding, the comparison operator in the second line will dispatch to a deterministic comparison, as both sides of the comparison are of type \type{Number}. The arithmetic engine replaces the grounded comparison term with atom \probloginline{true}.
\end{example}

\begin{example} Consider now  a \dcproblogsty code snippet where a comparison predicate is present that receives as one of its arguments a term of type \type{RandomVariable}.
	\begin{problog*}{linenos}
people ~ poisson(6).
large_group:- people>6.
	\end{problog*}
	During grounding the comparison predicate dispatches to a probabilistic comparison. Instead of replacing the comparison by either \probloginline{true} or \probloginline{false}, the arithmetic engine defers the evaluation and ultimately adds a \probloginline{test(people>6)} term to the set of measurable Boolean queries.
\end{example}

\subsection{Valid DC-PLP Programs}

\pedro{\url{https://youtu.be/rdsPMCYMcZA?t=1510} stuart russell makes an interesting point about restricting the language.}

\begin{definition}
	A DC-PLP program \dcpprogram consists of:
	\begin{enumerate}
		\item A countable set $P$ of random variables.
		\item A countable set $Q$ of measureable Boolean queries over the random variables.
		\item A logic program $L$, which has access to the set of Boolean queries.
	\end{enumerate}
\end{definition}

\ak{\begin{itemize}
		\item says two conditions for a list of 3 (also check use of "second" in the text afterwards; not sure total models link to finite support)
		\item point 1: doesn't seem fully defined, as it only assigns ranks to RVs, but not all $d_i$ are necessarily RVs
		\item point 2: "every model of L" -- what is a model of a program with access to Boolean queries? needs to be defined
		\item point 3: what does it mean for a model to be "derivable" from a set of queries?
	\end{itemize}
}

\begin{definition} \label{def:valid_dcplp}
	A DC-PLP \dcpprogram program is called valid  if the following two conditions hold.
	\begin{enumerate}
		\item The program \dcpprogram is distribution stratisfied, that is, there exists a function $rank(\cdot)$ that maps ground random variables, of the form \probloginline{rv(rand,dist(d@$_1$@,...,d@$_n$@))}, to $\mathbb{N}$ and that satisfies $rank(\text{\probloginline{rand}})>rank(\text{\probloginline{d@$_i$@}})$ for every $i$ ($1\leq i \leq n$). 
		\item The logic program $L$ is {\em sound}~\citep{riguzzi2013well}, \ie every model of $L$ is a two-valued well-founded model~\citep{van1991well,riguzzi2013well}.
		\item Each model of $L$ is derivable from a finite set of Boolean queries in $Q$.
	\end{enumerate}
\end{definition}
The first condition ensures that programs do not exhibit cyclic functional dependencies, disallowing programs of the form:
\begin{problog*}{linenos}
x ~ normal(y,1).
y ~ normal(x,1).
q:- y>3.
query(q).
\end{problog*}
The second condition is necessary to satisfy the finite support condition of the distribution semantics~\citep{sato1995statistical}. Definition~\ref{def:valid_dcplp} follows closely the semantics of \problogsty~\citep{fierens2015inference}, which guarantees that a \dcproblogsty program without distributional facts is in fact a \problogsty program.

\subsubsection{Negation}

In order to add negation to the hybrid probabilistic logic programming language Distributional Clauses, \citet{nitti2016probabilistic} resorted to {\em negation as failure} within a {\em Selective Linear Definite (SLD) resolution}~\citep{sterling1994art} inference scheme to patch the initial (negation-free) semantics given in~\citep{gutmann2011magic}. The semantics then follow~\citep{przymusinski1988perfect}'s perfect model semantics.

The well-founded semantics already account for negation in logic programs and we do not have to add any extra machinery to add negation. On a side note: in the absence of negation the well-founded model of a logic program is identical to the {\em least Herbrand model}.

\subsubsection{Probability of a DC-PLP Program}

\begin{definition}
	Let $MOD(L)$ be the set of all two-valued well-founded models of $L$, which is the logic program of a valid DC-PLP program \dcpprogram, and $Q(l)$ the set of Boolean queries used to derive $l \in MOD(L)$. Furthermore, let $var(Q(l))$ be the set of random variables on which $Q(l)$ depends.
	The probability of the model $l$ is then given by:
	\begin{align}
		p(l) = \int \left( \prod_{q \in Q(l)} \ive{q} \right) \underbrace{\left( \prod_{x_l \in var(Q(l))} D_l(x_l)  \right)}_{\eqqcolon D_l(\xvars_l)} \underbrace{\left( \prod_{x_l \in var(Q(l))} d x_l  \right)}_{\eqqcolon d\xvars_l}
	\end{align}
	$D_l(x_l)$ is the probability distribution according to which the variable $x_l$ is distributed and which is given by the countable set $P$.

	The probability of the DC-PLP program \dcpprogram is given by:
	\begin{align}
		p(\dcpprogram) = \sum_{l \in MOD(L)} p(l) = \sum_{l \in MOD(L)} \int \prod_{q \in Q(l)} \ive{q} D_l(\xvars_l) d\xvars_l \label{eq:definition_probability_dcplp}
	\end{align}
\end{definition}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/dcplp_old_16122020.tex
aij dcproblog paper

\section{DC-PLP}\label{sec:semantics}
We introduce DC-PLP, a probabilistic logic programming language with both discrete and continuous random variables following Sato's distribution semantics \citep{sato1995statistical}. DC-PLP is an assembler language with minimal syntax whose  purpose is to define the declarative semantics of the language independently of existing syntax, inference algorithms or systems.

Poole has characterised probabilistic programming as independent choices plus deterministic systems, with (discrete) probabilistic logic programming being the case where the deterministic system is given by a logic program \cite{poole2010probabilistic}. DC-PLP builds upon this idea, but relaxes the independence assumption by explicitly distinguishing two types of dependencies, namely those expressed by the logic program /  deterministic system, and those expressed by using random variables as parameters of distributions for other random variables. It further allows for both discrete and continuous random variables.

In the context of of probabilistic logic programming, \cite{gutmann2011magic} have introduced Distributional Clauses (DC), which also combines continuous random variables with (definite) logic programs. We deviate from their proposal in the following ways:
\begin{itemize}
    \item we provide a simpler framework that clearly exposes all dependencies at no loss in expressivity
    \item we refine their validity criteria to address two known issues (logical cycles and parameter cycles)
    \item we support negation from the start
    \item we provide explicit connections to a range of existing PLP languages
\end{itemize}

DC-PLP serves as a stepping stone towards DC-ProbLog, which will combine the semantics defined here with a higher level modeling language, and a system combining that modeling language with inference algorithms that have been or will be developed in the context of related languages from the ProbLog family.

The key idea behind DC-PLP is to build the semantics of a program in three steps:
\begin{enumerate}
\item Define a countable set of random variables (denoted by terms in the logic program) and associate a unique parametric distribution (denoted by a term from a language $P$ in the logic program) to each random variable through a set of ground atoms of the form \verb|rv(name,dist)|. For valid programs, this will give us a 
unique joint probability measure over these random variables that factors into the individual distributions.
\item Define a countable set of measurable Boolean queries (denoted by terms from a language $Q$) over the random variables.
\item Define a logic program that computes consequences of the Boolean queries. To include a  Boolean query in the body of a clause, it gets wrapped into an atom of the form \verb|holds(query)|, whose purpose is to explicitly transfer the externally determined truth value of the query into the truth value of an atom within the logic program.\footnote{This is for simplicity of exposition; higher level modeling languages such as DC-ProbLog would usually provide comparison predicates on terms directly.}
\end{enumerate}
The first step defines all random choices, the second and third step form the deterministic system. 
As the second and third step are deterministic functions of the random variables, the measure defined in the first step extends to the Boolean queries and from there to the atoms in the logic program.

We restrict the definition of \verb|rv/2| to be a database of facts here purely for simplicity; this can be relaxed to a logic program that does not refer to any \verb|holds| atoms or any predicates defined in the third component (which can be mapped back to a fact database corresponding to its well-founded model).

Note  that while we implicitly assume  ``atomic'' random variables here, the principles also apply to ``compound'' random variables (such as multivariate Gaussians), provided usual laws of probability are respected and suitable syntax for such variables is introduced.

In contrast to DC, where several conditional distributional clauses can jointly define the distribution of a single RV through a complex logical factorisation, DC-PLP uses a single fact to define a random variable unconditionally and to associate a single parameterised distribution to it. As we still have the full power of normal logic programs to model the deterministic system, this does not restrict expressivity (as we will see in Section~\ref{sec:dc}), but simplifies the formal definition of the semantics. When modeling conditional distributions, users have a range of choices depending on the syntax provided, from encoding the entire CPD into a single term within the defining \verb|rv/2| fact all the way to a set of individually named choice RVs with minimal use of RVs as parameters tied together by a deterministic system, or any intermediate factorisation. As already discussed by Poole \cite{poole2010probabilistic}, not all choices work equally well with all inference algorithms. We provide these choices within the same semantical framework, leaving it to system developers to focus the syntax towards a subset of choices and/or to exploit the different choices when integrating different inference approaches into their system.

\subsection{Syntax}
A key element of DC-PLP are reserved functors\footnote{We consider constants to be functors of arity 0.} with a specific external interpretation, which are used to define distributions and evaluable expressions over random variables. While the exact choice of these functors is up to the designers of a specific DC-PLP instance, they have to obey certain rules which we specify next.

The set of functors used in a DC-PLP program can be partitioned into the following types:
\begin{enumerate}
\item \emph{reserved functors} with typed arguments and an output type, further partitioned into 
\begin{enumerate}
    \item \emph{distribution functors}, \eg \verb|poisson/1| with input and output type \emph{natural number}
    \item \emph{arithmetic functors}, \eg, the (zero-arity) functor \verb|17| is of return type number, \verb|max/2| takes two numbers and returns a number
    \item \emph{Boolean query functors} with output type Boolean, \eg, \verb|</2| with two arguments of type number
\end{enumerate}
\item \emph{user-defined functors}, whose arguments do not need to be typed, further partitioned into
\begin{enumerate}
\item \emph{random variable functors} with type declared  in the program through association with a distribution term as detailed below
    \item \emph{regular functors} do not need to be typed
\end{enumerate}
\end{enumerate}
We refer to a term with top level functor of type X as a term of type X, \eg, \verb|poisson(6)| is a distribution term, and \verb|x<5| with random variable \verb|x| a Boolean query term. 
In the following, we assume a fixed set of reserved functors.

DC-PLP further uses two reserved, typed predicates:
\begin{enumerate}
    \item \verb|rv/2|, where the  first argument has to be a random variable term and the second argument has to be a distribution term, and a fact \verb|rv(x,d)| associates the random variable \verb|x| with the distribution term \verb|d|, thus also setting the type of \verb|x| to the output type of \verb|d|
    \item \verb|holds/1|, where the argument has to be a Boolean query term, and the truth value of a ground atom \verb|holds(q)| is the result of evaluating query \verb|q|
\end{enumerate}
As in the non-probabilistic case of logic programming versus Prolog, we leave the introduction of the external arithmetic evaluation predicate \verb|is/2| to DC-ProbLog, but note that the \verb|holds/1| predicate has a similar role as external evaluation interface (which includes evaluating any arithmetic functors that are part of its argument).

A DC-PLP program $P=(D,R)$ consists of
\begin{itemize}
    \item a countable set $D=\{\mathtt{rv}(x_1,d_1),\mathtt{rv}(x_2,d_2),\ldots\}$ of ground well-typed \verb|rv/2| facts,  where the $x_i$ are all different 
    \item a normal logic program $R$ where the heads of clauses cannot use reserved predicates, and any occurrence of reserved predicates in clause bodies is well-typed
\end{itemize}

Consider the following example DC-PLP program:
\begin{example}[DC-PLP Program] \label{ex:dcproblog_program_without_dc_plpdc}
	This example program models the correct working of a machine.

\begin{problog*}{linenos}
rv(hot, flip(0.2)). 
rv(cooling(1), flip(0.99)). 
rv(temperature(hot), normal(27,5)).
rv(temperature(not_hot),normal(20,5)).

machine(1).
works(N):- machine(N), test(cooling(N)=:=1).
works(N):-
	machine(N),
	test(temperature(hot)<25.0),
	test(hot=:=1). 
works(N):-
	machine(N),
	test(temperature(not_hot)<25.0),
	test(hot=:=0).

query(works(1)).
\end{problog*}
	We are now able to clearly identify the two components of a DC-PLP program. In the first part we see the random variables (representing independent choices). Here we also also included the probabilistic facts from the initial \dcproblogsty program. DC-PLP encodes probabilistic facts using the \probloginline{flip}\lstinline[columns=fixed]|/2| random function symbol, which returns either $1$ or $0$ with the probability given as the argument. The second block of code  contains the logic program.
\end{example}
\todo{be consistent on test vs holds}\ak{from a declarative / LP perspective, I prefer holds over test}

Here, \verb|flip/1| and \verb|normal/2| are distribution functors, \verb|=:=/2| and \verb|</2| are Boolean query functors written in infix notation, \verb|hot/0|, \verb|cooling/1|, and \verb|temperature/1| are random variable functors, and we do not have  arithmetic and regular functors. \todo{change / expand example?}

\subsection{Validity}\label{sec:dcplpvalidity}
For a DC-PLP program $P=(D,R)$, let $V=\{v_1,v_2,\ldots\}$  be the set of  random variable terms appearing as first arguments in the facts in $D$, 
let $Q=\{q_1,q_2,\ldots\}$ be the set of all well-typed ground Boolean query terms  that can be built from the reserved vocabulary and $V$, and let $F=\{\mathtt{holds}(q_i)~|~q_i\in Q\}$ be the set of possible holds-atoms for these queries.

Given a program $P$ with facts \verb|rv|$(v_p,d_p)$ and \verb|rv|$(v_c,d_c)$ in $D$, the random variable $v_p$  is a \emph{parent} of the random variable $v_c$ if and only if  $v_p$ appears in $d_c$. We define \emph{ancestor} to be the transitive closure of \emph{parent}.\todo{example}

\begin{definition}\label{def:well-defd-facts}
The set $D$ is \emph{well-defined} if and only if it satisfies the following criteria:
\begin{enumerate}
\item For each atom \verb|rv|$(v_i,d_i)$ in $D$, $d_i$ is a well-defined distribution term for all possible values of the parents. \todo{is well-typed sufficient for this? if so, drop this point}
\item The ancestor relation is acyclic.
\item Each $v_i\in V$ has a finite set of ancestors.
\end{enumerate}  
\end{definition}

Let $\omega$ be an assignment that maps each random variable $v_i \in V$ to a possible value $\omega(v_i)$ from the associated distribution $d_i$. 
Such an $\omega$ induces a unique truth value assignment $b_{\omega}$ for the Boolean queries in $Q$, \ie, $b_{\omega}(q_i)=1$ if $q_i$ is true after setting all random variables to their values under $\omega$, and $b_{\omega}(q_i)=0$ otherwise. 
It further induces a unique truth value assignment $I_{\omega}$ to facts in $F$, \ie, $I_{\omega}(\mathtt{holds}(q_i))=b_{\omega}(q_i)$. We also write $F_{\omega}=\{\mathtt{holds}(q_i)~|~b_{\omega}(q_i)=1\}$, and refer to such a $F_{\omega}$ as a \emph{consistent fact database}.

A program is \emph{valid} if and only if it satisfies the following criteria:
\begin{description}
\item[(C1)] $D$ is well-defined. 
\item[(C2)] All Boolean queries in $Q$ are Lebesgue-measurable.
\item[(C3)] For each consistent fact database $F_{\omega}$, the logic program $F_{\omega}\cup R$ has a two-valued well-founded model.  
\end{description}

\subsection{Semantics}
We follow the common practice of defining the semantics with respect to ground programs; the semantics of a program with non-ground $R$ is defined as the semantics of its grounding with respect to the  Herbrand universe.

Let $P$ be a valid DC-PLP program as defined above.

If $D$ is empty, \ie, $P$ does not define any random variables,  the semantics of $P$ is the well-founded model of $R$. Thus, normal logic programs are a special case of DC-PLP.\footnote{Note that in the absence of random variables, all well-typed Boolean queries, and thus all holds-atoms, have deterministic truth values.} In the following, we assume that $D$ is not empty.

Our definition of the semantics closely follows Sato's distribution semantics, \ie, we also establish a probability measure over interpretations of a countable set of facts $F$ (the \verb|holds/1|-atoms), and extend this to one over Herbrand interpretations. The key differences are that (1) our measure over facts is in itself induced by a measure over value assignments to the random random variables and (2) we use normal logic programs with the well-founded model as canonical model instead of definite clause programs.

\subsubsection{Probability measure over random variables}
The (non-empty) set of facts $D$ introduces a  countable set of random variables $V=\{v_1,v_2,\ldots\}$, each associated with a distribution $d_i$.  Condition C1 ensures that these distributions are well-defined for every combination of parent values. It further ensures finite numbers of ancestors per random variable and an acyclic ancestor relation. As shown by  \citeauthor{kersting2000bayesian} \cite[Theorem 4.9]{kersting2000bayesian}, under these conditions, there exists a unique probability measure $P_V$ over the space of possible assignments to $V$ that for every finite set of random variables $V'\subseteq V$ closed under the ancestor relation factors into the distributions of the random variables in $V'$.

\subsubsection{Probability measure over (consistent) fact databases}
We now show that the probability measure $P_V$ over assignments to the random variables uniquely induces a probability measure $P_F$ over truth value assignments to the facts $F$; this is the \emph{basic distribution} in the distribution semantics.

We fix an arbitrary enumeration  $A_1=\mathtt{holds}(q_1),A_2=\mathtt{holds}(q_2),...$ of the atoms in $F$. 
Each $A_i$ \emph{depends} on a finite set $V_i$ of random variables, namely  those mentioned in $q_i$ as well as all their ancestors. We write $V_{\leq n}=\bigcup_{1\leq j\leq n}V_j $ for the union of random variables the first $n$ atoms in the enumeration depend on, and $P_{V_{\leq n}}$ for $P_V$ restricted to this set.

By condition C2, all queries $q_i$ are  Lebesgue-measurable, and we thus get a family of distributions 
$$P_F^{(n)}(A_1=x_1,\ldots,A_n=x_n) = \int_{\Omega(V_{\leq n})} \mathbf{1}_{[q_1=x_1\wedge\ldots\wedge q_n=x_n]}(\omega)dP_{V_{\leq n}}(\omega) $$ where the $x_i$ are $0$ or $1$, $P_{V_{\leq n}}$ factorizes over the random variables in $V_{\leq n}$ as discussed above, $\Omega(V_{\leq n})$ denotes the space of possible assignments for variables in $V_{\leq n}$, and $\mathbf{1}_{[\varphi]}$ is the indicator function, \ie, equals $1$ if $\varphi$ is true and $0$ otherwise. The definition in terms of an indicator function and the measurability of the underlying Boolean queries ensures that this family of distributions is of the form required for the distribution semantics, \ie, they are well-defined probability distributions satisfying the compatibility condition. There thus exists a completely additive probability measure $P_F$ over the space of truth value assignments to $F$ such that for any $n$, we have
$$P_F(A_1=x_1,\ldots,A_n=x_n) =P_F^{(n)}(A_1=x_1,\ldots,A_n=x_n)$$
We note that by construction, this measure assigns mass to consistent fact databases only.

\subsubsection{Probability measure over Herbrand interpretations}
In the last step, we follow Sato's construction to obtain a probability measure $P_P$ over Herbrand interpretations from $P_F$. To do so, we fix an enumeration $A_1,A_2,\ldots$ of all atoms in the Herbrand base, which includes those in $F$. 
Condition C3 ensures that for every consistent fact database $F_{\omega}$, the logic program $F_{\omega}\cup R$ has a total well-founded model $M_{\omega}$, and we can thus define   
$$P_P^{(n)}(A_1=x_1,\ldots,A_n=x_n) = P_F(\{F_{\omega}~|~ M_{\omega}\models A_1^{x_1}\wedge\ldots\wedge A_n^{x_n}\})$$ where $A^1=A$ and $A^0=\neg A$, and it again follows that there is a completely additive probability measure $P_P$ over Herbrand interpretations, which completes the definition of the semantics.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/dcproblog.tex
aij dcproblog paper

\section{\dcproblogsty Semantics}\label{sec:dcproblog}
While DC-PLP is an expressive probabilistic programming language for which we have defined fully declarative semantics, using DC-PLP in practice might turn out to be challenging. The syntax is simply too cumbersome. Therefore, we will now introduce semantics for the higher-level \dcproblogsty language whose syntax we already introduced in Section~\ref{sec:syntax}.

A \dcproblogsty program consists of a countable set of ground distributional facts, a countable set of ground probabilistic facts, and a logic program (i.e. a finite set of facts and normal clauses), where bodies of normal clauses can use comparison predicates.

\begin{definition} A \dcproblogsty program \dcpprogram is {\bf valid} if and only if it encodes a unique probability distribution.
\end{definition}

This raises the question: what are the conditions we have to impose on \dcproblogsty programs such that they are valid, or in other words such that they encode unique probability distributions? We will answer this question in Theorem~\ref{theo:valid_dcproblog}. But first we are going to define a program transformation that maps a \dcproblogsty program to a DC-PLP program.

\begin{definition} \label{def:dcproblog2dcplp}

\todo{ make transformations formal with substitutions}

\pedro{@Angelika, I believe you mentioned you had this transformation ready already?}
\ak{see copy from old below for probabilistic facts; LP stays the same (modulo dropping the test-wrapping?); distributional facts should be just rv(v,d) to v\probloginline{~}d (the complicated part would be directly mapping full DC)}

\end{definition}

\ak{begin copy}
\subsubsection{ProbLog with probabilistic facts only}
ProbLog with probabilistic facts (no ADs) restricts $P$ to discrete distributions over Boolean truth values, and $Q$ to equality comparisons between a random variable and the constant \verb|true|. The translation is modular, mapping each normal logic clause to itself, and each ground probabilistic fact \verb|p::f| to a program fragment
\begin{verbatim}
rv(f,flip(p)).
f :- holds(f=true).
\end{verbatim}
where \verb|flip(p)| denotes the distribution that returns \verb|true| with probability \verb|p| and \verb|false| otherwise.\footnote{This assumes that all probabilistic facts are syntactically different and associated to a single probability; for the multi-set view of the current system, the name of the random variable has to be made unique through different means, which is straightforward.} The correspondence in semantics is straightforward.

\subsubsection{ProbLog with ADs}
Adding ADs to ProbLog adds discrete distributions over finite sets $\{1,...,n\}$ to $P$, and equality comparisons between random variables and the elements of these sets to $Q$.
A ground AD \verb|p1::h1;...;pn::hn :- body| maps to a program fragment
\begin{verbatim}
rv(id,[p1:1,...pn:n]).
h1 :- body, holds(id=1).
...
hn :- body, holds(id=n).
\end{verbatim}
where \verb|id| is a fresh term not used anywhere else. The correspondence in semantics is again straightforward. 
\ak{end copy}

\begin{theorem} \label{theo:valid_dcproblog}
A \dcproblogsty program is valid if the following two validity conditions hold.
	\begin{itemize}
	    \item[V1] No two distributional facts can introduce the same random variable, \ie use the same term on the left hand side.
		\item[V2] The program \dcpprogram is distribution stratified, that is, there exists a function $rank(\cdot)$ that maps ground terms to $\mathbb{N}$ and that satisfies the following properties:
		\begin{enumerate}			
			\item for each ground instance of a distributional fact of the form \linebreak \probloginline{rand~dist(d@$_1$@,...,d@$_n$@)} it holds that   $rank(\text{\probloginline{rand}})>rank(\text{\probloginline{d@$_i$@}})$ for every $i$ ($1\leq i \leq n$).
			\item for each ground instance of another program clause \probloginline{h:- b@$_1$@,...,b@$_n$@} it holds that $rank(\text{\probloginline{h}})\geq rank(\text{\probloginline{b@$_i$@}})$, for all $i$.
			\item for each ground term \probloginline{b} that contains a random variable \probloginline{rand}, $rank(\text{\probloginline{b}})\geq rank(\text{\probloginline{rand~dist}})$ (with \probloginline{rand~dist} the head of the distributional clause defining \probloginline{rand}).
		\end{enumerate}
	    \item[V3] \todo{Lebesgue measurabiltiy of comparisons}
	\end{itemize}
\end{theorem}

\proof{\todo{}

\begin{itemize}
    \item use transformation from \dcproblogsty to DC-PLP and show uniqueness under conditions
    \item we need to show that the transformation is unique (order does not matter) and results in a valid DC-PLP program
\end{itemize}
}

\todo{Discuss complexity/decidability of determining whether program is valid, cf. \citep[Chapter 4.5]{milch2006probabilistic}. discuss sufficiency not necessity of validity conditions. maybe disucss in dc-plp section }

\section{\dcproblogsty with Distributional Clauses}

\begin{example}[\dcproblogsty Program] \label{ex:dcproblog_distributional_clause}
	This example program models the correct working of a machine. The probability distribution of the \probloginline{temperature} of the machine depends on whether it is a \probloginline{hot} day or not.
	\begin{problog*}{linenos}
machine(1).

0.2::hot.
0.99::cooling(1).

temperature ~ normal(27,5):- hot.
temperature ~ normal(20,5):- \+hot.

works(N):- machine(N), cooling(N).
works(N):- machine(N), temperature<25.0.

query(works(1)).
	\end{problog*}
	Note that we introduced {\em distributional clauses} in this example, which let us write down conditional probability distributions. 
\end{example}

\begin{definition}
	Distributional clauses are syntactic sugar to concisely write down conditional probability distributions of random variables. \todo{more elaborate}

\end{definition}

\begin{definition}

\todo{formal transformation: \dcproblogsty with distributional clauses to \dcproblogsty without}
	A \dcproblogsty program containing a random variable \probloginline{rand} involved in $N$ distributional clauses
    \begin{problog}
rand ~ dist@$_i$@ :- b@$_i$@
    \end{problog}
	with mutually logically inconsistent bodies $\text{\probloginline{b@$_i$@}}$ ($1{\leq} i {\leq} N$)
	is equivalent to a \dcproblogsty program with $N$ fresh random variables \probloginline{rand@$_i$@} ($1{\leq} i {\leq} N$) involved in $N$ distinct distributional facts
    \begin{problog}
rand@$_i$@ ~ dist@$_i$@.
    \end{problog}
    Each clause initially containing \probloginline{rand} is replaced in the equivalent program that does not contain distributional clauses for \probloginline{rand} by $N$ clauses, where \probloginline{rand} is substituted by $\text{\probloginline{rand@$_i$@}}$ and its body is conjoined with \probloginline{b@$_i$@} for every i with $1{\leq} i {\leq} N$.

\end{definition}

Mapping distributional clauses to distributional facts is akin to mapping intentional probabilistic facts (e.g \probloginline{0.2::a:- b.}) to probabilistic facts (and deterministic rules). Note that \dcproblogsty allows, just as \problogsty, the usage of intentional probabilistic facts, with the same semantics as in \problogsty. In absence of any distributional facts and clauses, a \dcproblogsty program reduces to a \problogsty program.

\begin{theorem}\label{theo:valid_dcproblog_w_clauses} A {\bf \dcproblogsty program} \dcpprogram{} {\bf with distributional clauses} is {\bf valid} if it can be mapped to an equivalent valid \dcproblogsty program. Such a mapping exists if the following three validity conditions hold.
	\begin{itemize}
		\item[V1] The bodies of ground distributional clauses with the same random variable in their heads are pairwise logically inconsistent.
		\item[V2] The program \dcpprogram is distribution stratified, that is, there exists a function $rank(\cdot)$ that maps ground terms to $\mathbb{N}$ and that satisfies the following properties:
		\begin{enumerate}			
			\item for each ground instance of a distributional clause of the form \linebreak \probloginline{rand~dist(d@$_1$@,...,d@$_n$@):-b} it holds that   $rank(\text{\probloginline{rand}})>rank(\text{\probloginline{d@$_i$@}})$ for every $i$ ($1\leq i \leq n$).
			\item for each ground instance of a distributional clause \probloginline{rand~dist:-b@$_1$@,...,b@$_n$@} it holds that $rank(\text{\probloginline{rand ~ dist}})>rank(\text{\probloginline{b@$_i$@}})$, for all $i$.
			\item for each ground instance of another program clause \probloginline{h:- b@$_1$@,...,b@$_n$@} it holds that $rank(\text{\probloginline{h}})\geq rank(\text{\probloginline{b@$_i$@}})$, for all $i$.
			\item for each ground term \probloginline{b} that contains a random variable \probloginline{rand}, $rank(\text{\probloginline{b}})\geq rank(\text{\probloginline{rand~dist}})$ (with \probloginline{rand~dist} the head of the distributional clause defining \probloginline{rand}).
		\end{enumerate}
	    \item[V3] \todo{Lebesgue measurabiltiy of comparisons}
	\end{itemize}

\proof{transform distributional clauses to distributional facts.

}
\end{theorem}

The criterion V1 in Theorem~\ref{theo:valid_dcproblog_w_clauses} is equivalent to the validity criterion V1 given in~\citep[Definition 3]{gutmann2011magic}, which requires that there is a unique ground distribution for each ground random variable \probloginline{rand}. V2 is a carbon copy of the second validity criterion of~\citep[Definition 3]{gutmann2011magic}, with the exception of the first point of V2. We added it, as distributional clauses with cyclic functional dependencies were not {\em explicitly} disallowed in ~\citep{gutmann2011magic}.

\subsection*{Difference to Distributional Clauses}

\ak{discussion on differences to DC: in how far is this semantics vs implementation? I believe we silently assumed (but never stated) that only defined RVs would be used, but I don't remember how Davide handled this when introducing negation to the semantics...}

This is also a good point to illustrate a major difference between the semantics of random variables in Distributional Clauses and \dcproblogsty. Take the code snippet below:
\begin{problog*}{linenos}
x ~ normal(20,2).

q(1):- x>20.
q(2):- y>20.

query(q(1)).
query(q(2)).
\end{problog*}
The first query succeeds and returns as an answer $p(\text{\probloginline{query(q(1)))}}) = 0.5$. The second query, however, throws an error as there is no term of type \type{RandomVariable} declared that is named \probloginline{y}. The second query would also fail if there were an \type{Atom} term named \probloginline{y}. In this case the lefthand side of \probloginline{y>20} would be ill-typed.

Let us now write the same program as a Distributional Clauses program:
\begin{adjustwidth}{1em}{0em}
	{
		\normalfont
		\begin{lstlisting}[breaklines,mathescape]
x ~ normal(20,3).

q(1)$\leftarrow$ $ {\simeq}(x)$<20.
q(2)$\leftarrow$  $ {\simeq}(y)$<20.

query(q(1)).
query(q(2)).
		\end{lstlisting}
	}
\end{adjustwidth}
The first query will again return $ p(\text{\lstinline|query(q(1))|}) =0.5$ (or an approximation thereof) but the second query will not throw an error but return a probability of zero. How does this work? The term \lstinline[mathescape]|$ {\simeq} (y)$| tries to unify with the value of the random variable \lstinline|y|. As there is no such random variable, this silently fails and the body of the clause whose head is \lstinline|q(2)| is simply not satisfied.

If we were to negate the comparison predicate in the body of the \lstinline|q(2)| clause, \ie write:
\begin{adjustwidth}{1em}{0em}
	{
		\normalfont
		\begin{lstlisting}[breaklines,mathescape]
q(2) $\leftarrow$  \+(${\simeq}(y)$<20).
		\end{lstlisting}
	}
\end{adjustwidth}
we would obtain for the second query: $ p(\text{\lstinline|query(q(1))|}) =1$. Negating the corresponding comparison predicate in \dcproblogsty would still result in an ill-typed term, \ie an error would be thrown.

The comparison we just performed does not only illustrate a semantic difference between \dcproblogsty and Distributional Clauses but also a conceptual one. In \dcproblogsty the values of of random variables never enter the realm of the logic program: values of random variables only live in the external arithmetic engine of \dcproblogsty. In contrast,  random variables in Distributional Clauses are part of the logic program: the values of random variables are stored in a logic data base and can be accessed through logical unification. Distributional Clauses currently does this via SLD resolution.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/early_examples.tex
aij dcproblog paper

\section{Motivating examples}
\dcproblogsty is a probabilistic logic programming language derived from two related but separate probabilistic logic programming languages,  \problogsty and Distributional Clauses (DC).
It inherits its Prolog syntax and the concept of probabilistic facts from \problogsty, and combines this with the ideas to define and query continuous random variables in DC, using DC's infix 
\probloginline{~}\lstinline[columns=fixed]|/2| predicate to define  {\em distributional facts}. Distributional facts allow a user to express random variables that are distributed according to a certain probability distribution specified by a random function symbol.

In the following example, the distributional fact in line 3 introduces the random variable \verb|temperature| and associates it with the distribution term \verb|normal(20,5)|. This random variable is then queried in the clause in line 8 using comparison predicate \verb|</2|, again written in infix notation. 
\begin{example}\label{example:dcproblog_machine}
	In this example we model the temperature as a continuous random variable distributed according to a normal distribution with mean $20$ and standard deviation $5$ (Line \ref{program:dcproblog_machines_temp}).
	\begin{problog*}{linenos}
machine(1). machine(2).

temperature ~ normal(20,5). @\label{program:dcproblog_machines_temp}@
0.99::cooling(1).
0.95::cooling(2).

works(N):- machine(N), cooling(N).
works(N):- machine(N), temperature<25.0. @\label{program:dcproblog_machines_work_temp}@

evidence(works(2)).  @\label{program:dcproblog_machines_evidence}@
query(works(1)).
	\end{problog*}
\end{example}

Distributional facts are not limited to describe continuous random variables but can, for example, also model discrete probability distributions, of which an example would be the Poisson distribution.
\begin{example}\label{example:dcproblog_poisson_population}
	In this example we show how to model the size of a group of people as a Poisson distribution and answer queries about the size of the group using the comparison predicates  \probloginline{>}\lstinline[columns=fixed]{/2} and \probloginline{=:=}\lstinline[{columns=fixed}]{/2}.
	\begin{problog*}{linenos}
n_people ~ poisson(6).

more_than_five:- n_people>5@.@
exactly_five:- n_people=:=5@.@

query(more_than_five).
query(exactly_five).
	\end{problog*}

\end{example}

We will formally introduce \dcproblogsty in Section~\ref{}. To introduce the semantics, in the next Section, we start by introducing DC-PLP, an abstract version of the language that omits syntactic sugar.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/implementation.tex
aij dcproblog paper

\section{Implementation} \label{sec:implementation}
\ak{needs an introduction (and I'd suggest reorganization into algorithms first, then system)}
We make two observations:
\begin{enumerate}
    \item Every implementation of \prologsty  (or any other logic programming language)  reserves a set of {\em  system predicates} executing code not written in pure \prologsty~\citep[Chapter 8]{sterling1994art}.
    System predicates complement pure \prologsty implementations and give users access to efficient algorithms that are cumbersome to implement in pure logic programming -- for instance algorithms for numeric computations.
    \item In Section~\ref{sec:semantics} we have seen that we can separate a probabilistic logic program into a set of independent random choices and a logic program. The independent choices and the logic program communicate to each other via comparison predicates. Comparison predicates are an instance of system predicates.
\end{enumerate}

These observations hint at the possibility of implementing \dcproblogsty as a deterministic logic programming language with access to an external engine with support for random variables.
On an implementation level, the key difference between a probabilistic logic programming language and a (deterministic) logic programming language is the availability of probability distributions in the external engine of the former.

Decoupling independent choices (random variables) and the deterministic system (logic program) does also mean that we should be able to take a probabilistic logic programming language that only supports finite discrete probability distributions and extend it with continuous probability distributions.
We will demonstrate how to achieve this by adding continuous random variables to the ProbLog2 system~\citep{dries2015problog2} (the most recent implementation of \problogsty). We dub our implementation \dcproblogsys.

% First, we will introduce a type system 
% (Section~\ref{sec:type_system}), followed by a description of how a type system aids the extension of the ProbLog2 system with continuous random variables (Section~\ref{sec:external_engine}).
% Followed by a description of the system architecture of \dcproblogsty in Section~\ref{sec:architecture}, which is based on the \problogsys architecture~\citep{dries2015problog2} and which we dub \dcproblogsys.

% To this end we will first introduce a type system (Section~\ref{sec:type_system}), followed by a description of how a type system aids the extension of the ProbLog2 system with continuous random variables (Sections~\ref{sec:multiple_dispatch} and~\ref{sec:external_engine}).
% Lastly, we will also discuss the implementation of ALW (cf. Section~\ref{sec:inference}) and how it uses the existing inference mechanisms already present in the ProbLog2 system.

\subsection{\dcproblogsty System Architecture}
\label{sec:architecture}

\dcproblogsys, is built on top of the existing \problogsys system. This means we inherit the general architecture of performing inference via a number of deterministic program transformations performed on the source program (written in \dcproblogsty). Figure~\ref{figure:dcproblog_sys} summarizes the program transformation steps in \dcproblogsys. These are grounding, cycle breaking, compiling, and evaluating. Conceptually the transformations are identical to the ones performed in the \problogsys implementation.
% We will briefly describe the four transformations and give a more detailed account of the compilation and evaluation steps in Section~\ref{sec:inference} as these are the steps where \dcproblogsys and \problogsys differ most.

\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/dcproblog_sys.pdf}
	\caption[Overview of the program transformation steps in the \dcproblogsys system]{Overview of the program transformation steps in the \dcproblogsys system. (Figure inspired by~\protect\citep[Figure 2]{zuidbergdosmartires2019transforming}.}
	\label{figure:dcproblog_sys}
\end{figure}

%%% moved from elsewhere
\subsection{Conditional \dcproblogsty Programs}
\ak{this is not the right place for the sec; just moved it from the language section as it is about system}

This lets us, for instance, compute the probability of the query in Example~\ref{example:dcproblog_machine}, where we saw that we can express a conditional probability by using the reserved \probloginline{evidence/1} predicate. \dcproblogsty also has a binary predicate \probloginline{evidence/2}, where the second argument is either the \probloginline{true} symbol or the \probloginline{false} symbol. The former is equivalent to using the unary version, while the latter allows one to express negative evidence. These predicates are also present in \problogsty.

While the \probloginline{evidence} predicate allows us to express conditional probabilities where we condition on Boolean random variables, its semantics does not extend directly to conditioning on continuous random variables: a continuous random variable can neither be true nor false. In order to allow a user to condition on continuous random variables in \dcproblogsty, we introduce the \probloginline{observation/2} predicate.

\begin{example}\label{example:dcproblog:observation}
	We model the size of a ball as a mixture of  different beta distributions, depending on whether the ball is made out of wood or metal (Line~\ref{program:dcproblog_machines_observation:ad})\footnote{Annotated disjunctions are used to concisely write down mutually exclusive Boolean random variables. Internally they are compiled down to probabilistic facts and deterministic rules.}.
	We would now like to know the probability of the ball being made out of wood given that we have a measurement of the size of the ball. In order to condition on a continuous random variable we introduce the \probloginline{observation/2} predicate, which has an analogous functionality as the evidence predicates for Boolean random variables.
	\begin{problog*}{linenos}
3/10::material(wood);7/10::material(metal).@\label{program:dcproblog_machines_observation:ad}@

size~beta(2,3):-material(metal)@\label{example:dcproblog:observation:beta23}@.
size~beta(4,2):-material(wood).

observation(size,4/10).
query(material(wood)).
	\end{problog*}
	This \dcproblogsty program encodes the conditional probability:
	\begin{align}
		p(\text{\probloginline{material(wood)}}| \text{\probloginline{size=4/10}})
	\end{align}
\end{example}

%%% end moved from elsewhere

\ak{if we keep this overview, include pointers to the sections that discuss details}
\paragraph{\bf Grounding}

\dcproblogsys is an extension of the \problogsys system: the logic programming components in both systems are identical. This means that we can simply bootstrap the logic grounding engine of the \problogsys system for \dcproblogsys. Grounding is performed using {\em selective linear definite clause} (SLD) resolution~\citep{kowalski1974predicate,sterling1994art}. SLD resolution allows for computing the relevant ground program for a specific query to the logic program, which avoids grounding of unnecessary predicates~\citep{kersting2000bayesian,fierens2015inference}.\ak{what about negation? SLDNF or something else?}

\paragraph{\bf Cycle Breaking} Logic programs may contain cycles. In order to perform inference these need to be removed from the ground program. As these cycles only occur in the logic program part of a \dcproblogsty program we can again bootstrap the cycle breaking available in \problogsys, which is currently a variation of the algorithm proposed in~\citep{janhunen2004representing} \citet{mantadelis2010dedicated} have also proposed an alternative cycle breaking algorithm.

\paragraph{\bf Compiling} The next step that \problogsys performs to compute probabilities is the compilation of acyclic ground programs. Acyclic ground programs have a one-to-one mapping to propositional logic formulas, which are then compiled into decision diagrams in a process dubbed knowledge compilation~\citep{darwiche2002knowledge}. \dcproblogsys differs in that acyclic ground \dcplpsty programs are not mapped to propositional logic formulas but to so-called {\em satisfiability modulo theory} (SMT) formulas (\eg $x>3$)~\citep{barrett2009handbook}, which are then compiled to decision diagrams that support algebraic constraints~\citep{sanner2011symbolic,zuidbergdosmartires2019exact}.

\paragraph{\bf Evaluating} The probability of a query to a (hybrid) logic program is obtained by evaluating the compiled logic formula (decision diagram). While in \problogsys this is performed using the weighted model counting (WMC)~\citep{chavira2008probabilistic} framework, \dcproblogsys uses weighted model integration~\citep{belle2015probabilistic} -- an extension of WMC that allows also for algebraic constraints on random variables.

\bigbreak

We provide the necessary background on knowledge compilation, weighted model counting, weighted model integration, and SMT formulas in~\ref{app:amc}, \ref{app:compilation}, and
\ref{app:wmi}.

\subsection{Grounding}
As already mentioned, grounding in \dcproblogsys is performed by bootstrapping the existing grounding engine of \problogsys. The sole minor difference concerns the grounding of comparison predicates, \eg (\probloginline{1<2}), (\probloginline{x>20}).
During the grounding of a \dcproblogsty program, the types of the comparison's arguments are checked and the comparison is considered to be proven if the types of the arguments match the type signature of the comparison predicate.
The possible types of a comparison's predicate arguments are dictated by the external arithmetic engine. For instance, \prologsty only allows for (deterministic) numbers, in contrast to \dcproblogsty where the external arithmetic engine also allows for comparing random variables.\ak{what if typecheck fails?}

More concretely, if a comparison predicate of the form (\probloginline{x>20}) is encountered during the grounding, the engine checks whether there are distributional facts of the form \probloginline{x ~ Dist} present in the program. If such a clause is present, the type of \probloginline{x} is inferred from the distribution \probloginline{Dist}. The comparison predicate is proven if the external engines allows for comparing the inferred types, \eg can the external arithmetic engine compare continuous random variables to real valued numbers. We give a more detailed account of typing and the external arithmetic engine in \ref{app:types}.

\subsection*{Evaluation of Comparison Predicates}

A distinctive property of comparison predicates in \dcproblogsty is that they have a potential infinite and uncountable number of groundings, \ie an infinite number of possible assignment to the variables in the arguments that would satisfy the comparison\ak{careful with grounding (replacing logical variable by ground terms) vs evaluating (replacing random variable by its value)}.
This cannot to be carried out using a logic grounding engine.
Therefore, we have to defer the grounding of comparison predicates to a later stage of the inference mechanism and at the same time retain a symbolic representation of the set containing all possible groundings. Probabilistic inference then constitutes of {\em measuring} the size of these sets using integration.
We call the grounding of a system predicate also an evaluation, as the external engines performs an arithmetic evaluation in the background.

In the special case of a `deterministic' comparison, \ie a comparison not containing any random variables in its arguments, we can interleave the grounding with the evaluation of the predicate.
That is, we can replace the atom with its corresponding Boolean value. For the comparison (\probloginline{1<2}) we would get \probloginline{true}. Such a type dependent eager evaluation during grounding can be regarded as an instance of multiple dispatch: differently typed predicates induce different run time behavior~\citep{castagna1995calculus}.

Eager evaluation of arithmetic expressions makes most sense when this leads to simplifications of the ground program. More concretely, when a comparison predicate is not satisfied this results in a \probloginline{false} atom and proving a goal during grounding can be stopped pre-maturely.
Eager evaluation of the comparison predicates allows \dcproblogsty to retain the usual behavior of proving and grounding in \problogsty where a proof fails if a (deterministic) comparison is not satisfied.

Depending on the arithmetic engine deployed, eager evaluation is also possible for evaluations of non-deterministic arithmetic expressions. If the evaluation of arithmetic expressions is purely sampling-based, the random variables in an arithmetic expression are replaced by sampled values. For each sample the arithmetic expressions can then be eagerly evaluated. This is the approach taken by~\citet{nitti2016probabilistic}.\ak{we need to be more explicit about this entire "external engine" stuff; right now, remarks about alternative engines are mingled into the description of a pipeline that seems to be fixed to one such engine (sampling here, Prolog earlier)}

% For symbolic arithmetic engines it does also make sense to perform eager evaluations to some degree. Consider the comparison (\probloginline{3+5>2+1}). A purely symbolic external engine would evaluate two symbolic expression trees and wrap the comparison into a \probloginline{holds}\lstinline[columns=fixed]|/1| predicate. Eagerly evaluation the expression tree however lets us replace the \probloginline{holds(3+5>2+1)} with \probloginline{false}.

\begin{example}[\dcproblogsty grounding] \label{example:grounding}
Consider the non-ground \dcproblogsty program below.

	\begin{problog*}{linenos}
0.2::hot.
0.99::cooling(1).

temperature(1) ~ normal(27,5):- hot.
temperature(1) ~ normal(20,5):- \+hot.

works(N):- cooling(N).
works(N):- temperature(N)<25.0.

query(works(1)).
	\end{problog*}
The grounding mechanism in \dcproblogsys is the same one as in \problogsys and only works on pure logic programs. However, the logic program and the random variables are still interleaved. In a first step we remove syntactic sugar and rewrite the distributional clauses and probabilistic facts  as distributional facts, cf. Section~\ref{sec:semantics}.

	\begin{dcplp*}{linenos}
rv_1_hot ~ flip(0.2). @\label{line:dist_hot}@
map(rv_hot,rv_1_hot).
hot:- map(rv_hot,R), R=:=1.

rv_1_cooling(1) ~ flip(0.99). @\label{line:dist_cooling}@
map(rv_cooling(1),rv1_cooling(1)).
cooling(1):- map(rv_cooling(1),R), R=:=1.

rv_1_temperature(1) ~ normal(27,5). @\label{line:dist_temp_1}@
map(rv_temperature(1), rv_1_temperature(1)):- hot.

rv_2_temperature(1) ~ normal(20,5). @\label{line:dist_temp2}@
map(rv_temperature(1), rv_2_temperature(1)):- \+hot.

works(N):- cooling(N).
works(N):- map(rv_temperature(N),R) , R<25.0.

query(works(1)).
	\end{dcplp*}
After removing the syntactic sugar we have a \dcplpsty program constituted of distributional facts in Lines~\ref{line:dist_hot}, \ref{line:dist_cooling}, \ref{line:dist_temp_1} and \ref{line:dist_temp2}, and a (non-probabilistic) logic program in the remaining lines. The logic  part of the \dcplpsty program can now be grounded using the \problogsys grounding engine:

	\begin{dcplp*}{linenos}
rv_1_hot ~ flip(0.2). 
rv_1_cooling(1) ~ flip(0.99). 
rv_1_temperature(1) ~ normal(27,5).
rv_2_temperature(1) ~ normal(20,5).

map(rv_hot,rv_1_hot).
map(rv_cooling(1),rv_1_cooling(1)).
map(rv_temperature(1), rv_1_temperature(1)):- hot.
map(rv_temperature(1), rv_2_temperature(1)):- \+hot.

hot:- map(rv_hot,), rv_1_hot=:=1.
cooling(1):-
    map(rv_cooling(1),rv_1_cooling(1)), 
    rv_1_cooling(1)=:=1.
works(1):- cooling(1).
works(1):-
    map(rv_temperature(1),rv_1_temperature(1)) ,
    rv_1_temperature(1)<25.0.
works(1):-
    map(rv_temperature(1),rv_2_temperature(1)) 
    rv_2_temperature(1)<25.0.

query(works(1)).
	\end{dcplp*}
As there are no cycles present in the \dcplpsty program above, we do not need to perform cycle breaking on the logic program part.
\end{example}

Note that for, the sake of clarity, we performed the rewriting of a \dcproblogsty program to a \dcplpsty program in Example~\ref{example:grounding} as a separate and preliminary step. In the \dcproblogsys implementation the elimination of syntactic sugar is interleaved with the grounding and performed on the fly. This avoids rewriting parts of the program that are irrelevant to computing the probability of a query.\ak{why/when is grounding on the fly possible?}

%%%%%%%%%%%%%%%%%%%%

% In probabilistic programming a recent trend has emerged of mixing eager evaluation (usually in combination with sampling based probabilistic inference) and lazy evaluation (also called delayed evaluation). Examples for this are the probabilistic programming languages Birch~\citep{murray2018automated} or the symbolic probabilistic programming language PSI~\citep{gehr2016psi}. Lazy evaluation of arithmetic expressions do also allow to simplify probabilistic programs such that inference becomes tractable after these simplifications. An obvious example for this is then rewriting a conjugate prior-likelihood pair into a single expression. Lazy evaluation combined with symbolic simplifications goes into the direction of compiling probabilistic programs, \eg~\citep{wu2016swift}.

% The key difference between the external arithmetic engine of \prologsty (or \problogsty) and the external arithmetic engine of \dcproblogsty is that the external engine of the latter is capable of performing arithmetic operations (perform evaluations) with random variables. Combining this innovation with multiple dispatch allows us to retain the same syntax (same functor) to perform arithmetic operations on \type{SymbolicRV} terms and/or \type{Number} terms.

% \subsubsection{Arithemtics of Random Variables}

% Evaluating arithmetic expressions that contain random variables should not to be confused with performing arithmetic operations on random variables. Writing \probloginline{x+y} in a comparison predicate, where \probloginline{x} and \probloginline{y} are random variables has the semantics of evaluating both random variables (for instance by drawing a sample) and adding up both evaluations.

% In order to perform arithmetic operations on random variables directly, we would need to introduce syntax and an arithmetic engine for random variables. This could, for example, be achieved in the following fashion:
% \begin{problog*}{linenos}
%  x ~ normal(20,5).
%  y ~ normal(25,6).
%  z ~ x+y.
% \end{problog*}
% In this case z would be a normal distribution with mean ($20{+}25$) and standard deviation ($5{+}6$). We leave arithmetic operations on random variables for future research.\pedro{maybe add citation of somebody who has done something like this already?}

%%%%%%%%%%%%%%%%%%%%

\subsection{Mapping Acyclic Ground \dcproblogsty to Weighted SMT Formulas}

In \problogsys, a query to a probabilistic program is mapped to a weighted propositional formula~\citep{fierens2015inference}. The probability of the query being satisfied is then the weight of the propositional formula. This weight is also called the {\em weighted model count}~\citep{darwiche2009modeling}.
Inference in \dcproblogsys follows a similar principle, with the difference that a hybrid probabilistic program is mapped to a weighted SMT formula, instead of a weighted propositional formula.
The weight of a weighted SMT formula is called the {\em weighted model integral}.
% Weighted model integration (WMI)~\citep{belle2015probabilistic} generalizes weighted model counting in that it also allows for continuous variables.

The probability of a query \probloginline{query(q)} in a \dcproblogsty program is obtained by computing the probability of an equivalent weighted SMT formula using weighted model integration:
\begin{align}
	p(\text{\probloginline{query(q)}})= p(\support_q) = \wmi(\support_q, w_q)
\end{align}
where $\support_q$ denotes the SMT formula and $\weight_q$ the associated weight.

% We first encode a query to a \dcproblogsty program as a weighed logic formula and present then algebraic likelihood weighting (ALW), a probabilistic inference algorithm that computes the weight of a logic formula in the discrete-continuous domain in the context of probabilistic logic programming.

% We just mapped the probability encoded by a \dcplpsty program to a weighted model integral. Inference in \dcplpsty/\dcproblogsty can hence be performed by any algorithm capable of computing the weighted model integral in question.

\begin{example}[Mapping \dcplpsty to WMI] Consider again the \dcplpsty program in Example~\ref{ex:dcproblog_program_without_dc_plpdc}. If we ground the purely logic component of the program we obtain the ground program .
\begin{dcplp*}{linenos}
hot ~ flip(0.2). 
cooling(1) ~ flip(0.99). 
temperature(hot) ~ normal(27,5).
temperature(not_hot) ~ normal(20,5).

machine(1).
works(1):- machine(1), cooling(1)=:=1.
works(1):-
	machine(1),
	temperature(hot)<25.0,
	hot=:=1. 
works(1):-
	machine(1),
	temperature(not_hot)<25.0,
	hot=:=0.

query(works(1)).
	\end{dcplp*}
	We map the acyclic ground logic program and the Boolean queries to an SMT formula $\support_q$:
	\begin{align}
		\support_q
		&\leftrightarrow \text{\dcplpinline{machine(1)}} \land (\text{\dcplpinline{cooling(1)}}=1) \lor \nonumber \\
		&\phantom{{}\leftrightarrow{}} \text{\dcplpinline{machine(1)}} \land (\text{\dcplpinline{temperature(hot)}}<25.0) \land (\text{\dcplpinline{hot}}=1) \lor \nonumber \\
		&\phantom{{}\leftrightarrow{}} \text{\dcplpinline{machine(1)}} \land (\text{\dcplpinline{temperature(not_hot)}}<25.0) \land  (\text{\dcplpinline{hot}}=0) \nonumber \\
		&\leftrightarrow (\text{\dcplpinline{cooling(1)}}=1) \lor \nonumber \\
		&\phantom{{}\leftrightarrow{}} (\text{\dcplpinline{temperature(hot)}}<25.0) \land (\text{\dcplpinline{hot}}=1) \lor \nonumber \\
		&\phantom{{}\leftrightarrow{}} (\text{\dcplpinline{temperature(not_hot)}}<25.0) \land  (\text{\dcplpinline{hot}}=0)  
	\end{align}
	The weight function $w_q$ is simply obtained by multiplying together the functions corresponding to the random variables declared in the \dcplpsty program.
\end{example}

\subsection{Correctness of the Transformations}

In Algorithm~\ref{algo:inference_dc_problog} we outline the steps that are performed to map a \dcplpsty program to a weighted SMT formula. They closely follows the steps presented in~\citep[Section 5]{fierens2015inference}. 
\ak{the query should be an inout of the alg as well (note that the Boolean queries in the program are the comparison atoms, not the inference query)}

\begin{mdframed}
	\begin{algo}[\dcplpsty to Weighted SMT] \label{algo:inference_dc_problog}
		The algorithm takes as input a \dcplpsty program \dcpprogram, consisting of a countable set $P$ of ground distributional facts, a countable set of measurable Boolean queries $Q$, and a purely logic program $L$. The algorithm then maps $\dcpprogram$ through the following steps to a weighted SMT formula $(\support_q, \weight_q)$.
		\begin{enumerate}
			\item Ground out the purely logic program $L$ of $\dcpprogram$ with respect to a query $q$ and obtain the ground program $L_g$. This will also ground the Boolean queries in $Q$, leading to $Q_g$.
    		\item Break cycles in the ground logic program $L_g$.
			\item Convert the ground logic program consisting of $L_g$ and the set of measurable queries $Q_g$ to an equivalent SMT formula $\support_q$.
			\item Define a weight function $\weight_q$, which corresponds to the countable set of random variables in $P$.
		\end{enumerate}
	\end{algo}
\end{mdframed}

\begin{theorem}
The probability of a query $q$ to a valid \dcplpsty program $\dcpprogram$ can be expressed as a weighted model integral of a weighted SMT formula $(\support_q, \weight_q)$.\ak{too vague for a thm? why not make it concrete?}
\end{theorem}

\proof{The proof consists of showing that we can formulate the relevant ground program $L_g$ of the program $\dcpprogram$ with respect to the query $q$ as an SMT formula, that we can map the  }

\todo{work this in after Section 2 and 3 are written}

Let us consider the definition of weighted model integration given in \ref{app:wmi} and in particular the formulation given in Equation~\ref{eqn:wmi_sum_ive_weight}.
\begin{align}
	\wmi(\support, \weight) = \sum_{\langle\wsupport_i,\wweight_i\rangle \in \wtuples_f} \sum_{\bvars} \int \ive{\wsupport_i(\xvars, \bvars)} \wweight_i(\xvars) d\xvars 
\end{align}

Comparing the definition of the probability of a \dcplpsty program to the definition weighted model integration, and in particular the form of WMI given in Equation~\ref{eqn:wmi_sum_ive_weight}:
\begin{align}
	\sum_{\langle\wsupport_i,\wweight_i\rangle \in \wtuples_f} \sum_{\bvars} \int \ive{\wsupport_i(\xvars, \bvars)} \wweight_i(\xvars) d\xvars 
\end{align}
We cannot but appreciate the striking similarity between both Equations. The only substantial difference is that Equation~\ref{eq:definition_probability_dcplp} uses Lebesgue integration to marginalize out Boolean random variables, whereas Equation~\ref{eqn:wmi_sum_ive_weight} uses a summation. Other than that it is straight forward to map one equation to the other:

\begin{itemize}
	\item We associate the sum over the models $l \in MOD(L)$ to the sum over different SMT(\lra) formulas $\wsupport_i$ in weighted model
	\item We associate a specific product of Iverson brackets of Boolean queries $\prod_{q \in Q(l)} \ive{q}$ to the Iverson bracket of a specific SMT(\lra) formula $\ive{\wsupport_i}$, which is a conjunction of atomic SMT(\lra) formulas.
	\item We associate the distributions in $ D_l$ with the weight function $\wweight_i$.
\end{itemize}
\todo{end}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/inference.tex
aij dcproblog paper

\section{Computing Probabilities}\label{sec:inference}

In the previous section we saw how to implement \dcproblogsty using components of the existing \problogsys system. More specifically, we delineated how the first two program transformations in the \dcproblogsys implementation (grounding and cycle breaking) relate to analogous steps in \problogsys.
In order to calculate probabilities, we will now show how the compilation and evaluation steps from \problogsys can be adapted in such a way that they can accommodate continuous random variables as well.

\subsection{Conditional Queries}

\ak{this section is semantics; at least the conceptual parts discussing different types of observations etc should probably move to sec \ref{sec:inference-tasks}}

For \dcproblogsty programs that include evidence on logic atoms the probability is simply obtained by computing two weighted model integrals, one for the evidence $\support_e$ and one for the conjunction of the evidence with the query $\support_e \land \support_q$:
\begin{equation}
	p(\support_q|\support_e)=\frac{p(\support_q,\support_e)}{p(\support_e)}=\frac{\wmi(\support_q \land \support_e,w_{qe})}{\wmi(\support_e,w_e)}
\end{equation}

As the probability of a random variable with uncountably many outcomes taking a specific value is zero, conditioning with \probloginline{observation/2} is slightly more complicated. Consider the program in Example~\ref{example:dcproblog:observation}. The probability of the the conditioning set is zero:
\begin{equation}
	p(\text{\probloginline{observation(size,4/10))}}=p(\text{\probloginline{size}=4/10})=0
\end{equation}
Were we to apply the same approach as for conditional queries with \probloginline{evidence}, we would obtain a division by zero and consequently an undefined conditional probability. 
This problem is related to the Borel-Kolomogorov paradox~\citep{kolmogorov1950foundations,gyenis2017conditioning}, which is again caused by a zero-division.

We will, analogously to \citet{nitti2016probabilistic},   follow~\citet{kadane2011principles} to resolve the division by zero issue.
We make the explicit distinction between the probability $P$ and the probability density function $p:\frac{d}{dx}P(X \leq x)=p(x)$. The conditional probability is defined as:
\begin{align}
	P(\support_q| \support_e)
	&=P(\support_q|rv=v) & (\support_e \leftrightarrow rv=v )\\
	&=\lim_{\Delta v\rightarrow 0} \frac{P(\support_q,rv \in [v-\nicefrac{\Delta v}{2},v+\nicefrac{\Delta v}{2}])}{ P(rv \in [v-\nicefrac{\Delta v}{2},v+\nicefrac{\Delta v}{2}])} \label{equ:borel_limit} \\
	&=\lim_{\Delta v\rightarrow 0} \frac{\int \ive{ \support_q(x,rv=v)} p(x,rv=v)dx \cancel{\Delta v}}{p(rv=v)\cancel{\Delta v}} \\
	&=\frac{\int \ive{ \support_q(x, rv=v)} p(x,rv=v)dx}{p(rv=v)}
\end{align}
This means that we do not divide anymore by zero but by the number obtained when evaluating the probability distribution at the observed point. For example, when evaluating \probloginline{beta(2,3)} (cf. Example~\ref{example:dcproblog:observation}, Line~\ref{example:dcproblog:observation:beta23}) at $4/10$  we get $1.7280$ instead of zero.
% Note that this is not a probability between zero and one.

In order to guarantee that the limit in Equation~\ref{equ:borel_limit} exists, we need to restrict the type signature of the \probloginline{observation} predicate. The  first argument is a {\em named} random variable and the second argument a number, cf. Example~\ref{example:dcproblog:observation}.
Relaxing this restriction might lead to the non-existence of the limit. Exactly this situation gives rise to the Borel-Kolmogorov paradox emerges.

Nevertheless, some languages lift this restriction, which results in possibly semantically ill-defined programs, such as in Distributional Clauses~\citep[Section 3.2]{nitti2016probabilistic}. A more sophisticated method to tackle the non-uniqueness of the conditional probability is {\em disintegration}~\citep{shan2017exact}, where, instead of returning a conditional probability, a family of conditional probabilities is returned (representing an uncountable number of limits). Other languages, such as BLOG, simply assume the existence of the limit~\citep{wu2018discrete}.

The type signature we impose on the \probloginline{observation/2} predicate might seem too restrictive for the purpose of probabilistic programming. Take for example the following conditional probability~\citep{nitti2016probabilistic}:
\begin{align}
	P(nationality| height {=} 160 \lor height {=} 180) \label{eqn:disj_conditioning}
\end{align}
This probability cannot be expressed using the \probloginline{observation/2} predicate. However, if we think about what a conditioning set represents in the context of probabilistic programming and machine learning, the example above appears quite odd. The conditioning set tells us which observations we have made about the world, \ie data that we collected using a physical device (\eg tape measure for the height of a person). The example in Equation~\ref{eqn:disj_conditioning} now tells us that our measuring device returned two numbers for a single measurement, which means that the measuring protocol is underspecified.

We just stated that a measurement consists of retrieving a number from a measuring device. This is not entirely correct. In the natural sciences a measurement consists always of a number and an upper and lower error bound, which effectively eliminates the problem of zero probability events. However, given the ubiquity of single point measurement, we feel that a probabilistic programming language that is not able to handle single point measurements is greatly hurt in its expressivity.

To conclude, in \dcproblogsty,  the \probloginline{observation/2} predicate allows a user to condition on zero probability events, while at the same time the imposed restrictions ensure that the encoded probability distribution is well defined.

\ak{conditional probabilities: just wondering: is it possible to condition on tests, \eg, observing that someone's height is at most 180? related to that, in (17), what if someone with bad handwriting measured the height and recorded it on a piece of paper, and I can't tell whether it is 160 or 180? :) }
\pedro{in this case the measuremnt is the pixels and you would have yes/no for each pixel}

%%%%%%%%%%%%%%%%%%%%

\subsection{Compiling SMT formulas}

State-of-the art methods that perform probabilistic inference with discrete random variables, divide inference into a computationally expensive step, which is followed by a computationally cheap one. The first step, which is also called knowledge compilation~\citep{darwiche2002knowledge}, takes a propositional logic formula and compiles it into a decision diagram (DD). The most prominent representatives of DDs are arguably {\em binary decisions diagrams}~\citep{bryant1986graph}. The advantage of compiling Boolean structures to decision diagrams is that probabilistic inference (the second step), \ie computing the weighted model count, becomes computationally tractable~\citep{chavira2008probabilistic}.

This idea of compiling logic formulas, in order to perform probabilistic inference, has also been mapped to (quantifier free) SMT formulas~\citep{kolb2018efficient,zuidbergdosmartires2019exact} and
has led to state-of-the art algorithms to perform inference in hybrid domains~\citep{derkinderen2020ordering}.

There are, however, two important differences between probabilistic inference via knowledge compilation using propositional logic formulas and SMT formulas. First, the possible target languages for the compilation step differ. We can use BDDs, d-DNNFs~\citep{darwiche2004ddnnf}, or SDD~\citep{darwiche2011sdd} for propositional logic but we need to resort to {\em  extended decision diagrams} (XDD) for SMT formulas~\citep{sanner2011symbolic,zuidbergdosmartires2019exact,kolb2019exploit}.

A second and more fundamental difference is the computational hardness of performing probabilistic inference in DDs vs. XDDs: computing the weighted model count on a DD can be done in polytime but computing the weighted model integral on an XDD is in general \#P-hard.

\subsection{Algebraic Likelihood Weighting}

Performing probabilistic inference with Boolean random variables boils down to performing computations in the so-called probability semiring. \citet{kimmig2017algebraic} formulated a generalization of probabilistic inference on Boolean formulas by using arbitrary semirings, which they dubbed algebraic model counting. 
\citet{zuidbergdosmartires2019exact} picked up this idea and also formulated probabilistic inference in the discrete-continuous domain (using SMT formulas) as a semiring computation by reformulating weighted model integration as an algebraic model counting problem. The advantage of this approach is that by simplifying an algberaic structure that adheres to the properties already proves the correctness of the inference algorithm.

% Probabilistic programming languages that successfully handle mixtures of discrete and continuous random variables, including conditioning with zero probability events deploy either purely symbolic inference algorithms~\citep{gehr2016psi,shan2017exact} or extend the likelihood weighting~\citep{fung1989weighing} algorithm leading to a hybrid symbolic-approximate inference algorithm~\citep{nitti2016probabilistic,wu2018discrete}.

In this section we frame the {\em lexicographic likelihood weighting} (LLW) algorithm~\citep{wu2018discrete} in an algebraic model counting context, which we dub algebraic likelihood weighting (\ialw). LLW is based on ideas presented in~\citep{nitti2016probabilistic} and combines sampling based probabilistic inference with symbolic inference to correctly handle mixtures of discrete and continuous random variables. The advantage of framing LLW  as an algebraic model counting problem is that the resulting algorithm becomes agnostic of the underlying inference mechanism (\eg Monte Carlo sampling, or symbolic integration). Moreover, we can simply reuse algorithms already developed in the WMI literature in order to perform inference in \dcproblogsty.

Following \citet{kimmig2017algebraic}, we first define a labelling function, which maps atomic SMT formulas to elements of a semiring, and continue by defining the semiring operations.

\begin{definition}[\ialw labeling function]
	\label{def:label_alpha_alw}
	Let be an atomic SMT formula. Then the label of the atomic formula $l$ is given by:
	\begin{align*}
		\alpha_{\ialw}(l) \coloneqq 
		\begin{cases}
% 			(p(l), 0)& \text{if $l$ is a Boolean variable} \\
			(p(x{=}v),1) & \text{if $l$ encodes \probloginline{observation(x,v)}, \ie $l\leftrightarrow (x{=}v)$} \\
			(\ive{l},0) & \text{otherwise} 
		\end{cases}
	\end{align*}
    In the first case $p(x{=}v)$ denotes the value of the probability distribution of the random variable $x$ evaluated at $v$.
	The label of a negated atomic SMT formula $ \neg l$ is given by:
	\begin{align*}
		\alpha_{\ialw}(\neg l) \coloneqq 
		\begin{cases}
% 			(1-p(l),0)& \text{if $l$ is a Boolean variable} \\
			(1,0) & \text{if $l$ encodes \probloginline{observation(x,v)}}\\
			(\ive{\neg l},0) & \text{otherwise} 
		\end{cases}
	\end{align*}
\end{definition}

\begin{example}[ALW labeling function]

\end{example}

\begin{definition}[\ialw semiring]
	The elements of the semiring $\mathcal{S}_{\ialw}$ are given by the set
	\begin{equation}\label{eqn:set_aws}
		\mathcal{A}_{\ialw}\coloneqq \left \{ (a, d) \right \}
	\end{equation}
	where $a$ denotes a real-valued weight (produce by the labeling function) and $d$ the number of probability densities (continuous probability distribution) that have contributed to $a$. The neutral elements $e^\oplus$ and $e^\otimes$ are defined as:
	\begin{align}
		e_{\ialw}^\oplus  \coloneqq (0,0)  &&
		e_{\ialw}^\otimes  \coloneqq (1,0)
	\end{align}
	For the addition and multiplication we define:
	\begin{align}
		(a_1,d_1) \oplus
		(a_2,d_2)
		&\coloneqq
		\begin{cases}
			(a_1+a_2,d_1) &\quad  \text{if $d_1=d_2$} \\
			(a_1,d_1) &\quad  \text{if $d_1<d_2$} \\
			(a_2,d_2) &\quad  \text{if $d_1>d_2$}
		\end{cases} \\
		(a_1,d_1) \otimes
		(a_2,d_2)
		&\coloneqq (a_1 \times a_2 , d_1+d_2)  &
	\end{align}
\end{definition}

\citet{zuidbergdosmartires2019exact} have shown that the weighted model integral is correctly computed on compiled logic structures that respect determinism (e.g XSDD, XADD), if the properties of a commutative semiring are respected and if the labeling function and the addition of the semiring form a neutral pair.

\begin{lemma} The structure $\mathcal{S}_{\ialw}=({\cal A}_{\ialw}, \oplus, \otimes, e^\oplus,e^\otimes)$ is a commutative semiring.
\end{lemma}

\begin{lemma} The pair  $(\oplus, \alpha_{\ialw})$ is neutral ($\alpha_{\ialw}(l)\oplus \alpha_{\ialw}(\neg l) = e_{\ialw}^{\otimes}$).
\end{lemma}

Performing a computation on a compiled SMT formula using the \ialw semiring gives a tuple of tuple of two elements, which we denote by 
$$\ialw(\support,\weight)=\Big( \ialw_1(\support, w), \ialw_2(\support, w) \Big)$$
We can now express probabilities of a conditional queries in terms of algebraic likelihood weights. 
\begin{align}
    p(\support_q|\support_e) &=
    \begin{dcases}
        \frac{\E[\ialw_1(\support_q \land \support_e, {\weight_{q,e}})]}{\E[\ialw_1(\support_e, \weight_e)]} & \text{if $\ialw_2(\support_q {\land} \support_e, \weight_{q,e})\geq \ialw_2(\support_e, \weight_e)$}\\
        0 & \text{else}
    \end{dcases} \\
     &=\begin{dcases}
        \frac{\wmi(\support_q \land \support_e, {\weight_{q,e}})]}{\wmi(\support_e, \weight_e)]} & \text{if $\ialw_2(\support_q {\land} \support_e, \weight_{q,e})\geq \ialw_2(\support_e, \weight_e)$}\\
        0 & \text{else}
    \end{dcases} 
\end{align}
$\E(\ialw_1(\cdot, \cdot))$ denotes the expected value of the first element of tuple obtained by the algebraic likelihood computation and is equal to the weighted model integral over the same SMT formula.

\begin{example}[ALW on a circuit]

\end{example}

\subsection{What about \problogsty's External Engine?}

Syntactically, the \dcplpsty assembler language does not make a difference between Boolean random variables (\probloginline{flip}\lstinline[columns=fixed]{/2}) and other random variables (\eg \probloginline{normal}\lstinline[columns=fixed]{/2}).
Furthermore, \dcplpsty is constituted of a logic program (deterministic system) and random variables (independent choices).
This indicates that implementations of probabilistic logic programming languages with and without continuous random variables should not be too different from each other.
After all, a probabilistic logic programming language with continuous random variables simply adds additional probability distributions besides the \probloginline{flip} distribution.

% Assume now that the algebraic model count of $\support_q \land \support_e$ and $\support_e$ evaluate respectively to:
% \begin{align}
% 	\AMC(\support_{q}\land \support_e , \alpha_{\ialw})&= (\amc_{qe}, d_{qe}) \label{eqn:amc_alw_qe} \\
% 	\AMC(\support_e , \alpha_{\ialw})&= (\amc_{e}, d_{e}) \label{eqn:amc_alw_e}
% \end{align}

% The conditional probability is then given by:
% \begin{align}
% 	p(\support_{q} | \support_e) =
% 	\begin{cases}
% 		\frac{\wmi(\support_q \land \support_e, \weight_{qe})}{\wmi(\support_e, \weight_{e})} & \text{if $d_{qe}=d_e$} \\
% 		0  & \text{if $d_{qe}>d_e$} 
% 	\end{cases} \label{eqn:conditional_wmi}
% \end{align}

% Note, in Equations~\ref{eqn:amc_alw_qe} to~\ref{eqn:conditional_wmi} we have omitted the dependence of the algebraic model counts and the weighted model integration on the Boolean and/or real variables.

% \todo{talk about difference BLOG and DC to ALW, make difference on algorithmic level}
% DC~\citep{nitti2016probabilistic} and BLOG~\citep{wu2018discrete} do return the correct probabilities as well, however, the advantage of using Sampo in conjunction with ALW gives515the correct result without drawing any samples.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/introduction.tex
aij dcproblog paper

\section{Introduction}
\ak{intro should get to the point more quickly; what is the exciting thing about this work? see Table \ref{tab:features} for a (partial) answer covering the language/semantics side, which would need to be extended to inference and implementation}

\begin{table}[ht]
    \centering
    \begin{tabular}{|p{0.3\textwidth}||c|c|c||c|c|}
    \hline
       Feature  & ProbLog & DC & DC & \dcproblogsty & \dcproblogsty  \\
       & & (Gutmann) & (Nitti) & (core) & (full)\\
       \hline\hline
       fully declarative semantics & y & y & n & y & y \\\hline
       negation & y & n & y & y & y \\\hline
       RVs with finite, countable and continuous domains & n & y & y & y & y\\\hline
       RVs with distributions parameterised by other RVs & n & y & y & y & y \\\hline
       RVs with distributions determined by logical context / (full) DCs & n & y & y & n & y \\\hline
       probabilistic facts / tuple-independent PDB  & y & n & n & n & y \\\hline
       annotated disjunctions / independent causation  & y & n & n & n & y \\\hline
    \end{tabular}
    \caption{Probabilistic logic programming languages and their features \todo{expand beyond our own languages / refine features further} \ak{adapting a comment from Luc in a different semantics discussion, the goal here is to show that we introduce the first PLP language that brings the range of modeling features / syntax elements together with the fully declarative semantics in a single language; then, the examples in sec 2 should illustrate how cool this is}}
    \label{tab:features}
\end{table}

\ak{I think these only motivate part of the whole, and should be adapted to illustrate the full power of the features in table \ref{tab:features} -- show what we can do and discuss that others can't rather than showing how others need to hack approximations }

Uncertainty is ubiquitous in real-world phenomena and successfully modelling such phenomena necessitates taking into account this uncertainty.
Expressing the knowledge we have about the world in a formal fashion constitutes a necessary first step in mechanizing the reasoning process for these probabilistic phenomena.
An emerging paradigm of doing so is probabilistic programming~\citep{meent2018introduction}.
Probabilistic programming languages (PPL) unify the expressive power of general purpose (Turing complete) programming languages and probabilistic modelling.

PPLs come in various different flavors, adhering for example to the declarative programming paradigm or the imperative programming paradigm.
All PPLs have in common that a user writes down a probabilistic program and the probabilistic inference engine, which is shipped along side the language, takes subsequently care of performing probabilistic inference.\footnote{One could argue that imperative probabilistic programming languages are to a degree declarative: the user only specifies/declares the probability distributions, and does not write any probabilistic inference algorithm.}

The particular PPL flavor that we investigate in this paper is probabilistic logic programming (PLP)~\citep{de2015probabilistic,riguzzi2018foundations}.
PLP can be seen as the continuation of a long line of research of unifying logic and probability theory~\citep{russell2015unifying}. With the introduction of {\em probabilistic Horn abduction} (PHA) by ~\citet{poole1993probabilistic}, a first important step had been made towards combining (non-probabilistic) logic programming and probability theory. PHA allowed~\citeauthor{poole1993probabilistic} to  construct the {\em Independent Choice Logic} (ICL) language~\citep{poole1997independent}, an expressive universal probabilistic logic programming language.
In a seminal paper, \citet{sato1995statistical} built on top of the ideas advanced by~\citet{poole1993probabilistic} and introduced the so-called {\em distribution semantics} (DS) for probabilistic logic programs.  
Since the inception of DS, a variety of  additional PLP languages have appeared, such as {\em PRISM}~\citep{sato1997prism}, {\em Logic Programs with Annotated Disjunctions}~\citep{vennekens2004logic}, and {\em \problogsty}~\citep{de2007problog,fierens2015inference}, to name a few.
\ak{careful with history claims (eg Dantsin)}

Probabilistic logic programming has first and foremost been concerned with random variables that have finite sample spaces -- mainly due to the fact that it is rather straightforward to turn deterministic logic atoms into probabilistic logic atoms.
As a consequence, PLP languages that allow for continuous random variables, such as \extendedprismsty~\citep{islam2012inference} and \dcsty\todo{cite on first mention}, have been proposed.
In the context of probabilistic logic programming such languages have been dubbed {\em hybrid}, as they allow for discrete and continuous random variables alike.
Unfortunately, the semantics that have been proposed so far for hybrid PLP languages have been imprecise, as pointed out by~\citet{azzolini2021semantics}, or exhibit some caveats. For example, the declarative semantics proposed in~\citep{islam2012inference} or~\citep{gutmann2011magic} do not allow for negation, and while the semantics proposed in~\citep{nitti2016probabilistic}  do allow negation, they are not declarative anymore but procedural. Our first contribution tackles these issues:
\begin{enumerate}
    \item We introduce the hybrid probabilistic logic programming language \dcproblogsty, for which we define {\bf declarative semantics} that cover negation and are based on \citeauthor{sato1995statistical}'s distribution semantics.
\end{enumerate}
The declarative semantics that we define for the hybrid PLP langugage \dcproblogsty extends the semantics given in~\citep{gutmann2011magic} such that negation is allowed. Furthermore, we simplify the definition of the semantics and eliminate some inconsistencies present in~\citep{gutmann2011magic}.

Eventually, a probabilistic programming language should be able to compute (conditional) probabilities of queries given by the user. This means that PLPs need to be equipped with inference mechanism. We show how inference for the hybrid PLP language \dcproblogsty can be mapped onto weighted model integration (WMI)~\citep{belle2015probabilistic}.

\begin{enumerate}
\setcounter{enumi}{1}
	\item We {\bf reduce inference} in \dcproblogsty to {\bf weighted model integration}, analogous to reducing inference in \problogsty to weighted model counting~\citep{chavira2008probabilistic} when dealing with discrete random variables only.
	\item We present {\em algebraic likelihood weighting}, an inference algorithm for hybrid probabilistic logic programs that combines {\bf knowledge compilation}~\citep{darwiche2002knowledge}, {\bf algebraic model counting}~\citep{kimmig2017algebraic}, and {\bf likelihood weighting}~\citep{fung1989weighing}.
\end{enumerate}
Until now only probabilistic programming languages targeting discrete random variables exclusively have leveraged knowledge compilation. This has allowed them to exploit structure present in the program when performing inference.
Algebraic likelihood weighting is the first inference algorithm in the hybrid domain exhibiting this feature and is consequently the first inference algorithm to exploit discrete structures in hybrid probabilistic programs.

Furthermore, we provide an implementation of \dcproblogsty that we dub \dcproblogsys and which is an extension of the ProbLog2 system~\citep{dries2015problog2}.
\begin{enumerate}
\setcounter{enumi}{3}
% 	\item We introduce a {\bf type system} for \dcproblogsty. The type system allows us to overload the existing \problogsty syntax and use it for hybrid probabilistic logic programming.
% 	While type systems are omnipresent in non-logic probabilistic programming languages, they have not received any attention in the PLP community.
    % To the best of our knowledge, we present the first type system for a probabilistic logic programming language.
	\item We present an {\bf implementation of \dcproblogsty} that uses algebraic likelihood weighting and reduces naturally  to \problogsty in the absence of continuous random variables.
% 	\item We give two implementations of AWL. An exact implementation that performs exact  probabilistic inference using PSI~\citep{gehr2016psi} in its back end and an approximate implementation using Pyro~\citep{bingham2019pyro}.
\end{enumerate}

% \begin{itemize}
%     \item fully unify discrete and continuous semantics
%     \item disentangle random variables from program
%     \item perform inference by mapping to WMI -> generalizes WMC
%     \item what is different to other approaches, non-logical? separation of combinatorial \#P-hardness and integration \#-hardness
% \end{itemize}

\subsection*{Paper Organization}
In Section~\ref{sec:motivation}, we motivate our work by giving introductory examples. We continue in Section~\ref{sec:semantics} by presenting \dcplpsty, a bare bones probabilistic logic programming language with minimal syntax.
\dcplpsty with its minimal syntax has the advantage of facilitating the definition of a declarative semantics. In Section~\ref{sec:dcproblog} we formally introduce \dcproblogsty.
The semantics of \dcproblogsty programs are obtained via deterministically transforming \dcproblogsty programs to \dcplpsty programs, for which we will have defined the semantics in Section~\ref{sec:semantics}.
In Section~\ref{sec:implementation} we describe \dcproblogsys, an implementation of \dcproblogsty built on top of the \problogsys system~\citep{fierens2015inference,dries2015problog2}.
Subsequently, we map inference in \dcproblogsty to weighted model integration ~\citep{belle2015probabilistic} and describe the implementation of the inference algorithm in \dcproblogsys. 
In Section~\ref{sec:related} we discuss \dcproblogsty in a broader probabilistic programming context and relate the language to other probabilistic programming languages such as BLOG~\citep{milch2005blog}.
We end with concluding remarks in Section~\ref{sec:conclusions}.

\subsection*{Prerequisites}

We expect the reader to be familiar with rudimentary language concepts and terms of (non-probabilistic) logic programming and more specifically of Prolog~\citep{lloyd2012foundations,flach1994simply}. We provide a brief summary of these concepts in~\ref{app:lp}. Furthermore, a working knowledge of probabilistic inference with discrete random variables using weighted model counting and integration would be beneficial. A refresher on probabilistic inference using WMC and WMI can be found in~\ref{app:amc}, ~\ref{app:compilation}, and \ref{app:wmi}.

\ak{decide on LP/Prolog citation(s), here vs start of next sec (and elsewhere?)}

\subsection{stuff from elsewhere}
\ak{these are bits and pieces that disturbed the flow of the semantics section but may still be relevant somewhere...}

\ak{the following message seems better placed in the intro/motivation: }
\citet{poole2010probabilistic} has characterised probabilistic programs as deterministic systems with  independent choices. \dcproblogsty relaxes \citeauthor{poole2010probabilistic}'s independent choices by allowing for connections between the independent choices by means of random variables appearing as parameters of distributions, thus introducing a kind of hierarchical system of independent choices, as illustrated in  Example~\ref{ex:random_choice_dependency}.
\dcproblogsty is the first probabilistic programming language to make the distinction between:
\begin{enumerate}
    \item structure expressed by the the deterministic system
    \item structure expressed by hierarchically parameterized distributions of random variables
\end{enumerate}
\ak{any better term to refer to the structure between RVs than "hierarchical"? }

\ak{We need to decide what to do with this; can we claim it's syntactic sugar and deal with it in that section? }
Note that while we implicitly assume   univariate (or {\em `atomic'}) random variables here, the principles do also apply to multivariate (or {\em `compound'}) random variables, such as multivariate Gaussians, provided that the usual laws of probability are respected and suitable syntax for such variables is introduced.

%Distributional facts are not limited to describe continuous random variables but can, for example, also model discrete probability distributions such as the Poisson distribution.
%\begin{example}\label{example:dcproblog_poisson_population}
%The following DC-ProbLog program models the size of a group of people as a Poisson distribution and
%defines predicates related to that size using the comparison predicates  \probloginline{>}\lstinline[columns=fixed]{/2} and \probloginline{=:=}\lstinline[{columns=fixed}]{/2}, where the latter is standard Prolog syntax for equality comparisons on numbers.
%\luc{=:= non-standard}\ak{it is standard Prolog}
%	\begin{problog*}{linenos}
%n_people ~ poisson(6).

%more_than_five:- n_people>5@.@
%exactly_five:- n_people=:=5@.@
%	\end{problog*}
%\end{example}

%We will provide further examples throughout the text, including showcase examples in Section~\ref{sec:showcase}. \ak{any good reason why we delay the showcases til then?}

% \section{Introduction}

% The mathematical discipline of logic constitutes the bedrock of a large body of work in artificial intelligence (AI) research, including sub-disciplines such as expert systems~\citep{jackson1998introduction}, inductive logic programming~\citep{muggleton1994inductive}, or more broadly knowledge representation and reasoning~\citep{levesque1986knowledge}. One of the main shortcomings of logic based AI systems is that they fail to capture the inherent uncertainty present in the world -- aleatoric and epistemic uncertainty alike~\citep{hullermeier2019aleatoric}. To express such uncertainties, AI researchers have set out to enhance pure logic with probabilities.

% As a matter of fact, the marriage of logic and probability is an enduring endeavor of AI research that predates AI itself~\citep{russell2015unifying}. Grandees of mathematics, philosophy, and economics such as Leibniz, Peirce, and Keynes, undertook early efforts to this end \citep{hailperin1984probability,howson2003probability}. In the field of AI, \citet{nilsson1986probabilistic} and~\citet{pearl1988probabilistic} presented seminal works investigating the combination of logic and probability, which heralded the era of {\em modern AI}: systems were now able to capture the inherent uncertainty present in the world as well as reason about it.

% \luc{Dantsin 89, PDBs, general PP ?}

% \citet{poole1993probabilistic} improved upon these ideas and introduced {\em probabilistic Horn abduction}, which allowed him to later on construct the {\em Independent Choice Logic} (ICL) language~\citep{poole1997independent}. ICL is an expressive universal probabilistic logic programming language (PLP) \citep{de2015probabilistic,riguzzi2018foundations} rooted in logic programming.
% Subsequently, \citet{sato1995statistical} built on top of the ideas advanced by~\citet{poole1993probabilistic} and introduced the so-called {\em distribution semantics} (DS) for probabilistic logic programs.  
% After the introduction of DS, a variety of  further PLP languages appeared, such as {\em PRISM}~\citep{sato1997prism}, {\em Logic Programs with Annotated Disjunctions}~\citep{vennekens2004logic}, and {\em \problogsty}~\citep{de2007problog,fierens2015inference}, to name a few.

% In this paper we focus on \problogsty, a probabilistic programming language inheriting its syntax from the (non-probabilistic) logic programming language \prologsty~\citep{sterling1994art}. \problogsty  allows one to express, in addition to logic rules and facts, so-called \textit{probabilistic facts}, which are logic facts labeled with the probability of them being satisfied.
% \begin{example}\label{example:problog_machine}
% 	We model two machines (Line \ref{program:problog_machines_dec} in the program below).
% 	We want to know the probability of the first machine working (Line \ref{program:problog_machines_query}), given that the second machine works (Line \ref{program:problog_machines_evidence}) and given a model that describes under which conditions the machines work (Lines \ref{program:problog_machines_work_cool} and \ref{program:problog_machines_work_temp}).
% 	Additionally the program models the outside temperature (Line \ref{program:problog_machines_temp})  and whether the cooling of each machine works (Lines \ref{program:problog_machines_cool1} and \ref{program:problog_machines_cool2}) as (Boolean) random variables (expressed as probabilistic facts).
% 	\begin{problog*}{linenos}
% machine(1). machine(2). @\label{program:problog_machines_dec}@

% 0.8::temperature(low). @\label{program:problog_machines_temp}@
% 0.99::cooling(1). @\label{program:problog_machines_cool1}@
% 0.95::cooling(2). @\label{program:problog_machines_cool2}@

% works(N):- machine(N), cooling(N). @\label{program:problog_machines_work_cool}@
% works(N):- machine(N), temperature(low). @\label{program:problog_machines_work_temp}@

% evidence(works(2)).  @\label{program:problog_machines_evidence}@
% query(works(1)). @\label{program:problog_machines_query}@
% 	\end{problog*}
% Running the program yields $P(\text{\probloginline{works(1)}}|\text{\probloginline{works(2)}})\approx 0.998$.
% \end{example}

% A major limitation of \problogsty becomes immediately apparent in Example~\ref{example:problog_machine}: we need to model the {\em temperature} as a discrete (Boolean) random variable rather than a continuous probability density.

% To repair this deficiency we present \dcproblogsty, a probabilistic logic programming language derived from \problogsty and Distributional Clauses (DC) -- two related but separate probabilistic logic programming languages.
% \dcproblogsty inherits its syntax and the concept of probabilistic facts from \problogsty. Moreover, \dcproblogsty adapts the idea of defining and querying continuous random variables using so-called {\em distributional facts}.
% Distributional facts allow a user to express random variables that are distributed according to a certain probability distribution specified by a random function symbol.

% In the following example, the distributional fact in Line~\ref{program:dcproblog_machines_work_temp} introduces the random variable \probloginline{temperature} and associates it with the distribution term \probloginline{normal(20,5)}. This is done using the \probloginline{~}\lstinline[columns=fixed]|/2| predicate in infix notation.
% In Line~\ref{program:dcproblog_machines_work_temp} the random variable is then queried using the comparison predicate \probloginline{<}\lstinline[columns=fixed]|/2|, again written in infix notation.

% \begin{example}\label{example:dcproblog_machine}
% 	We model the temperature as a continuous random variable distributed according to a normal distribution with mean $20$ and standard deviation $5$ (Line \ref{program:dcproblog_machines_temp}).
% 	\begin{problog*}{linenos}
% machine(1). machine(2).

% temperature ~ normal(20,5). @\label{program:dcproblog_machines_temp}@
% 0.99::cooling(1).
% 0.95::cooling(2).

% works(N):- machine(N), cooling(N).
% works(N):- machine(N), temperature<25.0. @\label{program:dcproblog_machines_work_temp}@

% evidence(works(2)).  @\label{program:dcproblog_machines_evidence}@
% query(works(1)).
% 	\end{problog*}
% \end{example}
% \luc{query(works(1)|works(2))}

% Distributional facts are not limited to describe continuous random variables but can, for example, also model discrete probability distributions, of which an example would be the Poisson distribution.
% \begin{example}\label{example:dcproblog_poisson_population}
% 	In this example we show how to model the size of a group of people as a Poisson distribution and answer queries about the size of the group using the comparison predicates  \probloginline{>}\lstinline[columns=fixed]{/2} and \probloginline{=:=}\lstinline[{columns=fixed}]{/2}.\luc{=:= non-standard}\ak{it is standard Prolog}
% 	\begin{problog*}{linenos}
% n_people ~ poisson(6).

% more_than_five:- n_people>5@.@
% exactly_five:- n_people=:=5@.@

% query(more_than_five).
% query(exactly_five).
% 	\end{problog*}

% \end{example}

% \citet{poole2010probabilistic} has characterized probabilistic programming as {\em independent choices plus deterministic systems}.
% In the case of probabilistic logic programming with discrete random variables the deterministic system is given by a logic program.
% \dcproblogsty builds on this idea but relaxes the independence assumption by explicitly distinguishing two types of dependencies 1) those expressed by the logic program (the deterministic system) and 2) those expressed by using random variables as parameters of distributions for other random variables (the independent random choices \ak{sounds like  independent random choices are a form of dependencies, which they aren't; in general, this is a line of thought that we repeat in too many places -- we have to pick one and explain it well there}). This explicit distinction is not made by any other probabilistic (logic) programming language including discrete and continuous random variables. Such a programming language is also called {\em hybrid}. \luc{what about BLOG?}

% When writing (hybrid) probabilistic programs, users have a range of options, from encoding entire probability distributions defined by the probabilistic program into a single choice random variable (no deterministic system), all the way to 
% a set of individual independent choice random variables with minimal use of random variables as parameters tied together by a deterministic system.
% As already discussed by~\citet{poole2010probabilistic}, not all options work equally well with all inference algorithms. \dcproblogsty provides these options within the same semantic framework, leaving it to system developers to focus the syntax towards a subset of options and/or to exploit the different options when integrating different inference approaches into their system.

% Developing the \dcproblogsty language, leads to the following specific contributions:
% \todo{update needed}

% \begin{enumerate}
% 	\item We introduce a {\bf type system} for \dcproblogsty, which allows us to introduce a {\bf neat and clean syntax} to extend \problogsty with function symbols for continuous random variables.
% 	\item We introduce a purely {\bf declarative semantics for \dcproblogsty} based on \citeauthor{sato1995statistical}'s distribution semantics that allows us to disentangle independent choices and the discrete system underlying a probabilistic program.
% 	\item We {\bf reduce inference} in \dcproblogsty to {\bf weighted model integration} in the algebraic model counting setting~\citep{kimmig2011algebraic,kimmig2017algebraic}, analogous to reducing inference in \problogsty to weighted model counting~\citep{chavira2008probabilistic}.
% 	\item We present an {\bf implementation of \dcproblogsty}, which reduces naturally to \problogsty in the absence of continuous random variables.
% \end{enumerate}
% \dcproblogsty is heavily influenced by two predecessor languages: 1) \problogsty and 2) Distributional Clauses~\citep{gutmann2011magic,nitti2016probabilistic}. The latter is a probabilistic logic programming languages that allows for continuous random variables. However, its semantics differ from the semantics of \problogsty such that it does not reduce to \problogsty in case of the absence of continuous random variables.

% \subsection*{Prerequisites}

% To fully appreciate the paper, we expect the reader to be familiar with rudimentary language concepts and terms of (non-probabilistic) logic programming and more specifically of Prolog~\citep{lloyd2012foundations,flach1994simply}. We provide a brief summary of these concepts in~\ref{app:lp}.

% Furthermore, a working knowledge of probabilistic inference with discrete random variables using weighted model counting would be beneficial (WMC)~\citep{chavira2008probabilistic,darwiche2009modeling}, as we present an inference algorithm for \dcproblogsty that extends WMC. A refresher on probabilistic inference in the discrete domain can be found in~\ref{app:amc} and ~\ref{app:compilation}.

% \subsection*{Paper Organization}
% \todo{}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/preliminaries.tex
aij dcproblog paper

\section{Preliminaries}

We expect the reader to be familiar with basic language concepts and terms of Prolog \todo{cite (Lloyd, 1989; Flach, 1994)}; a quick summary can be found in~\ref{app:lp}. \ak{this sentence and appendix copied from plp concepts paper; should go earlier given we already have examples there?}

\subsection{Probabilistic Logic Programming}
\ak{how much do we need here, given that we later introduce ProbLog constructs through their mappings to the new language? how much detail on the distribution semantics is needed for completeness? (the semantics sec is heavily linked to the Sato paper, but we can't repeat all of that)}

\subsection{Weighted Model Integration}

\todo{introduce all the wmc, amc, wmi gadgets}

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/syntax_types.tex
aij dcproblog paper

\section{\dcproblogsty Syntax and Type System} \label{sec:syntax}
\ak{I propose to move the type system and multiple dispatch stuff to a separate implementation section as much as possible;\ie, from here onwards: 1) introduce DC-ProbLog programs as normal logic program + probabilistic facts + distributional facts + comparison predicates instead of wrapped query terms, 2) show that these map to DC-PLP, 3) add further syntactic sugar (is/2, distributional clauses, anything else? are annotated disjunctions part of DC-ProbLog?) and show how that maps back onto the basic version, 4) talk about multiple dispatch and all that }

\todo{this is just moved from earlier sec 3, needs linking to DC-PLP and merging with semantics in next section; the goal is to introduce DC-ProbLog as a higher level modeling language that makes concrete choices for the reserved functors in DC and introduces nicer syntax / syntactic sugar, but for its semantics maps back to (an instance of) DC-PLP}

\subsection{Type System}
Before we formally specify the syntax of \dcproblogsty, we specify a hierarchical type system. To this end we will modify the type system given in~\citep{sterling1994art}, one of the main references on \prologsty. The base type in the type system of~\citep{sterling1994art} is the \type{Term} type, from which other types are subtyped. In other words: everything is a \type{Term}.
\begin{enumerate}
	\item[]   \type{Term}
	\begin{itemize}		
		\item \type{Atomic}
		\begin{itemize}
			\item \type{Atom}
			\item \type{Number}
		\end{itemize}
		\item \type{Compound}
	\end{itemize}
\end{enumerate}

{\em Logical variables} (technically referred to as {\em meta-logical variables}~\citep[Chapter 10]{sterling1994art}) are omitted from the type system hierarchy, as they can ground to any type, depending on the context. In Figure~\ref{fig:type_system_prolog} we give a diagrammatic representation of the \prologsty type system.
Note that \problogsty uses a different nomenclature: atom terms in \problogsty correspond to the union of atom terms and compound terms in the \prologsty nomenclature of~\citep{sterling1994art}. In a first step we are going to change the \prologsty type system to match the nomenclature of \problogsty.
\begin{enumerate}
	\item[]   \type{Term}
	\begin{itemize}		
		\item \type{Number}
	\end{itemize}
	\begin{itemize}
		\item \type{Atom}
	\end{itemize}
\end{enumerate}

In Table~\ref{table:comp_type_system} we give a side by side comparison of the two type systems and in Figure~\ref{fig:type_system_prolog} and~\ref{fig:type_system_problog} we give diagrammatic representations of both type systems.

\begin{table}[h]
	\caption{Comparison of the \prologsty type system and the \problogsty type system.}
	\label{table:comp_type_system}
	\begin{center}
		\begin{tabular}{c||c|c|c}
			example \type{Term} & \multicolumn{2}{c|}{\prologsty} & \problogsty \\ 
			\hline
			\probloginline{3} & \type{Number} & \multirow{2}{*}{\type{Atomic}}  &  \type{Number}  \\
			\cline{1-1} \cline{4-4}
			\probloginline{foo} & \type{Atom} &   & \multirow{2}{*}{\type{Atom}}  \\
			\cline{1-3}
			\probloginline{bar(foo)} &  \multicolumn{2}{c|}{\type{Compound}}  &   \\
			%	\probloginline{foo} & \begin{tabular}{@{}c|c@{}}
			%		\type{Atom} &\multirow{0}{c}{\phantom{\type{Atomic}}} 
			%	\end{tabular}  &  \type{Atom} \\
		\end{tabular}
	\end{center}
\end{table}

\begin{figure}[h!]
	\centering
	\begin{minipage}{.45\textwidth}
		\centering
		\includegraphics[width=0.85\linewidth]{../figures/type_hierarchy_prolog.pdf}
	\end{minipage}%
	\hfill
	\begin{minipage}{.45\textwidth}
		\centering
		\includegraphics[width=0.85\linewidth]{../figures/type_hierarchy_problog.pdf}
	\end{minipage}
	\par
	\medskip
	\noindent
	\begin{minipage}[t]{.45\textwidth}
		\captionof{figure}[Diagrammatic representation of the hierarchical \prologsty type system.]{Diagrammatic representation of the hierarchical \prologsty type system~\citep{sterling1994art}. The \type{Number} type is further subtyped into \type{Integer} and \type{Float}.}
		\label{fig:type_system_prolog}
	\end{minipage}%
	\hfill
	\begin{minipage}[t]{.45\textwidth}
		\captionof{figure}[Diagrammatic representation of the hierarchical \problogsty type system.]{Diagrammatic representation of the hierarchical \problogsty type system. The \type{Number} type is further subtyped into \type{Integer} and \type{Float}.}
		\label{fig:type_system_problog}
	\end{minipage}
\end{figure}

Analogously to the just introduced type system of \problogsty, the base type for \dcproblogsty is again the \type{Term} type. At this point we modify the \problogsty type system by subtyping terms with a \type{Distrbution} type and a \type{Symbol} type. 
Dividing \dcproblogsty terms into \type{Distribution} terms and \type{Symbol} terms reflects the characterization of~\citet{poole2010probabilistic} of probabilistic programs as being constituted of independent choices (\type{Distribution}) and a deterministic system (\type{Symbol}).

\begin{definition}\label{def:type_system}
	The \dcproblogsty type system is a hierarchical type system with each type being a disjoint union of its subtypes.
	\begin{enumerate}
		\item[] \type{Term} 
		\begin{enumerate}
			\item[1.] \type{Distribution}
			\begin{itemize}
				\item \type{DiscreteD}
				\item \type{ContinuousD}
				\item \type{MixtureD}
			\end{itemize}
			\item[2.] \type{Symbol}
			\begin{itemize}		
				\item \type{Arithmetic}
				\begin{itemize}
					\item  \type{RandomVariable}
					%			\begin{itemize}
					%				\item \type{DiscreteRV}
					%				\item \type{ContinuousRV}
					%				\item \type{MixtureRV}
					%			\end{itemize}
					\item \type{Number}
				\end{itemize}
				\item \type{Atom}
			\end{itemize}
		\end{enumerate}
	\end{enumerate}
\end{definition}

A diagrammatic overview of the \dcproblogsty type system is given in Figure~\ref{fig:type_system}.

\begin{figure}[t]
	\begin{center}
		\includegraphics[width=\linewidth]{../figures/type_hierarchy.pdf}
	\end{center}
	\caption[Diagrammatic representation of the hierarchical \dcproblogsty type system.]{Diagrammatic representation of the hierarchical \dcproblogsty type system. Compared to Definition~\ref{def:type_system},  the \type{Number} type is further subtyped into \type{Integer} and \type{Float} types and the \type{RandomVariable} type is subtyped into \type{DiscreteRV}, \type{ContinuousRV}, and \type{MixtureRV}. These subtypes of \type{RandomVariable} mimic the subtypes of the \type{Distribution} type.}
	\label{fig:type_system}
\end{figure}

%A type is a disjoint union of its subtytpes, for example:
%$$\type{Term}=\type{Distribution} \cup \type{Symbol}$$

\subsubsection{Distribution Terms}
As already seen in Examples~\ref{example:dcproblog_machine} and~\ref{example:dcproblog_poisson_population}, \type{Distribution} terms determine the probability distribution according to which a random variable is distributed.

\begin{definition}[\type{Distribution}]
	Terms of type \type{Distribution} are constructed by using a reserved random function symbol and applying it to a (possible empty) tuple of input arguments, which maps the input tuple to a distribution term.
	Distribution terms are only allowed to be invoked as a second argument to the reserved distribution predicate \probloginline{is}\lstinline[columns=fixed]|/2|.
	This also means that random function symbols are only used there as well. \dcproblogsty subtypes the \type{Distribution} type into the \type{DiscreteD}, \type{ContinuousD}, and \type{MixtureD} types.
\end{definition}

\probloginline{poisson(6)} and \probloginline{normal(20,5)} are for instance of types \type{DiscreteD} and  \type{ContinuousD}, respectively.
Note that distributions of type \type{ContinuousD} denote {\em absolutely continuous} probability distributions\footnote{\dcproblogsty does not support {\em singular continuous} probability distributions such as the Cantor distribution.}. \ak{would be nice to have a hint on why this restriction is there} We give an example of a distribution of type \type{MixtureD} in Example~\ref{ex:indian_gpa_function}.
%\begin{example} \probloginline{mixed} is a random variables that is a mixture of a continuous part (the truncated normal between $0$ and $4$ with mean $3$ and standard deviation $1$) and a discrete part. 
%\begin{problog}
%mixted~Mix((0.998,TruncNormal(3,1,0,4)),(0.001,4),(0.001,4))
%\end{problog}
%\end{example}

%For and example, see Example~\ref{example:dcproblog_poisson_population}: we have the random function symbol \probloginline{poisson}\lstinline[columns=fixed]{/1}, apply it to the input tuple \probloginline{(6)}, and obtain a term of type \type{Distribution}.

\subsubsection{Symbol Terms}
\type{Symbol} terms are the building blocks of the discrete structure that makes up a logic program and as such determine also the deterministic system of a probabilistic logic program. Beyond the reserved syntax, which is used to define the distribution terms, the user is free to use any identifier (following \prologsty conventions) to declare \type{Symbol} terms -- this includes standard \prologsty terms such as lists.
%The subtyping of the \type{Symbol} type (shown in Figure~\ref{fig:type_system}) follows the convention given in~\citep[Chapter 9]{sterling1994art}.

\begin{definition}[\type{Arithmetic}]
	Terms of type \type{Symbol} for which arithmetic operations are defined are of type \type{Arithmetic}. \type{Arithmetic} terms can be used as second argument to the builtin \probloginline{is}\lstinline[columns=fixed]|/2| functor. \type{Arithmetic} terms are subtyped with \type{RandomVariable} terms and \type{Number} terms.
%	\ak{Def 6.3 Arithmetic terms: if (as stated in def 6.1) each type is the disjoint union of its subtypes, that means that "arithmetic" is made up exclusively of numbers and RVs, \ie, terms like 3-2 or min(x,5) for a RV x are not allowed, whereas the first two sentences of def 6.3 suggests they might be allowed (only having numbers and RVs directly on the rhs of is/2 seems quite boring...). Which is it?
%		Related to this, defs 6.8 and 6.10 talk about RVs and Numbers, why not simply "arithmetic"?}
\end{definition}
We talk in more detail about \type{Arithmetic} terms and their evaluation in Subsection~\ref{sec:arithmetic}.

\begin{definition}[\type{RandomVariable}] \label{def:random_variable}
	\type{RandomVariable} terms are named by the programmer using distributional facts with the binary infix predicate \probloginline{is}\lstinline[columns=fixed]|/2| and inherit their subtype from the distribution term. All random variables present in a \dcproblogsty program are named. \type{RandomVariable} terms are a subtype of the \type{Arithmetic} type. \ak{Def 6.4: why do you restrict random variables to be arithmetic? What about color(Ball)~uniform([red,green,blue])? Of course, such finite discrete domains could be encoded using integers and then translated back in the LP, so theoretically speaking it's probably not a restriction to only have numbers, but from a modeling perspective, it is surprising to me.}
\end{definition}

Examples of terms of type \type{RandomVariable} would be \probloginline{temperature} and \probloginline{n_people} in Examples~\ref{example:dcproblog_machine} and~\ref{example:dcproblog_poisson_population}, where we declared the random variables through so-called {\em distributional facts}: 
\begin{problog}
temperature ~ normal(20,5).
n_people ~ poisson(6).
\end{problog}
As mentioned in Defintion~\ref{def:random_variable}, \type{RandomVariable} terms inherit their type from the associated \type{Distribution} term, \eg a \type{ContinuousD} term induces a random variable term of type \type{ContinuousRV}.

Note that not all probabilistic programming languages enforce the naming of random variables. In the functional probabilistic programming language \anglicansty~\citep{wood2014approach}, for instance, the user can declare anonymous random variables, much like anonymous functions, \ie \textlambda-functions.
\ak{Discussion of named/unnamed between Defs 6.4 and 6.5: would be nice to have a brief hint on why you made that choice}

\begin{definition}[\type{Number}] Terms of type \type{Number} are used to express numerical data types such as integers and floating point numbers.	
\end{definition}

\ak{terminology mix-up; predicates never form terms}
\begin{definition}[\type{Atom}]
	An atom is of the form \probloginline{p(t@$_1$@,@\dots @, t@$_n$@ )} where \probloginline{p} is a predicate of arity $n$ and the \probloginline{t@$_i$@} are \type{Symbol} terms~\citep{fierens2015inference}. Predicates used to construct \type{Atom} terms must not clash with the builtin predicates and reserved random function symbols used to declare \type{Distribution} terms. Terms of type \type{Atoms} constitute the building blocks of the discrete structure of a probabilistic logic program.
\end{definition}

Examples of atoms terms would be \probloginline{foo} or \probloginline{bar(foo,42)}. The latter is constructed using the predicate \lstinline{bar/2} and the tuple of \type{Symbol} terms \probloginline{(foo,42)}. \probloginline{foo} is of type \type{Atom} and \probloginline{42} is a \type{Number} term.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Syntax}
%Based on the type system, we now define the syntactic components of DC-ProbLog and its syntax.
\begin{definition}[Deterministic Rules and Facts]
	A (normal) logic program is a set of rules. A rule is a universally quantified expression of the form
	\begin{problog}
h:- b@$_1$@, @\dots@, b@$_n$@.
	\end{problog}
	where \probloginline{h} is an \type{Atom} term and $\{ \text{\probloginline{b@$_1$@}}, \text{\dots},\text{\probloginline{b@$_n$@}} \}$ is a set of $n$ literals.
	Literals are either \type{Atom} terms, or their negation.
	The \type{Atom} term \probloginline{h} is called the head of the rule and \probloginline{b@$_1$@, @\dots@, b@$_n$@} the body. The latter represents the
	conjunction $\text{\probloginline{b@$_1$@}} \land \cdots \land  \text{\probloginline{b@$_n$@}}$. A fact is a rule that has \probloginline{true} as its body and is simply written as
	\begin{problog}
h.
	\end{problog}
\end{definition}

\begin{definition}[Probabilistic Facts]
	Probabilistic facts are of the form \probloginline{p::fact}. \probloginline{p} is a term of type \type{Number} or \type{RandomVariable}, with $0{\leq} \text{\probloginline{p}} {\leq} 1$  and \probloginline{fact} is a ground atom. 
	%\begin{definition} A {\bf probabilistic fact} is a ground definite fact labeled with a probability.
\end{definition}

\begin{definition}[Distributional Facts]
	Distributional facts introduce random variable terms and associate them with a distribution using the reserved binary predicate \probloginline{~}\lstinline[columns=fixed]|/2| in infix notation. A distributional fact is a ground fact of the form \probloginline{rand ~ distribution}, where \probloginline{distribution} is a well-typed distribution term with output type \type{DistributionX} and \probloginline{rand} is a ground term whose functor is not reserved. The distributional fact introduces \probloginline{rand} as a random variable of type \type{RandomVariableX}, \ie the type of the random variable term mimics the type of the distribution term.
\end{definition}
Distributional facts can be thought of as the link between the independent choices (in form of distributions) and the discrete structure (in form of rules and facts).

\begin{definition}[Comparison predicates] \label{def:comparison}
	\dcproblogsty (similar to \prologsty and \problogsty) uses a fixed set of builtin binary predicates whose truth values depend on the \emph{evaluation} of its arguments. For terms that can be evaluated (\type{RandomVariable} terms and \type{Number} terms) we use the set
	$\bowtie=\{ \text{\probloginline{=:=}}, \text{\probloginline{=\=}}, \text{\probloginline{<}}, \text{\probloginline{>}},  \text{\probloginline{=<}}, \text{\probloginline{>=}} \}$
	of binary predicates written in infix notation.
	Each of these has the usual \prologsty arithmetic interpretation of evaluating each side and comparing the results.
	Just as for \prologsty arithmetic, truth values of atoms using these predicates are determined through an evaluation process that is external to the (probabilistic) logic program. Such atoms can be used in clause bodies, but never in clause heads (or as facts).
\end{definition}

%We will further discuss arithmetic evaluations in Subsection~\ref{sec:arithmetic}.

\begin{example} Comparison predicates allow for the comparison of random variables to numbers. 
	\begin{problog*}{linenos}
x ~ normal(0,3).
q:- x>0.
query(q).
	\end{problog*}
\end{example}

%\begin{definition}\label{def:distributional_fact}
%	A {\bf distributional fact} is a ground definite fact with the reserved binary predicate \probloginline{is}\lstinline[columns=fixed]|/2| used in infix notation.
%\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiple Dispatch} \label{ref:multiple_dispatch}
The attentive reader might have noticed that in Definition~\ref{def:comparison} we used the same comparison predicates, for \type{RandomVariable} terms as well as \type{Number} terms. In \dcproblogsty we resolve this clash through {\em multiple dispatching}\citep{castagna1995calculus}.
{ Multiple dispatching} is the process of dynamically selecting at run time\footnote{In the context of logic programming we understand as run time the time it takes to ground the logic program part of a \dcproblogsty program.} a specific implementation of a function based on the types of the arguments that are fed as input to the corresponding function symbol.
Such functions are referred to as {\em multimethods}. 
An early example of multiple dispatch is the {\em Common Lisp Object System}~\citep{keene1989object}. Multiple dispatch is also the programming paradigm on which the \juliasty programming language is built~\citep{bezanson2017julia}. We reuse this idea in the context of probabilistic logic programming with random function symbols to retain a simple and neat syntax.
Consider, for instance, the declaration of the normal distribution in Example~\ref{example:dcproblog_machine}.
The definition of the function is the following:
$$ \text{\probloginline{normal}}: \type{Number} \times \type{Number} \rightarrow \type{ContinuousD}$$
However, the arguments of a normal distribution (its mean and standard deviation) might as well be random variables themselves.
Multiple dispatching then allows us to {\em overload} the \probloginline{normal} function symbol by using arguments of a different type\footnote{For simplicity we omit that the standard deviation has to be a strictly positive (real) number, which can be enforced by extending the type system.}:
\begin{align*}
	\text{\probloginline{normal}}&: \type{RandomVariable} \times \type{Number} \rightarrow \type{ContinuousD} \\
	\text{\probloginline{normal}}&: \type{Number} \times \type{RandomVariable} \rightarrow \type{ContinuousD} \\
	\text{\probloginline{normal}}&: \type{RandomVariable} \times \type{RandomVariable} \rightarrow \type{ContinuousD}
\end{align*}
The exact implementation of these random function symbols is left to the designer of the implementation of the language.
%In the next Subsection~\ref{sec:arithmetic} we will encounter further usages of multiple dispatching in \dcproblogsty.

\subsection{Arithmetic Evaluation}\label{sec:arithmetic}

\ak{Sec 6.1.4: you say the correct functor is determined during grounding, but doesn't grounding (at least to some extend) depend on the type of inference? \eg, if we ground during sampling, I'd expect the second line of ex 6.4 to ground the X to different numbers in different samples / runs, whereas exact inference techniques (including techniques to find the relevant ground program of a query) would probably ground to something symbolic? Theoretically speaking, grounding usually refers to grounding wrt the Herbrand universe, \ie, all the symbols in the program (plus in the case of arithmetic the countably many outcomes of arithmetic evaluations), but continuous distributions are extending that potential universe into the uncountable, so we need to be careful (and I need to think about this again for the journal version...)}

\prologsty implementations reserve a set of predicate names for system-related procedures. These system predicates are usually deployed when the goal is to perform efficient arithmetic operations on numbers, either by using specialized algortihms or specialized hardware. For example, when a user wants to add up the integers \probloginline{3} and \probloginline{5} they simply write \probloginline{Sum is 3+5}. The predicate \probloginline{is}\lstinline[columns=fixed]|/2| takes the second argument, passes it to an external evaluation engine, which performs integer addition, and unifies the result with the logical variable \probloginline{Sum}. It is rather obvious that the concept of multiple dispatch comes in handy here: the same functors can be used to perform the same operations on terms of different types. The correct functor is only dertermined during the grounding of the logic program, \ie during runtime.

The key difference between the external arithmetic engine of \prologsty (or \problogsty) and the external arithmetic engine of \dcproblogsty is that the external engine of the latter is capable of performing arithmetic operations (perform evaluations) with random variables. Combining this innovation with multiple dispatch allows us to retain the same syntax (same functor) to perform arithmetic operations on \type{RandomVariable} terms and/or \type{Number} terms.

\begin{example} A \dcproblogsty code snippet showing how random variable terms can be used in an arithmetic expression. 
	\begin{problog*}{linenos}
x ~ normal(0,3).
q:- X is x+3, X>3.
query(q).
	\end{problog*}
\end{example}

Multiple dispatch allows \dcproblogsty to call specific implementations of the comparison predicates in the external arithmetic engine.
In other words, multiple dispatch allows for using the same predicate for semantically different terms.
Comparing two terms of type \type{Number}, for example, materializes as a deterministic fact, whereas a comparison between terms of types \type{Number} and \type{RandomVariable} materializes as a probabilistic query. We will see how this extremely useful mechanism works in the next Section~\ref{sec:semantics}, where we also introduce the semantics of \dcproblogsty.

\subsection{Relationship of Multiple Dispatch to Typing in \prologsty}

Traditionally \prologsty is {\em weakly typed}. This means that only at run time (grounding time) it is checked whether a relation with certain arguments is present in the logic database, which is a set of ground relations. If the relation is not present, \prologsty implementations do usually not throw an error but the proof of the goal that is currently proven fails. For builtin predicates such as \probloginline{is}\lstinline[columns=fixed]|/2| or any comparison predicates, the run time type system is more strict and the \prologsty implementation might actually throw an error. For instance, the presence of the relation \probloginline{three=:=3} in a proof would throw an error.
Such typed arithmetic relations can be thought of as {\em multirelations}, which have a different interpretations based on their type at run time. In other words, they are the logic programming equivalent to multimethods: their interpretation depends on their predicate and the types of their arguments.

In \dcproblogsty we extend this idea of multirelations using the framework of multiple dispatch. The major difference being that the typed relations in \dcproblogsty are not only used to select the right computation in the external arithmetic engine (\eg floating point addition vs. integer addition) but the multirelations in \dcproblogsty can also affect the structure of the probabilistic program. We will encounter examples thereof in Section~\ref{sec:dcproblog}, where we rewrite a \dcproblogsty program as a DC-PLP program and where we have rewrite rules that do not only depend on the predicate of a relation but also on the types of the relation's arguments.

Note that in \dcproblogsty not only the builtin comparison predicates and the \probloginline{is}\lstinline[columns=fixed]|/2| have a typed signature but also the random function symbols, \eg \probloginline{normal/2}. We can say that interpreted predicated and functors, that is predicates and functors that have a system internal meaning, possess a type signature. This results in a type dependent calculus.

\subsubsection*{Strongly Typed \prologsty}

While \prologsty is traditionally not strongly/statically typed, there have been efforts to extend \prologsty with such a type system. The most interesting effort is probably the one presented in~\citet{Schrijvers2008TowardsTP}, where the authors develop the idea of partially typing a \prologsty program, \ie typing is optional and untyped code is interpreted instead of compiled.  Our efforts to utilize multiple dispatch within a (probabilistic) logic programming language can be understood as an effort orthogonal to the efforts towards statically typed \prologsty. Introducing predicates with strongly typed signatures can also be regarded as adding interpreted predicates to the internal \prologsty engine.

### 2024-09-10 202409_aij_dc_problog_aij_second/files_old/types.tex
aij dcproblog paper

--- 2024-09-10 202409_aij_dc_problog_aij_second/files_old/syntax_types.tex
@@
-\section{\dcproblogsty Syntax and Type System} \label{sec:syntax}
-\ak{I propose to move the type system and multiple dispatch stuff to a separate implementation section as much as possible;\ie, from here onwards: 1) introduce DC-ProbLog programs as normal logic program + probabilistic facts + distributional facts + comparison predicates instead of wrapped query terms, 2) show that these map to DC-PLP, 3) add further syntactic sugar (is/2, distributional clauses, anything else?
-are annotated disjunctions part of DC-ProbLog?) and show how that maps back onto the basic version, 4) talk about multiple dispatch and all that }
+\section{System Predicates and Arithmetic Evaluations}
 
-\todo{this is just moved from earlier sec 3, needs linking to DC-PLP and merging with semantics in next section; the goal is to introduce DC-ProbLog as a higher level modeling language that makes concrete choices for the reserved functors in DC and introduces nicer syntax / syntactic sugar, but for its semantics maps back to (an instance of) DC-PLP}
+Every Prolog implementation reserves some predicate names for
+system-related procedures. Queries to these predicates, called system
+predicates, are handled by special code in the implementation in contrast
+to calls to predicates denned by pure Prolog programs.
 
-\subsection{Type System}
-Before we formally specify the syntax of \dcproblogsty, we specify a hierarchical type system.
-To this end we will modify the type system given in~\citep{sterling1994art}, one of the main references on \prologsty.
-The base type in the type system of~\citep{sterling1994art} is the \type{Term} type, from which other types are subtyped.
-In other words: everything is a \type{Term}.
+builtin predicates
+
+The role of the system predicates for arithmetic introduced in Prolog is
+to provide an interface to the underlying arithmetic capabilities of the
+computer in a straightforward way. The price paid for this efficiency is
+that some of the machine-oriented arithmetic operations are not as gen-
+general as their logical counterparts. The interface provided is an arithmetic
+evaluator, which uses the underlying arithmetic facilities of the com-
+computer.
+
+\subsubsection{Arithmetic Evaluation}
+\todo{write a paragraph introducing the use of is/2 and define its semantics}
+
+\ak{for the cases where the lhs of is/2 ends up in a comparison atom (such as Example \ref{ex:is}), it should be enough to replace is/2 by unification and thereby delay evaluation until the comparison is externally evaluated, but if we also have "regular" uses of is/2, where the (deterministic?) result is \eg returned as an answer, then things get a bit more complex; as I don't know what uses are intended (and valid), this is for someone else to draft...
+(or it might be that this is really an engine rather than a language thing, in which case it should only be discussed later with the general multiple dispatch stuff)}
+
+\ak{this sec is for Pedro to fix / integrate -- I think a lot of the Prolog type system discussion is not needed for the purpose of this paper; note that the old tex file is still there if needed again}
+
+\dcproblogsty's type system refines the one of DC-PLP to instantiate a conrete version of the abstract language.  Specifically, we have the following refinements:
 \begin{enumerate}
-	\item[]   \type{Term}
-	\begin{itemize}		
-		\item \type{Atomic}
-		\begin{itemize}
-			\item \type{Atom}
-			\item \type{Number}
-		\end{itemize}
-		\item \type{Compound}
-	\end{itemize}
+    \item distribution functors are further partitioned into \emph{discrete distribution functors}. \emph{continuous distribution functors} and \emph{mixture distribution functors}
+    \item arithmetic functors are further partitioned into \emph{operation functors} and \emph{number functors}, with the latter in turn partitioned into \emph{integers} and \emph{floats} \todo{do we have / want a complete list of operations?
+should we at least give some examples?}
+    \item we use Boolean query functors 	$\bowtie=\{ \text{\probloginline{=:=}}, \text{\probloginline{=\=}}, \text{\probloginline{<}}, \text{\probloginline{>}},  \text{\probloginline{=<}}, \text{\probloginline{>=}} \}$
+	in infix notation (and lift them to predicate level as discussed before)
 \end{enumerate}
 
-{\em Logical variables} (technically referred to as {\em meta-logical variables}~\citep[Chapter 10]{sterling1994art}) are omitted from the type system hierarchy, as they can ground to any type, depending on the context.
-In Figure~\ref{fig:type_system_prolog} we give a diagrammatic representation of the \prologsty type system.
-Note that \problogsty uses a different nomenclature: atom terms in \problogsty correspond to the union of atom terms and compound terms in the \prologsty nomenclature of~\citep{sterling1994art}.
-In a first step we are going to change the \prologsty type system to match the nomenclature of \problogsty.
-\begin{enumerate}
-	\item[]   \type{Term}
-	\begin{itemize}		
-		\item \type{Number}
-	\end{itemize}
-	\begin{itemize}
-		\item \type{Atom}
-	\end{itemize}
-\end{enumerate}
-
-In Table~\ref{table:comp_type_system} we give a side by side comparison of the two type systems and in Figure~\ref{fig:type_system_prolog} and~\ref{fig:type_system_problog} we give diagrammatic representations of both type systems.
-
-\begin{table}[h]
-	\caption{Comparison of the \prologsty type system and the \problogsty type system.}
-	\label{table:comp_type_system}
-	\begin{center}
-		\begin{tabular}{c||c|c|c}
-			example \type{Term} & \multicolumn{2}{c|}{\prologsty} & \problogsty \\ 
-			\hline
-			\probloginline{3} & \type{Number} & \multirow{2}{*}{\type{Atomic}}  &  \type{Number}  \\
-			\cline{1-1} \cline{4-4}
-			\probloginline{foo} & \type{Atom} &   & \multirow{2}{*}{\type{Atom}}  \\
-			\cline{1-3}
-			\probloginline{bar(foo)} &  \multicolumn{2}{c|}{\type{Compound}}  &   \\
-			%	\probloginline{foo} & \begin{tabular}{@{}c|c@{}}
-			%		\type{Atom} &\multirow{0}{c}{\phantom{\type{Atomic}}} 
-			%	\end{tabular}  &  \type{Atom} \\
-		\end{tabular}
-	\end{center}
-\end{table}
-
-\begin{figure}[h!]
-	\centering
-	\begin{minipage}{.45\textwidth}
-		\centering
-		\includegraphics[width=0.85\linewidth]{../figures/type_hierarchy_prolog.pdf}
-	\end{minipage}%
-	\hfill
-	\begin{minipage}{.45\textwidth}
-		\centering
-		\includegraphics[width=0.85\linewidth]{../figures/type_hierarchy_problog.pdf}
-	\end{minipage}
-	\par
-	\medskip
-	\noindent
-	\begin{minipage}[t]{.45\textwidth}
-		\captionof{figure}[Diagrammatic representation of the hierarchical \prologsty type system.]{Diagrammatic representation of the hierarchical \prologsty type system~\citep{sterling1994art}.
-The \type{Number} type is further subtyped into \type{Integer} and \type{Float}.}
-		\label{fig:type_system_prolog}
-	\end{minipage}%
-	\hfill
-	\begin{minipage}[t]{.45\textwidth}
-		\captionof{figure}[Diagrammatic representation of the hierarchical \problogsty type system.]{Diagrammatic representation of the hierarchical \problogsty type system.
-The \type{Number} type is further subtyped into \type{Integer} and \type{Float}.}
-		\label{fig:type_system_problog}
-	\end{minipage}
-\end{figure}
-
-Analogously to the just introduced type system of \problogsty, the base type for \dcproblogsty is again the \type{Term} type.
-At this point we modify the \problogsty type system by subtyping terms with a \type{Distrbution} type and a \type{Symbol} type.
-Dividing \dcproblogsty terms into \type{Distribution} terms and \type{Symbol} terms reflects the characterization of~\citet{poole2010probabilistic} of probabilistic programs as being constituted of independent choices (\type{Distribution}) and a deterministic system (\type{Symbol}).
-
-\begin{definition}\label{def:type_system}
-	The \dcproblogsty type system is a hierarchical type system with each type being a disjoint union of its subtypes.
-	\begin{enumerate}
-		\item[] \type{Term} 
-		\begin{enumerate}
-			\item[1.] \type{Distribution}
-			\begin{itemize}
-				\item \type{DiscreteD}
-				\item \type{ContinuousD}
-				\item \type{MixtureD}
-			\end{itemize}
-			\item[2.] \type{Symbol}
-			\begin{itemize}		
-				\item \type{Arithmetic}
-				\begin{itemize}
-					\item  \type{RandomVariable}
-					%			\begin{itemize}
-					%				\item \type{DiscreteRV}
-					%				\item \type{ContinuousRV}
-					%				\item \type{MixtureRV}
-					%			\end{itemize}
-					\item \type{Number}
-				\end{itemize}
-				\item \type{Atom}
-			\end{itemize}
-		\end{enumerate}
-	\end{enumerate}
-\end{definition}
-
-A diagrammatic overview of the \dcproblogsty type system is given in Figure~\ref{fig:type_system}.
-
-\begin{figure}[t]
-	\begin{center}
-		\includegraphics[width=\linewidth]{../figures/type_hierarchy.pdf}
-	\end{center}
-	\caption[Diagrammatic representation of the hierarchical \dcproblogsty type system.]{Diagrammatic representation of the hierarchical \dcproblogsty type system.
-Compared to Definition~\ref{def:type_system},  the \type{Number} type is further subtyped into \type{Integer} and \type{Float} types and the \type{RandomVariable} type is subtyped into \type{DiscreteRV}, \type{ContinuousRV}, and \type{MixtureRV}.
-These subtypes of \type{RandomVariable} mimic the subtypes of the \type{Distribution} type.}
-	\label{fig:type_system}
-\end{figure}
-
-%A type is a disjoint union of its subtytpes, for example:
-%$$\type{Term}=\type{Distribution} \cup \type{Symbol}$$
-
-\subsubsection{Distribution Terms}
-As already seen in Examples~\ref{example:dcproblog_machine} and~\ref{example:dcproblog_poisson_population}, \type{Distribution} terms determine the probability distribution according to which a random variable is distributed.
-
-\begin{definition}[\type{Distribution}]
-	Terms of type \type{Distribution} are constructed by using a reserved random function symbol and applying it to a (possible empty) tuple of input arguments, which maps the input tuple to a distribution term.
-	Distribution terms are only allowed to be invoked as a second argument to the reserved distribution predicate \probloginline{is}\lstinline[columns=fixed]|/2|.
-	This also means that random function symbols are only used there as well. \dcproblogsty subtypes the \type{Distribution} type into the \type{DiscreteD}, \type{ContinuousD}, and \type{MixtureD} types.
-\end{definition}
+\ak{I don't think we need to sub-type random variable functors here, as rv terms inherit the \emph{output} type of the distribution only, which should be integer, float etc, not "discrete distribution"}
 
@@
 \ak{would be nice to have a hint on why this restriction is there} We give an example of a distribution of type \type{MixtureD} in Example~\ref{ex:indian_gpa_function}.
-%\begin{example} \probloginline{mixed} is a random variables that is a mixture of a continuous part (the truncated normal between $0$ and $4$ with mean $3$ and standard deviation $1$) and a discrete part.
-%\begin{problog}
-%mixted~Mix((0.998,TruncNormal(3,1,0,4)),(0.001,4),(0.001,4))
-%\end{problog}
-%\end{example}
 
-%For and example, see Example~\ref{example:dcproblog_poisson_population}: we have the random function symbol \probloginline{poisson}\lstinline[columns=fixed]{/1}, apply it to the input tuple \probloginline{(6)}, and obtain a term of type \type{Distribution}.
-
-\subsubsection{Symbol Terms}
-\type{Symbol} terms are the building blocks of the discrete structure that makes up a logic program and as such determine also the deterministic system of a probabilistic logic program.
-Beyond the reserved syntax, which is used to define the distribution terms, the user is free to use any identifier (following \prologsty conventions) to declare \type{Symbol} terms -- this includes standard \prologsty terms such as lists.
-%The subtyping of the \type{Symbol} type (shown in Figure~\ref{fig:type_system}) follows the convention given in~\citep[Chapter 9]{sterling1994art}.
-
-\begin{definition}[\type{Arithmetic}]
-	Terms of type \type{Symbol} for which arithmetic operations are defined are of type \type{Arithmetic}.
-\type{Arithmetic} terms can be used as second argument to the builtin \probloginline{is}\lstinline[columns=fixed]|/2| functor.
-\type{Arithmetic} terms are subtyped with \type{RandomVariable} terms and \type{Number} terms.
-%	\ak{Def 6.3 Arithmetic terms: if (as stated in def 6.1) each type is the disjoint union of its subtypes, that means that "arithmetic" is made up exclusively of numbers and RVs, \ie, terms like 3-2 or min(x,5) for a RV x are not allowed, whereas the first two sentences of def 6.3 suggests they might be allowed (only having numbers and RVs directly on the rhs of is/2 seems quite boring...).
-Which is it?
-%		Related to this, defs 6.8 and 6.10 talk about RVs and Numbers, why not simply "arithmetic"?}
-\end{definition}
-We talk in more detail about \type{Arithmetic} terms and their evaluation in Subsection~\ref{sec:arithmetic}.
-
-\begin{definition}[\type{RandomVariable}] \label{def:random_variable}
-	\type{RandomVariable} terms are named by the programmer using distributional facts with the binary infix predicate \probloginline{is}\lstinline[columns=fixed]|/2| and inherit their subtype from the distribution term.
-All random variables present in a \dcproblogsty program are named.
-\type{RandomVariable} terms are a subtype of the \type{Arithmetic} type.
-\ak{Def 6.4: why do you restrict random variables to be arithmetic?
-What about color(Ball)~uniform([red,green,blue])?
-Of course, such finite discrete domains could be encoded using integers and then translated back in the LP, so theoretically speaking it's probably not a restriction to only have numbers, but from a modeling perspective, it is surprising to me.}
-\end{definition}
-
-Examples of terms of type \type{RandomVariable} would be \probloginline{temperature} and \probloginline{n_people} in Examples~\ref{example:dcproblog_machine} and~\ref{example:dcproblog_poisson_population}, where we declared the random variables through so-called {\em distributional facts}:
-\begin{problog}
-temperature ~ normal(20,5).
-n_people ~ poisson(6).
-\end{problog}
-As mentioned in Defintion~\ref{def:random_variable}, \type{RandomVariable} terms inherit their type from the associated \type{Distribution} term, \eg a \type{ContinuousD} term induces a random variable term of type \type{ContinuousRV}.
-
-Note that not all probabilistic programming languages enforce the naming of random variables.
-In the functional probabilistic programming language \anglicansty~\citep{wood2014approach}, for instance, the user can declare anonymous random variables, much like anonymous functions, \ie \textlambda-functions.
-\ak{Discussion of named/unnamed between Defs 6.4 and 6.5: would be nice to have a brief hint on why you made that choice}
-
-\begin{definition}[\type{Number}] Terms of type \type{Number} are used to express numerical data types such as integers and floating point numbers.	
-\end{definition}
-
-\ak{terminology mix-up; predicates never form terms}
-\begin{definition}[\type{Atom}]
-	An atom is of the form \probloginline{p(t@$_1$@,@\dots @, t@$_n$@ )} where \probloginline{p} is a predicate of arity $n$ and the \probloginline{t@$_i$@} are \type{Symbol} terms~\citep{fierens2015inference}.
-Predicates used to construct \type{Atom} terms must not clash with the builtin predicates and reserved random function symbols used to declare \type{Distribution} terms.
-Terms of type \type{Atoms} constitute the building blocks of the discrete structure of a probabilistic logic program.
-\end{definition}
-
-Examples of atoms terms would be \probloginline{foo} or \probloginline{bar(foo,42)}.
-The latter is constructed using the predicate \lstinline{bar/2} and the tuple of \type{Symbol} terms \probloginline{(foo,42)}.
-\probloginline{foo} is of type \type{Atom} and \probloginline{42} is a \type{Number} term.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Syntax}
-%Based on the type system, we now define the syntactic components of DC-ProbLog and its syntax.
-\begin{definition}[Deterministic Rules and Facts]
-	A (normal) logic program is a set of rules. A rule is a universally quantified expression of the form
-	\begin{problog}
-h:- b@$_1$@, @\dots@, b@$_n$@.
-	\end{problog}
-	where \probloginline{h} is an \type{Atom} term and $\{ \text{\probloginline{b@$_1$@}}, \text{\dots},\text{\probloginline{b@$_n$@}} \}$ is a set of $n$ literals.
-	Literals are either \type{Atom} terms, or their negation.
-	The \type{Atom} term \probloginline{h} is called the head of the rule and \probloginline{b@$_1$@, @\dots@, b@$_n$@} the body. The latter represents the
-	conjunction $\text{\probloginline{b@$_1$@}} \land \cdots \land  \text{\probloginline{b@$_n$@}}$. A fact is a rule that has \probloginline{true} as its body and is simply written as
-	\begin{problog}
-h.
-	\end{problog}
-\end{definition}
-
-\begin{definition}[Probabilistic Facts]
-	Probabilistic facts are of the form \probloginline{p::fact}.
-\probloginline{p} is a term of type \type{Number} or \type{RandomVariable}, with $0{\leq} \text{\probloginline{p}} {\leq} 1$  and \probloginline{fact} is a ground atom.
-	%\begin{definition} A {\bf probabilistic fact} is a ground definite fact labeled with a probability.
-\end{definition}
-
-\begin{definition}[Distributional Facts]
-	Distributional facts introduce random variable terms and associate them with a distribution using the reserved binary predicate \probloginline{~}\lstinline[columns=fixed]|/2| in infix notation.
-A distributional fact is a ground fact of the form \probloginline{rand ~ distribution}, where \probloginline{distribution} is a well-typed distribution term with output type \type{DistributionX} and \probloginline{rand} is a ground term whose functor is not reserved.
-The distributional fact introduces \probloginline{rand} as a random variable of type \type{RandomVariableX}, \ie the type of the random variable term mimics the type of the distribution term.
-\end{definition}
-Distributional facts can be thought of as the link between the independent choices (in form of distributions) and the discrete structure (in form of rules and facts).
-
-\begin{definition}[Comparison predicates] \label{def:comparison}
-	\dcproblogsty (similar to \prologsty and \problogsty) uses a fixed set of builtin binary predicates whose truth values depend on the \emph{evaluation} of its arguments.
-For terms that can be evaluated (\type{RandomVariable} terms and \type{Number} terms) we use the set
-	$\bowtie=\{ \text{\probloginline{=:=}}, \text{\probloginline{=\=}}, \text{\probloginline{<}}, \text{\probloginline{>}},  \text{\probloginline{=<}}, \text{\probloginline{>=}} \}$
-	of binary predicates written in infix notation.
-	Each of these has the usual \prologsty arithmetic interpretation of evaluating each side and comparing the results.
-	Just as for \prologsty arithmetic, truth values of atoms using these predicates are determined through an evaluation process that is external to the (probabilistic) logic program.
-Such atoms can be used in clause bodies, but never in clause heads (or as facts).
-\end{definition}
-
-%We will further discuss arithmetic evaluations in Subsection~\ref{sec:arithmetic}.
-
-\begin{example} Comparison predicates allow for the comparison of random variables to numbers. 
-	\begin{problog*}{linenos}
-x ~ normal(0,3).
-q:- x>0.
-query(q).
-	\end{problog*}
-\end{example}
-
-%\begin{definition}\label{def:distributional_fact}
-%	A {\bf distributional fact} is a ground definite fact with the reserved binary predicate \probloginline{is}\lstinline[columns=fixed]|/2| used in infix notation.
-%\end{definition}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Multiple Dispatch} \label{ref:multiple_dispatch}
@@
 
-\begin{example} A \dcproblogsty code snippet showing how random variable terms can be used in an arithmetic expression. 
+\begin{example}\label{ex:is} A \dcproblogsty code snippet showing how random variable terms can be used in an arithmetic expression. 
 	\begin{problog*}{linenos}
@@
 \subsubsection*{Strongly Typed \prologsty}
+\ak{move to related work?}
 

### 2024-09-10 202409_aij_dc_problog_aij_second/main.tex
aij dcproblog paper

%\documentclass[preprint]{elsarticle}
\documentclass[final]{elsarticle}

\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{}%
 \let\@evenfoot\@oddfoot}
\makeatother

\usepackage{lineno}
\modulolinenumbers[5]

\usepackage{hyperref}

%% `Elsevier LaTeX' style
% \bibliographystyle{elsarticle-num}
\bibliographystyle{plainnat}
% \biboptions{authoryear}

\newcommand\hmmax{0}
\newcommand\bmmax{0}

%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsmath}
\usepackage{amssymb,amsthm}

\usepackage{stmaryrd}
\usepackage[noend]{algpseudocode}
\usepackage[dvipsnames,table,xcdraw]{xcolor}
\usepackage{xspace}
% \usepackage[cachedir=minted-cache,newfloat]{minted}
% \usepackage[finalizecache=true,cachedir=minted-cache,newfloat]{minted}
\usepackage[frozencache=true,cachedir=minted-cache,newfloat]{minted}
% \usepackage[utf8]{inputenc}
\usepackage{caption}
\usepackage{listings}
\usepackage{nicefrac}
\usepackage{multirow}
\usepackage{textcomp} % nice greek alphabet
\usepackage{textgreek}
\usepackage{changepage}
\usepackage{cancel}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{arrows, calc,shapes.geometric,shapes.multipart,positioning,tikzmark}
\usetikzlibrary{arrows,shapes,trees}
\usepackage{enumitem}
\usepackage{multicol}
\usepackage{bm}

\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{mdframed}
\usepackage{scrextend}
\usepackage{blindtext}
\usepackage[section]{placeins}

\usepackage{xassoccnt}
\usepackage{chngcntr}
\usepackage{thm-restate}
\usepackage{xpatch}
\usepackage{fixmath}
\usepackage{soul}

\usepackage[htt]{hyphenat}

\hyphenation{data-base ground-ers}

\usepackage[linesnumbered,ruled,vlined,algosection]{algorithm2e}
\DontPrintSemicolon
\SetKwInput{KwInput}{Input}                % Set the Input
\SetKwInput{KwOutput}{Output}              % set the Output
\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{gray}{#1}}
\SetCommentSty{mycommfont}

% \usepackage[bbgreekl]{mathbbol}

\graphicspath{{figures/}}

%%%%%%%%%%%%%%%%%%%%%%
\input{macros}
%%%%%%%%%%%%%%%%%%%%%%%

\numberwithin{equation}{section}

%Commands definitions
\newcommand{\setbackgroundcolour}{\pagecolor[rgb]{0.19,0.19,0.19}}
\newcommand{\settextcolour}{\color[rgb]{0.77,0.77,0.77}}
\newcommand{\invertbackgroundtext}{\setbackgroundcolour\settextcolour}

%Command execution. 
%If this line is commented, then the appearance remains as usual.
% \invertbackgroundtext

\setlist[itemize]{itemsep=0pt}
\setlist[enumerate]{itemsep=0pt}

\begin{document}

\setcounter{page}{1}
\renewcommand{\thepage}{\roman{page}}

\setcounter{section}{0}
\renewcommand{\thesection}{R\arabic{section}}

\renewcommand{\theequation}{R\arabic{section}.\arabic{equation}}

% \input{rebuttal}
% \clearpage

\setcounter{page}{1}
\renewcommand{\thepage}{\arabic{page}}

\setcounter{section}{0}
\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\theequation}{\arabic{section}.\arabic{equation}}

\begin{frontmatter}

    \title{Declarative Probabilistic Logic Programming\\in Discrete-Continuous Domains}
    % \tnotetext[mytitlenote]{}

    \author[orebro]{Pedro Zuidberg Dos Martires}
    \author[orebro,kuleuven,leuvenai]{Luc De Raedt}
    \author[kuleuven,leuvenai]{Angelika Kimmig}

    \address[orebro]{Centre for Applied Autonomous Sensor Systems, Örebro University, Sweden}
    \address[kuleuven]{Department of Computer Science, KU Leuven, Belgium}
    \address[leuvenai]{Leuven.AI, Belgium}

    % \begin{abstract}
    % A core strength of probabilistic logic programming languages, which belong to the best understood probabilistic programming languages in the scientific literature, are their declarative semantics. However, this is only true for programs limited to random variables with finite sample spaces. When extending PLP languages to the discrete-continuous domain, procedural semantics have usually been preferred over declarative semantics.
    % We introduce \dcproblogsty, a PLP language capable of representing random variables in the discrete-continuous domain, which we equip with purely declarative semantics.
    % Additionally, we develop the first inference algorithm for a probabilistic programming language in the discrete-continuous domain based on knowledge compilation. As such, we generalize conventional PLP in a twofold fashion. On the hand, \dcproblogsty extends the declarative semantics to random variables with infinite (and uncountable) sample spaces, instead of finite ones only. On the other hand, we generalize the state-of-the-art knowledge compilation approach towards inference in conventional PLP languages to the discrete-continuous domain.
    % \end{abstract}

    \begin{abstract}
        Over the past three decades, the logic programming paradigm has been successfully expanded
        to support probabilistic modeling, inference and learning. The resulting paradigm
        of probabilistic logic programming (PLP) and its programming languages owes much of its success to a declarative semantics,
        the so-called distribution semantics. However, the distribution semantics is limited to discrete random variables only.
        While PLP has been extended in various ways for supporting hybrid, that is, mixed discrete and continuous
        random variables, we are still lacking a declarative semantics for hybrid PLP that not only generalizes
        the distribution semantics and the modeling language but also the standard inference algorithm
        that is based on knowledge compilation.
        We contribute the {\em measure semantics} together with the hybrid PLP language
        \dcproblogsty (where DC stands for distributional clauses) and its inference engine {\em infinitesimal algebraic likelihood weighting} (IALW).
        These have the original distribution semantics, standard PLP languages
        such as \problogsty, and standard inference engines for PLP based on knowledge compilation as special cases.
        Thus, we generalize the state of the art of PLP towards hybrid PLP in three different aspects: semantics, language and inference.
        Furthermore, IALW is the first inference algorithm for hybrid probabilistic programming based on knowledge compilation.
    \end{abstract}

    \begin{keyword}
        Probabilistic Programming \sep Declarative Semantics \sep Discrete-Continuous Distributions \sep Likelihood Weighting \sep Logic Programming \sep Knowledge Compilation \sep Algebraic Model Counting
    \end{keyword}

\end{frontmatter}

% \linenumbers

\input{files_main/introduction}

\input{files_main/panorama}

\input{files_main/semantics}
%
\input{files_main/dcproblog}
%
\input{files_main/tasks}
%
\input{files_main/dc2smt}
%
\input{files_main/evaluating}

\input{files_main/related_work}

\input{files_main/conclusions}

\section*{Acknowledgement}

This research received funding from the Wallenberg AI, Autonomous Systems and Software Program (WASP) of the Knut and Alice Wallenberg Foundation, the Flemish Government (AI Research Program),
the KU Leuven Research Fund, the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No [694980] SYNTH: Synthesising Inductive Data Models), and
the Research Foundation - Flanders.

\bibliography{references}

\newpage
\clearpage

\appendix

% \setcounter{section}{0}
% \renewcommand{\thesection}{\Roman{section}}
% \renewcommand{\theequation}{\Roman{section}.\arabic{equation}}
% \renewcommand{\thetheorem}{\Roman{section}.\arabic{theorem}}
% \renewcommand{\thedefinition}{\Roman{section}.\arabic{definition}}

% \input{files_appendix1/lp}

% \input{files_appendix1/amc}

% \input{files_appendix1/compilation}

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}
\renewcommand{\theequation}{\Alph{section}.\arabic{equation}}
\renewcommand{\thetheorem}{\Alph{section}.\arabic{theorem}}
\renewcommand{\thedefinition}{\Alph{section}.\arabic{definition}}

\input{files_appendix2/lp_new}

\input{files_appendix2/app_table}

\input{files_appendix2/semantics_proofs}

\input{files_appendix2/syntactic_sugar_semantics.tex}

\input{files_appendix2/detailed_comp_to_GutmanNittiDC}

% \input{files_appendix2/smart_grounding}

\input{files_appendix2/transformation_proofs}

% \input{files_appendix2/labeled_propositional_to_SMT}

% \input{files_appendix2/implementation}

\

\end{document}

### 2024-09-10 202409_aij_dc_problog_aij_second/experiment_dc/build.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+set -exu
+docker build --tag local/dc_problog -f Dockerfile ./

### 2024-09-10 202409_aij_dc_problog_aij_second/experiment_dc/faulty.pl

@@ -1,0 +1,93 @@
+%%% -*- Mode: Prolog; -*-
+
+:- use_module('../distributionalclause.pl').
+:- use_module('../random/sampling.pl').
+:- use_module(library(lists)).
+
+:- set_options(default),set_inference(backward(lw)), set_query_propagation(true). % enable query propagation
+
+
+
+
+
+
+
+mode ~ finite([0.3:mode1,0.7:mode2]).
+
+pfaulty ~ val(_).
+faulty ~ finite([PF:true,NPF:false]):= pfaulty~=PF, NPF is 1-PF.
+
+
+temperature ~ gaussian(0.5,1.0) := faulty=false, mode~=mode1.
+temperature ~ gaussian(2.0,2.0) := faulty~=false, mode~=mode2.
+temperature ~ val(2.0) := faulty~=true.
+
+
+
+	
+fco(N,PF) :-
+	init,
+	eval_query([temperature~=2.0,pfaulty~=PF],[],faulty~=true,N,P,A,B), 
+	write('probability faulty~=true: '),writeln(P).
+
+
+all_experiments :-
+	init,
+	findall(
+		(N,PF), 
+		(
+			member(N,[10,100,1000,10000,100000]),
+			member(PF,[0.1,0.01,0.001,0.0001,0.00001])
+		),
+		List
+		),
+	experiment(100,List),
+	true.
+
+
+experiment(_,[]):- true.
+
+experiment(Runs, [(N,PF) | Tail]) :-
+	File='faulty.txt',
+	Q=[faulty~=true],
+	E=[temperature~=2.0,pfaulty~=PF],
+	experiment_lw(File,Q,E,N,Runs,PF),
+	experiment(Runs, Tail),
+	writeln(N),
+	writeln(PF),
+	true.
+
+
+experiment_lw(File,Q,E,N,Repeat,PF) :-
+	I is cputime,
+	findall(
+		P,
+		(
+			between(
+				1,
+				Repeat,
+				_
+			),
+			distributionalclause:eval_query_backward_lw(
+				E,[],Q,1,N,P1,_,SSum
+			),
+			(
+				SSum>0 -> P=P1;P is random
+			)
+		),
+		List
+	),
+	T is (cputime-I)/Repeat,
+	(
+		ground(AVG)->
+			variance(List,AVG,V);
+			avgvar(List,AVG,V)
+	),
+	
+	
+	open(File,'append',F),
[…]

### 2024-09-10 202409_aij_dc_problog_aij_second/problog.py

@@ -1,0 +1,315 @@
+# -*- coding: utf-8 -*-
+"""
+    Lexers for ProbLog.
+    :license: BSD, see LICENSE for details.
+"""
+
+import re
+from pygments import token
+
+from pygments.util import guess_decode
+from pygments.lexer import _encoding_map
+from pygments.filter import apply_filters
+from pygments.lexer import RegexLexer, bygroups, words, include
+from pygments.token import (
+    Text,
+    Comment,
+    Operator,
+    Keyword,
+    Name,
+    String,
+    Escape,
+    Number,
+    Punctuation,
+)
+
+__all__ = ["ProbLogLexer"]
+
+
+class ProbLogLexer(RegexLexer):
+    """
+    Lexer for Prolog files.
+    """
+
+    name = "ProbLog"
+    aliases = ["problog"]
+    filenames = ["*.ecl", "*.prolog", "*.pro", "*.pl"]
+    mimetypes = ["text/x-problog"]
+
+    flags = re.UNICODE | re.MULTILINE
+
+    def predicates_callback(lexer, match):
+        print("predicates_callback")
+        print(match)
+        # print(match.group(0), 0),
+        # print(match.group(1), 1)
+        # print(match.group(2), 2)
+        # print(match.group(3), 3)
+
+        if match.group(1) in lexer.random_variables:
+            token = Punctuation
+        else:
+            token =Punctuation
+
+        return bygroups(token, Text, Punctuation)(lexer, match)
+
+    def atoms_callback(lexer, match):
+
+        print("atoms_callback")
+        print(match)
+        if match.group(0) in lexer.random_variables:
+            token = Punctuation
+        else:
+            token = Punctuation
+
+        yield match.start(0), token, match.group(0)
+
+    def atoms_callback_bang(lexer, match):
+        print("atoms_callback_bang")
+        print(match)
+        if match.group(0) in lexer.random_variables:
+            token = Punctuation
+        else:
+            token = Punctuation
+
+        yield match.start(0), token, match.group(0)
+
+    tokens = {
+        "root": [
+            (r"/\*", Comment.Multiline, "nested-comment"),
+            (r"%.*", Comment.Single),
+            # character literal
+            (r"\@(.*?)\@", Escape),
+            include("comparison_builtins"),
+            include("operators"),
+            (r"0\'.", String.Char),
+            include("numbers"),
+            include("builtins"),
[…]

### 2025-06-02 202507_uai_puncs/iclr2025/content.tex
added uai

\maketitle

\begin{abstract}
	By recursively nesting sums and products, probabilistic circuits (PCs) have emerged in recent years as an attractive class of generative models as they enjoy, for instance, polytime marginalization of random variables.
	However, as these PCs form monotone functions, specifically the parameters are constrained to positive real values, their expressive power is limited.
	Drawing inspiration from the PC literature and quantum information theory we introduce a novel class of probabilistic models, which we dub \textit{positive operator circuits} (\pocs).
	Contrary to the parametrization of PCs (constrained to positive real values), \pocs use complex valued parameters.
	By leveraging concepts from quantum information theory, we show that \pocs encode proper probability distributions, while retaining the tractable marginalization property of PCs.
	We also give a  prescription for constructing
	an efficient parametrization of \pocs that enables us to perform parameter learning on hardware accelerators using standard gradient-based optimization.
	Lastly, we provide an implementation and experimental evaluation for \pocs, which establishes \pocs as an attractive class of tractable probabilistic models.
\end{abstract}

\section{Introduction}

Probabilistic circuits (PCs)~\citep{darwiche2003differential,poon2011sum} belong to an unusual class of probabilistic models: they are highly expressive but at the same time also tractable.
For instance, so-called decomposable probabilistic circuits~\citep{darwiche2001decomposable} allow for the computation of marginals in time polynomial in the size of the circuit.
From an abstract algebra perspective, PCs constitute computation graphs that perform their computations within the probability semiring. That is, they sum and multiply elements from the $[0,1]$ interval with each other.

\citet{zhang2020relationship} noted that it is exactly this restriction to positive values that limits the expressive efficiency (or succinctness) of PCs~\citep{martens2014expressive,decolnet2021compilation}. In particular, the positivity constraint on the set of elements that PCs operate on prevents them from modelling negative correlations between variables.
Circuits that are incapable of modelling negative correlations, \ie circuits that can only combine probabilities in an additive fashion, are also called monotone circuits~\citep{shpilka2010arithmetic}.
This restricted expressiveness can be combatted by the use of so-called \textit{non-monotone} circuits, where subtractions are allowed as a third operation (besides sums and products). Interestingly, \citet{valiant1979negation} showed that a mere single subtraction can render non-monotone circuits exponentially more expressive than monotone circuits.

As shown by \citet{harviainen2023inference}, non-monotone circuits do, however, introduce an important complication: if non-monotone circuits are not designed carefully, verifying whether a circuit encodes a valid probability distribution or not is an NP-hard problem. This does also render learning the parameters of a circuit practically infeasible.

We will formulate, in a first instance, the class of layered probabilistic circuits (\cf \citet{peharz2019random,peharz2020einsum}) in terms of, what we call, partition circuits (Section~\ref{sec:pc}).
This will allow us, Together with a slight generalization of concepts from quantum information theory (Section~\ref{sec:qit}),
to concisely introduce \textit{positive operator circuits} (\pocs) --
a novel non-monotone circuit class that uses complex valued parameters (instead of positive real ones) and that forms by construction valid probability distributions (Section~\ref{sec:pocs}).
Lastly, we demonstrate the benefit of modelling probability distributions using positive operator circuits (Section~\ref{sec:experiments}), in which \pocs outperform competing (tractable) density estimators on datasets from the MNIST family but where we also point out important open questions.
In Section~\ref{sec:related} we study the related work, and we give concluding remarks in Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Partition Circuits}
\label{sec:pc}

\begin{figure}[t]

	\centering
	\input{tex_input/partition}

	\caption{Left: Layered PC over four binary variables $X_i$ with $i \in \{0,1,2,3\}$ taking values $x_i^0$ or $x_i^1$.
		Within each partition in the bottom layer three mixtures for each of the four $X_i$'s are constructed using weighted sums (weights are not shown in the graphical representation but are present on the edges feeding into sum units).
		The sum and product units in the circuit are the elemental constitute the elemental computing units and make the partitions.
		The circuit in Figure~\ref{fig:circuit} has, except at the very top and bottom, three components in each partition.
		As increasing the number of components per partition increases the number of parameters (weights on edges feeding into sum units), the number of components per partition controls the capacity of a circuit.
		Right: partition tree abstracting the layered PC on the right.}
	\label{fig:circuit}
\end{figure}

Inspired by simple feed-forward neural networks, \citet{peharz2019random,peharz2020einsum} introduced the concept of layered probabilistic circuits.
These layered circuits are amenable to trivial parallelization and can be run on modern discrete GPUs. In Figure~\ref{fig:circuit} we give a graphical representation of such a layered circuit where we follow the construction introduced by \citet{shih2021hyperspns}. For the sake of conciseness we describe here only so-called \textit{structured decomposable and smooth} PCs. For a broader overview we refer the reader to the excellent work of \citet{vergari2021compositional}.

Layers within a layered PC constitute blocks of computational units that are processed sequentially in a bottom-up fashion. Layers consist themselves of so-called partitions. The circuit in Figure~\ref{fig:circuit} has four partitions in the leaf layer (very bottom), two in the subsequent sum and product layers, and a single partition in the last product layer and in the final root node (top most sum). By construction, partitions in the same layer have disjoint scopes. That is, they are functions over disjoint sets of variables. This property is called structured decomposability in the PC literature~\citep{darwiche2011sdd}. Note also that layered PCs are by construction smooth~\citep{darwiche2001tractable}: the union of the scopes of the partitions within a layer is exhaustive. That is, the union of the scopes equals the set of variables given as input to the circuit.

A (layered) PC is then parametrized by weighing the inputs to the sum units with positive real numbers -- giving rise to an unnormalized probability distribution over the input variables. Due to the properties of (structured) decomposability and smoothness the distribution can be normalized in time polynomial in the size of the circuit~\citep{peharz2015theoretical}.
Thanks to the properties of smoothness and (structured) decomposability we can also marginalize out single variables from a PC. Again in time polynomial in the size of the circuit.

Recently, \citet{zuidberg2024probabilistic} introduced the concept of \textit{partition trees} to abstract away certain aspects of smooth and structured decomposable probabilistic circuits. We give such a partition tree on the right side of Figure~\ref{fig:circuit}. We now take this abstraction a step further and define the computations performed by the probabilistic circuit not on the atomic computation units of the circuit itself (\ie sum and product units) but use the nodes in the partition tree as the elemental computation units. In other words, we regard the partition tree as a computation graph. To make this distinction explicit we dub these computation graphs \textit{partition circuits}. For the sake of conciseness, we will limit ourselves in this paper to balanced binary partition trees and will assume that the number of input variables is a power of two (unless indicated otherwise). Consequently, we will study partition circuits of depth $\log_2 M+1$ where $M$ is the number of input variables.

\begin{definition}[Partition Circuit]
	\label{def:partition_circuit}
	A partition circuit over a set of variables is a parametrized computation graph taking the form of a binary tree. The partition circuit consists of three kinds of computation units:
	\textit{leaf} and \textit{internal} units, as well as a single \textit{root}.
	Units at the same distance from the root form a layer.
	Furthermore, let $\circuit_n$ denote the root unit or an internal unit. The unit $\circuit_n$ then receives its inputs from two units in the previous layer, which we denote by $\circuit_{n_l}$ and $\circuit_{n_r}$. Each computation unit is input to exactly one other unit, except the root unit, which is the input to no other unit.
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Adapting Quantum Information Theory for Probabilistic Modelling}
\label{sec:qit}

A widely used and elegant framework to describe measurements of quantum systems is the so-called \textit{positive operator-valued measure} (POVM) formalism. While POVMs have physical interpretations in terms of quantum information and quantum statistics, we will only be interested in their mathematical properties as we will use them to show that positive operator circuits (defined in Section~\ref{sec:pocs}) form valid probability distributions.
We refer the reader to \citep{nielsen2001quantum} for an in-depth exposition on the topic, as well as quantum computing and quantum information theory in general.

\begin{definition}[Positive Semidefinite]
	An $N {\times} N$ Hermitian matrix $H$ is called positive semidefinite (PSD) if and only if $\forall \mathbf {x} {\in}  \mathbb {C} ^{N}: \mathbf{x}^{*} H \mathbf{x} {\geq} 0$, where $\mathbf{x}^{*}$ denotes the conjugate transpose and $\mathbb {C}^{N}$ the N-dimensional space of complex numbers.
\end{definition}
% We will refer to Hermitian PSD matrices as HPSD in the remainder of the paper.
\begin{definition}[{POVM~\citep[Page 90]{nielsen2001quantum}}]
	\label{def:povm}
	A positive operator-valued measure
	% with a finite number of elements acting on a finite-dimensional Hilbert space $\mathcal{H}$,
	is a set of PSD  matrices $\{E(i)\}_{i=0}^{V-1}$ ($V\in \mathbb{N}$ being the number of possible measurement outcomes) that sum to the identity:
	\begin{talign}
		\sum_{i=0}^{V-1} E(i) = \mathbb{1},
		\label{eq:povm_normalized}
	\end{talign}
\end{definition}
Before defining the probability of a specific $i$ occurring, we need the notion of a density matrix~\citep{neumann1927wahrscheinlich}:
\begin{definition}[{Density Matrix~\citep[Page 102]{nielsen2001quantum}}]
	\label{def:density_matrix}
	A density matrix $\rho$ is a PSD matrix of trace one, \ie $\Tr [\rho]=1$.
\end{definition}
\begin{definition}[{Event Probability~\citep[Page 102]{nielsen2001quantum}}]
	\label{def:eventprob}
	Let $\rho$ be a density matrix and let $i$ denote an event with $E(i)$ being the corresponding element from the POVM. The probability of the event $i$ happening, i.e. measuring the outcome $i$, is given by
	\begin{talign}
		p(i) = \Tr [ \rho E(i)   ]
		\label{eq:povm_prob}
	\end{talign}
\end{definition}

\begin{restatable}{proposition}{proppovmprob}
	\label{prop:povmprob}
	The expression in Equation~\ref{eq:povm_prob} defines a valid probability distribution.
\end{restatable}

\begin{proof}
	While this is a well-known result we were not able to identify a concise proof in the literature and provide therefore, for the sake of completeness, a proof in Appendix~\ref{sec:proof:prop:povmprob}
\end{proof}

From a formal perspective, POVMs constitute an elegant way of defining probability distributions. However, from a computational perspective the constraint that $\sum_i E(i)=  \mathbb{1}$ causes some issues. While one could enforce the constraint using eigenvalue or singular value decompositions, performing these within a learning loop we would need to differentiate through the decomposition algorithms, which could lead to numerical stability issues. Especially, when computing gradients.\footnote{For instance, the PyTorch documentation mentions some of these concerns for singular value decompostions
	\href{https://pytorch.org/docs/stable/generated/torch.linalg.svd.html}{(link)} and for eigenvalue value decompositions \href{https://pytorch.org/docs/stable/generated/torch.linalg.eig.html}{(link)}.
}
To circumvent these issues we (slightly) generalize Definition~\ref{def:eventprob}.

\begin{definition}[Z-Normalized Event Probability]
	\label{def:Zeventprob}
	Let $\rho$ be a PSD matrix and let $i$ denote an event with $E(i)$ being the associated PSD matrix. The probability of the event $i$ happening, i.e. measuring the outcome $i$, is given by
	\begin{talign}
		\Trace \left[ \rho E(i)   \right] \big/ Z , \qquad \text{where } Z = \sum_{i=0}^{V-1} \Trace \left[ \rho E(i)  \right].
		\label{eq:Zeventprob:def}
	\end{talign}
\end{definition}

In comparison to Definition~\ref{def:eventprob}, Definition~\ref{def:Zeventprob} does not only drop the requirement that $\sum_i E(i)=  \mathbb{1}$ but also that $\Trace [\rho] =1$.

\begin{restatable}{proposition}{propZpovmprob}
	\label{prop:Zpovmprob}
	The expression in Equation~\ref{eq:Zeventprob:def} defines a valid probability distribution.
\end{restatable}

\begin{proof}
	In order to prove this we need to show that $\forall: \Trace [E(i) \rho ]\geq 0$. This then implies that $Z>0$ (assuming that at least one event has non-zero probability). Note also that the constraint $ \sum_i \Trace [E(i) \rho ] \big/ Z  =1$ is trivially satisfied by construction.
	Showing that  $\forall: \Trace [E(i) \rho ]\geq 0$ amounts to showing that the trace of a matrix-matrix product of two PSD matrices is positive real-valued. We already prove this as a sub-step in the proof for Proposition~\ref{prop:povmprob} (\cf Equation~\ref{eq:proof:trpsdprod} in the appendix). This finishes the proof.
\end{proof}

\section{Positive Operator Circuits}
\label{sec:pocs}
With the slightly generalized definition of an event probability at hand we are now ready to define positive operator circuits (using partition circuits).
\begin{definition}[Positive Operator Circuit]
	\label{def:poc}
	Let $\Xvars=\{\Xvar_0,\dots ,\Xvar_{M-1}  \}$ (taking values  $\xvars=\{\xvar_0,\dots ,\xvar_{M-1}  \}$) be a set of categorical random variables of sample space size $N$. Furthermore, let $B$ be a positive integer and let $\rho \in \mathbb{C}^{B\times B}$ be a PSD matrix.
	We define a positive operator circuit as a partition circuit whose computation units take the following functional form:
	\begin{align}
		{\circuit}_k(\xvars_k)=
		\begin{cases}
			F_{k} \times  e_{\xvar_k} \otimes  e^*_{\xvar_k} \times F_{k}^*,
			\quad e_{\xvar_k} \in  \mathbb{C}^{N},  F_{k} \in \mathbb{C}^{B\times N}
			 & \text{if $k$ leaf unit}
			\\
			G_k \times \Bigl( \circuit_{k_l}(\xvars_{k_l}) \odot  \circuit_{k_r} (\xvars_{k_r}) \Bigr) \times G_{k}^*,
			\quad G_k \in \mathbb{C}^{B\times B}
			 & \text{if $k$ internal unit}
			\\
			\Trace \Bigl[  \rho \times \Bigl( \circuit_{k_l}(\xvars_{k_l}) \odot  \circuit_{k_r} (\xvars_{k_r}) \Bigr)  \Bigr] \Big/ Z  ,
			\quad \rho \in \mathbb{C}^{B\times B},Z \in \mathbb{R}_{>0}
			 & \text{if $k$ root unit.}
		\end{cases}
		\label{eq:poc:def}
	\end{align}
	Here we use the symbols $\times$, $\otimes$, $\odot$ and $\cdot$ to denote the matrix product, Kronecker product, Hadamard product, respectively.
	Furthermore, $\{e_{\xvar_n} \}_{n=0}^{N-1}$ is a set of $N$-dimensinoal orthonormal basis vectors associating each outcome of the random variable $\Xvar_k$ to a specific basis vector.
	The symbol $Z$ denotes the normalization constant for which we require:
	\begin{align}
		Z
		=
		\sum_{\xvars \in \Omega(\Xvars)}
		\Trace \Bigl[  \rho \times \Bigl( \circuit_{root_l}(\xvars_{root_l}) \odot  \circuit_{root_r} (\xvars_{root_r}) \Bigr)  \Bigr]
	\end{align}

	% \begin{align}
	% 	\forall \Xvar_n: \sum_{\xvar\in\Omega(\Xvar_n)} f_n(\xvar) \otimes  f^*_n(\xvar) =\mathbb{1}_{B\times B},
	% 	\qquad
	% 	\forall n: U_n \times U_n^* =\mathbb{1}_{B\times B},
	% 	\label{eq:def:punc:con}
	% \end{align}
	% where $\mathbb{1}_{B\times B}$ denotes the $B\times B$ identity matrix.
\end{definition}
% Note that the second normalization condition ensures that the $U_n$ matrices are unitary. From which we derive the eponymous probabilistic circuits' name. Using the POVM formalism we can also show that \puncs form probability distributions.

\begin{restatable}{theorem}{theopocvalid}
	\label{theo:pocvalid}
	\pocs are valid probability distributions.
\end{restatable}

\begin{proof}
	We need to prove that at the root we have $\circuit_{root}(\xvars)\geq 0$, for every $\xvars \in \Omega(\Xvars)$ and that $\sum_{\xvars \in \Omega(\Xvars_{root})} \circuit_{root}(\xvars)=1$.
	The latter requirement is satisfied by construction via the normalization constant $Z$. For the former, let us first define:
	\begin{align}
		\mu_{root}(\xvars) \coloneqq \circuit_{root_l}(\xvars_{l}) \odot  \circuit_{root_r}(\xvars_{r}).
	\end{align}
	By identifying each element $\xvars \in \Omega(\Xvars)$ with an event $i$ from Definition~\ref{def:Zeventprob}, consequently $\mu_{root}(\xvars)$ with $E(i)$, we observe that the root of a \poc has the functional form of the probability of an element from a (generalized) POVM.
	In order to prove that $\forall \xvars \in \Omega(\Xvars): \Trace[\mu_{root}(\xvars) \rho]{>}0$,
	we now only need to show that $\mu_{root}(\xvars)$ is PSD,
	as $\rho$ is already PSD by assumption.
	We show this in Appendix~\ref{sec:proof:pocvalid}.
\end{proof}

\subsection{Discussion on Computational Complexity}
\label{sec:complexity}

In order to determine the computational complexity of evaluating a positive operator circuit, we simply study the cost of the individual computation units. We start at the leaves where we first have a Kronecker product $ e_{\xvar_k} \otimes  e_{\xvar_k}^*$, which can be performed in $\bigO (N^2)$. The matrix products with $F_k$ and $F_k^*$ can then be performed in $\bigO (B N^2)$. In the internal layers we perform the Hadamard product in $\bigO (B^2)$, followed again by two matrix-matrix products in $\bigO (B^3)$. In the root we compute the numerator of the fraction with a Hadamard product ($\bigO(B^2)$). The trace of the resulting matrix-matrix product can then be obtained in $\bigO (B)$. This means that evaluating a circuit (ignoring the normalization constant for now) has a cost of $\bigO (BN^2)$ per computation unit (assuming $B<N$). Furthermore, it is easy to show that partition circuits, where computation units have two inputs, have a total of $\bigO (M)$ computation units, with $M$ being the number of input variables. We conclude that the computational complexity of evaluating a positive operator circuit is $\bigO (MBN^2)$.

In the discussion above we assumed that the normalization constant $Z$ was given. Usually, this is a strong assumption as computing the normalization constant is in general computationally hard. However, for \pocs we can, similarly to probabilistic circuits, exploit the decomposability property and compute $Z$ in polytime via tractable marginalization.

\begin{restatable}{proposition}{proppocmargtract}
	\label{prop:pocmargtract}
	\pocs allow for tractable marginalization.
\end{restatable}

\begin{proof}
	The proof is trivial, as marginalizing out a variable $\Xvar_m$ from a \poc simply amounts to pushing down the summation to the corresponding leaf. This is possible as we can exchange at the root the summation and the trace, and exploit that the internal computation units are multilinear functions in terms of the input variables. At the leaf $\circuit_m (\xvar)$ we then have:
	\begin{align}
		\sum_{\xvar\in \Omega(\Xvar_m)} F_{m} \times  e_{\xvar} \otimes  e^*_{\xvar} \times F_{m}^*
		=
		F_{m} \times  \left( \sum_{\xvar\in \Omega(\Xvar_m)}  e_{\xvar} \otimes  e^*_{\xvar} \right) \times F_{m}^*
		 &
		=
		F_{m} \times  \mathbb{1} \times F_{m}^*
		\nonumber
		\\
		 &
		=
		F_{m} \times  F_{m}^*,
	\end{align}
	where we exploit the fact that the $e_n$'s are orthonormal basis vectors, which allows us to replace the sum with the identity matrix.
\end{proof}

If we marginalize out all variables $\Xvars$ using the prescription described in the proof of Proposition~\ref{prop:pocmargtract}, we obtain a circuit that does not dependent anymore on any inputs but which is still evaluable in time $\bigO (MBN^2)$ (using identical reasoning to the argument for the input dependent case).
We conclude that we can compute $Z$ in time polynomial in the size of the circuit and that positive operator circuits can be evaluated in $\bigO (MBN^2)$ time. We show next how to bring this down to $\bigO (MBN)$.

\subsection{The Vector Representation of Positive Operator Circuits}
\label{sec:vpoc}

The cubic computational dependency of the computation cost per computation unit $bigO (BN^2)$ is due to the presence of matrix-matrix multiplications. By rearranging the operations performed in the positive operator circuit we can avoid these matrix-matrix multiplications entirely and obtain matrix-vector multiplications instead. We call these alternative representation of \pocs \textit{positive vector circuits}.

\begin{definition}[Positive Vector Circuit]
	\label{def:vpox}
	Let $\Xvars=\{\Xvar_0,\dots ,\Xvar_{M-1}  \}$ (taking values  $\xvars=\{\xvar_0,\dots ,\xvar_{M-1}  \}$) be a set of categorical random variables of sample space size $N$ and let $\{e_{\xvar_n} \}_{n=0}^{N-1}$ be a set of orthonormal basis vectors with elements from the sample space being bijectively associated to basis vectors. Furthermore, let $B$ be a positive integer and let $\rho \in \mathbb{C}^{B\times B}$ be a PSD matrix.
	We define a positive vector circuit as a partition circuit whose computation units take the following functional form:
	\begin{align}
		{\xircuit}_k(\xvars_k)=
		\begin{cases}
			F_{k} \times  e_{\xvar_k},
			 & \text{if $n$ leaf unit}
			\\
			G_k \times \Bigl( \xircuit_{n_l}(\xvars_{k_l}) \odot  \xircuit_{k_r} (\xvars_{k_r}) \Bigr),
			 & \text{if $k$ internal unit}
			\\
			\nu_{k}(\xvars_k)  \cdot \nu^*_k(\xvars_k),
			 & \text{if $k$ root unit,}
		\end{cases}
		\label{eq:vector_units}
	\end{align}
	where we have $\odot$, $\times$, and $\cdot$ denoting the Hadamard product, the matrix-vector product, and the dot product, respectively. Furthermore, we used:
	\begin{align}
		\nu_{root}(\xvars_{root})
		\coloneqq
		\gamma
		\times
		\Bigl(
		\xircuit_{k_l}(\xvars_{root_l})
		\odot
		\xircuit_{k_r}(\xvars_{root_r})
		\Bigr)
		\big/ \sqrt{Z},
		\label{eq:pvc:def}
	\end{align}
	with $\gamma \in \mathbb{C}^{B\times B}$ such that $\rho = \gamma \times \gamma^*$. The matrices $F_k$ and $G_k$ are defined as in Definition~\ref{def:poc}.
\end{definition}
Following a similar train of thought as in Section~\ref{sec:complexity} we can derive the computational cost of evaluating a positive vector circuit to be $\bigO (M B N)$ (assuming that the normalization constant has been pre-computed and which is a one-time cost).
The major difference between positive operator circuits and positive vector circuits is that for the former the internal computation units output matrices ($\circuit_k(\xvars_k)\in \mathbb{C}^{B\times B}$) while for the latter vectors are outputted ($\xircuit_k(\xvars_k)\in \mathbb{C}^{B}$).

\begin{restatable}{proposition}{proppqequ}
	\label{prop:pqequ}
	The root nodes of positive operator circuits and positive vector circuits compute the same probability. That is,
	$\circuit_{root}(\xvars)=\xircuit_{root}(\xvars)$ if both circuits use the same $F_k$'s, $G_k$'s, and $\rho$.
\end{restatable}
\begin{proof}
	We show this in Appendix~\ref{sec:proof:pqequ}.
\end{proof}

A similar observation and ensuing proof has been given by \citet{loconte2024subtractive}, although for a more restricted setting. We describe the relationship to their work in more detail in Section~\ref{sec:related}.

\subsection{Parametrizing Positive Operator Circuits}

So far we have only given the computational structure of \pocs and constraints that have to be satisfied for \pocs to form valid probability distributions, \eg positive semidefiniteness. A parametrization can be constructed as follows: at the root unit we need an (unnormalized) density matrix $\rho$, \ie we need to construct a PSD matrix. As already hinted at in Definition~\ref{def:vpox} this is easily achieved by setting
\begin{align}
	\rho = \gamma \times  \gamma^* ,
\end{align}
where $\gamma \in \mathbb{C}^{B\times B}$ is an arbitrary $B\times B$ complex valued matrix.

For the parameter matrices $F_k$ and $G_k$ in the leaf and internal units we can simply pick arbitrary complex valued matrices with adequate dimensions.
For the set of  basis vectors $\{e_{\xvar_n} \}_{n=0}^{N-1}$ in the leaves we simply choose the set of standard basis vectors, \ie one-hot unit vectors, with each basis vector $e_n$ corresponding to a random outcome. Although different choices of basis vectors would be valid as well, \eg the columns of the orthonormalized discrete Fourier transform matrix.

% For the internal computation units of a \punc we need to ensure unitarity of the $U_n$ matrices. We take a rather simple approach and pick the following matrix-matrix product:
% \begin{align}
% 	U_n = \frac{1}{\sqrt{B}} F_B \times D_{n}.
% \end{align}
% Here $F_B$ denotes the $B$-point discrete Fourier transform (DFT) matrix. The factor $\frac{1}{\sqrt{B}}$ orthonormalizes it, thereby giving us a unitary matrix. Note that the DFT matrix has no learnable parameters and all learnable parameters are present in the second matrix of the product. The second matrix is a diagonal matrix having as entries complex numbers of the form $D_{njj}=e^{i \phi_{nj}}$, where $i$ denotes the imaginary unit. The $\phi_{nj}$'s constitute our parameters and off-diagonal entries of $D_n$ are zero.
% Using the fact that $D_n$ is a unitary matrix for any value of the $\phi_{nj}$'s and the fact that the product of two unitary matrices is again unitary,  we do not have to constrain the $\phi_{nj}$'s in any way in order to guarantee unitarity of $U_n$.
% This mean we can optimize the $\phi_{nj}$'s using unconstrained gradient methods.

% What remains is to parametrize the leaf units. For the sake of conciseness, we assume that the categorical random variable $X_n$ present in each of the leaves has a sample space size of $|\Omega(\Xvar_n)| =D$.
% We propose the following functional form for $f_n(\xvar)$:
% \begin{align}
% 	f_n(\xvar) =  A_n \times \xvar
% \end{align}
% We restrict $A_n$ to be a real-valued $B\times D$ matrix. Furthermore, as $\xvar_n$ is categorical we can encode its sample space using the set of standard basis vectors $\{\epsilon_0, \epsilon_1, \dots, \epsilon_{D-1} \}$,\ie one-hot unit vectors, with each basis vector $\epsilon_i$ corresponding to a random outcome. This choice of encoding the sample space immediately gives us:
% \begin{align}
% 	\sum_{\xvar\in \Omega(\Xvar_n)}  \xvar \otimes  \xvar^{*}
% 	=
% 	\sum_{d=0}^{D} \epsilon_d \otimes  \epsilon_d^{T}
% 	=
% 	\mathbb{1}_{D\times D}
% \end{align}
% Consequently, we have:
% $
% 	\sum_{\xvar\in \Xvar_n} f_n(\xvar)\otimes f_n^*(\xvar)
% 	=
% 	A_n \left( \sum_{\xvar\in \Xvar_n} \xvar \otimes \xvar^* \right) A_n^*
% 	=
% 	A_n  \times A_n^*
% $. This means that $A_n$ has to be an \textit{isometric} matrix if we want to satisfy the condition $\sum_{\xvar\in \Xvar_n} f_n(\xvar)\otimes f_n^*(\xvar) =\mathbb{1}_{B\times B}$ from Equation~\ref{eq:def:punc:con}, \ie $A_n A_n^* = \mathbb{1}_{B\times B}$.
% Isometry of $A_n$ can be achieved by assuming the following factorization:
% \begin{align}
% 	A_n = L_n^{-1} B_n, \qquad \text{with  $B_n B_n^* = L_n  L_n^*$ and $L_n$ being a lower triangular matrix.}
% \end{align}
% In other words, we pick $B_n\in \mathbb{R}_{B\times D}$ to be an arbitrary matrix and $L_n \mathbb{R}_{B\times B}$ such that $L_n L_n^*$ is the Cholesky decomposition of $B_n B_n^*$, which is a differentiable operation. With this we now have completely parametrized \puncs using unconstrained parameters and can use off-the-shelf optimizers for learning the parameters.

Even though complex numbers are well-supported in modern deep learning libraries and the gradients can be computed with automated differentiation using Wirtinger calculus~\cite{kreutz2009complex,wirtinger1927formalen}, the issue of numerical stability with complex numbers has to be handled more carefully.
Similar to computations with classic probabilities, computations with positive operator and vector circuits have to be performed in log-space.

A naive implementation of complex numbers would, similar to probabilities, quickly result in the absolute value $r$ of a complex number $r e^{i \phi}$ either under- or overflowing when performing repeated multiplications.
Therefore, we represent complex-valued parameters using the polar form $r e^{i \phi}$ in the log domain using a tuple of real parameters: $(\log{r},  \phi)$.
In this representation $\phi$ and $\log r$ are both unconstrained and real-valued.
Multiplications of two number can now be computed in a straightforward fashion by adding up the elements of the tuples:
$(\log{r_1},  \phi_1) \times (\log{r_2},  \phi_2) = (\log{r_1}+\log{r_2},  \phi_1+ \phi_2) $.

In order to perform complex-valued matrix-matrix and matrix-vector multiplications in the log-domain, we adapt the \textit{LogEinsumExp-trick} (a generalization of the LogSumExp trick) introduced by \citet{peharz2020einsum} for positive real-valued PCs to the setting of complex numbers. This generalizes log-domain circuit evaluations with only positive reals, as well as circuit evaluations that allow for negative reals~\citep{correia2019towards,maua2018robustifying}.
The main idea is to perform the LogSumExp trick on the magnitude of the complex number (in log-space) $\log r$ only, and leave the phase $\phi$ alone.
We refer the reader to our implementation of \pocs/\pvcs for further details.

% At this point we would like to note that computations in probabilsitic unitary circuits have to be, for reasons of numerical stability, performed in log-space. This is similar to the use of the log-probability semiring when dealing with standard probabilities.
% To this end we we represent complex-valued parameters using the polar form in the log domain: $\log z=\log{ r e^{i\phi} } = \log {r }+ i\phi$. In this representation $\phi$ can be chosen freely between $0$ and $2\pi$ and $\log r$ is unconstrained and real-valued. Similar to probabilities, the absolute value $r$ can become small fast when multiplying complex numbers. Using the log-domain prevents numerical instabilities. In order to perform complex-valued matrix-matrix and matrix-vector multiplications in the log-domain, we adapt the \textit{logeinsumexp-trick} introduced by \citet{peharz2020einsum} for positive real-valued PCs to the setting of complex numbers. This generalizes log-domain circuit evaluations with only positive reals, as well as circuit evaluations that allow for negative reals~\citep{correia2019towards,maua2018robustifying}.

% So far we have only given the computational structure of \puncs and constraints that have to be satisfied for \puncs to form valid probability distributions, \eg unitarity of th $U_n$ matrices. Starting at the root unit we need to find a parametrization of the density matrix, \ie we need to construct a PSD matrix with unit trace. This can easily be achieved as follows:
% \begin{align}
% 	\rho =  \nicefrac{\sigma^* \sigma }{\Tr (\sigma^*  \sigma   )},
% \end{align}
% where $\sigma$ is an arbitrary $B\times B$ complex valued matrix.

% For the internal computation units of a \punc we need to ensure unitarity of the $U_n$ matrices. We take a rather simple approach and pick the following matrix-matrix product:
% \begin{align}
% 	U_n = \frac{1}{\sqrt{B}} F_B \times D_{n}.
% \end{align}
% Here $F_B$ denotes the $B$-point discrete Fourier transform (DFT) matrix. The factor $\frac{1}{\sqrt{B}}$ orthonormalizes it, thereby giving us a unitary matrix. Note that the DFT matrix has no learnable parameters and all learnable parameters are present in the second matrix of the product. The second matrix is a diagonal matrix having as entries complex numbers of the form $D_{njj}=e^{i \phi_{nj}}$, where $i$ denotes the imaginary unit. The $\phi_{nj}$'s constitute our parameters and off-diagonal entries of $D_n$ are zero.
% Using the fact that $D_n$ is a unitary matrix for any value of the $\phi_{nj}$'s and the fact that the product of two unitary matrices is again unitary,  we do not have to constrain the $\phi_{nj}$'s in any way in order to guarantee unitarity of $U_n$.
% This mean we can optimize the $\phi_{nj}$'s using unconstrained gradient methods.

% What remains is to parametrize the leaf units. For the sake of conciseness, we assume that the categorical random variable $X_n$ present in each of the leaves has a sample space size of $|\Omega(\Xvar_n)| =D$.
% We propose the following functional form for $f_n(\xvar)$:
% \begin{align}
% 	f_n(\xvar) =  A_n \times \xvar
% \end{align}
% We restrict $A_n$ to be a real-valued $B\times D$ matrix. Furthermore, as $\xvar_n$ is categorical we can encode its sample space using the set of standard basis vectors $\{\epsilon_0, \epsilon_1, \dots, \epsilon_{D-1} \}$,\ie one-hot unit vectors, with each basis vector $\epsilon_i$ corresponding to a random outcome. This choice of encoding the sample space immediately gives us:
% \begin{align}
% 	\sum_{\xvar\in \Omega(\Xvar_n)}  \xvar \otimes  \xvar^{*}
% 	=
% 	\sum_{d=0}^{D} \epsilon_d \otimes  \epsilon_d^{T}
% 	=
% 	\mathbb{1}_{D\times D}
% \end{align}
% Consequently, we have:
% $
% 	\sum_{\xvar\in \Xvar_n} f_n(\xvar)\otimes f_n^*(\xvar)
% 	=
% 	A_n \left( \sum_{\xvar\in \Xvar_n} \xvar \otimes \xvar^* \right) A_n^*
% 	=
% 	A_n  \times A_n^*
% $. This means that $A_n$ has to be an \textit{isometric} matrix if we want to satisfy the condition $\sum_{\xvar\in \Xvar_n} f_n(\xvar)\otimes f_n^*(\xvar) =\mathbb{1}_{B\times B}$ from Equation~\ref{eq:def:punc:con}, \ie $A_n A_n^* = \mathbb{1}_{B\times B}$.
% Isometry of $A_n$ can be achieved by assuming the following factorization:
% \begin{align}
% 	A_n = L_n^{-1} B_n, \qquad \text{with  $B_n B_n^* = L_n  L_n^*$ and $L_n$ being a lower triangular matrix.}
% \end{align}
% In other words, we pick $B_n\in \mathbb{R}_{B\times D}$ to be an arbitrary matrix and $L_n \mathbb{R}_{B\times B}$ such that $L_n L_n^*$ is the Cholesky decomposition of $B_n B_n^*$, which is a differentiable operation. With this we now have completely parametrized \puncs using unconstrained parameters and can use off-the-shelf optimizers for learning the parameters.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Related Work}
\label{sec:related}

This work is inspired by theoretical observations made in the statistical relational AI literature. Specifically, \citet{buchman2017rules}, and \citet{kuzelka2020complex} noted that using only real-valued parametrizations (including negatives~\citep{buchman2017negative}) does not allow for fully expressive models.
% Interestingly, the approach of \citet{kuzelka2020complex} for parametrizing so-called Markov logic networks~\citep{richardson2006markov} relies, similar to our parametrization of \puncs, on the discrete Fourier transform.
In contrast to our work, both of these works are not concerned with learning and are of a rather theoretical nature.
We discuss next three approaches from the physics and the machine learning literature that have taken a more practical approach.

\subsection*{Tensor Networks}

A popular technique to model systems in condensed matter physics are so-called tensor networks~\citep{orus2014practical,white1992density}, and in recent years they have been applied to tackle problems in supervised as well as unsupervised machine learning~\citep{cheng2019tree,han2018unsupervised,stoudenmire2016supervised} -- with the works of \citet{han2018unsupervised} and \citet{cheng2019tree} on tree tensor networks being most related to \pocs. Similar to \pocs, tree tensor networks perform inference in a hierarchical fashion. The conceptual difference, however, lies with the formulation of \pocs in the POVM formalism. In this regard and given that tensor networks originate in the physics community, it is rather surprising that tensor networks have so far, and to the best of our knowledge, not been formulated using POVMs.

Using our generalized POVM formulation for \pocs is, however, not only theoretically elegant but has also practical benefits: using this formalism leads to circuits that are normalized by construction (using a normalization constant), and we can perform learning simply by optimizing the likelihood. We contrast this to the sweeping algorithms that are usually deployed in the tensor network literature, where blocks of variables are optimized one at the time while the remaining variables are held constant. It appears that this approach is inspired by the variational ansatz taken in the density-matrix renormalization group algorithm~\citep{white1992density}, which is the original algorithm for tensor networks.

We would also like to point out a theoretical result from the tensor network literature stating that picking a complex-valued parametrization instead of a real-valued one can lead to an arbitrarily large reduction in the number of parameters~\citep{glasser2019expressive}. While this result is formulated with respect to (exact) low rank tensor decomposition and does not apply directly to the problem of learning parameters via gradient descent, we consider this observation to be a strong theoretical indicator for the superiority of complex numbers over real numbers when parametrizing probabilistic circuits. A similar argument has also been made by \citet{gao2022enhancing} in the context of hidden Markov models.

\subsection*{Squared Non-Monotonic Probabilistic Circuits}
\citet{loconte2024subtractive} recently introduced a class of tractable probabilistic models called \textit{squared non-monotonic PCs} (\snpcs). The main idea behind \snpcs is that one can parametrize a probabilistic circuit using negative and positive reals and simply square the final output.
They then use the circuit squaring algorithm from \citep{vergari2021compositional} to guarantee tractable marginalization and compute the normalization constant in time polynomial in the size of the circuit.
Squaring the output of a circuit in order to guarantee positivity and normalizing via a normalization constant is reminiscent of the positive vector circuits from Section~\ref{sec:vpoc}.

\begin{restatable}{proposition}{proppocrankone}
	\label{prop:pocrankone}
	\snpcs are real-valued positive vector circuits with a density matrix of rank-one.
\end{restatable}

\begin{proof}
	We show this in Appendix~\ref{sec:proof:prop:pocrankone}.
\end{proof}

Concerning the parametrization, the biggest difference is our use of the entire complex plane compared to real valued parameters only for \snpcs. While we do not have a theoretical proof that this complex parametrization results in improved expressive power, results from the tensor network literature~\citep{glasser2019expressive} and the statistical relation AI~\citep{buchman2017rules,kuzelka2020complex} literature hint in this direction.

Using Proposition~\ref{prop:pocrankone} and the fact that \snpcs only use real-valued parameters we can conclude that positive vector circuits generalize \snpcs two-fold. First, \pvcs use full-rank matrices as their (unnormalized) density matrix. Second, \pvcs utilize the entire complex plane. In our experiments we show that these generalizations do have an impact.
Interestingly the restriction to rank-one density matrices is also customary in the tensor network literature~\citep{cheng2019tree,han2018unsupervised}.

\subsection*{Positive Semidefinite Kernels}

Recently, the use of PSD matrices has also been studied in the kernel literature~\citep{marteau2020non,rudi2021psd}. Specifically, given a set of $N$ data samples $\xvars^{(0)}, \dots, \xvars^{(N-1)}$. An unnormalized probability distribution $q(\xvars)$ can be defined as follows:
$
	% \begin{align}
	q(\xvars) = \boldsymbol{\kappa}^{T}(\xvars)  A \boldsymbol{\kappa}(\xvars),
	\label{eq:psdk}
	% \end{align}
$
where $A  \in \mathbb{R}^{N \times N}$ is a PSD matrix and where
$\boldsymbol{\kappa}(\xvars) \in  \mathbb{R}^{N}$ is a vectorized kernel: $\boldsymbol{\kappa}_i(\xvars)= \kappa(\xvars, \xvars^{i} )$ and $\kappa$ being the base kernel, \eg radial basis function.
Comparing this to the root computation unit of a \pvc, we can  observe a superficial resemblance: in both cases we have a quadratic form. Important differences exist, however. Contrary to \pvcs the size of the PSD kernels grows with the number of data points. Furthermore, the distribution induced by PSD kernels is unnormalized.

\section{Experimental Evaluation}
\label{sec:experiments}

\textbf{Experimental Setup}

For our experimental evaluation we used the MNIST family of datasets. That is, the original MNIST~\citep{deng2012mnist}, FashionMNIST~\citep{xiao2017fashion}, and also EMNIST~\citep{cohen2017emnist}.
For the implementation we used PyTorch together with the Einops library~\citep{rogozhnikov2022einops}. We set up our experiments in Lightning\footnote{\url{https://lightning.ai/}} and ran them on a DGX-2 machine with V100 Nvidia cards.

We compare different methods using the \textit{bits per dimension} metric, which is calculated from the average negative log-likelihood ($\overline{NLL}$) as follows: $bpd {=} \nicefrac{\overline{NLL}}{(\log{2} \times D)}$ ($D{=}28^2$ for MNIST datasets).

\textbf{Training Positive Operator Circuits}

In our experiments we estimate the density of the different MNIST datasets by maximizing a parametrized likelihood function. We construct this  function using a positive vector circuit.
In order to build the underlying partition circuit,
we follow the approach described by \citet{zuidberg2024probabilistic}: we start with a grid of $28\times 28$ pixels and merge, in an alternating fashion, rows and columns of the image.
We also allow for merging three partitions into  a single partition, thereby breaking the binary character of the partition trees used so far.
Having three instead of two partitions being merged can easily be accommodated for by using a Hadamard product with three factors instead of two in the internal units of a \pvc. Ultimately, this allows us to handle layers with an uneven number of partitions.
At the leaves we encode pixel values by associating each value to one of $256$ standard basis vector of $\mathbb{R}^{256}$.

We performed training by minimizing the negative log-likelihood using Adam~\citep{kingma2014adam} with default hyperparameters, batch size of $50$, over $100$ epochs.
The best model was selected using a $90-10$ train-validation data split where we monitored the negative log-likelihood on the validation set.
We did not perform any hyperparameter search.
Further details can be found in the configuration files of the experiments.

\textbf{State-of-the-Art Baselines}

We compare \pvcs to three state-of-the-art tractable density estimators from the literature: quadrature of probabilistic circuits (QPCs)~\citep{gala2024probabilistic}, hidden Chow-Liu trees (HCLTs)~\citep{liu2021tractable}, and sparse HCLTs (SHCLTs)~\citep{dang2022sparse}.
All three methods are probabilistic circuits and they all use hidden Chow-Liu trees as their underlying structure. This tree is learned and dataset specific. The difference between QPCs and HCLTs is that the former is obtained by approximating a continuous latent variable extension of probabilistic circuits using numerical quadrature. The difference between HCLTs and SHCLTs is that the latter allows for dynamically adapting the number of components per partition during parameter learning. This means that SHCLTs can focus their computational budget on information-rich parts of the circuit.

\textbf{Q1: How Do the Different \pvc Parametrizations Fair Against Each Other?}

We construct \pvcs following the prescription of \cite{zuidberg2024probabilistic}, as described above. Furthermore, we use $B=128$ components per partition.
As for the parametrization we study four different variants:
$\mathrm{PVX}_{0}^{FR}$,
$\mathrm{PVX}_{2\pi}^{FR}$,
$\mathrm{PVX}_{0}^{R1}$, and
$\mathrm{PVX}_{2\pi}^{R1}$.
The subscripts ($0$ or $2\pi$) indicate whether we limit the phase to be zero or whether we allow the phase to be a learnable parameter. Note that having a tunable phase doubles the number of (real-valued) parameters. The superscript indicates whether the parametrization uses a full-rank ($FR$) density matrix in the root or a rank-one ($R1$) density matrix.
As such, $\mathrm{PVX}_{0}^{R1}$ is equivalent to the  squared monotonic probabilistic circuits (\smpcs) used in ~\cite{loconte2024subtractive}, and $\mathrm{PVX}_{2\pi}^{R1}$ generalizes their \snpcs from the real domain to the entire complex domain. Concretely, \smpcs are rank-one \pvcs with the phase fixed to zero and \snpcs are rank-one \pvcs with the phase fixed to values in the set $\{0, \pi \}$. We do not experiment on the latter.

We report the obtained bpds for the different \pvc parametrizations in the first four columns of Table~\ref{tab:mnist}. In general, we see that the complex-valued parametrizations are either on par or outperform the corresponding zero-phase parametrizations.
The notable exception is the FashionMNIST benchmark where the two parametrizations with no phase ($\mathrm{PVX}_{0}^{FR}$ and $\mathrm{PVX}_{0}^{R1}$) perform best.
Comparing full-rank to rank-one parametrizations we see a more important impact for the zero-phase parametrization than for the complex-valued parametrization.

Overall, $\mathrm{PVX}_{2\pi}^{FR}$ constitutes the best performing parametrization being best-in-class on all benchmarks but FashionMNIST.

\input{tex_input/mnist_table}

\textbf{Q2: How Do \pvcs Fair Against the State of the Art?}

In order to compare \pvcs to state-of-the-art circuits all methods (\pvc, QPC, HCLT, SHCLT) were given a computational budget of $B=128$ components per partition (on average for SHCLT).
The results for the competing methods were taken from the respective papers -- except for HCLTs. For HCLTs we took the bpds reported by \citet{gala2024probabilistic} as they achieved stronger results with their implementation compared to the originally reported performance~\citep{liu2021tractable}.

We see in Table~\ref{tab:mnist} that QPCs and HCLTs are in general outperformed by \pvc -- in particular $\mathrm{PVX}_{2\pi}^{FR}$. The FashionMNIST benchmark constitutes again an outlier in this regard.
The only methods outperforming \pvcs are the SHCLTs, which is due to them being able to dynamically allocate compute budget to informative parts of the circuit.
It is noteworthy that for the EMNIST-BYCLASS benchmark all four  \pvc parametrization exhibit strong performance with  $\mathrm{PVX}_{2\pi}^{FR}$ and $\mathrm{PVX}_{2\pi}^{R1}$ even outcompeting SHCLTs.

\textbf{Discussion}

In our experimental evaluation we did not perform an explicit comparison to the method of \citet{loconte2024subtractive} as they were not able to find any improvements of allowing non-negative parameters in probabilistic circuit for discrete data. They stipulated that
\say{simple categorical distribution can already capture any discrete distribution with finite support and a (subtractive) mixture thereof might not yield additional benefits}.
In our experimental evaluation we refute this conjecture, and show that using complex-valued parameters leads to noticeable gains when performing density estimation on discrete data.
The exception being of course the FashionMNIST benchmark. We believe that this is due to our optimization method that was not tuned in any way.
In this regard we believe that developing tailor-made optimization algorithms for \pvcs might further boost their performance.
Furthermore, one can envisage, similar to SHCLTs, dynamically adapting the compute budget per partition, which should again improve the density estimation capacities of \pvcs.

\section{Conclusions \& Future Work}
\label{sec:conclusions}

Based on first principles from quantum information theory, we constructed positive operator circuits -- a novel class of probabilistic tractable models:
Their construction as partition circuits allows for layer-wise parallelization and execution on modern machine learning hardware.
Using unconstrained gradient-descent we showed that \pocs, and \pvcs specifically, constitute a promising addition to the zoo of tractable probabilistic models.

In future work we would like to investigate in more detail theoretical properties of \pocs and how they differ from other tractable models. Ideally one would find an exponential separation between real-valued and complex-valued circuits. On the practical side it remains, however, to be seen whether such a separation has an equally important impact on real-world datasets.

This relates to another open question,  that of learning in \pocs. We presented a rather simple learning approach for parameters tailored towards neural architecture. It might be the case that more sophisticated techniques have to be deployed for large-scale \pocs, as they might exhibit the problem of barren plateaus~\citep{ragone2023unified} -- a well known issue in quantum machine learning~\citep{biamonte2017quantum,huggins2019towards}.
Furthermore, and as already pointed out by~\citet{loconte2024subtractive}, computing the normalization constant $Z$ requires the evaluation of the circuit in the \poc representation. This gives rise to a rather expensive cubic computation cost during learning (when compared to the quadratic cost of \pvc evaluations). Avoiding this issue would allow to drastically scale \pvcs. In order, to scale \pvcs it might also be necessary to use more sophisticated structures than binary trees with every partition having the same number of components $B$ per partition, \cf SHCLTs.

\clearpage

\bibliography{references}
\bibliographystyle{iclr2025_conference}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

\setcounter{equation}{0}
\renewcommand{\theequation}{A\arabic{equation}}

\appendix

\section{Probabilistic Circuits as Partition Circuits}
\label{sec:lpc}

\begin{definition}[Layered Probabilistic Circuit]
	\label{def:lpc}
	Let $\Xvars=\{\Xvar_0,\dots ,\Xvar_{M-1}  \}$ (taking values  $\xvars=\{\xvar_0,\dots ,\xvar_{M-1}  \}$) be a set of categorical random variables. Furthermore, let $B$ be a positive integer.
	We define a layered probabilistic circuit as a partition circuit whose computation units take the following functional form:
	\begin{align}
		{\circuit}_k(\xvars_k)=
		\begin{cases}
			f_k(\xvar_k),
			\quad f_k(\xvar_k) \in \mathbb{R}_{\geq 0}^B
			 & \text{if $n$ leaf unit, \ie $\xvars_k = \{\xvar_k \}$}
			\\
			W_k \times \Bigl( \circuit_{n_l}(\xvars_{n_l}) \odot  \circuit_{n_r} (\xvars_{n_r}) \Bigr),
			\quad W_k \in \mathbb{R}_{\geq 0}^{B \times B}
			 & \text{if $n$ internal unit}
			\\
			\rho \cdot \Bigl( \circuit_{n_l}(\xvars_{n_l}) \odot  \circuit_{n_r} (\xvars_{n_r}) \Bigr) \big/ Z,
			\quad \rho \in \mathbb{R}_{\geq 0}^{B}
			 & \text{if $n$ root unit.}
		\end{cases}
	\end{align}
	Here we use the symbols $\times$, $\odot$ and $\cdot$ to denote the matrix product, Hadamard product and dot product, respectively.
	Additionally, we necessitate:
	\begin{align}
		Z
		=
		\sum_{\xvars \in \Omega(\Xvars)}
		\circuit_{root}(\xvars)
	\end{align}
	% Additionally, we necessitate the following normalization constraints:
	% \begin{align}
	% 	\forall \Xvar_k: \sum_{\xvar\in\Omega(\Xvar_k)} f_k(\xvar)=\mathbf{1}_B,
	% 	\qquad
	% 	\forall n, i : \sum_{j=0}^{B-1} W_{kij}=1,
	% 	\qquad
	% 	\sum_{j=0}^{D-1} \rho_{j}=1,
	% \end{align}
	where $\Omega(\Xvars)$ denotes the sample space of $\Xvars$.
	% where $\mathbf{1}_B$ denotes the B-dimensional vector having as entries only ones,
	% and where the subscripts $i$ and $j$ index the elements of the matrices $W_k$ and the vector $\rho$.
\end{definition}

\begin{restatable}{theorem}{theolpcvalid}
	\label{theo:lpcvalid}
	Layered PCs are valid probability distributions.
\end{restatable}

\begin{proof}
	If $\circuit_{root} (\xvars)$ is the computation unit at the root of the layered PC, the unit forms a probability distribution if $\circuit(x)\geq 0$, for every $\xvars \in \Omega(\Xvars)$ and if $\sum_{\xvars \in \Omega(\Xvars)} \circuit(\xvars)=1$. The first condition is trivially satisfied as the circuit only performs linear operations on matrices and vectors ($f_k(\xvar_k), W_k, \rho$) with positive entries only.
	The second requirement is satisfied by construction using the normalization constant Z.
	% For the second condition we use the fact that the root unit and the internal unit are multilinear functions, which allows us to push the summation down to the leaf units. We then use that $\sum_{\xvar\in\Omega(\Xvar_n)} f_n(\xvar)=\mathbf{1}_B$. Evaluating the circuit with these $\mathbf{1}_B$ vectors and using the normalization constraints on $W_n$ and $\rho$ trivially returns $1$. Hence, both conditions are satisfied and layered PCs form valid probability distributions.
\end{proof}

\section{Proof of Theorems and Propositions}

\subsection{Proof of Proposition~\ref{prop:povmprob}}
\label{sec:proof:prop:povmprob}

\proppovmprob*
\begin{proof}
	First we show that $p(i)\geq 1$, for each $i$:
	\begin{align}
		p(i)
		%  &
		=
		\Tr [ E(i) \rho ]
		% \nonumber
		% \\
		%  &
		=
		\Tr [ F(i) F^*(i)  \rho ]
		% \nonumber
		% \\
		%  &
		=
		\Tr [ F(i)^*  \rho F(i) ].
	\end{align}
	Here we used the fact that $E(i)$ is PSD and factorized it into the product $F(i) F^*(i)$. Then we used the fact that the trace is invariant under cyclical shifts. Next we write out the trace using matrix elements:
	\begin{align}
		\Tr [ F(i)^*  \rho F(i)]
		%  &
		=
		\sum_{klmn}  \delta_{kn} F^*_{lk}(i)\rho_{lm} F_{mn}(i)
		% \nonumber
		% \\
		%  &
		=
		\sum_{k}   F^*_{k}(i)\rho F_{k}(i).
		\label{eq:proof:trpsdprod}
	\end{align}
	In the equation above $F_{k}(i)$ is the $k$-th column of $F(i)$. As $\rho$ is PSD we have that each term of the sum over $k$ is positive, which means in turn that $p(i)$ is a positive real number, \ie $p(i){\geq} 0$ for every $i$.

	Secondly, we show that $p(i)$ is normalized:
	\begin{align}
		\sum_{i=1}^N p(i)
		%  &
		=
		\sum_{i=1}^N \Tr [ E(i) \rho ]
		% \nonumber
		% \\
		%  &
		=
		\Tr [ \sum_{i=1}^N E(i) \rho ]
		% \nonumber
		% \\
		%  &
		=
		\Tr [  \rho ],
	\end{align}
	where we used Equation~\ref{eq:povm_normalized}.
	Exploiting the fact that the trace of a density matrix is one gives us indeed $\sum_{i=1}^N p(i)=1$, and we can conclude that $p(i)$ is a valid probability distribution.
\end{proof}

\subsection{Proof of Theorem~\ref{theo:pocvalid}}
\label{sec:proof:pocvalid}

\theopocvalid*

\begin{proof}
	We need to prove that at the root we have $\circuit_{root}(\xvars)\geq 0$, for every $\xvars \in \Omega(\Xvars_{root})$ and that $\sum_{\xvars \in \Omega(\Xvars_{root})} \circuit_{root}(\xvars)=1$.
	The latter requirement is satisfied by construction via the normalization constant $Z$. For the former, let us first define:
	\begin{align}
		\mu_{root}(\xvars) \coloneqq \circuit_{root_l}(\xvars_{l}) \odot  \circuit_{root_r}(\xvars_{r}).
	\end{align}
	By identifying each element $\xvars \in \Omega(\Xvars)$ with an event $i$ from Definition~\ref{def:Zeventprob}, consequently $\mu_{root}(\xvars)$ with $E(i)$, we observe that the root of a \poc has the functional form of the probability of an element from a (generalized) POVM.
	In order to prove that $\forall \xvars \in \Omega(\Xvars): \Trace[\mu_{root}(\xvars) \rho]{>}0$,
	we now only need to show that $\mu_{root}(\xvars)$ is PSD,
	as $\rho$ is already PSD by assumption.

	To this end, we first observe that in the leaf units of a \poc we have the cross product between a vector and its conjugate transpose, which produces a PSD matrix. These PSD matrices from the leaves are then fed into the first layer of internal units where we first perform Hadamard products, which is again a PSD matrix via Schur's product theorem~\citep[p. 14, Theorem VII]{schur1911bemerkungen}. For units $k$ in the first internal layer we then have:
	\begin{align}
		\circuit_k(\xvars_k)
		=
		G_k \times P_k(\xvars_k) \times G_{k}^*,
	\end{align}
	where $P_k(\xvars_k)$ is the matrix obtained by performing the Hadamard product in the leaf. Using the fact that $P_k(\xvars)$ is PSD allows for a square root decomposition, and we can write:
	\begin{align}
		\circuit_k(\xvars_h) = G_k P^{\nicefrac{1}{2}}_k(\xvars_n) P^{\nicefrac{1}{2}}_k(\xvars_k) G_k^*
		=
		\left( G_k P^{\nicefrac{1}{2}}_k(\xvars_k) \right)
		\left( G_k P^{\nicefrac{1}{2}}_k(\xvars_k) \right)^*,
	\end{align}
	which is clearly PSD. By repeating this argument recursively for all the layers up to the root we can conclude that $\circuit_{root}(\xvars)$ is PSD for all $\xvars \in \Omega(\Xvars_{root})$, which also concludes the proof.
\end{proof}

% \begin{proof}
% 	We need to prove that at the root we have $\circuit_{root}(\xvars)\geq 0$, for every $\xvars \in \Omega(\Xvars)$ and that $\sum_{\xvars \in \Omega(\Xvars_{root})} \circuit_{root}(\xvars)=1$.
% 	To this end, let us first define:
% 	\begin{align}
% 		\delta_{root}(\xvars) \coloneqq \dual_{root_l}(\xvars_{l}) \odot  \dual_{root_r}(\xvars_{r}).
% 	\end{align}
% 	By identifying $\delta_{root}(\xvars)$ with $E(i)$ from Definition~\ref{def:eventprob} we see that the root of a \punc has the functional form of the probability of an element from a POVM.
% 	As $rho$ in Definition~\ref{def:punc} is a density matrix by definition, we only need to show that the following two conditions hold:
% 	\begin{align}
% 		\forall \xvars \in \Omega(\Xvars_{root}): \delta_{root}(\xvars) & \text{ is PSD},
% 		\label{eq:con_psd:proof}
% 		\\
% 		\sum_{\xvars \in \Omega(\Xvars_{root})} \delta_{root}(\xvars)   & =\mathbb{1}_{B\times B}.
% 		\label{eq:con_one:proof}
% 	\end{align}
% 	For the condition in Equation~\ref{eq:con_psd} we first observe that in the leaf units we have the cross product between a vector and its conjugate transpose, which produces a PSD matrix. These PSD matrices from the leaves are then fed into the first layer of internal units where we first perform Hadamard products, which is again a PSD matrix via Schur's product theorem~\citep[p. 14, Theorem VII]{schur1911bemerkungen}. For units $n$ in the first internal layer we then have:
% 	\begin{align}
% 		\dual_n(\xvars_n) = U_n P_n(\xvars_n) U_n^*
% 	\end{align}
% 	where $P_n(\xvars_n)$ is the matrix obtained performing the Hadamard product. Using the fact that $P_n(\xvars)$ is PSD allows for a square root decomposition, and we can write:
% 	\begin{align}
% 		\dual_n(\xvars_n) = U_n P^{\nicefrac{1}{2}}_n(\xvars_n) P^{\nicefrac{1}{2}}_n(\xvars_n) U_n^*
% 		=
% 		\left( U_n P^{\nicefrac{1}{2}}_n(\xvars_n) \right)
% 		\left( U_n P^{\nicefrac{1}{2}}_n(\xvars_n) \right)^*,
% 	\end{align}
% 	which is clearly PSD. By repeating this argument recursively for all the layers up to the root we can conclude that $\delta_{root}(\xvars)$ is PSD for all $\xvars \in \Omega(\Xvars_{root})$.
% 	Next, for the condition in Equation~\ref{eq:con_one} we note again that we can push down the summation over the variables $\xvars \in \Omega(\Xvars)$ down to the individual leaf units where we have by definition $\sum_{\xvar \in \Omega(\Xvar_n)} f_n(\xvar) \otimes f_n^*(\xvar)=\mathbb{1}_{B\times B}$. For the internal nodes this then recursively gives us:
% 	\begin{align}
% 		\sum_{\xvars_n \in \Omega(\Xvars_n)} \dual_n(\xvars_n)
% 		 & =
% 		U_n
% 		\times
% 		\left(
% 		\sum_{\xvars_{n_l} \in \Omega(\Xvars_{n_l})} \dual_{n_l}(\xvars_{n_l})
% 		\odot
% 		\sum_{\xvars_{n_r}\in \Omega(\Xvars_{n_r})}  \dual_{n_r} (\xvars_{n_r})
% 		\right)
% 		\times
% 		U_{n}^*
% 		\\
% 		 &
% 		=
% 		U_n \times \left(\mathbb{1}_{B\times B} \odot \mathbb{1}_{B\times B}  \right) \times U_{n}^*=\mathbb{1}_{B\times B},
% 	\end{align}
% 	Consequently, we also have $\sum_{\xvars \in \Omega(\Xvars_{root})} \delta_{root}(\xvars){=}\mathbb{1}_{B\times B}$. With Equations~\ref{eq:con_psd} and~\ref{eq:con_one} satisfied we can conclude that probabilistic unitary circuits are indeed valid probability distributions.

% \end{proof}

\subsection{Proof of Proposition~\ref{prop:pqequ}}
\label{sec:proof:pqequ}

\proppqequ*

\begin{proof}
	We start the proof by trivially rewriting the computation units in the leaves of a \poc  (Equation~\ref{eq:poc:def}):
	\begin{align}
		\circuit_k(\xvar_k)
		=
		\left(F_{k} \times  e_{\xvar_k} \right)
		\otimes
		\left(   F_{k}  \times  e_{\xvar_k}\right)^*
		=
		\xircuit_k (\xvar_k) \otimes \xircuit_k^* (\xvar_k),
		\label{eq:leaf:vec}
	\end{align}
	This means that we can construct the matrix $\circuit_k(\xvars_k)$ from the vector $\xircuit_k(\xvars_k)$, and we need only to pass on the vector to the next layer. Coincidentally, $\xircuit_k(\xvars_k)$ is the exact computation performed by the \pvc at unit $k$. This also means that we only need to perform the computation for $\xircuit_k(\xvars_k)$ and not its conjugate transpose.

	In the next layer, we plug in the right-most side of Equation~\ref{eq:leaf:vec} into the expression of an internal computational unit of a \poc and we get:
	\begin{align}
		\circuit_k(\xvars_k)
		=
		G_k
		\times
		\Bigl[
			\Bigl(
			\xircuit_{k_l}(\xvars_{k_l})
			\otimes
			\xircuit^*_{k_l}(\xvars_{k_l})
			\Bigr)
			\odot
			\Bigl(
			\xircuit_{k_r}(\xvars_{k_r})
			\otimes
			\xircuit^*_{k_r}(\xvars_{k_r})
			\Bigr)
			\Bigr]
		\times
		G_{k}^*.
	\end{align}
	We now exploit the mixed-product property of the Hadamard product and the Kronecker product to obtain:
	\begin{align}
		\circuit_k(\xvars_k)
		 &
		=
		G_k
		\times
		\Bigl[
			\xircuit_{k_l}(\xvars_{k_l})
			\odot
			\xircuit_{k_r}(\xvars_{k_r})
			\Bigl)
			\otimes
			\Bigl(
			\xircuit^*_{k_l}(\xvars_{k_l})
			\odot
			\xircuit^*_{k_r}(\xvars_{k_r})
			\Bigr)
			\Bigr]
		\times
		G_{k}^*
		\label{eq:proof:pqeq:mixprod}
		\\
		 &
		=
		\Bigl[
			G_k
			\times
			\Bigl(
			\xircuit_{k_l}(\xvars_{k_l})
			\odot
			\xircuit_{k_r}(\xvars_{k_r})
			\Bigr)
			\Bigr]
		\otimes
		\Bigl[
			G_k
			\times
			\Bigl(
			\xircuit_{k_l}(\xvars_{k_l})
			\odot
			\xircuit_{k_r}(\xvars_{k_r})
			\Bigr)
			\Bigr]^*
		\label{eq:proof:pqeq:asso}
		\\
		 &
		=
		\xircuit_k(\xvars_k) \otimes \xircuit^*_k(\xvars_k).
		\label{eq:proof:pqeq:krodeco}
	\end{align}
	Going from Equation~\ref{eq:proof:pqeq:mixprod} to Equation~\ref{eq:proof:pqeq:asso} we made use of associativity.

	We can now recursively plug in in these Kronecker decompositions (\cf Equation~\ref{eq:proof:pqeq:krodeco}) into subsequent computation units and will obtain for each internal layer the one-to-one correspondence of:
	\begin{align}
		\circuit_k(\xvars_k) =\xircuit_k(\xvars_k) \otimes \xircuit^*_k(\xvars_k),
	\end{align}
	which means that we can evaluate the internal units of a \poc by evaluating the corresponding internal units in the \pvc.

	Finally, at the root layer we plug in again the Kronecker decomposition from the last internal layer and obtain:
	\begin{align}
		\circuit_k(\xvars_k)
		 &
		=
		\Trace
		\Bigl[
			\rho
			\times
			\Bigl(
			\xircuit_{k_l}(\xvars_{k_l}) \otimes \xircuit^*_{k_l}(\xvars_{k_l})
			\Bigr)
			\odot
			\Bigl(
			\xircuit_{k_r}(\xvars_{k_r}) \otimes \xircuit^*_{k_r}(\xvars_{k_r})
			\Bigr)
			\Bigr]
		\Big/ Z
		\\
		 &
		=
		\Trace
		\Bigl[
			\rho
			\times
			\Bigl(
			\xircuit_{k_l}(\xvars_{k_l}) \odot \xircuit_{k_r}(\xvars_{k_r})
			\Bigr)
			\otimes
			\Bigl(
			\xircuit^*_{k_l}(\xvars_{k_l}) \odot \xircuit^*_{k_r}(\xvars_{k_r})
			\Bigr)
			\Bigr]
		\Big/ Z
		\\
		%  &
		% =
		% \Trace
		% \Bigl[
		% 	\gamma
		% 	\times
		% 	\Bigl(
		% 	\xircuit_{k_l}(\xvars_{k_l}) \odot \xircuit_{k_r}(\xvars_{k_r})
		% 	\Bigr)
		% 	\otimes
		% 	\Bigl(
		% 	\xircuit^*_{k_l}(\xvars_{k_l}) \odot \xircuit^*_{k_r}(\xvars_{k_r})
		% 	\Bigr)
		% 	\times
		% 	\gamma^*
		% 	\Bigr]
		% \Big/ Z
		% \\
		 &
		=
		\Trace
		\Bigl[
			\gamma^*
			\times
			\gamma
			\times
			\Bigl(
			\xircuit_{k_l}(\xvars_{k_l}) \odot \xircuit_{k_r}(\xvars_{k_r})
			\Bigr)
			\otimes
			\Bigl(
			\xircuit^*_{k_l}(\xvars_{k_l}) \odot \xircuit^*_{k_r}(\xvars_{k_r})
			\Bigr)
			\Bigr]
		\Big/ Z
		\\
		 &
		=
		\Trace
		\Bigl[
			\gamma
			\times
			\Bigl(
			\xircuit_{k_l}(\xvars_{k_l}) \odot \xircuit_{k_r}(\xvars_{k_r})
			\Bigr)
			\otimes
			\Bigl(
			\xircuit^*_{k_l}(\xvars_{k_l}) \odot \xircuit^*_{k_r}(\xvars_{k_r})
			\Bigr)
			\times
			\gamma^*
			\Bigr]
		\Big/ Z.
	\end{align}
	We can write this expression now easily in terms of $\nu_k$
	\begin{align}
		\circuit_k(\xvars_k)
		 &
		=
		\Trace \bigl[ \nu_{k}(\xvars_k) \otimes \nu^*_{k}(\xvars_k) \bigr]
		\\
		 &
		= \nu_{k}(\xvars_k) \cdot \nu^*_{k}(\xvars_k),
	\end{align}
	which proves the equality of a \poc and its corresponding \pvc at the root.
\end{proof}

% \subsection{Proof of Proposition~\ref{prop:puncmargtract}}
% \label{sec:proof:prop:puncmargtract}

% \proppuncmargtract*
% \begin{proof}
% 	The proof is trivial, as marginalizing out a variable $\Xvar_n$ from a \punc simply amounts to pushing down the summation to the corresponding leaf and replacing $\sum_{\xvar\in \Xvar_n} f_n(\xvar)\otimes f_n^*(\xvar)$ with $\mathbb{1}_{B\times B}$.
% \end{proof}

\subsection{Proof of Proposition~\ref{prop:pocrankone}}
\label{sec:proof:prop:pocrankone}

\proppocrankone*

\begin{proof}
	To  see this consider the computation at the root of a \pvc:
	\begin{align}
		\nu_{root}(\xvars)
		=
		\gamma
		\times
		\Bigl(
		\xircuit_{root_l}(\xvars_{l})
		\odot
		\xircuit_{root_r}(\xvars_{r})
		\Bigr)
		\big/ \sqrt{Z}
	\end{align}
	Let us first define $\eta_{root}(\xvars) \coloneqq 			\Bigl(
		\xircuit_{root_l}(\xvars_{l})
		\odot
		\xircuit_{root_r}(\xvars_{r})
		\Bigr)
		\big/ \sqrt{Z} $.
	This then gives us:
	\begin{align}
		\nu_{root}(\xvars)
		=
		\gamma
		\times
		\eta_{root}(\xvars)
	\end{align}
	Here $\gamma$ is an arbitrarily complex-valued matrix -- potentially full-rank.
	If we, however, restrict $\gamma$ to be a rank-one matrix, \ie $\gamma = \delta \otimes \delta^*$, with $\delta$ being a complex-valued vector, we get:
	\begin{align}
		\nu_{root}(\xvars)
		=
		(
		\delta
		\otimes
		\delta^*
		)
		\times
		\eta_{root}(\xvars).
	\end{align}
	Plugging this into the expression for $\xircuit_{root}$ (\cf Equation~\ref{eq:pvc:def}) we obtain:
	\begin{align}
		\xircuit_{root}(\xvars)
		=
		\Bigl[
			(
			\delta
			\otimes
			\delta^*
			)
			\times
			\eta_{root}(\xvars)
			\Bigr]
		\cdot
		\Bigl[
			(
			\delta
			\otimes
			\delta^*
			)
			\times
			\eta_{root}(\xvars)
			\Bigr]^*
	\end{align}
	Slightly rearranging the factors gives us:
	\begin{align}
		\xircuit_{root}(\xvars)
		=
		\Bigl[
			\delta
			\times
			\underbrace{
				(
				\delta^*
				\cdot
				\eta_{root}(\xvars)
				)}_{\eqqcolon \epsilon_{root}(\xvars)}
			\Bigr]
		\cdot
		\Bigl[
			\delta
			\times
			(
			\delta^*
			\cdot
			\eta_{root}(\xvars)
			)
			\Bigr]^*
	\end{align}
	As $\delta^* \cdot \eta_{root}(\xvars)$ is a  simple dot product between two vectors  $\epsilon_{root}(\xvars)$ must be a scalar. This lets us further rearrange factors to result in:
	\begin{align}
		\xircuit_{root}(\xvars)
		=
		\epsilon_{root}(\xvars) \epsilon^*_{root}(\xvars) ( \delta \cdot \delta^*)
	\end{align}
	Here, the dot product $\delta \cdot \delta^*$ between two vectors results again  in a scalar, which is also positive.
	We absorb this scalar for simplicity into $\epsilon_{root}(\xvars)$ with $\widehat{\epsilon}_{root}(\xvars) = \epsilon_{root}(\xvars) \sqrt{ \delta \cdot \delta^*}$.
	Lastly, we can compute the probability using the product of two scalars:
	\begin{align}
		\xircuit_{root}(\xvars) = \widehat{\epsilon}_{root}(\xvars) \widehat{\epsilon}_{root}^*(\xvars),
	\end{align}
	As \citet{loconte2024subtractive} only formulated \snpcs for real-valued parameters, we need to impose a further assumption:
	$\widehat{\epsilon}_{root}(\xvars)= \widehat{\epsilon}_{root}^*(\xvars)$, which only holds for strictly real-valued scalars. Finally, we can write the probability computed by a \pvc as the square of a scalar:
	\begin{align}
		\xircuit_{root}(\xvars) = \widehat{\epsilon}_{root}^2(\xvars)
	\end{align}
	We conclude that we recover \snpcs as a special case of \pvc by assuming $\gamma$ and consequently $\rho$ to be a rank-one matrix, and furthermore assuming that we only have real-valued parameters.
\end{proof}

### 2025-06-02 202507_uai_puncs/icml2025/appendix.tex
added uai

\section*{\LARGE Appendix}

\section{Proofs}

\subsection{Proof of Proposition \ref{prop:povmprob}}
\label{sec:proof:prop:povmprob}

\proppovmprob*

\begin{proof}
	While this is a well-known result we were not able to identify a concise proof in the literature and provide therefore, for the sake of completeness here. To this end, we
	first show that $p(i)\geq 1$, for each $i$.
	\begin{align}
		p(i)
		=
		\Tr [ E(i) \rho ]
		% \nonumber
		% \\
		 &
		=
		\Tr [ D(i) D^*(i)  \rho ]
		% \nonumber
		\\
		 &
		=
		\Tr [ D(i)^*  \rho D(i) ]. \nonumber
	\end{align}
	Here we used the fact that $E(i)$ is PSD and factorized it into the product $D(i) D^*(i)$. Then we used the fact that the trace is invariant under cyclical shifts.
	Clearly, the matrix $ D(i)^*  \rho D(i)$ is PSD as we have for every vector $x$:
	\begin{align}
		x^* D(i)^*  \rho D(i) x = y^* \rho y \geq 0 \quad \text{with } y = Dx.
		\label{eq:def:psdPOVM}
	\end{align}
	As the trace of a PSD matrix is positive we have that $p(i)$ is a positive real number, \ie $p(i){\geq} 0$ for every $i$.

	Secondly, we show that $p(i)$ is normalized:
	\begin{align}
		\sum_{i=1}^N p(i)
		%  &
		=
		\sum_{i=1}^N \Tr [ E(i) \rho ]
		% \nonumber
		% \\
		%  &
		=
		\Tr [ \sum_{i=1}^N E(i) \rho ]
		% \nonumber
		% \\
		%  &
		{=}
		\Tr [  \rho ],
	\end{align}
	where we used Equation~\ref{eq:povm_normalized}.
	Exploiting the fact that the trace of a density matrix is one gives us indeed $\sum_{i=1}^N p(i)=1$, and we can conclude that $p(i)$ is a valid probability distribution.
\end{proof}

\subsection{Proof of Proposition \ref{prop:qopunital}}
\label{sec:proof:prop:qopunital}

\propqopunital*

\begin{proof}
	The (sufficient and necessary) condition that $\sum_j \kraus_j^* \kraus_j \leq \mathbb{1}$ stems from fact that we wish to bound the probability of a state $\qop (E(i))$ is less than or equal to $1$, \cf \citep[Proof of Theorem 8.1]{nielsen2001quantum}. That is, we wish to have:
	\begin{align}
		\Tr [ \qop (E(i)) \rho] \leq 1. \label{eq:proof:unitalvalid:bound}
	\end{align}
	We will now show that for unital quantum operations this holds by construction and that the condition $\sum_j K_j^* K_j$ is implied.

	We start with the probability for the state $\sum_i E(i)=\mathbb{1}$:
	\begin{align}
		\Tr [ \qop (\mathbb{1}) \rho]
		=
		\Tr [ \sum_j \kraus_j  \mathbb{1} \kraus_j^* \rho ]
		=
		\Tr [ \sum_j \kraus_j  \kraus_j^* \rho ]
		=
		\Tr [  \rho ] =1
	\end{align}
	Alternatively, we write this also as:
	\begin{align}
		\Tr [ \qop (\mathbb{1}) \rho]
		=
		\Tr [ \sum_j \kraus_j  \mathbb{1} \kraus_j^* \rho ]
		=
		\sum_j \Tr [ \kraus_j  \mathbb{1} \kraus_j^* \rho ]
		=
		\sum_j \Tr [\gamma^* \kraus_j  \mathbb{1} \kraus_j^* \gamma ]
	\end{align}
	with $\rho=\gamma \gamma^*$.

	Similarly we also write for the probability of the arbitrary state $\sigma$ denoting the sum over any subset of the POVM $\{ E(i) \}_{i=1}^{N}$:
	\begin{align}
		\Tr [ \qop (\sigma) \rho]
		=
		\sum_j \Tr [\gamma^* \kraus_j \sigma \kraus_j^* \gamma ].
	\end{align}
	We now need that $\Tr [ \qop (\sigma) \rho]\leq1$, which is equivalent to:
	\begin{align}
		1 - \Tr [ \qop (\sigma) \rho] \geq 0
		 & \Leftrightarrow
		\Tr [ \qop (\mathbb{1}) \rho] - \Tr [ \qop (\sigma) \rho] \geq 0
		\\
		 & \Leftrightarrow
		\sum_j \Tr [\gamma^* \kraus_j  \mathbb{1} \kraus_j^* \gamma ]
		-
		\sum_j \Tr [\gamma^* \kraus_j \sigma \kraus_j^* \gamma ] \geq 0
		\\
		 & \Leftrightarrow
		\sum_j \Tr \Bigl[\gamma^* \kraus_j  \Bigl( \mathbb{1} -\sigma \Bigr)  \kraus_j^* \gamma \Bigr] \geq 0
	\end{align}
	The last line only holds if $\mathbb{1}- \sigma$ is PSD, which is indeed the case as $\sigma$ only sums over a subset of the POVM. Substracting the sum of this subset from $\mathbb{1}$ leaves us with a sum over the remaining elements of the POVM. As this is a a sum over PSD matrices the sum over the remaining elements is again PSD.
	This concludes the proof as we have shown that Equation~\ref{eq:proof:unitalvalid:bound} is satisfied by construction.
\end{proof}

\subsection{Proof of Proposition \ref{prop:puncPOVM}}
\label{sec:proof:prop:puncPOVM}

\proppuncPOVM*

\begin{proof}
	Given that positive unital circuits are by definition positive operator circuits we already have that:
	\begin{align}
		\forall \xvars \in \Omega(\Xvars): \punccircuit(\xvars) \text{ is PSD}.
	\end{align}
	Next we show that $\sum_{\xvars\in \Omega(\Xvars)} \punccircuit(\xvars)=\mathbb{1}$. Here we observe that in the computation units we have:
	\begin{align}
		\sum_{\xvars_k \in \Omega(\Xvars_k)} \punccircuit_k(\xvars_k)
		 & = \sum_{\xvars_{k_l}} \sum_{\xvars_{k_r}}    \qop(\punccircuit_k(\xvars_{k_l}) \otimes \punccircuit_k(\xvars_{k_r})  )
		\\
		 & =
		\qop \left(
		\sum_{\xvars_{k_l}}   \punccircuit_k(\xvars_{k_l})
		\otimes
		\sum_{\xvars_{k_r}}\punccircuit_k(\xvars_{k_r})
		\nonumber
		\right)
	\end{align}
	This lets us push down the summation of a specific variable to the corresponding leaf where the variable is given as input, where we then have:
	\begin{align}
		\sum_{\xvars_k \in \Omega(\Xvars_k)} \punccircuit_k(\xvars_k)
		= \sum_{\xvar_k\in \Omega(X_k)} E_{\xvar_k}
		= \mathbb{1_k}
	\end{align}
	We now exploit that the completely positive maps in a positive unital circuit are unital, which gives us indeed
	$
		\sum_{\xvars\in \Omega(\Xvars)} \punccircuit(\xvars)=\mathbb{1}.
	$
\end{proof}

% \subsection{Proof of Proposition \ref{prop:computationalcomplexity}}
% \label{sec:proof:prop:computationalcomplexity}

% \propcomputationalcomplexity*

% \begin{proof}
% 	We determine the computational cost of evaluating the circuit by determining the cost of the individual computation units. We start at the leaves where we first have a Kronecker product $ e_{\xvar_k} \otimes  e_{\xvar_k}^*$, which can be performed in $\bigO (\samplespacesize^2)$.
% 	The matrix products with $A_k$ and $A_k^*$ can then be performed in $\bigO (\numbond^2 \samplespacesize)$, which implies $\bigO(\bigO (\samplespacesize^2))$ if $B\leq N$.
% 	In the bilinear form (\cf Equation~\ref{eq:polyadicbilinear}) we perform four matrix-matrix product in $\bigO (\numbond^3)$. This is followed by a Hadamard product in $\bigO (\numbond^2)$,
% 	This means that we have for the individual computation units in the partition a circuit a cost of $\bigO (\samplespacesize^3)$.

% 	It is also easy to show that for binary tree partition circuits we have a total of $\bigO (\numvar)$ computation units. This results in a computational complexity of $\bigO (\numvar  \samplespacesize^3)$.

% 	We note the computation of trace of the matrix-matrix multiplication between $\punccircuit_{root}(\xvars)$ and the density matrix $\rho$ necessary for obtaining a probability can be done in $\bigO(\numbond^2)$.

% 	Given that we can evaluate \puncs $\bigO ( \numvar \samplespacesize^3)$ and also that we can marginalize out variables by pushing the summation to the leaves (\cf the proof of Proposition~\ref{prop:puncPOVM}), we can conclude that marginal inference has complexity $\bigO(\numvar \samplespacesize^3)$, as well.
% \end{proof}

\subsection{Proof of Proposition \ref{prop:Oveq}}
\label{sec:proof:prop:Oveq}

\propOveq*

\begin{proof}
	We start the proof at the leaf where we have
	\begin{align}
		\Ocircuit_k(\xvar_k)
		=
		U_k \left( e_{\xvar_k} \otimes e_{\xvar_k}^*  \right) U_k^*
		=
		\left( U_k e_{\xvar_k} \right) \otimes \left( e_{\xvar_k}^* U_k^* \right)
	\end{align}
	and Equation~\ref{eq:def:opvec_equivalent} holds almost by definition. In the computation units into which the leaves feed, we then have
	\begin{align}
		\Ocircuit_k
		 & = U_k \left( \Ocircuit_{k_l} \otimes \Ocircuit_{k_r} \right) U_k^*
		\nonumber
		\\
		 & = U_k \left( \vcircuit_{k_l} \otimes \vcircuit^*_{k_l} \otimes (\vcircuit_{k_r} \otimes \vcircuit^*_{k_r} ) \right) U_k^*
		\nonumber
		\\
		 & = \Bigl( U_k \left( \vcircuit_{k_l} \otimes \vcircuit_{k_r} \right) \Bigr)
		\otimes
		\Bigl( \left( \vcircuit^*_{k_l}  \otimes \vcircuit^*_{k_r}  \right) U_k^* \Bigr)
		\nonumber
		\\
		 & = v_k \otimes v_k^*.
	\end{align}
	where we omitted the explicit dependencies on the variables $\xvar_k$, $\xvar_{k_l}$, and $\xvar_{k_r}$.
	Repeating this argument recursively until the root of the circuit concludes the proof.
\end{proof}

\subsection{Proof of Proposition \ref{prop:voequiprob}}
\label{sec:proof:prop:voequiprob}

\propvoequiprob*

\begin{proof}
	The proof starts by simply plugging in the vector representation of $\Ocircuit_{\text{root}}$ (obtained in the proof of Proposition~\ref{prop:Oveq}) into the expression $\Tr [\Ocircuit_{\text{root}} \rho]$ and rather straightforwardöy get:
	\begin{align}
		\Tr [\Ocircuit_{\text{root}} \rho]
		 &
		=
		\Tr \left[
			\Bigl(\vcircuit_{\text{root}} \otimes \vcircuit^*_{\text{root}}   \Bigr) \times \rho
			\right]
		\\
		 & = \vcircuit^*_{\text{root}} \times  \rho \times\vcircuit_{\text{root}},
	\end{align}
	which is indeed the same probability as defined in Defintion~\ref{def:vpoc}.
\end{proof}

% \subsection{Proof of Proposition \ref{prop:computationalcomplexityvec}}
% \label{sec:proof:prop:computationalcomplexityvec}

% \propcomputationalcomplexityvec*

% \begin{proof}
% 	Following a similar train of thought as the proof in Section~\ref{sec:proof:prop:computationalcomplexity} we can derive the computational cost of evaluating a positive vector circuit to be $\bigO (\numvar \numbond \samplespacesize)$ implying
% 	$\bigO (\numvar  \samplespacesize^2)$
% 	when $\numbond \leq \samplespacesize$.
% 	The major difference between matrix representation of operator circuits and their vector representation is that for the former the internal computation units output matrices ($\Ocircuit_k(\xvars_k)\in \mathbb{C}^{\numbond \times \numbond}$) while for the latter vectors are outputted ($\vcircuit_k(\xvars_k)\in \mathbb{C}^{\numbond}$).

% 	As $\vcircuit_{root}(\Xvar_{rooot})$ is computable in $\bigO (\numvar \numbond \samplespacesize)$ and $\lVert \gamma \times \vcircuit_{\text{root}}\rVert^2$ is computable in $\bigO(\numbond^2)$, we can conclude that the probability $p_\Xvars(\xvars)$ can be computed in $\bigO (\numvar \samplespacesize^2)$.
% \end{proof}

### 2025-06-02 202507_uai_puncs/icml2025/content.tex
added uai

\begin{abstract}
	By recursively nesting sums and products, probabilistic circuits have emerged in recent years as an attractive class of generative models as they enjoy, for instance, polytime marginalization of random variables.
	In this note we study these machine learning models using the framework of quantum information theory, leading to the introduction of \textit{positive unital circuits} (\puncs),
	which generalize circuit evaluations over positive real-valued probabilities to circuit evaluation over positive semidefinite matrices.
	As a consequence, \puncs strictly generalize probabilistic circuits as well as the recently introduced class of PSD circuits.
\end{abstract}

\section{Introduction}

Probabilistic circuits (PCs)~\citep{darwiche2003differential,poon2011sum} belong to an unusual class of probabilistic models: they are highly expressive but at the same time also tractable.
For instance, so-called decomposable probabilistic circuits~\citep{darwiche2001decomposable} allow for the computation of marginals in time polynomial in the size of the circuit.
\citet{zhang2020relationship} noted that it is exactly this restriction to positive values that limits the expressive efficiency (or succinctness) of PCs~\citep{martens2014expressive,decolnet2021compilation}. In particular, the positivity constraint on the set of elements that PCs operate on prevents them from modelling negative correlations between variables.
Circuits that are incapable of modelling negative correlations, \ie circuits that can only combine probabilities in an additive fashion, are also called monotone circuits~\citep{shpilka2010arithmetic}.
This restricted expressiveness can be combatted by the use of so-called \textit{non-monotone} circuits, where subtractions are allowed as a third operation (besides sums and products). Interestingly, \citet{valiant1979negation} showed that a mere single subtraction can render non-monotone circuits exponentially more expressive than monotone circuits a result that has recently been refined for decomposable circuits~\citep{loconte2024sum}.

As shown in \citep{harviainen2023inference,agarwalprobabilistic}, non-monotone circuits do, however, introduce an important complication: if non-monotone circuits are not designed carefully, verifying whether a circuit encodes a valid probability distribution or not is an NP-hard problem. This does also render learning the parameters of a circuit practically infeasible.

Using the concept of \textit{positive operator valued measures} from quantum information theory, which encode random event as positive semidefinite matrices we are able to construct non-monotone circuits that nonetheless encode proper (normalized) probability distributions by construction.
Our work falls into a line of recent works presented in the circuit literature ~\citep{sladek2023encoding,loconte2024subtractive,wangrelationship,loconte2024sum}. However, our work is the first that establishes this deep connection between concepts in quantum information theory and tractable probabilistic models.
Furthemorer, the circuits class of probabilistic unital circuits that we introduce generalizes both the probabilistic circuits as well as PSD circuits~~\citep{sladek2023encoding,loconte2024subtractive,loconte2024sum}\citep{sladek2023encoding}\footnote{PSD circuits were later on rebranded as sum of squares circuits~\citep{loconte2024sum}}.

% \begin{enumerate*}
% 	\item We generalize probabilistic circuits over scalars to operator-valued circuits and introduce \puncs (Section~\ref{sec:puncs}).
% 	\item We use the concept of noise in quantum information theory to construct deep operator mixture circuits, dubbed \noisepuncs (Section~\ref{sec:noisecircuits}).
% 	\item We provide an efficient parametrization such that the constructed probability distributions using \puncs are normalized by construction (Section~\ref{sec:efficient_param}).
% \end{enumerate*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{A Primer on Quantum Information Theory}
\label{sec:qit}

A widely used and elegant framework to describe measurements of quantum systems is the so-called \textit{positive operator-valued measure} (POVM) formalism. While POVMs have physical interpretations in terms of quantum information and quantum statistics, we will only be interested in their mathematical prperties as we use them to show that  circuits (defined in Section~\ref{sec:puncs}) form valid probability distributions.
We refer the reader to \citep{nielsen2001quantum} for an in-depth exposition on the topic, as well as quantum computing and quantum information theory in general.
\begin{definition}[Positive Semidefinite]
	An $\numbond {\times} \numbond$ Hermitian matrix $H$ is called positive semidefinite (PSD) if and only if $\forall \xvars {\in}  \mathbb {C} ^{\numbond}: \xvars^{*} H \xvars {\geq} 0$, where $\xvars^{*}$ denotes the conjugate transpose and $\mathbb {C}^{\numbond}$ the $\numbond$-dimensional space of complex numbers.
\end{definition}
\begin{definition}[{POVM~\citep[Page 90]{nielsen2001quantum}}]
	\label{def:povm}
	A positive operator-valued measure
	% with a finite number of elements acting on a finite-dimensional Hilbert space $\mathcal{H}$,
	is a set of PSD  matrices $\{E(i)\}_{i=0}^{\numevents-1}$ ($I$ being the number of possible measurement outcomes) that sum to the identity:
	\begin{align}
		\sum_{i=0}^{\numevents-1} E(i) = \mathbb{1},
		\label{eq:povm_normalized}
	\end{align}
\end{definition}
Before defining the probability of a specific $i$ occurring, we need the notion of a density matrix~\citep{neumann1927wahrscheinlich}:
\begin{definition}[{Density Matrix~\citep[Page 102]{nielsen2001quantum}}]
	\label{def:density_matrix}
	A density matrix $\rho$ is a PSD matrix of trace one, \ie $\Tr [\rho]=1$.
\end{definition}
\begin{definition}[{Event Probability~\citep[Page 102]{nielsen2001quantum}}]
	\label{def:eventprob}
	Let $\rho$ be a density matrix and let $i$ denote an event with $E(i)$ being the corresponding element from the POVM. The probability of the event $i$ happening, \ie measuring the outcome $i$, is given by
	\begin{align}
		p(i) = \Tr [ \rho E(i)]
		\label{eq:povm_prob}
	\end{align}
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{restatable}{proposition}{proppovmprob}
	\label{prop:povmprob}
	The expression in Equation~\ref{eq:povm_prob} defines a valid probability distribution.
\end{restatable}

\begin{proof}
	While this is a well-known result we were not able to identify a concise proof in the literature. We therefore provide one in Appendix~\ref{sec:proof:prop:povmprob}.
\end{proof}

Given that the $E(i)$'s completely describe the event $i$ such that its event probability can be computed, they represent the quantum state of a system. This quantum state (represented by a matrix) lives in a certain Hilbert space. The changes that a quantum state can undergo are then described by so-called \textit{quantum operations} acting on the Hilbert space. We can construct such operations using Kraus' theorem.

\todo{change theorm a bt more to make it more different from wikipedia}
\begin{theorem}[Kraus' Theorem\citep{kraus1983states}]
	Let $\mathcal{H}$  and $\mathcal {G}$ be Hilbert spaces of dimension $N$ and $M$ respectively, and $\qop$ be a quantum operation between $\mathcal{H}$  and $\mathcal {G}$. Then, there are matrices
	$\{ \kraus_j \}_{j=1}^{D}$ (with $D\leq NM$)
	mapping $\mathcal{H}$ to $\mathcal {G}$ such that for any state $E(i)$
	\begin{align}
		\qop(E(i)) = \sum_{j=1}^{D} \kraus_j E(i) \kraus_j^*
	\end{align}
	provided that $ \sum _{j} \kraus_{j}^{*} \kraus_{j}\leq \mathbb {1} $ (in the Loewner order sense).
\end{theorem}

\begin{proof}
	See \citep[Chapter 8]{nielsen2001quantum}
\end{proof}
The $\kraus_j$ matrices are usually referred to as Kraus operators.

% \todo{talk about Loewner order}

% \citep[Corollary 4.2]{baksalary1989some}

\section{Positive Unital Circuits}
\label{sec:puncs}

\begin{figure}[t]

	\centering
	\input{tex_input/partition}

	\caption{
		Partition circuit over four binary variables $x_i$ with $i \in \{0,1,2,3\}$, which are given as inputs to the circuit. The internal nodes of the partition circuit correspond the computation units.}
	\label{fig:circuit}
\end{figure}

A popular subclass of probabilistic circuits are so-called structured decomposable  probabilistic circuits~\citep{darwiche2011sdd} that are also smooth~\citep{darwiche2001tractable}. The advantage of this circuit subclass is that they can be implemented in a rather straightforward fashion on modern AI accelerators, as demonstrated by \citet{peharz2019random,peharz2020einsum}.
For the sake of exposition, we will limit ourselves in first instance to such circuits that adhere to structured decomposability and will generalize to (non-structured) decomposable circuits in Section~\ref{sec:nonstructdecomp}. For a detailed account on these different circuit properties we refer the reader to~\citep{vergari2021compositional}.

\citet{zuidberg2024probabilistic} introduced an abstraction for these smooth structured decomposable circuits in the form of partition trees. We further refine this introducing the concept of \textit{partition circuits}.
We give such a circuit in Figure~\ref{fig:circuit}.
\footnote{The concept of a partition tree, and hence a partition circuit, is related to the concept of a variable tree~\citep{pipatsrisawat2008new} and a pseudo-tree~\citep{dechter2007and}. However, partition circuits emphasize an interpretation as computation graphs, which the others do not.}

\begin{definition}[Partition Circuit]
	\label{def:partition_circuit}
	A partition circuit over a set of variables is a parametrized computation graph taking the form of a binary tree. The partition circuit consists of two kinds of computation units:
	\textit{leaf} and \textit{internal} units, as well as a single \textit{root}.
	Units at the same distance from the root form a layer.
	Furthermore, let $\circuit_k$ denote the root unit or an internal unit. The unit $\circuit_k$ then receives its inputs from two units in the previous layer, which we denote by $\circuit_{k_l}$ and $\circuit_{k_r}$. Each computation unit is input to exactly one other unit, except the root unit, which is the input to no other unit.
\end{definition}

\subsection{Positive Operator Circuits}

Using the concept of partition circuits we construct positive operator circuits. Positive operator circuits can be thought of as generalizing circuit evaluations with probabilities to circuit evaluations with PSD matrices. Note, in the definition below we use $\Ocircuit_k$ instead of $\circuit_k$ to make this generalization explicit.
\begin{definition}[Positive Operator Circuit]
	\label{def:poc}
	Let   $\xvars{=}\{\xvar_0,\dots ,\xvar_{\numvar{-}1}  \}$ be a set of $M$ categorical variables with domains of size $\samplespacesize$.
	We define an operator circuit as a partition circuit whose computation units take the following functional form:
	\begin{align}
		 & {\Ocircuit}_k(\xvars_k){=}
		\begin{cases}
			E_{\xvar_k},
			% A_{k} \times  e_{\xvar_k} \otimes  e^*_{\xvar_k} \times A_{k}^*,
			 & \text{if $k$ is leaf}
			\\
			\qop_k \Bigl(\Ocircuit_{k_l}(\xvars_{k_l}) \otimes \Ocircuit_{k_l}(\xvars_{k_l}) \Bigr)
			 & \text{else},
		\end{cases}
		\label{eq:poc:def}
	\end{align}
	where the  $E_{\xvar_k}$'s are quantum state matrices, and where the $\qop_k$ quantum operations.
\end{definition}
Note that using the Kronecker product between $\Ocircuit_{k_l}(\xvars_{k_l})$ and  $\Ocircuit_{k_l}(\xvars_{k_l})$ is a sensible choice as it describes the joint state of both subsystems.

% \begin{definition}
% 	We call an operator circuit \textit{positive} if we have that the bilinear forms $\bilinear_k$ are such that
% 	\begin{align}
% 		\bilinear_k \Bigl(\Ocircuit_{k_l}(\xvars_{k_l}), \Ocircuit_{k_l}(\xvars_{k_l}) \Bigr) \quad \text{is a PSD matrix}
% 		\label{def:eq:bilinear_psd}
% 	\end{align}
% 	if $\Ocircuit_{k_l}(\xvars_{k_l})$ and $\Ocircuit_{k_l}(\xvars_{k_l})$ are PSD.
% \end{definition}

\begin{restatable}{proposition}{proppocpsd}
	\label{prop:pocpsd}
	Positive operator circuits are PSD.
\end{restatable}

\begin{proof}
	We know that all the leaves carry PSD matrices as they describe quantum states. Passing these on recursively to the quantum operations in the internal units retains the positive semi-defiteness as the Kronecker product between two PSD matrices is again PSD.
\end{proof}

\subsection{Constructing a Probability Distribution}

In Section~\ref{sec:qit} we saw that we can construct a probability distribution using a density matrix $\rho$ and a positive operator-valued measure, with the latter being a set of PSD matrices (\cf Definition~\ref{def:povm}) that sum to the unit matrix. Using a positive operator circuit $\Ocircuit(\xvars)$ we indeed have a set of PSD matrices. Namely, on for each instantiation of the $\xvars$ variables. We now introduce \textit{positive unit preserving operator circuits} for which also the summation to the unit matrix holds.

\begin{definition}
	\label{def:cp_unital}
	We call a quantum operation \textit{unital} if we have that
	\begin{align}
		\qop_k(\mathbb{1}_{k_l} \otimes \mathbb{1}_{k_r})
		=\qop_k(\mathbb{1}_{k_l k_r})
		=  \mathbb{1}_k,
	\end{align}
	where $\mathbb{1}_{k}$, $\mathbb{1}_{k_l}$, $\mathbb{1}_{k_l k_r}$, and $\mathbb{1}_{k_r}$  denote unit matrices of appropriate size,
\end{definition}

\begin{restatable}{proposition}{propqopunital}
	\label{prop:qopunital}
	Unital quantum operations are valid in the sense that the inequality $\sum_j \kraus_{j}^* \kraus_{j} \leq \mathbb{1}$ holds for all unital quantum operations.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:qopunital}
\end{proof}

\begin{definition}
	We call a positive operator circuit \textit{unital} if we have that the quantum operation $\qop_k$ are unital, and if the sets $\{ E_{\xvar_k} \}_{\xvar_k \in \Omega(\Xvar_k)}$ form a POVM for each random variables $\Xvar_k$.
\end{definition}
We will also refer to positive unital operator circuits as positive unital circuits, or \puncs.

\begin{restatable}{proposition}{proppuncPOVM}
	\label{prop:puncPOVM}
	Let $\Xvars$ denote a set of random variables with sample space $\Omega(\Xvars)$.
	Then the set $\{ \punccircuit(\xvars)\}_{\xvars \in \Omega(\Xvars)}$ of positive unital circuits forms a POVM.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:puncPOVM}
\end{proof}

\begin{restatable}{theorem}{theopuncprobdist}
	\label{theo:puncprobdist}
	Let $\rho$ be a density matrix and $\punccircuit(\xvars)$ a positive unital circuit. The function
	\begin{align}
		p_\Xvars(\xvars) = \Tr [\punccircuit(\xvars) \rho]
		\label{eq:theo:prob_operator}
	\end{align}
	is a proper probability distribution over the random variables $\Xvars$ with sample space $\Omega(\Xvars)$.
\end{restatable}

\begin{proof}
	This follows from Proposition~\ref{prop:povmprob} and Proposition~\ref{prop:puncPOVM}
\end{proof}

One of the outstanding properties of probabilisitc circuits is that they are tractable -- in the sense that they allow for polytime marginalization of random variables. Positive unital circuits retain this property.

\begin{proposition}
	Positive unital circuits allow for tractable marginalization.
\end{proposition}

\begin{proof}
	(Sketch) The proof is rather straightforward and hinges on the fact that the quantum operations in the internal coputation units are computable in poytime and on the fact that the marginalization of a random variables is performed by pushing the sum to the corresponding leaf in the partition circuit. Anologous to the proof of Proposition~\ref{prop:puncPOVM}.
\end{proof}

\section{Special Cases}
\label{sec:special_cases}

We will now make certain structural assumption on the matrices representing the quantum states and the functional form of the quantum operations $\qop$. By doing so, we obtain the PSD circuits introduced by~\citet{sladek2023encoding} and the (strucutred-decomposable) probabilistic circuits as described by \citet{peharz2020einsum} as special cases of positive unital circuits (Section~\ref{sec:purestate} and Section~\ref{sec:mixeigenstate} respectively).

First, however, we note our formulation of \puncs already encompasses canonical polyadic tensor decomposition ~\citep{carroll1970analysis} -- a popular choice in the circuit literature \citep{shih2021hyperspns,loconte2024relationship} to merge partitions that uses the Hadamard product instead of the Kronecker product.

Specifically, we observe that the Hadamard product between two matrices $A$ and $B$ can be rewritten using Kronecker product;
\begin{align}
	A \circ B = P (A \otimes B) P^*,
\end{align}
where $P$ is the semi-unitary partial permutation matrix selecting a principal submatrix \citep[Corollary 2]{visick2000quantitative}.
This also means that a quantum operation involving q Hadamard product can be rewritten using a Kronecker product:

\begin{align}
	\qop (A \circ B )
	 & = \sum_i K_i (A\circ B) K_i^*
	\nonumber
	\\
	 & = \sum_i K_i P  (A\otimes B) P^* K_i^*
	\nonumber
	\\
	 & =  \sum_i K_i'  (A\otimes B) K_i'^* = \qop'(A \otimes B)
\end{align}
Note that, for $\qop'$ to be unital it suffices that $\sum_i K_i K_i^* = \mathbb{1}$ as $P$ is semi-unitary ($PP^* = \mathbb{1}$).

From the discussion above we conclude that we can safely limit the discussion to circuits with Kronecker products as circuits with Hadamard products follow as a special case.

\subsection{Pure Quantum States}
\label{sec:purestate}

As the matrices that represent quantum states are PSD we can decompose them as follows using the spectral theorem:
\begin{align}
	\qcircuit = \vcircuit_j \otimes \vcircuit_j^*,
\end{align}
with the $vcircuit_j$'s denoteing the eigenvector.
A special case are then so-called pure states for which we have
\begin{align}
	\qcircuit = \lambda \vcircuit \otimes \vcircuit^*,
\end{align}
We will show now that by restricting \puncs to performing operations on pure quantum states gives us the special case of PSD circuits as introduced by \citet{sladek2023encoding}, which we define first using a partition circuit.
\begin{definition}
	\label{def:vpoc}
	Let   $\xvars{=}\{\xvar_0,\dots ,\xvar_{\numvar{-}1}  \}$ be a set of $\numvar$ categorical variables with domains of size $\samplespacesize$.
	A PSD circuit is a partition circuit whose computation units take the following functional form:
	\begin{align}
		{\vcircuit}_k(\xvars_k){=}
		\begin{cases}
			U_{k} \times  e_{\xvar_k},
			 & \text{if $k$ leaf}
			\\
			U_{k} {\times}  \left( \vcircuit_{k_l}  (\xvars_k)  \otimes   \vcircuit_{k_r} (\xvars_k) \right),
			 & \text{else}
		\end{cases}
		\label{eq:vector_units}
	\end{align}
	where the $U_k$'s are semi-unitary matrices.
	The probability $p(\xvars)$ is computed via
	\begin{align}
		p(\xvars) = {\vcircuit}^*_{root}(\xvars) \times  \rho \times {\vcircuit}_{root}(\xvars),
	\end{align}
	where $\rho$ is a density matrix.
\end{definition}
Note that in the original formulation \citet{sladek2023encoding} used non-semi-unitary matrices. However, \citet{loconte2024faster} have recently shown that there is no loss in expressiveness when using semi-unitary matrices.

To show that PSD circuits are a special case of \puncs we now impose the following restriction on the quantum operations $\qop_k$:
\begin{align}
	\qop_k (\Ocircuit_{k_l} \otimes \Ocircuit_{k_l} )
	 & =
	\kraus_{k} \left( \Ocircuit_{k_l}  \otimes  \Ocircuit_{k_r} \right) \kraus_{k}^*
	\label{eq:pureinternal}
\end{align}
That is, we limit the quantum operation to having  only a single pair of Kraus operators. For the quantum operation to be unital we need to have $\kraus_{k1} \kraus_{k1}^*=\mathbb{1}$. That is, $\kraus_{k1}$ has to be semi-unitary,

Furthermore, we make the following choice in the leaves:
\begin{align}
	E_{\xvar_k} = K_{k} \left( e_{\xvar_k} \otimes e_{\xvar_k}^* \right) \kraus_{k}^*,
	\label{eq:pureleaf}
\end{align}
where the set  $\{ e_{\xvar_k} \}_{\xvar_k \in \Omega(\Xvar_k)}$ is a complete set of orthonormal basis vectors, and $\kraus_{k}$ is again semi-unitary.

We can show that this choice for $E_{\xvar_k}$ forms a POVM. Firstly, by observing that each $E_{\xvar_k}$ is PSD. Secondly, by verifying the completeness of the set of operators:
\begin{align}
	\sum_{\xvar_k \in \Omega(\Xvar_k)} E_{\xvar_k}
	 & = \sum_{\xvar_k \in \Omega(\Xvar_k)} K_{k} \left( e_{\xvar_k} \otimes e_{\xvar_k}^*  \right) K_{k}^*
	\nonumber
	\\
	 & =
	K_{k} \left(  \sum_{\xvar_k \in \Omega(\Xvar_k)}  e_{\xvar_k} \otimes e_{\xvar_k}^*  \right) K_{k}^*
	\nonumber
	\\
	 & =
	K_{k} \mathbb{1} K_{k}^*
	\nonumber
	\\
	 & = \mathbb{1}
\end{align}

\begin{definition}
	We call a positive unital circuit pure if Equation~\ref{eq:pureinternal} and Equation~\ref{eq:pureleaf} hold.
\end{definition}

\begin{restatable}{proposition}{propOveq}
	\label{prop:Oveq}
	For computation units of a pure positive unital circuit and a PSD circuit it holds that
	\label{prop:Oveq}
	\begin{align}
		\forall k: \Ocircuit_k(\xvars_k) = \vcircuit_{k}(\xvars_{k}) \otimes \vcircuit^*_{k}(\xvars_{k}).
		\label{eq:def:opvec_equivalent}
	\end{align}
	given that $U_k=\kraus_k$
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:Oveq}
\end{proof}

\begin{corollary}
	Pure positive unital circuits perform operations on pure quantum states exclusively.
\end{corollary}

\begin{proof}
	This follows immediately from Proposition~\ref{prop:Oveq}.
\end{proof}

\begin{restatable}{proposition}{propvoequiprob}
	\label{prop:voequiprob}
	A PSD circuit and a pure \punc encode the same probability distribution if $U_k=\kraus_k$ for each unit $k$.
\end{restatable}

\begin{proof}
	See Section~\ref{sec:proof:prop:voequiprob}
\end{proof}

In this section we have shown that by making specific choices in the functional form of the leaves and the internal units of a positive unital circuit that we recover the the special case PSD circuits and its variants \citep{loconte2024subtractive,loconte2024sum}. Furthermore, our analysis also provides the rather satisfying interpretation of PSD circuits as quantum circuits acting on pure states exclusively.

% Lastly,  for the set of  basis vectors $\{e_{\xvar_s} \}_{s=0}^{\samplespacesize-1}$ in the leaves we choose the set of standard basis vectors, \ie one-hot unit vectors, with each basis vector $e_s$ corresponding to a random outcome. Although different choices of basis vectors would be valid as well, For the $A_k$ we use, as already mentioned, the orthonormal discrete Fourier transform matrix, which we truncate if $\numbond<\samplespacesize$.

% In order to parametrize positive unitary circuits, we need to find a functional form of the bilinear forms such that Equation~\ref{def:eq:bilinear_psd} and Equation~\ref{def:eq:bilinear_unitpreserving} hold, and such that we have $A_k A_k^*=\mathbb{1}$. This means that we have to chose $A_k$ such that it is semi-unitary.

% For $\gamma$ this is rather straightforward as we can simply use:
% $
% 	\gamma = \frac{\tilde{\gamma}}{\sqrt{\Tr [\tilde{\gamma} \tilde{\gamma}^*]}},
% $
% where $\tilde{\gamma}\in \mathbb{C}^{B\times B}$ is an arbitrary complex values matrix. Note that we have almost by definition $\Tr [\rho] = 1$.

% \begin{restatable}{proposition}{propcomputationalcomplexity}
% 	\label{prop:computationalcomplexity}
% 	Let $\numvar$ be the number of variables $\xvars_k$,  having all the  domain size $\samplespacesize$, and let the $L_k$ and $R_k$ matrices be $\numbond\times \numbond$ square matrices, with $\numbond\leq \samplespacesize$.
% 	We can perform marginal inference in \puncs in $\bigO(\numvar \samplespacesize^3)$ when using the bilinear form from Equation~\ref{eq:polyadicbilinear}.
% \end{restatable}

% \begin{proof}
% 	See Append \ref{sec:proof:prop:computationalcomplexity}
% \end{proof}

% \begin{restatable}{proposition}{propcomputationalcomplexityvec}
% 	\label{prop:computationalcomplexityvec}
% 	\puncs in their vector representation compute joint probabilities in time $\bigO(\numvar \samplespacesize^2)$
% \end{restatable}

% \begin{proof}
% 	See Section~\ref{sec:proof:prop:computationalcomplexityvec}
% \end{proof}

% Note that the vector representation can only be used for computing joint probabilities as marginalizing out variables forces us to go to the matrix representation of operator circuits.

% Note also that the vector representation of \puncs can be seen as a PSD circuits~\cite{sladek2023encoding} or equivalently sum of squares circuits~\citep{loconte2024sum}, which we obtained by imposing a specific functional structure on the bilinear forms.

\subsection{Statistical Mixtures of Eigenstates}
\label{sec:mixeigenstate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Dropping Structured Decomposability}
\label{sec:nonstructdecomp}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Related Work}
\label{sec:related}

\subsection{Squared Cirucuits}

The work closest related to ours are the \text{sum of compatible square circuits} by \citet{loconte2024sum}.
As shown in Section~\ref{sec:poss_func_form} we recover sum of square circuits (introduced earlier as PSD circuits by \citet{sladek2023encoding}) if we use the functional form in Equation~\ref{eq:kroneckerbilinear}.
A non-polyadic decomposition would have resulted in a computation graph closer to that of \textit{inception circuits} introduced by \citet{wangrelationship}. Such a choice, however, increases the computation cost per unit in the partition circuit from quadratic (using the vector representation) to cubic (using the operator representation).

Our positive unit preserving operator circuits provide also a different perspective on constructing non-monotone circuits. While the methods described in ~\citep{sladek2023encoding,loconte2024subtractive,loconte2024sum,wangrelationship} regard such circuits as sum of squares, we interpret them as probabilistic events described by positive semidefinite matrices that are combined within a circuit using unit preserving bilinear forms. As such, we also establish a strong link between the circuit literature and quantum information theory.

% From a practical perspective, the (sub-complete) distributions encoded with \puncs and \noisepuncs tackle one of the main shortcoming of sum of squares type circuits: we do not need to compute a normalization constant to ensure that event probabilities are bounded. This especially useful during training, as it means that updating the parameters of the model still guarantees the event probability to be bounded. Otherwise one would have to compute a normalization constant and differentiate through it at every iteration of the parameter updates. This potentially hinder scaling up non-monotone models as computing the normalization constant cannot be done in the vector representation (quadratic cost) but requires matrix-matrix multiplications with cubic cost. For \puncs computing this normalization is not necessary at all and for \noisepunc we only need to compute the normalization constant once training has finished as post-processing step.

\subsection{Tensor Networks}

As already pointed out by \citet{loconte2024subtractive,loconte2024sum} squared circuits and monotone circuits share many similarities with \textit{tensor networks}~\citep{orus2014practical,white1992density}
developed in the condensed matter physics community and have in recent years also been applied to tackle problems in supervised as well as unsupervised machine learning~\citep{cheng2019tree,han2018unsupervised,stoudenmire2016supervised}. In this regard and given that tensor networks originate in the physics community, it is rather surprising that tensor networks have so far, and to the best of our knowledge, not been formulated using POVMs.

Using our POVM formulation for \puncs is, however, not only theoretically elegant but has also practical benefits: using this formalism leads to circuits that are normalized by construction, and we can perform learning simply by optimizing the likelihood. We contrast this to the sweeping algorithms that are usually deployed in the tensor network literature, where blocks of variables are optimized one at the time while the remaining variables are held constant. It appears that this approach is inspired by the variational ansatz taken in the density-matrix renormalization group algorithm~\citep{white1992density}, which is the original algorithm for tensor networks.
Alternatives to this sweeping approach have also been proposed, such as an intricate Riemannian gradient descent optimization scheme that conserves unitarity of matrices, but are rather costly to run.

We would also like to point out a theoretical result from the tensor network literature stating that picking a complex-valued parametrization instead of a real-valued one can lead to an arbitrarily large reduction in the number of parameters~\citep{glasser2019expressive}. While this result is formulated with respect to (exact) low rank tensor decomposition and does not apply directly to the problem of learning parameters via gradient descent, we consider this observation to be a strong theoretical indicator for the superiority of complex numbers over real numbers when parametrizing probabilistic circuits. A similar argument has also been made by \citet{gao2022enhancing} in the context of hidden Markov models.

\subsection{Theoretical Studies}

First theoretical results on expressive power of polynomial functions, \ie circuits and tensor networks, were presented in the tensor network literature in the context of tensor decomposition \citep{glasser2019expressive} and complex-valued hidden Markov models \citep{gao2022enhancing}. Recently, the works of \citet{loconte2024subtractive,loconte2024sum} and \citet{wangrelationship} have studied the relationship of different circuits classes more carefully, as well. Additionally, \citet{loconte2024sum} pointed out links between tensor networks and the circuit literature and were able to generalize earlier results from the tensor network literature by~\citet{glasser2019expressive}.

Finally, we would also like to point out theoretical results n the statistical relational AI literature. Specifically, \citet{buchman2017rules}, and \citet{kuzelka2020complex} noted that using only real-valued parametrizations (including negatives~\citep{buchman2017negative}), \ie monotone functions exclusively, does not allow for fully expressive models.

% \section{Experimental Evaluation}
% \label{sec:experiments}

% \textbf{Experimental Setup}

% For our experimental evaluation we used the MNIST family of datasets. That is, the original MNIST~\citep{deng2012mnist}, FashionMNIST~\citep{xiao2017fashion}, and also EMNIST~\citep{cohen2017emnist}.
% For the implementation we used PyTorch together with the Einops library~\citep{rogozhnikov2022einops}. We set up our experiments in Lightning\footnote{\url{https://lightning.ai/}} and ran them on a DGX-2 machine with V100 Nvidia cards.

% We compare different methods using the \textit{bits per dimension} metric, which is calculated from the average negative log-likelihood ($\overline{NLL}$) as follows: $bpd {=} \nicefrac{\overline{NLL}}{(\log{2} \times D)}$ ($D{=}28^2$ for MNIST datasets).

% \textbf{Training Positive Operator Circuits}

% In our experiments we estimate the density of the different MNIST datasets by maximizing a parametrized likelihood function. We construct this  function using a positive vector circuit.
% In order to build the underlying partition circuit,
% we follow the approach described by \citet{zuidberg2024probabilistic}: we start with a grid of $28\times 28$ pixels and merge, in an alternating fashion, rows and columns of the image.
% We also allow for merging three partitions into  a single partition, thereby breaking the binary character of the partition trees used so far.
% Having three instead of two partitions being merged can easily be accommodated for by using a Hadamard product with three factors instead of two in the internal units of a \pvc. Ultimately, this allows us to handle layers with an uneven number of partitions.
% At the leaves we encode pixel values by associating each value to one of $256$ standard basis vector of $\mathbb{R}^{256}$.

% We performed training by minimizing the negative log-likelihood using Adam~\citep{kingma2014adam} with default hyperparameters, batch size of $50$, over $100$ epochs.
% The best model was selected using a $90-10$ train-validation data split where we monitored the negative log-likelihood on the validation set.
% We did not perform any hyperparameter search.
% Further details can be found in the configuration files of the experiments.

% \textbf{State-of-the-Art Baselines}

% We compare \pvcs to three state-of-the-art tractable density estimators from the literature: quadrature of probabilistic circuits (QPCs)~\citep{gala2024probabilistic}, hidden Chow-Liu trees (HCLTs)~\citep{liu2021tractable}, and sparse HCLTs (SHCLTs)~\citep{dang2022sparse}.
% All three methods are probabilistic circuits and they all use hidden Chow-Liu trees as their underlying structure. This tree is learned and dataset specific. The difference between QPCs and HCLTs is that the former is obtained by approximating a continuous latent variable extension of probabilistic circuits using numerical quadrature. The difference between HCLTs and SHCLTs is that the latter allows for dynamically adapting the number of components per partition during parameter learning. This means that SHCLTs can focus their computational budget on information-rich parts of the circuit.

% \textbf{Q1: How Do the Different \pvc Parametrizations Fair Against Each Other?}

% We construct \pvcs following the prescription of \cite{zuidberg2024probabilistic}, as described above. Furthermore, we use $B=128$ components per partition.
% As for the parametrization we study four different variants:
% $\mathrm{PVX}_{0}^{FR}$,
% $\mathrm{PVX}_{2\pi}^{FR}$,
% $\mathrm{PVX}_{0}^{R1}$, and
% $\mathrm{PVX}_{2\pi}^{R1}$.
% The subscripts ($0$ or $2\pi$) indicate whether we limit the phase to be zero or whether we allow the phase to be a learnable parameter. Note that having a tunable phase doubles the number of (real-valued) parameters. The superscript indicates whether the parametrization uses a full-rank ($FR$) density matrix in the root or a rank-one ($R1$) density matrix.
% As such, $\mathrm{PVX}_{0}^{R1}$ is equivalent to the  squared monotonic probabilistic circuits (\smpcs) used in ~\cite{loconte2024subtractive}, and $\mathrm{PVX}_{2\pi}^{R1}$ generalizes their \snpcs from the real domain to the entire complex domain. Concretely, \smpcs are rank-one \pvcs with the phase fixed to zero and \snpcs are rank-one \pvcs with the phase fixed to values in the set $\{0, \pi \}$. We do not experiment on the latter.

% We report the obtained bpds for the different \pvc parametrizations in the first four columns of Table~\ref{tab:mnist}. In general, we see that the complex-valued parametrizations are either on par or outperform the corresponding zero-phase parametrizations.
% The notable exception is the FashionMNIST benchmark where the two parametrizations with no phase ($\mathrm{PVX}_{0}^{FR}$ and $\mathrm{PVX}_{0}^{R1}$) perform best.
% Comparing full-rank to rank-one parametrizations we see a more important impact for the zero-phase parametrization than for the complex-valued parametrization.

% Overall, $\mathrm{PVX}_{2\pi}^{FR}$ constitutes the best performing parametrization being best-in-class on all benchmarks but FashionMNIST.

% \input{tex_input/mnist_table}

% \textbf{Q2: How Do \pvcs Fair Against the State of the Art?}

% In order to compare \pvcs to state-of-the-art circuits all methods (\pvc, QPC, HCLT, SHCLT) were given a computational budget of $B=128$ components per partition (on average for SHCLT).
% The results for the competing methods were taken from the respective papers -- except for HCLTs. For HCLTs we took the bpds reported by \citet{gala2024probabilistic} as they achieved stronger results with their implementation compared to the originally reported performance~\citep{liu2021tractable}.

% We see in Table~\ref{tab:mnist} that QPCs and HCLTs are in general outperformed by \pvc -- in particular $\mathrm{PVX}_{2\pi}^{FR}$. The FashionMNIST benchmark constitutes again an outlier in this regard.
% The only methods outperforming \pvcs are the SHCLTs, which is due to them being able to dynamically allocate compute budget to informative parts of the circuit.
% It is noteworthy that for the EMNIST-BYCLASS benchmark all four  \pvc parametrization exhibit strong performance with  $\mathrm{PVX}_{2\pi}^{FR}$ and $\mathrm{PVX}_{2\pi}^{R1}$ even outcompeting SHCLTs.

% \textbf{Discussion}

% In our experimental evaluation we did not perform an explicit comparison to the method of \citet{loconte2024subtractive} as they were not able to find any improvements of allowing non-negative parameters in probabilistic circuit for discrete data. They stipulated that
% \say{simple categorical distribution can already capture any discrete distribution with finite support and a (subtractive) mixture thereof might not yield additional benefits}.
% In our experimental evaluation we refute this conjecture, and show that using complex-valued parameters leads to noticeable gains when performing density estimation on discrete data.
% The exception being of course the FashionMNIST benchmark. We believe that this is due to our optimization method that was not tuned in any way.
% In this regard we believe that developing tailor-made optimization algorithms for \pvcs might further boost their performance.
% Furthermore, one can envisage, similar to SHCLTs, dynamically adapting the compute budget per partition, which should again improve the density estimation capacities of \pvcs.

\section{Conclusions \& Future Work}
\label{sec:conclusions}

Based on first principles from quantum information theory, we constructed positive operator circuits -- a novel class of probabilistic tractable models.
A main hurdle to voercome to make \puncs practical machine learning models is to find expressive yet efficient functional forms of the bilinear forms and their parametrizations. A key challenge to overcome in this respect is the efficient parametrization of (expressive) unitary matrices as this is a notoriously difficult feat~\citep{kiani2022projunn,jing2017tunable,lezcano2019cheap,mhammedi2017efficient,wisdom2016full}.
Furthermore, we have only presented a specific functional form for the bilinear forms.

\todo{cite riguzzi quantum stuff here}

\todo{future work, are there tractable circuits for which a exponential quantum advantage exists?}

% Their construction as partition circuits allows for layer-wise parallelization and execution on modern machine learning hardware.
% Using unconstrained gradient-descent we showed that \pocs, and \pvcs specifically, constitute a promising addition to the zoo of tractable probabilistic models.

% In future work we would like to investigate in more detail theoretical properties of \pocs and how they differ from other tractable models. Ideally one would find an exponential separation between real-valued and complex-valued circuits. On the practical side it remains, however, to be seen whether such a separation has an equally important impact on real-world datasets.

% This relates to another open question,  that of learning in \pocs. We presented a rather simple learning approach for parameters tailored towards neural architecture. It might be the case that more sophisticated techniques have to be deployed for large-scale \pocs, as they might exhibit the problem of barren plateaus~\citep{ragone2023unified} -- a well known issue in quantum machine learning~\citep{biamonte2017quantum,huggins2019towards}.
% Furthermore, and as already pointed out by~\citet{loconte2024subtractive}, computing the normalization constant $Z$ requires the evaluation of the circuit in the \poc representation. This gives rise to a rather expensive cubic computation cost during learning (when compared to the quadratic cost of \pvc evaluations). Avoiding this issue would allow to drastically scale \pvcs. In order, to scale \pvcs it might also be necessary to use more sophisticated structures than binary trees with every partition having the same number of components $B$ per partition, \cf SHCLTs.

### 2025-06-02 202507_uai_puncs/icml2025/noise_puncstuff/appendix.tex
added uai

\section*{\LARGE Appendix}

\section{Proofs}

\subsection{Proof of Proposition \ref{prop:povmprob}}
\label{sec:proof:prop:povmprob}

\proppovmprob*

\begin{proof}
	While this is a well-known result we were not able to identify a concise proof in the literature and provide therefore, for the sake of completeness here. To this end, we
	first show that $p(i)\geq 1$, for each $i$.
	\begin{align}
		p(i)
		=
		\Tr [ E(i) \rho ]
		% \nonumber
		% \\
		 &
		=
		\Tr [ D(i) D^*(i)  \rho ]
		% \nonumber
		\\
		 &
		=
		\Tr [ D(i)^*  \rho D(i) ]. \nonumber
	\end{align}
	Here we used the fact that $E(i)$ is PSD and factorized it into the product $D(i) D^*(i)$. Then we used the fact that the trace is invariant under cyclical shifts.
	Clearly, the matrix $ D(i)^*  \rho D(i)$ is PSD as we have for every vector $x$:
	\begin{align}
		x^* D(i)^*  \rho D(i) x = y^* \rho y \geq 0 \quad \text{with } y = Dx.
		\label{eq:def:psdPOVM}
	\end{align}
	As the trace of a PSD matrix is positive we have that $p(i)$ is a positive real number, \ie $p(i){\geq} 0$ for every $i$.

	Secondly, we show that $p(i)$ is normalized:
	\begin{align}
		\sum_{i=1}^N p(i)
		%  &
		=
		\sum_{i=1}^N \Tr [ E(i) \rho ]
		% \nonumber
		% \\
		%  &
		=
		\Tr [ \sum_{i=1}^N E(i) \rho ]
		% \nonumber
		% \\
		%  &
		{=}
		\Tr [  \rho ],
	\end{align}
	where we used Equation~\ref{eq:povm_normalized}.
	Exploiting the fact that the trace of a density matrix is one gives us indeed $\sum_{i=1}^N p(i)=1$, and we can conclude that $p(i)$ is a valid probability distribution.
\end{proof}

\subsection{Proof of Proposition \ref{prop:pocpsd}}
\label{sec:proof:prop:pocpsd}

\proppocpsd*

\begin{proof}
	This can be seen by first inspecting the leaves where we have $A_{k} \times  e_{\xvar_k} \otimes  e^*_{\xvar_k} \times A_{k}^*$. Using a similar argument as made in Equation~\ref{eq:def:psdPOVM} we know that all the leaves carry PSD matrices. Passing these on recursively to the bilinear forms in the internal units retains the positive semi-defiteness (imposed by definition). We conclude that computation units in of positive operator circuit output PSD matrices.
\end{proof}

\subsection{Proof of Proposition \ref{prop:puncPOVM}}
\label{sec:proof:prop:puncPOVM}

\proppuncPOVM*

\begin{proof}
	Given that positive unitary circuits are by definition positive operator circuits we already have that:
	\begin{align}
		\forall \xvars \in \Omega(\Xvars): \punccircuit(\xvars) \text{ is PSD}.
	\end{align}
	Next we show that $\sum_{\xvars\in \Omega(\Xvars)} \punccircuit(\xvars)=\mathbb{1}$. Here we observe that in the computation units we have:
	\begin{align}
		\sum_{\xvars_k \in \Omega(\Xvars_k)} \punccircuit_k(\xvars_k)
		 & = \sum_{\xvars_{k_l}} \sum_{\xvars_{k_r}}    \bilinear(\punccircuit_k(\xvars_{k_l}), \punccircuit_k(\xvars_{k_r})  )
		\\
		 & =
		\bilinear \left(
		\sum_{\xvars_{k_l}}   \punccircuit_k(\xvars_{k_l}),
		\sum_{\xvars_{k_r}}\punccircuit_k(\xvars_{k_r})
		\nonumber
		\right)
	\end{align}
	This lets us push down the summation of a specific variable to the corresponding leaf where the variable is given as input, where we then have:
	\begin{align}
		\sum_{\xvars_k \in \Omega(\Xvars_k)} \punccircuit_k(\xvars_k)
		 & = \sum_{x_k\in \Omega(X_k)} A_k e_{x_k} \otimes e^*_{x_k} A_k^*
		\nonumber
		\\
		 & = A_k \left( \sum_{x_k\in \Omega(X_k)}  e_{x_k} \otimes e^*_{x_k} \right) A_k^*
		\nonumber
		\\
		 & = 		A_k \mathbb{1_{\Omega(X_k)}} A_k^* = \mathbb{1_k}
	\end{align}
	We now exploit that the bilinear forms in a positive unitary circuit are unit preserving (\cf Definition~\ref{def:bilinear_unitpreserving}), which gives us indeed
	$
		\sum_{\xvars\in \Omega(\Xvars)} \punccircuit(\xvars)=\mathbb{1}.
	$
\end{proof}

\subsection{Proof of Proposition \ref{prop:noisypovmdist}}
\label{sec:proof:prop:noisypovmdist}

\propnoisypovmdist*

\begin{proof}
	Let $\rho$ be a density matrix and $\{E(i)\}_{i=0}^{I-1} $ be a POVM such that $\sum_{i=0}^{I-1} E(i)=M$. We then have
	\begin{align}
		\Tr[(\mathbb{1}-M) \rho]\geq 0,
	\end{align}
	holds as $\mathbb{1}-M$ is PSD by definition.
	Pushing the trace over the subtraction and rearranging terms yields:
	\begin{align}
		\Tr[\mathbb{1} \rho]- 	\Tr[M \rho]\geq 0 \Leftrightarrow \Tr[M \rho]\leq 1.
	\end{align}
	This means that the induced probability distribution $p(i)= \Tr[E(i)\rho]$ is indeed sub-complete.
\end{proof}

\subsection{Proof of Proposition \ref{prop:noisepunddist}}
\label{sec:proof:prop:noisepunddist}

\propnoisepunddist*

\begin{proof}
	Let $\Xvars$ be a set of random variables and  $\Qcircuit(\xvars)= \qcircuit(\xvars)\punccircuit(\xvars)$ with $\xvars\in \Omega(\Xvars)$ be a \noisepunc inducing a probability distribution
	\begin{align}
		\pi_\Xvars(\xvars) = \Tr [ \qcircuit(\xvars)\punccircuit(\xvars) \rho].
	\end{align}
	As $\qcircuit(\xvars)$ is a scalar we can write
	\begin{align}
		\pi_\Xvars(\xvars) = \qcircuit(\xvars) \Tr [ \punccircuit(\xvars) \rho] = \qcircuit(\xvars) p_X(\xvars),
	\end{align}
	where $p_X(\xvars)$ is the (complete) probability distribution induced by the set $\{ \punccircuit(\xvars) \}_{\xvars \in \Omega(\Xvars)}$. As $0\leq q(\xvars)\leq 1$ (by definition) and $0\leq p_\Xvars(\xvars)\leq 1$ we have also that $0\leq \pi_\Xvars(\xvars)\leq 1$. This shows that each event $\xvars\in \Omega(\Xvars)$ has a positive probability.

	For the sub-completeness we observe that the sum of positive terms
	\begin{align}
		\sum_{\xvars \in \Omega(\Xvars)} p_\Xvars(\xvars)=1.
	\end{align}
	If we weigh each term in the sum with a scalar $q(\xvars) \in [0,1]$ we immediately conclude that $\sum_{\xvars \in \Omega(\Xvars)} \pi_\Xvars (\xvars)\leq$ and that we have indeed a sub-complete probability distribution.
\end{proof}

\subsection{Proof of Proposition \ref{prop:softcountingcondition}}
\label{sec:proof:prop:softcountingcondition}

\propsoftcountingcondition*

\begin{proof}
	We consider the two extreme cases where on teh one hand the $W_k$ in the leaves have only zeros as entries and on the other hand only ones. For the first case it is easy to see that multiplying any vector $\text{one\_hot}(\xvar_k)$ with a zero matrix results in a zero vector $\qcircuit_k(\xvar_k)$. Evaluting the circuit with these zero vector from the leaves trivially gives us $\circuit_{root}(\xvars_{root})=0$.

	For the second case we observe that multiplying any vector $\text{one\_hot}(\xvar_k)$ with a matrix having only ones as entries results in a vector $\qcircuit_k(\xvar_k)$ having only ones as entries. As the matrices $W_k$ in the internal nodes and the root node are row-normalized this is preserved throughout the circuit and we have $\circuit_{root}(\xvars_{root})=1$.

	Finally, as soft counting circuits are monotone functions all other cases fall between these two extremes and we conclude that in general $0\leq \qcircuit(\xvars)\leq 1$.
\end{proof}

\subsection{Proof of Proposition \ref{prop:computationalcomplexity}}
\label{sec:proof:prop:computationalcomplexity}

\propcomputationalcomplexity*

\begin{proof}
	We determine the computational cost of evaluating the circuit by determining the cost of the individual computation units. We start at the leaves where we first have a Kronecker product $ e_{\xvar_k} \otimes  e_{\xvar_k}^*$, which can be performed in $\bigO (\samplespacesize^2)$.
	The matrix products with $A_k$ and $A_k^*$ can then be performed in $\bigO (\numbond^2 \samplespacesize)$, which implies $\bigO(\bigO (\samplespacesize^2))$ if $B\leq N$.
	In the bilinear form (\cf Equation~\ref{eq:polyadicbilinear}) we perform four matrix-matrix product in $\bigO (\numbond^3)$. This is followed by a Hadamard product in $\bigO (\numbond^2)$,
	This means that we have for the individual computation units in the partition a circuit a cost of $\bigO (\samplespacesize^3)$.

	It is also easy to show that for binary tree partition circuits we have a total of $\bigO (\numvar)$ computation units. This results in a computational complexity of $\bigO (\numvar  \samplespacesize^3)$.

	We note the computation of trace of the matrix-matrix multiplication between $\punccircuit_{root}(\xvars)$ and the density matrix $\rho$ necessary for obtaining a probability can be done in $\bigO(\numbond^2)$.

	Given that we can evaluate \puncs $\bigO ( \numvar \samplespacesize^3)$ and also that we can marginalize out variables by pushing the summation to the leaves (\cf the proof of Proposition~\ref{prop:puncPOVM}), we can conclude that marginal inference has complexity $\bigO(\numvar \samplespacesize^3)$, as well.
\end{proof}

\subsection{Proof of Proposition \ref{prop:voequiprob}}
\label{sec:proof:prop:voequiprob}

\begin{restatable}{lemma}{proppOveq}
	For the circuits defined in Definition~\ref{def:poc} and Definition~\ref{def:vpoc} it holds that
	\label{prop:Oveq}
	\begin{align}
		\forall k: \Ocircuit_k(\xvars_k) = \vcircuit_{k}(\xvars_{k}) \otimes \vcircuit^*_{k}(\xvars_{k}).
		\label{eq:def:opvec_equivalent}
	\end{align}
\end{restatable}

\begin{proof}
	We start the proof at the leave where we have
	\begin{align}
		\Ocircuit_k(\xvar_k)
		=
		A_k e_{\xvar_k} \otimes  e_{\xvar_k}^* A_k^*
	\end{align}
	and Equation~\ref{eq:def:opvec_equivalent} holds by definition. In the computation untis into which the leaves feed, we then have
	\begin{align}
		\Ocircuit_k
		 & = L_k \Ocircuit_{k_l} L_k^* \odot R_k \Ocircuit_{k_r} R_k^*
		\nonumber
		\\
		 & = L_k (\vcircuit_{k_l} \otimes \vcircuit^*_{k_l} ) L_k^*
		\odot
		R_k (\vcircuit_{k_r} \otimes \vcircuit^*_{k_r} ) R_k^*,
	\end{align}
	where we omited the explicit dependencies on the variables $\xvar_k$, $\xvar_{k_l}$, and $\xvar_{k_r}$. Next we use the mixed-product property of the Hadamard product and Kronecker product to get
	\begin{align}
		\Ocircuit_k
		 & =
		L_k \vcircuit_{k_l} \odot  R_k \vcircuit_{k_r}
		\otimes
		\vcircuit^*_{k_l} L_k^* \odot \vcircuit^*_{k_r} R_k^*
		\nonumber
		\\
		 & = v_k \otimes v_k^*.
	\end{align}
	Repeating this argument recursively until the root of the circuit concludes the proof.
\end{proof}

\propvoequiprob*

\begin{proof}
	The proof starts by simply plugging in the vector representation of $\Ocircuit_{\text{root}}$ into the expression $\Tr [\Ocircuit_{\text{root}} \rho]$ and rather straightforward.
	\begin{align}
		\Tr [\Ocircuit_{\text{root}} \rho]
		 &
		=
		\Tr \left[
			\Bigl(\vcircuit_{\text{root}} \otimes \vcircuit^*_{\text{root}}   \Bigr) \times \rho
			\right]
		\\
		 & = \vcircuit^*_{\text{root}} \times  \rho \times\vcircuit_{\text{root}}
		\nonumber
		\\
		 & = \vcircuit^*_{\text{root}} \times  \gamma^* \times \gamma \times \vcircuit_{\text{root}}
		\nonumber
		% \\
		% \nonumber
		%  & 
		= \lVert \gamma \times \vcircuit_{\text{root}}\rVert^2
	\end{align}

\end{proof}

\subsection{Proof of Proposition \ref{prop:computationalcomplexityvec}}
\label{sec:proof:prop:computationalcomplexityvec}

\propcomputationalcomplexityvec*

\begin{proof}
	Following a similar train of thought as the proof in Section~\ref{sec:proof:prop:computationalcomplexity} we can derive the computational cost of evaluating a positive vector circuit to be $\bigO (\numvar \numbond \samplespacesize)$ implying
	$\bigO (\numvar  \samplespacesize^2)$
	when $\numbond \leq \samplespacesize$.
	The major difference between matrix representation of operator circuits and their vector representation is that for the former the internal computation units output matrices ($\Ocircuit_k(\xvars_k)\in \mathbb{C}^{\numbond \times \numbond}$) while for the latter vectors are outputted ($\vcircuit_k(\xvars_k)\in \mathbb{C}^{\numbond}$).

	As $\vcircuit_{root}(\Xvar_{rooot})$ is computable in $\bigO (\numvar \numbond \samplespacesize)$ and $\lVert \gamma \times \vcircuit_{\text{root}}\rVert^2$ is computable in $\bigO(\numbond^2)$, we can conclude that the probability $p_\Xvars(\xvars)$ can be computed in $\bigO (\numvar \samplespacesize^2)$.
\end{proof}

\section{Computing the Normalization for \noisepuncs}
\label{sec:noisepuncnormalization}

As we dscribe in Equation~\ref{eq:deepoperatirmix} multiplying a soft counting circuit $\qcircuit(\xvars)$ with a \punc $\punccircuit(\xvars)$ gives us in the internal nodes a mixture of operator:

\begin{align}
	\Qcircuit_{ki}(\xvars) = \sum_j w_{kij}  \tildeQcircuit_{kj},
\end{align}
with $ \tildeQcircuit_{kj} = \bilinear_k(  \qcircuit_{k_l j} \punccircuit_{k_l},  \qcircuit_{k_r j} \punccircuit_{k_r} )$. For the specific case of polyadic decompositions in the bilinear form we get:
\begin{align}
	\tildeQcircuit_{kj}
	=
	\Bigl(
	D_k^{\nicefrac{1}{2}} U_{k} \times  \qcircuit_{k_l j} \punccircuit_{k_l} \times U_{k}^* D_k^{\nicefrac{1}{2}}
	\Bigr)
	\odot
	\Bigl(
	D_k^{-\nicefrac{1}{2}} V_{k} \times \qcircuit_{k_r j} \punccircuit_{k_r} \times  V_{k}^* D_k^{-\nicefrac{1}{2}}
	\Bigr).
\end{align}
If we now apply Equation~\ref{eq:deepoperatirmix} recursively, we end up in the leaves. Here we first write the matrix vector multiplications in the leaves of the soft counting circuit element-wise:
\begin{align}
	                & \
	\qcircuit_{k}(\xvar_k) = W_k \times  \text{one\_hot}(\xvar_k)
	\nonumber
	\\
	\Leftrightarrow & \
	\qcircuit_{ki}(\xvar_k) = \sum_j w_{kij} \iverson{\xvar_k=j}
\end{align}
where we write the dependency on the $\xvar_k$ variable again explicitly. This gives us for the leaves of a \noisepunc:
\begin{align}
	\Qcircuit_{ki}(\xvar_k)
	 & = \qcircuit_{ki}(\xvar_k)  \punccircuit_{k}(\xvar_k),
	\nonumber
	\\
	 & = \left( \sum_j w_{kij} \iverson{\xvar_k=j} \right)
	A_k \left( e_{x_k} \otimes e_{x_k}^* \right) A_k^*
	\nonumber
	\\
	 & = w_{ki x_k} A_k \left( e_{x_k} \otimes e_{x_k}^* \right) A_k^*
\end{align}
As taking the product of two circuit with the identical partition tree results again in a circuit with the same partition tree, we can again push the summation over the variables $\xvars\in \Omega(\Xvars)$ to the respective leaves, analogous to the proof in Appendix~\ref{sec:proof:prop:puncPOVM}.

\begin{align}
	\sum_{\xvar_k \in \Omega(\Xvar_k)}\Qcircuit_{ki}(\xvar_k)
	=
	\sum_{\xvar_k \in \Omega(\Xvar_k)} \qcircuit_{ki}(\xvar_k) \punccircuit_k(\xvar_k)
	 & =
	\sum_{\xvar_k \in \Omega(\Xvar_k) }w_{ki x_k} A_k \left( e_{x_k} \otimes e_{x_k}^* \right)A_k^*
	\nonumber
	\\
	 & =
	A_k \left( \sum_{\xvar_k \in \Omega(\Xvar_k) }w_{ki x_k}  \left( e_{x_k} \otimes e_{x_k}^* \right) \right) A_k^*
	\nonumber
	\\
	 & =
	A_k \times
	\begin{pmatrix}
		w_{ki0} & 0       & \cdots & 0                          \\
		0       & w_{ki1} & \cdots & 0                          \\
		\vdots  & \vdots  & \ddots & \vdots                     \\
		0       & 0       & \cdots & w_{ki(\samplespacesize-1)} \\
	\end{pmatrix}
	\times  A_k^*
	\nonumber
	\\
	 & =
	A_k \times  \diag(w_{ki}) \times  A_k^*
\end{align}
We can now also see that when we have in the leaves that $W_{kij}=1$ for every $k$, $i$, and $j$ that we recover the (non-noisy) \puncs as a special case.

For computing the normalization constant of a \noisepunc this means that we first compute in the leaves the prouct of $ \diag(w_{ki})$ sandwidched between $A_k$ and $A_k^*$. For each of the leaves we have $\numcomponents$ of these products. Hence, the computation of a single leaf can be done in $\bigO (\numcomponents \numbond^2 \samplespacesize)$. Using agina that $\numbond\leq \samplespacesize$ we get $\bigO (\numcomponents \samplespacesize^3)$.

In the computation units we then compute $C$ times a bilinear form giving us $\bigO(\numcomponents \samplespacesize^3)$. This is followed by the computation of a mixture of operators for which we need $ \bigO(\numcomponents^2 \samplespacesize^2)$.
Given that there are $\bigO(\numvar)$ computation units we get that the computation of the normalization constant takes $\big(\numvar \numcomponents \samplespacesize^3 + \numcomponents^2 \samplespacesize^2)$. The advantage of \noisepuncs over other circuit architectures~\cite{loconte2024sum,wangrelationship} is that this normalization constant had to be computed once training has finished and does not need to be differentiated through.

### 2025-06-02 202507_uai_puncs/icml2025/noise_puncstuff/content.tex
added uai

\begin{abstract}
	By recursively nesting sums and products, probabilistic circuits (PCs) have emerged in recent years as an attractive class of generative models as they enjoy, for instance, polytime marginalization of random variables.
	However, as these PCs form monotone functions, specifically the parameters are constrained to positive real values, their expressive power is limited.
	Drawing inspiration quantum information theory we introduce \textit{positive unitary circuits} (\puncs),
	which generalize circuit evaluations over positive real-valued probabilities to circuit evaluation over positive semidefinite matrices.
	A key feature of \puncs is that they form normalized probability distributions by construction, which avoids the computation of a spurious normalization constant.
	Motivated by noisy quantum systems we generalize \puncs to \noisepuncs, which can be interpreted as deep mixtures of \puncs.
	Lastly, we provide an efficient parametrization of \puncs that enables us to perform parameter learning on hardware accelerators using standard and unconstrained gradient-based optimization.
\end{abstract}

\section{Introduction}

Probabilistic circuits (PCs)~\citep{darwiche2003differential,poon2011sum} belong to an unusual class of probabilistic models: they are highly expressive but at the same time also tractable.
For instance, so-called decomposable probabilistic circuits~\citep{darwiche2001decomposable} allow for the computation of marginals in time polynomial in the size of the circuit.
\citet{zhang2020relationship} noted that it is exactly this restriction to positive values that limits the expressive efficiency (or succinctness) of PCs~\citep{martens2014expressive,decolnet2021compilation}. In particular, the positivity constraint on the set of elements that PCs operate on prevents them from modelling negative correlations between variables.
Circuits that are incapable of modelling negative correlations, \ie circuits that can only combine probabilities in an additive fashion, are also called monotone circuits~\citep{shpilka2010arithmetic}.
This restricted expressiveness can be combatted by the use of so-called \textit{non-monotone} circuits, where subtractions are allowed as a third operation (besides sums and products). Interestingly, \citet{valiant1979negation} showed that a mere single subtraction can render non-monotone circuits exponentially more expressive than monotone circuits a result that has recently been refined for decomposable circuits~\citep{loconte2024sum}.

As shown in \citep{harviainen2023inference,agarwalprobabilistic}, non-monotone circuits do, however, introduce an important complication: if non-monotone circuits are not designed carefully, verifying whether a circuit encodes a valid probability distribution or not is an NP-hard problem. This does also render learning the parameters of a circuit practically infeasible.

Using the concept of \textit{positive operator valued measures} from quantum information theory, which encode random event as positive semidefinite matrices we are able to construct non-monotone circuits that nonetheless encode proper (normalized) probability distributions by construction.
Our work falls into a line of recent works presented in the circuit literature ~\citep{sladek2023encoding,loconte2024subtractive,wangrelationship,loconte2024sum}. However, our work is the first that establishes this deep connection between concepts in quantum information theory and tractable probabilistic models. We make three concrete contributions
\begin{enumerate*}
	\item We generalize probabilistic circuits over scalars to operator-valued circuits and introduce \puncs (Section~\ref{sec:puncs}).
	\item We use the concept of noise in quantum information theory to construct deep operator mixture circuits, dubbed \noisepuncs (Section~\ref{sec:noisecircuits}).
	\item We provide an efficient parametrization such that the constructed probability distributions using \puncs are normalized by construction (Section~\ref{sec:efficient_param}).
\end{enumerate*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{A Primer on Quantum Information Theory}
\label{sec:qit}

A widely used and elegant framework to describe measurements of quantum systems is the so-called \textit{positive operator-valued measure} (POVM) formalism. While POVMs have physical interpretations in terms of quantum information and quantum statistics, we will only be interested in their mathematical prperties as we use them to show that  circuits (defined in Section~\ref{sec:puncs}) form valid probability distributions.
We refer the reader to \citep{nielsen2001quantum} for an in-depth exposition on the topic, as well as quantum computing and quantum information theory in general.
\begin{definition}[Positive Semidefinite]
	An $\numbond {\times} \numbond$ Hermitian matrix $H$ is called positive semidefinite (PSD) if and only if $\forall \xvars {\in}  \mathbb {C} ^{\numbond}: \xvars^{*} H \xvars {\geq} 0$, where $\xvars^{*}$ denotes the conjugate transpose and $\mathbb {C}^{\numbond}$ the $\numbond$-dimensional space of complex numbers.
\end{definition}
\begin{definition}[{POVM~\citep[Page 90]{nielsen2001quantum}}]
	\label{def:povm}
	A positive operator-valued measure
	% with a finite number of elements acting on a finite-dimensional Hilbert space $\mathcal{H}$,
	is a set of PSD  matrices $\{E(i)\}_{i=0}^{\numevents-1}$ ($I$ being the number of possible measurement outcomes) that sum to the identity:
	\begin{talign}
		\sum_{i=0}^{\numevents-1} E(i) = \mathbb{1},
		\label{eq:povm_normalized}
	\end{talign}
\end{definition}
Before defining the probability of a specific $i$ occurring, we need the notion of a density matrix~\citep{neumann1927wahrscheinlich}:
\begin{definition}[{Density Matrix~\citep[Page 102]{nielsen2001quantum}}]
	\label{def:density_matrix}
	A density matrix $\rho$ is a PSD matrix of trace one, \ie $\Tr [\rho]=1$.
\end{definition}
\begin{definition}[{Event Probability~\citep[Page 102]{nielsen2001quantum}}]
	\label{def:eventprob}
	Let $\rho$ be a density matrix and let $i$ denote an event with $E(i)$ being the corresponding element from the POVM. The probability of the event $i$ happening, \ie measuring the outcome $i$, is given by
	\begin{talign}
		p(i) = \Tr [ \rho E(i)]
		\label{eq:povm_prob}
	\end{talign}
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{restatable}{proposition}{proppovmprob}
	\label{prop:povmprob}
	The expression in Equation~\ref{eq:povm_prob} defines a valid probability distribution.
\end{restatable}

\begin{proof}
	While this is a well-known result we were not able to identify a concise proof in the literature. We therefore provide one in Appendix~\ref{sec:proof:prop:povmprob}.
\end{proof}

\todo{talk about Loewner order}

\citep[Corollary 4.2]{baksalary1989some}

\section{Positive Unitary Circuits}
\label{sec:puncs}

\begin{figure}[t]

	\centering
	\input{tex_input/partition}

	\caption{
		Partition circuit over four binary variables $x_i$ with $i \in \{0,1,2,3\}$, which are given as inputs to the circuit. The internal nodes of the partition circuit correspond the computation units.}
	\label{fig:circuit}
\end{figure}

A popular subclass of probabilistic circuits are so-called structured decomposable  probabilistic circuits~\citep{darwiche2011sdd} that are also smooth~\citep{darwiche2001tractable}. The advantage of this circuit subclass is that they can be implemented in a rather straightforward fashion on modern AI accelerators, as demonstrated by \citet{peharz2019random,peharz2020einsum}.
\todo{cite LSE paper}
\citet{zuidberg2024probabilistic} introduced an abstraction for these smooth structured decomposable circuit dubbed \textit{partition circuits}.
We give such a circuit in Figure~\ref{fig:circuit}.

\todo{cite lse paper}
\begin{definition}[Partition Circuit~\citep{LSE}]
	\label{def:partition_circuit}
	A partition circuit over a set of variables is a parametrized computation graph taking the form of a binary tree. The partition circuit consists of two kinds of computation units:
	\textit{leaf} and \textit{internal} units, as well as a single \textit{root}.
	Units at the same distance from the root form a layer.
	Furthermore, let $\circuit_k$ denote the root unit or an internal unit. The unit $\circuit_k$ then receives its inputs from two units in the previous layer, which we denote by $\circuit_{k_l}$ and $\circuit_{k_r}$. Each computation unit is input to exactly one other unit, except the root unit, which is the input to no other unit.
\end{definition}

\subsection{Operator Circuits}

Using the concept of partition circuits we now generalize the probabilistic circuits to (unnormalized) positive operator circuits. Positive operator circuits can be thought of as generalizing circuit evaluations with probabilities to circuit evaluations with PSD matrices. Note, in the definition below we use $\Ocircuit_k$ instead of $\circuit_k$ to make this generalization explicit.
\begin{definition}[Operator Circuit]
	\label{def:poc}
	Let   $\xvars{=}\{\xvar_0,\dots ,\xvar_{\numvar{-}1}  \}$ be a set of $M$ categorical variables with domains of size $\samplespacesize$.
	We define an operator circuit as a partition circuit whose computation units take the following functional form:
	\begin{align}
		 & {\Ocircuit}_k(\xvars_k){=}
		\begin{cases}
			A_{k} \times  e_{\xvar_k} \otimes  e^*_{\xvar_k} \times A_{k}^*,
			 & \text{if $k$ is leaf}
			\\
			\bilinear_k \Bigl(\Ocircuit_{k_l}(\xvars_{k_l}), \Ocircuit_{k_l}(\xvars_{k_l}) \Bigr)
			 & \text{else}
		\end{cases}
		\label{eq:poc:def}
	\end{align}
	Here we use the symbol $\otimes$ to denote the Kronecker product.
	Furthermore, $\{e_{\xvar_k} \}_{n=0}^{\samplespacesize-1}$ is a set of $\samplespacesize$-dimensinoal orthonormal basis vectors associating each of the $\samplespacesize$ possible values for $\xvar_k$ to a specific basis vector. The functions $\bilinear_k(\cdot, \cdot)$ denote bilinear forms.
\end{definition}

\begin{definition}
	We call an operator circuit \textit{positive} if we have that the bilinear forms $\bilinear_k$ are such that
	\begin{align}
		\bilinear_k \Bigl(\Ocircuit_{k_l}(\xvars_{k_l}), \Ocircuit_{k_l}(\xvars_{k_l}) \Bigr) \quad \text{is a PSD matrix}
		\label{def:eq:bilinear_psd}
	\end{align}
	if $\Ocircuit_{k_l}(\xvars_{k_l})$ and $\Ocircuit_{k_l}(\xvars_{k_l})$ are PSD.
\end{definition}

\begin{restatable}{proposition}{proppocpsd}
	\label{prop:pocpsd}
	Positive operator circuits are PSD.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:pocpsd}.
\end{proof}

\subsection{Constructing a Probability Distribution}

In Section~\ref{sec:qit} we saw that we can construct a probability distribution using a density matrix $\rho$ and a positive operator-valued measure, with the latter being a set of PSD matrices (\cf Definition~\ref{def:povm}) that sum to the unit matrix. Using a positive operator circuit $\Ocircuit(\xvars)$ we indeed have a set of PSD matrices. Namely, on for each instantiation of the $\xvars$ variables. We now introduce \textit{positive unit preserving operator circuits} for which also the summation to the unit matrix holds.

\begin{definition}
	\label{def:bilinear_unitpreserving}
	We call a positive operator circuit \textit{unit preserving} if we have that the bilinear forms $\bilinear_k$ are such that
	\begin{align}
		\bilinear_k(\mathbb{1}_{k_l}, \mathbb{1}_{k_r})= \mathbb{1}_k,
		\label{def:eq:bilinear_unitpreserving}
	\end{align}
	where $\mathbb{1}_{k}$, $\mathbb{1}_{k_l}$ , and $\mathbb{1}_{k_r}$  denote unit matrices of appropriate size, and if $A_{k}A_{k}^{*}= \mathbb{1}_{k}$ in the leaves. That is, the $A_{k}$ are semi-unitary matrices.
\end{definition}
We will also refer to positive unit preserving operator circuits as positive unitary circuits, or \puncs.

\begin{restatable}{proposition}{proppuncPOVM}
	\label{prop:puncPOVM}
	Let $\Xvars$ denote a set of random variables with sample space $\Omega(\Xvars)$.
	Then the set $\{ \punccircuit(\xvars)\}_{\xvars \in \Omega(\Xvars)}$ of positive unitary circuis forms a POVM.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:puncPOVM}
\end{proof}

\begin{restatable}{theorem}{theopuncprobdist}
	\label{theo:puncprobdist}
	Let $\rho$ be a density matrix and $\punccircuit(\xvars)$ a positive unitary circuit. The function
	\begin{align}
		p_\Xvars(\xvars) = \Tr [\punccircuit(\xvars) \rho]
		\label{eq:theo:prob_operator}
	\end{align}
	is a proper probability distribution over the random variables $\Xvars$ with sample space $\Omega(\Xvars)$.
\end{restatable}

\begin{proof}
	This follows from Proposition~\ref{prop:povmprob} and Proposition~\ref{prop:puncPOVM}
\end{proof}

\section{Quantum Noise and Circuits}
\label{sec:noisecircuits}

\subsection{Sub-Complete Probability Distributions}

An important concept in quantum information theory that generalizes the standard POVM (\cf Definition~\ref{def:povm}) is the so-called noisy POVM, which enables for instance modelling imperfect measurement devices. In the context of machine learning this might be an improperly labeled data point of an error in the detector, \eg a pixel flip in the camera that took a picture.
\begin{definition}
	\label{def:noisypovm}
	A noisy positive operator-valued measure
	is a set of PSD  matrices $\{E(i)\}_{i=0}^{I-1}$ ($\numevents$ being the number of possible measurement outcomes) such that:
	\begin{talign}
		\sum_{i=0}^{V-1} E(i) = \subcompletemeasrure < \mathbb{1},
	\end{talign}
	where the inequality sign is interpreted using the Loewner ordering of PSD matrices,
	\ie $\subcompletemeasrure < \mathbb{1} \Leftrightarrow\mathbb{1}-\subcompletemeasrure$ is PSD.
\end{definition}

\begin{definition}[Sub-complete Probability Distribution]
	Let $\Xvars$ be a set of random variables with sample space $\Omega(\Xvars)$. We call a probability distribution \textit{sub-complete} if $\forall \xvars \in \Omega(\Xvars): p(\xvars)\geq 0$ and if $\sum{\xvars \in \Omega(\Xvars)} p(\xvars)\leq 1$. We call a probability distribution \textit{striclty sub-complete} if the latter inequality holds strictly and \textit{complete} if equality holds.
\end{definition}

\begin{restatable}{proposition}{propnoisypovmdist}
	\label{prop:noisypovmdist}
	Noisy POVMs induce a sub-complete probability distributions.
\end{restatable}

\begin{proof}
	See Appendix \ref{prop:noisypovmdist}
\end{proof}

To construct sub-complete probability distributions using operator circuits we now introduce \textit{noisy positive unit preserving operator circuits} or \noisepuncs.

\begin{definition}
	\label{def:noisepunc}
	A \noisepunc $\Qcircuit(\xvars)$ is of the form
	\begin{align}
		\Qcircuit(\xvars)= \qcircuit(\xvars) \punccircuit(\xvars),
	\end{align}
	where $\punccircuit(\xvars)$ is a positive unit preserving operator circuit and $\qcircuit(\xvars)$ is real-valued and belongs to the  $[0,1]$ interval for every $\xvars$.
\end{definition}

\begin{restatable}{proposition}{propnoisepunddist}
	\label{prop:noisepunddist}
	\noisepuncs induce sub-complete probability distributions.
\end{restatable}

\begin{proof}
	See Appendix \ref{sec:proof:prop:noisepunddist}
\end{proof}

\subsection{Noise Variables}

All of this is a bit unnerving. How is it possible that we have probabilistic events $\xvars\in \Omega(\Xvars)$ that violate the second of Kolmogorov's probability axioms (by having sub-complete distributions)? The resolution to this problem lies in interpreting sub-complete distributions as joint distributions not only over the set of variables $\Xvars$ but an extra set of variables $\Yvars$ representing extra noise~\citep{wiseman2009quantum}
\begin{align}
	\pi_{\Xvars}(\xvars) = p_{\Xvars \Yvars}(\xvars, \yvars),
\end{align}
for which we indeed have
\begin{talign}
	\sum_{\xvars\in \Omega(\Xvars)} \sum_{\yvars\in \Omega(\Yvars)}  p_{\Xvars \Yvars}(\xvars, \yvars) =1.
\end{talign}
The problem with the random variables $\Yvars$ is that they are not accessible to the observer. In the sense that they are observable in principle but in practice not, \eg a defect in a sensor.\footnote{Note these practically unobservable variables should not be confused with the concept of (local) hidden variables, which lead to violations of Bell's inequality.}
However, the probability we are actually interested in is the one for $\Xvars$ given $\Yvars$
\begin{align}
	p_{\Xvars|\Yvars}(\xvars\mid \yvars )
	 & =
	\nicefrac{
	p_{\Xvars \Yvars}(\xvars,  \yvars )
	}
	{
	\sum_{\xvars\in \Omega(\Xvars)}p_{\Xvars \Yvars}(\xvars, \yvars )
	}
	\nonumber
	\\
	 & = \nicefrac{\pi_{\Xvars}(\xvars)}{\sum_{\xvars\in \Omega(\Xvars)}  \pi_{\Xvars}(\xvars)}
\end{align}
The question now became whether we can choose $\qcircuit(\xvars)$ such that the summation in the denominator is tractable, \ie can be performed in time polynomial in the number of random variables $\Xvars$.

\subsection{Soft Counting Circuits}

\todo{use probabislitc circuits instead of soft counting seems to work better}

In the probabilistic circuit literature in has been shown that the product of two circuits is representable as a single circuit within polytime and polyspace if the two circuits are compatible~\citep{khosravi2019tractable,vergari2021compositional}. Importantly, this single circuit then allows for summing out the variables $\xvars$.
We will now show that this also holds when the $\qcircuit(\xvars)$ is a circuit and $\punccircuit(\xvars)$ is a \punc, and both share the same partition tree. To this end we also introduce a new circuit type that satisfies the $0\leq \qcircuit(\xvars)\leq 1$ constraint imposed in Definition~\ref{def:noisepunc}.

\begin{definition}[Soft Counting Circuit]
	\label{def:lpc}
	Let   $\xvars{=}\{\xvar_0,\dots ,\xvar_{\numvar{-}1}  \}$ be a set of $\numvar$ categorical variables with domains of size $\samplespacesize$.
	We define an operator circuit as a partition circuit whose computation units take the following functional form:
	\begin{align}
		{\qcircuit}_k(\xvars_k)=
		\begin{cases}
			W_k \times  \text{one\_hot}(\xvar_k),
			 & \text{if $k$ leaf}
			\\
			W_k \times \Bigl( \qcircuit_{k_l}(\xvars_{k_l}) \odot  \qcircuit_{k_r} (\xvars_{k_r}) \Bigr)
			 & \text{else}.
		\end{cases}
		\nonumber
	\end{align}
	Here we use the symbols $\times$ and $\odot$ to denote the matrix product and Hadamard product, respectively.
	In the leaves we have $W_k\in \mathbb{R}_{\geq 0}^{\numcomponents \times \samplespacesize}$ with $\numcomponents$ being a positive integer determining the capacity of the circuit. We also require the entries of $W_k$ in the leaves to be upper bounded by $1$.
	For the $W_k$ in the internal units and the root unit we require their entries to be positive reals and that they are row-normalized. That is, $\forall k,i: \sum_j W_{kij}=1$, where the $i$ and $j$ indices index the matrix. For internal units we have $W_k\in \mathbb{R}_{\geq 0}^{\numcomponents \times \numcomponents}$ and for the root we have $W_{root} \in \mathbb{R}_{\geq 0}^{1 \times \numcomponents}$.
\end{definition}

For the sake of simplicity we have chosen here a canonical polyadic form~\citep{carroll1970analysis} for the internal units by using a Hadamard product to decompose variable sets. We refer the reader to ~\citep{loconte2024relationship} for discussions on alternative decomposition strategies.

\begin{restatable}{proposition}{propsoftcountingcondition}
	\label{prop:softcountingcondition}
	A soft counting circuit $\qcircuit(\xvars)$ satisfies $0 \leq \qcircuit(\xvars) \leq 1$.
\end{restatable}

\begin{proof}
	See Appendix \ref{sec:proof:prop:softcountingcondition}
\end{proof}

We note that soft counting circuits share many similarities with probabilistic circuits, in that they are both monotone by having weights matrices $W_k$ that have positive entries only. Furthermore, both circuit types have row-normalized weight matrices in the internal units and the root unit. The only difference, really, is that in the leaves, probabilistic circuits have row-normalized weight matrices as well, while the entries of weight matrices of soft counting circuits are bounded between zero and one. This has as a consequence that for a probabilistic circuit  $\circuit(\xvars)$ we have $\sum_{\xvars \in \Omega(\Xvars)} p(\xvars)=1$, while we have for soft counting circuits $0\leq \sum_{\xvars \in \Omega(\Xvars)} p(\xvars)\leq \samplespacesize^{\numvar}$. With $\numvar$ being the number of variables and   $\samplespacesize$ being the domain size of the variables.

\subsection{\noisepuncs as Deep Operator Mixtures}

An interesting consequence of modelling noise in the system via a soft counting circuit $\qcircuit(\xvars)$ is that it effectively results in a model that can be interpreted as a recursive mixture of positive operators. To see this, let us first write the matrix vector product in the computation units of $\qcircuit(\xvars)$ using explicit indices:
\begin{talign}
	\qcircuit_k
	= W_k \times (\qcircuit_{k_l} \odot \qcircuit_{k_r})
	\Leftrightarrow  \qcircuit_{ki} = \sum_j w_{kij}  \qcircuit_{k_l j}  \qcircuit_{k_r j}.
\end{talign}
Here $\qcircuit_{ki}$ denotes the $i$-th entry of the vector $\qcircuit_{k}$ and $w_{kij}$ denotes the elements of the matrix $W_k$.
We also dropped the explicit dependency on $\xvars_k$.

We now multiply each $\qcircuit_{ki}\in [0,1]$ with the corresponding operator $\punccircuit_k$
\begin{talign}
	\qcircuit_{ki} \punccircuit_k
	& = \left(\sum_j w_{kij}  \qcircuit_{k_l j}  \qcircuit_{k_r j} \right)  \bilinear_k( \punccircuit_{k_l}, \punccircuit_{k_r} )
	\nonumber
	\\
	& =\sum_j w_{kij}  \underbrace{\bilinear_k(  \qcircuit_{k_l j} \punccircuit_{k_l},  \qcircuit_{k_r j} \punccircuit_{k_r} )}_{=\tildeQcircuit_{kj}}
	\label{eq:deepoperatirmix}
\end{talign}
This means that we have for each computation unit $k$ not one but multiple operators, indexed by $i$, and each of the  $i$ operators is a mixture of operators:
\begin{talign}
	\Qcircuit_{ki}(\xvars_k) =\sum_j w_{kij} \tildeQcircuit_{kj}(\xvars_k).
	\label{eq:operatormix}
\end{talign}
We illustrate the situation in Figure~\ref{fig:operatormixture}. Such mixtures of operators can be regarded as generalizing standard mixture models of probability distributions, which can be recovered by considering the special case of operators of dimension one, in other words positive real-valued scalars.

\begin{figure}[t]
	\centering
	\input{tex_input/operatormix}
	\caption{
	Graphical representation of operator mixing as described in Equation~\ref{eq:operatormix}. The operators $\Qcircuit_{k0}$, $\Qcircuit_{k1}$, and $\Qcircuit_{k2}$ are all PSD matrices and constructed using the operators  $\tildeQcircuit_{k0}$, $\tildeQcircuit_{k1}$, and $\tildeQcircuit_{k2}$ using weighted sums. For $\Qcircuit_{k0}$ we also indicate the mixing weights $w_{k00}$, $w_{k01}$, and $w_{k02}$, which are positive real-valued scalars and satisfy $w_{k00}+ w_{k01}+w_{k02}=1$.
	}
	\label{fig:operatormixture}
\end{figure}

Recursively nesting such weighted sums of operators allows for representing a polynomial with exponentially many terms~\citep{martens2014expressive}.
However, in practice we only need to materialize these mixtures when computing $\sum_{\xvars \in \Omega(\Xvars) \pi_\Xvars(\xvars)}$ in order to normalize the probabilities.
If we evaluate a \noisepunc it suffices to evaluate the soft counting circuit $\qcircuit(\xvars)$ and the \punc $\punccircuit(\xvars)$ separately, and use the individual circuit evaluations to compute the unnormalized but bounded probability.

\citet{loconte2024relationship} proposed a similar combination of monotone and non-monotone circuits. This was motivated by their observation that the expressive power of monotone and squared circuits (a special case of positive operator circuits) are incomparable \citep{decolnet2021compilation}. That is, each of these circuit classes can express circuits that would lead to an exponential blow-up in the respective other circuit class.
Interestingly, we arrived at a similar construction by starting from noise in quantum circuits.

\section{Parametrizing \puncs}

In order to parametrize positive unitary circuits, we need to find a functional form of the bilinear forms such that Equation~\ref{def:eq:bilinear_psd} and Equation~\ref{def:eq:bilinear_unitpreserving} hold, and such that we have $A_k A_k^*=\mathbb{1}$.
For the latter we simply pick $A_k$ to be the (possibly truncated) orthonormal discrete Fourier transform matrix. This means that the $A_k$ are identical for all the leaves and that the leaves are parameter free but guarantees that we indeed have $A_kA_k^*?\mathbb{1}$.

For the bilinear forms we take inspiration from the probabilistic circuit literature and use a canonical polyadic ansatz~\citep{carroll1970analysis}.
\begin{align}
	\bilinear_k (\Ocircuit_{k_l}, \Ocircuit_{k_l} )
	=
	L_k \Ocircuit_{k_l} L_k^* \odot R_k \Ocircuit_{k_r} R_k^*,
	\label{eq:polyadicbilinear}
\end{align}
where $L_k$ and $R_k$ are complex valued matrices satisfying $L_k L_k^* \odot R_k  R_k^* = \mathbb{1}_{B\times B}$.

Assuming $O_{k_l}$ and $O_{k_r}$ are PSD we also have that $L_k \Ocircuit_{k_l} L_k^*$ and $R_k \Ocircuit_{k_r} R_k^*$ are PSD. Furthermore, using Schur's product theorem~\citep[p. 14, Theorem VII]{schur1911bemerkungen} that states that the Hadamard product of two PSD matrices is again PSD lets us conclude that parametrizing operator circuits with bilinear forms from Equation~\ref{eq:polyadicbilinear} results in the circuit being positive. For a discussion on the constraint $L_k L_k^* \odot R_k  R_k^* = \mathbb{1}_{B\times B}$ we point to Section~\ref{sec:efficient_param}.

\begin{restatable}{proposition}{propcomputationalcomplexity}
	\label{prop:computationalcomplexity}
	Let $\numvar$ be the number of variables $\xvars_k$,  having all the  domain size $\samplespacesize$, and let the $L_k$ and $R_k$ matrices be $\numbond\times \numbond$ square matrices, with $\numbond\leq \samplespacesize$.
	We can perform marginal inference in \puncs in $\bigO(\numvar \samplespacesize^3)$ when using the bilinear form from Equation~\ref{eq:polyadicbilinear}.
\end{restatable}

\begin{proof}
	See Append \ref{sec:proof:prop:computationalcomplexity}
\end{proof}

\subsection{The Vector Representation}
\label{sec:vpoc}

The matrix-matrix multiplication present in Equation~\ref{eq:polyadicbilinear} lead to the cubic computation cost per computation unit $\bigO (\samplespacesize^3)$ (\cf Section~\ref{sec:proof:prop:computationalcomplexity}). By exploiting the particular structure of the bilinear form in Equation~\ref{eq:polyadicbilinear} and by rearranging the operations performed in the operator circuit, we can avoid these matrix-matrix multiplications entirely and obtain matrix-vector multiplications instead. We call this alternative representation the \textit{vector} representation of an operator circuit.

\begin{definition}[Vector Representation]
	\label{def:vpoc}
	Let   $\xvars{=}\{\xvar_0,\dots ,\xvar_{\numvar{-}1}  \}$ be a set of $\numvar$ categorical variables with domains of size $\samplespacesize$.
	We define the vector representation of an operator circuit as a partition circuit whose computation units take the following functional form:
	\begin{align}
		{\vcircuit}_k(\xvars_k){=}
		\begin{cases}
			A_{k} \times  e_{\xvar_k},
			 & \text{if $k$ leaf}
			\\
			L_k {\times}  \vcircuit_{k_l} (\xvars_k)  \odot R_k {\times}  \vcircuit_{k_r} (\xvars_k),
			 & \text{else}
		\end{cases}
		\label{eq:vector_units}
	\end{align}
\end{definition}

\begin{restatable}{proposition}{propvoequiprob}
	\label{prop:voequiprob}
	The expression $\lVert \gamma \times \vcircuit_{\text{root}}(\xvars)\rVert^2$, with $\rho = \gamma \times \gamma^*$ computes the same probability as $\Tr [\Ocircuit_{\text{root}}(\xvars) \rho]$.
\end{restatable}

\begin{proof}
	See Section~\ref{sec:proof:prop:voequiprob}
\end{proof}

\begin{restatable}{proposition}{propcomputationalcomplexityvec}
	\label{prop:computationalcomplexityvec}
	\puncs in their vector representation compute joint probabilities in time $\bigO(\numvar \samplespacesize^2)$
\end{restatable}

\begin{proof}
	See Section~\ref{sec:proof:prop:computationalcomplexityvec}
\end{proof}

Note that the vector representation can only be used for computing joint probabilities as marginalizing out variables forces us to go to the matrix representation of operator circuits. Note also that the vector representation of \puncs can be seen as a PSD circuits~\cite{sladek2023encoding} or equivalently sum of squares circuits~\citep{loconte2024sum}, which we obtained by imposing a specific functional structure on the bilinear forms.

Having an efficient evaluation for \puncs in their vector representation also means that we can evaluate \noisepuncs more efficiently as we can compute $\qcircuit_{root}(\xvars)$ and $\pi_\Xvars(\xvars)$ separately, with the former having a computational complexity of $\bigO(\numvar \numcomponents \samplespacesize + \numvar \numcomponents^2)$.
Computing the normalization constant for a \noisepunc is more involved, and we discuss it along with its computational complexity in Appendix~\ref{sec:noisepuncnormalization}. However, we note that it is still feasible in polytime.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{An Efficient Parametrization}
\label{sec:efficient_param}

Having specified the functional form of the positive unitary circuits using polyadic bilinear forms we would also like to find an efficient parametrization of the $\gamma$, $L_k$, and $R_k$ matrices that is amenable to unconstrained gradient descent optimization. For $\gamma$ this is rather straightforward as we can simply use:
$
	\gamma = \frac{\tilde{\gamma}}{\sqrt{\Tr [\tilde{\gamma} \tilde{\gamma}^*]}},
$
where $\tilde{\gamma}\in \mathbb{C}^{B\times B}$ is an arbitrary complex values matrix. Note that we have almost by definition $\Tr [\rho] = 1$.

% Efficiently parametrizing the bilinear form such that they are amenable to unconstrained gradient based optimization is a bit more intricate. We refer the reader to Appendix~\ref{sec:bilinearparametrization}. The main idea is to use products of diagonal and circulant matrices~\citep{araujo2020understanding}, which can be parametrized using discrete Fourier transform matrices.

For the $L_k$ and $R_k$ matrices we make some further specifying choices. Consider the real-valued diagonal matrix $D_k \in \mathbb{R^{B\times B}}$ and the two unitary matrices $U_k \in \mathbb{C^{B\times B}}$ and $V_k \in \mathbb{C^{B\times B}}$, which we use to construct the $L_k$ and $R_k$ matrices:
\begin{align}
	L_k  = D_k^{\nicefrac{1}{2}} U_k,
	\quad \text{and} \quad
	R_k  = D_k^{-\nicefrac{1}{2}} V_k.
\end{align}

We can verify that this parametrization satisfies the unit preserving condition of for the bilinear form (\cf Definition~\ref{def:bilinear_unitpreserving}).
\begin{align}
	B_k(\mathbb{1}_{B\times B}, \mathbb{1}_{B\times B})
	 & = L_k \mathbb{1}_{B\times B} L_k^* \odot R_k \mathbb{1}_{B\times B} R_k^*
	\nonumber
	% \\
	%  & = L_k L_k^* \odot R_k R_k^*
	% \nonumber
	\\
	 & =D_k^{\nicefrac{1}{2}} U_k  U_k^* D_k^{\nicefrac{1}{2}}
	{\odot}
	D_k^{-\nicefrac{1}{2}} V_k  V_k^* D_k^{-\nicefrac{1}{2}}
	\nonumber
	\\
	%  & =D_k \odot D_k^{-1}
	% \nonumber
	% \\
	 & = \mathbb{1}_{B\times B}
\end{align}
While we now have a prescription to parametrize \puncs, one problem remains: unitary matrices are notoriously difficult to parametrize efficiently~\citep{kiani2022projunn,jing2017tunable,lezcano2019cheap,mhammedi2017efficient,wisdom2016full}.

We will take a practical approach here and parametrize the unitary matrices using a product of a diagonal and circulant matrix:
\begin{align}
	U_k = \Lambda_{U_k} F_B \Sigma_{U_k} F_B^{*}
	\quad \text{and} \quad
	V_k = \Lambda_{V_k} F_B \Sigma_{V_k} F_B^{*}.
\end{align}
Here $\Lambda_{U_k}$, $\Lambda_{V_k}$,  $\Sigma_{U_k}$, and $\Sigma_{V_k}$ are diagonal matrices whose entries have no magnitude, \ie they are of the form $e^{i\phi}$ with $\phi \in \mathbb{R}$. Furthermore, $F_B$ denotes the orthonormalized B-point discrete Fourier transform matrix, and we exploit the fact any circulant $B\times B$ matrix can be written as a diagonal matrix sandwidched between $F_B$ and $F_B^*$. Recalling that $F_B$ is unitary it is again straightforward to show that this parametrization yields unitary matrices. We also refer the interested reader to ~\citet{araujo2020understanding} for a in depth discussion of diagonal-circulant parametrization of neural networks.

While using this diagonal-circulant parametrization does not cover the entire space of unitary matrices it has the advantage of being parametrizable with unconstrained parameters. Furthermore, using the fact that mulitplying a discrete Fourier transform matrix with a vector can be done on $\bigO(\numbond \log \numbond)$ brings down the computational cost of computing the bilinear form from $\bigO(\numbond^2)$.

Lastly,  for the set of  basis vectors $\{e_{\xvar_s} \}_{s=0}^{\samplespacesize-1}$ in the leaves we choose the set of standard basis vectors, \ie one-hot unit vectors, with each basis vector $e_s$ corresponding to a random outcome. Although different choices of basis vectors would be valid as well, For the $A_k$ we use, as already mentioned, the orthonormal discrete Fourier transform matrix, which we truncate if $\numbond<\samplespacesize$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Related Work}
\label{sec:related}

\subsection{Squared Cirucuits}

The work closest related to ours are the \text{sum of compatible square circuits} by \citet{loconte2024sum}.
As shown in Section~\ref{sec:vpoc} we recover sum of square circuits (introduced earlier as PSD circuits by \citet{sladek2023encoding}) if we use a polyadic decomposition in the bilinear form, \cf Equation~\ref{eq:polyadicbilinear}.
A non-polyadic decomposition would have resulted in a computation graph closer to that of \textit{inception circuits} introduced by \citet{wangrelationship}. Such a choice, however, increases the computation cost per unit in the partition circuit from quadratic (using the vector representation) to cubic (using the operator representation).

Our positive unit preserving operator circuits provide also a different perspective on constructing non-monotone circuits. While the methods described in ~\citep{sladek2023encoding,loconte2024subtractive,loconte2024sum,wangrelationship} regard such circuits as sum of squares, we interpret them as probabilistic events described by positive semidefinite matrices that are combined within a circuit using unit preserving bilinear forms. As such, we also establish a strong link between the circuit literature and quantum information theory.

From a practical perspective, the (sub-complete) distributions encoded with \puncs and \noisepuncs tackle one of the main shortcoming of sum of squares type circuits: we do not need to compute a normalization constant to ensure that event probabilities are bounded. This especially useful during training, as it means that updating the parameters of the model still guarantees the event probability to be bounded. Otherwise one would have to compute a normalization constant and differentiate through it at every iteration of the parameter updates. This potentially hinder scaling up non-monotone models as computing the normalization constant cannot be done in the vector representation (quadratic cost) but requires matrix-matrix multiplications with cubic cost. For \puncs computing this normalization is not necessary at all and for \noisepunc we only need to compute the normalization constant once training has finished as post-processing step.

\subsection{Tensor Networks}

As already pointed out by \citet{loconte2024subtractive,loconte2024sum} squared circuits and monotone circuits share many similarities with \textit{tensor networks}~\citep{orus2014practical,white1992density}
developed in the condensed matter physics community and have in recent years also been applied to tackle problems in supervised as well as unsupervised machine learning~\citep{cheng2019tree,han2018unsupervised,stoudenmire2016supervised}. In this regard and given that tensor networks originate in the physics community, it is rather surprising that tensor networks have so far, and to the best of our knowledge, not been formulated using POVMs.

Using our POVM formulation for \puncs is, however, not only theoretically elegant but has also practical benefits: using this formalism leads to circuits that are normalized by construction, and we can perform learning simply by optimizing the likelihood. We contrast this to the sweeping algorithms that are usually deployed in the tensor network literature, where blocks of variables are optimized one at the time while the remaining variables are held constant. It appears that this approach is inspired by the variational ansatz taken in the density-matrix renormalization group algorithm~\citep{white1992density}, which is the original algorithm for tensor networks.
Alternatives to this sweeping approach have also been proposed, such as an intricate Riemannian gradient descent optimization scheme that conserves unitarity of matrices, but are rather costly to run.

We would also like to point out a theoretical result from the tensor network literature stating that picking a complex-valued parametrization instead of a real-valued one can lead to an arbitrarily large reduction in the number of parameters~\citep{glasser2019expressive}. While this result is formulated with respect to (exact) low rank tensor decomposition and does not apply directly to the problem of learning parameters via gradient descent, we consider this observation to be a strong theoretical indicator for the superiority of complex numbers over real numbers when parametrizing probabilistic circuits. A similar argument has also been made by \citet{gao2022enhancing} in the context of hidden Markov models.

\subsection{Theoretical Studies}

First theoretical results on expressive power of polynomial functions, \ie circuits and tensor networks, were presented in the tensor network literature in the context of tensor decomposition \citep{glasser2019expressive} and complex-valued hidden Markov models \citep{gao2022enhancing}. Recently, the works of \citet{loconte2024subtractive,loconte2024sum} and \citet{wangrelationship} have studied the relationship of different circuits classes more carefully, as well. Additionally, \citet{loconte2024sum} pointed out links between tensor networks and the circuit literature and were able to generalize earlier results from the tensor network literature by~\citet{glasser2019expressive}.

Finally, we would also like to point out theoretical results n the statistical relational AI literature. Specifically, \citet{buchman2017rules}, and \citet{kuzelka2020complex} noted that using only real-valued parametrizations (including negatives~\citep{buchman2017negative}), \ie monotone functions exclusively, does not allow for fully expressive models.

\section{Experimental Evaluation}
\label{sec:experiments}

\textbf{Experimental Setup}

For our experimental evaluation we used the MNIST family of datasets. That is, the original MNIST~\citep{deng2012mnist}, FashionMNIST~\citep{xiao2017fashion}, and also EMNIST~\citep{cohen2017emnist}.
For the implementation we used PyTorch together with the Einops library~\citep{rogozhnikov2022einops}. We set up our experiments in Lightning\footnote{\url{https://lightning.ai/}} and ran them on a DGX-2 machine with V100 Nvidia cards.

We compare different methods using the \textit{bits per dimension} metric, which is calculated from the average negative log-likelihood ($\overline{NLL}$) as follows: $bpd {=} \nicefrac{\overline{NLL}}{(\log{2} \times D)}$ ($D{=}28^2$ for MNIST datasets).

\textbf{Training Positive Operator Circuits}

In our experiments we estimate the density of the different MNIST datasets by maximizing a parametrized likelihood function. We construct this  function using a positive vector circuit.
In order to build the underlying partition circuit,
we follow the approach described by \citet{zuidberg2024probabilistic}: we start with a grid of $28\times 28$ pixels and merge, in an alternating fashion, rows and columns of the image.
We also allow for merging three partitions into  a single partition, thereby breaking the binary character of the partition trees used so far.
Having three instead of two partitions being merged can easily be accommodated for by using a Hadamard product with three factors instead of two in the internal units of a \pvc. Ultimately, this allows us to handle layers with an uneven number of partitions.
At the leaves we encode pixel values by associating each value to one of $256$ standard basis vector of $\mathbb{R}^{256}$.

We performed training by minimizing the negative log-likelihood using Adam~\citep{kingma2014adam} with default hyperparameters, batch size of $50$, over $100$ epochs.
The best model was selected using a $90-10$ train-validation data split where we monitored the negative log-likelihood on the validation set.
We did not perform any hyperparameter search.
Further details can be found in the configuration files of the experiments.

\textbf{State-of-the-Art Baselines}

We compare \pvcs to three state-of-the-art tractable density estimators from the literature: quadrature of probabilistic circuits (QPCs)~\citep{gala2024probabilistic}, hidden Chow-Liu trees (HCLTs)~\citep{liu2021tractable}, and sparse HCLTs (SHCLTs)~\citep{dang2022sparse}.
All three methods are probabilistic circuits and they all use hidden Chow-Liu trees as their underlying structure. This tree is learned and dataset specific. The difference between QPCs and HCLTs is that the former is obtained by approximating a continuous latent variable extension of probabilistic circuits using numerical quadrature. The difference between HCLTs and SHCLTs is that the latter allows for dynamically adapting the number of components per partition during parameter learning. This means that SHCLTs can focus their computational budget on information-rich parts of the circuit.

\textbf{Q1: How Do the Different \pvc Parametrizations Fair Against Each Other?}

We construct \pvcs following the prescription of \cite{zuidberg2024probabilistic}, as described above. Furthermore, we use $B=128$ components per partition.
As for the parametrization we study four different variants:
$\mathrm{PVX}_{0}^{FR}$,
$\mathrm{PVX}_{2\pi}^{FR}$,
$\mathrm{PVX}_{0}^{R1}$, and
$\mathrm{PVX}_{2\pi}^{R1}$.
The subscripts ($0$ or $2\pi$) indicate whether we limit the phase to be zero or whether we allow the phase to be a learnable parameter. Note that having a tunable phase doubles the number of (real-valued) parameters. The superscript indicates whether the parametrization uses a full-rank ($FR$) density matrix in the root or a rank-one ($R1$) density matrix.
As such, $\mathrm{PVX}_{0}^{R1}$ is equivalent to the  squared monotonic probabilistic circuits (\smpcs) used in ~\cite{loconte2024subtractive}, and $\mathrm{PVX}_{2\pi}^{R1}$ generalizes their \snpcs from the real domain to the entire complex domain. Concretely, \smpcs are rank-one \pvcs with the phase fixed to zero and \snpcs are rank-one \pvcs with the phase fixed to values in the set $\{0, \pi \}$. We do not experiment on the latter.

We report the obtained bpds for the different \pvc parametrizations in the first four columns of Table~\ref{tab:mnist}. In general, we see that the complex-valued parametrizations are either on par or outperform the corresponding zero-phase parametrizations.
The notable exception is the FashionMNIST benchmark where the two parametrizations with no phase ($\mathrm{PVX}_{0}^{FR}$ and $\mathrm{PVX}_{0}^{R1}$) perform best.
Comparing full-rank to rank-one parametrizations we see a more important impact for the zero-phase parametrization than for the complex-valued parametrization.

Overall, $\mathrm{PVX}_{2\pi}^{FR}$ constitutes the best performing parametrization being best-in-class on all benchmarks but FashionMNIST.

\input{tex_input/mnist_table}

\textbf{Q2: How Do \pvcs Fair Against the State of the Art?}

In order to compare \pvcs to state-of-the-art circuits all methods (\pvc, QPC, HCLT, SHCLT) were given a computational budget of $B=128$ components per partition (on average for SHCLT).
The results for the competing methods were taken from the respective papers -- except for HCLTs. For HCLTs we took the bpds reported by \citet{gala2024probabilistic} as they achieved stronger results with their implementation compared to the originally reported performance~\citep{liu2021tractable}.

We see in Table~\ref{tab:mnist} that QPCs and HCLTs are in general outperformed by \pvc -- in particular $\mathrm{PVX}_{2\pi}^{FR}$. The FashionMNIST benchmark constitutes again an outlier in this regard.
The only methods outperforming \pvcs are the SHCLTs, which is due to them being able to dynamically allocate compute budget to informative parts of the circuit.
It is noteworthy that for the EMNIST-BYCLASS benchmark all four  \pvc parametrization exhibit strong performance with  $\mathrm{PVX}_{2\pi}^{FR}$ and $\mathrm{PVX}_{2\pi}^{R1}$ even outcompeting SHCLTs.

\textbf{Discussion}

In our experimental evaluation we did not perform an explicit comparison to the method of \citet{loconte2024subtractive} as they were not able to find any improvements of allowing non-negative parameters in probabilistic circuit for discrete data. They stipulated that
\say{simple categorical distribution can already capture any discrete distribution with finite support and a (subtractive) mixture thereof might not yield additional benefits}.
In our experimental evaluation we refute this conjecture, and show that using complex-valued parameters leads to noticeable gains when performing density estimation on discrete data.
The exception being of course the FashionMNIST benchmark. We believe that this is due to our optimization method that was not tuned in any way.
In this regard we believe that developing tailor-made optimization algorithms for \pvcs might further boost their performance.
Furthermore, one can envisage, similar to SHCLTs, dynamically adapting the compute budget per partition, which should again improve the density estimation capacities of \pvcs.

\section{Conclusions \& Future Work}
\label{sec:conclusions}

Based on first principles from quantum information theory, we constructed positive operator circuits -- a novel class of probabilistic tractable models:
Their construction as partition circuits allows for layer-wise parallelization and execution on modern machine learning hardware.
Using unconstrained gradient-descent we showed that \pocs, and \pvcs specifically, constitute a promising addition to the zoo of tractable probabilistic models.

In future work we would like to investigate in more detail theoretical properties of \pocs and how they differ from other tractable models. Ideally one would find an exponential separation between real-valued and complex-valued circuits. On the practical side it remains, however, to be seen whether such a separation has an equally important impact on real-world datasets.

This relates to another open question,  that of learning in \pocs. We presented a rather simple learning approach for parameters tailored towards neural architecture. It might be the case that more sophisticated techniques have to be deployed for large-scale \pocs, as they might exhibit the problem of barren plateaus~\citep{ragone2023unified} -- a well known issue in quantum machine learning~\citep{biamonte2017quantum,huggins2019towards}.
Furthermore, and as already pointed out by~\citet{loconte2024subtractive}, computing the normalization constant $Z$ requires the evaluation of the circuit in the \poc representation. This gives rise to a rather expensive cubic computation cost during learning (when compared to the quadratic cost of \pvc evaluations). Avoiding this issue would allow to drastically scale \pvcs. In order, to scale \pvcs it might also be necessary to use more sophisticated structures than binary trees with every partition having the same number of components $B$ per partition, \cf SHCLTs.

### 2025-06-02 202507_uai_puncs/icml2025/noise_puncstuff/tex_input/operatormix.tex
added uai

\resizebox{0.7\columnwidth}{!}{
	\centering

	\begin{tikzpicture}[node distance=2cm, auto]

		% Upper layer nodes with increased size
		\node[draw, circle, minimum size=0.5cm,inner sep=1pt, line width=1.5pt] (A1) at (0, 2) {$+$};
		\node[draw, circle, minimum size=0.5cm,inner sep=1pt] (A2) at (2, 2) {$+$};
		\node[draw, circle, minimum size=0.5cm,inner sep=1pt] (A3) at (4, 2) {$+$};
		% Lower layer nodes with increase size
		\node[draw, circle, minimum size=0.5cm] (B1) at (0, 0) {};
		\node[draw, circle, minimum size=0.5cm] (B2) at (2, 0) {};
		\node[draw, circle, minimum size=0.5cm] (B3) at (4, 0) {};

		% Labels for the upper layer
		\node at (0, 2.5) {$\Qcircuit_{k0}$};
		\node at (2, 2.5) {$\Qcircuit_{k1}$};
		\node at (4, 2.5) {$\Qcircuit_{k2}$};

		% Labels for the lower layer
		\node at (0, -0.5) {$\tildeQcircuit_{k0}$};
		\node at (2, -0.5) {$\tildeQcircuit_{k1}$};
		\node at (4, -0.5) {$\tildeQcircuit_{k2}$};

		% Connecting upper layer to lower layer
		\foreach \i in {1,2,3}{
				\foreach \j in {1,2,3}{
						\draw (A\i) -- (B\j);
					}
			}

		% Thicker edges connected to the upper-left node
		\draw[line width=1.5pt] (A1) -- (B1) node[midway, left]{$w_{k00}$};
		\draw[line width=1.5pt] (A1) -- (B2) node[midway, left] {$w_{k01}$};
		\draw[line width=1.5pt] (A1) -- (B3) node[midway, left] {$w_{k02}$};

	\end{tikzpicture}
}

### 2025-06-02 202507_uai_puncs/neurips2024/content.tex
added uai

--- 2025-06-02 202507_uai_puncs/iclr2025/content.tex
@@
 
+\todo{add back this shit at the end for replication and all that super long annoying list of silly questions}
+
 Probabilistic circuits (PCs)~\citep{darwiche2003differential,poon2011sum} belong to an unusual class of probabilistic models: they are highly expressive but at the same time also tractable.
@@
 \end{definition}
+
+For the interested reader we give the definition of a layered probabilistic circuit in terms of a partition
+circuit, as well as a proof that they form valid probability distributions, in Appendix \ref{sec:pc}.
 
@@
 \bibliography{references}
-\bibliographystyle{iclr2025_conference}
 

### 2025-06-02 202507_uai_puncs/submission/appendix.tex
added uai

\maketitle

\section{Proofs for Section~\ref{sec:qit}}

\subsection{Proof of Proposition \ref{prop:povmprob}}
\label{sec:proof:prop:povmprob}

\proppovmprob*

\begin{proof}
	While this is a well-known result we were not able to identify a concise proof in the literature and provide therefore, for the sake of completeness, one here. To this end, we
	first show that $p(i)\geq 1$, for each $i$.
	\begin{align}
		p(i)
		=
		\Tr [ E(i) \rho ]
		% \nonumber
		% \\
		 &
		=
		\Tr [ D(i) D^*(i)  \rho ]
		% \nonumber
		\\
		 &
		=
		\Tr [ D(i)^*  \rho D(i) ]. \nonumber
	\end{align}
	Here we used the fact that $E(i)$ is PSD and factorized it into the product $D(i) D^*(i)$. Then we used the fact that the trace is invariant under cyclical shifts.
	Clearly, the matrix $ D(i)^*  \rho D(i)$ is PSD as we have for every vector $x$:
	\begin{align}
		x^* D(i)^*  \rho D(i) x = y^* \rho y \geq 0 \quad \text{with } y = Dx.
		\label{eq:def:psdPOVM}
	\end{align}
	As the trace of a PSD matrix is positive we have that $p(i)$ is a positive real number, \ie $p(i){\geq} 0$ for every $i$.

	Secondly, we show that $p(i)$ is normalized:
	\begin{align}
		\sum_{i=1}^N p(i)
		%  &
		=
		\sum_{i=1}^N \Tr [ E(i) \rho ]
		% \nonumber
		% \\
		%  &
		=
		\Tr [ \sum_{i=1}^N E(i) \rho ]
		% \nonumber
		% \\
		%  &
		{=}
		\Tr [  \rho ],
	\end{align}
	where we used Equation~\ref{eq:povm_normalized}.
	Exploiting the fact that the trace of a density matrix is $1$ gives us indeed $\sum_{i=1}^N p(i)=1$, and we can conclude that $p(i)$ is a valid probability distribution.
\end{proof}

\section{Proofs for Section \ref{sec:puncs}}

\subsection{Proof of Proposition \ref{prop:qopunital}}
\label{sec:proof:prop:qopunital}

\propqopunital*

\begin{proof}
	The (sufficient and necessary) condition that $\sum_j \kraus_j^* \kraus_j \leq \mathbb{1}$ stems from fact that we wish to bound the probability of a state $\qop (E(i))$ to be less than or equal to $1$, \cf \citep[Proof of Theorem 8.1]{nielsen2001quantum}. That is, we wish to have:
	\begin{align}
		\Tr [ \qop (E(i)) \rho] \leq 1. \label{eq:proof:unitalvalid:bound}
	\end{align}
	We will now show that for unital quantum operations this holds by construction and that the condition $\sum_j K_j^* K_j$ is implied.

	We start with the probability for the state $\sum_i E(i)=\mathbb{1}$:
	\begin{align}
		\Tr [ \qop (\mathbb{1}) \rho]
		=
		\Tr [ \sum_j \kraus_j  \mathbb{1} \kraus_j^* \rho ]
		=
		\Tr [ \sum_j \kraus_j  \kraus_j^* \rho ]
		=
		\Tr [  \rho ] =1
	\end{align}
	Alternatively, we write this also as:
	\begin{align}
		\Tr [ \qop (\mathbb{1}) \rho]
		=
		\Tr [ \sum_j \kraus_j  \mathbb{1} \kraus_j^* \rho ]
		=
		\sum_j \Tr [ \kraus_j  \mathbb{1} \kraus_j^* \rho ]
		=
		\sum_j \Tr [\gamma^* \kraus_j  \mathbb{1} \kraus_j^* \gamma ]
	\end{align}
	with $\rho=\gamma \gamma^*$.

	Similarly, we also write for the probability of the arbitrary state $\sigma$ denoting the sum over any subset of the POVM $\{ E(i) \}_{i=1}^{N}$:
	\begin{align}
		\Tr [ \qop (\sigma) \rho]
		=
		\sum_j \Tr [\gamma^* \kraus_j \sigma \kraus_j^* \gamma ].
	\end{align}
	We now need that $\Tr [ \qop (\sigma) \rho]\leq1$, which is equivalent to:
	\begin{align}
		1 - \Tr [ \qop (\sigma) \rho] \geq 0
		 & \Leftrightarrow
		\Tr [ \qop (\mathbb{1}) \rho] - \Tr [ \qop (\sigma) \rho] \geq 0
		\\
		 & \Leftrightarrow
		\sum_j \Tr [\gamma^* \kraus_j  \mathbb{1} \kraus_j^* \gamma ]
		-
		\sum_j \Tr [\gamma^* \kraus_j \sigma \kraus_j^* \gamma ] \geq 0
		\\
		 & \Leftrightarrow
		\sum_j \Tr \Bigl[\gamma^* \kraus_j  \Bigl( \mathbb{1} -\sigma \Bigr)  \kraus_j^* \gamma \Bigr] \geq 0
	\end{align}
	The last line only holds if $\mathbb{1}- \sigma$ is PSD, which is indeed the case as $\sigma$ only sums over a subset of the POVM. Subtracting the sum of this subset from $\mathbb{1}$ leaves us with a sum over the remaining elements of the POVM. As this is a sum over PSD matrices the sum over the remaining elements is again PSD.
	This concludes the proof as we have shown that Equation~\ref{eq:proof:unitalvalid:bound} is satisfied by construction.
\end{proof}

\subsection{Proof of Proposition \ref{prop:puncPOVM}}
\label{sec:proof:prop:puncPOVM}

\proppuncPOVM*

\begin{proof}
	Given that positive unital circuits are by definition positive operator circuits we already have that:
	\begin{align}
		\forall \xvars \in \Omega(\Xvars): \Ocircuit(\xvars) \text{ is PSD}.
	\end{align}
	Next we show that $\sum_{\xvars\in \Omega(\Xvars)} \Ocircuit(\xvars)=\mathbb{1}$. Here we observe that in the computation units we have:
	\begin{align}
		\sum_{\xvars_k \in \Omega(\Xvars_k)} \Ocircuit_k(\xvars_k)
		 & = \sum_{\xvars_{k_l}} \sum_{\xvars_{k_r}}    \qop(\Ocircuit_k(\xvars_{k_l}) \otimes \Ocircuit_k(\xvars_{k_r})  )
		\\
		 & =
		\qop \left(
		\sum_{\xvars_{k_l}}   \Ocircuit_k(\xvars_{k_l})
		\otimes
		\sum_{\xvars_{k_r}}\Ocircuit_k(\xvars_{k_r})
		\nonumber
		\right)
	\end{align}
	This lets us push down the summation of a specific variable to the corresponding leaf where the variable is given as input, where we then have:
	\begin{align}
		\sum_{\xvars_k \in \Omega(\Xvars_k)} \Ocircuit_k(\xvars_k)
		= \sum_{\xvar_k\in \Omega(X_k)} E_{\xvar_k}
		= \mathbb{1_k}
	\end{align}
	We now exploit that the completely positive maps in a positive unital circuit are unital, which gives us indeed
	$
		\sum_{\xvars\in \Omega(\Xvars)} \Ocircuit(\xvars)=\mathbb{1}.
	$
\end{proof}

\section{Proofs for Section \ref{sec:special_cases}}

\subsection{Proof of Proposition \ref{prop:Oveq}}
\label{sec:proof:prop:Oveq}

\propOveq*

\begin{proof}
	We start the proof at the leaf where we have
	\begin{align}
		\Ocircuit_k(\xvar_k)
		=
		U_k \left( e_{\xvar_k} \otimes e_{\xvar_k}^*  \right) U_k^*
		=
		\left( U_k e_{\xvar_k} \right) \otimes \left( e_{\xvar_k}^* U_k^* \right)
	\end{align}
	and Equation~\ref{eq:def:opvec_equivalent} holds almost by definition. In the computation units into which the leaves feed, we then have
	\begin{align}
		\Ocircuit_k
		 & = U_k \left( \Ocircuit_{k_l} \otimes \Ocircuit_{k_r} \right) U_k^*
		\nonumber
		\\
		 & = U_k \left( (\Vcircuit_{k_l} \otimes \Vcircuit^*_{k_l}) \otimes (\Vcircuit_{k_r} \otimes \Vcircuit^*_{k_r} ) \right) U_k^*
		\nonumber
		\\
		 & = \Bigl( U_k \left( \Vcircuit_{k_l} \otimes \Vcircuit_{k_r} \right) \Bigr)
		\otimes
		\Bigl( \left( \Vcircuit^*_{k_l}  \otimes \Vcircuit^*_{k_r}  \right) U_k^* \Bigr)
		\nonumber
		\\
		 & = v_k \otimes v_k^*.
	\end{align}
	where we omitted the explicit dependencies on the variables $\xvar_k$, $\xvar_{k_l}$, and $\xvar_{k_r}$.
	Repeating this argument recursively until the root of the circuit concludes the proof.
\end{proof}

\subsection{Proof of Proposition \ref{prop:voequiprob}}
\label{sec:proof:prop:voequiprob}

\propvoequiprob*

\begin{proof}
	The proof starts by simply plugging in the vector representation of $\Ocircuit_{\text{root}}$ (obtained in the proof of Proposition~\ref{prop:Oveq}) into the expression $\Tr [\Ocircuit_{\text{root}} \rho]$ and rather straightforwardly get:
	\begin{align}
		\Tr [\Ocircuit_{\text{root}} \rho]
		 &
		=
		\Tr \left[
			\Bigl(\Vcircuit_{\text{root}} \otimes \Vcircuit^*_{\text{root}}   \Bigr) \times \rho
			\right]
		\\
		 & = \Vcircuit^*_{\text{root}} \times  \rho \times\Vcircuit_{\text{root}},
	\end{align}
	which is indeed the same probability as defined in Definition~\ref{def:vpoc}.
\end{proof}

\subsection{Proof of Proposition \ref{prop:circuit_is_prob}}
\label{app:circuit_is_prob}

\proplpcvalid*

\begin{proof}
	If $\Pcircuit_{root} (\xvars)$ is the computation unit at the root of the layered PC, each entry $i$ of $\Pcircuit_{root} (\xvars)$ forms a probability distribution if $\forall i: \Pcircuit_{root, i}(\xvars)\geq 0$, for every $\xvars \in \Omega(\Xvars)$ and if $\forall i: \sum_{\xvars \in \Omega(\Xvars)} \Pcircuit_{root, i}(\xvars)=1$. The first condition is trivially satisfied as the circuit only performs linear operations on matrices and vectors ($P_{\xvar_k}, W_k$) with positive entries only.
	For the condition $\sum_{\xvars \in \Omega(\Xvars)} \Pcircuit_{root}(\xvars)=1$ we observe that we can simply push down the summation for each variable to the respective leaf unit. This yields the following summation in the leaves:
	\begin{align}
		\sum_{\xvar_k \in \Omega(\Xvar_k)} W_k \times P_{\xvar_k}
		=
		W_k \times \sum_{\xvar_k \in \Omega(\Xvar_k)} P_{\xvar_k}
		=
		W_k \times \eta_{k} = \eta_{k}.
	\end{align}
	Here $\eta$ denotes a vector having as entries only ones (with appropriate dimensions). For the last step in the equation above we exploited the fact that the weight matrices $W_k$ are row-normalized.

	Passing on the marginalized leaves to the parent units gives us:
	\begin{align}
		W_k \times \Bigl( \eta_{k_l} \otimes \eta_{k_r} \Bigr)
		=
		W_k \times \eta_{k}  =  \eta_k
	\end{align}
	Repeating this process until we reach the root will eventually result in $\sum_{\xvars \in \Omega(\Xvars)} \Pcircuit_{root}(\xvars)=\eta_{root}$. This means that all the entries of all the vectors $\pcircuit_k(\xvars_k)$ equal to $1$ when marginalized, which means in turn that the second condition is satisfied.
\end{proof}

\subsection{Proof of Proposition \ref{prop:diagpunc}}
\label{sec:proof:prop:diagpunc}

\propdiagpunc*

\begin{proof}
	The proof is relatively straightforward as in the leaves we already have diagonal matrices by definition. For an internal unit $k$ we have that
	\begin{align}
		\Ocircuit_k = \sum_j J_j D_{kj} \left(\Ocircuit_{kl}\otimes \Ocircuit_{kr} \right)  D_{kj}^* J_j^*{kj},
	\end{align}
	which can be written as
	\begin{align}
		\Ocircuit_k = \sum_j J_j D_{kj} J_j^*{kj},
	\end{align}
	where $D_{kj}$ is diagonal and PSD. Carrying out the remaining matrix products then gives us:
	\begin{align}
		\Ocircuit_k = \sum_j H_j \Tr [D_{kj}],
	\end{align}
	which is clearly a diagonal matrix (and also PSD).
\end{proof}

\subsection{Proof of Proposition \ref{prop:PCPunciso}}
\label{sec:proof:prop:PCPunciso}

\propPCPunciso*

\begin{proof}
	In order to show this we need to show that each computation unit in a probabilistic circuit can be mapped to a computation unit in a diagonal \punc -- and vice versa. Starting at the leaves for both circuits we have
	\begin{align}
		P_{\xvar_k}
	\end{align}
	for the probabilistic circuit and
	\begin{align}
		\Delta_{\xvar_k}  \Delta_{\xvar_k}^*,
	\end{align}
	for the diagonal \punc. Given that $\Delta_{x_k}$ is a diagonal matrix the product $\Delta_{\xvar_k}  \Delta_{\xvar_k}^*$ is also a diagonal matrix with positive entries only. Taking into consideration the completeness constraints $\sum_{x_k} P_{x_k} = \eta_k $ ($\eta$ being a vector with only ones) and $\sum_{x_k} \Delta_{x_k} \Delta_{x_k}^* = \mathbb{1} $ is trivial to see that in the leaves probabilistic circuits can be mapped onto diagonal \puncs and vice versa.

	For the internal computation units we need to show that
	\begin{align}
		\phantom{\Leftrightarrow} &
		\Pcircuit_k
		=
		W_k {\times} \Bigl(\Pcircuit_{k_l} {\otimes} \Pcircuit_{k_l}\Bigr)
		\\
		\Leftrightarrow           &
		\Pcircuit_{ki}
		=
		\sum_j W_{kij} {\times} \Bigl(\Pcircuit_{k_l} {\otimes} \Pcircuit_{k_l}\Bigr)_j
		\label{eq:proof:iso:vecunit}
	\end{align}
	is equivalent to
	\begin{align}
		\Ocircuit_k = \qop (\Ocircuit_{k_l} \otimes \Ocircuit_{k_r})
		=
		\sum_j J_j D_{kj} (\Ocircuit_{k_l} \otimes \Ocircuit_{k_r}) D^*_{kj} J_j^*.
		\label{eq:proof:iso:sumdiag}
	\end{align}
	We first note that for diagonal \puncs all the operators $\Ocircuit_k$ are representable as diagonal matrices. This allows us to rewrite Equation~\ref{eq:proof:iso:sumdiag} as
	\begin{align}
		\Ocircuit_k = \qop (\Ocircuit_{k_l} \otimes \Ocircuit_{k_r})
		=
		\sum_j  J_j  \underbrace{ D_{kj} D^*_{kj}}_{\eqqcolon \Lambda_{kj}} (\Ocircuit_{k_l} \otimes \Ocircuit_{k_r})  J_j^*.
	\end{align}
	When writing out this sum of matrix product on the right-hand side explicitly it is trivial to observe that the diagonal element $\Ocircuit_{kjj}$ can be written as:
	\begin{align}
		\Ocircuit_{kjj}
		=
		\Tr \left[  \Lambda_{kj} \times (\Ocircuit_{k_l} \otimes \Ocircuit_{k_r}) \right],
	\end{align}
	with all the off-diagonal elements being zero. This can in turn be written as
	\begin{align}
		\Ocircuit_{kjj}
		=
		\sum_i  \diagvec ( \Lambda_{kj} )_i  \Bigl( \diagvec (\Ocircuit_{k_l}) \otimes \diagvec (\Ocircuit_{k_r}) \Bigr)_i,
	\end{align}
	where $\diagvec (\Lambda_{kj})$ denotes a vector whose entries corresponds to the diagonal elements of $\Lambda_{kj}$.

	Switching around the names of the indices $i$ and $j$ we can write:
	\begin{align}
		\diagvec (\Ocircuit_{k} )_i
		=
		\sum_j  \diagvec ( \Lambda_{ki} )_j  \Bigl( \diagvec (\Ocircuit_{k_l}) \otimes \diagvec (\Ocircuit_{k_r}) \Bigr)_j.
		\label{eq:proof:iso:opunit}
	\end{align}
	We now identify the expressions in Equation~\ref{eq:proof:iso:vecunit} and Equation~\ref{eq:proof:iso:opunit} as follows with each other:
	\begin{align}
		\diagvec (\Ocircuit_{k} )_i   & = \Pcircuit_{ki}
		\\
		\diagvec (\Lambda_{ki} )_j    & = W_{kij}
		\\
		\diagvec (\Ocircuit_{k_l} )_i & = \Pcircuit_{k_l i}
		\\
		\diagvec (\Ocircuit_{k_r} )_i & = \Pcircuit_{k_r i}.
	\end{align}
	This -- together with the equivalence between $\forall k: \Tr [D_{ki}D^{*}_{ki}]= \Tr [\Lambda_{ki}] =1$ and the row-normalization of the $W_k$'s -- establishes the isomorphism between the two circuit classes.
\end{proof}

\section{Noisy and Block-Diagonal \puncs}
\label{sec:noisyblockdiagonalpuncs}

In order to increase the expressive power of non-monotone circuits, \citet{loconte2025sum} proposed to multiply a pure \punc with a probabilistic circuit.
This was motivated by their observation that the expressive power of monotone and squared circuits (a special case of pure \puncs) are incomparable \citep{decolnet2021compilation}. That is, each of these circuit classes can express circuits that would lead to an exponential blow-up in the respective other circuit class. The idea is relatively simple: they express an unnormalized probability distribution as the product of a probabilistic circuit and a traced \punc:
\begin{align}
	\qcircuit(\xvars) \Tr [\Ocircuit(\xvars) \rho].
\end{align}
They called such a model a \msocs. We will put these models in a quantum information theoretic framework. This will then lead to the realization that \msocss are nothing but block-diagonal operator circuits.

\subsection{Sub-Complete Probability Distributions}

An important concept in quantum information theory that generalizes the standard POVM (\cf Definition~\ref{def:povm}) is the so-called noisy POVM, which enables for instance modelling imperfect measurement devices. In the context of machine learning this might be an improperly labeled data point of an error in the detector, \eg a pixel flip in the camera that took a picture.
\begin{definition}
	\label{def:noisypovm}
	A noisy positive operator-valued measure
	is a set of PSD  matrices $\{E(i)\}_{i=0}^{I-1}$ ($\numevents$ being the number of possible measurement outcomes) such that:
	\begin{talign}
		\sum_{i=0}^{I-1} E(i) = \subcompletemeasrure < \mathbb{1},
	\end{talign}
	where the inequality sign is interpreted using the Loewner ordering of PSD matrices,
	\ie $\subcompletemeasrure < \mathbb{1} \Leftrightarrow\mathbb{1}-\subcompletemeasrure$ is PSD.
\end{definition}

\begin{definition}[Sub-complete Probability Distribution]
	Let $\Xvars$ be a set of random variables with sample space $\Omega(\Xvars)$. We call a probability distribution \textit{sub-complete} if $\forall \xvars \in \Omega(\Xvars): p(\xvars)\geq 0$ and if $\sum_{\xvars \in \Omega(\Xvars)} p(\xvars)\leq 1$. We call a probability distribution \textit{strictly sub-complete} if the latter inequality holds strictly and \textit{complete} if equality holds.
\end{definition}

\begin{restatable}{proposition}{propnoisypovmdist}
	\label{prop:noisypovmdist}
	Noisy POVMs induce a sub-complete probability distributions.
\end{restatable}

\begin{proof}
	Let $\rho$ be a density matrix and $\{E(i)\}_{i=0}^{I-1} $ be a POVM such that $\sum_{i=0}^{I-1} E(i)=M$. We then have
	\begin{align}
		\Tr[(\mathbb{1}-M) \rho]\geq 0,
	\end{align}
	holds as $\mathbb{1}-M$ is PSD by definition.
	Pushing the trace over the subtraction and rearranging terms yields:
	\begin{align}
		\Tr[\mathbb{1} \rho]- 	\Tr[M \rho]\geq 0 \Leftrightarrow \Tr[M \rho]\leq 1.
	\end{align}
	This means that the induced probability distribution $p(i)= \Tr[E(i)\rho]$ is indeed sub-complete.
\end{proof}

To construct sub-complete probability distributions using operator circuits we now introduce \textit{noisy positive unital circuits} or \noisepuncs.

\begin{definition}
	\label{def:noisepunc}
	A \noisepunc $\Qcircuit(\xvars)$ is of the form
	\begin{align}
		\Qcircuit(\xvars)= \qcircuit(\xvars) \Ocircuit(\xvars),
	\end{align}
	where $\Ocircuit(\xvars)$ is a \punc and $\qcircuit(\xvars)$ is real-valued and belongs to the  $[0,1]$ interval for every $\xvars$.
\end{definition}

\begin{restatable}{proposition}{propnoisepunddist}
	\label{prop:noisepunddist}
	\noisepuncs induce sub-complete probability distributions.
\end{restatable}

\begin{proof}
	Let $\Xvars$ be a set of random variables and  $\Qcircuit(\xvars)= \qcircuit(\xvars)\Ocircuit(\xvars)$ with $\xvars\in \Omega(\Xvars)$ be a \noisepunc inducing a probability distribution
	\begin{align}
		\pi_\Xvars(\xvars) = \Tr [ \qcircuit(\xvars)\Ocircuit(\xvars) \rho].
	\end{align}
	As $\qcircuit(\xvars)$ is a scalar we can write
	\begin{align}
		\pi_\Xvars(\xvars) = \qcircuit(\xvars) \Tr [ \Ocircuit(\xvars) \rho] = \qcircuit(\xvars) p_X(\xvars),
	\end{align}
	where $p_X(\xvars)$ is the (complete) probability distribution induced by the set $\{ \Ocircuit(\xvars) \}_{\xvars \in \Omega(\Xvars)}$. As $0\leq q(\xvars)\leq 1$ (by definition) and $0\leq p_\Xvars(\xvars)\leq 1$ we have also that $0\leq \pi_\Xvars(\xvars)\leq 1$. This shows that each event $\xvars\in \Omega(\Xvars)$ has a positive probability.

	For the sub-completeness we observe that the sum of positive terms
	\begin{align}
		\sum_{\xvars \in \Omega(\Xvars)} p_\Xvars(\xvars)=1.
	\end{align}
	If we weigh each term in the sum with a scalar $q(\xvars) \in [0,1]$ we immediately conclude that $\sum_{\xvars \in \Omega(\Xvars)} \pi_\Xvars (\xvars)\leq$ and that we have indeed a sub-complete probability distribution.
\end{proof}

\subsection{Noise Variables}

All of this is a bit unnerving. How is it possible that we have probabilistic events $\xvars\in \Omega(\Xvars)$ that violate the second of Kolmogorov's probability axioms (by having sub-complete distributions)? The resolution to this problem lies in interpreting sub-complete distributions as joint distributions not only over the set of variables $\Xvars$ but an extra set of variables $\Yvars$ representing extra noise~\citep{wiseman2009quantum}
\begin{align}
	\pi_{\Xvars}(\xvars) = p_{\Xvars \Yvars}(\xvars, \yvars),
\end{align}
for which we indeed have
\begin{align}
	\sum_{\xvars\in \Omega(\Xvars)} \sum_{\yvars\in \Omega(\Yvars)}  p_{\Xvars \Yvars}(\xvars, \yvars) =1.
\end{align}
The problem with the random variables $\Yvars$ is that they are not accessible to the observer. In the sense that they are observable in principle but in practice not, \eg a defect in a sensor.\footnote{Note these practically unobservable variables should not be confused with the concept of (local) hidden variables, which lead to violations of Bell's inequality.}
However, the probability we are actually interested in is the one for $\Xvars$ given $\Yvars$
\begin{align}
	p_{\Xvars|\Yvars}(\xvars\mid \yvars )
	 & =
	\frac{
	p_{\Xvars \Yvars}(\xvars,  \yvars )
	}
	{
	\sum_{\xvars\in \Omega(\Xvars)}p_{\Xvars \Yvars}(\xvars, \yvars )
	}
	\nonumber
	\\
	 & = \frac{\pi_{\Xvars}(\xvars)}{\sum_{\xvars\in \Omega(\Xvars)}  \pi_{\Xvars}(\xvars)}
\end{align}
The question now became whether we can choose $\qcircuit(\xvars)$ such that the summation in the denominator is tractable, \ie can be performed in time polynomial in the number of random variables $\Xvars$.

In the probabilistic circuit literature it has been shown that the product of two circuits is representable as a single circuit within polytime and polyspace if the two circuits are compatible~\citep{khosravi2019tractable,vergari2021compositional}. Importantly, this single circuit then allows for summing out the variables $\xvars$.
As pointed out also by \citet{loconte2025sum}, all we need is that $\qcircuit(\xvars)$ and $\Ocircuit(\xvars)$ are both structured decomposable according to the same partition tree (or variable tree). Furthermore, we pick $\qcircuit(\xvars)$ to be a probabilistic circuit and $\Ocircuit(\xvars)$ to be a \punc.
For the sake of simplicity we construct the circuit $\qcircuit(\xvars)$ (with an underlying partition circuit) using
Hadamard products instead of Kronecker products.
\begin{align}
	{\qcircuit}_k(\xvars_k)=
	\begin{cases}
		f_{\xvar_k},
		 & \text{if $k$ leaf}
		\\
		W_k \times \Bigl( \qcircuit_{k_l}(\xvars_{k_l}) \odot  \qcircuit_{k_r} (\xvars_{k_r}) \Bigr)
		 & \text{else}.
	\end{cases}
	\nonumber
\end{align}

\subsection{\noisepuncs as Deep Operator Mixtures}

An interesting consequence of modelling noise in the system via a probabilistic circuit $\qcircuit(\xvars)$ is that it effectively results in a model that can be interpreted as a recursive mixture of positive operators. To see this, let us first write the matrix vector product in the computation units of $\qcircuit(\xvars)$ using explicit indices:
\begin{talign}
	\qcircuit_k
	= W_k \times (\qcircuit_{k_l} \odot \qcircuit_{k_r})
	\Leftrightarrow  \qcircuit_{ki} = \sum_j w_{kij}  \qcircuit_{k_l j}  \qcircuit_{k_r j}.
\end{talign}
Here $\qcircuit_{ki}$ denotes the $i$-th entry of the vector $\qcircuit_{k}$ and $w_{kij}$ denotes the elements of the matrix $W_k$.
We also dropped the explicit dependency on $\xvars_k$.

We now multiply each $\qcircuit_{ki}\in [0,1]$ with the corresponding operator $\Ocircuit_k$
\begin{talign}
	\qcircuit_{ki} \Ocircuit_k
	& = \left(\sum_j w_{kij}  \qcircuit_{k_l j}  \qcircuit_{k_r j} \right)  \qop_k( \Ocircuit_{k_l} \otimes \Ocircuit_{k_r} )
	\nonumber
	\\
	& =\sum_j w_{kij}  \underbrace{\qop_k(  \qcircuit_{k_l j} \Ocircuit_{k_l} \otimes  \qcircuit_{k_r j} \Ocircuit_{k_r} )}_{=\tildeQcircuit_{kj}}
	\label{eq:deepoperatirmix}
\end{talign}
This means that we have for each computation unit $k$ not one but multiple operators, indexed by $i$, and each of the  $i$ operators is a mixture of operators:
\begin{talign}
	\Qcircuit_{ki}(\xvars_k) =\sum_j w_{kij} \tildeQcircuit_{kj}(\xvars_k).
	\label{eq:operatormix}
\end{talign}
We illustrate the situation in Figure~\ref{fig:operatormixture}. Such mixtures of operators can be regarded as generalizing standard mixture models of probability distributions, which can be recovered by considering the special case of operators of dimension one -- in other words positive real-valued scalars. With this interpretation we can follow a similar reasoning to viewing probabilistic circuits as \puncs closed over diagonal matrices. But now we view \noisepuncs as \puncs closed over block-diagonal matrices.

\begin{figure}[t]
	\centering
	\input{tex_input/operatormix}
	\caption{
	Graphical representation of operator mixing as described in Equation~\ref{eq:operatormix}. The operators $\Qcircuit_{k0}$, $\Qcircuit_{k1}$, and $\Qcircuit_{k2}$ are all PSD matrices and constructed using the operators  $\tildeQcircuit_{k0}$, $\tildeQcircuit_{k1}$, and $\tildeQcircuit_{k2}$ using weighted sums. For $\Qcircuit_{k0}$ we also indicate the mixing weights $w_{k00}$, $w_{k01}$, and $w_{k02}$, which are positive real-valued scalars and satisfy $w_{k00}+ w_{k01}+w_{k02}=1$.
	}
	\label{fig:operatormixture}
\end{figure}

\section{Proofs for Section \ref{sec:nsdnmcircuit}}

\subsection{Proof for Theorem \ref{theo:properprobdpunc}}
\label{sec:proof:theo:properpropdpunc}

\theoproperprobdpunc*

\begin{proof}
	The proof is relatively straightforward as positivity is guaranteed by the fact that we perform computations on members of POVMs that all preserve positive semi-deifiniteness (Kronecker product in the product units and convex combination of quantum operations in the sum units). This guarantees that $\ocircuit(\xvars)$ is PSD and hence $ \Tr [\ocircuit(\xvars) \rho]>0$. The argument is similar for the completeness of $\Tr [\ocircuit(\xvars) \rho]$ as here we need to show that $\sum_{\xvars \in \Omega(\Xvars)} \ocircuit(\xvars) =\mathbb{1}$. For this we push down the summations to the corresponding leaves where we obtain unit matrices. The circuit then performs again computation with these unit matrices that are all unital, which guarantees indeed that
	\begin{align}
		\sum_{\xvars \in \Omega(\Xvars)} \Tr [\ocircuit(\xvars) \rho]
		 & =
		\Tr \left[
			\left(
			\sum_{\xvars \in \Omega(\Xvars)}\ocircuit(\xvars)
			\right)
			\rho
			\right]
		\\
		 & =
		\Tr \left[  \mathbb{1} \rho\right]
		\nonumber
		\\
		 & =1
		\nonumber
	\end{align}
\end{proof}

\subsection{Proof of Proposition \ref{prop:sdpuncsubsetdpunc}}
\label{sec:proof:prop:sdpuncsubsetdpunc}

\propsdpuncsubsetdpunc*

\begin{proof}
	Consider the expression of a sum unit in a \dpunc
	\begin{align}
		\sum_{j\in\inputs(k)} \weight_{kj} \qop_{kj} (\ocircuit_j(\xvars_j) ).
	\end{align}
	Writing also the operation that is performed in the product units we obtain:
	\begin{align}
		\sum_{j\in\inputs(k)} \weight_{kj} \qop_{kj} \Bigl(\ocircuit_{j_l}(\xvars_{j_l}) \otimes \ocircuit_{j_r}(\xvars_{j_r})   \Bigr).
	\end{align}
	Given that all product units decompose in the same way we can write this as
	\begin{align}
		\sum_{j\in\inputs(k)} \weight_{kj} \qop_{kj} \Bigl(\ocircuit_{j_l}(\xvars_{k_l}) \otimes \ocircuit_{j_r}(\xvars_{k_r})   \Bigr),
	\end{align}
	where the index on $\xvars_j$, $\xvars_{j_l}$, and $\xvars_{j_r}$ is now on $k$ and not on $j$. Given that the expression above is a completely positive map \citep{nielsen2001quantum} that maps positive semi-definite matrices to positive semi-definite matrices, we can write the expression using a single quantum operation:
	\begin{align}
		\qop_{k} \Bigl(\ocircuit_{k_l}(\xvars_{k_l}) \otimes \ocircuit_{k_r}(\xvars_{k_r})   \Bigr).
	\end{align}
	Following the convention from Section~\ref{sec:puncs} and using upper case letter for the circuit we have:
	\begin{align}
		\qop_{k} \Bigl(\Ocircuit_{k_l}(\xvars_{k_l}) \otimes \Ocircuit_{k_r}(\xvars_{k_r})   \Bigr),
	\end{align}
	which corresponds exactly to the computations performed in an \sdpunc and thereby concludes the proof.
\end{proof}

### 2025-06-02 202507_uai_puncs/submission/content.tex
added uai

\begin{abstract}
	By recursively nesting sums and products, probabilistic circuits have emerged in recent years as an attractive class of generative models as they enjoy, for instance, polytime marginalization of random variables.
	In this work we study these machine learning models using the framework of quantum information theory, leading to the introduction of \textit{positive unital circuits} (\puncs),
	which generalize circuit evaluations over positive real-valued probabilities to circuit evaluations over positive semi-definite matrices.
	As a consequence, \puncs strictly generalize probabilistic circuits as well as recently introduced circuit classes such as PSD circuits.
\end{abstract}

\section{Introduction}

Probabilistic circuits (PCs)~\citep{darwiche2003differential,poon2011sum} belong to an unusual class of probabilistic models: they are highly expressive but at the same time also tractable.
For instance, so-called decomposable probabilistic circuits~\citep{darwiche2001decomposable} encode probability distributions using nested sums and products over positive real-valued numbers and allow for the computation of marginals in time polynomial in the size of the circuit.
\citet{zhang2020relationship} noted that it is exactly this restriction to positive values that limits the expressive efficiency (or succinctness) of PCs~\citep{martens2014expressive,decolnet2021compilation}. In particular, the positivity constraint on the set of elements that PCs operate on prevents them from modelling negative correlations between variables.

Circuits that are incapable of modelling negative correlations, \ie circuits that can only combine probabilities in an additive fashion, are also called monotone circuits~\citep{shpilka2010arithmetic}.
This restricted expressiveness can be combatted by the use of so-called \textit{non-monotone} circuits, where subtractions are allowed as a third operation (besides sums and products). Interestingly, \citet{valiant1979negation} showed that a mere single subtraction can render non-monotone circuits exponentially more expressive than monotone circuits -- a result that has recently been refined for a subclass of decomposable circuits~\citep{loconte2025sum}.

As shown by \citet{harviainen2023inference} and \citet{agarwalprobabilistic}, non-monotone circuits do, however, introduce an important complication: if non-monotone circuits are not designed carefully, verifying whether a circuit encodes a valid probability distribution or not is an NP-hard problem. This does also render learning the parameters of a circuit practically infeasible.

Using the concept of \textit{positive operator valued measures} from quantum information theory, which encode random events as positive semi-definite matrices, we are able to devise non-monotone circuits that nonetheless encode proper (normalized) probability distributions by construction.
Our approach extends a line of recent works presented in the circuit literature ~\citep{sladek2023encoding,loconte2024subtractive,wangrelationship,loconte2025sum}. However, our work is the first that establishes this deep connection between concepts in quantum information theory and tractable probabilistic models.
Furthermore, the non-monotone circuits that we introduce generalize probabilistic circuits and PSD circuits~~\citep{sladek2023encoding,loconte2024subtractive,loconte2025sum}\footnote{PSD circuits were later on rebranded as sum of compatible squares circuits~\citep{loconte2025sum}}.

The remainder of the paper is structured as follows. We introduce in Section \ref{sec:qit} the necessary concepts from quantum information theory. In Section \ref{sec:puncs} we then use these concepts to construct tractable probability distributions using positive operator circuits. In Section \ref{sec:special_cases} we impose specific restrictions on the functional form of the computation units in \puncs and show how these restrictions lead to circuit classes known in the literature, \eg probabilistic circuits in Section \ref{sec:diagcircuits}.

In Section~\ref{sec:nsdnmcircuit} we then drop the so-called property of \textit{structured decomposability}, which has been imposed so far on all non-monotone circuit models. As such, we introduce the first circuit model that is non-monotone and only adheres to the weaker property of \textit{decomposability}. We discuss related work in Section~\ref{sec:related} and end the paper with concluding remarks in Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{A Primer on Quantum Information Theory}
\label{sec:qit}

A widely used and elegant framework to describe measurements of quantum systems is the so-called \textit{positive operator-valued measure} (POVM) formalism. While POVMs have physical interpretations in terms of quantum information and quantum statistics, we will only be interested in their mathematical properties as we use them to show that  circuits (defined in Section~\ref{sec:puncs}) form valid probability distributions.
We refer the reader to \citep{nielsen2001quantum} for an in-depth exposition on the topic, as well as quantum computing and quantum information theory in general.
\begin{definition}[Positive Semidefinite]
	A $\numbond {\times} \numbond$ Hermitian matrix $H$ is called positive semi-definite (PSD) if and only if $\forall \xvars {\in}  \mathbb {C} ^{\numbond}: \xvars^{*} H \xvars {\geq} 0$, where $\xvars^{*}$ denotes the conjugate transpose and $\mathbb {C}^{\numbond}$ the $\numbond$-dimensional space of complex numbers.
\end{definition}
\begin{definition}[{POVM~\citep[Page 90]{nielsen2001quantum}}]
	\label{def:povm}
	A positive operator-valued measure
	% with a finite number of elements acting on a finite-dimensional Hilbert space $\mathcal{H}$,
	is a set of PSD  matrices $\{E(i)\}_{i=0}^{\numevents-1}$ ($I$ being the number of possible measurement outcomes) that sum to the identity:
	\begin{talign}
		\sum_{i=0}^{\numevents-1} E(i) = \mathbb{1},
		\label{eq:povm_normalized}
	\end{talign}
\end{definition}
Before defining the probability of a specific $i$ occurring, we need the notion of a density matrix~\citep{neumann1927wahrscheinlich,landau1927dampfungsproblem}:
\begin{definition}[{Density Matrix~\citep[Page 102]{nielsen2001quantum}}]
	\label{def:density_matrix}
	A density matrix $\rho$ is a PSD matrix of trace one, \ie $\Tr [\rho]=1$.
\end{definition}
\begin{definition}[{Event Probability~\citep[Page 102]{nielsen2001quantum}}]
	\label{def:eventprob}
	Let $\rho$ be a density matrix and let $i$ denote an event with $E(i)$ being the corresponding element from the POVM. The probability of the event $i$ happening, \ie measuring the outcome $i$, is given by
	\begin{talign}
		p(i) = \Tr [ \rho E(i)]
		\label{eq:povm_prob}
	\end{talign}
\end{definition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{restatable}{proposition}{proppovmprob}
	\label{prop:povmprob}
	The expression in Equation~\ref{eq:povm_prob} defines a valid probability distribution.
\end{restatable}

\begin{proof}
	While this is a well-known result we were not able to identify a concise proof in the literature. We therefore provide one in Appendix~\ref{sec:proof:prop:povmprob}.
\end{proof}

Given that the $E(i)$'s completely describe the event $i$ such that its event probability can be computed, they represent the quantum state of a system. This quantum state (represented by a matrix) lives in a certain Hilbert space. The changes that a quantum state can undergo are then described by so-called \textit{quantum operations} acting on the Hilbert space. We can construct such operations using Kraus' theorem.

\begin{theorem}[Kraus' Theorem~\citep{kraus1983states}]
	Let $\mathcal{H}$  and $\mathcal {G}$ be Hilbert spaces of dimension $N$ and $M$ respectively, and $\qop$ be a quantum operation between $\mathcal{H}$  and $\mathcal {G}$. Then, there are matrices
	$\{ \kraus_j \}_{j=1}^{D}$ (with $D\leq NM$)
	mapping $\mathcal{H}$ to $\mathcal {G}$ such that for any state $E(i)$
	\begin{talign}
		\qop(E(i)) = \sum_{j=1}^{D} \kraus_j E(i) \kraus_j^*
		\label{eq:def:kraus}
	\end{talign}
	provided that $ \sum _{j} \kraus_{j}^{*} \kraus_{j}\leq \mathbb {1} $ (in the Loewner order sense).
\end{theorem}

\begin{proof}
	See \citep[Chapter 8]{nielsen2001quantum}
\end{proof}
The $\kraus_j$ matrices are usually referred to as Kraus operators.

% \todo{talk about Loewner order}

% \citep[Corollary 4.2]{baksalary1989some}

\section{Positive Unital Circuits}
\label{sec:puncs}

A popular subclass of probabilistic circuits are so-called structured decomposable  probabilistic circuits~\citep{darwiche2011sdd} that are also smooth~\citep{darwiche2001tractable}. The advantage of this circuit subclass is that they can be implemented in a rather straightforward fashion on modern AI accelerators, as demonstrated by \citet{peharz2019random,peharz2020einsum}.
For the sake of exposition, we will limit ourselves in a first instance to such circuits that adhere to structured decomposability and will generalize to (non-structured) decomposable circuits in Section~\ref{sec:nsdnmcircuit}. For a detailed account on these different circuit properties we refer the reader to~\citep{vergari2021compositional}.

\begin{figure}[t]
	\centering
	\input{tex_input/partition}

	\caption{
		Partition circuit over four binary variables $x_i$ with $i \in \{0,1,2,3\}$, which are given as inputs to the circuit. The internal nodes of the partition circuit correspond the computation units.}
	\label{fig:circuit}
\end{figure}

\citet{zuidberg2024probabilistic} introduced an abstraction for these smooth structured decomposable circuits in the form of partition trees.
We further refine this by introducing the concept of a \textit{partition circuit}.
We give such a circuit in Figure~\ref{fig:circuit}.
Note, the concept of a partition tree, and hence a partition circuit, is related to the concept of a variable tree~\citep{pipatsrisawat2008new}. However, partition circuits emphasize an interpretation as computation graphs, unlike variable trees.

\begin{definition}[Partition Circuit]
	\label{def:partition_circuit}
	A partition circuit over a set of variables is a parametrized computation graph taking the form of a binary tree. The partition circuit consists of two kinds of computation units:
	\textit{leaf} and \textit{internal} units (including a single \textit{root}).
	Units at the same distance from the root form a layer.
	Furthermore, let $\circuit_k$ denote the root unit or an internal unit. The unit $\circuit_k$ then receives its inputs from two units in the previous layer, which we denote by $\circuit_{k_l}$ and $\circuit_{k_r}$. Each computation unit is input to exactly one other unit, except the root unit, which is the input to no other unit.
\end{definition}

\subsection{Positive Operator Circuits}

Using the concept of partition circuits we construct positive operator circuits. Positive operator circuits can be thought of as generalizing circuit evaluations with probabilities to circuit evaluations with PSD matrices.
% Note, in the definition below we use $\Ocircuit_k$ instead of $\circuit_k$ to make this generalization explicit.
\begin{definition}[Positive Operator Circuit (Partition Circuit)]
	\label{def:poc}
	Let   $\xvars{=}\{\xvar_0,\dots ,\xvar_{\numvar{-}1}  \}$ be a set of discrete variables.
	We define an operator circuit as a partition circuit whose computation units take the following functional form:
	\begin{align}
		 & {\Ocircuit}_k(\xvars_k){=}
		\begin{cases}
			E_{\xvar_k}
			% A_{k} \times  e_{\xvar_k} \otimes  e^*_{\xvar_k} \times A_{k}^*,
			 & \text{if $k$ is leaf}
			\\
			\qop_k \Bigl(\Ocircuit_{k_l}(\xvars_{k_l}) \otimes \Ocircuit_{k_r}(\xvars_{k_r}) \Bigr)
			 & \text{else},
		\end{cases}
		\label{eq:poc:def}
	\end{align}
	where the  $E_{\xvar_k}$'s are quantum state matrices, and where the $\qop_k$ quantum operations.
\end{definition}
Note that using the Kronecker product between $\Ocircuit_{k_l}(\xvars_{k_l})$ and  $\Ocircuit_{k_r}(\xvars_{k_r})$ is a sensible choice as it describes the joint state of both subsystems.

\begin{restatable}{proposition}{proppocpsd}
	\label{prop:pocpsd}
	Positive operator circuits are PSD.
\end{restatable}

\begin{proof}
	We know that all the leaves carry PSD matrices as they describe quantum states. Passing these on recursively to the quantum operations in the internal units retains the positive semi-definiteness as the Kronecker product between two PSD matrices is again PSD.
\end{proof}

\subsection{Constructing a Probability Distribution}

In Section~\ref{sec:qit} we saw that we can construct a probability distribution using a density matrix $\rho$ and a positive operator-valued measure, with the latter being a set of PSD matrices (\cf Definition~\ref{def:povm}) that sum to the unit matrix. Using a positive operator circuit $\Ocircuit(\xvars)$ we indeed have a set of PSD matrices. Namely, one for each instantiation of the $\xvars$ variables. We now introduce \textit{positive unital (operator) circuits} (\puncs) for which also the summation to the unit matrix holds.

\begin{definition}
	\label{def:cp_unital}
	We call a quantum operation \textit{unital} if
	\begin{align}
		\qop_k(\mathbb{1}_{k_l} \otimes \mathbb{1}_{k_r})
		=\qop_k(\mathbb{1}_{k_l k_r})
		=  \mathbb{1}_k,
	\end{align}
	where $\mathbb{1}_{k}$, $\mathbb{1}_{k_l}$, $\mathbb{1}_{k_l k_r}$, and $\mathbb{1}_{k_r}$  denote unit matrices of appropriate size,
\end{definition}

\begin{restatable}{proposition}{propqopunital}
	\label{prop:qopunital}
	Unital quantum operations are valid in the sense that the inequality $\sum_j \kraus_{j}^* \kraus_{j} \leq \mathbb{1}$ holds for all unital quantum operations.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:qopunital}
\end{proof}

\begin{definition}
	We call a positive operator circuit \textit{unital} if the quantum operations $\qop_k$ are unital, and if the sets $\{ E_{\xvar_k} \}_{\xvar_k \in \Omega(\Xvar_k)}$ form a POVM for each $\Xvar_k$.
\end{definition}
% We will also refer to positive unital operator circuits as positive unital circuits, or \puncs.

\begin{restatable}{proposition}{proppuncPOVM}
	\label{prop:puncPOVM}
	Let $\Xvars$ denote a set of random variables with sample space $\Omega(\Xvars)$.
	Then the set $\{ \Ocircuit(\xvars)\}_{\xvars \in \Omega(\Xvars)}$ of positive unital circuits forms a POVM.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:puncPOVM}
\end{proof}

\begin{restatable}{theorem}{theopuncprobdist}
	\label{theo:puncprobdist}
	Let $\rho$ be a density matrix and $\Ocircuit(\xvars)$ a positive unital circuit. The function
	\begin{align}
		p_\Xvars(\xvars) = \Tr [\Ocircuit(\xvars) \rho]
		\label{eq:theo:prob_operator}
	\end{align}
	is a proper probability distribution over the random variables $\Xvars$ with sample space $\Omega(\Xvars)$.
\end{restatable}

\begin{proof}
	This follows from Propositions~\ref{prop:povmprob} and~\ref{prop:puncPOVM}
\end{proof}

One of the outstanding properties of probabilistic circuits is that they are tractable -- in the sense that they allow for polytime marginalization of random variables. Positive unital circuits retain this property.

\begin{proposition}
	\label{prop:efficientmarg_sdpunc}
	Positive unital circuits allow for tractable marginalization.
\end{proposition}

\begin{proof}
	(Sketch) The proof is rather straightforward and hinges on the fact that the quantum operations in the internal units are assumed to be computable in polytime and on the fact that the marginalization of a random variable is performed by pushing the sum to the corresponding leaf in the partition circuit. Analogous to the proof of Proposition~\ref{prop:puncPOVM}.
\end{proof}

\section{Special Cases}
\label{sec:special_cases}

We will now make certain structural assumptions on the matrices representing the quantum states and the functional form of the quantum operations $\qop$. By doing so, we obtain the PSD circuits introduced by~\citet{sladek2023encoding} and (structured decomposable) probabilistic circuits as described by \citet{peharz2020einsum} as special cases (Section~\ref{sec:purestate} and Section~\ref{sec:diagcircuits} respectively).

\subsection{Hadamard Product Units}

First, however, we note that our formulation of \puncs already encompasses canonical polyadic tensor decompositions~\citep{carroll1970analysis} -- a popular choice in the circuit literature \citep{shih2021hyperspns,loconte2025relationship} to merge partitions that uses the Hadamard product instead of the Kronecker product.

Specifically, we observe that the Hadamard product between two matrices $A$ and $B$ can be rewritten using a Kronecker product;
\begin{align}
	A \circ B = P (A \otimes B) P^*,
\end{align}
where $P$ is the semi-unitary partial permutation matrix selecting a principal sub-matrix \citep[Corollary 2]{visick2000quantitative}.
This also means that a quantum operation involving a Hadamard product can be rewritten using a Kronecker product:
\begin{talign}
	\qop (A \circ B )
	& = \sum_i K_i (A\circ B) K_i^*
	\nonumber
	\\
	& = \sum_i K_i P  (A\otimes B) P^* K_i^*
	\nonumber
	\\
	& =  \sum_i K_i'  (A\otimes B) K_i'^* = \qop'(A \otimes B)
\end{talign}
Note that, for $\qop'$ to be unital it suffices that $\sum_i K_i K_i^* = \mathbb{1}$ as $P$ is semi-unitary ($PP^* = \mathbb{1}$).

From the discussion above we conclude that we can safely limit the discussion to circuits with Kronecker products as circuits with Hadamard products follow as a special case.

\subsection{Pure Quantum States}
\label{sec:purestate}

As the matrices that represent quantum states are PSD, we can decompose them as follows using the spectral theorem:
\begin{talign}
	\Ocircuit = \sum_j \Vcircuit_j \otimes \Vcircuit_j^*,
\end{talign}
with the $\Vcircuit_j$'s denoting the eigenvectors.
As a special case we then have so-called pure states. That is quantum states constructed with a single eigenvector:
\begin{align}
	\Ocircuit = \Vcircuit \otimes \Vcircuit^*,
\end{align}
We will show now that by restricting \puncs to performing operations on pure quantum states gives us the special case of PSD circuits as introduced by \citet{sladek2023encoding}, which we define first using a partition circuit.
\begin{definition}
	\label{def:vpoc}
	Let   $\xvars{=}\{\xvar_0,\dots ,\xvar_{\numvar{-}1}  \}$ be a set of $\numvar$ discrete variables.
	A PSD circuit is a partition circuit whose computation units take the following functional form:
	\begin{align}
		{\Vcircuit}_k(\xvars_k){=}
		\begin{cases}
			U_{k} \times  e_{\xvar_k},
			 & \text{if $k$ leaf}
			\\
			U_{k} {\times}  \left( \Vcircuit_{k_l}  (\xvars_k)  \otimes   \Vcircuit_{k_r} (\xvars_k) \right),
			 & \text{else}
		\end{cases}
		\label{eq:vector_units}
	\end{align}
	where the $U_k$'s are semi-unitary matrices.
	The probability $p_\Xvars(\xvars)$ is computed via
	\begin{align}
		p_\Xvars(\xvars) = {\Vcircuit}^*_{root}(\xvars) \times  \rho \times {\Vcircuit}_{root}(\xvars),
	\end{align}
	where $\rho$ is a density matrix.
\end{definition}
Note that in the original formulation \citet{sladek2023encoding} used non-semi-unitary matrices. However, \citet{loconte2024faster} have recently shown that there is no loss in expressiveness with such a restriction.

To show that PSD circuits are a special case of \puncs we now impose the following restriction on the quantum operations $\qop_k$:
\begin{align}
	\qop_k (\Ocircuit_{k_l} \otimes \Ocircuit_{k_l} )
	 & =
	\kraus_{k} \left( \Ocircuit_{k_l}  \otimes  \Ocircuit_{k_r} \right) \kraus_{k}^*
	\label{eq:pureinternal}
\end{align}
That is, we limit the quantum operation to having  only a single pair of Kraus operators. For the quantum operation to be unital we need to have $\kraus_{k} \kraus_{k}^*=\mathbb{1}$. That is, $\kraus_{k}$ has to be semi-unitary,

Furthermore, we make the following choice in the leaves:
\begin{align}
	E_{\xvar_k} = K_{k} \left( e_{\xvar_k} \otimes e_{\xvar_k}^* \right) \kraus_{k}^*,
	\label{eq:pureleaf}
\end{align}
where the set  $\{ e_{\xvar_k} \}_{\xvar_k \in \Omega(\Xvar_k)}$ is a complete set of orthonormal basis vectors, and $\kraus_{k}$ is again semi-unitary.

We can show that this choice for $E_{\xvar_k}$ forms a POVM. Firstly, by observing that each $E_{\xvar_k}$ is PSD. Secondly, by verifying the completeness of the set of operators:
\begin{talign}
	\sum_{\xvar_k \in \Omega(\Xvar_k)} E_{\xvar_k}
	& = \sum_{\xvar_k \in \Omega(\Xvar_k)} K_{k} \left( e_{\xvar_k} \otimes e_{\xvar_k}^*  \right) K_{k}^*
	\nonumber
	\\
	& =
	K_{k} \left(  \sum_{\xvar_k \in \Omega(\Xvar_k)}  e_{\xvar_k} \otimes e_{\xvar_k}^*  \right) K_{k}^*
	\nonumber
	\\
	& =
	K_{k} \mathbb{1} K_{k}^*= \mathbb{1}
\end{talign}

\begin{definition}
	We call a positive unital circuit pure if Equation~\ref{eq:pureinternal} and Equation~\ref{eq:pureleaf} hold.
\end{definition}

\begin{restatable}{proposition}{propOveq}
	\label{prop:Oveq}
	For computation units of a pure positive unital circuit and a PSD circuit it holds that
	\begin{align}
		\forall k: \Ocircuit_k(\xvars_k) = \Vcircuit_{k}(\xvars_{k}) \otimes \Vcircuit^*_{k}(\xvars_{k}).
		\label{eq:def:opvec_equivalent}
	\end{align}
	given that $U_k=\kraus_k$
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:Oveq}
\end{proof}

\begin{corollary}
	Pure positive unital circuits perform operations on pure quantum states exclusively.
\end{corollary}

\begin{proof}
	This follows immediately from Proposition~\ref{prop:Oveq}.
\end{proof}

\begin{restatable}{proposition}{propvoequiprob}
	\label{prop:voequiprob}
	A PSD circuit and a pure \punc encode the same probability distribution if $U_k=\kraus_k$ for each unit~$k$.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:voequiprob}
\end{proof}

In this subsection we have shown that by making specific choices in the functional form of the leaves and the internal units of a positive unital circuit we recover the special case of PSD circuits and its variants \citep{loconte2024subtractive,loconte2025sum}. Furthermore, our analysis also provides the rather satisfying interpretation of PSD circuits as quantum circuits acting on pure states exclusively. While this connection has already been pointed out informally by \citet{wangrelationship}, we give a formal argument.

\subsection{Diagonal Positive Unital Circuits}
\label{sec:diagcircuits}

The oldest and most widely used class of tractable circuits fall into the model class of probabilistic circuits. These probabilistic circuits are well understood, and their properties have been mapped out comprehensively \citep{vergari2021compositional}. We now show how we retrieve probabilistic circuits as a special case from \puncs by imposing specific constraints on the functional form of the computation units of \puncs. Specifically, by ensuring that the computations performed in \puncs are closed over diagonal matrices. Before imposing diagonal closedness on \puncs, we start by giving a definition of probabilistic circuits in terms of a partition circuits.

\begin{definition}[Probabilistic Circuit (Partition Circuit)]
	\label{def:sdprobabilisticcircuit}
	Let   $\xvars = \{\xvar_0, \allowbreak \dots ,\allowbreak \xvar_{\numvar{-}1}  \}$ be a set of discrete variables.
	% with domains of size $\samplespacesize$.
	We define a probabilistic circuit as a partition circuit whose computation units take the following functional form:
	\begin{align}
		 & {\Pcircuit}_k(\xvars_k){=}
		\begin{cases}
			P_{\xvar_k}
			 & \text{if $k$ is leaf}
			\\
			W_k {\times} \Bigl(\Pcircuit_{k_l}(\xvars_{k_l}) {\otimes} \Pcircuit_{k_l}(\xvars_{k_l}) \Bigr)
			 & \text{else},
		\end{cases}
		\nonumber
	\end{align}
	where the  $\Pcircuit_{\xvar_k}$'s are real-valued positive vectors such that summing over $\xvar_k$ gives a vector with exclusively ones as entries ($\sum_{\xvar_k} P_{\xvar_k}= [ 1, \dots, 1 ]^T$), and where the $W_k$'s are row-normalized matrices with positive entries only, \ie $\forall k,i: \sum_j W_{kij}=1$, where the $i$ and $j$ indices index the matrix. Furthermore, the dimensions of the $W_k$'s, $\Pcircuit_{\xvar_k}$'s and $\Pcircuit_k(\xvars_k)$ are such that they match the matrix-vector products in the computation units.
\end{definition}

\begin{restatable}{proposition}{proplpcvalid}
	\label{prop:circuit_is_prob}
	Every entry of a vector $\Pcircuit_k(\xvars_k)$ forms a valid probability distribution.
\end{restatable}

\begin{proof}
	The proof is included for the sake of completeness in Appendix~\ref{app:circuit_is_prob} and follows a similar structure to those found in the literature, \eg \citep{peharz2015theoretical}.
\end{proof}

Next, we define \textit{diagonal} \puncs. For this, consider the following functional form of a quantum operation:
\begin{talign}
	\qop (\Ocircuit)
	=
	\sum_j J_j D_{j}\Ocircuit D^*_{j} J_j^*,
	\label{eq:diagonalinternal}
\end{talign}
where the  $D_{j}$'s are diagonal matrices such that $\Tr [D_{j}D^{*}_{j}]=1$.
The $J_j$ are sparse matrices that are zero everywhere but in the $j$-th row where all their entries are $1$. For instance, if we assume that $J_2$ is a $3$ by $3$ matrix it takes the following form:
$
	J_2 =
	\left(
	\begin{smallmatrix}
			0 & 0 & 0 \\
			1 & 1 & 1 \\
			0 & 0 & 0
		\end{smallmatrix}
	\right).
$
Together, $J_j$ and $D_j$ represent the Kraus operators $K_j {=} J_j D_j$.

Given that $\Ocircuit$ is PSD it is obvious that also $\qop(\Ocircuit)$ in Equation~\ref{eq:diagonalinternal} is PSD: the individual term of the sum on the right-hand side are all PSD and the sum of PSD matrices is again a PSD matrix. For showing that the quantum operation is also unital, we rewrite the expression as follows:
\begin{talign}
	\qop(\mathbb{1}) = \sum_j J_j \diagmat (w_j) J_j^*,
\end{talign}
where $w_j$ is a vector whose entries correspond to the diagonal elements of the matrix $D_jD_j^*$. Simply carrying out the matrix products results in:
\begin{talign}
	\qop(\mathbb{1}) = \sum_j H_j  \sum_i w_{ji},
\end{talign}
where $H_j$ is a square matrix that is zero everywhere but on the $j$-th entry of the diagonal where it is $1$. From the condition that $\Tr [D_{j}D^{*}_{j}]=1$ it immediately follows that $ \sum_i w_{ji}=1$. This leaves us with:
\begin{talign}
	\qop(\mathbb{1}) = \sum_j H_j  = \mathbb{1}.
\end{talign}

Furthermore, in the leaves we pick
\begin{align}
	E_{\xvar_k} = \Delta_{\xvar_k}  \Delta_{\xvar_k}^*,
	\label{eq:diagonalleaf}
\end{align}
such that the $\Delta_{\xvar_k}$'s are diagonal matrices and such that $\sum_{\xvar_k} \Delta_{\xvar_k} \Delta_{\xvar_k}^*=\mathbb{1}$.

\begin{definition}
	We call a positive unital circuit diagonal if Equation~\ref{eq:diagonalinternal} and Equation~\ref{eq:diagonalleaf} hold.
\end{definition}

\begin{restatable}{proposition}{propdiagpunc}
	\label{prop:diagpunc}
	All operators $\Ocircuit_k (\xvars_k)$ in a diagonal \punc can be represented as diagonal matrices.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:diagpunc}
\end{proof}

\begin{restatable}{proposition}{propPCPunciso}
	\label{prop:PCPunciso}
	Probabilistic circuits and diagonal \puncs are isomorphic.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:PCPunciso}
\end{proof}

This last proposition tells us that every (structured decomposable) probabilistic circuits can be represented as a diagonal \punc (and vice versa).

\subsection{Block-Diagonal \puncs}

In order to combat theoretical limitations of pure and diagonal \puncs, \citet{loconte2025sum} introduced a circuit class dubbed \textit{\textmu SOCS}. In Appendix~\ref{sec:noisyblockdiagonalpuncs} we show how this circuit class can be represented with \puncs over block-diagonal matrices. Interestingly, this allows us to discuss the concept of noise in quantum information theory.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Decomposable \puncs}
\label{sec:nsdnmcircuit}

Using the concept of partition circuits we were able to define the different circuits in terms of matrix-matrix or matrix-vector multiplications. This was, however, only possible because the circuits we have studied so far obey the property of \textit{structured decomposability}~\citep{pipatsrisawat2008new,darwiche2011sdd}.
We will now study circuits adhering to the weaker property of (non-structured) decomposability. To this end, we first define probabilistic circuits in the usual way using nested sum and product units \citep{vergari2021compositional} (and not partition circuits).

\begin{definition}[Probabilistic Circuit]
	\label{def:probabilisticcircuit}
	Let   $\xvars = \{\xvar_0, \allowbreak \dots ,\allowbreak \xvar_{\numvar{-}1}  \}$ be a set of discrete variables.
	A probabilistic circuit is a computation graph consisting of three kinds of computational units:
	\textit{leaf}, \textit{product}, and \textit{sum}.
	Each product or sum unit receives inputs from a set of input units denoted by $\inputs(k)$.
	Each unit $k$ encodes a function $\pcircuit_{k}(\xvars_k)$ with $\xvars_k {\subseteq} \xvars$ as follows:
	\begin{align*}
		{\pcircuit}_k(\xvars_k)=
		\begin{cases}
			f_k({\xvar_k})                                                & \text{if $k$ leaf unit}    \\
			\pcircuit_{k_l}(\xvars_{k_l})  \pcircuit_{k_r} (\xvars_{k_r}) & \text{if $k$ product unit} \\
			\sum_{j\in\inputs(k)} \weight_{kj} \pcircuit_j(\xvars_j)      & \text{if $k$ sum unit}
		\end{cases}
		% \label{eq:def:circuit}
	\end{align*}
	where $f_{\xvar_k}$  denotes a parametrized function such that $\sum_{\xvar_k} f_{\xvar_k}=1$ and where $\forall k: \sum_{j\in\inputs(k)} \weight_{kj} =1$.
\end{definition}

\begin{proposition}
	Let $\Xvars_k$ be a set of random variables with sample space $\Omega(\Xvars_k)$ equal to the possible values for $\xvars_k$.
	A probabilistic circuit $\pcircuit(\xvars_k)$ then defines a proper probability distribution as
	$
		p_{\Xvars_k}(\xvars_k) = \pcircuit_k (\xvars_k)
	$.
\end{proposition}

\begin{proof}
	We need to show that $\forall \xvars_k {\in} \Omega(\Xvars_k): p_{\Xvars_k}(\xvars_k){\geq} 0$ and that $\sum_{\xvars_k \in \Omega(\Xvars_k)} p_{\Xvars_k}(\xvars_k)=1$. Both of which are trivial.
\end{proof}

The sums and products that we write explicitly in Definition~\ref{def:probabilisticcircuit} are also implicitly present in Definition~\ref{def:sdprobabilisticcircuit}, namely within the matrix-vector product of the internal computation units of the underlying partition circuit.

\subsection{A Primer on Decomposability}

We can now define the concept of (non-structured) decomposability using the concept of a scope function.

\begin{definition}[Scope]
	\label{def:scope:cond}
	The scope of a unit $k$, denoted by $\scope(k)$, is the set of random variables $\Xvars_k$ for which the function $\pcircuit_k (\cdot)$ encodes a probability distribution.
\end{definition}

\begin{definition}[Decomposability]
	A circuit is decomposable if the inputs of every product unit
	$k$ encode distributions over disjoint sets of random variables:
	$\scope (k_l) \cap \scope (k_r) = \emptyset$ with $\{k_l, k_r\}= \inputs (k)$.
\end{definition}

It is precisely this decomposability property that enables tractable (any-order) marginalization in probabilistic circuits, as well as positive operator circuits and relaxing the property of decomposability leads necessarily to a decrease in tractability \citep{choi2020probabilistic,vergari2021compositional,zuidberg2024probabilistic}. Usually the smoothness property is also assumed to hold
\begin{definition}[Smoothness]
	A circuit is smooth if for every sum unit $k$ its inputs encode distributions over the same random variables:
	$\forall j_1, j_2 {\in} \inputs(k)$ it holds that $\scope(j_1){=}\scope(j_2) $.
\end{definition}
Note that for circuits constructed using a partition tree, the decomposability and smoothness properties hold by construction. We give a graphical representation of a decomposable circuit in Figure \ref{fig:dcircuit}.

\begin{figure}
	\centering

	\input{tex_input/dcircuit.tex}
	\caption{Graphical representation of a (non-structured) decomposable probabilistic circuit. We have four binary random variables in the leaves at the bottom. These are first passed individually through a set of sum units. In the second layer of the circuit we then have for blocks of computation units $\kappa_1$, $\kappa_2$, $\kappa_3$, and $\kappa_4$ (from lest to right). Each with their own scope ($\scope(\kappa_1)= \{ \Xvar_1, \Xvar_2  \}$, $\scope(\kappa_2)= \{ \Xvar_1, \Xvar_3  \}$, $\scope(\kappa_3)= \{ \Xvar_2, \Xvar_4  \}$, and $\scope(\kappa_4)= \{ \Xvar_3, \Xvar_4  \}$). At the root we have a block of computations $\kappa_{root}$ with $\scope(\kappa_{root}) = \{ \Xvar_1, \Xvar_2, \Xvar_3, \Xvar_4 \}$. Note that we slightly abuse notation and applied the scope function on sets of computation units instead of single computations units.
	}
	\label{fig:dcircuit}
\end{figure}

\begin{definition}[Structured Decomposability]
	A circuit is structured decomposable if the circuit is decomposable and product units with identical scope decompose identically: let $k$ and $j$ be two product nodes. If we have that $\scope (k_l) {=} \scope(j_l)$ and $\scope (k_r) {=} \scope(j_r)$ for every pair of product nodes where $\scope(k) {=} \scope(j)$, then we call a circuit structured decomposable.
\end{definition}

For partition circuits, we have again that the property of structured decomposability is respected by construction.
We give a structured decomposable circuit in Figure \ref{fig:sdcircuit}.

We can now compare the two circuits in Figure~\ref{fig:dcircuit} and Figure~\ref{fig:sdcircuit}. Specifically, we point to the computational blocks at the root of the circuits, where we have for each circuit four product units and a single sum unit. In both circuits the scope of all the product units (and the sum unit) is the same:

\begin{align}
	\scope(\pcircuit_{k}^{NSD}) & = {\Xvar_1,\Xvar_2,\Xvar_3,\Xvar_4 }
	\\
	\scope(\pcircuit_{k}^{SD})  & = {\Xvar_1,\Xvar_2,\Xvar_3,\Xvar_4 },
\end{align}
where $NSD$ stands for non-structured decomposable and $SD$ for structured decomposable. The index $k \in \{1,2,3,4 \}$ specifies in both cases the individual product units at the root (going from left to right). For the structured decomposable circuit in Figure~\ref{fig:sdcircuit} we have that the scope arises from the same union of random variables for all four units: $\scope(\pcircuit_{k}^{SD}) =  \{\Xvar_1,\Xvar_2 \} \cup \{\Xvar_3,\Xvar_4 \}$, for $k \in \{1,2,3,4 \}$.

The situation for the non-structured decomposable circuit presents itself differently. Here we have $\scope(\pcircuit_{k}^{SD}) =  \{\Xvar_1,\Xvar_2\} \cup \{\Xvar_3,\Xvar_4 \}$, for $k \in \{2,3\}$ and $\scope(\pcircuit_{k}^{NSD}) =  \{\Xvar_1,\Xvar_3\} \cup \{\Xvar_2,\Xvar_4 \}$, for $k \in \{1,4 \}$ -- showing that not all product units with the same scope decompose their respective scopes in the same fashion.

\citet{pipatsrisawat2008new} showed that dropping the requirement that the product nodes with the same scope decompose in the same way leads to exponential gains in expressiveness. In the next subsection we show how we can construct non-structured decomposable non-monotone circuits. This is in contrast to all the non-monotone circuits discussed in the previous sections and the non-monotone circuits discussed in the literature \citep{sladek2023encoding,loconte2025sum,loconte2024subtractive,wangrelationship}.

\subsection{Dropping Structured Decomposability}
\label{sec:nonstructdecomp}

\begin{definition}[Positive Unital Circuit]
	\label{def:dpunc}
	Let   $\xvars = \{\xvar_0, \allowbreak \dots ,\allowbreak \xvar_{\numvar{-}1}  \}$ be a set of discrete variables.
	A positive unital circuit is a computation graph consisting of three kinds of computational units:
	\textit{leaf}, \textit{product}, and \textit{sum}.
	Each product or sum unit receives inputs from a set of input units denoted by $\inputs(k)$.
	Each unit $k$ encodes a function $\ocircuit_{k}(\xvars_k)$ with $\xvars_k {\subseteq} \xvars$ as follows:
	\begin{align*}
		{\ocircuit}_k(\xvars_k)=
		\begin{cases}
			e_{\xvar_k}                                                           & \text{if $k$ leaf unit}    \\
			\ocircuit_{k_l}(\xvars_{k_l}) \otimes \ocircuit_{k_r}  (\xvars_{k_r}) & \text{if $k$ product unit} \\
			\sum_{j\in\inputs(k)} \weight_{kj} \qop_{kj} (\ocircuit_j(\xvars_j) ) & \text{if $k$ sum unit}
		\end{cases}
		% \label{eq:def:circuit}
	\end{align*}
	where $e_{\xvar_k}$  denotes an element of a POVM. The $\qop_{kj}$ are unital quantum operations and the $w_{kj}$ are positive real-valued scalars and obey $\forall k: \sum_j w_{kj} {=}1$.
\end{definition}

Note how the sum units form convex combinations of quantum operation, \ie quantum mixtures.
From now on we will denote (non-structured) decomposable \puncs by \dpuncs (\cf Definition~\ref{def:dpunc}) and structured decomposable \puncs by \sdpuncs (\cf Section \ref{sec:puncs}).

\begin{restatable}{theorem}{theoproperprobdpunc}
	\label{theo:properprobdpunc}

	Let $\Xvars_k$ be a set of random variables with sample space $\Omega(\Xvars_k)$ equal to the possible values for $\xvars_k$.
	A \dpunc $\ocircuit(\xvars_k)$ and a density matrix $\rho$ then define a proper probability distribution as
	$
		p_\Xvars(\xvars) = \Tr [\ocircuit(\xvars) \rho]
		\label{eq:theo:prob_operator_nsd}
	$.
\end{restatable}
\begin{proof}
	See Appendix~\ref{sec:proof:theo:properpropdpunc}
\end{proof}

\begin{restatable}{proposition}{propsdpuncsubsetdpunc}
	\label{prop:sdpuncsubsetdpunc}
	\sdpuncs are a proper subset of \dpuncs.
\end{restatable}

\begin{proof}
	See Appendix~\ref{sec:proof:prop:sdpuncsubsetdpunc}
\end{proof}

\begin{proposition}
	(Non-structured) decomposable probabilistic circuits are a proper subset of \dpuncs.
\end{proposition}

\begin{proof}
	This follows trivially from the observation that restricting the circuit elements $\ocircuit_k$ to  $1{\times}1$ dimensional PSD matrices we are left with positive real-valued scalars. In this case the Kronecker product becomes the usual product over the reals and quantum operations simplify to identity operations, and Definition~\ref{def:dpunc} is equivalent to that of a probabilistic circuit (Definition~\ref{def:probabilisticcircuit}).
\end{proof}

\begin{proposition}
	\dpuncs allow for tractable marginalization.
\end{proposition}

\begin{proof}
	The proof follows a similar rationale as the one for Proposition~\ref{prop:efficientmarg_sdpunc}
\end{proof}

\begin{figure}
	\centering

	\input{tex_input/sdcircuit.tex}
	\caption{
		Graphical representation of a structured decomposable probabilistic circuit. We have four binary random variables in the leaves at the bottom. These are first passed individually through a set of sum units. In the second layer of the circuit we then have two blocks of computation units $\kappa_1$, $\kappa_2$ with scopes ($\scope(\kappa_1)= \{ \Xvar_1, \Xvar_2  \}$ and $\scope(\kappa_2)= \{ \Xvar_3, \Xvar_4  \}$). At the root we have a block of computations $\kappa_{root}$ with $\scope(\kappa_{root}) = \{ \Xvar_1, \Xvar_2, \Xvar_3, \Xvar_4 \}$. One can easily see how such a circuit maps on to a partition circuits by associating each block of computation units to a unit in a partition circuit, \cf Figure~\ref{fig:circuit}.
	}
	\label{fig:sdcircuit}
\end{figure}

To the best of our knowledge, we present with \dpuncs the first non-monotone tractable circuit class that encodes (together with a denisty matrix) a positive function and that does not adhere to structured decomposability but only (non-structured) decomposability, regardless of the input.
This means that similarly to how non-monotone structured decomposable circuits, \eg \sdpuncs and the works of \citep{sladek2023encoding,loconte2025sum,wangrelationship}, generalize the language of (weighted) \textit{SDNNFs} \citep{pipatsrisawat2008new}, \dpuncs generalize the strictly larger language of \textit{DNNFs} \citep{darwiche2001decomposable}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Related Work}
\label{sec:related}
\puncs extend recent advancements in the probabilistic circuit literature (PSD circuits \citep{sladek2023encoding}, SOCS \citep{loconte2025sum}, and inception circuits \citep{wangrelationship}),
which in turn extend the traditionally monotone circuits to the non-monotone setting.\footnote{We refer the reader to \citep[Section 5]{loconte2025sum} for a discussion on the relationship between inception circuits and SOCS.}.
The main differentiator of \puncs with regard to these works is that we can use decomposability instead of structured decomposability.

Furthermore, positive unital circuits provide also a different perspective on constructing non-monotone circuits. While the methods described in ~\citep{sladek2023encoding,loconte2024subtractive,loconte2025sum,wangrelationship} regard such circuits as sum of (nested) squares, we interpret them as probabilistic events described by positive semi-definite matrices that are combined within a circuit using unit preserving quantum operations. As such, we also establish a strong link between quantum information theory and the circuit literature.

As already pointed out by \citet{loconte2024subtractive,loconte2025sum,loconte2025relationship} structured decomposable circuits share many aspects with \textit{tensor networks}~\citep{orus2014practical,white1992density} -- a class of statistical models developed in the condensed matter physics community, which have in recent years also been applied to supervised and unsupervised machine learning~\citep{cheng2019tree,han2018unsupervised,stoudenmire2016supervised}. In this regard, and given that tensor networks originate in the physics community, it is rather surprising that tensor networks have so far not been formulated using POVMs and quantum information theory.

First results on the expressive power of tensor networks and by extension of non-monotone circuits, were presented in the tensor network literature in the context of tensor decompositions \citep{glasser2019expressive} and complex-valued hidden Markov models \citep{gao2022enhancing}. Recently, the works of \citet{loconte2024subtractive,loconte2025sum} and \citet{wangrelationship} have studied the relationship of different circuit classes more carefully, as well. Additionally, \citet{loconte2025sum} pointed out links between tensor networks and the circuit literature and were able to generalize earlier results from the tensor network literature by~\citet{glasser2019expressive}.
As the circuits studied by \citet{loconte2025sum} and \citet{wangrelationship} are special cases of \dpuncs their results do also partially apply to \dpuncs.

Lastly, we point out theoretical results in the statistical relational AI literature. Specifically, \citet{buchman2017rules}, and \citet{kuzelka2020complex} noted that using only real-valued parametrizations (including negatives~\citep{buchman2017negative}), does not allow for fully expressive models.

\section{Conclusions \& Future Work}
\label{sec:conclusions}

Based on first principles from quantum information theory, we constructed positive unital circuits -- a novel class of probabilistic tractable models (Section~\ref{sec:puncs}).
In a first instance, we then showed how structured decomposable \puncs encompass all existing non-monotone circuit classes by enforcing certain constraints on the functional form of the quantum operations (Section ~\ref{sec:special_cases}).
Then we continued in Section~\ref{sec:nsdnmcircuit} with showing how the formalism of
positive unital circuits is effortlessly relaxed to (non-structured) decomposable \puncs -- thereby creating a new circuit class.

In future work we would like to investigate in detail the expressive power of \dpuncs compared to \sdpuncs and (non-structured) decomposable probabilistic circuits. Specifically, we make the following conjecture.

\begin{conjecture}
	There is an exponential separation in expressive efficiency between \dpuncs and \sdpuncs.
\end{conjecture}

Effectively, this would expand the research initiated by \citet{loconte2024subtractive} for mapping out the relationships between the different structured decomposable circuits. Ideally, one would like to create an analogue to \citeauthor{darwiche2002knowledge}' knowledge compilation map \citep{darwiche2002knowledge} but for positive operator circuits and relate it to the framework of algebraic model counting \citep{kimmig2017algebraic}.
This would also allow for establishing possible separations between sum-of-squares PCs~\citep{loconte2024subtractive}, product of monotonic by SOCS~\citep[Definition 5]{loconte2024subtractive}, and inception networks~\citep{wangrelationship}.
A first discussion on these issues can also be fund in \citep[Appendix A]{wangrelationship}.

A more audacious goal would then be to investigate whether \puncs can be run efficiently on quantum computers and whether there are speed-ups to be had. We can formulate this question more directly as a conjecture,

\begin{conjecture}
	There are circuit-query pairs that are tractable on a quantum computer but not on a classical computer.
\end{conjecture}

We stipulate that a positive answer to this question would involve Fourier transforms as they are the key mechanism underlying Shor's algorithm \citep{shor1994algorithms} leading to exponential quantum speed-up.
% -- the only known algorithm (modulo derivatives) providing an exponential quantum speed-up.
Note that this question seems also related to the work of \citet{riguzzi2024quantum}. Although the question in that work is on performing a computationally hard weighted model count using Grover's algorithm \citep{grover1996fast}, which does not provide an exponential speed-up.

Finally, a more practical research avenue consists of finding parametrizations of \puncs that sidestep the expensive dense matrix-matrix multiplications and replace them with structured sparse matrices. Monarch matrices seem to hold promise \citep{dao2022monarch,ZhangCoLoRAI25b}.

### 2025-07-05 202507_uai_puncs/arxiv/content.tex
y54wytr

--- 2025-06-02 202507_uai_puncs/submission/content.tex
@@
 We discuss related work in Section~\ref{sec:related} and end the paper with concluding remarks in Section~\ref{sec:conclusions}.
+
+% \begin{enumerate*}
+% 	\item We generalize probabilistic circuits over scalars to operator-valued circuits and introduce \puncs (Section~\ref{sec:puncs}).
+% 	\item We use the concept of noise in quantum information theory to construct deep operator mixture circuits, dubbed \noisepuncs (Section~\ref{sec:noisecircuits}).
+% 	\item We provide an efficient parametrization such that the constructed probability distributions using \puncs are normalized by construction (Section~\ref{sec:efficient_param}).
+% \end{enumerate*}
 
@@
 	\end{align}
-	where the  $E_{\xvar_k}$'s are quantum state matrices, and where the $\qop_k$ quantum operations.
+	where the  $E_{\xvar_k}$'s are quantum state matrices, and where the $\qop_k$ are quantum operations.
 \end{definition}
 Note that using the Kronecker product between $\Ocircuit_{k_l}(\xvars_{k_l})$ and  $\Ocircuit_{k_r}(\xvars_{k_r})$ is a sensible choice as it describes the joint state of both subsystems.
+
+% \begin{definition}
+% 	We call an operator circuit \textit{positive} if we have that the bilinear forms $\bilinear_k$ are such that
+% 	\begin{align}
+% 		\bilinear_k \Bigl(\Ocircuit_{k_l}(\xvars_{k_l}), \Ocircuit_{k_l}(\xvars_{k_l}) \Bigr) \quad \text{is a PSD matrix}
+% 		\label{def:eq:bilinear_psd}
+% 	\end{align}
+% 	if $\Ocircuit_{k_l}(\xvars_{k_l})$ and $\Ocircuit_{k_l}(\xvars_{k_l})$ are PSD.
+% \end{definition}
 
@@
 \begin{align}
-	\qop_k (\Ocircuit_{k_l} \otimes \Ocircuit_{k_l} )
+	\qop_k (\Ocircuit_{k_l} \otimes \Ocircuit_{k_r} )
 	 & =
@@
 
+% Lastly,  for the set of  basis vectors $\{e_{\xvar_s} \}_{s=0}^{\samplespacesize-1}$ in the leaves we choose the set of standard basis vectors, \ie one-hot unit vectors, with each basis vector $e_s$ corresponding to a random outcome.
+Although different choices of basis vectors would be valid as well, For the $A_k$ we use, as already mentioned, the orthonormal discrete Fourier transform matrix, which we truncate if $\numbond<\samplespacesize$.
+
+% In order to parametrize positive unitary circuits, we need to find a functional form of the bilinear forms such that Equation~\ref{def:eq:bilinear_psd} and Equation~\ref{def:eq:bilinear_unitpreserving} hold, and such that we have $A_k A_k^*=\mathbb{1}$.
+This means that we have to chose $A_k$ such that it is semi-unitary.
+
+% For $\gamma$ this is rather straightforward as we can simply use:
+% $
+% 	\gamma = \frac{\tilde{\gamma}}{\sqrt{\Tr [\tilde{\gamma} \tilde{\gamma}^*]}},
+% $
+% where $\tilde{\gamma}\in \mathbb{C}^{B\times B}$ is an arbitrary complex values matrix. Note that we have almost by definition $\Tr [\rho] = 1$.
+
+% \begin{restatable}{proposition}{propcomputationalcomplexity}
+% 	\label{prop:computationalcomplexity}
+% 	Let $\numvar$ be the number of variables $\xvars_k$,  having all the  domain size $\samplespacesize$, and let the $L_k$ and $R_k$ matrices be $\numbond\times \numbond$ square matrices, with $\numbond\leq \samplespacesize$.
+% 	We can perform marginal inference in \puncs in $\bigO(\numvar \samplespacesize^3)$ when using the bilinear form from Equation~\ref{eq:polyadicbilinear}.
+% \end{restatable}
+
+% \begin{proof}
+% 	See Append \ref{sec:proof:prop:computationalcomplexity}
+% \end{proof}
+
+% \begin{restatable}{proposition}{propcomputationalcomplexityvec}
+% 	\label{prop:computationalcomplexityvec}
+% 	\puncs in their vector representation compute joint probabilities in time $\bigO(\numvar \samplespacesize^2)$
+% \end{restatable}
+
+% \begin{proof}
+% 	See Section~\ref{sec:proof:prop:computationalcomplexityvec}
+% \end{proof}
+
+% Note that the vector representation can only be used for computing joint probabilities as marginalizing out variables forces us to go to the matrix representation of operator circuits.
+
+% Note also that the vector representation of \puncs can be seen as a PSD circuits~\cite{sladek2023encoding} or equivalently sum of squares circuits~\citep{loconte2025sum}, which we obtained by imposing a specific functional structure on the bilinear forms.
+
 \subsection{Diagonal Positive Unital Circuits}
@@
 			\\
-			W_k {\times} \Bigl(\Pcircuit_{k_l}(\xvars_{k_l}) {\otimes} \Pcircuit_{k_l}(\xvars_{k_l}) \Bigr)
+			W_k {\times} \Bigl(\Pcircuit_{k_l}(\xvars_{k_l}) {\otimes} \Pcircuit_{k_r}(\xvars_{k_r}) \Bigr)
 			 & \text{else},
@@
 
+% The major difference between these works and \puncs is that \puncs do not require the property of structured decomposability.
+% to hold but only the weaker (non-structured) decomposability property.
+
 Furthermore, positive unital circuits provide also a different perspective on constructing non-monotone circuits.
@@
 Specifically, \citet{buchman2017rules}, and \citet{kuzelka2020complex} noted that using only real-valued parametrizations (including negatives~\citep{buchman2017negative}), does not allow for fully expressive models.
+
+% From a practical perspective, the (sub-complete) distributions encoded with \puncs and \noisepuncs tackle one of the main shortcoming of sum of squares type circuits: we do not need to compute a normalization constant to ensure that event probabilities are bounded.
+This especially useful during training, as it means that updating the parameters of the model still guarantees the event probability to be bounded.
+Otherwise one would have to compute a normalization constant and differentiate through it at every iteration of the parameter updates.
+This potentially hinder scaling up non-monotone models as computing the normalization constant cannot be done in the vector representation (quadratic cost) but requires matrix-matrix multiplications with cubic cost.
+For \puncs computing this normalization is not necessary at all and for \noisepunc we only need to compute the normalization constant once training has finished as post-processing step.
+
+% \subsection{Tensor Networks}
+
+% Using our POVM formulation for \puncs is, however, not only theoretically elegant but has also practical benefits: using this formalism leads to circuits that are normalized by construction, and we can perform learning simply by optimizing the likelihood.
+We contrast this to the sweeping algorithms that are usually deployed in the tensor network literature, where blocks of variables are optimized one at the time while the remaining variables are held constant.
+It appears that this approach is inspired by the variational ansatz taken in the density-matrix renormalization group algorithm~\citep{white1992density}, which is the original algorithm for tensor networks.
+% Alternatives to this sweeping approach have also been proposed, such as an intricate Riemannian gradient descent optimization scheme that conserves unitarity of matrices, but are rather costly to run.
+
+% We would also like to point out a theoretical result from the tensor network literature stating that picking a complex-valued parametrization instead of a real-valued one can lead to an arbitrarily large reduction in the number of parameters~\citep{glasser2019expressive}.
+While this result is formulated with respect to (exact) low rank tensor decomposition and does not apply directly to the problem of learning parameters via gradient descent, we consider this observation to be a strong theoretical indicator for the superiority of complex numbers over real numbers when parametrizing probabilistic circuits.
+A similar argument has also been made by \citet{gao2022enhancing} in the context of hidden Markov models.
+
+% \subsection{Theoretical Studies}
+
+% \section{Experimental Evaluation}
+% \label{sec:experiments}
+
+% \textbf{Experimental Setup}
+
+% For our experimental evaluation we used the MNIST family of datasets.
+That is, the original MNIST~\citep{deng2012mnist}, FashionMNIST~\citep{xiao2017fashion}, and also EMNIST~\citep{cohen2017emnist}.
+% For the implementation we used PyTorch together with the Einops library~\citep{rogozhnikov2022einops}.
+We set up our experiments in Lightning\footnote{\url{https://lightning.ai/}} and ran them on a DGX-2 machine with V100 Nvidia cards.
+
+% We compare different methods using the \textit{bits per dimension} metric, which is calculated from the average negative log-likelihood ($\overline{NLL}$) as follows: $bpd {=} \nicefrac{\overline{NLL}}{(\log{2} \times D)}$ ($D{=}28^2$ for MNIST datasets).
+
+% \textbf{Training Positive Operator Circuits}
+
+% In our experiments we estimate the density of the different MNIST datasets by maximizing a parametrized likelihood function. We construct this  function using a positive vector circuit.
+% In order to build the underlying partition circuit,
+% we follow the approach described by \citet{zuidberg2024probabilistic}: we start with a grid of $28\times 28$ pixels and merge, in an alternating fashion, rows and columns of the image.
+% We also allow for merging three partitions into  a single partition, thereby breaking the binary character of the partition trees used so far.
+% Having three instead of two partitions being merged can easily be accommodated for by using a Hadamard product with three factors instead of two in the internal units of a \pvc.
+Ultimately, this allows us to handle layers with an uneven number of partitions.
+% At the leaves we encode pixel values by associating each value to one of $256$ standard basis vector of $\mathbb{R}^{256}$.
+
+% We performed training by minimizing the negative log-likelihood using Adam~\citep{kingma2014adam} with default hyperparameters, batch size of $50$, over $100$ epochs.
+% The best model was selected using a $90-10$ train-validation data split where we monitored the negative log-likelihood on the validation set.
+% We did not perform any hyperparameter search.
+% Further details can be found in the configuration files of the experiments.
+
+% \textbf{State-of-the-Art Baselines}
+
+% We compare \pvcs to three state-of-the-art tractable density estimators from the literature: quadrature of probabilistic circuits (QPCs)~\citep{gala2024probabilistic}, hidden Chow-Liu trees (HCLTs)~\citep{liu2021tractable}, and sparse HCLTs (SHCLTs)~\citep{dang2022sparse}.
+% All three methods are probabilistic circuits and they all use hidden Chow-Liu trees as their underlying structure.
+This tree is learned and dataset specific.
+The difference between QPCs and HCLTs is that the former is obtained by approximating a continuous latent variable extension of probabilistic circuits using numerical quadrature.
+The difference between HCLTs and SHCLTs is that the latter allows for dynamically adapting the number of components per partition during parameter learning.
+This means that SHCLTs can focus their computational budget on information-rich parts of the circuit.
+
+% \textbf{Q1: How Do the Different \pvc Parametrizations Fair Against Each Other?}
+
+% We construct \pvcs following the prescription of \cite{zuidberg2024probabilistic}, as described above. Furthermore, we use $B=128$ components per partition.
+% As for the parametrization we study four different variants:
+% $\mathrm{PVX}_{0}^{FR}$,
+% $\mathrm{PVX}_{2\pi}^{FR}$,
+% $\mathrm{PVX}_{0}^{R1}$, and
+% $\mathrm{PVX}_{2\pi}^{R1}$.
+% The subscripts ($0$ or $2\pi$) indicate whether we limit the phase to be zero or whether we allow the phase to be a learnable parameter.
+Note that having a tunable phase doubles the number of (real-valued) parameters.
+The superscript indicates whether the parametrization uses a full-rank ($FR$) density matrix in the root or a rank-one ($R1$) density matrix.
+% As such, $\mathrm{PVX}_{0}^{R1}$ is equivalent to the  squared monotonic probabilistic circuits (\smpcs) used in ~\cite{loconte2024subtractive}, and $\mathrm{PVX}_{2\pi}^{R1}$ generalizes their \snpcs from the real domain to the entire complex domain.
+Concretely, \smpcs are rank-one \pvcs with the phase fixed to zero and \snpcs are rank-one \pvcs with the phase fixed to values in the set $\{0, \pi \}$.
+We do not experiment on the latter.
+
+% We report the obtained bpds for the different \pvc parametrizations in the first four columns of Table~\ref{tab:mnist}.
+In general, we see that the complex-valued parametrizations are either on par or outperform the corresponding zero-phase parametrizations.
+% The notable exception is the FashionMNIST benchmark where the two parametrizations with no phase ($\mathrm{PVX}_{0}^{FR}$ and $\mathrm{PVX}_{0}^{R1}$) perform best.
+% Comparing full-rank to rank-one parametrizations we see a more important impact for the zero-phase parametrization than for the complex-valued parametrization.
+
+% Overall, $\mathrm{PVX}_{2\pi}^{FR}$ constitutes the best performing parametrization being best-in-class on all benchmarks but FashionMNIST.
+
+% \input{tex_input/mnist_table}
+
+% \textbf{Q2: How Do \pvcs Fair Against the State of the Art?}
+
+% In order to compare \pvcs to state-of-the-art circuits all methods (\pvc, QPC, HCLT, SHCLT) were given a computational budget of $B=128$ components per partition (on average for SHCLT).
+% The results for the competing methods were taken from the respective papers -- except for HCLTs.
+For HCLTs we took the bpds reported by \citet{gala2024probabilistic} as they achieved stronger results with their implementation compared to the originally reported performance~\citep{liu2021tractable}.
+
+% We see in Table~\ref{tab:mnist} that QPCs and HCLTs are in general outperformed by \pvc -- in particular $\mathrm{PVX}_{2\pi}^{FR}$.
+The FashionMNIST benchmark constitutes again an outlier in this regard.
+% The only methods outperforming \pvcs are the SHCLTs, which is due to them being able to dynamically allocate compute budget to informative parts of the circuit.
+% It is noteworthy that for the EMNIST-BYCLASS benchmark all four  \pvc parametrization exhibit strong performance with  $\mathrm{PVX}_{2\pi}^{FR}$ and $\mathrm{PVX}_{2\pi}^{R1}$ even outcompeting SHCLTs.
+
+% \textbf{Discussion}
+
+% In our experimental evaluation we did not perform an explicit comparison to the method of \citet{loconte2024subtractive} as they were not able to find any improvements of allowing non-negative parameters in probabilistic circuit for discrete data.
+They stipulated that
+% \say{simple categorical distribution can already capture any discrete distribution with finite support and a (subtractive) mixture thereof might not yield additional benefits}.
+% In our experimental evaluation we refute this conjecture, and show that using complex-valued parameters leads to noticeable gains when performing density estimation on discrete data.
+% The exception being of course the FashionMNIST benchmark. We believe that this is due to our optimization method that was not tuned in any way.
+% In this regard we believe that developing tailor-made optimization algorithms for \pvcs might further boost their performance.
+% Furthermore, one can envisage, similar to SHCLTs, dynamically adapting the compute budget per partition, which should again improve the density estimation capacities of \pvcs.
 
@@
 This would also allow for establishing possible separations between sum-of-squares PCs~\citep{loconte2024subtractive}, product of monotonic by SOCS~\citep[Definition 5]{loconte2024subtractive}, and inception networks~\citep{wangrelationship}.
-A first discussion on these issues can also be fund in \citep[Appendix A]{wangrelationship}.
+A first discussion on these issues can also be found in \citep[Appendix A]{wangrelationship}.
+
+% The big challenge to overcome in this regard will be to generalize algebraic model counting with commutative semirings (\eg using scalar probabilities) to non-commutative semirings with PSD matrices as their elements.
+% This non-commutativity is inherent to quantum information theory as changing the order in which quantities are measured changes the measurement outcome.
+This is also the reason why sampling in monotone circuits can be parallelized but for non-monotone circuit one needs to sample in an autoregressive fashion \citep{loconte2024subtractive} (at least on a classical computer).
 
@@
 Monarch matrices seem to hold promise \citep{dao2022monarch,ZhangCoLoRAI25b}.
+
+% A main hurdle to overcome to make \puncs practical machine learning models is to find expressive yet efficient functional forms of the bilinear forms and their parametrizations.
+A key challenge to overcome in this respect is the efficient parametrization of (expressive) unitary matrices as this is a notoriously difficult feat~\citep{kiani2022projunn,jing2017tunable,lezcano2019cheap,mhammedi2017efficient,wisdom2016full}.
+% Furthermore, we have only presented a specific functional form for the bilinear forms.
+
+% \todo{cite Chiara Marletto?}
+
+% Their construction as partition circuits allows for layer-wise parallelization and execution on modern machine learning hardware.
+% Using unconstrained gradient-descent we showed that \pocs, and \pvcs specifically, constitute a promising addition to the zoo of tractable probabilistic models.
+
+% In future work we would like to investigate in more detail theoretical properties of \pocs and how they differ from other tractable models.
+Ideally one would find an exponential separation between real-valued and complex-valued circuits.
+On the practical side it remains, however, to be seen whether such a separation has an equally important impact on real-world datasets.
+
+% This relates to another open question,  that of learning in \pocs.
+We presented a rather simple learning approach for parameters tailored towards neural architecture.
+It might be the case that more sophisticated techniques have to be deployed for large-scale \pocs, as they might exhibit the problem of barren plateaus~\citep{ragone2023unified} -- a well known issue in quantum machine learning~\citep{biamonte2017quantum,huggins2019towards}.
+% Furthermore, and as already pointed out by~\citet{loconte2024subtractive}, computing the normalization constant $Z$ requires the evaluation of the circuit in the \poc representation.
+This gives rise to a rather expensive cubic computation cost during learning (when compared to the quadratic cost of \pvc evaluations).
+Avoiding this issue would allow to drastically scale \pvcs.
+In order, to scale \pvcs it might also be necessary to use more sophisticated structures than binary trees with every partition having the same number of components $B$ per partition, \cf SHCLTs.
